### 2013-10-15 doc/proposal/abstract.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% abstract.tex
%%---------------------------------------------------------------------------%%

\hrule
\vskip10pt
\noindent
{\bf\large Abstract}.

The next generation of computational science applications will require
numerical solvers that are capable of high performance on proposed exascale
platforms.  In order to meet this goal, solvers must be resilient to soft and
hard system failures, provide high concurrency on heterogeneous hardware
configurations, and retain numerical accuracy and efficiency.  In light of
these requirements, a natural avenue of inquiry would be to adapt the current
stable of numerically efficient solvers to this new high-performance computing
regime.  However, an alternative approach would be to investigate different
classes of algorithms that can address issues of resiliency, particularly
fault tolerance and hard processor failures, naturally.  In this proposal we
will investigate new stochastic methods for solving linear systems, otherwise
termed Monte Carlo Resilient, Exascale (MCREX) solvers.  The family of methods
that we have proposed builds on the sequential Monte Carlo work of Halton,
1962. While showing significant promise, this class of solvers has not made
inroads into the broader computational science community.  The methods that we
have initially developed use Monte Carlo to accelerate a fixed-point
iteration; therefore, we have called them Monte Carlo Synthetic Acceleration
(MCSA). Preliminary work using MCSA has demonstrated that they are at least as
efficient as Jacobi-preconditioned Conjugate Gradient (PCG) on sparse, SPD
systems.  These initial results demonstrate that, because MCSA does not
require symmetry or positive definiteness, very good efficiency could be
attained on non-symmetric systems, thus making MCSA an ideal solver in
non-linear Newton schemes.  Furthermore, Monte Carlo methods have the benefit
of addressing resiliency in a natural way; soft errors can be treated as high
variance samples and lost histories from processor failures can be easily
discarded without affecting the quality of the solution.

%%---------------------------------------------------------------------------%%
%% end of abstract.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/bios/benzi.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% benzi.tex
%%---------------------------------------------------------------------------%%

\subsection{Biographical Sketch for Michele Benzi}

% This section is not called out in the RFP, so it could be omitted if
% you are tight on space.
%\vspace*{-1ex}
\subsubsection*{Address}
\begin{tabular}{l}
  Department of Mathematics and Computer Science \\
  Emory University \\
  400 Dowman Drive \\
  Atlanta, GA 30322 \\
  Email: {\tt benzi@mathcs.emory.edu}
\end{tabular}

%%---------------------------------------------------------------------------%%

\vspace*{-1ex}
\subsubsection*{Education and Training}
\vspace*{-1ex}
\begin{tabbing}
  \hspace*{1ex}
  xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxx \kill
  University of Bologna, Italy \> Mathematics \>  B.S., 1987 \\
  North Carolina State University \> Mathematics \> M.S., 1991 \\
  North Carolina State University \> Applied Mathematics \> Ph.D., 1993
\end{tabbing}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Research and Professional Experience}
\vspace*{-1ex}
\begin{tabbing}
  \hspace*{1ex} 
  %% 
  \= 2012 -- present \hspace*{2ex} \= Samuel Candler Dobbs Professor, Emory
  University\\
  %%
  \> 2006 -- 2012 \> Professor, Emory University\\ 
  %% 
  \> 2000 -- 2006 \> Associate Professor, Emory University \\
  %%
  \> 1998 -- 2000 \> Staff member, Scientific Computing Group, LANL \\ 
  %%
  \> 1997 -- 1998 \> Director-funded postdoctoral associate, Scientific 
  Computing Group, LANL\\
  %%
  \> 1996 -- 1997 \> Postdoctoral Researcher, Parallel Algorithms Group,
  CERFACS, Toulouse, France\\
  %%
  \> 1993 -- 1996 \>  Researcher, Department of Mathematics, University of Bologna,
  Italy
\end{tabbing}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Related Publications}
\begin{enumerate}
\parskip = -2pt

\item M.~Benzi and V.~Kuhlemann, ``Restricted additive Schwarz methods
  for Markov chains'', Numer.~Linear Algebra Appl., {\bf 18} (2011), 
  pp.~1011--1029.

\item M.~Benzi, M.~K.~Ng, W.~Niu and Z.~Wang, ``A relaxed dimensional
  factorization preconditioner for the Navier--Stokes equations'', 
  J.~Comp.~Phys., {\bf 230} (2011), pp.~6185--6202.

\item S.~P.~Hamilton, M.~Benzi and E.~Haber, ``New multigrid
  smoothers for the Oseen Problem'', Numer.~Linear Algebra Appl., {\bf 17}
  (2010), pp.~557--576.

\item S.~P.~Hamilton, M.~Benzi and J.~S.~Warsa, ``Negative-flux fixups in 
  discontinuous finite element $S_N$ transport'', 
  Proceedings of the International Conference on Mathematics, 
  Computational Methods \& Reactor Physics 2009 (M\&C 2009), 
  American Nuclear Society, Vol. 4 (2009), pp. 2529-2538, 
  Saratoga Springs, NY, 2009.

\item M.~Noskov, M.~Benzi, and M.~D.~Smooke, ``An implicit compact 
  scheme solver for two-dimensional multicomponent flows'', Computers 
  \& Fluids, {\bf 36} (2007), pp.~376--397.

\item M.~Benzi, G.~H.~Golub, and J.~Liesen, ``Numerical solution of 
  saddle point problems'', Acta Numerica, {\bf 14} (2005), pp.~1--137.

\item J.~Warsa, M.~Benzi, T.~Wareing and J.~Morel, ``Preconditioning a 
  mixed discontinuous finite element method for radiation diffusion'', 
  Numer.~Linear Algebra Appl., {\bf 11} (2004), pp.~795--811.

\item M.~Benzi, ``Preconditioning techniques for large linear systems:
  a survey'', J.~Comp.~Phys., {\bf 182} (2002), pp.~418--477.

\item M.~Benzi and M.~Tuma, 
  ``A parallel solver for large-scale Markov chains'',
  Appl.~Numer.~Math., {\bf 41} (2002), pp.~135--153.

\item M.~Benzi, J.~Marin and M.~Tuma, ``A two-level parallel preconditioner
  based on sparse approximate inverses'', in Iterative Methods in Scientific 
  Computation IV, D. R. Kincaid and A. C. Elster, eds., IMACS Series in 
  Computational and Applied Mathematics, Vol.~5, IMACS, New Brunswick, 
  NJ (1999), pp.~167--178.

\end{enumerate}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Synergistic Activities}
\vspace{-1ex}
%
\begin{enumerate}
  \parskip = -2pt
\item Chair, SIAM Activity Group on Linear Algebra (2010-present); 
\item Member of the SIAM Council (2009-present);
\item SIAM Fellow (Class of 2012); 
\item Program Committeee Co-chair, 2012 SIAM Applied Linear Conference
  (Valencia, Spain, June 18-22, 2012);
\item Program Committeee Co-chair, SIAM Annual Meeting (Minneapolis, MN, July
  9-13, 2012).
\item Editor-in-Charge, SISC Copper Mountain Special Issue (2012-present).
\end{enumerate}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Co-Authors and Collaborators (last 48 months)}
\vspace{-1ex}

{\parindent 0.2in \narrower 
E.~Agichtein (Emory), Z.-Z.~Bai (CAS, Beijing), 
P.~Boito (Limoges), M.~Challacombe (LANL), F.~Chen (Beijing),
E.~Estrada (Strathclyde), L.~Ferragut (Saragoza), 
X.-P.~Guo (Shanghai), E.~Haber (UBC, Canada)
S.~Hamilton (ORNL), N.~Hatano (Tokyo), M.~K.~Ng (Hong Kong), 
Q.~Niu (Zuhai, PRC), M.~Pennacchio (Pavia), M.~Olshanskii (Moscow), 
N.~Razouk (Emory), L.~Rebholz (Clemson), V.~Simoncini 
(Bologna), Y.~Wang (Emory), Z.~Wang (ORNL), Z.-Q.~Wang (Shanghai),
J.~Warsa (LANL).
}

%%\vspace*{1ex}
\subsubsection*{Graduate and Postdoctoral Advisors and Advisees}
% List all advisors and advisees for the past 5 years
\vspace*{-1ex}
%%{\parindent 0.2in \narrower 
Advisors:
\begin{enumerate}
\item Wayne Joubert (Postdoc, LANL)
\item Iain Duff (Postdoc, CERFACS)
\item Carl Meyer (PhD, NCSU)
\end{enumerate} 
\vspace*{-1ex}
Advisees: 
\begin{enumerate}
\item Paola Boito (Postdoc) 
\item Bora Ucar (Postdoc) 
\item Jia Liu (PhD)
\item Nader Razouk (PhD)
\item Lauren Taralli (PhD)
\item Steven Hamilton (PhD)
\item Zhen Wang (PhD)
\item Christine Klymko (PhD)
\item Verena Kuhlemann (PhD)
\item Cheng-yi Zhang (PhD)
\end{enumerate}

\pagebreak
\endinput

%%---------------------------------------------------------------------------%%
%% end of benzi.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/bios/engelmann.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% engelmann.tex
%%---------------------------------------------------------------------------%%

\subsection{Biographical sketch for Christian Engelmann}

\begin{tabular*}{\textwidth}
{@{}l@{\extracolsep{\fill}}r@{}}
Christian Engelmann                       &
(865) 574-3132                            \\
Computer Science Research Group           &
engelmannc@ornl.gov                       \\
Computer Science and Mathematics Division &
www.christian-engelmann.info              \\
Oak Ridge National Laboratory             &
P.O. Box 2008, Oak Ridge, TN 37831-6173   \\
\end{tabular*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Accomplishments
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Professional Accomplishments}
\vspace*{-1ex}

\begin{tabular*}{\textwidth}
{@{}r@{~}
    l@{~~~}
    r@{~}
    l@{~~~}
    r@{~}
    l@{}}
8  & Research grants (2 as Lead)      &
6  & Peer-reviewed journal articles   &
35 & Invited talks and seminars       \\
8  & Co-advised M.Sc. theses          &
33 & Peer-reviewed conference papers  &
70 & Committees at 34 conferences     \\
2  & Mentored summer faculty          &
27 & Peer-reviewed workshop papers    &
20 & Article \& book proposal reviews \\
10 & Direct reports over 7 years      &
8  & Peer-reviewed conference posters &
11 & Conference booth exhibitions     \\
   & Erd\"{o}s number of 5            &
   & 860+ Total publication citations &
   & H-index of 16 / G-index of 27    \\
\end{tabular*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Experience
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Professional Experience}
\vspace*{-1ex}

\begin{compactitem}[0000--Present :]
  \item[2009--Present\hfill :] \textbf{R\&D Staff,
                                       Oak Ridge National Laboratory }
                            \begin{compactitem}
                            \item HPC hard-/software co-design:
                                  performance/resilience/power modeling \&
                                  simulation
                            \item Soft-error injection for vulnerability
                                  analysis of scientific applications and
                                  CMOS logic
                            \item HPC resiliency system software for monitoring,
                                  fault prediction, and fault avoidance
                            \item HPC checkpoint storage virtualization and
                                  MPI-level redundancy for HPC applications
                            \item HPC hard-/software co-design:
                                  performance modeling \& simulation
                            \item Light-weight simulation of extreme scale HPC
                                  architectures ($\sim$100,000,000 cores)
                            \end{compactitem}
  \item[2004--2009\hfill :] \textbf{Associate R\&D Staff,
                                    Oak Ridge National Laboratory }
                            \begin{compactitem}
                            \item Fault tolerance support for MPI: Scalable
                                  membership, job pause, and process migration
                            \item 99.9997\% high availability for HPC
                                  head/service-node services: Torque and PVFS MDS
                            \item Ph.D. thesis research: Symmetric active/active
                                  high availability for HPC system services
                            \item Virtual system environments for
                                  ``plug-and-play'' supercomputing with HPC
                                  hypervisors
                            \item Enhancing application development productivity
                                  via a common view across platforms
                            \end{compactitem}
  \item[2001--2004\hfill :] \textbf{Post-Master's Research Associate,
                                    Oak Ridge National Laboratory }
                            \begin{compactitem}
                            \item Harness Distributed Virtual Machine: Pluggable, 
                                  lightweight, adaptive, and fault tolerant
                            \item Light-weight simulation of HPC architectures
                                  at large scale ($\sim$1,000,000 cores)
                            \end{compactitem}
  \item[2000--2001\hfill :] \textbf{Software Developer,
                                    Oak Ridge National Laboratory }
                            \begin{compactitem}
                            \item M.Sc. thesis research: Distributed peer-to-peer 
                                  control for Harness (a resilient runtime)
                            \end{compactitem}
\end{compactitem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Education
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Education}
\vspace*{-1ex}

\begin{compactitem}[0000]
  \item[2008] Ph.D. in Computer Science,
              University of Reading, UK
  \item[2001] M.Sc. in Computer Science,
              University of Reading, UK
  \item[2001] Dipl.-Ing. (FH) in Computer Systems Engineering,
              University of Applied Sciences Berlin, Germany
\end{compactitem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Important Peer-reviewed Publications
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Important Peer-reviewed Publications}
\vspace*{-1ex}

\begin{compactenum}[{[}1{]}]
\item D.~Fiala, F.~Mueller, C.~Engelmann, K.~Ferreira, R.~Brightwell, and R.~Riesen. Detection and correction of silent data corruption for large-scale high-performance computing. In \emph{Intl. Conf. on High Performance Computing, Networking, Storage and Analysis (SC)}, 2012.
\item J.~Elliott, K.~Kharbas, D.~Fiala, F.~Mueller, K.~Ferreira, and C.~Engelmann. Combining partial redundancy and checkpointing for HPC. In \emph{Intl. Conf. on Distributed Computing Systems (ICDCS)}, pages 615--626, 2012.
\item C.~Wang, F.~Mueller, C.~Engelmann, and S.~Scott. Proactive process-level live migration and back migration in HPC environments. \emph{J. of Parallel and Distributed Computing (JPDC)}, 72\penalty0 (2):\penalty0 254--267, 2012.
\item C.~Wang, S.~Vazhkudai, X.~Ma, F.~Meng, Y.~Kim, and C.~Engelmann. NVMalloc: Exposing an aggregate SSD store as a memory partition in extreme-scale machines. In \emph{Intl. Parallel and Distributed Processing Symp. (IPDPS)}, pages 957--968, 2012.
\item S.~B{\"o}hm and C.~Engelmann. xSim: The extreme-scale simulator. In \emph{Intl. Conf. on High Performance Computing and Simulation (HPCS)}, pages 280--286, 2011.
\item M.~Li, S.~Vazhkudai, A.~Butt, F.~Meng, X.~Ma, Y.~Kim, C.~Engelmann, and G.~Shipman. Functional partitioning to optimize end-to-end performance on many-core architectures. In \emph{ Intl. Conf. on High Performance Computing, Networking, Storage and Analysis (SC)}, pages 1--12, 2010.
\item X.~He, L.~Ou, C.~Engelmann, X.~Chen, and S.~Scott. Symmetric active/active metadata service for high availability parallel file systems. \emph{J. of Parallel and Distributed Computing (JPDC)}, 69\penalty0 (12):\penalty0 961--973, 2009.
\item A.~Nagarajan, F.~Mueller, C.~Engelmann, and S.~Scott. Proactive fault tolerance for HPC with Xen virtualization. In \emph{Intl. Conf. on Supercomputing (ICS)}, pages 23--32, 2007.
\item C.~Wang, F.~Mueller, C.~Engelmann, and S.~Scott. A job pause service under {LAM/MPI+BLCR} for transparent fault tolerance. In \emph{Intl. Parallel and Distributed Processing Symp. (IPDPS)}, pages 1--10, 2007.
\item J.~Varma, C.~Wang, F.~Mueller, C.~Engelmann, and S.~Scott. Scalable, fault-tolerant membership for MPI tasks on HPC systems. In \emph{Intl. Conf. on Supercomputing (ICS)}, pages 219--228, 2006.
\end{compactenum}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Important Professional Activities
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Important Professional Activities}
\vspace*{-1ex}

\begin{compactitem}[0000--0000 :]
  \item[2010--2012\hfill :]
    PC chair:
    Workshop on Latest Advances in Scalable Algorithms
    for Large-Scale Systems (ScalA) at the Intl.
    Conf. on High Perf. Computing, Networking,
    Storage and Analysis (SC)
  \item[2009--2012\hfill :]
    PC chair:
    Workshop on Resiliency in
    High-Perf. Computing (Resilience) at the
    Intl. Symp. on
    High Perf, Distributed Computing (HPDC), the
    Intl. Symp. on
    Cluster Computing and the Grid (CCGrid), and the
    European Conf. on
    Parallel and Distributed Computing (Euro-Par)
  \item[2006--2010\hfill :]
    PC chair:
    HPC Resiliency Summit at the
    Los Alamos CS Symp. (LACSS),
    and the
    High Availability and Perf. Workshop (HAPCW)
    at the Los Alamos CS Inst. (LACSI) Symp.
  \item[2012\hfill :]
    PC vice-chair:
    IEEE Intl. Symp. on Parallel and Distributed
    Processing and Apps. (ISPA)
  \item[2009\hfill :]
    PC vice-chair:
    IEEE Intl. Conf. on Networking, Architecture, and
    Storage (NAS)
\end{compactitem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Memberships
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Professional Memberships}
\vspace*{-1ex}

USENIX,
ACM + SIGHPC/SIGOPS,
IEEE + ComSoc/CS/RL,
IEEE CS TCFT/TCDP/TCPP/TCSC

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Collaborators and Co-editors
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Collaborators and Co-editors in Past 48 Months
          (Excl. Advisors, Advisees, Junior Pers., and Non-US)}
\vspace*{-1ex}

\textbf{ANL:}
  P. Beckman,
  A. Chien,
  R. Gupta,
  R. Ross,
  M. Snir;
\textbf{Cray:}
  L. Kaplan;
\textbf{DOE:}
  L. Nowell;
\textbf{Emory U:}
  M. Benzi;
\textbf{IBM:}
  M. Elnozahy;
\textbf{Intel:}
  S. Borkar;
\textbf{ISI:}
  B. Lucas;
\textbf{LA Tech:}
  C. Leangsuksun;
\textbf{LANL:}
  N. DeBardeleben;
\textbf{LBNL:}
  P. Hargrove,
  E. Roman,
  J. Wu;
\textbf{LLNL:}
  M. Schulz,
  B. Still;
\textbf{NCSU:}
  X. Ma,
  F. Mueller;
\textbf{ORNL:}
  S. Atchley,
  D. Bernholdt,
  D. Dillow,
  T. Evans,
  S. Hamilton,
  Y. Kim,
  T. Naughton,
  B. Park,
  G. Shipman,
  G. Vall\'ee,
  S. Vazhkudai;
\textbf{PNNL:}
  S. Krishnamoorthy,
  A. Vishnu;
\textbf{SNL:}
  A. Gentile,
  J. Brandt,
  R. Brightwell,
  R. Clay,
  K. Ferreira,
  M. Heroux;
\textbf{TN Tech:}
  S. Gafoor;
\textbf{UTK:}
  G. Bosilca,
  J. Dongarra;
\textbf{VA Tech:} 
  A. Butt

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Graduate and Postdoctoral Advisors
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Graduate and Postdoctoral Advisors}
\vspace*{-1ex}

Prof. V. Alexandrov, BSC, Spain;
A. Geist, ORNL;
Prof. S. Scott, TN Tech \& ORNL;
Prof. U. Metzler, U of Applied Sciences Berlin, Germany

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Graduate and Postdoctoral Advisees
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\subsubsection*{Graduate and Postdoctoral Advisees}
\vspace*{-1ex}

R. Baumann, TI, Germany;
S. B{\"o}hm, ORNL;
I. Jones, Ocado, UK;
F. Lauer, UTK;
A. Litvinova, Gresham Computing, UK;
B. K{\"o}nning, TU of Berlin;
K. Uhlemann, Coca Cola, Germany;
M. Weber, TU of Dresden

\pagebreak
\endinput

%%---------------------------------------------------------------------------%%
%% end of hamilton.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/bios/hamilton.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% hamilton.tex
%%---------------------------------------------------------------------------%%

\subsection{Biographical sketch for Steven P. Hamilton}

\subsubsection*{Address}
\begin{tabular}{l}
  Oak Ridge National Laboratory \\
  Radiation Transport Group\\
  Reactor and Nuclear Systems Division\\
  One Bethel Valley Road\\
  P.O.~Box 2008, MS-6170\\
  Phone:  (865) 576-3535, Fax:    (865) 574-9619	\\
  Email:  \href{mailto:hamiltonsp@ornl.gov}{\texttt{hamiltonsp@ornl.gov}}
\end{tabular}

%%---------------------------------------------------------------------------%%

\vspace*{-1ex}
\subsubsection*{Education and Training}
\vspace*{-1ex}
\begin{tabbing}
  \hspace*{1ex} 
  xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxx \kill
  Georgia Institute of Technology \> Nuclear Engineering       \> B.S. 2006 \\
  Georgia Institute of Technology \> Nuclear Engineering       \> M.S. 2007 \\
  Emory University                \> Computational Mathematics \> PhD  2011
\end{tabbing}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Research and Professional Experience}
\vspace*{-1ex}
\begin{tabbing}
  \hspace*{1ex} 
  %%
  \= 2011 -- present \hspace*{2ex} \= R\&D Staff, Oak Ridge National Laboratory\\
  %%
  \> 2009 -- 2010 \> Instructor,Mathematics and Computer Science Department,
  Emory University\\
  %%
  \> 2008 \> Research Practicum, Computational Physics (CCS-2), Los Alamos
  National Laboratory\\
  %%
  \> 2006 \> Intern, Reactor Physics Group, Oak Ridge National Laboratory
\end{tabbing}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Related Publications}
\begin{enumerate}
\parskip = -2pt

\item S. Hamilton, K. Clarno, B. Philip, M. Berrill, R. Sampath and S. Allu.
   Integrated Radiation Transport and Nuclear Fuel Performance for
   Assembly-Level Simulations.  \textit{Proc. PHYSOR}, 2012.

\item S. Hamilton, K. Clarno, B. Philip, M. Berrill, R. Sampath and S. Allu.
   Coupled Radiation Transport and Thermomechanics using the AMP and Denovo
   Codes.  \textit{Proc. 11th Copper Mtn. Conference on Iterative Methods}, 
   2012.

\item S. Hamilton and K. Clarno.  Mathematical Framework for Coupling the AMP 
   and Denovo Codes.  \textit{Trans. Am. Nucl. Soc.}, {\bf 105}, 515--517, 2011.

\item S. Hamilton and M. Benzi.  A Davidson Method for the k-Eigenvalue Problem.
   \textit{Trans. Am. Nucl. Soc.}, {\bf 105}, 432--424, 2011.

\item S. Hamilton and M. Benzi.  Eigensolvers for Radiation Transport
   Applications.  \textit{Proc. 11th Copper Mountain Conference on 
   Iterative Methods}, 2010.

\item S. Hamilton, M. Benzi and J. Warsa.  Negative Flux Fixups in 
   Discontinuous Finite Element SN Transport.  \textit{Proc. International
   Conference on Mathematics, Computational Methods and Reactor Physics},
   2009.

\item S. Hamilton, M. Benzi and E. Haber.  New Multigrid Smoothers for the Oseen 
   Problem. Num. Lin. Alg. with Apps. {\bf 17}, 557--576, 2010.

\item S. Hamilton, M. Benzi and E. Haber.  New Smoothers for the Oseen 
   Problem. \textit{Proc. 14th Copper Mtn. Conference on Multigrid Methods},
   2009.

\item S. Hamilton, K. Clarno and C. de Oliveira.  Error Control in a 
   Time-Dependent Slice-Balance Method.  \textit{Trans. Am. Nucl. Soc.},
   {\bf 97}, 533--535, 2007.

\item W. Stacey et. al.  Advances in the Subcritical,
   Gas-Cooled Transmutation Reactor Concept.  Nucl. Tech. {\bf 159}, 2007.
\end{enumerate}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Synergistic Activities}
\vspace{-1ex}
%
\begin{enumerate}
\item Developer, Denovo parallel $S_N$ transport code and the Advanced
  Multi-Physics (AMP) Nuclear Fuel Performance package at Oak Ridge National
  Laboratory.
\item Alumnus, DOE Computational Science Graduate Fellowship program.
\item Member, American Nuclear Society.
\end{enumerate}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Co-Authors and Collaborators (last 48 months)}
\vspace{-1ex}

{\bf LANL:} James Warsa
{\bf ORNL:} Srikanth Allu, Mark Baird, Kevin Clarno, Greg Davidson, Tom Evans,
Josh Jarrell, Wayne Joubert, Bobby Philip, Rahul Sampath,
{\bf Univ. Arizona}: Barry Ganapol,
{\bf Univ. British Columbia}: Eldad Haber,
{\bf Univ. New Mexico}: Cassiano de Oliveira,
{\bf Univ. Tennessee}: James Banfield

\vspace*{-2ex}
\subsubsection*{Graduate Advisor}
\vspace{-1ex}
Michele Benzi, {\bf Emory University}.

\pagebreak
\endinput

%%---------------------------------------------------------------------------%%
%% end of hamilton.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/bios/joubert.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% joubert.tex
%%---------------------------------------------------------------------------%%

\subsection{Biographical sketch for Wayne Joubert}

\subsubsection*{Address}
\begin{tabular}{l}
  Oak Ridge National Laboratory \\
  Scientific Computing Group\\
  Oak Ridge Leadership Computing Facility\\
  1 Bethel Valley Road, MS-6008\\
  Phone:  (865) 241-2835, Fax:    (865) 281-2850	\\
  Email:  \href{mailto:joubert@ornl.gov}{\texttt{joubert@ornl.gov}}
\end{tabular}

%%---------------------------------------------------------------------------%%

\vspace*{-1ex}
\subsubsection*{Education and Training}
\vspace*{-1ex}
\begin{tabbing}
  \hspace*{1ex} 
  xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxx \kill
  University of Louisiana \> Mathematics \> B.S.  1981 \\
  Univerity of Texas      \> Mathematics \> Ph.D. 1990
\end{tabbing}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Research and Professional Experience}
\vspace*{-1ex}
\begin{tabbing}
  \hspace*{1ex} 
  %%
  \= 2008 -- present \hspace*{2ex} \= Senior R\&D Staff, Oak Ridge National Laboratory\\
  %%
  \> 2007 -- 2008 \> Computational Scientist, Engineer Research and Development Center\\
  %%
  \> 2002 -- 2007 \> Computer Scientist, Xiylon Software\\
  %%
  \> 1999 -- 2001 \> Computer Scientist, LizardTech, Inc.\\
  %%
  \> 1998 -- 1999 \> Team Leader, Los Alamos National Laboratory\\
  %%
  \> 1992 -- 1998 \> Technical Staff, Los Alamos National Laboratory\\
\end{tabbing}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Related Publications}
\begin{enumerate}
\parskip = -2pt

\item C. Baker, G. Davidson, T. Evans, S. Hamilton, J. Jarrell and W. Joubert.
   High Performance Radiation Transport Simulations: Preparing for TITAN.
   \textit{International Conference for High Performance Computing, Networking, Storage and Analysis (SC12)}, 2012.

\item W. Joubert and S. Su.
   An Analysis of Computational Workloads on the Jaguar Cray XT System.
   \textit{International Conference on Supercomputing}, 2012.

\item W. Joubert, D. Kothe and H. Nam.
      PREPARING FOR EXASCALE:  ORNL Leadership Computing Application Requirements and Strategy.
   \textit{ORNL/TM-2009/308}, 2009.

\item B. Philip, M. Pernice, W. Joubert and B. Lally.
      A Comparison of Multilevel Preconditioners for Solving Multimaterial Equilibrium Radiation Diffusion Problems on Locally Refined Grids.
   \textit{Nuclear Weapons Highlights}, 2007.

\item W. Joubert and J. Cullum.
   Scalable Algebraic Multigrid on 3500 Processors.
   \textit{Electronic Transactions in Numerical Analysis}, {\bf 23}, 2006.

\item W. Joubert and G. Carey.
   PCG: A Software Package for the Iterative Solution of Linear Systems on Scalar, Vector and Parallel Computers.
   \textit{Scalable High Performance Computing Conference}, 1994.

\item W. Joubert and T. Oppe.
   Improved SSOR and Incomplete Cholesky Solution of Linear Equations on Shared Memory and Distributed Memory Parallel Computers.
   \textit{Numerical Linear Algebra with Applications}, {\bf 1}, 1994.

\item  W. Joubert.
   Lanczos Methods for the Solution of Nonsymmetric Systems of Linear Equations.
   \textit{SIAM Journal on Matrix Analysis and Applications}, {\bf 13}, 1992.
\end{enumerate}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Synergistic Activities}
\vspace{-1ex}
%
\begin{enumerate}
\item Performance Task Lead, Oak Ridge Leadership Computing Facility.
\item ORNL Center for Advanced Application Readiness, readiness lead for Denovo code, Titan system.
\item Requirements analysis for 100-250 PF ORNL OLCF-4 system.
\item Member, Society for Industrial and Applied Mathematics.
\end{enumerate}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Co-Authors and Collaborators (last 48 months)}
\vspace{-1ex}

{\bf LANL:} David Moulton.
{\bf Mellanox:} Richard Graham.
{\bf Cray:} Jeff Larkin, John Levesque, Norm Troullier.
{\bf ORNL:} Chris Baker, W. Michael Brown, Kevin Clarno, Greg Davidson, Tom Evans, Steven Hamilton, Rebecca Hartman-Baker, Oscar Hernandez, Josh Jarrell, Ricky Kendall, Doug Kothe, Don Maxwell, Bronson Messer, Hai Ah Nam, Bobby Philip, Steve Poole, Ramanan Sankaran, Shiquan Su, Arnold Tharrington, John Turner, Bogdan Vacaliuc.
{\bf SNL:} Richard Barrett.
{\bf University of Colorado:} Samuel Skillman.

\vspace*{-2ex}
\subsubsection*{Graduate Advisor}
\vspace{-1ex}
David M. Young, {\bf University of Texas}.

\pagebreak
\endinput

%%---------------------------------------------------------------------------%%
%% end of joubert.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/bios/tme.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% tme.tex
%%---------------------------------------------------------------------------%%

\subsection{Biographical sketch for Thomas M. Evans}

\subsubsection*{Address}
\begin{tabular}{l}
  Oak Ridge National Laboratory \\
  Radiation Transport Group\\
  Reactor and Nuclear Systems Division\\
  One Bethel Valley Road\\
  P.O.~Box 2008, MS-6170\\
  Phone:  (865) 576-3535, Fax:    (865) 574-9619	\\
  Email:  \href{mailto:evanstm@ornl.gov}{\texttt{evanstm@ornl.gov}}
\end{tabular}

%%---------------------------------------------------------------------------%%

\vspace*{-1ex}
\subsubsection*{Education and Training}
\vspace*{-1ex}
\begin{tabbing}
  \hspace*{1ex}
  xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxx \kill
  Haverford College \> Physics and Astronomy \> B.S. 1992 \\
  Georgia Institute of Technology \> Health Physics \> M.S. 1994 \\
  Georgia Institute of Technology \> Nuclear Engineering \> Ph.D. 1997 
\end{tabbing}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Research and Professional Experience}
\vspace*{-1ex}
\begin{tabbing}
  \hspace*{1ex} 
  %%
  \= 2012 -- present \hspace*{2ex} \= Distinguished R\&D Staff, Oak Ridge National
  Laboratory\\ 
  %%
  \> 2007 -- 2012 \> Senior R\&D Staff, Oak Ridge National Laboratory\\
  %%
  \> 2003 -- 2007 \>  Project Leader , Radiation Transport Group,
  Los Alamos National Laboratory\\
  %%
  \> 1997 -- 2003 \> Techical Staff Member, Los Alamos National Laboratory\\
  %%
  \> 1997 -- 1997 \> Post-doctoral Fellow, Los Alamos National Laboratory
\end{tabbing}

%%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Related Publications}
\begin{enumerate}
\parskip = -2pt

\item J.C. Wagner, S.W. Mosher, T.M. Evans, D.E. Peplow, and
  J.A. Turner, {\em Hybrid and Parallel Domain-Decomposition Methods Development to
  Enable Monte Carlo for Reactor Analyses,} {Prog. Nuc. Sci. Tech.},
  {2}:815-820, 2011.
\item G.G. Davidson, T.M. Evans, R.N. Slaybaugh, and C.G. Baker, {\em Massively
  Parallel Solutions to the k-Eigenvalue Problem},
  {Trans. Am. Nucl. Soc.}, {103}, 2010.
\item T.M. Evans, K.T. Clarno, and J.E. Morel, {\em A transport acceleration
  scheme for multigroup discrete ordinates with
  upscattering}, {Nuc. Sci. Eng.}, { 165}:1-13, 2010.
\item T.M. Evans, A.S. Stafford, R.N. Slaybaugh, and K.T. Clarno, 
  {\em Denovo---A
  new three-dimensional parallel discrete ordinates code in
  SCALE}, {Nuc. Tech.}, { 171}:171-200, 2010.
\item T.M. Evans and S.W. Mosher, {\em A Monte Carlo Synthetic
  Acceleration Method for the Non-Linear, Time-Dependent Diffusion
  Equation},  {International Conference on Mathematics,
    Computational Methods and Reactor Physics}, Saratoga Springs, NY,
  ISBN: 978-0-89448-069-0, American Nuclear Society, LaGrange Park,
  IL, 2009.
\item Ryan G. McClarren, Thomas M. Evans, Robert B. Lowrie, and
  Jefferey D. Densmore, {\em Semi-Implicit Time Integration for $P_N$
  Thermal Radiative Transfer},  {J. Comp. Phys.}, 
  { 227}:7561-7586, 2008.
\item J.D. Densmore, T.M. Evans, and M.W. Buksas, {\em A Hybrid
  Transport-Diffusion Algorithm for Monte Carlo Radiation-Transport
  Simulations on Adaptive-Refinement Meshes in $XY$ Geometry},
  {Nuc. Sci. Eng.}, { 159}:1-22, 2008.
\item T.M. Evans and J.D. Densmore {\em Methods for Coupling Radiation,
  Ion, and Electron Energies In Grey Implicit Monte Carlo},
  {J. Comp. Phys.}, { 225}:1695-1720, 2007.
\item T.A. Brunner, T.J. Urbatsch, T.M. Evans, and N.A. Gentile,
  {\em Comparison of Four Parallel Algorithms for Domain Decomposed
  Implicit Monte Carlo}, {J. Comp. Phys.}, { 212}:527--539, 2006.
\item T.M. Evans, T.J. Urbatsch, H.Lichtenstein, and J.E. Morel. A
  residual Monte Carlo Method for Discrete Thermal Radiative
  Diffusion. \textit{J. Comp. Phys.}, {\bf 189}(2), 539--556, 2003.
\end{enumerate}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Synergistic Activities}
\vspace{-1ex}
%
\begin{enumerate}
\parskip = -2pt
\item Project Leader, Denovo parallel $S_N$ and {\em Shift} Monte Carlo
  transport codes at Oak Ridge National Laboratory, which are part of the
  SCALE package.
\item PI, Deterministic $S_N$ Transport, CASL Modeling and Simulation Hub at
  Oak Ridge
\item PI, 2012--2013 INCITE Project {\em The Solution of Three-Dimensional PWR
    Neutronics Benchmark Problems for CASL}.
\item Member, technical committee for several Math and Computation Division
  (MCD) meetings of the American Nuclear Society 
\item Officer, MCD Vice President, 2012--2013; incoming MCD President,
  2013--2014; MCD Secretary, 2000--2001.
\end{enumerate}

%%---------------------------------------------------------------------------%%

\vspace*{-3ex}
\subsubsection*{Co-Authors and Collaborators (last 48 months)}
\vspace{-1ex}

{\bf LANL:}  J. Densmore, A.~Hungerford, R.~Lowrie, T.~Urbatsch,
{\bf LLNL:} T.~Brunner, N.~Gentile, 
{\bf Michigan:}  E.~Larsen, B.~Martin, 
{\bf MIT:} B.~Forget,
{\bf ORNL:} C.~Baker, R.~Bartlett, K.~Clarno, G.~Davidson, S.~Hamilton,
J.~Jarrell, S.~Johnson, D.~Kothe, S.~Mosher, D.~Peplow, J.~Turner, J.~Wagner,
C.~Webster, M.~Williams,
{\bf Oregon:} T.~Palmer,
{\bf SNL:} M.~Heroux, R.~Pawlowski
{\bf Texas A\&M:} R.~McClarren, J.~Morel, J.~Ragusa, 
{\bf Wisconsin:}  P.~Wilson

\vspace*{-2ex}
\subsubsection*{Graduate Advisor}
\vspace{-1ex}
C.K.~Wang, {\bf Georgia Institute of Technology}

\vspace*{-2ex}
\subsubsection*{PhD Dissertation Advisees}
\vspace{-1ex}
Rachel Slaybaugh, {\bf University of Wisconsin}
Stuart Slattery, {\bf University of Wisconsin}

\pagebreak
\endinput

%%---------------------------------------------------------------------------%%
%% end of tme.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/budgets.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% budgets.tex
%%---------------------------------------------------------------------------%%

\addcontentsline{toc}{section}{\protect\numberline{}Budgets}
\includepdf[pages=-,pagecommand={\section*{Budgets}\subsection*{ORNL Budget}
  \thispagestyle{fancy}}, noautoscale=true,scale=0.9] {supp/doe_4620_1}
\addcontentsline{toc}{subsection}{\protect\numberline{}ORNL Budget}
\includepdf[pages=-,pagecommand={\thispagestyle{fancy}},
noautoscale=true,scale=0.9] {supp/doe_4620_2}
\includepdf[pages=-,pagecommand={\thispagestyle{fancy}},
noautoscale=true,scale=0.9] {supp/doe_4620_3}
\includepdf[pages=-,pagecommand={\thispagestyle{fancy}},
noautoscale=true,scale=0.9] {supp/doe_4620_4}

\subsection*{ORNL Budget Explanation}
\addcontentsline{toc}{subsection}{\protect\numberline{}ORNL Budget Explanation}

Cost estimates presented in this proposal have been reclassified to be
comparable with proposals from other research institutions.  At the Oak Ridge
National Laboratory (ORNL), actual costs will be collected and reported in
accordance with the Department of Energy (DOE) approved cost accounting
system.  Total cost presented in this proposal and actual costs will be
equivalent as will the subtotal of direct and indirect costs.  Details of the
budget breakdown of all budget categories, including salaries, benefits and
fringe, and direct costs are described below.  In consideration of cost
estimates for ORNL staff support, it is important to understand that all ORNL
costs are covered by proposals such as this one.  There is no ``base'' funding
for DOE Office of Science multiprogram national laboratories, including ORNL.

\begin{description}

\item {\bf A. (1-6) Senior Personnel}
  %\vspace{2ex}

  The ORNL cost accounting system incorporates wage pools which are built
  around actual salaries for staff with similar salaries.  The salary figure
  listed for Senior Personnel represents the average salary for a staff
  scientist within a specific wage pool.  For budgeting purposes, one calendar
  month is assumed to be 150.0 hours.

  Support is requested for Drs. Christian Engelmann, Thomas Evans, Steven
  Hamilton and Wayne Joubert as detailed in the budget spreadsheets.
  Dr. Thomas Evans will serve as Research Coordinator and Project PI.

\item {\bf B. (1-6) Other Personnel}
  %\vspace{2ex}

  No funds are allocated toward other staff for this proposal.

\item {\bf C. Fringe Benefits}
  %\vspace{2ex}

  Fringe benefits are included in the ORNL employees wage pool rate and are
  estimated at 59.0\% for FY2012, decreasing to 48.0\% for FY2013--2017.

\item {\bf D. Permanent Equipment}
  %\vspace{2ex}

  No funds for permanent equipment are allocated.

\item {\bf E. Travel}
  %\vspace{2ex}

  Travel in the amount of \$4,000 per year is being included in the proposal
  in order to interact with M. Benzi and his students and present results at
  seminars.  This sum also allows travel to 1 to 2 meetings per year in order
  to present results of the project.

\item {\bf G. (1-6) Other Direct Costs}
  %\vspace{2ex}

  \begin{description}

  \item {\bf 1. Materials \& Supplies}
    %\vspace{2ex}

    Materials and supplies include general and miscellaneous supplies directly
    related to the project.  The additional budget for the first year is
    provided for software licenses for both toolkits needed to complete work
    for the proposal and collaboration software that will enable a virtual
    ``one-roof'' collaboration between ORNL and Emory.

  \item {\bf 6. Other --- Organization Burden Administration}
    %\vspace{2ex}

    Organizational Burden Administration costs include utilities (purchased
    utilities as well as laboratory staff associated with maintaining the
    utility systems), space charges (building maintenance), division
    managerial oversight, technical and administrative support, and other
    support personnel such as plant and equipment, instrumentation and
    controls, environmental, safety, and health, finance and budget, quality,
    and health physics provided for the general benefit of all staff and R\&D
    activities.  Inclusion of these costs is necessary to provide a full
    accounting of estimated cost for the project period.  All cost will be
    collected and reported in ORNL's cost accounting system, as approved by
    DOE.

  \end{description}

\end{description}

\pagebreak

%%---------------------------------------------------------------------------%%

\includepdf[pages=1,pagecommand={\subsection*{Emory Budget}
  \thispagestyle{fancy}}, noautoscale=true,scale=0.7] {supp/emory_budget_pages}
\addcontentsline{toc}{subsection}{\protect\numberline{}Emory Budget}
\includepdf[pages=2-3,pagecommand={\thispagestyle{fancy}},
noautoscale=true,scale=0.7] {supp/emory_budget_pages}
\includepdf[pages=4,pagecommand={\thispagestyle{fancy}},
noautoscale=true,scale=0.7] {supp/emory_budget_pages}
\includepdf[pages=5-6,pagecommand={\thispagestyle{fancy}},
noautoscale=true,scale=0.7] {supp/emory_budget_pages}
\includepdf[pages=7,pagecommand={\thispagestyle{fancy}},
noautoscale=true,scale=0.7] {supp/emory_budget_pages}
\includepdf[pages=8-10,pagecommand={\thispagestyle{fancy}},
noautoscale=true,scale=0.7] {supp/emory_budget_pages}

\subsection*{Emory Budget Explanation}
\addcontentsline{toc}{subsection}{\protect\numberline{}Emory Budget
  Explanation}

\noindent {\bf Salary}

One month summer salary in each year of the project is requested to free the
PI from teaching in order to conduct a significant portion of the proposed
research. Fringe benefits for faculty members at Emory are currently 30\% of
salary. A 3\% raise per year is projected.

The proposed research provides an excellent opportunity for a motivated Ph.D.
student working on a dissertation in large-scale scientific computing on
emerging architectures.  The Ph.D. program in Computational Mathematics at
Emory University has a strong history of attracting top graduate students,
many of whom went on to successful careers in academia, government, and
industry, including several females.  Funds to support a full-time (academic
year) Ph.D.  student are requested for each of the three years of the project.
Student salary is \$19,500 for the first year, with a 3\% raise per year.
Additionally, funds to cover the student's enrollment fees totalling
\$2,813/year are being requested.

\medskip
\par\noindent 
{\bf Travel}

Travel support to attend meetings with the other project investigators plus
one relevant conference per year, for a total of \$6,116 in year one, \$6,353
year two, and \$6,598 in year three is requested.

\pagebreak
\endinput

%%---------------------------------------------------------------------------%%
%% end of budgets.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/emory/bio-benzi.tex
Moved documents into new mcrex repo.

--- 2013-10-15 doc/proposal/bios/benzi.tex
@@
-%%---------------------------------------------------------------------------%%
-%% benzi.tex
-%%---------------------------------------------------------------------------%%
+% THIS MUST ALL FIT IN 2 PAGES
 
-\subsection{Biographical Sketch for Michele Benzi}
+\subsubsection{Biographical Sketch for Michele Benzi}
 
@@
 \subsubsection*{Address}
+\vspace*{-1ex}
 \begin{tabular}{l}
@@
 
-%%---------------------------------------------------------------------------%%
-
-\vspace*{-1ex}
+%\vspace*{-1ex}
 \subsubsection*{Education and Training}
 \vspace*{-1ex}
+\vspace*{-2ex}
 \begin{tabbing}
-  \hspace*{1ex}
-  xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxx \kill
-  University of Bologna, Italy \> Mathematics \>  B.S., 1987 \\
-  North Carolina State University \> Mathematics \> M.S., 1991 \\
-  North Carolina State University \> Applied Mathematics \> Ph.D., 1993
+\hspace*{3em}  \= \\
+\> Ph.D., 1993, Applied Mathematics, North Carolina State University \\
+\> M.S., 1991,  Mathematics, North Carolina State University\\
+\> B.S., 1987,  Mathematics, University of Bologna, Italy\\
 \end{tabbing}
+\vspace*{-4ex}
 
-%%---------------------------------------------------------------------------%%
-
-\vspace*{-3ex}
+%\vspace*{-1ex}
 \subsubsection*{Research and Professional Experience}
 \vspace*{-1ex}
+\vspace*{-2ex}
 \begin{tabbing}
-  \hspace*{1ex} 
-  %% 
-  \= 2012 -- present \hspace*{2ex} \= Samuel Candler Dobbs Professor, Emory
-  University\\
-  %%
-  \> 2006 -- 2012 \> Professor, Emory University\\ 
-  %% 
-  \> 2000 -- 2006 \> Associate Professor, Emory University \\
-  %%
-  \> 1998 -- 2000 \> Staff member, Scientific Computing Group, LANL \\ 
-  %%
-  \> 1997 -- 1998 \> Director-funded postdoctoral associate, Scientific 
+\hspace*{3em} \= \\
+\> Samuel Candler Dobbs Professor (2012--present), Emory University \\ 
+\> Professor (2006--present), Emory University\\ 
+\> Associate Professor (2000--2006), Emory 
+   University \\
+\> Staff member (1998--2000), Scientific Computing Group, LANL \\ 
+\> Director-funded postdoctoral associate (1997--1998), Scientific 
   Computing Group, LANL\\
-  %%
-  \> 1996 -- 1997 \> Postdoctoral Researcher, Parallel Algorithms Group,
-  CERFACS, Toulouse, France\\
-  %%
-  \> 1993 -- 1996 \>  Researcher, Department of Mathematics, University of Bologna,
-  Italy
+\> Postdoctoral Researcher, Parallel Algorithms Group,
+  CERFACS, Toulouse, France (1996--1997)\\
+\> Researcher, Department of Mathematics, University of Bologna,
+   Italy (1993--1996)\\
 \end{tabbing}
+\vspace*{-4ex}
 
-%%---------------------------------------------------------------------------%%
-
-\vspace*{-3ex}
-\subsubsection*{Related Publications}
-\begin{enumerate}
-\parskip = -2pt
+%\vspace*{-1ex}
+\subsubsection*{Publications} 
+% List up to ten relevant publications.
+%\vspace*{-1ex}
+\begin{enumerate*}
 
@@
 \item S.~P.~Hamilton, M.~Benzi and J.~S.~Warsa, ``Negative-flux fixups in 
-  discontinuous finite element $S_N$ transport'', 
+discontinuous finite element S_N transport'', 
 Proceedings of the International Conference on Mathematics, 
-  Computational Methods \& Reactor Physics 2009 (M\&C 2009), 
+Computational Methods & Reactor Physics 2009 (M&C 2009), 
 American Nuclear Society, Vol. 4 (2009), pp. 2529-2538, 
@@
 
-\end{enumerate}
+\end{enumerate*}
 
-%%---------------------------------------------------------------------------%%
+%\pagebreak
 
-\vspace*{-3ex}
+%\vspace*{-1ex}
 \subsubsection*{Synergistic Activities} 
-\vspace{-1ex}
-%
-\begin{enumerate}
-  \parskip = -2pt
+%\vspace*{-1ex}
+\begin{enumerate*}
 \item Chair, SIAM Activity Group on Linear Algebra (2010-present); 
-\item Member of the SIAM Council (2009-present);
+\item Member
+of the SIAM Council (2009-present); 
 \item SIAM Fellow (Class of 2012); 
-\item Program Committeee Co-chair, 2012 SIAM Applied Linear Conference
-  (Valencia, Spain, June 18-22, 2012);
-\item Program Committeee Co-chair, SIAM Annual Meeting (Minneapolis, MN, July
-  9-13, 2012).
+\item Program Committeee Co-chair,
+2012 SIAM Applied Linear Conference (Valencia, Spain, June 18-22, 2012); 
+\item Program Committeee Co-chair, SIAM Annual Meeting (Minneapolis, MN,
+July 9-13, 2012).
 \item Editor-in-Charge, SISC Copper Mountain Special Issue (2012-present).
-\end{enumerate}
+\end{enumerate*}
 
-%%---------------------------------------------------------------------------%%
-
-\vspace*{-3ex}
-\subsubsection*{Co-Authors and Collaborators (last 48 months)}
-\vspace{-1ex}
+%\vspace*{-1ex}
+\subsubsection*{Collaborators and Co-editors}
+% List Collaborators for the past 48 months and co-editors for the
+% past 24 months.
+\vspace*{-1ex}
 
@@
 Advisors:
-\begin{enumerate}
+\begin{enumerate*}
 \item Wayne Joubert (Postdoc, LANL)
@@
 \item Carl Meyer (PhD, NCSU)
-\end{enumerate} 
+\end{enumerate*} 
 \vspace*{-1ex}
 Advisees: 
-\begin{enumerate}
+\begin{enumerate*}
 \item Paola Boito (Postdoc) 
@@
 \item Cheng-yi Zhang (PhD)
-\end{enumerate}
-
-\pagebreak
-\endinput
-
-%%---------------------------------------------------------------------------%%
-%% end of benzi.tex
-%%---------------------------------------------------------------------------%%
+\end{enumerate*}

### 2013-10-15 doc/proposal/emory/budget.tex
Moved documents into new mcrex repo.

\section*{Budget Justification }

%\bigskip

\noindent {\bf Salary}

One month summer salary in each year of the project
is requested to free
the PI from teaching in order to conduct a significant portion of
the proposed research. Fringe benefits for faculty members at Emory
are currently 30\% of salary. A 3\% raise per year is projected.

The proposed research provides an excellent opportunity 
for a motivated PhD student working on a dissertation in
large-scale scientific computing on emerging architectures. 
The PhD program in Computational 
Mathematics at Emory University has a strong history of attracting
top graduate students, many of whom went on to successful careers
in academia, government, and industry, including several females.
Funds to support a full-time (academic year) PhD student 
are requested for each of the three years of the project.
Student salary is \$19,500 for the first year (plus 
7.65\% FICA), with a 3\% raise per year.
Additionally, funds to cover the student's 
enrollment fees 
totalling \$2,500/year are being requested.

\medskip
\par\noindent
{\bf Travel}

Travel support to attend meetings with the other project
investigators plus one relevant conference per year,
for a total of \$4,000/year, is requested.

\end{document}

### 2013-10-15 doc/proposal/emory/facilities.tex
Moved documents into new mcrex repo.

The Department of Mathematics and Computer Science at 
Emory University maintains appropriate computing facilities for our
research needs. These include
a shared memory, multi-processor system suited for multi-threaded, 
CPU-intensive programs. The current hardware is a Sun Microsystems 
SunFire V40z, with 4 Dual Core AMD Opteron(tm) Processors and 32 GB of 
memory running Linux.

The department also maintains a second shared memory, multi-processor system 
 suited for multi-threaded, CPU intensive programs. The current hardware 
is a Sun Microsystems SunFire V880, with 8 CPUs and 16 GB of memory running 
Solaris.

Additionally,
the department maintains {\em Puma},
a high performance cluster with 32 nodes and 
128 processor cores. Each node has two dual core AMD 2214 2.2 GHz 
Opteron CPUs, 4 GB RAM and an 80 GB drive. 
The nodes are connected via a High Performance InfiniBand network and also 
Gigabit Ethernet. The nodes are running Linux CentOS.

The department also maintains a parallel computing cluster with 
32 CPUs and 16 GB of memory split across 16 identical systems. 
All nodes use RedHat Linux/Intel for their operating system. 
Nodes are interconnected with switched, 100 Mb ethernet. 
Two build environments are available for MPI programs. 
One uses the GNU compiler suite [gcc/g++/g77]; 
the other uses the Portland Group (PGI) compiler suite.

The department also utilizes workstations from the teaching 
lab to form a compute grid for off-hour and weekend use. 
The pool of nodes normally includes 32 identical systems each 
with 2 SPARC CPUs; by arrangement, 16 additional systems could 
be recruited to bring the total CPU count up to 96.

Emory researchers have also access to a 
256 dual-core, dual-socket AMD Opteron-based compute cluster,
with a total of 1024 CPUs.
Each node uses AMD 2218 processors and has
8 GB DDR2 RAM per node (or 2 GB per core),
250 GB SATA drive local storage per node with 
Gigabit Ethernet interconnect.
The total global storage (parallel file system)
is 8 TB.

Further resources for Emory researchers are available at the
Emerson Center for Scientific Computation.

### 2013-10-15 doc/proposal/emory/support-benzi.tex
Moved documents into new mcrex repo.

\subsubsection{Michele Benzi}

\subsubsubsection{Pending}
\begin{itemize}

\item \FundingWithTotal
{DOE ASCR}

{MCREX: Using Monte Carlo algorithms to achieve resiliency 
and performance at scale for linear and non-linear solver
applications (this proposal)}
{8.33}
{FY13--16}
{\$???}

\item \FundingWithTotal
{DOE ASCR}

{A Forest for the Trees: An Ecosystem of Generalized
N-Body Solvers}
{8.33}
{FY13--16}
{\$???}

\subsubsubsection{Current}
\begin{itemize}

\item \FundingWithTotal
{NSF}

{Numerical Linear Algebra Tools for the Analysis of Complex Networks}
{14}
{08/01/11--07/31/14}
{\$303,045}

\end{itemize}

### 2013-10-15 doc/proposal/facilities.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% facilities.tex
%%---------------------------------------------------------------------------%%

\section{Facilities \& resources}

Due to ORNL's world class resources for computing, ORNL will be the prinicpal
provider of facilities and resources for this collaborative proposal.
However, facility descriptions are provided for each institution in the
proposal. 
%
\nopagebreak
\subsection{Oak Ridge National Laboratory}
\begin{center}
\includegraphics[scale=0.965]{supp/ORNL_fac1}
\end{center}

\includepdf[pages=-,pagecommand={\thispagestyle{fancy}},noautoscale=true,scale=0.99]
{supp/ORNL_Facility}

\subsection{Emory University}
\input{supp/emory}

\pagebreak
\endinput

%%---------------------------------------------------------------------------%%
%% end of facilities.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/narrative.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% narrative.tex
%%---------------------------------------------------------------------------%%

%%---------------------------------------------------------------------------%%
\section{Introduction}
\label{sec:introduction}
%%---------------------------------------------------------------------------%%

In nearly all computational engineering and physics fields, linear and
non-linear solvers form the core components of modeling and simulation
applications. Recent focus on multiphysics coupling adds additional complexity
to common linear and non-linear systems as solution strategies change when
physical models are coupled. Furthermore, a desire for predictive simulations
to enhance the safety and performance of engineered systems creates a need for
extremely high fidelity computations to be performed for these coupled systems
as a means to capture effects not modeled by coarser methods. In order to
achieve this high fidelity, state-of-the-art computing facilities must be
leveraged in a way that is both efficient and considerate of hardware-related
issues. As scientific computing moves towards exascale facilities, with
machines of $O(1,000,000)$ cores already coming online, new algorithms to
solve these complex problems must be developed to leverage this new
hardware. Issues such as resiliency to node failures and scaling to large
numbers of heterogeneous computing elements (CPUs and GPUs) will be pertinent
to robust algorithms aimed at this new hardware.

A natural path of investigation in addressing these issues would be to analyze
and adapt the existing class of state-of-the-art solvers (Krylov, multigrid,
etc.) in order to make them efficient and resilient on current and future
hardware.  Instead, we pose the question, ``Do algorithms exist that naturally
enable resiliency while preserving accuracy, convergence, and robustness?''
Our objective is to answer this question by proposing a novel group of
stochastic methods to advance solution techniques for linear and non-linear
problems with a focus on resiliency against both hard and soft errors.

For many decades, the particle transport community has been utilizing Monte
Carlo methods for the solution of transport problems. The partial differential
equation (PDE) community has focused on various deterministic methods for
solutions to linear problems. In between these two areas are a not widely
known small group of stochastic methods for solving sparse linear systems
\cite{hammersley_1964, halton_1962, halton_1994}. In recent years, we have
further developed these methods for transport problems in the form of Monte
Carlo Synthetic-Acceleration (MCSA) that have yet to be applied to more
general sparse linear systems. Compared to other methods in these regimes,
MCSA offers three attractive qualities: (1) the linear operator need not be
symmetric or positive-definite, thereby reducing preconditioning complexity,
(2) parallelization using modern methods developed by the transport community
\cite{Wagner:2011wc} is possible, and (3) the stochastic nature of the
solution method provides a natural solution to the issue of resiliency.

In addition to linear solver advancements, non-linear solvers may also benefit
from a general and parallel MCSA scheme. In the engineering community,
non-linear problems are often addressed by linearizing the problem and using
traditional iterative or direct methods. In the mathematics community, various
Newton methods have been popular \cite{kelley_1995}. Recently, Jacobian-Free
Newton-Krylov (JFNK) schemes \cite{knoll_2004} have been utilized in
multiphysics and advanced single physics codes. The benefits of JFNK schemes
are that the Jacobian is never formed, simplifying the implementation, and a
Krylov solver is leveraged (typically GMRES or Conjugate Gradient), providing
excellent convergence properties for well-conditioned and well-scaled
systems. However, there are two potential drawbacks to these methods for high
fidelity predictive simulations: (1) the Jacobian is approximated by a
first-order differencing method on the order of machine precision such that
the error can grow beyond that of those in a fine-grained system and (2) for
systems that are not symmetric positive-definite (which will be the case for
most multiphysics systems and certainly for most preconditioned systems) the
Krylov subspace generated by methods utilizing a long recurrence relation such
as GMRES may become prohibitively large. To address these issues, this work
proposes novel methods for non-linear systems based on the MCSA
method. Although the Jacobian must be explicitly formed to use MCSA, for
problems that take more than a few GMRES iterations to converge the storage
required for the Krylov subspace will likely grow beyond that of the
Jacobian. Finally, using MCSA for the linear solve provides its benefits for
preconditioning, parallelism, and resiliency.

To present the key and novel components of this work, we first provide
background on stochastic methods for solving linear problems and give results
from our initial work in this area. We then provide a research plan that aims
to drive the development of these methods with an emphasis on addressing the
issue of resiliency and performance at the exascale.

%%---------------------------------------------------------------------------%%
\section{Background}
\label{sec:background}
%%---------------------------------------------------------------------------%%

The idea of using Monte Carlo methods (random walks) to invert linear systems
is not new.  The earliest referenceable work dates to a 1950 paper by Forsythe
and Leibler \cite{forsythe}.  Their paper credits the idea to unpublished work
by J.v. Neumann and S.M. Ulam dating back to the 1940's.  The basic principles
of Monte Carlo matrix inversion were further elucidated in Hammersley and
Handscomb's 1964 text \cite{hammersley_1964}.  These early methods are
distinguished by very slow, statistically noisy convergence properties; thus,
they have not made any significant impact in the linear solver community.

In order to address the convergence issues plaguing Monte Carlo solvers,
Halton \cite{halton_1962,halton_1994} proposed a Sequential Monte Carlo
method.  This algorithm demonstrated dramatically improved convergence over
regular Monte Carlo.  Nonetheless, this method has not gained widespread use
as a production-quality linear solver.  In the next sections, we will briefly
describe the Monte Carlo methods that can be used to solve linear systems.
Furthermore, we will show preliminary results from our extension of Halton's
Sequential Monte Carlo method, the Monte Carlo Synthetic Acceleration Method
(MCSA) \cite{mc2009}.

%%---------------------------------------------------------------------------%%

\subsection{Monte Carlo Solver Methods}
\label{sec:monte-carlo-solver}

To establish the mathematical framework for Monte Carlo linear solvers, we
consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\
    &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius of the iteration matrix ($\vH$) is less than 1, we
can expand $\vA^{-1}$ using the \latin{Neumann Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} =  \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as a
series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\
    &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that solves
Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb \in \mathcal{R}^{N}$
when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte Carlo methods that
are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

%%---------------------------------------------------------------------------%%

\subsubsection{Direct Method}
\label{sec:direct-method}

Now we consider a Monte Carlo method that can be used to estimate the solution
to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First, rewrite
Eq.~(\ref{eq:neumann_series}) in order to calculate a component of $\vx$,
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
The terms in the Neumann series can be interpreted as a series of transitions
from $i_{m-1}\rightarrow i_m$ that can be simulated by a random walk.  Let $X$
be a random variable sampled from a random walk with $k$ events that initiates
in state $i$,
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\
    &= \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each random
walk starts with unit weight. At every step, contributions to each component
of $X$ are generated only in the starting state of the history, $i_0$, by the
source term in the current state, $b_{i_m}$. Therefore, for a given random
walk permutation, each state must be used as a starting value in order to
calculate its component of the expectation value (i.e. a state of size $N$
will have $N$ histories per random walk permutation). When there is a
transition between states, for example $i\rightarrow j$, the weight is
multiplied by the transition factor.
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$ to
$j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\
    &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore, the
estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator of the
components of $\vx$ provided $\rho(\vH) < 1$.

We are left to define the transition probabilities. The most straightforward
approach is to set
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_{j}|h_{ij}|}\:.
  \label{eq:probability}
\end{equation}
Each row of the transition probability matrix, $\vP$, represents a discrete
probability density function that can be sampled to select a new state $j$,
given that the current state is $i$.  Random walks can be terminated in two
ways: the matrix can be augmented with a terminating event equation that
describes the probability that a history ends its walk, or the random walk can
be terminated by weight cutoff, $W_c$.  Generally, we choose to terminate
random walks using a weight cutoff as opposed to augmenting the matrix such
that the random walk terminates when $W_m < W_c$.

%%---------------------------------------------------------------------------%%

\subsubsection{Adjoint Method}
\label{sec:adjoint-method}

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk. In this
method, the weight change from state $i\rightarrow j$ is
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}
The transition probabilities may be calculated as
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}
Note that the indices are reversed so that the probabilities are
normalized over a column, as opposed to the forward method in which
the probabilities are normalized over a row.  This is equivalent to
forming the Neumann series in reverse order.  Correspondingly, the
estimator for this method is
\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\
    &= \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
                                  w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}
Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in state $i_0$.
The Kronecker delta implies that tallies are only made in the state where the
random walk currently resides.  This is the common approach in standard Monte
Carlo transport simulations. Unlike the direct method, because tallies are
made in the state in which the random walk currently resides, we are no longer
required to start a history in each state for each random walk
permutation. Instead, sampling the source is sufficient and requires only one
random walk per permutation.

Similar to the direct method, the adjoint method random walk process requires
a terminating condition.  In all of the work that follows we utilize a
relative weight cutoff.  The weight cutoff is defined as a fraction of the
starting weight such that
\begin{equation}
  W_f = W_c\hat{b}_{i_0}\:,
  \label{eq:weight_cutoff}
\end{equation}
where $W_f$ is the terminating weight of the random walk and $W_c$ is the
input relative weight cutoff. The random walk terminates on the $m^\text{th}$
step if $W_m < W_f$.

%%---------------------------------------------------------------------------%%

\subsection{Monte Carlo Synthetic-Acceleration}
\label{sec:iter-refin-monte}

The methods presented in \S~\ref{sec:background} are characterized by slow
convergence rates bound by the Central Limit Theorem.
Halton~\cite{halton_1962,halton_1994} proposed a staged residual scheme called
Sequential Monte Carlo to speed up the convergence of these methods.  A
variant of this scheme has been successfully applied to the 1D non-linear
thermal radiation diffusion equation in Ref.~\cite{evans_2003}.

Concisely, the Sequential Monte Carlo method solves Eq.~(\ref{eq:Ax=b}) using
the adjoint solution technique described in \S~\ref{sec:adjoint-method}.  The
following iteration scheme is applied:
\begin{subequations}
  \begin{gather}
    \vr^{l} = \vb - \vA\vx^{l}\:, \\
    \vA\delta\vx^{l+1} = \vr^{l}\:, \label{eq:residual_solve}\\
    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
In this scheme, the Monte Carlo adjoint method is used to estimate the
solution to Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
convergence.  This iteration sequence is closely related to iterative
refinement; the exception being that there is no update to $x^{l+1}$
between residual iterations.

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration for the fixed-point iteration given by
Eq.~(\ref{eq:richardson}). We begin by subtracting Eq.~(\ref{eq:richardson})
from Eq:~(\ref{eq:point_iteration}):
\begin{equation}
  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
  \label{eq:synthetic_error}
\end{equation}
where
\begin{equation}
  \delta \vx^{l} = \vx - \vx^{l}
\end{equation}
is defined as the error at iteration $l$. We then subtract $(\vI-\vA) \delta
\vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}), giving a linear system to
solve for the error:
\begin{equation}
  \begin{split}
    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\
    &= \vr^{l+1}\:.
  \end{split}
  \label{eq:synthetic_error2}
\end{equation}

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\
    %%
    \vA \delta \vx^{l+1} = \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    %% 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$ directly in
Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
create a new iterative scheme by using the Monte Carlo method to estimate the
solution error. With this in hand, the Fixed-Point Monte Carlo
Synthetic-Acceleration (MCSA) method can be defined as follows:
\begin{subequations}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\
    %%
    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
    %%
    \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
    %% 
    \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
Monte Carlo solution only approximately inverts this operator.  Thus, we
have defined a scheme in which the initial estimate of $\vx$ in each iteration
is updated using a single fixed-point iteration\footnote{Although, a Krylov
  method could be used instead of a stationary method here.}.  The residual is
calculated and is used as a source to estimate the error, $\delta\vx^{l+1/2}$,
by solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the solution
vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence based on
the following stopping criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

%%---------------------------------------------------------------------------%%

\subsubsection{Preliminary Results}
\label{sec:preliminary-results}

The MCSA method can use either the direct or adjoint method to solve
Eq.~(\ref{eq:MCSA-residual_solve}). We have performed a preliminary study that
demonstrates the superiority of the adjoint method within the context of MCSA
using the two-dimensional time-dependent Poisson equation as a simple model
problem with second and fourth-order finite-difference stencils. A timing and
convergence study is used to demonstrate the effectiveness of the adjoint
method as compared to the direct method.

We see clearly in Fig.~\ref{fig:adjoint_v_direct}a that using the adjoint
\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \subfigure[]{\includegraphics[width=3in]{Adjoint_Direct_CPU_Time}} &
      \subfigure[]{\includegraphics[width=3in]{Adjoint_Direct_Iterations}} 
    \end{tabular}
    \caption{Comparison of Direct and Adjoint MCSA methods for varying problem
      size specified by an $N\times N$ grid, (a) time to solution, (b)
      iterations required for solution.}
    \label{fig:adjoint_v_direct}
  \end{center}
\end{figure}
solver with MCSA results in several orders of magnitude speedup over the
direct solver while the number of iterations required to converge is of a
similar scale. We expect this for several reasons. First, with an equivalent
number of histories specified for both solvers and a system of size $N \times
N$, the direct solver will compute $N \times N$ random walks per permutation
while the adjoint solver will only compute one. Initially, this may not seem
like a fair comparison, however, we see from Fig.~\ref{fig:adjoint_v_direct}b
that the number of iterations required to converge is approximately the same
and therefore the extra computations performed by the direct method do not
improve the error estimation.

In order to gauge the performance of MCSA we have used it to solve the
thermal-equilibrium radiation diffusion equation,
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi}^{n+1} + \sign\phi^{n} = \qn\:,
  \label{eq:equil-diffusion}
\end{equation}
where $\phi$ is the energy-integrated radiation intensity, $\sign$ is the
effective absorption plus reemission, $\Dn$ is the diffusion coefficient, and
$\qn$ is the source at $t^n$.  In operator form,
Eq.~(\ref{eq:equil-diffusion}) is $\ve{D}\bphi = \ve{q}$ where $\ve{D}$ is an
SPD operator.  Using left-preconditioning, the MCSA algorithm that is applied
at each timestep is
\begin{align*}
  %%
  \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\
  %%
  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
  \quad & (\text{compute residual})\\
  %%
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\
  %%
  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
  (\text{update}\ \bphi)
\end{align*}
As noted above, the residual acts as the source for this simulation.  There
are two Monte Carlo interpretations that can be applied to this system.  The
first follows the mathematical presentation given in \S~\ref{sec:background}.
Namely, we form probabilities from the iteration matrix and perform random
walks to generate unbiased estimates of the solution vector,
$\delta\bphi^{l+1/2}$, using the estimator in Eq.~(\ref{eq:adjoint-tally}).

However, a more natural interpretation is to consider the random walk as a
transport process; in the problem above this is equivalent to a particle
transport process.  In this connotation, the machinery described in
\S~\ref{sec:adjoint-method} is used to define Probability Distribution
Functions (PDFs) that determine the transport of state transitions through the
linear system (matrix). The random walk is a Monte Carlo transport calculation
that uses Eq.~(\ref{eq:adjoint-weight}) to calculate the weight change at each
transition. Equation~(\ref{eq:adjoint-probability}) gives the probability of
transmission to an adjacent state in the system. We tally the contribution to
the solution ($\phi$ in the above example) in each state using
Eq.~(\ref{eq:adjoint-tally}).  The transport process is terminated using a
relative weight cutoff that is defined in Eq.~(\ref{eq:weight_cutoff}).

There are two principal degrees of freedom when utilizing the MCSA method: (1)
the number of histories per iteration ($N_p$), and (2) the weight cutoff
($W_c$).  Figure~\ref{fig:Np}a shows the CPU time as a function of the
requested number of histories per iteration for a relative weight cutoff of
$\sn{1}{-4}$ for a Marshak wave problem \cite{larsen_1980}.  For less than 5
histories per iteration the method does not converge.  Figure~\ref{fig:Np}b
shows the variation in the maximum number of iterations per cycle with number
of histories.  Analysis of these two figures shows that increasing the number
of histories reduces the total number of iterations required to converge the
solution.  However, the cost of the Monte Carlo transport is high; thus,
better performance is attained by running more iterations with fewer histories
per iteration \cite{evans_2003}.
\begin{figure}[h]
  \begin{center}
    \begin{tabular}{cc}
      \subfigure[]{\includegraphics[width=3in,clip]{mrshk_np_CPU.pdf}} &
      \subfigure[]{\includegraphics[width=3in,clip]{mrshk_np_Ns.pdf}}
    \end{tabular}
  \end{center}
  \caption{(a) CPU time versus number of histories per iteration, and (b)
    maximum number of iterations per cycle versus number of histories per
    iteration for the Marshak problem. The number of histories is the
    requested number.  In each iteration the exact number can vary depending
    upon the number of cells and source strength per cell.}
  \label{fig:Np}
\end{figure}

A similar analysis was done comparing different weight cutoffs. These results
indicate that the performance of the MCSA method is relatively insensitive to
the weight cutoff.  This is especially true when running small numbers of
histories per iteration.  When larger numbers of histories are used the weight
cutoff has a more pronounced effect on the efficiency of the method.

We have compared the MCSA solution algorithm with both Jacobi-preconditioned
Conjugate Gradient (PCG) and Jacobi iteration (JACOBI). The model problem is a
dog-legged duct through a thick wall where the radiation flows into a thin
region bound by a foil on one side. The size of the spatial grid was $60^3$
(216,000 spatial cells).  All three methods, give numerically identical
answers when using a stopping criterion of $\sn{1}{-8}$.  The geometry and
time-evolution of the temperature is shown at 4 edit points in
Fig.~\ref{fig:edit-points}.
\begin{figure}[h]
  \begin{center}
    \begin{tabular}{m{2in}m{4in}}
      \subfigure[Edit points]{\includegraphics[width=2in,clip]{points}} &
      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat}}
    \end{tabular}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature
    at each point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}
Table~\ref{tab:multimat_comparison} shows the timing results for the
multi-material problem using PCG, PFIX, and MCSA.  The results
correspond roughly with the results from the Marshak problem.  The
MCSA is marginally faster than PCG and 43\% faster than PFIX.
\begin{table}[h]
  \caption{
    Comparison of solution methods for the multi-material problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=1000$ and $W_c = \sn{1}{-3}$.  The problem
    was run to an elapsed time of 1000~ns.}
  \label{tab:multimat_comparison}
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & 
      \multicolumn{1}{c}{Max Iterations} & 
      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
      %% 
      PCG & 18 & 1.03 \\
      %% 
      PFIX & 38 & 1.43 \\
      %% 
      MCSA & 20 & 1.00 \\
      \hline\hline
    \end{tabular}
  \end{center}
\end{table}

%%---------------------------------------------------------------------------%%
\section{Research Plan}
\label{sec:research-plan}
%%---------------------------------------------------------------------------%%

As stated in \S~\ref{sec:introduction}, we plan to analyze and extend MCSA in
order to achieve numerical efficiency, robustness, and accuracy combined with
resiliency and performance on current and future HPC architectures.  The
objective is to obtain resiliency in a manner that naturally fits into the
complete MCSA algorithm.  Numerical characterization and improvements to MCSA
will be performed using the tools of standard numerical analysis.  We intend to
verify solver resiliency through \textit{fault injection},
described in the following sections.  Likewise, we will extrapolate
performance estimates to proposed future HPC architectures by developing a
\textit{performance model}.

%%---------------------------------------------------------------------------%%

\subsection{Numerical and Algorithmic Investigations}
\label{sec:numer-char}

Numerical investigations of MCSA will be performed in two phases (a)
demonstrating convergence, accuracy, efficiency, and robustness, and (b)
determining the applicability of subspace iteration schemes.  
Algorithmic investigations will
be focused on improving parallel performance.  Both numerical and algorithmic
research will investigate resiliency aspects of the solution process.  
We define resiliency as the ability to
recover from both hard and soft errors.  Soft errors consist of logical,
floating-point, and other processor evaluation faults short of process
failure.  Hard errors are full process failures.  We will show how we can
address each failure type algorithmically through numerical analysis and
parallel redundancy.

All numerical and algorithmic investigations will be demonstrated using
linear and non-linear forms of the Advection-Diffusion-Reaction (ADR) equations,
\begin{equation}
  \partial_t u+ \grad\cdot\ve{f}(\ve{x},u - \grad\cdot\bigl[
  d(\ve{x},u)\grad u\bigr] + r(\ve{x},u)u = q(\ve{x},t)\:.
\end{equation}
This system is representative of the wide class of problems that the
algorithms and methods proposed in this study should solve efficiently.  In
order to verify the methods on real models, we will also consider the
non-linear consistent, incompressible Navier-Stokes equations
\cite{Evans:2006dd,Evans:2007el,Pernice:2001th}. All investigations will
consider two and three-dimensional forms of the models in order to demonstrate
broad applicability to relevant science and engineering problems.

\subsubsection{Numerical Investigations}

The first stage of our numerics research will be demonstrating convergence,
accuracy, efficiency, and robustness of the MCSA algorithm.  The preliminary
results described in \S~\ref{sec:preliminary-results} indicate that the
fundamental solver converges and is efficient for sparse SPD systems.
However, significant analysis remains to determine
\begin{itemize}
\parskip = -2pt
\item its convergence properties on non-symmetric systems;
\item its effectiveness as the linear solver in non-linear Newton methods;
\item an analytical framework that relates weight cutoff and number of
  histories to problem parameters;
\item the effect of high variance events on convergence and robustness.
\end{itemize}
All of our preliminary studies have focused on sparse SPD systems, in which we
have shown MCSA to be competitive with PCG.  Because MCSA has no requirements
on symmetry or positive-definiteness, it is anticipated that the method will
compare favorably to GMRES, BICGSTAB, or TFQMR.

Once the properties of MCSA have been scoped, we can begin investigations
using MCSA as the inner solver in non-linear Newton methods.  The inner part
of each Newton iteration requires the solution to the following linear system,
\begin{equation}
  \ve{J}(\ve{u}^k)\delta\ve{u}^{k+1} = -\bff(\ve{u}^k)\:.
\end{equation}
As opposed to Krylov methods, MCSA requires the formation of the matrix
elements to build the probabilities that are used to execute random walks.
Thus, the standard inexact Jacobian-Free Newton-Krylov (JFNK)\cite{knoll_2004}
methods that rely on numerical approximations of the matrix-vector product,
\begin{equation}
  \ve{J}\ve{v} \approx \frac{\bff(\ve{u}+\epsilon\ve{v}) - 
    \bff(\ve{u})}{\epsilon}\:,
\end{equation}
are not feasible with Newton-MCSA.  Instead, we can numerically approximate
the matrix elements of the Jacobian, $J_{ij} = \partial \ff_i(\ve{u})/\partial
u_j$ directly using automatic differentiation
\cite{Bartlett_2006,Notz:2010wk}.  This technology is available in Trilinos'
Phalanx package \cite{Pawlowski:2008vy}.

For both direct application in linear systems and as an inner solver for
non-linear systems, a mathematical framework must be established that can be
used to analytically determine the runtime parameters of the solver for a
given problem.  The problem-dependent runtime parameters for the MCSA solver
are number of histories per iteration and weight cutoff.  Because the cost of
the Monte Carlo is proportional to the total number of random walks, the goal
is to minimize the total number of histories.  As indicated in
Fig.~\ref{fig:Np} this involves a balance between number of iterations and
number of histories per iteration.

The effect of high variance events has significant implications for both
robustness and resiliency to soft errors.  All of the Monte Carlo methods we
are investigating rely on the assumption that
\begin{equation}
    E[X] = \sum_{\nu}P_\nu X_\nu\:,
    \label{eq:expectation-X}
\end{equation}
is an unbiased estimator of $X$. Soft errors will result in random walk events
that appear as high-variance histories.  In order to determine algorithmic
resiliency we must determine the scale at which high-variance events
invalidate Eq.~(\ref{eq:expectation-X}).  We expect that most soft errors will
manifest as high-variance events that will be ameliorated during successive
random walks.  Thus, the MCSA algorithm provides a natural means of dealing
with the majority of soft errors.  Research must be performed to determine the
robustness of the method to high-variance events that result from the numerics
of the linear system or from soft errors.  We will use the simulator described
in \S~\ref{sec:resiliency} to validate solver robustness to these events.

In addition to investigating the numerical features of the proposed MCSA
algorithm, we also intend to investigate new iteration sequences that may
prove more efficient.  In the current MCSA method, Monte Carlo is used to
accelerate a stationary fixed-point iteration.  Instead, significant
performance gains could possibly be achieved by casting MCSA as a 
preconditioner to a Krylov subspace method.  Because the statistical 
convergence of Monte Carlo methods will inherently lead to variations in 
the preconditioning operator from one iteration to the next, it will likely 
be necessary to appeal to flexible Krylov methods\cite{saad93}. Although
the idea of using Krylov subspace methods to accelerate Monte Carlo
calculations is not new (see Ref.~\cite{conlin11} for an example applied
to eigenvalue calculations in neutron transport), significant effort 
will be necessary to quantify convergence properties of such an approach.

\subsubsection{Algorithmic Investigations}

In general, deterministic and Monte Carlo solution methods have competing
requirements with regard to achieving concurrency; efficient parallelism in
deterministic methods is often achieved by decomposing the global phase space
as much as possible, whereas in Monte Carlo efficiency is often the result of
replicating as much of the phase space as possible.  An additional constraint
beyond pure performance considerations is the best decomposition strategy that
can aid resiliency, particularly to hard system errors.  Balancing these
requirements, we propose to use a recently developed algorithm for Monte Carlo
called Multi-Set, Overlapping Domain (MSOD) \cite{Wagner:2011wc} (originally
named multi-level, overlapping region in Ref.~\cite{Wagner:2011wc}).  An
illustration of the MSOD algorithm is displayed in Fig.~\ref{fig:msod} for the
ADR equations.
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=3.5in]{msod}
  \end{center}
  \caption{Schematic of the MSOD decomposition algorithm. The spatial grid
    (matrix) is decomposed into 4 blocks (blue, green, pink, violet); the
    spatial decomposition is replicated in 3 sets yielding 12 total domains, $
    N_\mathrm{domains} = N_\mathrm{blocks}\times N_\mathrm{sets}$.  The entire
    matrix/spatial grid is contained in each set.  Random walks through the
    matrix do not cross set boundaries, only block boundaries.  }
    \label{fig:msod}
\end{figure}
The spatial domain (matrix) is decomposed into blocks such that the set
$\{b_0,b_1,\ldots,b_{N_\mathrm{blocks}}\}$ constitutes the complete grid.
Each set of $N_\mathrm{blocks}$ is replicated in $N_\mathrm{sets}$.  Thus, the
total number of domains for any given problem is $N_\mathrm{blocks}\times
N_\mathrm{sets}$.

The potential advantages of this parallel algorithm are
\begin{itemize}
  \parskip = -2pt
\item the intrinsic concurrency patterns of both Monte Carlo and deterministic
  algorithms are, in part, preserved;
\item the algorithm limits to both full \textit{domain decomposition} (single
  set---multiple blocks) and full \textit{domain replication} (multiple
  sets---single block);
\item problems can be scaled to arbitrarily large size by increasing multiple
  sets;
\item a natural partitioning of work to heterogeneous hardware components is
  provided, e.g. blocks on GPUs, sets on cores;
\item the algorithm enables \textit{redundancy} by replicating the solution
  domain.
\end{itemize}
The MSOD algorithm should provide natural resiliency to hard errors through
redundancy.  In any given MCSA iteration, histories can be rejected as long as
this is done in an unbiased manner.  Hard errors result in part of the problem
domain being unsampled.  The MSOD algorithm provides a mechanism for rejecting
a complete set in these cases, thus preserving an unbiased random walk.
Additionally, a small bias may be tolerable by only rejecting the blocks that
encounter hard errors.  The amount of bias that negligibly affects the
solution will need to be determined.

%%---------------------------------------------------------------------------%%

\subsection{Resiliency Modeling}
\label{sec:resiliency}

The resilience of the proposed algorithm will be modeled though a series of
fault injection campaigns. The impact of a hard or soft error on the algorithm
will be investigated by artificially creating such errors during algorithm
execution and by observing the resulting runtime behavior, execution time, and
output. The two most likely errors in an extreme-scale system are a process
failure, e.g., a detected but unrecoverable error leading to process
termination, and silent data corruption (SDC), e.g., an undetected bit flip in
a memory latch~\cite{cappello09toward, elnozahy08system, geist09major,
  kogge08exascale, schroeder07understanding,
  Schroeder:2009:DEW:1555349.1555372, Hwang:2012:CRD:2189750.2150989,
  10.1109/TC.2011.106}. Both error types will be injected based on their
probability, i.e., with a certain frequency and probability distribution,
targeting specific algorithm vulnerabilities.

The error probability will be modeled using historical log data from ORNL
systems and assumptions communicated by HPC vendors, such as Cray and Intel,
regarding the reliability of future-generation systems. The algorithm
vulnerabilities will be modeled using a whitebox analysis for identifying the
most vulnerable execution paths and data structures in combination with
runtime profiling for identifying system component (processor, memory, and
network) usage. Correlating algorithm vulnerabilities with system component
usage provides a number of injection points at which a specific error class is
most likely and is expected to result in a significant impact, such as a
process fault during a critical collective operation or SDC in the most
significant bits of an important value. The fault injection campaigns will
utilize different existing technologies to model the resilience of the
algorithm on today's and on future-generation HPC systems.

The fault injection campaigns will provide the data points needed to model the
resilience of the proposed algorithm, such that the probability of computing a
correct result within a certain time to solution can be estimated for both
error classes depending on the reliability properties of the HPC system, the
scale of the application run, and the problem size.

\subsubsection{Hard Errors on Today's HPC Systems}

For investigation of hard errors on today's HPC systems, a fault-tolerant MPI
solution will be used to implement a fully functional resilient algorithm atop
a message passing runtime environment that is capable of detecting and
surviving process failures and notifying an application about each failure. Injecting a
hard error is as simple as manually killing a process. The fault-tolerant MPI
automatically reconfigures to survive the process failure and notifies the
application about the failure. The notification can be used to trigger any
reconfiguration of the application, including algorithm-level recovery. The
PIs have access to two different fault-tolerant MPI solutions, FT-MPI
developed by the University of Tennessee in 2001~\cite{fagg01harness,
  fagg01fault} and the prototype for the MPI 3.x standard with fault tolerance
support developed by ORNL in 2011~\cite{hursey12:parco:hips, hursey11:ft_ring,
  hursey11:ft_coll}. This fault injection campaign will observe the impact of
hard errors on the algorithm's execution time and result.

\subsubsection{Soft Errors on Today's HPC Systems}

For investigation of soft errors on today's HPC systems, the redundant MPI
implementation, redMPI~\cite{fiala12detection2, elliott12combining,
  boehm12file, engelmann11redundant, engelmann09case}, will be used to perform
comparative studies at runtime. redMPI transparently executes an application
in a redundant fashion by utilizing the MPI performance tool interface, PMPI,
to transparently intercept MPI calls from an application and to hide all
redundancy-related mechanisms. A redundantly executed application runs with
$r*m$ native MPI processes, where $r$ is the number of MPI ranks visible to
the application and $m$ is the replication degree.

Under normal operation, messages between redundant nodes are replicated and
compared for hard and soft error detection and correction. For fault injection
campaigns~\cite{fiala12detection2}, the message replication and comparison
protocol performs detection only. This permits the original parallel
application to be tainted with data corruption, while the fully redundant
parallel application serves as a correct live control for close observation of
the error impact, including its propagation, detection, and
masking. Differences between the tainted execution and the control are
detected and recorded at runtime by the message replication and comparison
protocol, providing detailed information about error sensitivity and
propagation.

\subsubsection{Hard and Soft Errors on Future HPC Systems}
\label{sec:hard-soft-errors}

For investigation of hard and soft errors on future-generation HPC systems,
the Extreme-scale Simulator (xSim) performance investigation toolkit will be
utilized. xSim~\cite{boehm11xsim, engelmann10facilitating, jones11simulation}
is a recently developed performance investigation toolkit that permits running
HPC applications in a controlled environment with millions of concurrent
execution threads. It allows observing application performance in a simulated
extreme-scale system for hardware/software co-design. Using a lightweight
parallel discrete event simulation, xSim executes an application on a much
smaller HPC system in an oversubscribed fashion with a virtual wall clock
time, such that performance data can be extracted based on a processor and a
network model with the appropriate simulation scalability/accuracy
trade-off. xSim is designed like a traditional performance tool, as an
interposition library that sits between the MPI application and the MPI layer,
using the MPI performance tool interface. It currently holds the world record
in extreme-scale simulation, running up to 134,217,728 ($2^{27}$)
communicating MPI tasks, each with its own process context, using just a
960-core Linux-based cluster.

As part of an LDRD project at ORNL (see Current and Pending Support for
Christian Engelmann in \S~\ref{sec:current--pending}), xSim is being
extended with advanced features to (1) permit the injection of different
faults, errors, and failures into the simulation, (2) model various
propagation, isolation, and detection properties of the simulated system, (3)
support a variety of avoidance, masking, and recovery strategies, (4) model
the power consumption of the entire simulated system, and (5) study the
performance, resilience, and power consumption impact with different parameter
sets for (1), (2), (3), and (4) using standardized methods and metrics. Within
this proposal, we will leverage the new xSim capabilities developed under the
LDRD project, such as the support for a simulated fault-tolerant MPI,
algorithm-based fault tolerance, and SDC injection, for investigating the
impact of hard and soft errors on the proposed algorithm in simulated
future-generation HPC systems. Both hard and soft errors will be injected
into the algorithm execution atop a simulated HPC system, while the simulator
allows close observation of the error impact, including its propagation,
detection, and masking as well as the algorithm's execution time and result.

%%---------------------------------------------------------------------------%%

\subsection{Performance Modeling}
\label{sec:performance-modeling}

The DARPA Exascale Report \cite{darpa-exascale} made clear the multiple
challenges facing software and algorithm developers on the path to exascale,
including the memory wall, the power wall, resiliency issues and the need for
unprecedented levels of parallelism.  Exascale application development is made
particularly challenging by the uncertainty regarding the final hardware
characteristics of these systems due to the multiple innovations that will be
required in the coming years.

The Oak Ridge Leadership Computing Facility (OLCF) has successfully navigated
application and algorithm migration across multiple generations of new HPC
system installations and upgrades, most recently with the Jaguar Istanbul
processor upgrade in 2009 and the OLCF-3 Titan installation which is currently
underway.  Central to this effort has been the performance modeling and
analysis of key science applications and their algorithms as input to the
system design and procurement process.  This effort has required the OLCF to
develop and validate predictive models for many of its heavily-used
application codes to understand their anticipated performance characteristics
on future hardware.  The performance models have successfully guided the
development teams' code and algorithm changes both for incremental hardware
upgrades as well as disruptive hardware changes such as the advent of
heterogeneous GPU-based systems.

To insure the success of the MCREX solvers on exascale hardware, a vigorous
effort will be undertaken to model the algorithm and code performance
characteristics as part of the algorithm design process.  This work will take
place in the setting of the following interrelated steps.

\begin{itemize}
  \parskip = -2pt
\item {\bf Algorithm Development.}  The MCREX solver algorithms will be
  designed with concern not only for the algorithm numerics and convergence
  properties but also with anticipated future hardware characteristics in
  mind, such as the presence of hard and soft errors, increasing time and
  power costs of communicating off-die and off-node, decreasing relative sizes of
  high-speed cache memories and register files, and the need to expose
  increasing amounts of thread-level parallelism.

\item {\bf Code Implementation.} The algorithms will be implemented in software,
  initially as prototypes and then as more full-fledged versions, to be
  evaluated and tested on current state-of-the-art HPC hardware such as the
  ORNL Titan system and follow-on hardware.

\item {\bf Performance Analysis.}  The performance characteristics of these
  codes will be determined, using tools such as VAMPIR, CrayPat and other
  profiling tools to understand the performance hot spots of the algorithms
  and understand how the performance depends on characteristics of the
  different system hardware components.

\item {\bf Performance Modeling.}  Quantitative models will be developed to
  predict the performance of the algorithms for different problem types of
  interest under the assumption of multiple potential future architecture
  scenarios, based on possible futures for exascale hardware.  These findings
  will not only provide a feedback loop for further algorithm design but also
  provide an assessment of the relative effectiveness of alternative system
  hardware designs for algorithms of this type. The fully developed
  performance model will be simulated using the performance investigation
  features of the xSim toolkit, described
  in \S~\ref{sec:hard-soft-errors}.

\end{itemize}

%%---------------------------------------------------------------------------%%
\section{Research Timetable and Tasks}
\label{sec:rese-timet-tasks}
%%---------------------------------------------------------------------------%%

The research tasks described in \S~\ref{sec:research-plan} are organized into
yearly milestones with concurrent tasks.  The yearly milestones for the
project are:
\begin{description}
  \parskip=-2pt
\item[Year 1]

  Demonstrate convergence properties and performance of MCSA algorithm on
  sparse symmetric and non-symmetric systems.

\item[Year 2]

  Show resiliency of MCSA algorithm for soft and hard errors while solving the
  linear advection-diffusion and non-linear Navier-Stokes equations.

\item[Year 3]

  Show parallel performance of MCSA algorithm on existing HPC
  architectures and demonstrate scaling to exascale systems using the
  MCSA performance model.
\end{description}
The full list of major tasks, correlated by yearly milestones, are
\begin{center}
  \small
  \begin{tabular}{|c|p{3in}|l|}\hline 
    {\bf Task} & \multicolumn{1}{|c|}{\bf Description} &
    \multicolumn{1}{c|}{\bf Milestone} \\\hline
    %%
    1 & Determine performance on non-symmetric systems & Y1 \\\hline
    2 & Implement Newton-MCSA method & Y1, Y2 \\\hline
    3 & Derive and implement linear ADR model equations & Y1, Y2 \\\hline
    4 & Derive and implement non-linear, incompressible Navier-Stokes
    equations & Y1, Y2 \\\hline
    5 & Investigate solver runtime parameters & Y1 \\\hline
    6 & Analyze robustness of unbiased estimators to high-variance events &
    Y1, Y2 \\\hline
    7 & Implement MSOD parallel algorithm & Y2, Y3 \\\hline
    8 & Develop performance model for MSOD/MCSA algorithm & Y3 \\\hline
    9 & Model algorithm resiliency using fault injection campaigns & Y2, Y3 \\\hline
    10 & Estimate algorithm performance on future systems using the
    performance model and xSim & Y3 \\\hline 
  \end{tabular}
\end{center}

The products of this research are twofold:
\begin{itemize}
  \parskip = -2pt
\item Journal and conference articles and lab reports;
\item Open-source distribution code base.
\end{itemize}
All code products produced \textit{directly} as part of this proposal will be
hosted on ORNL-managed GIT repositories.  All code products will be available
through open-source distribution, but access to the development repositories
will be controlled through ORNL.

%%---------------------------------------------------------------------------%%
\section{Management Plan}
\label{sec:management-plan}
%%---------------------------------------------------------------------------%%

This is a highly focused proposal with a limited scope. Accordingly, the
management plan specifies an agile development model in order to optimize the
work output and allow maximum flexibility in pursuing new avenues of
investigation.  The project has two PIs, Drs. Thomas Evans, ORNL and Michele
Benzi, Emory University.  Dr. Evans will serve as the Research Coordinator for
the project, which will be led at ORNL.  The project member roles and
responsibilities are shown in Fig.~\ref{fig:roles}.
\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=6.3in]{roles}
  \end{center}
  \caption{MCREX project  member roles and responsibilities.}
  \label{fig:roles}
\end{figure}
Also illustrated are the institutional roles.  ORNL will host the project
management site (\textsf{http://fogbugz.ornl.gov}, see
\textsf{http://www.fogcreek.com/fogbugz} for details) and GIT software and
document repositories.  An electronic notebook server using NoteShare
(\textsf{http://www.aguaminds.com}) will be maintained at ORNL that allows
instant sharing of progress and results.

Because this project only involves 5 professional staff plus one student, the
process should be agile so that subsequent investigations can be easily
modified to reflect current work.  Accordingly, the project lifecycle model
that will be employed is based on an agile Kanban\cite{Kniberg:2010wx}
process. All milestones and tasks will be tracked on a Kanban board hosted at
ORNL.  This project will leverage experience gained in the CASL DOE NE Hub
(CASL, \textsf{http://www.casl.gov}) for managing an inter-institutional
project under a virtual ``one roof.''  The relatively close proximity between
ORNL and Emory makes travel between the institutions feasible.  Furthermore,
bi-weekly and as-needed conference calls/interactive coding sessions using
ReadyTalk (\textsf{http://www.readytalk.com}) can be used to easily facilitate
collaboration.  As noted above, electronic notebooks will be used to provide
instant access to current status and results of ongoing research.

All code produced in this project will be constructed using a test-driven
development model~\cite{evans_ieee}.  Test-driven code development is not only
essential for production-level code project; it is required for research in
order to verify that output is the result of numerics, as opposed to code
defects.

\pagebreak
\endinput

%%---------------------------------------------------------------------------%%
%% end of narrative
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/templates/cv_template.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% CV For DOE Proposals
%%---------------------------------------------------------------------------%%

\documentclass[11pt]{article}

%%---------------------------------------------------------------------------%%

\pagestyle{empty}

\setlength{\textheight}{9.5in}
\setlength{\textwidth}{6.5in}
\setlength{\footskip}{0.5in}
\setlength{\headheight}{0.0in}
\setlength{\topmargin}{0.0in}
\setlength{\headsep}{0.0in}

\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}

\setlength{\parindent}{0in}

%%---------------------------------------------------------------------------%%

\begin{document}

%%---------------------------------------------------------------------------%%

{\bf Thomas M. Evans}\\
Radiation Transport Group, Reactor and Nuclear Systems Division, Oak Ridge
National Laboratory, P.O. Box 2008, Bldg. 5700, Oak Ridge, TN 37831-6170,
$<$evanstm@ornl.gov$>$\\

%%---------------------------------------------------------------------------%%

{\bf PROFESSIONAL EDUCATION}
\begin{tabbing}
  xxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxx \= xxxxxxxxx \kill
  Haverford College              \> Physics and Astronomy \> B.S. 1992 \\
  Georgia Insitute of Technology \> Health Physcs         \> M.S. 1994 \\
  Georgia Insitute of Technology \> Nuclear Engineering   \> PhD  1997 
\end{tabbing}

%%---------------------------------------------------------------------------%%

{\bf RESEARCH AND PROFESSIONAL EXPERIENCE}\\

\underline{Radiation Transport Group, Oak Ridge National Laboratory}\hfill
2007--present\\
\textit{Senior R\&D Staff}\\
Develop computational methods, algorithms, and parallel application codes in
the Reactor and Nuclear Systems Division for ORNL customers.  Develop and
submit proposals for scientific funding.\\

\underline{Radiation Transport Group, Los Alamos National Laboratory}\hfill
1997--2007\\
\textit{Project Leader (2003--2007)}; \textit{Techical Staff Member
  (1997--2003)};
\textit{Postdoctory Fellow (1997--1997)}\\
Project Leader of the Marmot and Jayenne projects. As a staff member,
contributed to radiation transport and radiative
transfer methods development in CCS-4 Transport Methods Group. \\

%%---------------------------------------------------------------------------%%

{\bf RELEVANT PUBLICATIONS}
\begin{enumerate}
  \setlength{\itemsep}{0in}
\item G.G. Davidson, T.M. Evans, R.N. Slaybaugh, and C.G. Baker.  Massively
  Parallel Solutions to the k-Eigenvalue Problem.
  \textit{Trans. Am. Nucl. Soc.}, {\bf 103}, 2010.
\item T.M. Evans, G.G. Davidson, and R.N. Slaybaugh.  Three-Dimensional Full
  Core Power Calculations for Pressurized Water Reactors.  \textit{Journal of
  Physics: Conference Series, SciDAC 2010}, accepted for publication, 2010.
\item T.M. Evans, K.T. Clarno, and J.E. Morel.  A transport acceleration
  scheme for multigroup discrete ordinates with
  upscattering. \textit{Nuc. Sci. Eng.}, {\bf 165}, 1--13, 2010.
\item T.M. Evans, A.S. Stafford, R.N. Slaybaugh, and K.T. Clarno.  Denovo---A
  new three-dimensional parallel discrete ordinates code in
  SCALE. \textit{Nuc. Tech.}, {\bf 171}, 171--200, 2010.
\item A.M. Ibrahim, D.E. Peplow, T.M. Evans, J.C. Wagner, and
  P.P.H. Wilson. Improving the Mesh Generation Capabilities in the
  SCALE Hybrid Shielding Analysis
  Sequence. \textit{Trans. Am. Nucl. Soc.}, {\bf 100}, 302--303, 2009.
\item T.M. Evans and S.W. Mosher.  A Monte Carlo Synthetic
  Acceleration Method for the Non-Linear, Time-Dependent Diffusion
  Equation.  \textit{International Conference on Mathematics,
    Computational Methods and Reactor Physics}, Saratoga Springs, NY,
  ISBN: 978-0-89448-069-0, American Nuclear Society, LaGrange Park,
  IL, 2009.
\item Ryan G. McClarren, Thomas M. Evans, Robert B. Lowrie, and
  Jefferey D. Densmore.  Semi-Implicit Time Integration for $P_N$
  Thermal Radiative Transfer.  \textit{J. Comp. Phys.}, {\bf 227},
  7561--7586, 2008.
\item J.D. Densmore, T.M. Evans, and M.W. Buksas.  A Hybrid
  Transport-Diffusion Algorithm for Monte Carlo Radiation-Transport
  Simulations on Adaptive-Refinement Meshes in $XY$ Geometry.
  \textit{Nuc. Sci. Eng.}, {\bf 159}, 1--22, 2008.
\item T.M. Evans and J.D. Densmore.  Methods for Coupling Radiation,
  Ion, and Electron Energies In Grey Implicit Monte Carlo.
  \textit{J. Comp. Phys.}, {\bf 225}, 1695--1720, 2007.
\item T.A. Brunner, T.J. Urbatsch, T.M. Evans, and N.A. Gentile.
  Comparison of Four Parallel Algorithms for Domain Decomposed
  Implicit Monte Carlo. \textit{J. Comp. Phys.}, {\bf 212},
  527--539, 2006.
\end{enumerate}

%%---------------------------------------------------------------------------%%

{\bf SYNERGISTIC ACTIVITIES}\\

Evans specializes in the development, implementation, and application of
computational radiation transport as applied in the areas of Nuclear
Engineering, radiation detection, astrophysics, high energy density physics,
and medical applications.  His interests include stochastic and deterministic
transport methods on massively parallel platforms, nonlinear and
time-dependent transport methods, coupled physics including
radiation-hydrodynamics, acceleration and preconditioning techniques,
optimization and performance analysis, and large-scale scientific software
design for parallel codes.  He has published over 50 refereed journal and
conference articles.  He is the primary developer of the Denovo parallel $S_N$
and {\em Shift} Monte Carlo transport codes at Oak Ridge National Laboratory,
which are part of the SCALE package.  He is currently a PI in the CASL
Modeling and Simulation Hub at Oak Ridge.\\

He has served on the technical committee for several Math and Computation
Division (MCD) meetings of the American Nuclear Society and is currently on
the MCD Executive Committee and was MCD Secretary in 2000-2001.\\

%%---------------------------------------------------------------------------%%

{\bf RECENT COLLABORATIONS AND CO-EDITORS}\\

Chris Baker, Oak Ridge National Laboratory;
Roscoe Bartlett, Sandia National Laboratory;
Tom Brunner, Lawerence Livermore National Laboratory;
Kevin Clarno, Oak Ridge National Laboratory;
Greg Davidson, Oak Ridge National Laboratory;
Jeff Densmore, Los Alamos National Laboratory;
Ben Forget, MIT;
Nick Gentile, Lawrence Livermore National Laboratory;
Mike Heroux, Sandia National Laboratory;
Aimee Hungerford, Los Alamos National Laboratory ;
Josh Jarrell, Oak Ridge National Laboratory;
Dana Knoll, Idaho National Laboratory;
Doug Kothe, Oak Ridge National Laboratory;
Ed Larsen, University of Michigan;
Robert Lowrie, Los Alamos National Laboratory;
Ryan McClarren, Texas A\&M University;
Jim Morel, Texas A\&M University;
Scott Mosher, Oak Ridge National Laboratory;
Todd Palmer, Oregan State University;
Roger Pawlowski, Sandia National Laboratory;
Douglas Peplow, Oak Ridge National Laboratory;
Jean Ragusa, Texas A\&M University;
John Turner, Oak Ridge National Laboratory;
Todd Urbatsch, Los Alamos National Laboratory;
John Wagner, Oak Ridge National Laboratory;
Mark Williams, Oak Ridge National Laboratory;
Paul Wilson, University of Wisconsin\\

\underline{PhD Advisees}\\
Rachel Slaybaugh (University of Wisconsin)\\

\underline{Students}:\\
Ahmed Ibrahim (University of Wisconsin);
Peter Maginot (Texas A\&M);
Stuart Slattery (University of Wisconsin)

\end{document}

%%---------------------------------------------------------------------------%%
%% end of CV
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/templates/support_template.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% Pending and support for DOE Proposals
%%---------------------------------------------------------------------------%%

\documentclass[11pt]{article}

%%---------------------------------------------------------------------------%%

\pagestyle{empty}

\setlength{\textheight}{9.5in}
\setlength{\textwidth}{6.5in}
\setlength{\footskip}{0.5in}
\setlength{\headheight}{0.0in}
\setlength{\topmargin}{0.0in}
\setlength{\headsep}{0.0in}

\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}

\setlength{\parindent}{0in}

%%---------------------------------------------------------------------------%%

\begin{document}

%%---------------------------------------------------------------------------%%

{\bf Thomas M. Evans}\\

%%---------------------------------------------------------------------------%%
{\bf Current Support}\\
%%---------------------------------------------------------------------------%%

Funding Organization: Department of Energy (NE)\\
Project Title: CASL Nuclear Modeling and Simulation Hub\\
Dates of the project: 6/1/10--6/1/15\\
Annual funding: \$25M\\
Level of effort: 0.66 FTE\\

%%---------------------------------------------------------------------------%%

Funding Organization: Oak Ridge National Laboratory (LDRD)\\
Project Title: Revolutionary Radiation Transport for Next-Generation
Predictive Multi-Physics Modeling and Simulation\\
Dates of the project:  10/1/09--10/1/11\\
Annual funding:  \$250K\\
Level of effort: 0.34 FTE\\

%%---------------------------------------------------------------------------%%
{\bf Pending Support}\\
%%---------------------------------------------------------------------------%%

Funding Organization:  U.S. Department of Energy, (ASCR/SciDAC)\\
Project Title: Isotope production toolkit (this proposal\\
Dates of the project: 10/1/11--9/30/16\\
Annual funding:  \$3M\\
Level of effort:  0.5 FTE\\

Funding Organization:  U.S. Department of Energy, (ASCR/SciDAC)\\
Project Title: Institute for Computational Design and Analysis at the extreme
scale (CODA) (SciDAC Institute)\\
Dates of the project: 10/1/11--9/30/16\\
Annual funding:  \$4.5M
Level of effort:  0.5 FTE\\

Funding Organization:  Oak Ridge National Laboratory (LDRD)\\
Project Title: Creating a virtual HFIR: A complete modeling and simulation
toolkit for isotope production in HFIR \\
Dates of the project: 10/1/11--10/1/13\\
Annual funding:  \$450K\\
Level of effort:  0.25 FTE\\

\end{document}

%%---------------------------------------------------------------------------%%
%% end of pending and support
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/white.tex
Moved documents into new mcrex repo.

%%---------------------------------------------------------------------------%%
%% white.tex
%%
%% White paper for MCREX proposal
%%---------------------------------------------------------------------------%%

\documentclass[10pt,letterpaper,english]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{tmath}

%%---------------------------------------------------------------------------%%

\newcommand{\vA}{\ensuremath{\ve{A}}}
\newcommand{\vb}{\ensuremath{\ve{b}}}
\newcommand{\vx}{\ensuremath{\ve{x}}}
\newcommand{\vr}{\ensuremath{\ve{r}}}
\newcommand{\vI}{\ensuremath{\ve{I}}}
\newcommand{\vH}{\ensuremath{\ve{H}}}
\newcommand{\vP}{\ensuremath{\ve{P}}}

%%---------------------------------------------------------------------------%%

\title{MCREX: Using Monte Carlo algorithms to achieve resiliency and
  performance at scale for linear and non-linear solver applications}
\author{Thomas Evans\thanks{PI}, Steven Hamilton, Wayne Joubert\\
  Oak Ridge National Laboratory
  \and Michele Benzi\\Emory University}
\date{July 10, 2012}

\begin{document}

\pagestyle{fancy}
\lhead{MCREX}
\rhead{T.M. Evans}

\maketitle

%%---------------------------------------------------------------------------%%

\section*{Proposal Objective}

The DOE LAB 12-742 FOA, {\em Resilient Extreme-Scale Solvers}, is looking for
research proposals in numerical solver algorithms that address three
components:
\begin{enumerate}
\item Advances in solvers
\item Fault tolerance and resilience at the algorithmic level
\item Performance of proposed algorithms
\end{enumerate} 
To address these topical areas we are proposing to investigate new stochastic
methods for solving linear systems or Monte Carlo Resilient, Exascale (MCREX)
solvers.  The family of methods that we have developed build on the sequential
Monte Carlo work of Halton \cite {halton_1962, halton_1994}. The methods that
we have developed use Monte Carlo to accelerate a fixed-point iteration;
therefore, we have called them Monte Carlo Synthetic Acceleration
(MCSA). Preliminary work \cite{rmc,mc} using MCSA has demonstrated that they
are at least as efficient as Jacobi-preconditioned Conjugate Gradient (PCG) on
sparse, SPD systems.

MCSA fits ideally into this call because, as described in the next section, we
can show how these algorithms can be used to address all three required
aspects of the FOA.  Furthermore, we will demonstrate how this can be achieved
by a small focused team.  We are seeking an award in the \$500K range, which
should be sufficient to perform algorithmic enhancements, demonstrate
resiliency, and do performance analysis.

%%---------------------------------------------------------------------------%%

\section*{Research Plan}

The Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method is defined,
for the linear system, $\vA\vx = \vb$, as follows:
\begin{subequations}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:, \\
    %%
    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,\\
    %%
    \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
    %%
    \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
  \label{eq:general-MCSA}
\end{subequations}
We use Monte Carlo to perform random walks through the elements of the matrix
to solve Eq.~(\ref{eq:MCSA-residual_solve}). The hat on $\vA$ indicates that
the Monte Carlo solution only approximately inverts this operator.  Thus, we
have defined a scheme in which the initial estimate of $\vx$ in each iteration
is updated using a single fixed-point iteration.  The residual is calculated
and is used as a source to estimate the error, $\delta\vx^{l+1/2}$.

As described in Refs.~\cite{rmc} and \cite{mc}, the power of this algorithm
derives from the fact that random walks are only performed where the residual
is large.  The components of the system that are converged do not require
further work.  Furthermore, as opposed to PCG, there are no symmetry
requirements for MCSA; thus, it could be an excellent alternative to GMRES,
especially for consistently-treated non-linear problems that are solved using
Newton's method.  Also, this algorithm naturally addresses resiliency because
random walk histories that are lost due to processor or system faults can be
neglected during any iteration.

The proposed research plan for this project is broken down to address each
component of the FOA:
\begin{enumerate}
\item Advances in solvers
  \begin{itemize}
  \item Investigate efficiency of MCSA on non-symmetric matrix systems.
  \item Investigate the efficiency of MCSA on non-linear solvers using
    automatic differentiation of the Jacobian matrix elements.  Numerical
    automatic differentiation will be provided by Trilinos.
  \item Investigate the extension of the method to non-stationary (Krylov
    sub-space) iteration schemes.
  \end{itemize}
\item Fault tolerance and resilience at the algorithmic level
  \begin{itemize}
  \item Incorporate MPI error signals into the algorithm to address cases of
    core failure.
  \end{itemize}
\item Performance of proposed algorithms
  \begin{itemize}
  \item Leverage work from CASL on the Multi-Set, Overlapping Domain (MSOD)
    Monte Carlo parallel decomposition method to test its efficiency for MCSA.
  \item Develop heuristic models to project its performance on projected
    heterogeneous, exa-scale computing platforms.
  \end{itemize}
\end{enumerate}
The model equations that we will utilize in all of these studies will be the
linear heat transfer equation and the non-linear, incompressible,
Navier-Stokes equations.

%%---------------------------------------------------------------------------%%

\section*{Summary}

We have proposed a highly focused, innovative project that naturally addresses
all three components of the FOA: novel algorithmic advances, resiliency, and
performance modeling.  The proposed team is 
\begin{center}
  \begin{tabular}{llr}\hline\hline
    \multicolumn{1}{c}{Team Member} & 
    \multicolumn{1}{c}{Role}        & 
    \multicolumn{1}{c}{Budget (K)} \\\hline
    %%
    Tom Evans       & PI                & 100  \\
    Steven Hamilton & Solver Algorithms & 175  \\
    Wayne Joubert   & Performance       & 100  \\
    Michele Benzi   & Solver Algorithms & 75   \\
    Other           & Resiliency        & 100  \\ \hline
    {\bf Total}     &                   & 550  \\
    \hline\hline
  \end{tabular}
\end{center}
The strength of this proposal is its focus on a single, innovative solver
technology that addresses aspects of all three pieces of the FOA.  Therefore,
we feel that a local, small team is both justified and appropriate.

%%---------------------------------------------------------------------------%%

\begin{thebibliography}{300}

\bibitem{halton_1962} 
  J.H. Halton. Sequential Monte Carlo, \textit{
    Proc. Camb. Phil. Soc.}, {\bf 58}, pp. 57-78, 1962.

\bibitem{halton_1994} 
  J.H. Halton. Sequential Monte Carlo Techniques for the
  Solution of Linear Systems, \textit{J. Sci. Comp.}, {\bf 9}(2), pp. 213-257,
  1994.

\bibitem{rmc}
  T.M. Evans, T.J. Urbatsch, H.Lichtenstein, and J.E. Morel. A
  residual Monte Carlo Method for Discrete Thermal Radiative
  Diffusion. \textit{J. Comp. Phys.}, {\bf 189}(2), 539--556, 2003.

\bibitem{mc} 
  T.M. Evans and S.W. Mosher.  A Monte Carlo Synthetic
  Acceleration Method for the Non-Linear, Time-Dependent Diffusion
  Equation.  \textit{International Conference on Mathematics,
    Computational Methods and Reactor Physics}, Saratoga Springs, NY,
  ISBN: 978-0-89448-069-0, American Nuclear Society, LaGrange Park,
  IL, 2009.

\end{thebibliography}

\end{document}

%%---------------------------------------------------------------------------%%
%% end of white.tex
%%---------------------------------------------------------------------------%%

### 2013-10-15 doc/proposal/figures/matrix.py

@@ -1,0 +1,54 @@
+##---------------------------------------------------------------------------##
+## matrix.py
+##---------------------------------------------------------------------------##
+
+import os, sys
+
+##---------------------------------------------------------------------------##
+## Build an 7-point stencil matrix
+##---------------------------------------------------------------------------##
+
+Nx = 10
+Ny = 10
+Nz = 5
+N  = Nx*Ny*Nz
+
+A = []
+
+for k in xrange(Nz):
+    for j in xrange(Ny):
+        for i in xrange(Nx):
+            row = [0] * N
+            
+            px = i + 1 + Nx * (j + k * Ny)
+            mx = i - 1 + Nx * (j + k * Ny)
+            py = i + Nx * (j + 1 + k * Ny)
+            my = i + Nx * (j - 1 + k * Ny)
+            pz = i + Nx * (j + (k + 1) * Ny)
+            mz = i + Nx * (j + (k - 1) * Ny)
+
+            c  = i + Nx * (j + k * Ny)
+            
+            row[c] = 1
+
+            if (i > 0) : row[mx] = 1
+            if (j > 0) : row[my] = 1
+            if (k > 0) : row[mz] = 1
+
+            if (i < Nx - 1) : row[px] = 1
+            if (j < Ny - 1) : row[py] = 1
+            if (k < Nz - 1) : row[pz] = 1
+
+            A.append(row)
+            
+for j in xrange(N):
+    for i in xrange(N):
+        print A[i][j],
+    print
+
+
+##---------------------------------------------------------------------------##
+## end of matrix.py
+##---------------------------------------------------------------------------##
+
+

### 2013-10-18 doc/ttsp2013/ttsp_2013.tex
adding ttsp spn paper

\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{tmadd,tmath}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[mathcal]{euscript} 
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery
\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

\date{\today}
\title{Monte Carlo Synthetic Acceleration for the Simplified $P_N$ Approximation}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\abstract

In this paper, we present Monte Carlo synthetic acceleration as a
linear solver technique and briefly derive the simplified $P_N$
($SP_N$) for fixed source and criticality problems. Using the
fully-formed linear operator for the transport problem, we explore
solutions to the $SP_N$ equations with Monte Carlo Synthetic
Acceleration using a challenging light water reactor fuel assembly
criticality calculation as the driving problem. Several difficulties
arise when applying MCSA to these problems that were not observed when
investigating the simpler transport systems and we apply additional
algebraic preconditioning techniques to achieve convergence. For
Jacobi-based preconditioners, convergence with MCSA is difficult if
not impossible for the ill-conditioned systems generated by light
water reactor problems with this behavior demonstrated using a simple
neutron diffusion problem. In order to effectively solve the $SP_N$
equations for the fuel assembly problem, a suite of preconditioners
along with a relaxation scheme is developed and studied within the
context of MCSA. Using these preconditioners, several additional
issues are observed and alleviated to a certain extent by applying the
reduced domain approximation. Finally performance is analyzed through
comparison with those same Krylov solvers in terms of both iterative
performance and CPU timing using the fuel assembly problem.

%%---------------------------------------------------------------------------%%
\section{Introduction}

The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \cite{evans_denovo:_2010}. For fission
reactor neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
method was either widely unknown, widely unused, or combination of
both even though numerical studies showed promising results with
better solutions than diffusion theory and a significant reduction in
computational time over more accurate methods such as discrete
ordinates. Why did this happen? A significant problem, pointed out by
Brantley and Larsen \cite{brantley_simplified_2000}, was that little
rigor had been applied to the formulation of the $SP_N$ equations
since their derivation through primarily heuristic arguments. Instead,
studies at that time focused on simply comparing the results of the
method to other contemporary transport solution strategies. In
addition, many problems of interest from the literature at the time
were either solved using nodal-type methods for reactor-sized problems
or $S_N$-type methods for benchmark problems with intricate material
configurations and potentially large flux gradients over small spatial
domains.

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment with both variational and
asymptotic derivations performed as a means of verification. In
addition, these equations have been more rigorously studied as
solution methods to MOX fuel problems and have been shown to provide
accurate solutions. With this mathematical literature to provide a
solid numerical footing for the method, we look at its application to
today's challenge problems in neutron transport for fission reactor
analysis. The reduction in numerical complexity of current
deterministic solution methods using the $S_N$ approximation could
mean significant savings in both compute time and memory required. In
addition, the characteristics of the solution to the transport problem
for a steady state reactor core permit diffusion theory to be used; a
staple of the nuclear industry since its inception. Therefore, if
diffusion theory is applicable, then finer grained solutions that
capture more of the physics contained in the transport equation should
be possible with the $SP_N$ method. In doing so, we also expect from
the literature to obtain computed responses on the order of accuracy
we would expect from an appropriately discretized $S_N$ method at a
fraction of the cost.

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for neutron transport problems, its current formulation
requires the physics operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems\cite{evans_simplified_2013}. By fully forming these
equations and formulating them as a linear algebra problem instead of
using the explicit iterative methods of the past, we now have access
to all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and algebraic
preconditioning methods.

This leads us to then research the applicability of Monte Carlo
Synthetic Acceleration methods as a possible solution method for the
$SP_N$ equations. In addition, solving the $SP_N$ equations in this
way also breaks away from the $S_N$ forms of parallelism where spatial
parallelism is achieved by an efficient parallel sweep, angular
efficiency achieved by pipe-lining, and energy parallelism achieved by
decoupling the groups. With the $SP_N$ equations as a full matrix
system, we now can parallelize the problem as prescribed by the Monte
Carlo algorithm, which may be significantly more scalable than current
$S_N$ transport practices. In this work we therefore apply Monte Carlo
Synthetic Acceleration to the $SP_N$ equations and research its
applicability, performance, and convergence properties in the context
of a light water reactor system.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}

Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \cite{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \cite{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers as well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. To derive this
method, we begin by splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therefore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where a Neumann-Ulam Monte Carlo method is used to generate the
solution correction from the residual and Richardson's iteration in
the first step has been rewritten as a residual correction. Using
Monte Carlo in this way achieves the same effect as Halton's method,
decoupling its convergence rate from the overall convergence rate of
the method. Here, the approximate Monte Carlo solution is not driven
to a particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the Neumann-Ulam method to
generate the correction. In addition, the fact that the scheme in
Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
is inverted exactly means that as more and more stochastic histories
are used to compute the correction and the error is reduced towards
zero, the number of iterations required for MCSA to converge should
decrease accordingly, thus accelerating the solution.

In addition to the Monte Carlo solver parameters dictating the number
of histories and weight cutoff, the outer MCSA iterations also have
the following stopping criteria:
\begin{equation}
  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
  \label{eq:mcsa_stopping_criteria}
\end{equation}
where $\epsilon$ is a user-defined parameter. As with any iterative
method, other stopping criteria using other vector norms could be
computed, however, for this work we will only use
Eq~(\ref{eq:mcsa_stopping_criteria}). We therefore have 3 parameters
to tune in an MCSA implementation: the number of Monte Carlo histories
computed in the Neumann-Ulam solve during each MCSA iteration, the
weight cutoff for those histories, and the total MCSA convergence
tolerance as specified by $\epsilon$.

%%---------------------------------------------------------------------------%%
\section{Simplified $P_N$ Approximation}

%%---------------------------------------------------------------------------%%
\section{Model Problem}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
bottom is given in Figure~\ref{fig:problem3_end}. On the top and
bottom of the assembly, vacuum conditions are used as well as on the
top and right boundaries in
Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized moderator. Each fuel pin is resolved by a $2 \times
      2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is moderator and darker
      purple/red material is fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is moderator and darker purple/red material is fuel.}}
  \label{fig:problem3_end}
\end{figure}

Significant geometric details are contained in the model including
spacer grids, fuel pins with homogenized cladding and gas gap, core
plenum, and moderator with boron. Group cross sections and other
discrete nuclear data are generated as needed by a cross-section
processing module dependent on the meshing parameters used to
discretize the geometry and single-dimension pin-cell calculations for
initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
gives the primary design parameters for the fuel assembly
calculations.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Parameter}} & 
      \multicolumn{1}{l}{\textbf{Value}} \\
      Power Level & 0 MW \\
      Inlet Temperature & 326.85C \\
      Fuel Temperature & 600C \\
      Boron Concentration & 1300 ppm \\
      Moderator Density & 0.743 g/cc \\
      Helium Density & \sn{1.79}{-4} g/cc \\
      Zirconium Density & 6.56 g/cc \\
      Stainless Steel Density & 8.0 g/cc \\
      Inconel Density & 8.19 g/cc \\
      UO2 Density & 10.257 g/cc \\
      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
      assembly criticality calculation.}}
  \label{tab:problem3_parameters}
\end{table}

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
computation\footnote{The operator will change if, for example, physics
  feedback through temperature or potentially burnup is
  considered. These additional physics will modify the cross sections
  used to assemble $\mathbf{M}$. The calculations presented here will
  consist of a single eigenvalue calculation with a static
  $\mathbf{M}$.} while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor while not
affecting Monte Carlo transport.

%%---------------------------------------------------------------------------%%
\section{Preconditioned MCSA}

In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. Preconditioning serves as a
means of achieving this by altering the eigenvalue spectrum of the
iteration matrix. It is possible to use general left, right, and left/right
preconditioning with MCSA by carefully considering the underlying
Monte Carlo problem that will be solved with the Neumann-Ulam
method. We consider here the general left/right preconditioned method
as the left or right preconditioned methods can be inferred from its
formulation.

We consider a left preconditioner $\ve{M_L}$ and a right
preconditioner $\ve{M_R}$. The left/right preconditioned linear
problem is then:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
  \label{eq:left_right_linear_problem}
\end{equation}
To handle the right preconditioning, the system is written with a
substitution of variables:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
  \label{eq:left_right_subs_problem}
\end{equation}
with
\begin{equation}
  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
  \label{eq:left_right_recover}
\end{equation}
To apply such a method to MCSA, we solve for the substituted variable
$\ve{u}$ during the iteration sequence:
\begin{subequations}
  \begin{gather}
    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
  \end{gather}
  \label{eq:left_right_mcsa}
\end{subequations}
and then recover the original solution vector with
Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
isolate the generation of the correction:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
  \label{eq:left_right_correction}
\end{equation}
and note that the preconditioned residual of the substituted variable
is now serving as the source and the new iteration matrix is:
\begin{equation}
  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
  \label{eq:left_right_iteration_matrix}
\end{equation}
As we require $(i,j)$ element-wise access to the iteration matrix in
order to construct probabilities and weights for the Monte Carlo
procedure from the Neumann-Ulam decomposition, the \textit{composite
  operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
matrix-matrix multiplication.

Several possible shortcomings of this preconditioning approach are
readily observed. First, the matrix-matrix multiplication operation
for sparse, parallel distributed matrices is significantly more
expensive than a matrix-vector multiplication operation. Second, each
preconditioner must be explicitly inverted, an operation in itself
that may be expensive and which prohibits the use of any
preconditioners which provide no mechanism to extract their
inverse. Third, for many modern preconditioning methods, this
inversion may yield dense matrices, destroying sparsity and further
impeding the performance of a matrix-matrix multiplication
operation. It is also interesting to note that the Monte Carlo problem
in the general left/right preconditioned scheme given by
Eq~(\ref{eq:left_right_correction}) is not fully left/right
preconditioned (meaning that we do not recover $\ve{x}$), but instead
part of a sequence for finding the substituted variable $\ve{u}$. We
do, however, gain the benefits of this general preconditioning by
building the iteration matrix in
Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
preconditioned linear operator. In addition, for MCSA to apply to
increasingly difficult problems, more advanced preconditioning
techniques that require the generation of the composite operator may
be necessary for convergence.

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. Even if
the problem may be ill-conditioned, we do expect convergence of
MCSA. To investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor converge is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

Based on these results, it then appears that even if the simple
criteria of a spectral radius of less than one is met and the
Richardson iteration will converge with the same preconditioning, MCSA
still may not converge. We then expect the issue to reside with the
Neumann-Ulam solve providing the correction as the Richardson
iteration is known to provide the correct result. Furthermore, the
fact that the spectral radius is less than one means that the
stochastic histories in the block Jacobi preconditioned Neumann-Ulam
method are eventually being terminated by the weight cutoff as no
artificial absorption was used for these preliminary
calculations. Based on this, the correction being generated by the
Neumann-Ulam solve is the component of MCSA causing a divergent
solution.

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi
preconditioned\footnote{For 1-group $SP_1$ problems the block size is
  1, giving a point Jacobi preconditioner.}  calculations used a
one-speed $SP_1$ discretization, effectively giving a diffusion
system. To study the breakdown of MCSA at iteration matrix spectral
radii near one, we will use the simpler homogeneous 2-dimensional
one-speed neutron diffusion system presented in
Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
this system, we can vary the cross sections while maintaining a fixed
grid in order to achieve varying spectral radii. For these studies, we
neglect fission as MCSA behavior is dictated by the transport operator
$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
generate a fixed source.

For each solver and estimator combination, the spectral radius of the
iteration matrix generated by the diffusion problem was varied by
changing the absorption cross section from 0.25 to 100 while fixing
the grid size at $100 \times 100$ with $h = 0.1$ and a fixed
scattering cross section of unity. For each parameter variation, a
minimum of one stochastic history per degree of freedom (DOF) in the
problem was used to compute the Monte Carlo correction. If the solver
could not converge in less than 100 iterations, the number of
histories was increased by increments of 5,000 until convergence was
achieved in less than 100 iterations. The number of iterations
required to converge MCSA and the time to converge was recorded as a
means to capture the breakdown.

Figure~\ref{fig:breakdown_iterations} gives the number of iterations
required to converge for the chosen number of histories per iteration
given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
with the collision and expected value estimators and the forward
solver with the collision estimator. For spectral radii less than
0.97, all MCSA problems converged with 1 history per DOF (10,000 for
this problem) with the number of iterations required to converge
increasing as a function of spectral radius. Near a spectral radius of
0.97, the number of histories required to converge MCSA in less than
100 iterations takes a dramatic rise that exhibits neither exponential
nor power law behavior. As the spectral radius approaches 1, the
number of histories required becomes significant and effectively
impractical to compute. Even with this simple diffusion problem, the
behavior is consistent with that observed for the fuel assembly
problem with $SP_1$ discretization. In that case, we estimated a
spectral radius of $\approx 0.998$, larger than any of the spectral
radii that could be computed within even 90 minutes of compute time
for this simple two dimensional problem. For that problem, even
1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
convergence. Even if single solve times of an hour can be tolerated,
dozens of solves are typically required to compute the multiplication
factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
difficult problems like the fuel assembly, making the method unusable.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{breakdown_iterations.pdf}
  \end{center}
  \caption{\textbf{Iterations required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_iterations}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{breakdown_histories.pdf}
  \end{center}
  \caption{\textbf{Histories per DOF required to converge as a
      function of spectral radius for the neutron diffusion problem.}
    \textit{The number of histories was increased to achieve
      convergence in less than 100 iterations. At least 10,000
      histories were used for each calculation with 10,000 DOFs in the
      problem.}}
  \label{fig:breakdown_histories}
\end{figure}

In addition to the significantly larger number of histories required
to achieve convergence for ill-conditioned problems another penalty is
paid due to histories that take longer to
compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
required to compute a single random walk averaged over the entire set
of histories run in the calculation over all iterations. As the
spectral radius increases (correlating to a higher ratio of scattering
in the system) the random walk lengths increase, using more CPU time
to finish the computation. Compared to spectral radii of 0.5, larger
spectral radii over 0.97 have histories that require two orders of
magnitude more computation time. This significant increase in
computation time per history coupled with the significant increase in
the number of histories required to converge is evidence that for
problems with spectral radii above $\approx 0.97$, using MCSA to solve
any problems of interest is entirely ineffective and not practical. We
therefore require a more expansive set of preconditioning techniques
to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
which MCSA is more applicable and in which performance is improved.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

\subsection{Algebraic Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}
For the fuel assembly criticality problem presented in the previous
section, the spectral radius was near one using Jacobi preconditioning
and in a regime in which MCSA breakdown is observed. In this regime
the number of stochastic histories required to converge MCSA increases
rapidly and the resulting poor performance is compounded by those
histories being increasingly expensive to compute. To overcome this
difficulty, advanced preconditioning strategies for the $SP_N$ problem
are required beyond simple Jacobi methods that can reduce the spectral
radius into a region of better MCSA performance. Several modern
algebraic preconditioning strategies will be presented here and used
within the explicit preconditioning framework given by
Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
solutions of the fuel assembly criticality problem will be presented.

\subsection{ILUT Preconditioning}
\label{subsec:spn_ilut_preconditioning}
Incomplete lower-upper (ILU) factorizations of the linear operator can
be used a simple mechanism to form an approximate inverse of a
preconditioner. To build the factorization, the sparse upper and lower
triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
that the residual matrix formed by the factorization
\cite{saad_iterative_2003}:
\begin{equation}
  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
  \label{eq:ilu_residual_matrix}
\end{equation}
has a specified sparsity pattern and element magnitude. When a
magnitude threshold is used, elements generated in the factors below
that magnitude are dropped, resulting in ILU threshold (ILUT)
preconditioning. The sparsity pattern in this case is determined from
the input matrix to be preconditioned and the number of elements
maintained in the factor is specified by a fill level parameter. A
fill level of 1 will generate the same number of elements as the
sparsity pattern of the input matrix while a fill level of 2 will
contain twice as many elements in the factorization, resulting in a
better representation of the true LU factorization. The inverse of the
lower and upper triangular factors may then be easily inverted by
means of simple elimination to produce the preconditioner. For MCSA,
$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
the right to precondition the system.

For modern subspace methods, only the action to the preconditioner on
a vector is required for efficient implementations and therefore a
triangular solve can be used for this purpose. For the explicit
preconditioning scheme presented in Eq~(\ref{eq:left_right_mcsa}), the
fully inverted operator must be generated in order to build the set of
probabilities and weights required for Monte Carlo
sampling. Therefore, for a linear system of size $N$, $N$ triangular
solves will be required in order to extract the inverse matrices from
production ILU implementations. In addition, parallel implementations
typically generate lower and upper factors that are only triangular
locally, providing an easily parallelizable mechanism to generate the
action of the preconditioner inverse. A consequence of this choice for
parallel scalability is a degradation of the preconditioning quality
as the size of the parallel system is increased. For serial
computations, the triangular factorization is potentially exact
depending on the parameters chosen while at thousands of parallel
tasks, the global triangular factors differ significantly from the true
factorization. As a result, more iterations are required at higher
levels of parallelism to converge the system and will ultimately
degrade overall scalability of the system with respect to total wall
time to converge.

MCSA preconditioned with ILUT was used to solve the fuel assembly
problem presented in the previous section for a 1-group $SP_1$
discretization. Unlike the Jacobi preconditioned strategy, convergence
was achieved with ILUT preconditioning. To study convergence
sensitivity to ILUT parameters, the fill level and drop tolerance were
parametrically varied with the number of iterations to converge an
eigenvalue iteration for the fuel assembly problem recorded along with
the maximum number of non-zero entries observed in all matrix rows for
the left/right preconditioned composite linear operator. To provide
some sparsity to the factorization, the ILUT drop tolerance was used
to drop elements in the extracted inverse triangular factors. For each
calculation, the number of iterations required to converge reported
was for a single eigenvalue iteration with \sn{3}{4} histories at each
MCSA iteration to compute the Monte Carlo correction using the adjoint
collision estimator. All ILUT calculations reported here were
performed on a single CPU and therefore this data does not take into
account the aforementioned effects of parallel decomposition on the
quality of the ILUT preconditioning.

Figure~\ref{fig:ilut_iterations} gives the number of iterations to
converge the fixed source problem to a tolerance of \sn{1}{8}. As
expected, the higher the fill level chosen for the ILUT factorization,
the fewer iterations are required to converge the problem (only 9 MCSA
iterations were required to converge the problem for the fill level of
3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
the iterations needed to converge were not as sensitive, signaling
that a larger drop tolerance can perhaps be used without a significant
degradation in iterative performance. At smaller fill levels, the
sensitivity to the ILUT drop tolerance is more significant. For all
fill levels used, convergence was not achieved for the fuel assembly
problem with a drop tolerance larger than \sn{1}{-3}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{ilut_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge an
      eigenvalue iteration for the fuel assembly problem with ILUT
      preconditioning as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the iteration behavior for a
      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
      were used.}}
  \label{fig:ilut_iterations}
\end{figure}

Unfortunately, gaining convergence (and excellent iterative
performance) with MCSA for the $SP_N$ fuel assembly problem comes at
an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
of non-zero entries in the composite linear operator generated by
preconditioning as a function ILUT drop tolerance for varying values
of fill level. For the 1-group $SP_1$ discretization, the original
linear operator will contain only a maximum of 7 non-zero entries per
row in the system. As observed in Figure~\ref{fig:ilut_size},
performing the explicit preconditioning yields composite linear
operators with $O(1,000)$ elements in a row for all combinations of
fill level and drop tolerance, over 10\% of the total row size for
this particular problem. This large number of row entries, observed
for a significant fraction of rows in the system, creates several
problems. First, sparsity is completely destroyed with each state in
the system now coupled to over 10\% of the total states in the system
through the composite iteration matrix. This means that Monte Carlo
sampling tables will be large, requiring significant memory to store
them and a substantial overhead in the sampling procedure during
transport. Second, the matrix-matrix multiply operations required to
build the composite operator see significant performance losses due to
the amount of data that must be handled. In parallel, losing sparsity
severely inhibits performance where now parallel operations require
communications among orders of magnitude more processors for
nearest-neighbor type algorithms.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{ilut_size.pdf}
  \end{center}
  \caption{\textbf{Maximum number of non-zero entries observed for all
      rows in the composite linear operator for the fuel assembly
      problem with ILUT preconditioning given as a function of ILUT
      drop tolerance.} \textit{Each colored curve represents the row
      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
      2.5, and 3.0 were used.}}
  \label{fig:ilut_size}
\end{figure}

As a means to assess the quality of the preconditioning, a simple
metric is developed to address these concerns and allow comparison to
future developments. For our studies, our core performance metric will
be iterative performance with the minimum number of MCSA iterations
required to converge the problem desired. For Monte Carlo
calculations, this improved performance is balanced by the creation of
a dense composite system and we seek to reduce the number of non-zero
entries to a minimum value in order to lower the amount of coupling
among states in the system and reduce the amount of memory used
along with the potential latency overhead. Given these two objectives,
the following metric is proposed:
\begin{equation}
  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
    \# of non-zero values})\:,
\end{equation}
where the highest quality preconditioning is one that minimizes this
metric. For the ILUT preconditioned data provided,
Figure~\ref{fig:ilut_quality} provides the computed metric as a
function of ILUT drop tolerance for each fill level used. In general,
the metric is similar to the data observed for the number of
iterations required to converge as the non-zero row entries changed by
a smaller fraction over drop tolerances tested.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{ilut_quality.pdf}
  \end{center}
  \caption{\textbf{ILUT preconditioning quality metric for the fuel
      assembly problem given as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the quality metric behavior
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
      and 3.0 were used.}}
  \label{fig:ilut_quality}
\end{figure}

It should be noted here that although only the behavior of the linear
solve during a single eigenvalue iteration is reported, the behavior
of the linear solver was observed to be consistent throughout the
eigenvalue iterations (25 total eigenvalue iterations were required to
converge the 1-group $SP_1$ problem). We expect this as the linear
operator (which is unchanging in the fuel assembly criticality
problem) dictates the convergence of MCSA. At each iteration the
fission source provided to the linear problem is changing, however, we
expect the same convergence behavior for all source vectors
$\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.

\subsection{Applying the Reduced Domain Approximation}
\label{subsec:spn_prec_rda}
For each preconditioning technique presented convergence was achieved
for the single fuel assembly problem. However, a primary concern is
the number of non-zero states in each row of the system generated by
the explicit preconditioning strategy. In many cases, orders of
magnitude more matrix elements were generated resulting in poor
scalability for domain decomposed algorithms and overall performance
issues for Monte Carlo. As outlined in
\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
approximation may be used as a mechanism to potentially alleviate this
problem by filtering elements of the composite matrix in each row that
fall below a certain threshold value or by maintaining the largest $N$
elements in each row where $N$ is a designated fill level.

For the ILUT, SPAINV, and SPAINV-ILUT preconditioning strategies the
reduced domain approximation will be applied to reduce the density of
the composite linear operator to more manageable levels. For each
preconditioner, the parameters that achieved the best quality metric
results from the previous analysis were used. These correspond to ILUT
parameters of a fill level of 5.0 and a drop tolerance of \sn{1}{-5},
SPAINV parameters of a sparsity level of 4 and a threshold of 0.1 and
SPAINV-ILUT with ILUT parameters of a fill level of 4.0 and a drop
tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level of 1
and a threshold of 1.0. The reduced domain threshold was set to
\sn{1}{-10} in order to eliminate any exceedingly small values from
the matrix (generally this is simply removing non-zero elements within
the floating point tolerance of zero). The reduced domain fill level
was then varied, starting with the largest non-zero entries per row
value observed for each of the preconditioning types in order to
assess its effects relative to the case where no reduced domain
approximation was applied.

Figure~\ref{fig:rda_iterations} gives the number of iterations
required to converge for each preconditioner type as a function of
reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
corresponding quality metric for each data point where the number of
non-zero entries used to compute the metric is equivalent to the
reduced domain fill level. We first note that SPAINV preconditioning
alone is significantly more sensitive to the reduction in domain size
over the ILUT-based methods, although the preconditioner was of
$O(100)$ non-zero entries per row without any approximation
applied. For the ILUT-based methods, performance was significantly
better with convergence achieved in less than 40 MCSA iterations with
only 10 non-zero entries in each row (vs. 7 non-zero entries for the
case with no preconditioning). SPAINV-ILUT iterative performance was
marginally better than ILUT alone for all reduced domain fill
levels. However, it should be noted that the ILUT level of fill used
for the ILUT only calculations was set to 4 for this case while the
SPAINV-ILUT preconditioner used an ILUT level of fill of 5 and
therefore the marginally better performance is more likely a result of
this addition of fill rather than the extra SPAINV
preconditioning. Looking at the quality metric data, we see a nice
power-law reduction in the quality metric as a function of the reduced
domain fill level, achieving 2 orders of magnitude reduction in the
quality metric for the ILUT-based preconditioning methods.

Although applying the reduced domain approximation results in a
successful recovery of sparsity for the Monte Carlo problem while
maintaining good convergence properties, there is still an issue of
forming the composite operator before applying the approximation and
potentially generating the transpose of this operator in the case of
the adjoint Monte Carlo method. Because of this, memory and scaling
issues may still be observed when building the probability and weight
matrices for the Monte Carlo problem. In addition, the expensive
extraction of the inverse of the preconditioning operators for the
explicit scheme creates a significant cost in overall
performance. Future work in this area should be considerate of these
important components of the preconditioned MCSA algorithm.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{rda_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge a
      single eigenvalue iteration for the fuel assembly problem with
      each preconditioning as a function of reduced domain
      approximation fill level.} \textit{The largest fill level for
      each preconditioning presented is that using the parameters that
      gave the best results without the approximation.}}
  \label{fig:rda_iterations}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{rda_quality.pdf}
  \end{center}
  \caption{\textbf{Preconditioning quality metric for the fuel
      assembly problem given as a function of reduced domain
      approximation fill level.} \textit{The largest fill level for
      each preconditioning presented is that using the parameters that
      gave the best results without the approximation.}}
  \label{fig:rda_quality}
\end{figure}

\subsection{MCSA Relaxation Parameters}
\label{subsec:spn_mcsa_relaxation}
As another means of preconditioning (or variance reduction), the
Richardson iteration upon which the Neumann-Ulam method is built may
be implemented with a scalar relaxation parameter that can be adjusted
to improve convergence:
\begin{equation}
  \mathbf{x} = \mathbf{x} + \omega \mathbf{r}\:,
  \label{eq:richardson_relaxation}
\end{equation}
where $\omega$ is the relaxation parameter. This is very similar to
point Jacobi preconditioning where the system is being scaled on the
left by a constant value in all rows. Analogously, these relaxation
parameter techniques can be applied to Monte Carlo to improve
convergence as demonstrated by Dimov \cite{dimov_new_1998}. In this
case, building an iteration matrix from
Eq~(\ref{eq:richardson_relaxation}) gives:
\begin{equation}
  \mathbf{H} = \mathbf{I} - \omega \mathbf{A}\:,
  \label{eq:relaxed_iteration_matrix}
\end{equation}
with the probabilities and weights for the Monte Carlo procedure
appropriately scaled. By inspection, such a scaling is a simple form
of preconditioning on the left where all rows in the system are scaled
by the same scalar parameter. For MCSA, we can stage this scheme with
two separate relaxation parameters; one for the outer Richardson
iteration and one for the inner Monte Carlo solve:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \omega_R \ve{r}^k\:,\\ \ve{r}^{k+1/2}
    = \ve{b} -
    \ve{A}\ve{x}^{k+1/2}\:,\\ \omega_N\ve{A}\delta\ve{x}^{k+1/2} =
    \omega_N\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
    \ve{x}^{k+1/2}\:,\\ \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa_relaxed}
\end{subequations}
where $\omega_R$ is the Richardson iteration relaxation parameter and
$\omega_N$ is the Neumann-Ulam Monte Carlo solve relaxation parameter.

We apply these relaxation parameters to the fuel assembly problem
along with the reduced domain approximation as a means of studying
their effects. For each calculation presented, the number of
iterations required to converge reported was for a single eigenvalue
iteration.  Again, \sn{3}{4} histories are used at each MCSA iteration
to compute the Monte Carlo correction using the adjoint collision
estimator. A reduced domain fill level of 100 is used with a threshold
of \sn{1}{-10} to filter small values. ILUT preconditioning with a
drop tolerance of \sn{1}{-5} and fill level of 5 is used as
well.

Figure~\ref{fig:relax_iters} gives the number of iterations required
to converge the fuel assembly problem for a 1-group SP1 discretization
using varying combinations of the relaxation parameters. First, both
parameters were fixed at the base case of 1 while the other parameter
was varied as shown in the left-hand side plots of
Figure~\ref{fig:relax_iters}. For the Richardson relaxation parameter,
using a value larger than 1 gave better iterative performance up to a
point, effectively providing a stronger extrapolation using the
residual at each iteration. For the Neumann relaxation parameter, a
value of less than 1 is observed to be ideal. Although initially
counter-intuitive, the fact that the correction computed by the Monte
Carlo solver has a stochastic error associated with it means that by
using a Neumann relaxation parameter less than 1, the correction and
its error are effectively dampened to improve iterative
performance. Considering the CPU time required to converge a single
eigenvalue iteration, similar results are also observed for the
relaxation parameters as given by Figure~\ref{fig:relax_time}. For the
base cases given by the plots on the left, it was found the a
Richardson relaxation parameter of 1.1 and a Neumann relaxation
parameter of 0.7 provided the fastest CPU time for convergence. Fixing
each parameter at these new values, the calculations were repeated as
shown for the plots on the right hand sides of both
Figures~\ref{fig:relax_iters} and \ref{fig:relax_time}. For each
repeated timing calculation, it was found that the same combination of
relaxation parameters found in the base cases performed the best
although they did not necessarily have the best iterative performance.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{relax_iters.pdf}
  \end{center}
  \caption{\textbf{Number of iterations to converge a single
      eigenvalue iteration of the fuel assembly problem as a function
      of the relaxation parameters.} \textit{Starting in upper left
      and moving counter-clockwise: Neumann relaxation parameter fixed
      at 1.0, Richardson relaxation parameter fixed at 1.0, Neumann
      relaxation parameter fixed at 0.7, Richardson relaxation
      parameter fixed at 1.1. For each calculation \sn{3}{4}
      stochastic histories were used to compute the MCSA correction at
      each iteration.}}
  \label{fig:relax_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{relax_time.pdf}
  \end{center}
  \caption{\textbf{CPU time in seconds to converge a single eigenvalue
      iteration of the fuel assembly problem as a function of the
      relaxation parameters.} \textit{Starting in upper left and
      moving counter-clockwise: Neumann relaxation parameter fixed at
      1.0, Richardson relaxation parameter fixed at 1.0, Neumann
      relaxation parameter fixed at 0.7, Richardson relaxation
      parameter fixed at 1.1. For each calculation \sn{3}{4}
      stochastic histories were used to compute the MCSA correction at
      each iteration.}}
  \label{fig:relax_time}
\end{figure}

%%---------------------------------------------------------------------------%%
\section{Results}

Through the numerical studies presented in this paper, we have
developed a set of preconditioning techniques that permit MCSA to be
used with the $SP_N$ form of the neutron transport equation and
applied to a difficult fuel assembly criticality problem. Along with
these preconditioning techniques, a set a varying MCSA parameters were
analyzed to find those that yielded the best performance for this
particular problem. In addition, several steps were taken to mitigate
the dense composite linear operators that arise when performing
explicit MCSA preconditioning in order to achieve convergence. Using
these results, we compare MCSA directly to conventional linear solvers
that would be typically used to solve the $SP_N$ form of the transport
equation as presented here. In particular, we will use a pair of
production Krylov solvers with which we will compare direct numerical
results in order to verify MCSA solutions in this section.

For the comparison calculations, all solvers will be preconditioned
with ILUT using a drop tolerance of \sn{1}{-5} and a fill level of
5. For GMRES, no restrictions were placed on the size of the subspace
and therefore no restarts occurred. When the collision estimator was
used with MCSA, \sn{2}{4} stochastic histories were used to compute
the correction for every energy group in the problem (\sn{4}{4} and
\sn{8}{4} histories total for the 2 and 4 group calculations
respectively) corresponding to approximately 1 stochastic history per
DOF. When the expected value estimator was used, \sn{5}{2} stochastic
histories used for each energy group (\sn{1}{3} and \sn{1.5}{3}
histories total for the 2 and 4 group calculations) giving
approximately 1 stochastic history for every 5 DOFs. The reduced
domain approximation was also applied in conjunction with the ILUT
preconditioning using a fill level of 100 and a threshold value of
\sn{1}{-10} to reduce the density of states in the Monte Carlo
problem. For relaxation parameters, all MCSA computations used a
Neumann relaxation parameter of 0.7 while a Richardson relaxation of
1.1 was used with the collision estimator and 1.0 with the expected
value estimator as determined by the previous analysis of the
relaxation parameters. Table~\ref{tab:spn_solver_defs} gives
definitions for the solvers used to generate the results in the
remainder of this section. For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
group 0).

To compare iterative performance, the number of linear solver
iterations required to converge each eigenvalue iteration was recorded
and then averaged over all eigenvalue iterations for each
solver. Figure~\ref{fig:spn_comparison_iterations} gives the average
number of linear solver iterations required to converge a single
eigenvalue iteration of the fuel assembly problem as a function of the
number of energy groups. In general, the iterative performance of all
the solvers tested is comparable with BiCGStab performing the best. We
expect this because not only are the multigroup $SP_N$ equations
elliptic and positive-definite, but they are also nearly symmetric and
therefore we expect the conjugate gradient-based methods to perform
well due to the structure of the projection scheme. It important to
note that the iterative performance of each solver is not a strong
function of the number of energy groups in the problem (and therefore
the problem size). Even more important, although BiCGStab produced the
best iterative results, MCSA did perform better than GMRES in many
cases with the Richardson iteration accelerated with Monte Carlo using
the collision estimator performing best.

For MCSA, we are linearly increasing the number of stochastic
histories used to compute the correction at each iteration as the
number of energy groups are increased and therefore the work performed
for the problem is of $O(N)$. It is also important to note the reduced
domain approximation parameters, particularly the fill level, were
fixed as the number of energy groups was increased. By fixing the fill
level, we are effectively fixing the amount of information contained
in the composite linear operator available for the Monte Carlo
problem. Limiting this information to 100 entries per row in the
preconditioned system did not appear to have a significant effect on
the iterative performance of the solver. For larger numbers of energy
groups (and therefore DOFs) this may be an issue and this number may
need to be increased to maintain iterative performance. Currently,
memory restrictions arising from the explicit preconditioning strategy
prevent larger energy groups to be analyzed using an appropriate
preconditioner for the fuel assembly problem.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{solver_iters.pdf}
  \end{center}
  \caption{\textbf{Average number of iterations required to converge
      each eigenvalue iteration for the fuel assembly problem as a
      function of energy groups.}
    \textit{Table~\ref{tab:spn_solver_defs} gives the description for
      each solver type presented in the legend.}}
  \label{fig:spn_comparison_iterations}
\end{figure}

Although iterative performance for MCSA was comparable to that
observed for production Krylov implementations using exactly the same
preconditioning scheme, timing performance is not as competitive. For
this comparison, the amount of time to perform each linear solver
iteration in every eigenvalue iteration was recorded and then averaged
over all linear solver
iterations. Figure~\ref{fig:spn_comparison_time} gives the average CPU
time per linear solver iteration required to converge the fuel
assembly problem with all linear solver iterations over all eigenvalue
iterations used to compute the average. The production Krylov solver
implementations are approximately an order of magnitude faster than
the MCSA implementations presented here. We expect these results for
several reasons. First, even when the reduced domain approximation is
used there are $O(100)$ elements in each row of the system and each of
those elements must be processed during a transition in the Monte
Carlo random walk sequence. In addition, although the preconditioned
fixed point iterations used in the MCSA sequence do not use the
composite linear operator but rather a sequence of matrix vector
multiplies to achieve the same preconditioning effect, they do use the
explicitly extracted inverse of the preconditioning matrices which
themselves are dense, leading to exceedingly slow computation times
even in the fixed point iteration. For the production Krylov methods,
the ILUT preconditioners are applied to a vector with two triangular
solves, one for each triangular factor. Also, the original linear
operator has $O(10)$ elements in each row of the system, an order of
magnitude less than that used in the reduced domain composite
operator. Second, the implementation presented here has not been
optimized and it is likely that several components of the algorithm
can be implemented in a more desirable time complexity. The
production Krylov solvers presented here have enjoyed decades of
professional development and optimization. However, it is most
important to note here that the CPU timing data presented in
Figure~\ref{fig:spn_comparison_time} shows that all methods have the
same time complexity; their differences are merely the manifestation
of a large time constant due to the effects of the explicit
preconditioning scheme. For transport systems where this is not a
problem, the literature explicitly shows that MCSA can be competitive
with these production Krylov methods using CPU time as a measure
\cite{evans_monte_2012}.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{solver_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per linear solver iteration in
      seconds for the fuel assembly problem as a function of energy
      groups.}  \textit{All linear solver iterations over all
      eigenvalue iterations were used to compute the
      average. Table~\ref{tab:spn_solver_defs} gives the description
      for each solver type presented in the legend.}}
  \label{fig:spn_comparison_time}
\end{figure}

Finally, not considered in Figure~\ref{fig:spn_comparison_time} is the
time required to actually form the explicit inverse preconditioner
matrices and the composite operator through matrix-matrix
multiplication. The timing numbers reported were simply to perform the
MCSA iteration procedure with the composite operator already
formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
presents the CPU times for MCSA convergence with the time to form the
inverse of the preconditioners and the composite linear operator
through matrix-matrix multiplication amortized over all iterations. As
is readily observed, including the costs of these operations increases
the MCSA computation time by another order of magnitude.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{solver_p_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per iteration in seconds for the
      fuel assembly problem as a function of energy groups with
      preconditioning time included for the MCSA methods.}
    \textit{All linear solver iterations over all eigenvalue
      iterations were used to compute the
      average. Table~\ref{tab:spn_solver_defs} gives the description
      for each solver type presented in the legend. Data labeled
      starting with PMCSA is identical to those labeled with MCSA
      except that they additionally include the cost of generating the
      inverse of the preconditioners and composite linear operator.}}
  \label{fig:spn_comparison_prec_time}
\end{figure}

Based on the performance results in this section, MCSA shows promise
as a competitive and perhaps even superior method for solutions to the
neutron transport problem discretized with the $SP_N$
approximation. Not only are the correct answers produced when compared
to production linear solvers, but the iterative performance is
comparable to production Krylov methods with identical preconditioning
that would typically be used every-day calculations. From a CPU timing
perspective, the methods yield the same time complexity as a function
of the number of energy groups in the problem when compared to the
production Krylov methods. Here, a large time constant is generated
due to the explicit preconditioning strategy and the generation of
dense linear operators as a result. To improve these results and put
general MCSA schemes into a performance regime where they are
competitive with Krylov methods for neutron transport problems,
significant research will be required to improve upon the explicit
preconditioning scheme presented here.

%%---------------------------------------------------------------------------%%
\section{Conclusion}

%%---------------------------------------------------------------------------%%
\pagebreak
\bibliographystyle{ieeetr} \bibliography{references}
\end{document}

### 2013-10-28 doc/ttsp2013/ttsp_2013.tex
adding ttsp updates

@@
 \usepackage{amsmath}
+\usepackage{tmath,tmadd}
 \usepackage{algpseudocode}
@@
 %%---------------------------------------------------------------------------%%
-\author{Stuart R. Slattery
+\author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton
 \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
@@
 \date{\today}
-\title{Monte Carlo Synthetic Acceleration for the Simplified $P_N$ Approximation}
+\title{Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation}
 \begin{document}
@@
 
-In this paper, we present Monte Carlo synthetic acceleration as a
-linear solver technique and briefly derive the simplified $P_N$
-($SP_N$) for fixed source and criticality problems. Using the
+In this paper, we present Monte Carlo Synthetic Acceleration (MCSA) as
+a linear solver technique for the simplified $P_N$ ($SP_N$)
+discretization of the Boltzmann neutron transport equation. Using the
 fully-formed linear operator for the transport problem, we explore
-solutions to the $SP_N$ equations with Monte Carlo Synthetic
-Acceleration using a challenging light water reactor fuel assembly
-criticality calculation as the driving problem. Several difficulties
-arise when applying MCSA to these problems that were not observed when
-investigating the simpler transport systems and we apply additional
-algebraic preconditioning techniques to achieve convergence. For
-Jacobi-based preconditioners, convergence with MCSA is difficult if
-not impossible for the ill-conditioned systems generated by light
-water reactor problems with this behavior demonstrated using a simple
-neutron diffusion problem. In order to effectively solve the $SP_N$
-equations for the fuel assembly problem, a suite of preconditioners
-along with a relaxation scheme is developed and studied within the
-context of MCSA. Using these preconditioners, several additional
-issues are observed and alleviated to a certain extent by applying the
-reduced domain approximation. Finally performance is analyzed through
-comparison with those same Krylov solvers in terms of both iterative
-performance and CPU timing using the fuel assembly problem.
+solutions to the $SP_N$ equations with MCSA using a challenging light
+water reactor fuel assembly criticality calculation as the driving
+problem. Several difficulties arise when applying MCSA to these
+problems that were not observed in previous work where the method was
+applied to more opaque thermal radiaton transport systems. In order to
+effectively solve the $SP_N$ equations for the fuel assembly problem,
+several algebraic preconditioners are studied within the context of
+MCSA. For Jacobi-based preconditioners, convergence of the $SP_N$
+system with MCSA is difficult if not impossible due to the lack of
+diagonal dominance observed as a result the scattering generated by
+light water reactor problems. To overcome these difficulties, we
+utilize basic incomplete factorizations as preconditioners to achieve
+convergence. Although the algebraic preconditioning strategy presented
+here will not be practical from a performance implementation
+standpoint, it does indicate the iterative performance of the method
+when the operator spectrum is properly conditioned. Using the
+preconditioners, MCSA performance is analyzed through comparison with
+production Krylov solvers that would typically be used to solve
+transport problems with an $SP_N$ discretization in terms of both
+iterations to converge and CPU timing.
 
@@
 \section{Simplified $P_N$ Approximation}
+I was thinking about dumping the SPN tech note derivation here. Its a
+lot but it would be a good place to formally present the Exnihilo
+discretization.
 
@@
 
-\subsection{Preliminary Jacobi Preconditioned Calculations}
+\subsection{Block Jacobi Preconditioned Results}
 \label{subsec:jacobi_prec_assembly_calc}
-Based on the success of block Jacobi preconditioning with the test
-problem used for the spectral radius parameter study, we use it first
-to solve the $17 \times 17$ fuel assembly problem. A single energy
-group problem was first solved with $SP_1$ discretization, effectively
-giving the one-speed neutron diffusion system for the fuel assembly
-resulting in 20,088 degrees of freedom in the
-problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
-infinity norm as a function of iteration for the MCSA linear solve in
-the first eigenvalue iteration using 25,000 stochastic histories at
-every iteration for the adjoint Neumann-Ulam solve.
+A single energy group problem was first computed with
+$SP_1$ discretization, effectively giving the one-speed neutron
+diffusion system for the fuel assembly resulting in 20,088 degrees of
+freedom in the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives
+the residual infinity norm as a function of iteration for the MCSA
+linear solve in the first eigenvalue iteration using 25,000 stochastic
+histories at every iteration for the adjoint Neumann-Ulam solve.
+
 \begin{figure}[t!]
@@
 \end{figure}
+
 Convergence was not achieved as noted by the rapid rise in the
@@
 \sn{1}{6} histories per iteration exhibited the same divergent
-behavior at a significant computational cost and compute time. Even if
-the problem may be ill-conditioned, we do expect convergence of
-MCSA. To investigate further, a block Jacobi preconditioned Richardson
+behavior at a significant computational cost and compute time. To
+investigate further, a block Jacobi preconditioned Richardson
 iteration was used to solve the same
@@
 linear solve in the first eigenvalue iteration.
+
 \begin{figure}[t!]
@@
 \end{figure}
-Poor converge is observed for the Richardson iteration, however,
+
+Poor convergence is observed for the Richardson iteration, however,
 convergence is achieved meaning that the preliminary eigenvalue
@@
 stochastic histories in the block Jacobi preconditioned Neumann-Ulam
-method are eventually being terminated by the weight cutoff as no
-artificial absorption was used for these preliminary
-calculations. Based on this, the correction being generated by the
-Neumann-Ulam solve is the component of MCSA causing a divergent
-solution.
+method are eventually being terminated by the weight cutoff. Based on
+this, the correction being generated by the Neumann-Ulam solve is the
+component of MCSA causing a divergent solution which we study next.
 
@@
 \label{subsec:mcsa_break_down}
-For the fuel assembly problem, the initial block Jacobi
-preconditioned\footnote{For 1-group $SP_1$ problems the block size is
-  1, giving a point Jacobi preconditioner.}  calculations used a
-one-speed $SP_1$ discretization, effectively giving a diffusion
-system. To study the breakdown of MCSA at iteration matrix spectral
-radii near one, we will use the simpler homogeneous 2-dimensional
-one-speed neutron diffusion system presented in
+To study the breakdown of MCSA at iteration matrix spectral radii near
+one, we will use the simpler homogeneous 2-dimensional one-speed
+neutron diffusion system presented in
 Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
@@
 solver with the collision estimator. For spectral radii less than
-0.97, all MCSA problems converged with 1 history per DOF (10,000 for
-this problem) with the number of iterations required to converge
-increasing as a function of spectral radius. Near a spectral radius of
-0.97, the number of histories required to converge MCSA in less than
-100 iterations takes a dramatic rise that exhibits neither exponential
-nor power law behavior. As the spectral radius approaches 1, the
-number of histories required becomes significant and effectively
-impractical to compute. Even with this simple diffusion problem, the
-behavior is consistent with that observed for the fuel assembly
-problem with $SP_1$ discretization. In that case, we estimated a
-spectral radius of $\approx 0.998$, larger than any of the spectral
-radii that could be computed within even 90 minutes of compute time
-for this simple two dimensional problem. For that problem, even
-1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
-convergence. Even if single solve times of an hour can be tolerated,
-dozens of solves are typically required to compute the multiplication
-factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
-difficult problems like the fuel assembly, making the method unusable.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{breakdown_iterations.pdf}
+0.97, all MCSA problems converged with 1 history per DOF (10,000 total
+histories for this problem) with the number of iterations required to
+converge increasing as a function of spectral radius. Near a spectral
+radius of 0.97, the number of histories required to converge MCSA in
+less than 100 iterations takes a dramatic rise that exhibits neither
+exponential nor power law behavior. As the spectral radius approaches
+1, the number of histories required becomes significant and
+effectively impractical to compute. Even with this simple diffusion
+problem, the behavior is consistent with that observed for the fuel
+assembly problem with $SP_1$ discretization. In that case, we
+estimated a spectral radius of $\approx 0.998$, larger than any of the
+spectral radii that could be computed within even 90 minutes of
+compute time for this simple two dimensional problem. For that large
+of a spectral radius, even 1,000,000 histories ($\approx 50$ per DOF)
+was not enough to provide convergence. Even if single solve times of
+an hour can be tolerated, dozens of solves are typically required to
+compute the multiplication factor and flux spectrum within the $SP_N$
+eigenvalue scheme for more difficult problems like the fuel assembly,
+making the method unusable.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4.5in]{breakdown_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{breakdown_histories.pdf}
+    \includegraphics[width=4.5in]{breakdown_histories.pdf}
   \end{center}
@@
 problems with spectral radii above $\approx 0.97$, using MCSA to solve
-any problems of interest is entirely ineffective and not practical. We
-therefore require a more expansive set of preconditioning techniques
-to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
-which MCSA is more applicable and in which performance is improved.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{breakdown_time.pdf}
+any problems of interest is entirely ineffective and not practical. At
+iteration matrix spectral radii this large, using significant numbers
+of stochastic histories in the Monte Carlo solve is not enough to
+reduce the stochastic uncertainty in the correction, causing the MCSA
+sequence to diverge instead of accelerating convergence. We therefore
+require a more expansive set of preconditioning techniques to move the
+eigenvalue spectrum of the $SP_N$ problem into a regime in which MCSA
+is more applicable and in which performance is improved.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4.5in]{breakdown_time.pdf}
   \end{center}
@@
 
-\subsection{Algebraic Preconditioning Strategies}
+\clearpage
+
+%%---------------------------------------------------------------------------%%
+\section{Advanced Preconditioning Strategies}
 \label{subsec:spn_advanced_preconditioning}
@@
 difficulty, advanced preconditioning strategies for the $SP_N$ problem
-are required beyond simple Jacobi methods that can reduce the spectral
-radius into a region of better MCSA performance. Several modern
-algebraic preconditioning strategies will be presented here and used
-within the explicit preconditioning framework given by
-Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
-solutions of the fuel assembly criticality problem will be presented.
+are required, beyond simple Jacobi methods, that can reduce the
+spectral radius into a region of better MCSA performance. Two modern
+algebraic preconditioning strategies, ILUT and SPAINV, will be
+presented here and used within the explicit preconditioning framework
+given by Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on
+MCSA solutions of the fuel assembly criticality problem will be
+provided.
 
@@
 \label{subsec:spn_ilut_preconditioning}
+ILUT preconditioning is chosen first due to its applicability to
+problems that have a dominating elliptic character such as the fuel
+assembly problem or other problems with a large scattering to
+absorption ratio \cite{saad_iterative_2003}. The algebraic formulation
+of ILUT requires the physics operator to be fully formed, also a
+requirement for MCSA, and is therefore easily incorporated into our
+solution sequence.
+
 Incomplete lower-upper (ILU) factorizations of the linear operator can
-be used a simple mechanism to form an approximate inverse of a
+be used as a simple mechanism to form an approximate inverse of a
 preconditioner. To build the factorization, the sparse upper and lower
@@
 \end{equation}
-has a specified sparsity pattern and element magnitude. When a
-magnitude threshold is used, elements generated in the factors below
+has a specified sparsity pattern and element magnitude. When a drop
+tolerance threshold is used, elements generated in the factors below
 that magnitude are dropped, resulting in ILU threshold (ILUT)
@@
 lower and upper triangular factors may then be easily inverted by
-means of simple elimination to produce the preconditioner. For MCSA,
-$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
-the right to precondition the system.
-
-For modern subspace methods, only the action to the preconditioner on
-a vector is required for efficient implementations and therefore a
-triangular solve can be used for this purpose. For the explicit
-preconditioning scheme presented in Eq~(\ref{eq:left_right_mcsa}), the
-fully inverted operator must be generated in order to build the set of
-probabilities and weights required for Monte Carlo
-sampling. Therefore, for a linear system of size $N$, $N$ triangular
-solves will be required in order to extract the inverse matrices from
-production ILU implementations. In addition, parallel implementations
-typically generate lower and upper factors that are only triangular
-locally, providing an easily parallelizable mechanism to generate the
-action of the preconditioner inverse. A consequence of this choice for
-parallel scalability is a degradation of the preconditioning quality
-as the size of the parallel system is increased. For serial
-computations, the triangular factorization is potentially exact
-depending on the parameters chosen while at thousands of parallel
-tasks, the global triangular factors differ significantly from the true
-factorization. As a result, more iterations are required at higher
-levels of parallelism to converge the system and will ultimately
-degrade overall scalability of the system with respect to total wall
-time to converge.
+means of simple elimination procedure to produce the
+preconditioner. For MCSA, $\mathbf{L}^{-1}$ will be used on the left
+and $\mathbf{U}^{-1}$ on the right to precondition the system.
+
+For the explicit preconditioning scheme presented in
+Eq~(\ref{eq:left_right_mcsa}), the fully inverted preconditioners must
+be generated in order to build the set of probabilities and weights
+required for Monte Carlo sampling. Therefore, for a linear system of
+size $N$, $N$ triangular solves with the basis of each row in the
+system as a solution vector will be required in order to extract the
+inverse matrices from production ILU implementations. In addition,
+parallel implementations typically generate lower and upper factors
+that are only triangular locally, providing an easily parallelizable
+mechanism to generate the action of the preconditioner inverse. A
+consequence of this choice for parallel scalability is a degradation
+of the preconditioning quality as the size of the parallel system is
+increased. For serial computations, the triangular factorization is
+potentially exact depending on the parameters chosen while at
+thousands of parallel tasks, the global triangular factors differ
+significantly from the true factorization. As a result, more
+iterations are required at higher levels of parallelism to converge
+the system and will ultimately degrade overall scalability of the
+system with respect to total wall time to converge.
 
@@
 parametrically varied with the number of iterations to converge an
-eigenvalue iteration for the fuel assembly problem recorded along with
-the maximum number of non-zero entries observed in all matrix rows for
-the left/right preconditioned composite linear operator. To provide
-some sparsity to the factorization, the ILUT drop tolerance was used
-to drop elements in the extracted inverse triangular factors. For each
-calculation, the number of iterations required to converge reported
-was for a single eigenvalue iteration with \sn{3}{4} histories at each
-MCSA iteration to compute the Monte Carlo correction using the adjoint
-collision estimator. All ILUT calculations reported here were
-performed on a single CPU and therefore this data does not take into
-account the aforementioned effects of parallel decomposition on the
-quality of the ILUT preconditioning.
+eigenvalue iteration for the fuel assembly problem recorded. In
+addition, the maximum number of non-zero entries observed in all rows
+for the left/right preconditioned composite linear operator was
+recorded to observe the effects of the explicit preconditioner
+inversion and matrix-matrix multiplication on sparsity. To provide
+some sparsity to the factorization, an additional step was added with
+the ILUT drop tolerance used to drop elements in the explicitly
+inverted preconditioners. For each calculation, the number of
+iterations reported was for a single eigenvalue iteration with
+\sn{3}{4} histories at each MCSA iteration to compute the Monte Carlo
+correction using the adjoint collision estimator. All ILUT
+calculations reported here were performed on a single CPU and
+therefore this data does not take into account the aforementioned
+effects of parallel decomposition on the quality of the ILUT
+preconditioning.
 
 Figure~\ref{fig:ilut_iterations} gives the number of iterations to
-converge the fixed source problem to a tolerance of \sn{1}{8}. As
+converge the fixed source problem to a tolerance of \sn{1}{-8}. As
 expected, the higher the fill level chosen for the ILUT factorization,
-the fewer iterations are required to converge the problem (only 9 MCSA
-iterations were required to converge the problem for the fill level of
-3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
-the iterations needed to converge were not as sensitive, signaling
-that a larger drop tolerance can perhaps be used without a significant
-degradation in iterative performance. At smaller fill levels, the
-sensitivity to the ILUT drop tolerance is more significant. For all
-fill levels used, convergence was not achieved for the fuel assembly
-problem with a drop tolerance larger than \sn{1}{-3}.
+the fewer iterations are required to converge the problem with only 9
+MCSA iterations were required to converge the problem for the fill
+level of 3 and drop tolerance of \sn{1}{-5} case. For higher levels of
+fill, the iterations needed to converge were not as sensitive,
+signaling that a larger drop tolerance can perhaps be used without a
+significant degradation in iterative performance. At smaller fill
+levels, the sensitivity to the ILUT drop tolerance is more
+significant. For all fill levels used, convergence was not achieved
+for the fuel assembly problem with a drop tolerance larger than
+\sn{1}{-3}.
 \begin{figure}[t!]
@@
 nearest-neighbor type algorithms.
+
 \begin{figure}[t!]
@@
 entries to a minimum value in order to lower the amount of coupling
-among states in the system and reduce the amount of memory used
-along with the potential latency overhead. Given these two objectives,
-the following metric is proposed:
+among states in the system and reduce the amount of memory used along
+with the potential latency overhead. Given these two objectives, the
+following metric is proposed:
 \begin{equation}
   \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
-    \# of non-zero values})\:,
+    \# non-zero values})\:,
 \end{equation}
 where the highest quality preconditioning is one that minimizes this
-metric. For the ILUT preconditioned data provided,
+metric. This metric gives equal importance to both iterative
+performance and sparsity with the purpose of simply measuring their
+combined effect rather than seeking some optimal relationship of
+importance. For the ILUT preconditioned data provided,
 Figure~\ref{fig:ilut_quality} provides the computed metric as a
@@
 iterations required to converge as the non-zero row entries changed by
-a smaller fraction over drop tolerances tested.
+a smaller fraction over drop tolerances tested. For higher levels of
+fill, the quality metric is observed to be relatively insensitive to
+the drop tolerance as maintaining more entries in the preconditioner
+has a stronger impact than filtering small values.
+
 \begin{figure}[t!]
@@
 
-It should be noted here that although only the behavior of the linear
-solve during a single eigenvalue iteration is reported, the behavior
-of the linear solver was observed to be consistent throughout the
-eigenvalue iterations (25 total eigenvalue iterations were required to
-converge the 1-group $SP_1$ problem). We expect this as the linear
-operator (which is unchanging in the fuel assembly criticality
-problem) dictates the convergence of MCSA. At each iteration the
-fission source provided to the linear problem is changing, however, we
-expect the same convergence behavior for all source vectors
-$\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.
-
-\subsection{Applying the Reduced Domain Approximation}
+\subsection{Sparse Approximate Inverse Preconditioning}
+\label{subsec:spn_spainv_preconditioning}
+Sparse approximate inverse (SPAINV) preconditioning is chosen next
+because the true inverse of the preconditioner is the product of the
+algorithm, avoiding the need for many inverse matrix applications as
+required by using ILUT with MCSA. In addition, the algorithm seeks to
+maintain the sparsity of the inverse preconditioner, potentially
+alleviating some of the dense operator issues observed with ILUT.
+
+SPAINV produces a sparse approximation of the inverse matrix directly
+by minimizing the Frobenius norm of the residual matrix
+\cite{saad_iterative_2003}:
+\begin{equation}
+  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
+  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
+  \label{eq:frobenius_norm_min}
+\end{equation}
+where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
+norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
+$j^{th}$ columns of the identity and preconditioning matrices
+respectively. Here, $\mathbf{M}$ is the actual inverse of the
+preconditioner and is formed directly by the algorithm. On the right
+hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
+represented as a effective sum of $N$ linear system residual norms.
+
+A few iterations of a subspace method or other iterative scheme can be
+used to effectively solve these systems to a loose tolerance and yield
+the columns of the approximate preconditioning matrix. At each
+iteration, threshold values are used to eliminate the components of
+each $\mathbf{m}_j$ column to maintain the desired level of
+sparsity. In addition, a sparsity pattern can be predetermined and
+enforced during this process. For the SPAINV implementation used for
+this work, the sparsity pattern was defined by levels such that a
+level of $N$ for an input operator $\mathbf{A}$ will generate a
+preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
+$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
+a factorization such as ILUT, parallel results are reproduced
+regardless of parallel problem size, meaning that a serial computation
+will converge in the same number of iterations as a computation with
+thousands of cores.
+
+MCSA preconditioned with SPAINV was used to solve the fuel assembly
+problem again for a 1-group $SP_1$ discretization. To study
+convergence sensitivity to SPAINV parameters, the number of levels in
+the sparsity pattern and threshold were parametrically varied with the
+number of iterations to converge a single eigenvalue iteration for the
+fuel assembly problem recorded along with the maximum number of
+non-zero entries observed in all matrix rows for the preconditioned
+composite linear operator. For each calculation, the number of
+iterations required to converge reported was for a single eigenvalue
+iteration with \sn{3}{4} histories at each MCSA iteration to compute
+the Monte Carlo correction using the adjoint collision estimator. All
+SPAINV calculations reported here were performed on a single CPU.
+
+Figure~\ref{fig:spainv_iterations} gives the number of MCSA iterations
+needed to converge the fuel assembly problem and
+Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
+per row observed in the composite linear operator as a function the
+SPAINV threshold. For this analysis, the number of levels in the
+sparsity pattern was varied from 3 to 7 with convergence of the fuel
+assembly problem not achieved for smaller level values. Compared to
+ILUT preconditioning, at higher levels we see comparable iterative
+performance with a relative invariance to the threshold value. The
+threshold did have a significant effect, however, on the time required
+to generate the preconditioner with a threshold of 0.001 over an order
+of magnitude slower than a threshold of 0.01 due to the inclusion of a
+significant number of extra values in the input operator,
+$\mathbf{A}$. In addition to comparable iterative performance, the
+sparsity of the composite operator is greatly improved with nearly an
+order of magnitude reduction in number of non-zero entries per row for
+lower level values.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4.25in]{spainv_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Number of MCSA iterations required to converge a
+      single eigenvalue iteration for the fuel assembly problem with
+      SPAINV preconditioning as a function of SPAINV threshold.}
+    \textit{Each colored curve represents the iteration behavior for a
+      different SPAINV level pattern. Levels of 3, 4, 5, 6, and 7 were
+      used.}}
+  \label{fig:spainv_iterations}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4.25in]{spainv_size.pdf}
+  \end{center}
+  \caption{\textbf{Maximum number of non-zero entries observed for all
+      rows in the composite linear operator for the fuel assembly
+      problem with SPAINV preconditioning given as a function of
+      SPAINV threshold.} \textit{Each colored curve represents the row
+      size for a different SPAINV level pattern. Levels of 3, 4, 5, 6,
+      and 7 were used.}}
+  \label{fig:spainv_size}
+\end{figure}
+
+Figure~\ref{fig:spainv_quality} gives the quality metric as a function
+of the threshold value for each of the sparsity levels computed. Not
+only do we see an improved quality metric over ILUT preconditioning
+(about an order of magnitude less), but we also note that the ideal
+sparsity level is not the largest nor the smallest. In fact, the
+largest and smallest levels (3 and 7) performed the worst in terms of
+the quality metric with a level of 4 performing the best.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4.25in]{spainv_quality.pdf}
+  \end{center}
+  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
+      assembly problem given as a function of SPAINV threshold.}
+    \textit{Each colored curve represents the quality metric behavior
+      for a different SPAINV level pattern. Levels of 3, 4, 5, 6, and
+      7 were used.}}
+  \label{fig:spainv_quality}
+\end{figure}
+
+\clearpage
+
+\subsection{The Reduced Domain Approximation}
 \label{subsec:spn_prec_rda}
-For each preconditioning technique presented convergence was achieved
-for the single fuel assembly problem. However, a primary concern is
-the number of non-zero states in each row of the system generated by
-the explicit preconditioning strategy. In many cases, orders of
-magnitude more matrix elements were generated resulting in poor
-scalability for domain decomposed algorithms and overall performance
-issues for Monte Carlo. As outlined in
-\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
-approximation may be used as a mechanism to potentially alleviate this
-problem by filtering elements of the composite matrix in each row that
-fall below a certain threshold value or by maintaining the largest $N$
-elements in each row where $N$ is a designated fill level.
-
-For the ILUT, SPAINV, and SPAINV-ILUT preconditioning strategies the
-reduced domain approximation will be applied to reduce the density of
-the composite linear operator to more manageable levels. For each
-preconditioner, the parameters that achieved the best quality metric
-results from the previous analysis were used. These correspond to ILUT
-parameters of a fill level of 5.0 and a drop tolerance of \sn{1}{-5},
-SPAINV parameters of a sparsity level of 4 and a threshold of 0.1 and
-SPAINV-ILUT with ILUT parameters of a fill level of 4.0 and a drop
-tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level of 1
-and a threshold of 1.0. The reduced domain threshold was set to
+Although ILUT and SPAINV preconditioning permitted MCSA solutions for
+the fuel assembly problem, a primary concern is the number of non-zero
+states in each row of the system generated by the explicit
+preconditioning strategy. In many cases, orders of magnitude more
+matrix elements were generated resulting in poor scalability for
+domain decomposed algorithms and overall performance issues for Monte
+Carlo. To alleviate this, we propose modifying the domain sampled by
+the Monte Carlo method to generate the correction within MCSA by
+removing terms in the iteration matrix through some predetermined
+criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
+stochastic realization of multiple matrix-vector multiplies, the
+criteria we choose should be based on maintaining those elements that
+will make the largest contributions to the multiplication and
+therefore recover as much of the original product as possible.
+
+For each row in the preconditioned system, we can then apply the
+reduced domain approximation to eliminate those elements that either
+fall below a specified threshold level, fall within a fill level $N$
+where we keep the $N$ largest elements in each row, or both giving:
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
+    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
+    \label{eq:rda_correction}\\
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
+  \end{gather}
+  \label{eq:rda_mcsa}
+\end{subequations}
+where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
+reduced domain approximation has been applied to the iteration matrix
+in the Monte Carlo method. If enough of the terms in the original
+iteration matrix have been maintained, the character of the original
+iteration matrix should be retained and continue to accelerate the
+iterative scheme.
+
+We next apply the reduced domain approximation to both ILUT and SPAINV
+preconditioned system to reduce the density of the composite linear
+operator to more manageable levels and study the resulting iterative
+performance. For each preconditioner, the parameters that achieved
+the best quality metric results from the previous analysis were
+used. These correspond to ILUT parameters of a fill level of 5.0 and a
+drop tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level
+of 4 and a threshold of 0.1. The reduced domain threshold was set to
 \sn{1}{-10} in order to eliminate any exceedingly small values from
-the matrix (generally this is simply removing non-zero elements within
-the floating point tolerance of zero). The reduced domain fill level
-was then varied, starting with the largest non-zero entries per row
-value observed for each of the preconditioning types in order to
-assess its effects relative to the case where no reduced domain
-approximation was applied.
+the matrix (often this is simply removing non-zero elements within the
+floating point tolerance of zero). The reduced domain fill level was
+then varied, starting with the largest non-zero entries per row value
+observed for each of the preconditioning types in order to assess its
+effects relative to the case where no reduced domain approximation was
+applied.
 
@@
 only 10 non-zero entries in each row (vs. 7 non-zero entries for the
-case with no preconditioning). SPAINV-ILUT iterative performance was
-marginally better than ILUT alone for all reduced domain fill
-levels. However, it should be noted that the ILUT level of fill used
-for the ILUT only calculations was set to 4 for this case while the
-SPAINV-ILUT preconditioner used an ILUT level of fill of 5 and
-therefore the marginally better performance is more likely a result of
-this addition of fill rather than the extra SPAINV
-preconditioning. Looking at the quality metric data, we see a nice
-power-law reduction in the quality metric as a function of the reduced
+case with no preconditioning). Looking at the quality metric data, we
+see a reduction in the quality metric as a function of the reduced
 domain fill level, achieving 2 orders of magnitude reduction in the
-quality metric for the ILUT-based preconditioning methods.
-
-Although applying the reduced domain approximation results in a
-successful recovery of sparsity for the Monte Carlo problem while
-maintaining good convergence properties, there is still an issue of
-forming the composite operator before applying the approximation and
-potentially generating the transpose of this operator in the case of
-the adjoint Monte Carlo method. Because of this, memory and scaling
-issues may still be observed when building the probability and weight
-matrices for the Monte Carlo problem. In addition, the expensive
-extraction of the inverse of the preconditioning operators for the
-explicit scheme creates a significant cost in overall
-performance. Future work in this area should be considerate of these
-important components of the preconditioned MCSA algorithm.
+quality metric for the ILUT preconditioning.
 
@@
 
-\subsection{MCSA Relaxation Parameters}
-\label{subsec:spn_mcsa_relaxation}
-As another means of preconditioning (or variance reduction), the
-Richardson iteration upon which the Neumann-Ulam method is built may
-be implemented with a scalar relaxation parameter that can be adjusted
-to improve convergence:
-\begin{equation}
-  \mathbf{x} = \mathbf{x} + \omega \mathbf{r}\:,
-  \label{eq:richardson_relaxation}
-\end{equation}
-where $\omega$ is the relaxation parameter. This is very similar to
-point Jacobi preconditioning where the system is being scaled on the
-left by a constant value in all rows. Analogously, these relaxation
-parameter techniques can be applied to Monte Carlo to improve
-convergence as demonstrated by Dimov \cite{dimov_new_1998}. In this
-case, building an iteration matrix from
-Eq~(\ref{eq:richardson_relaxation}) gives:
-\begin{equation}
-  \mathbf{H} = \mathbf{I} - \omega \mathbf{A}\:,
-  \label{eq:relaxed_iteration_matrix}
-\end{equation}
-with the probabilities and weights for the Monte Carlo procedure
-appropriately scaled. By inspection, such a scaling is a simple form
-of preconditioning on the left where all rows in the system are scaled
-by the same scalar parameter. For MCSA, we can stage this scheme with
-two separate relaxation parameters; one for the outer Richardson
-iteration and one for the inner Monte Carlo solve:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \omega_R \ve{r}^k\:,\\ \ve{r}^{k+1/2}
-    = \ve{b} -
-    \ve{A}\ve{x}^{k+1/2}\:,\\ \omega_N\ve{A}\delta\ve{x}^{k+1/2} =
-    \omega_N\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
-    \ve{x}^{k+1/2}\:,\\ \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:mcsa_relaxed}
-\end{subequations}
-where $\omega_R$ is the Richardson iteration relaxation parameter and
-$\omega_N$ is the Neumann-Ulam Monte Carlo solve relaxation parameter.
-
-We apply these relaxation parameters to the fuel assembly problem
-along with the reduced domain approximation as a means of studying
-their effects. For each calculation presented, the number of
-iterations required to converge reported was for a single eigenvalue
-iteration.  Again, \sn{3}{4} histories are used at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. A reduced domain fill level of 100 is used with a threshold
-of \sn{1}{-10} to filter small values. ILUT preconditioning with a
-drop tolerance of \sn{1}{-5} and fill level of 5 is used as
-well.
-
-Figure~\ref{fig:relax_iters} gives the number of iterations required
-to converge the fuel assembly problem for a 1-group SP1 discretization
-using varying combinations of the relaxation parameters. First, both
-parameters were fixed at the base case of 1 while the other parameter
-was varied as shown in the left-hand side plots of
-Figure~\ref{fig:relax_iters}. For the Richardson relaxation parameter,
-using a value larger than 1 gave better iterative performance up to a
-point, effectively providing a stronger extrapolation using the
-residual at each iteration. For the Neumann relaxation parameter, a
-value of less than 1 is observed to be ideal. Although initially
-counter-intuitive, the fact that the correction computed by the Monte
-Carlo solver has a stochastic error associated with it means that by
-using a Neumann relaxation parameter less than 1, the correction and
-its error are effectively dampened to improve iterative
-performance. Considering the CPU time required to converge a single
-eigenvalue iteration, similar results are also observed for the
-relaxation parameters as given by Figure~\ref{fig:relax_time}. For the
-base cases given by the plots on the left, it was found the a
-Richardson relaxation parameter of 1.1 and a Neumann relaxation
-parameter of 0.7 provided the fastest CPU time for convergence. Fixing
-each parameter at these new values, the calculations were repeated as
-shown for the plots on the right hand sides of both
-Figures~\ref{fig:relax_iters} and \ref{fig:relax_time}. For each
-repeated timing calculation, it was found that the same combination of
-relaxation parameters found in the base cases performed the best
-although they did not necessarily have the best iterative performance.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{relax_iters.pdf}
-  \end{center}
-  \caption{\textbf{Number of iterations to converge a single
-      eigenvalue iteration of the fuel assembly problem as a function
-      of the relaxation parameters.} \textit{Starting in upper left
-      and moving counter-clockwise: Neumann relaxation parameter fixed
-      at 1.0, Richardson relaxation parameter fixed at 1.0, Neumann
-      relaxation parameter fixed at 0.7, Richardson relaxation
-      parameter fixed at 1.1. For each calculation \sn{3}{4}
-      stochastic histories were used to compute the MCSA correction at
-      each iteration.}}
-  \label{fig:relax_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{relax_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time in seconds to converge a single eigenvalue
-      iteration of the fuel assembly problem as a function of the
-      relaxation parameters.} \textit{Starting in upper left and
-      moving counter-clockwise: Neumann relaxation parameter fixed at
-      1.0, Richardson relaxation parameter fixed at 1.0, Neumann
-      relaxation parameter fixed at 0.7, Richardson relaxation
-      parameter fixed at 1.1. For each calculation \sn{3}{4}
-      stochastic histories were used to compute the MCSA correction at
-      each iteration.}}
-  \label{fig:relax_time}
-\end{figure}
+Although applying the reduced domain approximation results in a
+successful recovery of sparsity for the Monte Carlo problem while
+maintaining good convergence properties with ILUT, there is still an
+issue of forming the composite operator before applying the
+approximation and potentially generating the transpose of this
+operator in the case of the adjoint Monte Carlo method. Because of
+this, memory and scaling issues may still be observed when building
+the probability and weight matrices for the Monte Carlo problem. In
+addition, the expensive extraction of the inverse of the
+preconditioning operators for the explicit scheme creates a
+significant cost in overall performance. Future work in this area
+should be considerate of these important components of the
+preconditioned MCSA algorithm.
 
 %%---------------------------------------------------------------------------%%
-\section{Results}
-
-Through the numerical studies presented in this paper, we have
+\section{MCSA Verification}
+\label{sec:spn_mcsa_verification}
+Through the numerical studies presented in this chapter, we have
 developed a set of preconditioning techniques that permit MCSA to be
 used with the $SP_N$ form of the neutron transport equation and
-applied to a difficult fuel assembly criticality problem. Along with
-these preconditioning techniques, a set a varying MCSA parameters were
-analyzed to find those that yielded the best performance for this
-particular problem. In addition, several steps were taken to mitigate
-the dense composite linear operators that arise when performing
-explicit MCSA preconditioning in order to achieve convergence. Using
-these results, we compare MCSA directly to conventional linear solvers
-that would be typically used to solve the $SP_N$ form of the transport
-equation as presented here. In particular, we will use a pair of
-production Krylov solvers with which we will compare direct numerical
-results in order to verify MCSA solutions in this section.
-
-For the comparison calculations, all solvers will be preconditioned
-with ILUT using a drop tolerance of \sn{1}{-5} and a fill level of
-5. For GMRES, no restrictions were placed on the size of the subspace
-and therefore no restarts occurred. When the collision estimator was
-used with MCSA, \sn{2}{4} stochastic histories were used to compute
-the correction for every energy group in the problem (\sn{4}{4} and
-\sn{8}{4} histories total for the 2 and 4 group calculations
-respectively) corresponding to approximately 1 stochastic history per
-DOF. When the expected value estimator was used, \sn{5}{2} stochastic
-histories used for each energy group (\sn{1}{3} and \sn{1.5}{3}
-histories total for the 2 and 4 group calculations) giving
-approximately 1 stochastic history for every 5 DOFs. The reduced
-domain approximation was also applied in conjunction with the ILUT
-preconditioning using a fill level of 100 and a threshold value of
-\sn{1}{-10} to reduce the density of states in the Monte Carlo
-problem. For relaxation parameters, all MCSA computations used a
-Neumann relaxation parameter of 0.7 while a Richardson relaxation of
-1.1 was used with the collision estimator and 1.0 with the expected
-value estimator as determined by the previous analysis of the
-relaxation parameters. Table~\ref{tab:spn_solver_defs} gives
-definitions for the solvers used to generate the results in the
-remainder of this section. For the energy group structures,
-Table~\ref{tab:spn_group_structure} gives the lower bounds of each
-group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
-group 0).
-
+applied to a difficult fuel assembly criticality problem. In addition,
+several steps were taken to mitigate the dense composite linear
+operators that arise when performing explicit MCSA preconditioning
+necessary for achieve convergence. Using these results, we now compare
+MCSA to conventional Krylov solvers that would be typically used to
+solve the $SP_N$ form of the transport equation as a means of
+verification.
+
+MCSA solutions to the fuel assembly problem with 1, 2, and 4 energy
+groups using $SP_1$ discretization will computed using both BiCGStab
+and GMRES implementations from the Trilinos scientific computing
+library Aztec \cite{heroux_overview_2005}. The neutron fluxes, the
+number of eigenvalue iterations required to converge the problem, and
+the eigenvalue itself will each be evaluated. For the flux values, we
+will look at the 2-norm and infinity norm of the relative difference
+vector between the reference case and the comparison case in each
+energy group:
+\begin{equation}
+  \text{Relative Group Flux Difference} = \frac{|\ve{\Phi}_{ref}^g -
+    \ve{\Phi}^g|}{|\ve{\Phi}_{ref}^g|}\:,
+  \label{eq:relative_difference_vector}
+\end{equation}
+where the absolute value, subtraction, and division operations are
+performed element-wise to construct the vector. For this work, the
+BiCGStab results will be taken as the reference case.
+
+All solvers will be preconditioned with ILUT using a drop tolerance of
+\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
+placed on the size of the subspace and therefore no restarts
+occurred. When the collision estimator was used with MCSA, \sn{2}{4}
+stochastic histories were used to compute the correction for every
+energy group in the problem (\sn{4}{4} and \sn{8}{4} histories total
+for the 2 and 4 group calculations respectively) corresponding to
+approximately 1 stochastic history per DOF. When the expected value
+estimator was used, \sn{5}{2} stochastic histories used for each
+energy group (\sn{1}{3} and \sn{1.5}{3} histories total for the 2 and
+4 group calculations) giving approximately 1 stochastic history for
+every 5 DOFs. The reduced domain approximation was also applied in
+conjunction with the ILUT preconditioning using a fill level of 100
+and a threshold value of \sn{1}{-10} to reduce the density of states
+in the Monte Carlo problem. For relaxation parameters, all MCSA
+computations used a Neumann relaxation parameter of 0.7 while a
+Richardson relaxation of 1.1 was used with the collision estimator and
+1.0 with the expected value estimator. When the collision estimator
+was used, an artificial absorption value of 0.2 as outlined in
+Appendix~\ref{chap:artificial_absorption} was also
+applied. Additionally, both the Richardson-based MCSA iteration given
+by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration
+given by Eq~(\ref{eq:mcsa_min_res}) will be used. A supplementary
+performance analysis for this set of MCSA variations is presented in
+Appendix~\ref{chap:spn_estimator_comparison}. Table~\ref{tab:spn_solver_defs}
+gives definitions for the solvers used to generate the results in the
+remainder of this section and Table~\ref{tab:spn_group_structure}
+gives the group structures with lower bounds of each group in eV.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{ll}\hline\hline
+      \multicolumn{1}{l}{\textbf{Name}} & 
+      \multicolumn{1}{l}{\textbf{Definition}} \\
+      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
+      GMRES-ILUT & GMRES preconditioned with ILUT \\
+      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
+                    & fixed point iteration and collision estimator \\
+      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using minimal residual \\
+                     & fixed point iteration and collision estimator \\
+      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
+                     & fixed point iteration and expected value estimator \\
+      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using minimal residual \\
+                      & fixed point iteration and expected value estimator \\
+      Richardson-ILUT & Richardson iteration preconditioned with ILUT \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Solver definitions used for MCSA verification and
+      performance analysis.} \textit{The ILUT preconditioner
+      parameters were identical for all calculations and solvers.}}
+  \label{tab:spn_solver_defs}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cl}\hline\hline
+      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
+      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
+      1 & \{ \sn{1}{-5} \} \\
+      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
+      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Energy group lower bounds for the multigroup fuel
+      assembly criticality problem in electron volts.} \textit{An
+      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
+  \label{tab:spn_group_structure}
+\end{table}
+
+For every combination of energy group structure and solver, the
+eigenvalue problem was converged with an eigensolver tolerance of
+\sn{1}{-6} and a linear solver tolerance of \sn{1}{-8} in a serial
+calculation. Table~\ref{tab:serial_ev_results} gives the eigenvalue
+computed by each solver for each problem as well as the number of
+eigenvalue iterations required to converge. For every problem, each
+solver converged to the same eigenvalue up to the number of digits
+reported by the eigensolver in the same number of eigenvalue
+iterations.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{c}{\textbf{Solver}} & 
+      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
+      \multicolumn{1}{c}{\textbf{Iterations}} &
+      \multicolumn{1}{c}{\textbf{Eigenvalue}} \\
+      \hline
+      BiCGStab-ILUT & 1 & 25 & 1.105987 \\
+      GMRES-ILUT & 1 & 25 & 1.105987 \\
+      MCSA-ILUT-R-C & 1 & 25 & 1.105987 \\
+      MCSA-ILUT-MR-C & 1 & 25 & 1.105987 \\
+      MCSA-ILUT-R-EV & 1 & 25 & 1.105987 \\
+      MCSA-ILUT-MR-EV & 1 & 25 & 1.105987 \\
+      Richardson-ILUT & 1 & 25 & 1.105987 \\
+      \hline
+      BiCGStab-ILUT & 2 & 35 & 1.155239 \\
+      GMRES-ILUT & 2 & 35 & 1.155239 \\
+      MCSA-ILUT-R-C & 2 & 35 & 1.155239 \\
+      MCSA-ILUT-MR-C & 2 & 35 & 1.155239 \\
+      MCSA-ILUT-R-EV & 2 & 35 & 1.155239 \\
+      MCSA-ILUT-MR-EV & 2 & 35 & 1.155239 \\
+      Richardson-ILUT & 2 & 35 & 1.155239 \\
+      \hline
+      BiCGStab-ILUT & 4 & 31 & 1.175850 \\
+      GMRES-ILUT & 4 & 31 & 1.175850 \\
+      MCSA-ILUT-R-C & 4 & 31 & 1.175850 \\
+      MCSA-ILUT-MR-C & 4 & 31 & 1.175850 \\
+      MCSA-ILUT-R-EV & 4 & 31 & 1.175850 \\
+      MCSA-ILUT-MR-EV & 4 & 31 & 1.175850 \\
+      Richardson-ILUT & 4 & 31 & 1.175850 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Serial eigensolver verification results for the
+      fuel assembly problem.} \textit{Each calculation was converged
+      with the same eigensolver and linear solver
+      tolerances. Table~\ref{tab:spn_solver_defs} gives the
+      description for each solver type presented.}}
+  \label{tab:serial_ev_results}
+\end{table}
+
+Table~\ref{tab:serial_differences_g1},
+Table~\ref{tab:serial_differences_g2}, and
+Table~\ref{tab:serial_differences_g4} give the infinity norms and
+2-norms of the flux relative difference vectors given by
+Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
+energy groups. In general, we see good agreement in the 2-norm between
+the MCSA variants and the GMRES calculations, meaning that they have
+approximately the same variation from the BiCGStab calculations. We do
+note larger infinity norm values of this vector for some of the MCSA
+variants relative to the GMRES results (i.e. group 1 in the 2-group
+calculation), however, this large maximum value is apparently a local
+phenomenon with good agreement in the 2-norm for the
+calculations. Based on the good agreement between MCSA and GMRES for
+the relative difference of the group fluxes and the eigenvalue results
+presented here, the MCSA implementation should be deemed correct for
+the fuel assembly problem for serial calculations with multiple energy
+groups.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{c}{\textbf{Solver}} & 
+      \multicolumn{1}{c}{\textbf{Group}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{\infty}$}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
+      \hline
+      GMRES-ILUT & 0 & \sn{4.562}{-6} & \sn{6.746}{-5} \\
+      MCSA-ILUT-R-C & 0 & \sn{2.099}{-5} & \sn{2.541}{-4} \\
+      MCSA-ILUT-MR-C & 0 & \sn{4.812}{-5} & \sn{4.672}{-4} \\
+      MCSA-ILUT-R-EV & 0 & \sn{2.371}{-5} & \sn{2.740}{-4} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{3.605}{-5} & \sn{2.554}{-4} \\
+      Richardson-ILUT & 0 & \sn{1.531}{-5} & \sn{2.771}{-4} \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Serial scalar flux verification results for the
+      1-group fuel assembly problem.} \textit{The infinity norm and
+      2-norm of the relative flux vector in each group is given with
+      BiCGStab used as the reference.}}
+  \label{tab:serial_differences_g1}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{c}{\textbf{Solver}} & 
+      \multicolumn{1}{c}{\textbf{Group}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{\infty}$}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
+      \hline
+      GMRES-ILUT & 0 & \sn{7.590}{-6} & \sn{1.234}{-4} \\
+      MCSA-ILUT-R-C & 0 & \sn{5.308}{-6} & \sn{1.066}{-4} \\
+      MCSA-ILUT-MR-C & 0 & \sn{6.388}{-5} & \sn{2.233}{-4} \\
+      MCSA-ILUT-R-EV & 0 & \sn{4.197}{-5} & \sn{2.168}{-4} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{2.736}{-5} & \sn{8.160}{-5} \\
+      Richardson-ILUT & 0 & \sn{3.976}{-6} & \sn{7.059}{-5} \\
+      \hline
+      GMRES-ILUT & 1 & \sn{7.579}{-6} & \sn{1.246}{-4} \\
+      MCSA-ILUT-R-C & 1 & \sn{1.645}{-5} & \sn{1.210}{-4} \\
+      MCSA-ILUT-MR-C & 1 & \sn{6.635}{-4} & \sn{9.435}{-4} \\
+      MCSA-ILUT-R-EV & 1 & \sn{1.797}{-4} & \sn{2.983}{-4} \\
+      MCSA-ILUT-MR-EV & 1 & \sn{1.556}{-4} & \sn{2.055}{-4} \\
+      Richardson-ILUT & 1 & \sn{4.031}{-6} & \sn{7.110}{-5} \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Serial scalar flux verification results for the
+      2-group fuel assembly problem.} \textit{The infinity norm and
+      2-norm of the relative flux vector in each group is given with
+      BiCGStab used as the reference.}}
+  \label{tab:serial_differences_g2}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{c}{\textbf{Solver}} & 
+      \multicolumn{1}{c}{\textbf{Group}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{\infty}$}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
+      \hline
+      GMRES-ILUT & 0 & \sn{2.376}{-5} & \sn{4.043}{-4} \\
+      MCSA-ILUT-R-C & 0 & \sn{2.639}{-5} & \sn{3.950}{-4} \\
+      MCSA-ILUT-MR-C & 0 & \sn{3.958}{-5} & \sn{4.635}{-4} \\
+      MCSA-ILUT-R-EV & 0 & \sn{2.749}{-5} & \sn{3.975}{-4} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{2.969}{-5} & \sn{4.054}{-4} \\
+      Richardson-ILUT & 0 & \sn{2.189}{-5} & \sn{3.791}{-4} \\
+      \hline
+      GMRES-ILUT & 1 & \sn{2.271}{-5} & \sn{3.952}{-4} \\
+      MCSA-ILUT-R-C & 1 & \sn{3.701}{-5} & \sn{3.933}{-4} \\
+      MCSA-ILUT-MR-C & 1 & \sn{3.803}{-5} & \sn{3.933}{-4} \\
+      MCSA-ILUT-R-EV & 1 & \sn{4.142}{-5} & \sn{4.011}{-4} \\
+      MCSA-ILUT-MR-EV & 1 & \sn{2.723}{-5} & \sn{4.030}{-4} \\
+      Richardson-ILUT & 1 & \sn{2.137}{-5} & \sn{3.768}{-4} \\
+      \hline
+      GMRES-ILUT & 2 & \sn{2.310}{-5} & \sn{4.021}{-4} \\
+      MCSA-ILUT-R-C & 2 & \sn{2.490}{-5} & \sn{3.956}{-4} \\
+      MCSA-ILUT-MR-C & 2 & \sn{2.214}{-5} & \sn{3.608}{-4} \\
+      MCSA-ILUT-R-EV & 2 & \sn{4.456}{-5} & \sn{3.833}{-4} \\
+      MCSA-ILUT-MR-EV & 2 & \sn{3.366}{-5} & \sn{3.797}{-4} \\
+      Richardson-ILUT & 2 & \sn{2.143}{-5} & \sn{3.767}{-4} \\
+      \hline
+      GMRES-ILUT & 3 & \sn{2.335}{-5} & \sn{4.029}{-4} \\
+      MCSA-ILUT-R-C & 3 & \sn{3.045}{-5} & \sn{3.963}{-4} \\
+      MCSA-ILUT-MR-C & 3 & \sn{2.696}{-5} & \sn{3.606}{-4} \\
+      MCSA-ILUT-R-EV & 3 & \sn{1.947}{-5} & \sn{4.283}{-4} \\
+      MCSA-ILUT-MR-EV & 3 & \sn{3.295}{-5} & \sn{3.837}{-4} \\
+      Richardson-ILUT & 3 & \sn{2.161}{-5} & \sn{3.772}{-4} \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Serial scalar flux verification results for the
+      4-group fuel assembly problem.} \textit{The infinity norm and
+      2-norm of the relative flux vector in each group is given with
+      BiCGStab used as the reference.}}
+  \label{tab:serial_differences_g4}
+\end{table}
+
+\clearpage
+
+%%---------------------------------------------------------------------------%%
+\section{MCSA Performance Comparison to Conventional Methods}
+\label{sec:spn_comparison}
+With the same set of solver parameters used for verification
+calculation, MCSA will now be compared to the same Krylov solvers in
+terms of both iterative performance and CPU timing. A Richardson
+iteration solution will also be used in order to assess the
+acceleration provided by the residual Monte Carlo component of MCSA.
 To compare iterative performance, the number of linear solver
@@
 and then averaged over all eigenvalue iterations for each
-solver. Figure~\ref{fig:spn_comparison_iterations} gives the average
-number of linear solver iterations required to converge a single
-eigenvalue iteration of the fuel assembly problem as a function of the
-number of energy groups. In general, the iterative performance of all
-the solvers tested is comparable with BiCGStab performing the best. We
-expect this because not only are the multigroup $SP_N$ equations
-elliptic and positive-definite, but they are also nearly symmetric and
-therefore we expect the conjugate gradient-based methods to perform
-well due to the structure of the projection scheme. It important to
-note that the iterative performance of each solver is not a strong
-function of the number of energy groups in the problem (and therefore
-the problem size). Even more important, although BiCGStab produced the
-best iterative results, MCSA did perform better than GMRES in many
-cases with the Richardson iteration accelerated with Monte Carlo using
-the collision estimator performing best.
-
-For MCSA, we are linearly increasing the number of stochastic
-histories used to compute the correction at each iteration as the
-number of energy groups are increased and therefore the work performed
-for the problem is of $O(N)$. It is also important to note the reduced
-domain approximation parameters, particularly the fill level, were
-fixed as the number of energy groups was increased. By fixing the fill
-level, we are effectively fixing the amount of information contained
-in the composite linear operator available for the Monte Carlo
+solver.
+
+Table~\ref{tab:spn_comparison_iterations} gives this average number of
+linear solver iterations required to converge a single eigenvalue
+iteration of the fuel assembly problem as a function of the number of
+energy groups. In general, the iterative performance of all the
+solvers tested aside from the Richardson iteration is comparable with
+BiCGStab performing the best. We expect this because not only are the
+multigroup $SP_N$ equations positive-definite, but they are also
+block-symmetric and therefore we expect the conjugate gradient-based
+methods to perform well due to the structure of the projection
+scheme. It is important to note that the iterative performance of each
+solver is not a strong function of the number of energy groups in the
+problem (and therefore the problem size). Even more important,
+although BiCGStab produced the best iterative results, MCSA did
+perform better than GMRES in many cases with the Richardson iteration
+clearly accelerated by the Monte Carlo solve. When compared to the
+preconditioned Richardson iteration, the acceleration provided by MCSA
+is clear with 3-4 times fewer iterations required for convergence.
+
+It is also important to note the reduced domain approximation
+parameters, particularly the fill level, were fixed as the number of
+energy groups was increased. By fixing the fill level, we are
+effectively fixing the amount of information contained in the
+composite linear operator available for the Monte Carlo
 problem. Limiting this information to 100 entries per row in the
@@
 memory restrictions arising from the explicit preconditioning strategy
-prevent larger energy groups to be analyzed using an appropriate
-preconditioner for the fuel assembly problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{solver_iters.pdf}
-  \end{center}
-  \caption{\textbf{Average number of iterations required to converge
-      each eigenvalue iteration for the fuel assembly problem as a
-      function of energy groups.}
-    \textit{Table~\ref{tab:spn_solver_defs} gives the description for
-      each solver type presented in the legend.}}
-  \label{fig:spn_comparison_iterations}
-\end{figure}
+prevent larger numbers of energy groups to be analyzed using an
+appropriate preconditioner for the fuel assembly problem.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{l}{Solver}&
+      \multicolumn{1}{c}{1 Group}&
+      \multicolumn{1}{c}{2 Groups}&
+      \multicolumn{1}{c}{4 Groups}\\
+      \hline
+      BiCGStab-ILUT & 11.6 & 11.6 & 12.4 \\ 
+      GMRES-ILUT & 18.1 & 17.9 & 18.9 \\
+      MCSA-ILUT-R-C & 14.6 & 15.4 & 17.6 \\
+      MCSA-ILUT-MR-C & 16.0 & 17.1 & 23.7 \\
+      MCSA-ILUT-R-EV & 18.3 & 19.4 & 16.8 \\
+      MCSA-ILUT-MR-EV & 19.6 & 22.4 & 17.5 \\
+      Richardson-ILUT & 60.9 & 60.4 & 63.4 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Average number of linear solver iterations
+      iterations required to converge the fixed source $SP_N$ problem
+      at each eigenvalue iteration for the fuel assembly problem for
+      1, 2 and 4 energy groups.} \textit{Values are rounded to the
+      first decimal place. Table~\ref{tab:spn_solver_defs} gives the
+      description for each solver type presented.}}
+  \label{tab:spn_comparison_iterations}
+\end{table}
 
 Although iterative performance for MCSA was comparable to that
-observed for production Krylov implementations using exactly the same
-preconditioning scheme, timing performance is not as competitive. For
-this comparison, the amount of time to perform each linear solver
-iteration in every eigenvalue iteration was recorded and then averaged
-over all linear solver
-iterations. Figure~\ref{fig:spn_comparison_time} gives the average CPU
-time per linear solver iteration required to converge the fuel
+observed for Krylov implementations using exactly the same
+preconditioning scheme, timing performance is not as
+competitive. Figure~\ref{fig:spn_comparison_time} gives the average
+CPU time per linear solver iteration required to converge the fuel
 assembly problem with all linear solver iterations over all eigenvalue
-iterations used to compute the average. The production Krylov solver
+iterations used to compute the average. The Krylov solver
 implementations are approximately an order of magnitude faster than
@@
 used there are $O(100)$ elements in each row of the system and each of
-those elements must be processed during a transition in the Monte
-Carlo random walk sequence. In addition, although the preconditioned
-fixed point iterations used in the MCSA sequence do not use the
-composite linear operator but rather a sequence of matrix vector
-multiplies to achieve the same preconditioning effect, they do use the
-explicitly extracted inverse of the preconditioning matrices which
-themselves are dense, leading to exceedingly slow computation times
-even in the fixed point iteration. For the production Krylov methods,
-the ILUT preconditioners are applied to a vector with two triangular
-solves, one for each triangular factor. Also, the original linear
-operator has $O(10)$ elements in each row of the system, an order of
-magnitude less than that used in the reduced domain composite
-operator. Second, the implementation presented here has not been
-optimized and it is likely that several components of the algorithm
-can be implemented in a more desirable time complexity. The
-production Krylov solvers presented here have enjoyed decades of
-professional development and optimization. However, it is most
-important to note here that the CPU timing data presented in
-Figure~\ref{fig:spn_comparison_time} shows that all methods have the
-same time complexity; their differences are merely the manifestation
-of a large time constant due to the effects of the explicit
-preconditioning scheme. For transport systems where this is not a
-problem, the literature explicitly shows that MCSA can be competitive
-with these production Krylov methods using CPU time as a measure
-\cite{evans_monte_2012}.
+those elements must be accessed for next-event sampling and tally
+vector modification during a transition in the Monte Carlo random walk
+sequence. In addition, although the preconditioned fixed point
+iterations used in the MCSA sequence do not use the composite linear
+operator but rather a sequence of matrix vector multiplies to achieve
+the same preconditioning effect, they do use the explicitly extracted
+inverse of the preconditioning matrices which themselves are dense,
+leading to exceedingly slow computation times even in the fixed point
+iteration.
+
+For the Krylov methods, the ILUT preconditioners are applied to a
+vector with two triangular solves, one for each triangular
+factor. Also, the original linear operator has $O(10)$ elements in
+each row of the system, an order of magnitude less than that used in
+the reduced domain composite operator. Second, the implementation
+presented here has not been optimized and it is likely that several
+components of the algorithm can be implemented in a more desirable
+time complexity. However, it is most important to note here that the
+CPU timing data presented in Figure~\ref{fig:spn_comparison_time}
+shows that all methods have the same time complexity; their
+differences are merely the manifestation of a large time constant due
+to the effects of the explicit preconditioning scheme. For transport
+systems where this is not a problem, the literature explicitly shows
+that MCSA can be competitive with these production Krylov methods
+using CPU time as a measure \cite{evans_monte_2012}.
 
@@
       average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend.}}
+      for each solver type presented in the legend with the Krylov
+      solvers represented in black and the MCSA solvers represented in
+      red.}}
   \label{fig:spn_comparison_time}
@@
       for each solver type presented in the legend. Data labeled
-      starting with PMCSA is identical to those labeled with MCSA
-      except that they additionally include the cost of generating the
-      inverse of the preconditioners and composite linear operator.}}
+      starting with PMCSA and shown in blue is identical to those
+      labeled with MCSA except that they additionally include the cost
+      of generating the inverse of the preconditioners and composite
+      linear operator.}}
   \label{fig:spn_comparison_prec_time}
@@
 Based on the performance results in this section, MCSA shows promise
-as a competitive and perhaps even superior method for solutions to the
-neutron transport problem discretized with the $SP_N$
-approximation. Not only are the correct answers produced when compared
-to production linear solvers, but the iterative performance is
-comparable to production Krylov methods with identical preconditioning
-that would typically be used every-day calculations. From a CPU timing
-perspective, the methods yield the same time complexity as a function
-of the number of energy groups in the problem when compared to the
-production Krylov methods. Here, a large time constant is generated
-due to the explicit preconditioning strategy and the generation of
-dense linear operators as a result. To improve these results and put
-general MCSA schemes into a performance regime where they are
-competitive with Krylov methods for neutron transport problems,
-significant research will be required to improve upon the explicit
-preconditioning scheme presented here.
+as a competitive method for solutions to the neutron transport problem
+discretized with the $SP_N$ approximation and could potentially be
+superior if the preconditioning strategy were improved to reduce time
+to solution and memory usage. Not only are the correct answers
+produced when compared to production linear solvers, but the iterative
+performance is comparable to Krylov methods with identical
+preconditioning that would typically be used most calculations. From a
+CPU timing perspective, the methods yield the same time complexity as
+a function of the number of energy groups in the problem when compared
+to the Krylov methods. Here, a large time constant is generated due to
+the explicit preconditioning strategy and the generation of dense
+linear operators as a result. To improve these results and put general
+MCSA schemes into a performance regime where they are competitive with
+Krylov methods for neutron transport problems, significant research
+will be required to improve upon the explicit preconditioning scheme
+presented here.
 
@@
 \section{Conclusion}
+\label{sec:conclusion}
+
+In this paper, the Monte Carlo Synthetic Acceleration method has
+been applied to the $SP_N$ discretization of the neutron transport
+equation and explored in the context of a difficult nuclear fuel
+assembly criticality problem. The following are the significant
+observations, findings, and contributions.
+
+\begin{itemize}
+\item MCSA can solve the asymmetric system generated by the $SP_N$
+  equations
+\item MCSA has been incorporated into the Exnihilo neutronics
+  production code base developed at Oak Ridge National Laboratory
+\item Light water reactor problems are difficult to solve with MCSA as
+  they have large spectral radii due to the neutron scattering in the
+  moderator
+\item Simple Jacobi-based preconditioning reduces the spectral radius
+  of the $SP_N$ system below unity, however, this alone is not
+  sufficient for MCSA convergence
+\item As the spectral radius approaches unity, MCSA was demonstrated
+  to break down with more stochastic histories required for
+  convergence and more CPU time required per history
+\item Advanced algebraic preconditioning strategies were applied to
+  the $SP_N$ equations to obtain convergence with ILUT chosen for
+  subsequent investigations
+\item The reduced domain approximation was introduced to mitigate the
+  memory and performance constraints of the explicit preconditioning
+  strategy although both remain significant obstacles
+\item MCSA was verified to produce the same k-eigenvalue for the fuel
+  assembly as production Krylov methods and gave good relative flux
+  agreement
+\item MCSA was observed to converge in fewer iterations per eigenvalue
+  iteration than GMRES for the fuel assembly criticality problem and
+  more than Bi-CGStab using the same preconditioning
+\item The explicit preconditioning strategy required to overcome the
+  MCSA spectral radius restriction creates run times $O(100)$ larger
+  than the production Krylov methods for the fuel assembly criticality
+  problem
+\end{itemize}
 

### 2014-01-11 doc/ttsp2013/ttsp_2013.tex
TTSP 2014 paper: *: Refactoring the introduction *: Cleaning out some of the preconditioning results *: Removing the solver verification results

@@
 
-\date{\today}
-\title{Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation}
+\date{\today} \title{A Preconditioned Monte Carlo
+  Synthetic-Acceleration Method for the Simplified $P_N$
+  Approximation}
 \begin{document}
@@
 water reactor fuel assembly criticality calculation as the driving
-problem. Several difficulties arise when applying MCSA to these
-problems that were not observed in previous work where the method was
-applied to more opaque thermal radiaton transport systems. In order to
-effectively solve the $SP_N$ equations for the fuel assembly problem,
-several algebraic preconditioners are studied within the context of
-MCSA. For Jacobi-based preconditioners, convergence of the $SP_N$
-system with MCSA is difficult if not impossible due to the lack of
-diagonal dominance observed as a result the scattering generated by
-light water reactor problems. To overcome these difficulties, we
-utilize basic incomplete factorizations as preconditioners to achieve
-convergence. Although the algebraic preconditioning strategy presented
-here will not be practical from a performance implementation
-standpoint, it does indicate the iterative performance of the method
-when the operator spectrum is properly conditioned. Using the
-preconditioners, MCSA performance is analyzed through comparison with
-production Krylov solvers that would typically be used to solve
-transport problems with an $SP_N$ discretization in terms of both
-iterations to converge and CPU timing.
+problem. Due to neutron scattering in the light water moderator,
+several difficulties arise when applying MCSA to these problems that
+were not observed in previous work where the method was applied to
+more opaque thermal radiaton transport systems. In order to
+effectively solve the $SP_N$ equations for the model problem, an
+algebraic preconditioning strategy is developed for MCSA where we
+utilize incomplete factorizations as preconditioners to achieve
+convergence.  Using the preconditioners, MCSA performance is compared
+with production impelementations of the preconditioned GMRES and
+BiCGSTAB methods. Although the algebraic preconditioning strategy
+presented in this work will not be practical from a high performance
+implementation standpoint with results demonstrating poor time to
+solution, it does indicate that the iterative performance of the
+method is competitive with the preconditioned Krylov methods when
+operator spectrum is properly conditioned.
 
@@
 
-The neutron transport problem is complicated. Solutions cover a large
-phase space and the problems of interest are often geometrically
-complex, very large, or both, requiring tremendous computational
-resources to generate an adequate solution. Modern deterministic
-methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \cite{evans_denovo:_2010}. For fission
-reactor neutronics simulations, the $S_N$ method requires potentially
-trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the responses of interest
-\cite{slaybaugh_acceleration_2011}. Other forms of the transport
-problem, including the $P_N$ method, take on a simpler form than the
-more common $S_N$ methods but lack in accuracy when compared while
-still requiring considerable computational resources for solutions in
-multiple dimensions.
-
-In the 1960's, Gelbard developed an ad-hoc multidimensional extension
-of the simple single dimension planar $P_N$ equations that created a
-system of coupled, diffusion-like equations known as the simplified
-$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
-method was either widely unknown, widely unused, or combination of
-both even though numerical studies showed promising results with
-better solutions than diffusion theory and a significant reduction in
-computational time over more accurate methods such as discrete
-ordinates. Why did this happen? A significant problem, pointed out by
-Brantley and Larsen \cite{brantley_simplified_2000}, was that little
-rigor had been applied to the formulation of the $SP_N$ equations
-since their derivation through primarily heuristic arguments. Instead,
-studies at that time focused on simply comparing the results of the
-method to other contemporary transport solution strategies. In
-addition, many problems of interest from the literature at the time
-were either solved using nodal-type methods for reactor-sized problems
-or $S_N$-type methods for benchmark problems with intricate material
-configurations and potentially large flux gradients over small spatial
-domains.
-
-So why reconsider the $SP_N$ equations? Starting in the 1990's and
-primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment with both variational and
-asymptotic derivations performed as a means of verification. In
-addition, these equations have been more rigorously studied as
-solution methods to MOX fuel problems and have been shown to provide
-accurate solutions. With this mathematical literature to provide a
-solid numerical footing for the method, we look at its application to
-today's challenge problems in neutron transport for fission reactor
-analysis. The reduction in numerical complexity of current
-deterministic solution methods using the $S_N$ approximation could
-mean significant savings in both compute time and memory required. In
-addition, the characteristics of the solution to the transport problem
-for a steady state reactor core permit diffusion theory to be used; a
-staple of the nuclear industry since its inception. Therefore, if
-diffusion theory is applicable, then finer grained solutions that
-capture more of the physics contained in the transport equation should
-be possible with the $SP_N$ method. In doing so, we also expect from
-the literature to obtain computed responses on the order of accuracy
-we would expect from an appropriately discretized $S_N$ method at a
-fraction of the cost.
+Predictive modeling and simulation capability requires the combination
+of high fidelity models, high performance computing hardware that can
+handle the intense computational loads required by these models, and
+modern algorithms for solving these problems that leverage this high
+performance hardware. For nuclear reactor analysis, this predictive
+capability can enable tighter design tolerances for improved thermal
+performance and efficiency, higher fuel burn-up and therefore
+reduction in generated waste, and high confidence in accident scenario
+models. Modern high fidelity deterministic methods for large scale
+problems are often variants on the discrete ordinates ($S_N$) method
+\cite{evans_denovo:_2010}. For fission reactor neutronics simulations,
+the $S_N$ method requires potentially trillions of unknown angular
+flux moments to be computed to achieve good accuracy for the responses
+of interest \cite{slaybaugh_acceleration_2011}.
+
+As machines begin to operate at hundreds of petaflops peak performance
+and beyond, trends toward reduced energy consumption will require
+incredibly high levels of concurrency to achieve the desired
+computation rates. The end result of these hardware changes is that
+the larger number of low-powered processors will be prone to both soft
+failures such as bit errors in floating point operations and hard
+failures where the data owned by that processor cannot be
+recovered. Because these failures are predicted to be common,
+resilient solver technologies are required to overcome these events at
+the application level. With solutions to deterministic transport
+problems based on Monte Carlo techniques, such issues are potentially
+alleviated by statistical arguments. In the case of soft failures,
+isolated floating point errors in the Monte Carlo simulation are
+absorbed within tally statistics while a hard failure is treated as a
+high variance event where some portion of the Monte Carlo histories
+lost.
+
+Considering these physics and hardware issues, this goal of this work
+is to research and develop Monte Carlo Synthetic Acceleration methods
+for neutron transport problems. These methods are based on using the
+Monte Carlo method to directly accelerate the convergence of
+traditional fixed-point iterative schemes for these problems. We aim
+to address both their benefits and their shortcomings in the context
+of the physics of interest and desire to identify areas in which
+improvements can be made. Using these numerical studies, we look
+forward to parallelizing them with the goal that they may be leveraged
+competitively on leadership-class computing platforms.
+
+In previous work \cite{evans_monte_2012}, we applied the Monte Carlo
+synthetic acceleration (MCSA) method the thermal radiation diffusion
+equation. For several transient problems we observed that MCSA
+performed competetively with preconditioned Krylov methods. In a
+continuation of these studies, we have applied the method light water
+reactor problems discretized by the $SP_N$ equations. This system has
+no time dependence, a large scattering component, and assymetry in the
+linear operator, creating a set of problems complementary to the
+transient, opaque and symmetric systems of our previous analyis.
 
@@
 technique for neutron transport problems, its current formulation
-requires the physics operator and all preconditioners to be explicitly
+requires the linear operator and all preconditioners to be explicitly
 formed. Recent developments in the Exnihilo neutronics package at Oak
@@
 system of equations for detailed neutronics models of fission
-reactor-based systems\cite{evans_simplified_2013}. By fully forming these
-equations and formulating them as a linear algebra problem instead of
-using the explicit iterative methods of the past, we now have access
-to all of the modern advancements in computational linear algebra
-including Krylov solvers for asymmetric systems and algebraic
-preconditioning methods.
-
-This leads us to then research the applicability of Monte Carlo
-Synthetic Acceleration methods as a possible solution method for the
-$SP_N$ equations. In addition, solving the $SP_N$ equations in this
-way also breaks away from the $S_N$ forms of parallelism where spatial
-parallelism is achieved by an efficient parallel sweep, angular
-efficiency achieved by pipe-lining, and energy parallelism achieved by
-decoupling the groups. With the $SP_N$ equations as a full matrix
-system, we now can parallelize the problem as prescribed by the Monte
-Carlo algorithm, which may be significantly more scalable than current
-$S_N$ transport practices. In this work we therefore apply Monte Carlo
-Synthetic Acceleration to the $SP_N$ equations and research its
-applicability, performance, and convergence properties in the context
-of a light water reactor system.
+reactor-based systems\cite{evans_simplified_2013}. By fully forming
+these equations and formulating them as a linear algebra problem
+instead of using an explicit iterative method, we now have access to
+modern advancements in computational linear algebra including Krylov
+solvers for asymmetric systems and algebraic preconditioning methods.
+
+The paper is organized as follows. First, we briefly introduce the
+model $SP_N$ problem and describe the application of the MCSA
+algorithm. Next, we demonstrate that the basic Jacobi preconditioning
+applied in our previous work is unsatisfactory to achieve convergence
+for neutron transport in light water reactors and subsquently develop
+a new preconditioning scheme. We then apply this preconditioning
+scheme to a light water reactor fuel assembly and compare the
+performance of the preconditioned MCSA method to preconditioned Krylov
+methods.
 
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic Acceleration}
-
+\section{Simplified $P_N$ Approximation}
+For criticality problems, the space, angle, and energy discretizations
+for the $SP_N$ equations can be applied to the general eigenvalue form
+of the transport equation given by
+Eq~(\ref{eq:eigenvalue_transport}):
+\begin{multline}
+  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
+    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
+    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
+    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
+    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
+    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
+  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
+  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
+  \label{eq:multigroup_spn_eigenvalue}
+\end{multline}
+with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
+fission matrix, $\mathbf{F}$. We apply the following change of
+variables to yield the set of multigroup pseudo-moments
+$\mathbb{U}_n$:
+\begin{subequations}
+  \begin{gather}
+    \mathbb{U}_1 = \mathbf{\Phi}_0 + 2\mathbf{\Phi}_2 \\
+    \mathbb{U}_2 = 3\mathbf{\Phi}_2 + 4\mathbf{\Phi}_4 \\
+    \mathbb{U}_3 = 5\mathbf{\Phi}_4 + 6\mathbf{\Phi}_6 \\
+    \mathbb{U}_4 = 7\mathbf{\Phi}_6 \:.
+  \end{gather}
+  \label{eq:spn7_subs}
+\end{subequations}
+where now the application of the fission matrix to the moment vectors
+will be expanded into a set of block matrices in identical fashion to
+the scattering matrices:
+\begin{equation}
+  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
+  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
+  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
+  \label{eq:spn_fission_matrix}
+\end{equation}
+To complete the spatial discretization, the gradient operators are
+resolved by a finite volume scheme as in
+\cite{evans_simplified_2013}. It should be noted here that with
+respect to a general MCSA scheme, the introduction of fission in the
+system will only affect the source vector in the linear system for the
+eigenvalue schemes used in this work. The linear operator, and
+therefore overall MCSA performance will be dictated by streaming and
+scattering as defined on the left-hand side.
+
+\subsection{MCSA Solution of the $SP_N$ Equations}
+\label{subsec:mcsa_solution}
 Using the ideas of Halton, Evans and Mosher recently developed a Monte
@@
 and CPU time. Their approach was instead to use residual Monte Carlo
-as a synthetic acceleration for a stationary method. To derive this
-method, we begin by splitting the operator in
-Eq~(\ref{eq:linear_problem})
-\begin{equation}
-  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
-  \label{eq:linear_split}
-\end{equation}
-With this we can then define the stationary method
-\textit{Richardson's iteration} as:
-\begin{equation}
-  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
-  \label{eq:richardsons_iteration}
-\end{equation}
-which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
-solution error at the $k^{th}$ iterate relative to the true solution:
-\begin{equation}
-  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
-  \label{eq:mcsa_error}
-\end{equation}
-Subtracting Eq~(\ref{eq:richardsons_iteration}) from
-Eq~(\ref{eq:linear_split}) we get:
-\begin{equation}
-  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
-  \label{eq:mcsa_setup_1}
-\end{equation}
-Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
-\begin{equation}
-  \begin{split}
-    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
-    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
-    \label{eq:mcsa_setup_2}
-  \end{split}
-\end{equation}
-Using this, we define the following scheme that will converge in one
-iteration if $\ve{A}$ is inverted exactly:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
-    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
-  \end{gather}
-  \label{eq:mcsa_setup_3}
-\end{subequations}
-However, $\ve{A}$ is only approximately inverted by our numerical
-methods and therefore we instead pose an iterative scheme in which the
-Monte Carlo solvers are used to invert the operator. The
+as a synthetic acceleration for a stationary method. The
 \textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
@@
 number of histories are required using the Neumann-Ulam method to
-generate the correction. In addition, the fact that the scheme in
-Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
-is inverted exactly means that as more and more stochastic histories
-are used to compute the correction and the error is reduced towards
-zero, the number of iterations required for MCSA to converge should
-decrease accordingly, thus accelerating the solution.
-
-In addition to the Monte Carlo solver parameters dictating the number
-of histories and weight cutoff, the outer MCSA iterations also have
-the following stopping criteria:
-\begin{equation}
-  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
-  \label{eq:mcsa_stopping_criteria}
-\end{equation}
-where $\epsilon$ is a user-defined parameter. As with any iterative
-method, other stopping criteria using other vector norms could be
-computed, however, for this work we will only use
-Eq~(\ref{eq:mcsa_stopping_criteria}). We therefore have 3 parameters
-to tune in an MCSA implementation: the number of Monte Carlo histories
-computed in the Neumann-Ulam solve during each MCSA iteration, the
-weight cutoff for those histories, and the total MCSA convergence
-tolerance as specified by $\epsilon$.
-
-%%---------------------------------------------------------------------------%%
-\section{Simplified $P_N$ Approximation}
-I was thinking about dumping the SPN tech note derivation here. Its a
-lot but it would be a good place to formally present the Exnihilo
-discretization.
-
-%%---------------------------------------------------------------------------%%
-\section{Model Problem}
-Fuel assembly calculations are a critical piece of nuclear engineering
-infrastructure for reactor core analysis and design. At this level,
-individual fuel pins may be resolved at fine resolution in a variety
-of configurations. As a sophisticated problem of interest to push the
-limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
-used with varying energy group structure and $SP_N$ discretization in
-a criticality calculation. A cross section along the vertical axis
-showing homogenized fuel pin materials and the associated grid is
-given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
-the materials configuration along the horizontal axis is given by
-Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
-bottom is given in Figure~\ref{fig:problem3_end}. On the top and
-bottom of the assembly, vacuum conditions are used as well as on the
-top and right boundaries in
-Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
-on the left and bottom boundaries of
-Figure~\ref{fig:problem3_radial_mat}, effectively giving a
-representation of one quarter of the assembly. For the spatial
-discretization, each fuel pin (gray regions in
-Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
-mesh with materials and cross sections homogenized over this region.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{problem3_radial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
-    \textit{Reflecting boundaries are used on the left and lower
-      boundaries to create a complete $17 \times 17$ assembly
-      geometry. Gray regions are homogenized fuel and red regions are
-      homogenized moderator. Each fuel pin is resolved by a $2 \times
-      2$ mesh.}}
-  \label{fig:problem3_radial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{problem3_axial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
-      geometry is subdivided along the axial direction into 50 zones
-      spaced to capture material boundaries. Important details include
-      spacer grids along the length of the fuel pins and reactor core
-      structures on the top and bottom of the assembly. Lighter purple
-      material in the center of the assembly is moderator and darker
-      purple/red material is fuel.}}
-  \label{fig:problem3_axial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{problem3_end.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry end detail.}
-    \textit{Reactor core structure including spacer grids and plenum
-      has been included. Lighter purple material on the right of the
-      figure is moderator and darker purple/red material is fuel.}}
-  \label{fig:problem3_end}
-\end{figure}
-
-Significant geometric details are contained in the model including
-spacer grids, fuel pins with homogenized cladding and gas gap, core
-plenum, and moderator with boron. Group cross sections and other
-discrete nuclear data are generated as needed by a cross-section
-processing module dependent on the meshing parameters used to
-discretize the geometry and single-dimension pin-cell calculations for
-initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
-gives the primary design parameters for the fuel assembly
-calculations.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Parameter}} & 
-      \multicolumn{1}{l}{\textbf{Value}} \\
-      Power Level & 0 MW \\
-      Inlet Temperature & 326.85C \\
-      Fuel Temperature & 600C \\
-      Boron Concentration & 1300 ppm \\
-      Moderator Density & 0.743 g/cc \\
-      Helium Density & \sn{1.79}{-4} g/cc \\
-      Zirconium Density & 6.56 g/cc \\
-      Stainless Steel Density & 8.0 g/cc \\
-      Inconel Density & 8.19 g/cc \\
-      UO2 Density & 10.257 g/cc \\
-      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
-      assembly criticality calculation.}}
-  \label{tab:problem3_parameters}
-\end{table}
+generate the correction. We refer the reader to
+\cite{evans_monte_2012} for a more rigorous defintion of the method.
 
@@
 iteration will dictate convergence and remain unchanged throughout the
-computation\footnote{The operator will change if, for example, physics
-  feedback through temperature or potentially burnup is
-  considered. These additional physics will modify the cross sections
-  used to assemble $\mathbf{M}$. The calculations presented here will
-  consist of a single eigenvalue calculation with a static
-  $\mathbf{M}$.} while the addition of fission to the system will only
+computation while the addition of fission to the system will only
 modify the source of neutrons and the multiplication factor while not
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Preconditioned MCSA}
-
+\subsection{Preconditioned MCSA}
+\label{subsec:preconditioning}
 In most cases, at least a minimal amount of \textit{preconditioning}
@@
 Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
-preconditioned linear operator. In addition, for MCSA to apply to
-increasingly difficult problems, more advanced preconditioning
-techniques that require the generation of the composite operator may
-be necessary for convergence.
+preconditioned linear operator.
+
+%%---------------------------------------------------------------------------%%
+\section{Fuel Assembly Model Problem}
+\label{sec:fuel_assembly}
+Fuel assembly calculations are a critical piece of nuclear engineering
+infrastructure for reactor core analysis and design. At this level,
+individual fuel pins may be resolved at fine resolution in a variety
+of configurations. As a sophisticated problem of interest to push the
+limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
+used with varying energy group structure and $SP_N$ discretization in
+a criticality calculation. A cross section along the vertical axis
+showing homogenized fuel pin materials and the associated grid is
+given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
+the materials configuration along the horizontal axis is given by
+Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
+bottom is given in Figure~\ref{fig:problem3_end}. On the top and
+bottom of the assembly, vacuum conditions are used as well as on the
+top and right boundaries in
+Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
+on the left and bottom boundaries of
+Figure~\ref{fig:problem3_radial_mat}, effectively giving a
+representation of one quarter of the assembly. For the spatial
+discretization, each fuel pin (gray regions in
+Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
+mesh with materials and cross sections homogenized over this region.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4in]{problem3_radial_mat.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
+    \textit{Reflecting boundaries are used on the left and lower
+      boundaries to create a complete $17 \times 17$ assembly
+      geometry. Gray regions are homogenized fuel and red regions are
+      homogenized moderator. Each fuel pin is resolved by a $2 \times
+      2$ mesh.}}
+  \label{fig:problem3_radial_mat}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6.0in]{problem3_axial_mat.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
+      geometry is subdivided along the axial direction into 50 zones
+      spaced to capture material boundaries. Important details include
+      spacer grids along the length of the fuel pins and reactor core
+      structures on the top and bottom of the assembly. Lighter purple
+      material in the center of the assembly is moderator and darker
+      purple/red material is fuel.}}
+  \label{fig:problem3_axial_mat}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6.0in]{problem3_end.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly geometry end detail.}
+    \textit{Reactor core structure including spacer grids and plenum
+      has been included. Lighter purple material on the right of the
+      figure is moderator and darker purple/red material is fuel.}}
+  \label{fig:problem3_end}
+\end{figure}
+
+Significant geometric details are contained in the model including
+spacer grids, fuel pins with homogenized cladding and gas gap, core
+plenum, and moderator with boron. Group cross sections and other
+discrete nuclear data are generated as needed by a cross-section
+processing module dependent on the meshing parameters used to
+discretize the geometry and single-dimension pin-cell calculations for
+initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
+gives the primary design parameters for the fuel assembly
+calculations.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{ll}\hline\hline
+      \multicolumn{1}{l}{\textbf{Parameter}} & 
+      \multicolumn{1}{l}{\textbf{Value}} \\
+      Power Level & 0 MW \\
+      Inlet Temperature & 326.85C \\
+      Fuel Temperature & 600C \\
+      Boron Concentration & 1300 ppm \\
+      Moderator Density & 0.743 g/cc \\
+      Helium Density & \sn{1.79}{-4} g/cc \\
+      Zirconium Density & 6.56 g/cc \\
+      Stainless Steel Density & 8.0 g/cc \\
+      Inconel Density & 8.19 g/cc \\
+      UO2 Density & 10.257 g/cc \\
+      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
+      assembly criticality calculation.}}
+  \label{tab:problem3_parameters}
+\end{table}
+
+%%---------------------------------------------------------------------------%%
+\section{Results}
+In this section we present solutions to the fuel assembly problem
+using MCSA preconditioning strategies.
 
@@
 \label{subsec:jacobi_prec_assembly_calc}
-A single energy group problem was first computed with
-$SP_1$ discretization, effectively giving the one-speed neutron
-diffusion system for the fuel assembly resulting in 20,088 degrees of
-freedom in the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives
-the residual infinity norm as a function of iteration for the MCSA
-linear solve in the first eigenvalue iteration using 25,000 stochastic
-histories at every iteration for the adjoint Neumann-Ulam solve.
+Previous work in applying MCSA to the thermal radiation diffusion
+equation had success in using Jacobi preconditioning
+\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
+equations given their similar diffusion-like form. A single energy
+group problem was first computed with $SP_1$ discretization,
+effectively giving the one-speed neutron diffusion system for the fuel
+assembly resulting in 20,088 degrees of freedom in the
+problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
+infinity norm as a function of iteration for the MCSA linear solve in
+the first eigenvalue iteration using 25,000 stochastic histories at
+every iteration for the adjoint Neumann-Ulam solve.
 
@@
 Convergence was not achieved as noted by the rapid rise in the
-residual over a few iterations. Based on the spectral radius
-computations performed, these results are not in line with
-expectations for this problem. Additional computations performed with
+residual over a few iterations. Additional computations performed with
 \sn{1}{6} histories per iteration exhibited the same divergent
@@
 investigate further, a block Jacobi preconditioned Richardson
-iteration was used to solve the same
-problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
-residual infinity norm as a function of iteration for the Richardson
-linear solve in the first eigenvalue iteration.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{block_jacobi_rich_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned Richardson solve during the first
-      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
-      problem.} \textit{A spectral radius near 1 was observed for the
-      iteration matrix.}}
-  \label{fig:block_jacobi_res_richardson}
-\end{figure}
-
-Poor convergence is observed for the Richardson iteration, however,
-convergence is achieved meaning that the preliminary eigenvalue
-criteria needed to satisfy MCSA convergence has also been met. The
-number of iterations required for the Richardson iteration to converge
-will give an approximation for the spectral radius via
-Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
-converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
-0.998$, nearing the limits of MCSA applicability and well beyond the
-spectral radii generated in the initial spectral analysis.
-
-Based on these results, it then appears that even if the simple
-criteria of a spectral radius of less than one is met and the
-Richardson iteration will converge with the same preconditioning, MCSA
-still may not converge. We then expect the issue to reside with the
-Neumann-Ulam solve providing the correction as the Richardson
-iteration is known to provide the correct result. Furthermore, the
-fact that the spectral radius is less than one means that the
-stochastic histories in the block Jacobi preconditioned Neumann-Ulam
-method are eventually being terminated by the weight cutoff. Based on
-this, the correction being generated by the Neumann-Ulam solve is the
-component of MCSA causing a divergent solution which we study next.
-
-\subsection{MCSA Breakdown}
-\label{subsec:mcsa_break_down}
+iteration was used to solve the same problem. Poor convergence was
+observed for the Richardson iteration, however, convergence was
+achieved meaning that the preliminary eigenvalue criteria needed to
+satisfy MCSA convergence has also been met. With 7291 iterations
+required to converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H})
+\approx 0.998$, nearing the limits of MCSA applicability. Based on
+these results, it then appears that even if the simple criteria of a
+spectral radius of less than one is met and the Richardson iteration
+will converge with the same preconditioning, MCSA still may not
+converge.
+
+\subsubsection{MCSA Breakdown}
+\label{subsubsec:mcsa_break_down}
 To study the breakdown of MCSA at iteration matrix spectral radii near
 one, we will use the simpler homogeneous 2-dimensional one-speed
-neutron diffusion system presented in
-Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
-this system, we can vary the cross sections while maintaining a fixed
-grid in order to achieve varying spectral radii. For these studies, we
-neglect fission as MCSA behavior is dictated by the transport operator
-$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
-generate a fixed source.
+neutron diffusion system in a rectilinear grid with vaccuum boundary
+conditions to isolate this behavior. In this system, we can vary the
+cross sections while maintaining a fixed grid in order to achieve
+varying spectral radii. For these studies, we neglect fission as MCSA
+behavior is dictated by the transport operator $\mathbf{M}$ in an
+eigenvalue scheme with the fission matrix used to generate a fixed
+source.
 
@@
   \begin{center}
-    \includegraphics[width=4.5in]{breakdown_iterations.pdf}
+    \includegraphics[width=3.5in]{breakdown_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.5in]{breakdown_histories.pdf}
+    \includegraphics[width=3.5in]{breakdown_histories.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.5in]{breakdown_time.pdf}
+    \includegraphics[width=3.5in]{breakdown_time.pdf}
   \end{center}
@@
 
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Advanced Preconditioning Strategies}
-\label{subsec:spn_advanced_preconditioning}
-For the fuel assembly criticality problem presented in the previous
-section, the spectral radius was near one using Jacobi preconditioning
-and in a regime in which MCSA breakdown is observed. In this regime
-the number of stochastic histories required to converge MCSA increases
-rapidly and the resulting poor performance is compounded by those
-histories being increasingly expensive to compute. To overcome this
-difficulty, advanced preconditioning strategies for the $SP_N$ problem
-are required, beyond simple Jacobi methods, that can reduce the
-spectral radius into a region of better MCSA performance. Two modern
-algebraic preconditioning strategies, ILUT and SPAINV, will be
-presented here and used within the explicit preconditioning framework
-given by Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on
-MCSA solutions of the fuel assembly criticality problem will be
-provided.
-
-\subsection{ILUT Preconditioning}
-\label{subsec:spn_ilut_preconditioning}
-ILUT preconditioning is chosen first due to its applicability to
-problems that have a dominating elliptic character such as the fuel
-assembly problem or other problems with a large scattering to
-absorption ratio \cite{saad_iterative_2003}. The algebraic formulation
-of ILUT requires the physics operator to be fully formed, also a
-requirement for MCSA, and is therefore easily incorporated into our
-solution sequence.
-
-Incomplete lower-upper (ILU) factorizations of the linear operator can
-be used as a simple mechanism to form an approximate inverse of a
-preconditioner. To build the factorization, the sparse upper and lower
-triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
-that the residual matrix formed by the factorization
-\cite{saad_iterative_2003}:
-\begin{equation}
-  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
-  \label{eq:ilu_residual_matrix}
-\end{equation}
-has a specified sparsity pattern and element magnitude. When a drop
-tolerance threshold is used, elements generated in the factors below
-that magnitude are dropped, resulting in ILU threshold (ILUT)
-preconditioning. The sparsity pattern in this case is determined from
-the input matrix to be preconditioned and the number of elements
-maintained in the factor is specified by a fill level parameter. A
-fill level of 1 will generate the same number of elements as the
-sparsity pattern of the input matrix while a fill level of 2 will
-contain twice as many elements in the factorization, resulting in a
-better representation of the true LU factorization. The inverse of the
-lower and upper triangular factors may then be easily inverted by
-means of simple elimination procedure to produce the
-preconditioner. For MCSA, $\mathbf{L}^{-1}$ will be used on the left
-and $\mathbf{U}^{-1}$ on the right to precondition the system.
-
-For the explicit preconditioning scheme presented in
-Eq~(\ref{eq:left_right_mcsa}), the fully inverted preconditioners must
-be generated in order to build the set of probabilities and weights
-required for Monte Carlo sampling. Therefore, for a linear system of
-size $N$, $N$ triangular solves with the basis of each row in the
-system as a solution vector will be required in order to extract the
-inverse matrices from production ILU implementations. In addition,
-parallel implementations typically generate lower and upper factors
-that are only triangular locally, providing an easily parallelizable
-mechanism to generate the action of the preconditioner inverse. A
-consequence of this choice for parallel scalability is a degradation
-of the preconditioning quality as the size of the parallel system is
-increased. For serial computations, the triangular factorization is
-potentially exact depending on the parameters chosen while at
-thousands of parallel tasks, the global triangular factors differ
-significantly from the true factorization. As a result, more
-iterations are required at higher levels of parallelism to converge
-the system and will ultimately degrade overall scalability of the
-system with respect to total wall time to converge.
-
-MCSA preconditioned with ILUT was used to solve the fuel assembly
-problem presented in the previous section for a 1-group $SP_1$
-discretization. Unlike the Jacobi preconditioned strategy, convergence
-was achieved with ILUT preconditioning. To study convergence
-sensitivity to ILUT parameters, the fill level and drop tolerance were
-parametrically varied with the number of iterations to converge an
-eigenvalue iteration for the fuel assembly problem recorded. In
-addition, the maximum number of non-zero entries observed in all rows
-for the left/right preconditioned composite linear operator was
-recorded to observe the effects of the explicit preconditioner
-inversion and matrix-matrix multiplication on sparsity. To provide
-some sparsity to the factorization, an additional step was added with
-the ILUT drop tolerance used to drop elements in the explicitly
-inverted preconditioners. For each calculation, the number of
-iterations reported was for a single eigenvalue iteration with
-\sn{3}{4} histories at each MCSA iteration to compute the Monte Carlo
-correction using the adjoint collision estimator. All ILUT
-calculations reported here were performed on a single CPU and
-therefore this data does not take into account the aforementioned
-effects of parallel decomposition on the quality of the ILUT
-preconditioning.
-
-Figure~\ref{fig:ilut_iterations} gives the number of iterations to
-converge the fixed source problem to a tolerance of \sn{1}{-8}. As
-expected, the higher the fill level chosen for the ILUT factorization,
-the fewer iterations are required to converge the problem with only 9
-MCSA iterations were required to converge the problem for the fill
-level of 3 and drop tolerance of \sn{1}{-5} case. For higher levels of
-fill, the iterations needed to converge were not as sensitive,
-signaling that a larger drop tolerance can perhaps be used without a
-significant degradation in iterative performance. At smaller fill
-levels, the sensitivity to the ILUT drop tolerance is more
-significant. For all fill levels used, convergence was not achieved
-for the fuel assembly problem with a drop tolerance larger than
-\sn{1}{-3}.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{ilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge an
-      eigenvalue iteration for the fuel assembly problem with ILUT
-      preconditioning as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
-      were used.}}
-  \label{fig:ilut_iterations}
-\end{figure}
-
-Unfortunately, gaining convergence (and excellent iterative
-performance) with MCSA for the $SP_N$ fuel assembly problem comes at
-an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
-of non-zero entries in the composite linear operator generated by
-preconditioning as a function ILUT drop tolerance for varying values
-of fill level. For the 1-group $SP_1$ discretization, the original
-linear operator will contain only a maximum of 7 non-zero entries per
-row in the system. As observed in Figure~\ref{fig:ilut_size},
-performing the explicit preconditioning yields composite linear
-operators with $O(1,000)$ elements in a row for all combinations of
-fill level and drop tolerance, over 10\% of the total row size for
-this particular problem. This large number of row entries, observed
-for a significant fraction of rows in the system, creates several
-problems. First, sparsity is completely destroyed with each state in
-the system now coupled to over 10\% of the total states in the system
-through the composite iteration matrix. This means that Monte Carlo
-sampling tables will be large, requiring significant memory to store
-them and a substantial overhead in the sampling procedure during
-transport. Second, the matrix-matrix multiply operations required to
-build the composite operator see significant performance losses due to
-the amount of data that must be handled. In parallel, losing sparsity
-severely inhibits performance where now parallel operations require
-communications among orders of magnitude more processors for
-nearest-neighbor type algorithms.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{ilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with ILUT preconditioning given as a function of ILUT
-      drop tolerance.} \textit{Each colored curve represents the row
-      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
-      2.5, and 3.0 were used.}}
-  \label{fig:ilut_size}
-\end{figure}
-
-As a means to assess the quality of the preconditioning, a simple
-metric is developed to address these concerns and allow comparison to
-future developments. For our studies, our core performance metric will
-be iterative performance with the minimum number of MCSA iterations
-required to converge the problem desired. For Monte Carlo
-calculations, this improved performance is balanced by the creation of
-a dense composite system and we seek to reduce the number of non-zero
-entries to a minimum value in order to lower the amount of coupling
-among states in the system and reduce the amount of memory used along
-with the potential latency overhead. Given these two objectives, the
-following metric is proposed:
-\begin{equation}
-  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
-    \# non-zero values})\:,
-\end{equation}
-where the highest quality preconditioning is one that minimizes this
-metric. This metric gives equal importance to both iterative
-performance and sparsity with the purpose of simply measuring their
-combined effect rather than seeking some optimal relationship of
-importance. For the ILUT preconditioned data provided,
-Figure~\ref{fig:ilut_quality} provides the computed metric as a
-function of ILUT drop tolerance for each fill level used. In general,
-the metric is similar to the data observed for the number of
-iterations required to converge as the non-zero row entries changed by
-a smaller fraction over drop tolerances tested. For higher levels of
-fill, the quality metric is observed to be relatively insensitive to
-the drop tolerance as maintaining more entries in the preconditioner
-has a stronger impact than filtering small values.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{ilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{ILUT preconditioning quality metric for the fuel
-      assembly problem given as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
-      and 3.0 were used.}}
-  \label{fig:ilut_quality}
-\end{figure}
-
-\subsection{Sparse Approximate Inverse Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-Sparse approximate inverse (SPAINV) preconditioning is chosen next
-because the true inverse of the preconditioner is the product of the
-algorithm, avoiding the need for many inverse matrix applications as
-required by using ILUT with MCSA. In addition, the algorithm seeks to
-maintain the sparsity of the inverse preconditioner, potentially
-alleviating some of the dense operator issues observed with ILUT.
-
-SPAINV produces a sparse approximation of the inverse matrix directly
-by minimizing the Frobenius norm of the residual matrix
-\cite{saad_iterative_2003}:
-\begin{equation}
-  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
-  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
-  \label{eq:frobenius_norm_min}
-\end{equation}
-where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
-norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
-$j^{th}$ columns of the identity and preconditioning matrices
-respectively. Here, $\mathbf{M}$ is the actual inverse of the
-preconditioner and is formed directly by the algorithm. On the right
-hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
-represented as a effective sum of $N$ linear system residual norms.
-
-A few iterations of a subspace method or other iterative scheme can be
-used to effectively solve these systems to a loose tolerance and yield
-the columns of the approximate preconditioning matrix. At each
-iteration, threshold values are used to eliminate the components of
-each $\mathbf{m}_j$ column to maintain the desired level of
-sparsity. In addition, a sparsity pattern can be predetermined and
-enforced during this process. For the SPAINV implementation used for
-this work, the sparsity pattern was defined by levels such that a
-level of $N$ for an input operator $\mathbf{A}$ will generate a
-preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
-$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
-a factorization such as ILUT, parallel results are reproduced
-regardless of parallel problem size, meaning that a serial computation
-will converge in the same number of iterations as a computation with
-thousands of cores.
-
-MCSA preconditioned with SPAINV was used to solve the fuel assembly
-problem again for a 1-group $SP_1$ discretization. To study
-convergence sensitivity to SPAINV parameters, the number of levels in
-the sparsity pattern and threshold were parametrically varied with the
-number of iterations to converge a single eigenvalue iteration for the
-fuel assembly problem recorded along with the maximum number of
-non-zero entries observed in all matrix rows for the preconditioned
-composite linear operator. For each calculation, the number of
-iterations required to converge reported was for a single eigenvalue
-iteration with \sn{3}{4} histories at each MCSA iteration to compute
-the Monte Carlo correction using the adjoint collision estimator. All
-SPAINV calculations reported here were performed on a single CPU.
-
-Figure~\ref{fig:spainv_iterations} gives the number of MCSA iterations
-needed to converge the fuel assembly problem and
-Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
-per row observed in the composite linear operator as a function the
-SPAINV threshold. For this analysis, the number of levels in the
-sparsity pattern was varied from 3 to 7 with convergence of the fuel
-assembly problem not achieved for smaller level values. Compared to
-ILUT preconditioning, at higher levels we see comparable iterative
-performance with a relative invariance to the threshold value. The
-threshold did have a significant effect, however, on the time required
-to generate the preconditioner with a threshold of 0.001 over an order
-of magnitude slower than a threshold of 0.01 due to the inclusion of a
-significant number of extra values in the input operator,
-$\mathbf{A}$. In addition to comparable iterative performance, the
-sparsity of the composite operator is greatly improved with nearly an
-order of magnitude reduction in number of non-zero entries per row for
-lower level values.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{spainv_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV preconditioning as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different SPAINV level pattern. Levels of 3, 4, 5, 6, and 7 were
-      used.}}
-  \label{fig:spainv_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{spainv_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV preconditioning given as a function of
-      SPAINV threshold.} \textit{Each colored curve represents the row
-      size for a different SPAINV level pattern. Levels of 3, 4, 5, 6,
-      and 7 were used.}}
-  \label{fig:spainv_size}
-\end{figure}
-
-Figure~\ref{fig:spainv_quality} gives the quality metric as a function
-of the threshold value for each of the sparsity levels computed. Not
-only do we see an improved quality metric over ILUT preconditioning
-(about an order of magnitude less), but we also note that the ideal
-sparsity level is not the largest nor the smallest. In fact, the
-largest and smallest levels (3 and 7) performed the worst in terms of
-the quality metric with a level of 4 performing the best.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{spainv_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
-      assembly problem given as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different SPAINV level pattern. Levels of 3, 4, 5, 6, and
-      7 were used.}}
-  \label{fig:spainv_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{The Reduced Domain Approximation}
-\label{subsec:spn_prec_rda}
-Although ILUT and SPAINV preconditioning permitted MCSA solutions for
-the fuel assembly problem, a primary concern is the number of non-zero
-states in each row of the system generated by the explicit
-preconditioning strategy. In many cases, orders of magnitude more
-matrix elements were generated resulting in poor scalability for
-domain decomposed algorithms and overall performance issues for Monte
-Carlo. To alleviate this, we propose modifying the domain sampled by
-the Monte Carlo method to generate the correction within MCSA by
-removing terms in the iteration matrix through some predetermined
-criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
-stochastic realization of multiple matrix-vector multiplies, the
-criteria we choose should be based on maintaining those elements that
-will make the largest contributions to the multiplication and
-therefore recover as much of the original product as possible.
-
-For each row in the preconditioned system, we can then apply the
-reduced domain approximation to eliminate those elements that either
-fall below a specified threshold level, fall within a fill level $N$
-where we keep the $N$ largest elements in each row, or both giving:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
-    \label{eq:rda_correction}\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:rda_mcsa}
-\end{subequations}
-where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
-reduced domain approximation has been applied to the iteration matrix
-in the Monte Carlo method. If enough of the terms in the original
-iteration matrix have been maintained, the character of the original
-iteration matrix should be retained and continue to accelerate the
-iterative scheme.
-
-We next apply the reduced domain approximation to both ILUT and SPAINV
-preconditioned system to reduce the density of the composite linear
-operator to more manageable levels and study the resulting iterative
-performance. For each preconditioner, the parameters that achieved
-the best quality metric results from the previous analysis were
-used. These correspond to ILUT parameters of a fill level of 5.0 and a
-drop tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level
-of 4 and a threshold of 0.1. The reduced domain threshold was set to
-\sn{1}{-10} in order to eliminate any exceedingly small values from
-the matrix (often this is simply removing non-zero elements within the
-floating point tolerance of zero). The reduced domain fill level was
-then varied, starting with the largest non-zero entries per row value
-observed for each of the preconditioning types in order to assess its
-effects relative to the case where no reduced domain approximation was
-applied.
-
-Figure~\ref{fig:rda_iterations} gives the number of iterations
-required to converge for each preconditioner type as a function of
-reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
-corresponding quality metric for each data point where the number of
-non-zero entries used to compute the metric is equivalent to the
-reduced domain fill level. We first note that SPAINV preconditioning
-alone is significantly more sensitive to the reduction in domain size
-over the ILUT-based methods, although the preconditioner was of
-$O(100)$ non-zero entries per row without any approximation
-applied. For the ILUT-based methods, performance was significantly
-better with convergence achieved in less than 40 MCSA iterations with
-only 10 non-zero entries in each row (vs. 7 non-zero entries for the
-case with no preconditioning). Looking at the quality metric data, we
-see a reduction in the quality metric as a function of the reduced
-domain fill level, achieving 2 orders of magnitude reduction in the
-quality metric for the ILUT preconditioning.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{rda_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      each preconditioning as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{rda_quality.pdf}
-  \end{center}
-  \caption{\textbf{Preconditioning quality metric for the fuel
-      assembly problem given as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_quality}
-\end{figure}
-
-Although applying the reduced domain approximation results in a
-successful recovery of sparsity for the Monte Carlo problem while
-maintaining good convergence properties with ILUT, there is still an
-issue of forming the composite operator before applying the
-approximation and potentially generating the transpose of this
-operator in the case of the adjoint Monte Carlo method. Because of
-this, memory and scaling issues may still be observed when building
-the probability and weight matrices for the Monte Carlo problem. In
-addition, the expensive extraction of the inverse of the
-preconditioning operators for the explicit scheme creates a
-significant cost in overall performance. Future work in this area
-should be considerate of these important components of the
-preconditioned MCSA algorithm.
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Verification}
-\label{sec:spn_mcsa_verification}
-Through the numerical studies presented in this chapter, we have
-developed a set of preconditioning techniques that permit MCSA to be
-used with the $SP_N$ form of the neutron transport equation and
-applied to a difficult fuel assembly criticality problem. In addition,
-several steps were taken to mitigate the dense composite linear
-operators that arise when performing explicit MCSA preconditioning
-necessary for achieve convergence. Using these results, we now compare
-MCSA to conventional Krylov solvers that would be typically used to
-solve the $SP_N$ form of the transport equation as a means of
-verification.
-
-MCSA solutions to the fuel assembly problem with 1, 2, and 4 energy
-groups using $SP_1$ discretization will computed using both BiCGStab
-and GMRES implementations from the Trilinos scientific computing
-library Aztec \cite{heroux_overview_2005}. The neutron fluxes, the
-number of eigenvalue iterations required to converge the problem, and
-the eigenvalue itself will each be evaluated. For the flux values, we
-will look at the 2-norm and infinity norm of the relative difference
-vector between the reference case and the comparison case in each
-energy group:
-\begin{equation}
-  \text{Relative Group Flux Difference} = \frac{|\ve{\Phi}_{ref}^g -
-    \ve{\Phi}^g|}{|\ve{\Phi}_{ref}^g|}\:,
-  \label{eq:relative_difference_vector}
-\end{equation}
-where the absolute value, subtraction, and division operations are
-performed element-wise to construct the vector. For this work, the
-BiCGStab results will be taken as the reference case.
+\subsection{ILUT Preconditioned Results}
+\label{subsec:spn_comparison}
+MCSA will now be compared to the same Krylov solvers in terms of both
+iterative performance and CPU timing. A Richardson iteration solution
+will also be used in order to assess the acceleration provided by the
+residual Monte Carlo component of MCSA.  To compare iterative
+performance, the number of linear solver iterations required to
+converge each eigenvalue iteration was recorded and then averaged over
+all eigenvalue iterations for each solver. MCSA solutions to the fuel
+assembly problem with 1, 2, and 4 energy groups using $SP_1$
+discretization will computed using both BiCGStab and GMRES
+implementations from the Trilinos scientific computing library Aztec
+\cite{heroux_overview_2005}.
 
@@
 
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{c}{\textbf{Iterations}} &
-      \multicolumn{1}{c}{\textbf{Eigenvalue}} \\
-      \hline
-      BiCGStab-ILUT & 1 & 25 & 1.105987 \\
-      GMRES-ILUT & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-R-C & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-MR-C & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-R-EV & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-MR-EV & 1 & 25 & 1.105987 \\
-      Richardson-ILUT & 1 & 25 & 1.105987 \\
-      \hline
-      BiCGStab-ILUT & 2 & 35 & 1.155239 \\
-      GMRES-ILUT & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-R-C & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-MR-C & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-R-EV & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-MR-EV & 2 & 35 & 1.155239 \\
-      Richardson-ILUT & 2 & 35 & 1.155239 \\
-      \hline
-      BiCGStab-ILUT & 4 & 31 & 1.175850 \\
-      GMRES-ILUT & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-R-C & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-MR-C & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-R-EV & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-MR-EV & 4 & 31 & 1.175850 \\
-      Richardson-ILUT & 4 & 31 & 1.175850 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial eigensolver verification results for the
-      fuel assembly problem.} \textit{Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_ev_results}
-\end{table}
-
-Table~\ref{tab:serial_differences_g1},
-Table~\ref{tab:serial_differences_g2}, and
-Table~\ref{tab:serial_differences_g4} give the infinity norms and
-2-norms of the flux relative difference vectors given by
-Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
-energy groups. In general, we see good agreement in the 2-norm between
-the MCSA variants and the GMRES calculations, meaning that they have
-approximately the same variation from the BiCGStab calculations. We do
-note larger infinity norm values of this vector for some of the MCSA
-variants relative to the GMRES results (i.e. group 1 in the 2-group
-calculation), however, this large maximum value is apparently a local
-phenomenon with good agreement in the 2-norm for the
-calculations. Based on the good agreement between MCSA and GMRES for
-the relative difference of the group fluxes and the eigenvalue results
-presented here, the MCSA implementation should be deemed correct for
-the fuel assembly problem for serial calculations with multiple energy
-groups.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{\infty}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{4.562}{-6} & \sn{6.746}{-5} \\
-      MCSA-ILUT-R-C & 0 & \sn{2.099}{-5} & \sn{2.541}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{4.812}{-5} & \sn{4.672}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{2.371}{-5} & \sn{2.740}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{3.605}{-5} & \sn{2.554}{-4} \\
-      Richardson-ILUT & 0 & \sn{1.531}{-5} & \sn{2.771}{-4} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      1-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference.}}
-  \label{tab:serial_differences_g1}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{\infty}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{7.590}{-6} & \sn{1.234}{-4} \\
-      MCSA-ILUT-R-C & 0 & \sn{5.308}{-6} & \sn{1.066}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{6.388}{-5} & \sn{2.233}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{4.197}{-5} & \sn{2.168}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{2.736}{-5} & \sn{8.160}{-5} \\
-      Richardson-ILUT & 0 & \sn{3.976}{-6} & \sn{7.059}{-5} \\
-      \hline
-      GMRES-ILUT & 1 & \sn{7.579}{-6} & \sn{1.246}{-4} \\
-      MCSA-ILUT-R-C & 1 & \sn{1.645}{-5} & \sn{1.210}{-4} \\
-      MCSA-ILUT-MR-C & 1 & \sn{6.635}{-4} & \sn{9.435}{-4} \\
-      MCSA-ILUT-R-EV & 1 & \sn{1.797}{-4} & \sn{2.983}{-4} \\
-      MCSA-ILUT-MR-EV & 1 & \sn{1.556}{-4} & \sn{2.055}{-4} \\
-      Richardson-ILUT & 1 & \sn{4.031}{-6} & \sn{7.110}{-5} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      2-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference.}}
-  \label{tab:serial_differences_g2}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{\infty}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{2.376}{-5} & \sn{4.043}{-4} \\
-      MCSA-ILUT-R-C & 0 & \sn{2.639}{-5} & \sn{3.950}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{3.958}{-5} & \sn{4.635}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{2.749}{-5} & \sn{3.975}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{2.969}{-5} & \sn{4.054}{-4} \\
-      Richardson-ILUT & 0 & \sn{2.189}{-5} & \sn{3.791}{-4} \\
-      \hline
-      GMRES-ILUT & 1 & \sn{2.271}{-5} & \sn{3.952}{-4} \\
-      MCSA-ILUT-R-C & 1 & \sn{3.701}{-5} & \sn{3.933}{-4} \\
-      MCSA-ILUT-MR-C & 1 & \sn{3.803}{-5} & \sn{3.933}{-4} \\
-      MCSA-ILUT-R-EV & 1 & \sn{4.142}{-5} & \sn{4.011}{-4} \\
-      MCSA-ILUT-MR-EV & 1 & \sn{2.723}{-5} & \sn{4.030}{-4} \\
-      Richardson-ILUT & 1 & \sn{2.137}{-5} & \sn{3.768}{-4} \\
-      \hline
-      GMRES-ILUT & 2 & \sn{2.310}{-5} & \sn{4.021}{-4} \\
-      MCSA-ILUT-R-C & 2 & \sn{2.490}{-5} & \sn{3.956}{-4} \\
-      MCSA-ILUT-MR-C & 2 & \sn{2.214}{-5} & \sn{3.608}{-4} \\
-      MCSA-ILUT-R-EV & 2 & \sn{4.456}{-5} & \sn{3.833}{-4} \\
-      MCSA-ILUT-MR-EV & 2 & \sn{3.366}{-5} & \sn{3.797}{-4} \\
-      Richardson-ILUT & 2 & \sn{2.143}{-5} & \sn{3.767}{-4} \\
-      \hline
-      GMRES-ILUT & 3 & \sn{2.335}{-5} & \sn{4.029}{-4} \\
-      MCSA-ILUT-R-C & 3 & \sn{3.045}{-5} & \sn{3.963}{-4} \\
-      MCSA-ILUT-MR-C & 3 & \sn{2.696}{-5} & \sn{3.606}{-4} \\
-      MCSA-ILUT-R-EV & 3 & \sn{1.947}{-5} & \sn{4.283}{-4} \\
-      MCSA-ILUT-MR-EV & 3 & \sn{3.295}{-5} & \sn{3.837}{-4} \\
-      Richardson-ILUT & 3 & \sn{2.161}{-5} & \sn{3.772}{-4} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      4-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference.}}
-  \label{tab:serial_differences_g4}
-\end{table}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Performance Comparison to Conventional Methods}
-\label{sec:spn_comparison}
-With the same set of solver parameters used for verification
-calculation, MCSA will now be compared to the same Krylov solvers in
-terms of both iterative performance and CPU timing. A Richardson
-iteration solution will also be used in order to assess the
-acceleration provided by the residual Monte Carlo component of MCSA.
-To compare iterative performance, the number of linear solver
-iterations required to converge each eigenvalue iteration was recorded
-and then averaged over all eigenvalue iterations for each
-solver.
-
 Table~\ref{tab:spn_comparison_iterations} gives this average number of
@@
 
+From a timing perspective, the greatest burden is the time required to
+actually form the explicit inverse preconditioner matrices and the
+composite operator through matrix-matrix multiplication. The timing
+numbers reported were simply to perform the MCSA iteration procedure
+with the composite operator already
+formed. Figure~\ref{fig:spn_comparison_time} additionally presents the
+CPU times for MCSA convergence with the time to form the inverse of
+the preconditioners and the composite linear operator through
+matrix-matrix multiplication amortized over all iterations. As is
+readily observed, including the costs of these operations increases
+the MCSA computation time by another order of magnitude.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in]{solver_p_time.pdf}
+  \end{center}
+  \caption{\textbf{Average CPU time per iteration in seconds for the
+      fuel assembly problem as a function of energy groups with
+      preconditioning time included for the MCSA methods.}
+    \textit{All linear solver iterations over all eigenvalue
+      iterations were used to compute the
+      average. Table~\ref{tab:spn_solver_defs} gives the description
+      for each solver type presented in the legend. Data labeled
+      starting with PMCSA and shown in blue is identical to those
+      labeled with MCSA except that they additionally include the cost
+      of generating the inverse of the preconditioners and composite
+      linear operator.}}
+  \label{fig:spn_comparison_time}
+\end{figure}
+
 For the Krylov methods, the ILUT preconditioners are applied to a
@@
 using CPU time as a measure \cite{evans_monte_2012}.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{solver_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per linear solver iteration in
-      seconds for the fuel assembly problem as a function of energy
-      groups.}  \textit{All linear solver iterations over all
-      eigenvalue iterations were used to compute the
-      average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend with the Krylov
-      solvers represented in black and the MCSA solvers represented in
-      red.}}
-  \label{fig:spn_comparison_time}
-\end{figure}
-
-Finally, not considered in Figure~\ref{fig:spn_comparison_time} is the
-time required to actually form the explicit inverse preconditioner
-matrices and the composite operator through matrix-matrix
-multiplication. The timing numbers reported were simply to perform the
-MCSA iteration procedure with the composite operator already
-formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
-presents the CPU times for MCSA convergence with the time to form the
-inverse of the preconditioners and the composite linear operator
-through matrix-matrix multiplication amortized over all iterations. As
-is readily observed, including the costs of these operations increases
-the MCSA computation time by another order of magnitude.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{solver_p_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per iteration in seconds for the
-      fuel assembly problem as a function of energy groups with
-      preconditioning time included for the MCSA methods.}
-    \textit{All linear solver iterations over all eigenvalue
-      iterations were used to compute the
-      average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend. Data labeled
-      starting with PMCSA and shown in blue is identical to those
-      labeled with MCSA except that they additionally include the cost
-      of generating the inverse of the preconditioners and composite
-      linear operator.}}
-  \label{fig:spn_comparison_prec_time}
-\end{figure}
 
@@
 
-In this paper, the Monte Carlo Synthetic Acceleration method has
-been applied to the $SP_N$ discretization of the neutron transport
-equation and explored in the context of a difficult nuclear fuel
-assembly criticality problem. The following are the significant
-observations, findings, and contributions.
-
-\begin{itemize}
-\item MCSA can solve the asymmetric system generated by the $SP_N$
-  equations
-\item MCSA has been incorporated into the Exnihilo neutronics
-  production code base developed at Oak Ridge National Laboratory
-\item Light water reactor problems are difficult to solve with MCSA as
-  they have large spectral radii due to the neutron scattering in the
-  moderator
-\item Simple Jacobi-based preconditioning reduces the spectral radius
-  of the $SP_N$ system below unity, however, this alone is not
-  sufficient for MCSA convergence
-\item As the spectral radius approaches unity, MCSA was demonstrated
-  to break down with more stochastic histories required for
-  convergence and more CPU time required per history
-\item Advanced algebraic preconditioning strategies were applied to
-  the $SP_N$ equations to obtain convergence with ILUT chosen for
-  subsequent investigations
-\item The reduced domain approximation was introduced to mitigate the
-  memory and performance constraints of the explicit preconditioning
-  strategy although both remain significant obstacles
-\item MCSA was verified to produce the same k-eigenvalue for the fuel
-  assembly as production Krylov methods and gave good relative flux
-  agreement
-\item MCSA was observed to converge in fewer iterations per eigenvalue
-  iteration than GMRES for the fuel assembly criticality problem and
-  more than Bi-CGStab using the same preconditioning
-\item The explicit preconditioning strategy required to overcome the
-  MCSA spectral radius restriction creates run times $O(100)$ larger
-  than the production Krylov methods for the fuel assembly criticality
-  problem
-\end{itemize}
+In this paper, the Monte Carlo Synthetic Acceleration method has been
+applied to the $SP_N$ discretization of the neutron transport equation
+and explored in the context of a difficult nuclear fuel assembly
+criticality problem. We observed that MCSA can solve the asymmetric
+system generated by the $SP_N$ equations by incorporating the solution
+technique into the Exnihilo neutronics code base developed at Oak
+Ridge National Laboratory. In contrast to the thermal radiation
+diffusion problem addressed in our previous work, light water reactor
+problems are difficult to solve with MCSA as they have large spectral
+radii due to the neutron scattering in the moderator. Because of this,
+simple Jacobi-based preconditioning reduces the spectral radius of the
+$SP_N$ system below unity, however, this alone was discovered not to
+be sufficient criteria alone for MCSA convergence. As the spectral
+radius approaches unity, MCSA was demonstrated to break down with more
+stochastic histories required for convergence and more CPU time
+required per history. To alleviate these effects, a left-right
+preconditioned form of the MCSA method was developed and applied to
+the $SP_N$ equations to obtain convergence. For the nuclear fuel
+assembly problem, MCSA was observed to converge in fewer iterations
+per eigenvalue iteration than GMRES for the fuel assembly criticality
+problem and more than Bi-CGStab using the same
+preconditioning. However, the explicit preconditioning strategy
+required to overcome the MCSA spectral radius restriction creates run
+times $O(100)$ larger than the production Krylov methods for the fuel
+assembly criticality problem, indicating that significant research in
+the area of Monte Carlo preconditioning implementations is required in
+order for the method to be competetive in terms of time to solution.
 
@@
 \pagebreak
-\bibliographystyle{ieeetr} \bibliography{references}
+\bibliographystyle{ieeetr} 
+\bibliography{references}
 \end{document}

### 2014-01-16 doc/ttsp2013/ttsp_2013.tex
updating conclusion to ttsp paper

@@
 capability can enable tighter design tolerances for improved thermal
-performance and efficiency, higher fuel burn-up and therefore
-reduction in generated waste, and high confidence in accident scenario
-models. Modern high fidelity deterministic methods for large scale
-problems are often variants on the discrete ordinates ($S_N$) method
-\cite{evans_denovo:_2010}. For fission reactor neutronics simulations,
-the $S_N$ method requires potentially trillions of unknown angular
-flux moments to be computed to achieve good accuracy for the responses
-of interest \cite{slaybaugh_acceleration_2011}.
+performance and efficiency, higher fuel utilization, and high
+confidence in accident scenario models. Modern high fidelity
+deterministic methods for large scale problems are often variants on
+the discrete ordinates ($S_N$) method \cite{evans_denovo:_2010}. For
+fission reactor neutronics simulations, the $S_N$ method requires
+potentially trillions of unknown angular flux moments to be computed
+to achieve good accuracy for the responses of interest
+\cite{slaybaugh_acceleration_2011}. In the 1960's, Gelbard developed
+an ad-hoc, multidimensional extension of the simple single dimension
+planar $P_N$ equations that created a system of coupled,
+diffusion-like equations known as the simplified $P_N$ ($SP_N$)
+equations (cite gelbard). The reduction in numerical complexity
+compared to current deterministic solution methods using the $S_N$
+approximation could mean significant savings in both compute time and
+memory required.
 
@@
 linear operator, creating a set of problems complementary to the
-transient, opaque and symmetric systems of our previous analyis.
-
-In order to leverage Monte Carlo Synthetic Acceleration as a solution
-technique for neutron transport problems, its current formulation
-requires the linear operator and all preconditioners to be explicitly
-formed. Recent developments in the Exnihilo neutronics package at Oak
-Ridge National Laboratory have permitted generation of the $SP_N$
-system of equations for detailed neutronics models of fission
-reactor-based systems\cite{evans_simplified_2013}. By fully forming
-these equations and formulating them as a linear algebra problem
-instead of using an explicit iterative method, we now have access to
-modern advancements in computational linear algebra including Krylov
-solvers for asymmetric systems and algebraic preconditioning methods.
+transient, opaque, and symmetric systems of our previous analyis. In
+addition, solving the $SP_N$ equations in this way also breaks away
+from the $S_N$ forms of parallelism where spatial parallelism is
+achieved by an efficient parallel sweep, angular efficiency achieved
+by pipe-lining, and energy parallelism achieved by decoupling the
+groups. With the $SP_N$ equations as a full matrix system, we now can
+parallelize the problem as prescribed by the Monte Carlo algorithm,
+which may be significantly more scalable than current $S_N$ transport
+practices.
 
@@
 \label{subsec:mcsa_solution}
-Using the ideas of Halton, Evans and Mosher recently developed a Monte
-Carlo solution method that was not prohibited severely by the quality
-of the initial guess for the system \cite{evans_monte_2009} and later
-applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \cite{evans_monte_2012}. With their new methods,
-they achieved identical numerical results as conventional Krylov
-solvers as well as comparable performance in both number of iterations
-and CPU time. Their approach was instead to use residual Monte Carlo
-as a synthetic acceleration for a stationary method. The
-\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
-is defined as:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:mcsa}
-\end{subequations}
-where a Neumann-Ulam Monte Carlo method is used to generate the
-solution correction from the residual and Richardson's iteration in
-the first step has been rewritten as a residual correction. Using
-Monte Carlo in this way achieves the same effect as Halton's method,
-decoupling its convergence rate from the overall convergence rate of
-the method. Here, the approximate Monte Carlo solution is not driven
-to a particular convergence as it merely supplies a correction for the
-initial guess generated by Richardson's iteration. Rather, only a set
-number of histories are required using the Neumann-Ulam method to
-generate the correction. We refer the reader to
-\cite{evans_monte_2012} for a more rigorous defintion of the method.
-
 To generate the multiplication factor and steady-state flux
@@
 affecting Monte Carlo transport.
+
+At each eigenvalue iteration, we then solve the linear system with the
+MCSA:
+\begin{subequations}
+  \begin{gather}
+    \ve{r}_{l} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
+    \ve{M}\mathbf{\Phi}^{n+1}_{l}\:,\\
+    \mathbf{\Phi}^{n+1}_{l+1/2} = \mathbf{\Phi}^{n+1}_l + \ve{r}_l\:,\\
+    \ve{r}_{l+1/2} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
+    \ve{M}\mathbf{\Phi}^{n+1}_{l+1/2}\:,\\
+    \ve{M}\delta\mathbf{\Phi}^{n+1}_{l+1/2} = \ve{r}_{l+1/2}\:,\\
+    \mathbf{\Phi}^{n+1}_{l+1} = \mathbf{\Phi}^{n+1}_{l+1/2} + \delta
+    \mathbf{\Phi}_{l+1/2}\:,
+  \end{gather}
+  \label{eq:mcsa}
+\end{subequations}
+where the $l$ subscript indicates the MCSA iteration number and $n$
+again indicates the eigenvalue iteration. We refer the reader to
+\cite{evans_monte_2012} for a more rigorous defintion of the method.
 
@@
 preconditioned linear operator.
+
+\subsection{Alternative Fixed Point Iterations}
+\label{subsubsec:alternative_fixed_point}
+In addition to preconditioning the basic Richardson iteration, the
+MCSA algorithm presented in Eq~(\ref{eq:mcsa}) can be used to
+accelerate any fixed point iteration which only depends on the state
+(typically the residual) of the last iteration. As an example, for
+positive-definite but not necessarily symmetric problems, the minimal
+residual iteration \cite{saad_iterative_2003} can be used in
+conjunction with MCSA:
+\begin{subequations}
+  \begin{gather}
+    \alpha = \frac{\langle \mathbf{A}\mathbf{r}^k, \mathbf{r}^k
+      \rangle}{\langle \mathbf{A}\mathbf{r}^k, \mathbf{A}\mathbf{r}^k
+      \rangle}\:, \\
+    \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
+    \label{eq:min_res_correction}\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
+    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta\ve{x}^{k+1/2}\:,\\ 
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
+  \end{gather}
+  \label{eq:mcsa_min_res}
+\end{subequations}
+where $\alpha$ is an optimal extrapolation parameter generated from
+the constraints. Interestingly, this scheme is nearly identical to the
+Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
+additional extrapolation parameter, $\alpha$, is applied to the
+residual correction in Eq~(\ref{eq:min_res_correction}) as a means of
+selecting a more optimal search direction at each iteration,
+potentially further improving convergence.
 
@@
 and explored in the context of a difficult nuclear fuel assembly
-criticality problem. We observed that MCSA can solve the asymmetric
-system generated by the $SP_N$ equations by incorporating the solution
-technique into the Exnihilo neutronics code base developed at Oak
-Ridge National Laboratory. In contrast to the thermal radiation
-diffusion problem addressed in our previous work, light water reactor
-problems are difficult to solve with MCSA as they have large spectral
-radii due to the neutron scattering in the moderator. Because of this,
-simple Jacobi-based preconditioning reduces the spectral radius of the
-$SP_N$ system below unity, however, this alone was discovered not to
-be sufficient criteria alone for MCSA convergence. As the spectral
-radius approaches unity, MCSA was demonstrated to break down with more
-stochastic histories required for convergence and more CPU time
-required per history. To alleviate these effects, a left-right
-preconditioned form of the MCSA method was developed and applied to
-the $SP_N$ equations to obtain convergence. For the nuclear fuel
-assembly problem, MCSA was observed to converge in fewer iterations
-per eigenvalue iteration than GMRES for the fuel assembly criticality
-problem and more than Bi-CGStab using the same
-preconditioning. However, the explicit preconditioning strategy
-required to overcome the MCSA spectral radius restriction creates run
-times $O(100)$ larger than the production Krylov methods for the fuel
-assembly criticality problem, indicating that significant research in
-the area of Monte Carlo preconditioning implementations is required in
-order for the method to be competetive in terms of time to solution.
+criticality problem. We observed that MCSA can in fact solve the
+asymmetric system generated by the $SP_N$ equations. In contrast to
+the thermal radiation diffusion problem addressed in our previous
+work, light water reactor problems are difficult to solve with MCSA as
+they have large spectral radii due to the neutron scattering in the
+moderator. Because of this, simple Jacobi-based preconditioning
+reduces the spectral radius of the $SP_N$ system below unity, however,
+this alone was discovered not to be sufficient criteria alone for MCSA
+convergence. As the spectral radius approaches unity, MCSA was
+demonstrated to break down with more stochastic histories required for
+convergence and more CPU time required per history. To alleviate these
+effects, a left-right preconditioned form of the MCSA method was
+developed and applied to the $SP_N$ equations to obtain
+convergence. For the nuclear fuel assembly problem, MCSA was observed
+to converge in fewer iterations per eigenvalue iteration than GMRES
+for the fuel assembly criticality problem and more than Bi-CGStab
+using the same preconditioning. However, the explicit preconditioning
+strategy required to overcome the MCSA spectral radius restriction
+creates run times $O(100)$ larger than the production Krylov methods
+for the fuel assembly criticality problem, indicating that significant
+research in the area of Monte Carlo preconditioning implementations is
+required in order for the method to be competetive in terms of time to
+solution.
 

### 2014-01-16 doc/coppermtn_multilevel/abstract/abstract.tex
adding copper mountain 2014 abstract for multilevel solvers

\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{tmadd,tmath}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{tmath,tmadd}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[mathcal]{euscript} 
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton
\\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}}
}

\date{\today} \title{Multilevel Monte Carlo Solvers for Linear Systems}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\abstract

Linear solvers based on Monte Carlo techniques are currntly being
studied as potentially resilient alternatives to conventional
iterative methods. These solvers produce stochastic estimates of the
solutions to linear systems by estimating the terms of the Neumann
series formed by the Richardson iteration; thus providing a stochastic
realization of a relaxation method. The random walks used to estimate
the solution are a variant of Markov Chain Monte Carlo (MCMC) methods
and recent work has aimed to adapt various techniques in MCMC to
linear solver applications.

In this work we consider leveraging recent developments in multilevel
MCMC where the space over which the Markov chains are generated is
represented by a sequence of coarse and fine grids. The estimates on
the grids are then combined via superposition to achieve results with
lower statistical uncertainty than those on a single grid. When
applied to Monte Carlo linear solvers the multilevel technique
resembles the multigrid method where the Monte Carlo scheme acts as a
stochastic smoother. We will present both the Monte Carlo method for
linear systems as well as the multilevel method. We will then compare
the performance of the multilevel method against that of the
traditional method and discuss applicabililty to various problems.

\end{document}

### 2014-01-16 doc/coppermtn_multilevel/abstract/abstract.tex
spell check on copper mountain abstract

@@
 
-Linear solvers based on Monte Carlo techniques are currntly being
+Linear solvers based on Monte Carlo techniques are currently being
 studied as potentially resilient alternatives to conventional
@@
 the performance of the multilevel method against that of the
-traditional method and discuss applicabililty to various problems.
+traditional method and discuss applicability to various problems.
 

### 2014-01-29 doc/coppermtn_multilevel/abstract/abstract.tex
adding copper mountain abstract corrections

@@
-\documentclass[letterpaper,12pt]{article}
+\\documentclass[letterpaper,12pt]{article}
 \usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
@@
 solutions to linear systems by estimating the terms of the Neumann
-series formed by the Richardson iteration; thus providing a stochastic
+series formed by the Richardson iteration, thus providing a stochastic
 realization of a relaxation method. The random walks used to estimate
-the solution are a variant of Markov Chain Monte Carlo (MCMC) methods
+the solution are a variant of Markov Chain Monte Carlo (MCMC) methods,
 and recent work has aimed to adapt various techniques in MCMC to
@@
 applied to Monte Carlo linear solvers the multilevel technique
-resembles the multigrid method where the Monte Carlo scheme acts as a
-stochastic smoother. We will present both the Monte Carlo method for
-linear systems as well as the multilevel method. We will then compare
-the performance of the multilevel method against that of the
-traditional method and discuss applicability to various problems.
+resembles the multigrid method, where the Monte Carlo scheme acts as a
+stochastic smoother. We present both the Monte Carlo method for linear
+systems as well as the multilevel method. We then compare the
+performance of the multilevel method against that of the traditional
+method and discuss applicability to various problems.
 

### 2014-01-29 doc/ttsp2013/ttsp_2013.tex
adding mc theory to the ttsp paper

@@
-\documentclass[letterpaper,12pt]{article}
+\documentclass[letterpaper,11pt]{article}
 \usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
@@
 \author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton
-\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
+\\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}}
 }
@@
 
-In this paper, we present Monte Carlo Synthetic Acceleration (MCSA) as
-a linear solver technique for the simplified $P_N$ ($SP_N$)
-discretization of the Boltzmann neutron transport equation. Using the
-fully-formed linear operator for the transport problem, we explore
-solutions to the $SP_N$ equations with MCSA using a challenging light
-water reactor fuel assembly criticality calculation as the driving
-problem. Due to neutron scattering in the light water moderator,
-several difficulties arise when applying MCSA to these problems that
-were not observed in previous work where the method was applied to
-more opaque thermal radiaton transport systems. In order to
-effectively solve the $SP_N$ equations for the model problem, an
-algebraic preconditioning strategy is developed for MCSA where we
-utilize incomplete factorizations as preconditioners to achieve
-convergence.  Using the preconditioners, MCSA performance is compared
-with production impelementations of the preconditioned GMRES and
-BiCGSTAB methods. Although the algebraic preconditioning strategy
-presented in this work will not be practical from a high performance
-implementation standpoint with results demonstrating poor time to
-solution, it does indicate that the iterative performance of the
-method is competitive with the preconditioned Krylov methods when
-operator spectrum is properly conditioned.
+We present Monte Carlo Synthetic Acceleration (MCSA) as a linear
+solver technique for the simplified $P_N$ ($SP_N$) discretization of
+the Boltzmann neutron transport equation. Using the fully-formed
+linear operator for the transport problem, we explore solutions to the
+$SP_N$ equations with MCSA using a challenging light water reactor
+fuel assembly criticality calculation as the driving problem. Due to
+neutron scattering in the light water moderator, several difficulties
+arise when applying MCSA to these problems that were not observed in
+previous work where the method was applied to more opaque thermal
+radiaton transport systems. Of most significance, it was discovered
+that simply reducing the spectral radius of the Monte Carlo problem
+below unity is not enough to achieve convergence as outlined in
+previous work. In order to effectively solve the $SP_N$ equations for
+the model problem, an algebraic preconditioning strategy is developed
+for MCSA where we utilize incomplete factorizations as preconditioners
+to achieve convergence. Using the preconditioners, MCSA performance is
+compared with production implementations of the preconditioned GMRES
+and BiCGSTAB methods. Studies indicate that the iterative performance
+of the method is competitive with the preconditioned Krylov methods
+when operator spectrum is properly conditioned.
 
@@
 approximation could mean significant savings in both compute time and
-memory required.
+memory required and permit larger simulations to be performed on the
+available computing platforms.
 
-As machines begin to operate at hundreds of petaflops peak performance
-and beyond, trends toward reduced energy consumption will require
-incredibly high levels of concurrency to achieve the desired
-computation rates. The end result of these hardware changes is that
-the larger number of low-powered processors will be prone to both soft
-failures such as bit errors in floating point operations and hard
-failures where the data owned by that processor cannot be
-recovered. Because these failures are predicted to be common,
-resilient solver technologies are required to overcome these events at
-the application level. With solutions to deterministic transport
-problems based on Monte Carlo techniques, such issues are potentially
-alleviated by statistical arguments. In the case of soft failures,
-isolated floating point errors in the Monte Carlo simulation are
-absorbed within tally statistics while a hard failure is treated as a
-high variance event where some portion of the Monte Carlo histories
-lost.
+Looking towards future computing platforms, as machines begin to
+operate at hundreds of petaflops peak performance and beyond, trends
+toward reduced energy consumption will require incredibly high levels
+of concurrency to achieve the desired computation rates. The end
+result of these hardware changes is that the larger number of
+low-powered processors will be prone to both soft failures such as bit
+errors in floating point operations and hard failures where the data
+owned by that processor cannot be recovered. Because these failures
+are predicted to be common, resilient solver technologies are required
+to overcome these events at the application level. With solutions to
+deterministic transport problems based on Monte Carlo techniques, such
+issues are potentially alleviated by statistical arguments. In the
+case of soft failures, isolated floating point errors in the Monte
+Carlo simulation are absorbed within tally statistics while a hard
+failure is treated as a high variance event where some portion of the
+Monte Carlo histories lost.
 
-Considering these physics and hardware issues, this goal of this work
+Considering these physics and hardware issues, the goal of this work
 is to research and develop Monte Carlo Synthetic Acceleration methods
@@
 of the physics of interest and desire to identify areas in which
-improvements can be made. Using these numerical studies, we look
-forward to parallelizing them with the goal that they may be leveraged
-competitively on leadership-class computing platforms.
-
-In previous work \cite{evans_monte_2012}, we applied the Monte Carlo
-synthetic acceleration (MCSA) method the thermal radiation diffusion
-equation. For several transient problems we observed that MCSA
-performed competetively with preconditioned Krylov methods. In a
-continuation of these studies, we have applied the method light water
-reactor problems discretized by the $SP_N$ equations. This system has
-no time dependence, a large scattering component, and assymetry in the
-linear operator, creating a set of problems complementary to the
-transient, opaque, and symmetric systems of our previous analyis. In
-addition, solving the $SP_N$ equations in this way also breaks away
-from the $S_N$ forms of parallelism where spatial parallelism is
-achieved by an efficient parallel sweep, angular efficiency achieved
-by pipe-lining, and energy parallelism achieved by decoupling the
-groups. With the $SP_N$ equations as a full matrix system, we now can
-parallelize the problem as prescribed by the Monte Carlo algorithm,
-which may be significantly more scalable than current $S_N$ transport
-practices.
+improvements can be made. In previous work \cite{evans_monte_2012}, we
+applied the Monte Carlo synthetic acceleration (MCSA) method to the
+thermal radiation diffusion equation. For several transient problems
+we observed that MCSA performed competetively with preconditioned
+Krylov methods. In a continuation of these studies, we apply the
+method to light water reactor problems discretized by the $SP_N$
+equations. This system has no time dependence, a large scattering
+component, and assymetry in the linear operator, creating a set of
+problems complementary to the transient, opaque, and symmetric systems
+of our previous analyis. In addition, we apply a new fixed point
+iteration in the MCSA sequence as well as a new estimator in order to
+study their effects on convergence.
 
@@
 
+\subsection{Monte Carlo Synthetic Acceleration}
+\label{subsec:mcsa}
+In this section we briefly review the Monte Carlo method for linear
+systems and Monte Carlo Synthetic Acceleration. Given a linear
+problem:
+\begin{equation}
+  \ve{A} \ve{x} = \ve{b}\:,
+  \label{eq:linear_problem}
+\end{equation}
+where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
+that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
+\mathbb{R}^N$ is the solution vector, and $\ve{b} \in
+\mathbb{R}^N$. To form the Monte Carlo method we define the linear
+system adjoint to Eq~(\ref{eq:linear_problem}):
+\begin{equation}
+  \ve{A}^T \ve{y} = \ve{d}\:,
+  \label{eq:adjoint_linear_problem}
+\end{equation}
+where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
+vectors respectively and $\ve{A}^T$ is the adjoint operator for
+$\ve{A} \in \mathbb{R}^{N \times N}$ which we hten split to get:
+\begin{equation}
+  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
+  \label{eq:adjoint_split_system}
+\end{equation}
+where $\ve{H}^T$ is the adjoint iteration matrix. The spectral radius
+of $\ve{H}^T$ must remain less than 1 for convergence. By defining the
+following adjoint inner product equivalence \cite{spanier_monte_1969}:
+\begin{equation}
+  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
+  \ve{x} \rangle\:.
+  \label{eq:adjoint_operator_product}
+\end{equation}
+it follows that:
+\begin{equation}
+  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
+  \label{eq:adjoint_vector_relation}
+\end{equation}
+We can acquire the adjoint solution by forming the Neumann series by
+writing Eq~(\ref{eq:adjoint_split_system}) as:
+\begin{equation}
+  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
+  \label{eq:adjoint_split_system_2}
+\end{equation}
+which in turn yields the Neumann series using the adjoint operator:
+\begin{equation}
+  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
+  \label{eq:adjoint_neumann_series}
+\end{equation}
+We expand this summation to again yield a series of transitions that
+can be approximated by a Monte Carlo random walk sequence by forming
+the Neumann series in reverse order:
+\begin{equation}
+  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
+  \label{eq:adjoint_neumann_solution}
+\end{equation}
+We can readily build an estimator for the adjoint solution from this
+series expansion, but we instead desire the solution to
+Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
+$\ve{d}$, and therefore we require two constraints to close the
+system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
+constraint and as a second constraint we select:
+\begin{equation}
+  \ve{d} = \boldsymbol{\delta}_j\:,
+  \label{eq:adjoint_second_constraint}
+\end{equation}
+where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
+$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
+we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
+constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
+convenient outcome:
+\begin{equation}
+  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
+  \boldsymbol{\delta}_j \rangle = x_j \:,
+  \label{eq:inner_product_constraint}
+\end{equation}
+meaning that if we compute the inner product of the original source
+and the adjoint solution using a delta function source, we recover one
+component of the original solution.
+
+In terms of radiation transport, this adjoint method is equivalent to
+a traditional forward method where the initial state $i_0$ of the
+random walk is determined by sampling the source vector $\ve{b}$ with
+probabilities:
+\begin{equation}
+  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
+  \label{eq:adjoint_source_probability}
+\end{equation}
+with a random walk starting weight of:
+\begin{equation}
+  W_0 = ||\ve{b}||_1 \frac{b_i}{|b_i|}\:,
+  \label{eq:adjoint_starting_weight}
+\end{equation}
+which gives the additional useful relation:
+\begin{equation}
+  b_{i_0} = W_0 P_{(i_0=i)}\:.
+  \label{eq:adjoint_source_definition}
+\end{equation}
+We build weights and probabilities for the random walk using the
+\textit{adjoint Neumann-Ulam decomposition} of $\ve{H}$:
+\begin{equation}
+  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
+  \label{eq:adjoint_neumann_ulam}
+\end{equation}
+where we are forming the decomposition with respect to the transpose
+of $\ve{H}$. $\ve{P}$ is defined element-wise as:
+\begin{equation}
+  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
+  \label{eq:adjoint_probability}
+\end{equation}
+such that we expect to select a new state, $j$, from the current state
+in the random walk, $i$, by sampling column-wise (or row-wise if an
+adjoint probability matrix is formed). Per
+Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
+defined as:
+\begin{equation}
+  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
+  \label{eq:adjoint_weight}
+\end{equation}
+Using the decomposition we can then define an expectation value for
+the adjoint method. The contribution to the solution in state $j$ from
+a particular random walk permutation of $k$ events is given by
+\textit{collision estimator}:
+\begin{equation}
+  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
+  \label{eq:adjoint_permutation_contribution}
+\end{equation}
+where the Kronecker delta indicates that the tally contributes only in
+the current state, $i_m$, of the random walk. The expectation value
+using all permutations is then:
+\begin{equation}
+  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:,
+  \label{eq:adjoint_expectation_value}
+\end{equation}
+which, if expanded utilizing Eq~(\ref{eq:adjoint_source_definition})
+to insert the source term, directly recovers the exact solution:
+\begin{equation}
+  \begin{split}
+    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
+    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
+    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
+  \end{split}
+  \label{eq:adjoint_expectation_expansion}
+\end{equation}
+therefore, also providing an unbiased Monte Carlo estimate of the
+solution. It should be noted here that
+Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
+component of our desired solution vector when really what we desire is
+the entire solution vector. In an adjoint Monte Carlo simulation using
+this estimator, the $w_{ij}$ elements that are added into the tally
+for each state are only selected if/when the random walk currently
+resides in that state. Much like a mesh tally in a particle transport
+simulation, we have $N$ simultaneous tallies for $\ve{A} \in
+\mathbb{R}^{N \times N}$ that will yield the entire solution vector.
+
+In addition to the collision estimator, an additional estimator is
+available due to the work of Okten \cite{okten_solving_2005} that uses
+the method of expected values as a means to improve the Monte Carlo
+estimate. As outlined by Spanier and Gelbard
+\cite{spanier_monte_1969}, the method of expected values is a
+deterministic averaging of events that may potentially occur in the
+Monte Carlo random walk sequence. Okten applied this principle
+directly to discrete Monte Carlo by forming the \textit{expected value
+  estimator}\footnote{The variance for this estimator is discussed in
+  Appendix~\ref{chap:estimator_variance}.} for a random walk of $k$
+events:
+\begin{equation}
+  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
+  \label{eq:expected_value_estimator}
+\end{equation}
+where now the contribution of the iteration matrix is
+deterministically averaged at step $m$ over all potential states $j$
+that may be reached from the current state $i_m$. Via Okten, the
+estimator can be shown to be unbiased through a comparison to the
+collision estimator. We can first rewrite the summation in
+Eq~(\ref{eq:expected_value_estimator}):
+\begin{equation}
+  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
+  \delta_{i_m,i} h_{ji}\,
+  \label{eq:unbiased_eval_1}
+\end{equation}
+where $N$ is the number of states in the system. Immediately, we see
+the collision estimator as defined by
+Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
+the expectation value as:
+\begin{equation}
+  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
+  \label{eq:unbiased_eval_2}
+\end{equation}
+which is equivalently is the $j^{th}$ component of
+Eq~(\ref{eq:richardson_split}):
+\begin{equation}
+  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
+  \label{eq:unbiased_eval_2}
+\end{equation}
+and is therefore an unbiased estimate. Compared to the collision
+estimator, the expected value estimator provides additional
+information at every step of the random walk, yielding potentially
+better statistics with the same amount of transport
+work. Conveniently, even if no Monte Carlo histories are computed, the
+expected value estimator still deterministically computes the first
+term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
+estimator will provide no information.
+
 \subsection{MCSA Solution of the $SP_N$ Equations}
@@
 again indicates the eigenvalue iteration. We refer the reader to
-\cite{evans_monte_2012} for a more rigorous defintion of the method.
+\cite{evans_monte_2012} for more a more complete definition of the
+Monte Carlo method for linear systems and MCSA.
 

### 2014-03-27 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding tech note for multilevel mc

\documentclass[note]{TechNote}
\usepackage[centertags]{amsmath}
\usepackage{amssymb,amsthm}
\usepackage[mathcal]{euscript}
\usepackage{tabularx}
\usepackage{cite}
\usepackage{c++}
\usepackage{tmadd,tmath}

%%---------------------------------------------------------------------------%%
\begin{document}

\refno{CSMD-00-000}
\subject{A Multilevel Monte Carlo Solver for Linear Systems}

\TIname{Mathematics and Computer Science Division}
\groupname{Computational Engineering and Energy Sciences Group}
\from{Stuart R. Slattery}
\date{\today}
\audience{
  \email{Stuart Slattery}{slatterysr@ornl.gov} \\
  \email{Tom Evans}{evanstm@ornl.gov} \\
  \email{Steven Hamilton}{hamiltonsp@ornl.gov}
}

%%---------------------------------------------------------------------------%%
\opening

\begin{abstract}
  Monte Carlo solvers for linear systems have been demonstrated to
  perform poorly for strongly elliptic problems. This poor performance
  is primarily due to the fact that both a large number of samples are
  required to obtain a good statistical error and that these samples
  require a large amount of time to compute. As a mechanism to reduce
  the computational complexity for such problems and thereby improve
  the figure of merit of the calculation, multilevel Monte Carlo has
  been introduced in finance problems, the solution of stochastic
  partial differential equations, and other algorithms that leverage
  Markov chain Monte Carlo. We adapt these ideas to form a multilevel
  Monte Carlo solver for linear systems which, in the form presented
  here, is effectively a stochastic realization of a geometric
  multigrid solver. Numerical studies indicate that the new multilevel
  method can reduce the time required to achieve a certain statistical
  error in the solution by at least two orders of magnitude, thereby
  dramatically reducing the computational complexity of the
  problem. Furthermore, the general formulation presented here
  indicates that this methodology is not restricted to geometric
  formulations of the multigrid method and could additionally be
  formulated using algebraic techniques.
\end{abstract}

%%---------------------------------------------------------------------------%%
\section{Introduction}
\label{sec:introduction}
Monte Carlo solvers for linear systems have been in existence for
decades as a stochastic alternative to iterative methods
\cite{forsythe_matrix_1950,wasow_note_1952,halton_sequential_1962,hammersley_monte_1964,spanier_monte_1969}. However,
these methods have failed to gain popularity both in the mathematics
and applications community partly due to their slow convergence bound
by the central limit theorem. Recent work has indicated that when used
as an acceleration in the Monte Carlo Synethetic Acceleration (MCSA)
method, exponential convergence rates may be achieved that are
competetive with contemporary iterative methods
\cite{evans_residual_2003,evans_monte_2009,evans_monte_2012,slattery_phd_2013}.
However, in this recent work it was discovered that for physics
problems that are largely elliptic (e.g. neutron transport in a light
water reactor), convergence of the Monte Carlo method is extremely
slow and prohibitive for the solution of larger systems. One avenue to
improve the time to solution for these calculations and to enable the
solution of more difficult problems is to study preconditioning
strategies as in \cite{slattery_phd_2013}. Another approach is to
instead focus on improving the time complexity of the Monte Carlo
sequence independent of the condition number of the linear problem.

Recent work in Monte Carlo methods for problems in finance and
stochastic partial differential equations has indicated that the
computationaly complexity of the problem can be dramatically reduced
by incorporating multigrid concepts into the solution scheme
\cite{heinrich_2001,giles_2008,cliffe_2011}. In this work, we adapt
those ideas and apply them to the Monte Carlo problem for linear
systems as a means of reducing the computational complexity of the
algorithm. To begin, we first introduce the elliptic model problem for
our numerical experiments and compare the behavior of the Monte Carlo
method to traditional iterative smoothers that would be used with the
multigrid method. Next, we present the multilevel Monte Carlo method
for linear systems using a general algebraic formulation. Finally, we
present results using the model problem the demonstrate the
superiority of the multilevel method.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Solver Fourier Analysis}
\label{sec:fourier_analysis}
The numerical analysis presented here closely follows that presented
in \cite{briggs_multigrid} for multigrid problems. For this analysis,
we will use the following one-dimensional, homogenous model problem:
\begin{equation}
  \nabla^2 x = 0\:.
  \label{eq:model_problem}
\end{equation}
We discretize the problem into $N$ discrete points
where now $\ve{x} \in \mathbb{R}^N$ with boundary conditions:
\begin{equation}
  \ve{x}_1 = 0,\ \ \ve{x}_N = 0\:.
  \label{eq:boundary_conditions}
\end{equation}
The Laplacian is discretized using a standard second-order finite
difference with a grid spacing of one:
\begin{equation}
  (\nabla \ve{u})_i = \ve{u}_{i-1} - 2 \ve{u}_{i} + \ve{u}_{i+1}\:,
  \label{eq:discrete_laplacian}
\end{equation}
which then gives the following linear problem:
\begin{equation}
  \ve{A}\ve{x} = \ve{0}\:.
  \label{eq:linear_problem}
\end{equation}
To bound the spectral radius of the problem we will use a Jacobi
preconditioner:
\begin{equation}
  \ve{M} = diag(\ve{A})\:,
  \label{eq:preconditioner}
\end{equation}
such that we instead solve the following linear problem:
\begin{equation}
  \ve{M}^{-1}\ve{A}\ve{x} = \ve{0}\:.
  \label{eq:precond_problem}
\end{equation}

To elucidate the effect of a given solution technique on a given error
mode in the problem, we can assign an initial guess of $\ve{x}^0$ to
be a chosen Fourier mode:
\begin{equation}
  \ve{x}^0_i = \sin\Bigg( \frac{ik\pi}{N} \Bigg)\:,
  \label{eq:fourier_mode}
\end{equation}
where $\ve{x}^0_i$ is the $i^{th}$ component of the initial guess and
k is the wave number of the chosen Fourier mode. First, we will look at
the performance Richardson's iteration as the smoother which is
equivalently a Jacobi iteration when Jacobi preconditioning is used:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I}-\ve{M}^{-1}\ve{A})\ve{x}^k\:,
\end{equation}
where $k$ is the iteration index. Figure~\ref{fig:richardson} gives
infinity norm of the error in the solution vector\footnote{The
  solution to the homogenous problem is zero and therefore
  $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$} as a function of
iterations for wave numbers of 1, 5, and 10 on a grid with N = 100.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{richardson.png}
  \end{center}
  \caption{\textbf{Convergence of Richardson's iteration as a function
      of iteration for a grid of size N = 100.} \textit{Richardson's
      iteration performs better for larger wave numbers and therefore
      more oscillatory modes.}}
  \label{fig:richardson}
\end{figure}
Immediately we note that the larger the wave number the better
Richardson's iteration performs. Per the spectral analysis in
\cite{briggs_multigrid}, this iteration sequence perfoms better for
higher wave numbers as the eigenvalue spectrum spans a smaller space
then less oscillatory modes. This behavior motivated the multigrid
approach where moving a smooth mode to a coarser grid makes that mode
appear more oscillatory relative to that coarser grid, thus improving
convergence for that particular mode.

Next we perform the same calculations using the adjoint Monte Carlo
solver presented in \cite{evans_monte_2012}. Before doing this
however, we must first modify the linear problem as the Monte Carlo
solver is a direct method and therefore a homogenous problem with a
right hand side of $\ve{0}$ will yield no samples. Instead, we will
solve the residual problem:
\begin{equation}
  \ve{A}\ve{d} = \ve{r}\:,
  \label{eq:residual_problem}
\end{equation}
where the residual of the homogenous problem is:
\begin{equation}
  \ve{r} = -\ve{A}\ve{x}^0\:,
  \label{eq:homogenous_residual}
\end{equation}
and the solution is computed as:
\begin{equation}
  \ve{x} = \ve{x}^0 + \ve{d}\:.
  \label{eq:residual_solution}
\end{equation}
Forming the problem in this way lets us directly apply the adjoint
Monte Carlo method to the homogenous problem and then apply the
effective correction, $\ve{d}$, to the initial guess to give the
solution. This approach is also equivalent to performing a single
iteration of Halton's method \cite{halton_sequential_1962}. Using this
formulation we can again solve the model problem with wave numbers of
1, 5 and 10 on a grid of size N=1 but this time we vary the number of
histories used to compute the solution instead of the number of
iterations. Figure~\ref{fig:adjoint_mc} gives the results of this
calculation.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{adjoint_mc.png}
  \end{center}
  \caption{\textbf{Convergence of the adjoint Monte Carlo method as a
      function of sampled histories for a grid of size N = 100.}
    \textit{Adjoint Monte Carlo performs better for smaller wave
      numbers and therefore smoother modes.}}
  \label{fig:adjoint_mc}
\end{figure}
Surprisingly, the Monte Carlo method performs better for smooth modes
then more oscillatory modes\footnote{This behavior was also observed
  for the forward Monte Carlo method presented in
  \cite{evans_monte_2012}}. The results of these calculations are
counterintuitive given the fact that the Monte Carlo solver is
effectively a stochastic realization of of Richardson's iteration and
therefore one should expect the same spectral behavior from the
results.

Looking at the timing results in Table~\ref{tab:mc_timing}, we see
that the behavior of the Monte Carlo solver is in fact consistent with
expectations.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cc}\hline\hline
      \multicolumn{1}{c}{\textbf{Wave Number}} & 
      \multicolumn{1}{c}{\textbf{Time per History (s)}} \\
      \hline
      1 & 1 \\
      5 & 0.85 \\
      10 & 0.83 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Normalized average CPU time per history.}}
  \label{tab:mc_timing}
\end{table}
Timing results show that the average time required to compute an
entire history in the Monte Carlo solver decreases as a function of
wave number. We showed analytically in \cite{slattery_2013} that the
length of the random walk is equivalent to the number of Richardson
iterations that would be required to achieve a given convergence
criteria. In Figure~\ref{fig:richardson}, we see that fewer iterations
are required to converge larger wave numbers and therefore we should
also expect shorter random walks in the Monte Carlo solver and
therefore a faster time to solution as observed in
Table~\ref{tab:mc_timing}.

%%---------------------------------------------------------------------------%%
\section{Multilevel Monte Carlo Algorithm}
\label{sec:algorithm}

%%---------------------------------------------------------------------------%%
\section{Results}
\label{sec:results}

%%---------------------------------------------------------------------------%%
\bibliographystyle{ieeetr}
\bibliography{references}

\closing
\caution
\end{document}

### 2014-03-27 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding fourier mode analysis to multilevel note

@@
 \usepackage{tmadd,tmath}
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
   solution to the homogenous problem is zero and therefore
-  $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$} as a function of
+  $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$.} as a function of
 iterations for wave numbers of 1, 5, and 10 on a grid with N = 100.
@@
 higher wave numbers as the eigenvalue spectrum spans a smaller space
-then less oscillatory modes. This behavior motivated the multigrid
+then less oscillatory modes. This behavior motivates the multigrid
 approach where moving a smooth mode to a coarser grid makes that mode
@@
   for the forward Monte Carlo method presented in
-  \cite{evans_monte_2012}}. The results of these calculations are
+  \cite{evans_monte_2012}.}. The results of these calculations are
 counterintuitive given the fact that the Monte Carlo solver is
@@
 therefore a faster time to solution as observed in
-Table~\ref{tab:mc_timing}.
+Table~\ref{tab:mc_timing}. However, this does not indicate why the
+Monte Carlo solvers perform better for smoother modes rather than
+oscillatory modes. Fortunately, the explanation for this behavior is
+simple. Consider the plots for the Fourier modes with $k = 1$ and $k =
+10$ given in Figure~\ref{fig:fourier_modes}.
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{mode_1.pdf}
+    \caption{\textbf{Fourier mode with $k = 1$.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.5\textwidth}
+    \includegraphics[width=\textwidth]{mode_10.pdf}
+    \caption{\textbf{Fourier mode with $k = 10$.}}
+  \end{subfigure}
+  \caption{\textit{The more oscillatory the mode the finer the
+      structure of the error. Finer structures require more samples to
+      resolve them.}}
+  \label{fig:fourier_modes}
+\end{figure}
+The higher the wave number is the more oscillatory the mode
+appears. To resolve these fine structures in the error we therefore
+naturally require more Monte Carlo samples to distinguish them from
+one another.
+
+Based on this, it may not seem that applying Monte Carlo in a
+multigrid context will be useful considering that error modes appear
+more oscillatory as the grid is coarsened. This would indicate that in
+order to achieve the same amount of error on a coarse grid
+representation of the problem, one would be required to compute many
+more samples to resolve the resulting finer structures. However, we
+may gain a computational advantage for two reasons. First, consider
+the timing results in Table~\ref{tab:mc_timing}. As the error
+structure becomes more oscillatory, the time it takes to compute a
+sample decreases. Second, if we coarsen the grid the time it takes to
+compute a sample will also decrease due to the decreased problem
+size. Therefore, it is plausible that we might observe an improvement
+in run times for the Monte Carlo problem by using a multigrid
+approach. This then motivates applying recently developed techniques
+in multilevel Markov chain Monte Carlo to the adjoint solver as a
+potential means of accelerating time to solution.
 

### 2014-03-27 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding multilevel results for m=2

@@
 \usepackage{subcaption}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 algorithm. To begin, we first introduce the elliptic model problem for
-our numerical experiments and compare the behavior of the Monte Carlo
-method to traditional iterative smoothers that would be used with the
-multigrid method. Next, we present the multilevel Monte Carlo method
-for linear systems using a general algebraic formulation. Finally, we
-present results using the model problem the demonstrate the
-superiority of the multilevel method.
+our numerical experiments and compare the spectral behavior of the
+Monte Carlo method to traditional iterative smoothers that would be
+used with the multigrid method. Next, we present the multilevel Monte
+Carlo method for linear systems using a general algebraic
+formulation. Finally, we present results using the model problem the
+demonstrate the superiority of the multilevel method in terms of time
+to solution.
 
@@
 \label{sec:fourier_analysis}
-The numerical analysis presented here closely follows that presented
-in \cite{briggs_multigrid} for multigrid problems. For this analysis,
-we will use the following one-dimensional, homogenous model problem:
+We would first like to analyze the behavior of Monte Carlo solvers in
+the context of error modes for a given model problem with the
+numerical analysis presented here closely following that presented in
+\cite{briggs_multigrid}. For this analysis, we will use the following
+one-dimensional, homogenous model problem:
 \begin{equation}
@@
 \end{equation}
-
 To elucidate the effect of a given solution technique on a given error
@@
 where $\ve{x}^0_i$ is the $i^{th}$ component of the initial guess and
-k is the wave number of the chosen Fourier mode. First, we will look at
-the performance Richardson's iteration as the smoother which is
-equivalently a Jacobi iteration when Jacobi preconditioning is used:
+k is the wave number of the chosen Fourier mode.
+
+Monte Carlo solvers are effectively a stochastic realization of
+Richardson's iteration and therefore we will first look at the
+performance ofRichardson's iteration as a smoother:
 \begin{equation}
@@
 \end{equation}
-where $k$ is the iteration index. Figure~\ref{fig:richardson} gives
-infinity norm of the error in the solution vector\footnote{The
-  solution to the homogenous problem is zero and therefore
-  $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$.} as a function of
-iterations for wave numbers of 1, 5, and 10 on a grid with N = 100.
+where $k$ is the iteration index and which is equivalently a Jacobi
+iteration when Jacobi preconditioning is
+used. Figure~\ref{fig:richardson} gives infinity norm of the error in
+the solution vector\footnote{The solution to the homogenous problem is
+  zero and therefore $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$.} as
+a function of iteration for wave numbers of 1, 5, and 10 on a grid
+with $G = 100$.
 \begin{figure}[h!]
@@
   \caption{\textbf{Convergence of Richardson's iteration as a function
-      of iteration for a grid of size N = 100.} \textit{Richardson's
+      of iteration for a grid of size $G = 100$.} \textit{Richardson's
       iteration performs better for larger wave numbers and therefore
@@
 formulation we can again solve the model problem with wave numbers of
-1, 5 and 10 on a grid of size N=1 but this time we vary the number of
-histories used to compute the solution instead of the number of
-iterations. Figure~\ref{fig:adjoint_mc} gives the results of this
-calculation.
+1, 5 and 10 on a grid of size $G = 100$ but this time we vary the
+number of histories used to compute the solution instead of the number
+of iterations. Figure~\ref{fig:adjoint_mc} gives the results of these
+calculations.
 \begin{figure}[h!]
@@
   \caption{\textbf{Convergence of the adjoint Monte Carlo method as a
-      function of sampled histories for a grid of size N = 100.}
+      function of sampled histories for a grid of size $G = 100$.}
     \textit{Adjoint Monte Carlo performs better for smaller wave
@@
 Surprisingly, the Monte Carlo method performs better for smooth modes
-then more oscillatory modes\footnote{This behavior was also observed
+than more oscillatory modes\footnote{This behavior was also observed
   for the forward Monte Carlo method presented in
@@
 counterintuitive given the fact that the Monte Carlo solver is
-effectively a stochastic realization of of Richardson's iteration and
+effectively a stochastic realization of Richardson's iteration and
 therefore one should expect the same spectral behavior from the
@@
 that the behavior of the Monte Carlo solver is in fact consistent with
-expectations.
+these expectations.
 \begin{table}[h!]
@@
 in run times for the Monte Carlo problem by using a multigrid
-approach. This then motivates applying recently developed techniques
-in multilevel Markov chain Monte Carlo to the adjoint solver as a
-potential means of accelerating time to solution.
+approach.
 
@@
 \label{sec:algorithm}
+A potential improvement in computational complexity motivates applying
+techniques in multilevel Markov chain Monte Carlo recently developed
+by Heinrich \cite{heinrich_2001} and Giles \cite{giles_2008}. We start
+first with the standard Monte Carlo estimator for the $i^{th}$
+component of the solution vector:
+\begin{equation}
+  \hat{\ve{x}}_i = \frac{1}{N} \sum_{m=1}^N x^m_i\:,
+  \label{eq:standard_estimator}
+\end{equation}
+where $i \in [1,N]$, $N$ samples are made and $x^m_i$ is the $m^{th}$
+observation. We next consider representing the problem in multiple
+levels designated by $l$ with each one finer than the next such that
+there are $L$ total levels and $l=L$ the coarsest level. We can
+combine expectation values of the estimates from each of these levels
+using the linearity of expectation such that:
+\begin{equation}
+  E(\ve{x}_{0}) = E(\ve{x}_{L}) + E(\ve{x}_{L-1} - \ve{x}_{L}) +
+  E(\ve{x}_{L-2} - \ve{x}_{L-1}) + \dots + E(\ve{x}_{0} -
+  \ve{x}_{(1)})\:,
+  \label{eq:linearity_of_expectation}
+\end{equation}
+where $E(\ve{x}_{L})$ is the expectation value of the solution vector
+at the coarsest level, $l = L$.  If we rewrite this summation as:
+\begin{equation}
+  E(\ve{x}_{0}) = E(\ve{x}_{L}) + \sum_{l=1}^L E(\ve{x}_{l} -
+  \ve{x}_{l+1})\:,
+  \label{eq:linearity_of_expectation_sum}
+\end{equation}
+we can then build a new \textit{multilevel estimator} for the $l^{th}$
+level as:
+\begin{equation}
+    \hat{\ve{y}}_{l} = \frac{1}{N_l} \sum_{m=1}^{N_l} (x^m_{l} -
+    x^m_{l+1})\:,
+    \label{eq:level_estimate}
+\end{equation}
+where $N_l$ are the number of samples computed at the level and
+$x^m_{l+1} = 0$ for $l=L$. Estimates from all levels can then be
+combined to give the final solution as:
+\begin{equation}
+  \hat{\ve{x}} = \sum_{l=0}^L \hat{\ve{y}}_{l}\:.
+  \label{eq:multilevel_estimator}
+\end{equation}
+It is critical to note here that the term $(x^m_{l} - x^m_{l+1})$
+in Eq~(\ref{eq:level_estimate}) consists of observations from the same
+Markov chain computed over the space of the $l^{th}$ level.
+
+Although multigrid methods for linear algebra problems indicate that a
+coarsening parameter of $M = 2$ is optimal such that every level is
+half the size of the finer level. However, Giles reports that values
+of 4 and larger give optimal perfomance for these particular
+estimators. In addition, as mentioned in
+\S~\ref{sec:fourier_analysis}, because coarsening the problem makes
+the modes appear more oscillatory, we therefore must compute more
+samples at coarser levels to balance the error amongst the levels. For
+this work we will use the result reported by Hienrich for computing
+the number of samples at each level:
+\begin{equation}
+  N_l = M^{-3(L-l-1)/2}N\:.
+  \label{eq:level_histories}
+\end{equation}
+It should be noted that Giles does not compute this quantity a priori
+but instead first estimates the variance at each level and then
+computes $N_l$ based on that variance.
+
+Using the estimator given by Eq~(\ref{eq:multilevel_estimator}) we can
+now define a multilevel Monte Carlo method for linear systems based on
+the adjoint Monte Carlo solver given in \cite{evans_monte_2012}. To do
+this, we must consider how to compute the expectation value
+$E(\ve{x}_{l} - \ve{x}_{l+1})$ at each level in the problem. We are
+required to construct observations on both grids from the same Markov
+chain. Computing observations on the $l^{th}$ grid in this case is
+simply the procedure for solving the adjoint Monte Carlo problem. In
+this procedure we can simply tally observations on the grid for the
+$l+1$ level through the use of a linear interpolation operator such
+that a tally in some state of level $l$ is also a tally in some state
+of level $l+1$ with that state determined by the linear operator such
+that the weight of the observation is preserved. If this is the case,
+then we can actually avoid tallying observations of the coarse grid
+altogether and simply apply the linear interpolation operator the fine
+grid results once all samples have been computed.
+
+Using this idea we then define a \textit{prolongation operator},
+$\ve{P}_{l+1}^l$, which maps a vector defined on grid $l+1$ to a
+vector defined on grid $l$ and a \textit{restriction operator},
+$\ve{R}_l^{l+1}$, which mpas a solution defined on grid $l$ to a
+vector defined on grid $l+1$. We can then define the following
+equivalent expecation value for a given level $l$:
+\begin{equation}
+  E(\ve{x}_{l} - \ve{x}_{(l+1)}) = \Big(\ve{I} - \ve{P}_{l+1}^l
+  \ve{R}_l^{l+1}\Big) \hat{\ve{x}}_{l}\:,
+  \label{eq:level_expectation}
+\end{equation}
+with $\hat{\ve{x}}_{l}$ given Eq~(\ref{eq:standard_estimator}) as the
+solution on grid $l$ and where now the action of the restriction
+operator maps the tally to the coarse level to compute the
+observations on that grid and the action of the prolongation operator
+on the coarse tally maps the result back to the fine
+level. Eq~(\ref{eq:level_expectation}) then gives the following
+multilevel Monte Carlo algorithm for linear systems now with arbitrary
+right hand side vector, $\ve{b}$:
+\begin{algorithm}[h!]
+  \caption{Multilevel Adjoint Monte Carlo}
+  \label{alg:mlamc}
+  \begin{algorithmic}[1]
+
+    \For{ l = 0...L }
+    \State $\ve{A}_l \ve{x}_l = \ve{b}_l$
+    \Comment{Solve the $l^{th}$ level problem with adjoint Monte
+      Carlo}
+    \If{ l != L }
+    \State $\ve{x}_l = (\ve{I} - \ve{P}_{l+1}^l\ve{R}_l^{l+1}) \ve{x}_{l}$
+    \Comment{Apply the multilevel tally}
+    \EndIf
+    \EndFor
+
+    \For{ l = L...1 }
+    \State $\ve{x}_{l-1} = ( \ve{I} + \ve{P}_{l}^{l-1} ) \ve{x}_{l}$
+    \Comment{Collapse the tallies to the finest grid}
+    \EndFor
+
+  \end{algorithmic}
+\end{algorithm}
+This algorithm permits each of the levels to be computed independently
+and then combined appropriately with the prolongation and restriction
+operators. Like other multigrid algorithms for linear system, we are
+required to produce a representation of the linear operator on each
+level, $\ve{A}^l$, as well as the right hand side, $\ve{b}^l$. The
+formulation of Algorithm~\ref{alg:mlamc} places no restrictions on the
+form of $\ve{A}^l$, $\ve{b}^l$, $\ve{P}_{l+1}^l$, or $\ve{R}_l^{l+1}$
+and therefore could be used with any multigrid scheme including those
+arising from algebraic multigrid formulations.
 
@@
 \label{sec:results}
+To demonstrate the multilevel Monte Carlo algorithm we solve the
+Poisson problem presented in \S~\ref{sec:fourier_analysis}, this time
+with a grid of size $G = 1024$. For the coarsening parameter we will
+first use a value of $M = 2$. In addition, we will introduce the
+following standard figure of merit metric to assess the quality of the
+Monte Carlo solver:
+\begin{equation}
+  FOM = \frac{1}{||\ve{e}||^2_{\infty} T}\:,
+  \label{eq:figure_of_merit}
+\end{equation}
+where $||\ve{e}||_{\infty}$ is a measure of the variance of the
+calculation and $T$ is the time required to solve the problem. The
+larger the figure of merit, the more useful the solution scheme
+is. For example, if two solutions schemes arrive at the same error but
+the first calculation takes twice as long as the second, the figure of
+merit will be twice as large for the second calculation.
+
+We again perform calculations with initial guesses of wave numbers 1,
+5 and 10 with Tables~\ref{tab:k1_results}, \ref{tab:k5_results}, and
+\ref{tab:k10_results} giving the results of these calculations. In
+these tables, the RFOM column indicates the relative figure of merit
+which has been normalized to the single level calculation.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      1 & 10,000 & 0.022 & 119.1 & 1 \\
+      2 & 13,535 & 0.026 & 72.0 & 1.14 \\
+      3 & 14,785 & 0.035 & 33.2 & 1.38 \\
+      4 & 15,226 & 0.039 & 13.7 & 2.63 \\
+      5 & 15,382 & 0.027 & 5.5 & 13.86 \\
+      6 & 15,437 & 0.036 & 2.3 & 19.21 \\
+      7 & 15,456 & 0.045 & 0.98 & 28.00 \\
+      8 & 15,462 & 0.081 & 0.41 & 20.66 \\
+      9 & 15,464 & 0.267 & 0.17 & 11.72 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Multilevel Monte Carlo results for $k = 1$ with $M
+      = 2$, $N = 10,000$ and run times reported in seconds.}}
+  \label{tab:k1_results}
+\end{table}
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      1 & 10,000 & 0.668 & 101.6 & 1 \\
+      2 & 13,535 & 0.381 & 60.0 & 5.21 \\
+      3 & 14,785 & 0.396 & 28.0 & 10.34 \\
+      4 & 15,226 & 0.599 & 11.9 & 10.64 \\
+      5 & 15,382 & 0.638 & 4.7 & 23.60 \\
+      6 & 15,437 & 1.052 & 1.8 & 23.15 \\
+      7 & 15,456 & 1.070 & 0.67 & 59.16 \\
+      8 & 15,462 & 1.180 & 0.23 & 144.10 \\
+      9 & 15,464 & 2.130 & 0.10 & 99.95 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Multilevel Monte Carlo results for $k = 5$ with $M
+      = 2$, $N = 10,000$ and run times reported in seconds.}}
+  \label{tab:k5_results}
+\end{table}
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      1 & 10,000 & 1.81 & 98.6 & 1 \\
+      2 & 13,535 & 1.67 & 59.6 & 1.94 \\
+      3 & 14,785 & 2.41 & 27.1 & 2.05 \\
+      4 & 15,226 & 3.20 & 11.5 & 2.74 \\
+      5 & 15,382 & 1.85 & 4.93 & 19.14 \\
+      6 & 15,437 & 3.89 & 1.95 & 10.95 \\
+      7 & 15,456 & 3.79 & 0.78 & 28.68 \\
+      8 & 15,462 & 7.08 & 0.30 & 21.62 \\
+      9 & 15,464 & 11.7 & 0.11 & 21.45 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
+      = 2$, $N = 10,000$ and run times reported in seconds.}}
+  \label{tab:k10_results}
+\end{table}
 

### 2014-03-27 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding multilevel results for m=2

@@
 a function of iteration for wave numbers of 1, 5, and 10 on a grid
-with $G = 100$.
+with $J = 100$.
 \begin{figure}[h!]
@@
   \caption{\textbf{Convergence of Richardson's iteration as a function
-      of iteration for a grid of size $G = 100$.} \textit{Richardson's
+      of iteration for a grid of size $J = 100$.} \textit{Richardson's
       iteration performs better for larger wave numbers and therefore
@@
 formulation we can again solve the model problem with wave numbers of
-1, 5 and 10 on a grid of size $G = 100$ but this time we vary the
+1, 5 and 10 on a grid of size $J = 100$ but this time we vary the
 number of histories used to compute the solution instead of the number
@@
   \caption{\textbf{Convergence of the adjoint Monte Carlo method as a
-      function of sampled histories for a grid of size $G = 100$.}
+      function of sampled histories for a grid of size $J = 100$.}
     \textit{Adjoint Monte Carlo performs better for smaller wave
@@
 Poisson problem presented in \S~\ref{sec:fourier_analysis}, this time
-with a grid of size $G = 1024$. For the coarsening parameter we will
-first use a value of $M = 2$. In addition, we will introduce the
+with a grid of size $J = 1024$. For the coarsening parameter we will
+first use a value of $M = 2$. For simplicity, we will assume a basic
+geometric multigrid formulation of the linear problem per
+\cite{briggs_multigrid} with the the following prolongation operator
+for $M=2$:
+\begin{gather}
+  \begin{subequations}
+    x^h_{2j} = x_j^{2h} \\
+    x^h_{2j+1} = \frac{1}{2} ( x_j^{2h} + x_{j+1}^{2h} )\:,
+  \end{subequations}
+  \label{eq:m2_prolongation}
+\end{gather}
+for $0 \leq j \leq \frac{J}{2} - 1$ and the restriction operator:
+\begin{equation}
+  x^{2h}_j = \frac{1}{4} ( x_{2j-1}^{h} + 2 x_{2j}^{2h} + x_{2j+1}^{2h} )
+  \label{eq:m2_restriction}
+\end{equation}
+for $1 \leq j \leq \frac{J}{2} - 1$.
+
+To measure performance of the algorithm, we will introduce the
 following standard figure of merit metric to assess the quality of the

### 2014-03-27 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
finished draft of multilevel note

@@
   Monte Carlo solver for linear systems which, in the form presented
-  here, is effectively a stochastic realization of a geometric
-  multigrid solver. Numerical studies indicate that the new multilevel
-  method can reduce the time required to achieve a certain statistical
-  error in the solution by at least two orders of magnitude, thereby
+  here, is effectively a stochastic realization of a multigrid
+  solver. Numerical studies indicate that the new multilevel method
+  can reduce the time required to achieve a certain statistical error
+  in the solution by at least two orders of magnitude, thereby
   dramatically reducing the computational complexity of the
@@
 by the central limit theorem. Recent work has indicated that when used
-as an acceleration in the Monte Carlo Synethetic Acceleration (MCSA)
+as an acceleration in the Monte Carlo Synthetic Acceleration (MCSA)
 method, exponential convergence rates may be achieved that are
-competetive with contemporary iterative methods
+competitive with contemporary iterative methods
 \cite{evans_residual_2003,evans_monte_2009,evans_monte_2012,slattery_phd_2013}.
@@
 stochastic partial differential equations has indicated that the
-computationaly complexity of the problem can be dramatically reduced
+computationally complexity of the problem can be dramatically reduced
 by incorporating multigrid concepts into the solution scheme
@@
 \cite{briggs_multigrid}. For this analysis, we will use the following
-one-dimensional, homogenous model problem:
+one-dimensional, homogeneous model problem:
 \begin{equation}
@@
 Richardson's iteration and therefore we will first look at the
-performance ofRichardson's iteration as a smoother:
+performance of Richardson's iteration as a smoother:
 \begin{equation}
@@
 used. Figure~\ref{fig:richardson} gives infinity norm of the error in
-the solution vector\footnote{The solution to the homogenous problem is
+the solution vector\footnote{The solution to the homogeneous problem is
   zero and therefore $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$.} as
@@
   \begin{center}
-    \includegraphics[width=5in]{richardson.png}
+    \includegraphics[width=4in]{richardson.png}
   \end{center}
@@
 Richardson's iteration performs. Per the spectral analysis in
-\cite{briggs_multigrid}, this iteration sequence perfoms better for
+\cite{briggs_multigrid}, this iteration sequence performs better for
 higher wave numbers as the eigenvalue spectrum spans a smaller space
@@
 however, we must first modify the linear problem as the Monte Carlo
-solver is a direct method and therefore a homogenous problem with a
+solver is a direct method and therefore a homogeneous problem with a
 right hand side of $\ve{0}$ will yield no samples. Instead, we will
@@
 \end{equation}
-where the residual of the homogenous problem is:
+where the residual of the homogeneous problem is:
 \begin{equation}
@@
 Forming the problem in this way lets us directly apply the adjoint
-Monte Carlo method to the homogenous problem and then apply the
+Monte Carlo method to the homogeneous problem and then apply the
 effective correction, $\ve{d}$, to the initial guess to give the
@@
   \begin{center}
-    \includegraphics[width=5in]{adjoint_mc.png}
+    \includegraphics[width=4in]{adjoint_mc.png}
   \end{center}
@@
 naturally require more Monte Carlo samples to distinguish them from
-one another.
+one another. In other words, statistical noise in the solution must be
+reduced by more samples in order to resolve fine structures in the
+error that would otherwise be buried in the noise.
 
-Based on this, it may not seem that applying Monte Carlo in a
+Based on this, it may not seem likely that applying Monte Carlo in a
 multigrid context will be useful considering that error modes appear
@@
 
+\clearpage
+
 %%---------------------------------------------------------------------------%%
@@
 by Heinrich \cite{heinrich_2001} and Giles \cite{giles_2008}. We start
-first with the standard Monte Carlo estimator for the $i^{th}$
-component of the solution vector:
+first with the standard Monte Carlo estimator for the solution vector:
 \begin{equation}
-  \hat{\ve{x}}_i = \frac{1}{N} \sum_{m=1}^N x^m_i\:,
+  \hat{\ve{x}} = \frac{1}{N} \sum_{m=1}^N x^m\:,
   \label{eq:standard_estimator}
 \end{equation}
-where $i \in [1,N]$, $N$ samples are made and $x^m_i$ is the $m^{th}$
-observation. We next consider representing the problem in multiple
-levels designated by $l$ with each one finer than the next such that
-there are $L$ total levels and $l=L$ the coarsest level. We can
-combine expectation values of the estimates from each of these levels
-using the linearity of expectation such that:
+where $x^m$ is the $m^{th}$ observation. We next consider representing
+the problem in multiple levels designated by $l$ with each one finer
+than the next such that there are $L$ total levels and $l=L$ is the
+coarsest level. We can combine expectation values of the estimates
+from each of these levels using the linearity of expectation such
+that:
 \begin{equation}
@@
   E(\ve{x}_{L-2} - \ve{x}_{L-1}) + \dots + E(\ve{x}_{0} -
-  \ve{x}_{(1)})\:,
+  \ve{x}_{1})\:,
   \label{eq:linearity_of_expectation}
@@
 where $E(\ve{x}_{L})$ is the expectation value of the solution vector
-at the coarsest level, $l = L$.  If we rewrite this summation as:
+at the coarsest level and $E(\ve{x}_{0})$ the expectation at the
+finest level. If we rewrite this summation as:
 \begin{equation}
@@
 Although multigrid methods for linear algebra problems indicate that a
-coarsening parameter of $M = 2$ is optimal such that every level is
-half the size of the finer level. However, Giles reports that values
-of 4 and larger give optimal perfomance for these particular
-estimators. In addition, as mentioned in
+coarsening parameter of $M = 2$ is optimal such that every coarse
+level is half the size of the previous finer level. However, Giles
+reports that values of 4 and larger give optimal performance for these
+particular estimators. In addition, as mentioned in
 \S~\ref{sec:fourier_analysis}, because coarsening the problem makes
-the modes appear more oscillatory, we therefore must compute more
-samples at coarser levels to balance the error amongst the levels. For
-this work we will use the result reported by Hienrich for computing
-the number of samples at each level:
+the modes appear more oscillatory, we must compute more samples at
+coarser levels to balance the error amongst the levels. For this work
+we will use the result reported by Heinrich for computing the number
+of samples at each level:
 \begin{equation}
-  N_l = M^{-3(L-l-1)/2}N\:.
+  N_l = M^{-3(L-l)/2}N\:,
   \label{eq:level_histories}
 \end{equation}
-It should be noted that Giles does not compute this quantity a priori
-but instead first estimates the variance at each level and then
-computes $N_l$ based on that variance.
+which provides a larger number of samples at the coarse levels. It
+should be noted that Giles does not compute this quantity a priori but
+instead first estimates the variance at each level and then computes
+$N_l$ based on that variance.
 
@@
 chain. Computing observations on the $l^{th}$ grid in this case is
-simply the procedure for solving the adjoint Monte Carlo problem. In
-this procedure we can simply tally observations on the grid for the
-$l+1$ level through the use of a linear interpolation operator such
-that a tally in some state of level $l$ is also a tally in some state
-of level $l+1$ with that state determined by the linear operator such
-that the weight of the observation is preserved. If this is the case,
-then we can actually avoid tallying observations of the coarse grid
-altogether and simply apply the linear interpolation operator the fine
-grid results once all samples have been computed.
+simply the procedure for solving the adjoint Monte Carlo problem. We
+could modify this procedure such that we tally observations on the
+grid for the $l+1$ level through the use of a weight preserving linear
+interpolation operator such that a tally in some state of level $l$ is
+also a tally in some state of level $l+1$ with that state determined
+by the operator. If this is the case, then we can actually avoid
+tallying observations of the coarse grid altogether and simply apply
+the linear interpolation operator the fine grid results once all
+samples have been computed instead of modifying the Monte Carlo
+sequence.
 
@@
 vector defined on grid $l$ and a \textit{restriction operator},
-$\ve{R}_l^{l+1}$, which mpas a solution defined on grid $l$ to a
+$\ve{R}_l^{l+1}$, which maps a vector defined on grid $l$ to a
 vector defined on grid $l+1$. We can then define the following
-equivalent expecation value for a given level $l$:
+equivalent expectation value for a given level $l$:
 \begin{equation}
@@
 \end{equation}
-with $\hat{\ve{x}}_{l}$ given Eq~(\ref{eq:standard_estimator}) as the
-solution on grid $l$ and where now the action of the restriction
-operator maps the tally to the coarse level to compute the
+with $\hat{\ve{x}}_{l}$ given by Eq~(\ref{eq:standard_estimator}) as
+the solution on grid $l$. In this estimator, the action of the
+restriction operator maps the tally to the coarse level to compute the
 observations on that grid and the action of the prolongation operator
-on the coarse tally maps the result back to the fine
-level. Eq~(\ref{eq:level_expectation}) then gives the following
-multilevel Monte Carlo algorithm for linear systems now with arbitrary
-right hand side vector, $\ve{b}$:
+on the coarse tally maps the result back to the fine level. Using this
+estimator, Algorithm~\ref{alg:mlamc} gives the multilevel Monte Carlo
+algorithm for linear systems now with an arbitrary right hand side
+vector, $\ve{b}$.
 \begin{algorithm}[h!]
@@
 and then combined appropriately with the prolongation and restriction
-operators. Like other multigrid algorithms for linear system, we are
+operators. Like other multigrid algorithms for linear systems, we are
 required to produce a representation of the linear operator on each
@@
 for $M=2$:
+\begin{subequations}
   \begin{gather}
-  \begin{subequations}
     x^h_{2j} = x_j^{2h} \\
     x^h_{2j+1} = \frac{1}{2} ( x_j^{2h} + x_{j+1}^{2h} )\:,
+  \end{gather}
+  \label{eq:m2_prolongation}
 \end{subequations}
-  \label{eq:m2_prolongation}
-\end{gather}
 for $0 \leq j \leq \frac{J}{2} - 1$ and the restriction operator:
@@
 
-We again perform calculations with initial guesses of wave numbers 1,
-5 and 10 with Tables~\ref{tab:k1_results}, \ref{tab:k5_results}, and
+We again perform calculations with $N=10,000$ and initial guesses of
+wave numbers 1, 5 and 10 and vary the number of levels used in the
+calculation. Tables~\ref{tab:k1_results}, \ref{tab:k5_results}, and
 \ref{tab:k10_results} giving the results of these calculations. In
 these tables, the RFOM column indicates the relative figure of merit
-which has been normalized to the single level calculation.
+which has been normalized to the single level calculation. The results
+indicate the performance of the multilevel method compared to the
+original adjoint Monte Carlo method is excellent, offering a dramatic
+improvement in time to solution and figure of merit values 10-100
+times that of the single level solver.
 \begin{table}[h!]
@@
 
+Next we perform the calculations with $M=4$ using the following
+prolongation operator:
+\begin{subequations}
+  \begin{gather}
+    x^h_{4j} = x_j^{4h} \\
+    x^h_{4j+1} = \frac{3}{4} x_j^{4h} + \frac{1}{4} x_{j+1}^{4h}\:\\
+    x^h_{4j+2} = \frac{1}{2} x_j^{4h} + \frac{1}{2} x_{j+1}^{4h}\:\\
+    x^h_{4j+3} = \frac{1}{4} x_j^{4h} + \frac{3}{4} x_{j+1}^{4h}\:,
+  \end{gather}
+  \label{eq:m4_prolongation}
+\end{subequations}
+for $0 \leq j \leq \frac{J}{4} - 1$ and the restriction operator:
+\begin{equation}
+  x^{4h}_j = \frac{1}{10} ( x_{4j-2}^{h} + 2 x_{4j-1}^{h} + 4
+  x_{4j}^{4h} + 2 x_{4j+1}^{4h} + x_{4j+2}^{4h} )
+  \label{eq:m4_restriction}
+\end{equation}
+for $1 \leq j \leq \frac{J}{4} - 1$. For these calculations we fix
+$k=10$ and vary $N$ to study performance for the more difficult larger
+wave numbers. Tables~\ref{tab:n4_results}, \ref{tab:n5_results}, and
+\ref{tab:n6_results} give the results for $N=10,000$, $N=100,000$, and
+$N=1,000,000$ respectively. Because of the time required to compute
+problems with a small number of levels and large numbers of samples,
+single level calculations were not performed for the last two
+cases. For these cases, relative figure of merit values were computed
+relative to the single level case with $N=10,000$.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      1 & 10,000 & 1.81 & 98.6 & 1 \\
+      2 & 11,250 & 2.64 & 18.2 & 2.53 \\
+      3 & 11,406 & 2.61 & 3.1 & 15.15 \\
+      4 & 11,425 & 4.56 & 0.47 & 33.04 \\
+      5 & 11,427 & 16.5 & 0.07 & 16.95 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
+      = 4$, $N = 10,000$ and run times reported in seconds.}}
+  \label{tab:n4_results}
+\end{table}
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      3 & 114,062 & 0.76 & 28.3 & 19.63 \\
+      4 & 114,275 & 1.53 & 4.4 & 31.43 \\
+      5 & 114,281 & 4.60 & 0.67 & 22.78 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
+      = 4$, $N = 100,000$ and run times reported in seconds.}}
+  \label{tab:n5_results}
+\end{table}
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      4 & 1,142,758 & 0.56 & 41.8 & 25.00 \\
+      5 & 1,142,822 & 1.42 & 6.49 & 24.68 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
+      = 4$, $N = 1,000,000$ and run times reported in seconds.}}
+  \label{tab:n6_results}
+\end{table}
+Although Giles reports that $M \geq 4$ provides good performance in
+the context of his calculations, the improvement is not as dramatic
+for these cases. A coarser discretization of $M=4$ does provide a
+small improvement in the time required to compute a single sample and
+therefore enables slightly higher figures of merit. However, these
+results do not rule out using $M=2$. This is an important result in
+that both geometric and algebraic multigrid methods utilize $M=2$ and
+therefore we may leverage the interpolation operators and other
+infrastructure developed for those methods within the multilevel Monte
+Carlo scheme presented here without a dramatic increase in
+computational complexity.
+
 %%---------------------------------------------------------------------------%%
+\clearpage
+
 \bibliographystyle{ieeetr}

### 2014-04-06 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding copper mountain multilevel slides

@@
 demonstrate the superiority of the multilevel method in terms of time
-to solution.
+to solution using both geometric and algebraic multigrid
+representations of the linear problem.
 
@@
 Using this idea we then define a \textit{prolongation operator},
-$\ve{P}_{l+1}^l$, which maps a vector defined on grid $l+1$ to a
+$\ve{P}_l$, which maps a vector defined on grid $l+1$ to a
 vector defined on grid $l$ and a \textit{restriction operator},
-$\ve{R}_l^{l+1}$, which maps a vector defined on grid $l$ to a
+$\ve{R}_l$, which maps a vector defined on grid $l$ to a
 vector defined on grid $l+1$. We can then define the following
@@
 \begin{equation}
-  E(\ve{x}_{l} - \ve{x}_{(l+1)}) = \Big(\ve{I} - \ve{P}_{l+1}^l
-  \ve{R}_l^{l+1}\Big) \hat{\ve{x}}_{l}\:,
+  E(\ve{x}_{l} - \ve{x}_{l+1}) = \Big(\ve{I} - \ve{P}_l
+  \ve{R}_l\Big) \hat{\ve{x}}_{l}\:,
   \label{eq:level_expectation}
@@
 algorithm for linear systems now with an arbitrary right hand side
-vector, $\ve{b}$.
+vector, $\ve{b}$, and an initial solution guess, $\ve{x}^0$.
 \begin{algorithm}[h!]
-  \caption{Multilevel Adjoint Monte Carlo}
+  \caption{Multilevel Monte Carlo Method}
   \label{alg:mlamc}
   \begin{algorithmic}[1]
-
     \For{ l = 0...L }
-    \State $\ve{A}_l \ve{x}_l = \ve{b}_l$
+    \State $\ve{P}_l = P(\ve{A}_l)$
+    \Comment{Build the prolongation and restriction operators for
+      the $l^{th}$ level.}
+    \State $\ve{R}_l = c \ve{P}_l^T$
+    \State $\ve{r}_l = \ve{b}_l - \ve{A}_l \ve{x}_l^0$
+    \Comment{Build the $l^{th}$ level residual.}
+    \State $\ve{d}_l = \hat{\ve{A}}_l^{-1} \ve{r}_l$
     \Comment{Solve the $l^{th}$ level problem with adjoint Monte
@@
     \If{ l != L }
-    \State $\ve{x}_l = (\ve{I} - \ve{P}_{l+1}^l\ve{R}_l^{l+1}) \ve{x}_{l}$
+    \State $\ve{d}_l = (\ve{I} - \ve{P}_l\ve{R}_l) \ve{d}_{l}$
     \Comment{Apply the multilevel tally}
+    \State $\ve{A}_{l+1} = \ve{R}_l \ve{A}_l \ve{P}_l$
+    \Comment{Construct the next level.}
+    \State $\ve{x}_{l+1}^0 = \ve{R}_l \ve{x}_l^0$
+    \State $\ve{b}_{l+1} = \ve{R}_l \ve{b}_l$
     \EndIf
@@
     \For{ l = L...1 }
-    \State $\ve{x}_{l-1} = ( \ve{I} + \ve{P}_{l}^{l-1} ) \ve{x}_{l}$
+    \State $\ve{d}_{l-1} = ( \ve{I} + \ve{P}_{l} ) \ve{d}_{l}$
     \Comment{Collapse the tallies to the finest grid}
     \EndFor
+
+    \State $\ve{x} = \ve{x}^0 + \ve{d}_0$
 
@@
 required to produce a representation of the linear operator on each
-level, $\ve{A}^l$, as well as the right hand side, $\ve{b}^l$. The
+level, $\ve{A}_l$, as well as the right hand side, $\ve{b}_l$. The
 formulation of Algorithm~\ref{alg:mlamc} places no restrictions on the
-form of $\ve{A}^l$, $\ve{b}^l$, $\ve{P}_{l+1}^l$, or $\ve{R}_l^{l+1}$
+form of $\ve{A}_l$, $\ve{b}_l$, $\ve{P}_l$, or $\ve{R}_l$
 and therefore could be used with any multigrid scheme including those
@@
 %%---------------------------------------------------------------------------%%
-\section{Results}
-\label{sec:results}
+\section{Geometric Multigrid Results}
+\label{sec:geometric_results}
 To demonstrate the multilevel Monte Carlo algorithm we solve the
@@
 %%---------------------------------------------------------------------------%%
+\section{Algebraic Multigrid Results}
+\label{sec:algebraic_results}
+
+%%---------------------------------------------------------------------------%%
 \clearpage

### 2014-04-07 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding ML results to multilevel tech note

@@
   dramatically reducing the computational complexity of the
-  problem. Furthermore, the general formulation presented here
-  indicates that this methodology is not restricted to geometric
-  formulations of the multigrid method and could additionally be
-  formulated using algebraic techniques.
+  problem. Furthermore, the general formulation and results presented
+  here indicate that this methodology may be used with both geometric
+  and algebraic formulations of the Galerkin mutlgrid method.
 \end{abstract}
@@
 and therefore could be used with any multigrid scheme including those
-arising from algebraic multigrid formulations.
+arising from algebraic multigrid formulations. It should be noted,
+however, that work to date has only tested the algorithm for the
+Galerkin formulation presented in Algorithm~\ref{alg:mlamc}.
 
@@
   \end{center}
-  \caption{\textbf{Multilevel Monte Carlo results for $k = 1$ with $M
-      = 2$, $N = 10,000$ and run times reported in seconds.}}
+  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
+      1$ with $M = 2$, $N = 10,000$ and run times reported in
+      seconds.}}
   \label{tab:k1_results}
@@
   \end{center}
-  \caption{\textbf{Multilevel Monte Carlo results for $k = 5$ with $M
-      = 2$, $N = 10,000$ and run times reported in seconds.}}
+  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
+      5$ with $M = 2$, $N = 10,000$ and run times reported in
+      seconds.}}
   \label{tab:k5_results}
@@
   \end{center}
-  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
-      = 2$, $N = 10,000$ and run times reported in seconds.}}
+  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
+      10$ with $M = 2$, $N = 10,000$ and run times reported in
+      seconds.}}
   \label{tab:k10_results}
@@
   \end{center}
-  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
-      = 4$, $N = 10,000$ and run times reported in seconds.}}
+  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
+      10$ with $M = 4$, $N = 10,000$ and run times reported in
+      seconds.}}
   \label{tab:n4_results}
@@
   \end{center}
-  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
-      = 4$, $N = 100,000$ and run times reported in seconds.}}
+  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
+      10$ with $M = 4$, $N = 100,000$ and run times reported in
+      seconds.}}
   \label{tab:n5_results}
@@
   \end{center}
-  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
-      = 4$, $N = 1,000,000$ and run times reported in seconds.}}
+  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
+      10$ with $M = 4$, $N = 1,000,000$ and run times reported in
+      seconds.}}
   \label{tab:n6_results}
@@
 \label{sec:algebraic_results}
+We next solve the same model problem, this time using an algebraic
+multigrid formulation of the Galerkin problem. For these calculations
+we used the ML multigrid library to construct $\ve{A}$, $\ve{P}$, and
+$\ve{R}$ on each level with $\ve{M} \approx 3$
+\cite{ml_manual}. Tables~\ref{tab:ml_k1_results},
+\ref{tab:ml_k5_results}, and \ref{tab:ml_k10_results} give the results
+of these calculations.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      1 & 10,000 & 0.022 & 119.1 & 1 \\
+      2 & 11,924 & 0.037 & 31.9 & 1.28 \\
+      3 & 12,294 & 0.055 & 7.6 & 2.40 \\
+      4 & 12,365 & 0.051 & 1.8 & 11.76 \\
+      5 & 12,378 & 0.094 & 0.5 & 12.01 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Algebraic Multilevel Monte Carlo results for $k =
+      1$ with $M = 3$, $N = 10,000$ and run times reported in
+      seconds.}}
+  \label{tab:ml_k1_results}
+\end{table}
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      1 & 10,000 & 0.668 & 101.6 & 1 \\
+      2 & 11,924 & 0.834 & 30.8 & 2.12 \\
+      3 & 12,294 & 1.140 & 7.5 & 4.63 \\
+      4 & 12,365 & 0.975 & 1.6 & 29.09 \\
+      5 & 12,378 & 2.240 & 0.4 & 25.10 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Algebraic Multilevel Monte Carlo results for $k =
+      5$ with $M = 3$, $N = 10,000$ and run times reported in
+      seconds.}}
+  \label{tab:ml_k5_results}
+\end{table}
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} & 
+      \multicolumn{1}{l}{\textbf{Samples}} & 
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
+      \multicolumn{1}{l}{\textbf{Time (s)}} & 
+      \multicolumn{1}{l}{\textbf{RFOM}} \\
+      \hline
+      1 & 10,000 & 1.81 & 98.6 & 1 \\
+      2 & 11,924 & 2.33 & 30.0 & 1.99 \\
+      3 & 12,294 & 3.19 & 7.6 & 4.20 \\
+      4 & 12,365 & 3.31 & 1.8 & 16.29 \\
+      5 & 12,378 & 9.98 & 0.5 & 7.21 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Algebraic Multilevel Monte Carlo results for $k =
+      10$ with $M = 3$, $N = 10,000$ and run times reported in
+      seconds.}}
+  \label{tab:ml_k10_results}
+\end{table}
+In general, the results show similar performance to the geometric
+multigrid method, however, gains in figure of merit were not as
+significant which may be in part due to the simplified selection of
+sample size at each level. In addition, the Dirichlet conditions for
+the model problem were explicitly included in the geometric
+interpolation operators while the level operators generated by the
+algebraic scheme did not. Most importantly, these results indicate
+that Algorithm~\ref{alg:mlamc} can be used with general Galerkin
+formulations of the multigrid problem.
 

### 2014-04-29 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
finished first draft of copper multilevel slides
adding some heterogenous poisson results to the multilevel mc note

--- 2014-04-06 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
@@
 Monte Carlo method to traditional iterative smoothers that would be
-used with the multigrid method. Next, we present the multilevel Monte
-Carlo method for linear systems using a general algebraic
-formulation. Finally, we present results using the model problem the
-demonstrate the superiority of the multilevel method in terms of time
-to solution using both geometric and algebraic multigrid
+used with the multigrid method. Finally, we present results using the
+model problem the demonstrate the superiority of the multilevel method
+in terms of time to solution using a geometric multigrid
 representations of the linear problem.
@@
     \For{ l = L...1 }
-    \State $\ve{d}_{l-1} = ( \ve{I} + \ve{P}_{l} ) \ve{d}_{l}$
+    \State $\ve{d}_{l-1} = \ve{d}_{l-1} + \ve{P}_{l} \ve{d}_{l}$
     \Comment{Collapse the tallies to the finest grid}
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Algebraic Multigrid Results}
-\label{sec:algebraic_results}
+\subsection{Heterogeneous Results}
+We next solve the same linear problem as before using $M=2$ but this
+time with a constant forcing of $\ve{b}=1.0$ such that:
+\begin{equation}
+  \ve{A}\ve{x} = \ve{b}\:.
+  \label{eq:heterogeneous_problem}
+\end{equation}
+We will choose an initial guess of $\ve{0}$ for these calculations as
+the solution is non-trivial. Studies indicate that residual norms are
+controlled by the convergence of the finest level. However, adding
+additional levels provides significant smoothing to the statistical
+error. For example, given a fixed amount of compute time of
+approximately 7 seconds, Eq~(\ref{eq:heterogeneous_problem}) was
+solved using 1 level and 10,000 samples and again using 9 levels and
+1,546,783 total samples ($N=1,000,000$). The 1 level calculation gave
+$||\ve{r}||_\infty = 0.027$ and $||\ve{r}||_2 = 0.230$ while the 1
+level calculation gave $||\ve{r}||_\infty = 0.129$ and $||\ve{r}||_2 =
+0.812$. Based on the residual norm values, it would seem as though the
+1 level result was superior as better convergence was achieved in the
+same amount of compute time. However, the structure of the actual
+solutions would indicate otherwise. Figure~\ref{fig:7_sec_1_level}
+gives the 1 level result and Figure~\ref{fig:7_sec_9_level} gives the
+9 level result, both plotted against an exact solution computed using
+a minimal residual iteration. Clearly, the estimate produced by the 9
+level computation is more representative of the exact solution than
+the single level scheme due to the well-converged results provided
+coarse level estimates.
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=4in]{7_sec_1_level.pdf}
+  \end{center}
+  \caption{\textbf{Solution of the Poisson problem using the geometric
+      multilevel Monte Carlo method with 1 level and 6.87 seconds of
+      compute time.}}
+  \label{fig:7_sec_1_level}
+\end{figure}
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=4in]{7_sec_9_level.pdf}
+  \end{center}
+  \caption{\textbf{Solution of the Poisson problem using the geometric
+      multilevel Monte Carlo method with 9 levels and 6.78 seconds of
+      compute time.}}
+  \label{fig:7_sec_9_level}
+\end{figure}
+To show that the 1 level calculation will eventually converge to the
+correct result, Figure~\ref{fig:converged_1_level} shows the solution
+with 1,000,000 samples. This calculation required 665.7 seconds to run
+and gave $||\ve{r}||_\infty = 0.0025$ and $||\ve{r}||_2 = 0.022$.
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=4in]{converged_1_level.pdf}
+  \end{center}
+  \caption{\textbf{Solution of the Poisson problem using the geometric
+      multilevel Monte Carlo method with 1 level and 665.7 seconds of
+      compute time.}}
+  \label{fig:converged_1_level}
+\end{figure}
+
+Clearly, these results indicate that quality of solution from the
+multilevel method may not lie entirely in a residual norm metric as
+they are global in nature. For the cases above we see that the
+multilevel estimator has signficantly better local convergence
+properties for a fixed amount of work when compared to single level
+methods. Therefore, we must determine if this improved local
+convergence is satisfactory to improve the convergence of MCSA or if
+convergence of global metrics is also required of the Monte Carlo method.
 

### 2014-04-29 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding heterogenous multilevel figures to mlmc tech note

@@
 We will choose an initial guess of $\ve{0}$ for these calculations as
-the solution is non-trivial. Studies indicate that residual norms are
-controlled by the convergence of the finest level. However, adding
-additional levels provides significant smoothing to the statistical
-error. For example, given a fixed amount of compute time of
-approximately 7 seconds, Eq~(\ref{eq:heterogeneous_problem}) was
-solved using 1 level and 10,000 samples and again using 9 levels and
-1,546,783 total samples ($N=1,000,000$). The 1 level calculation gave
-$||\ve{r}||_\infty = 0.027$ and $||\ve{r}||_2 = 0.230$ while the 1
-level calculation gave $||\ve{r}||_\infty = 0.129$ and $||\ve{r}||_2 =
-0.812$. Based on the residual norm values, it would seem as though the
-1 level result was superior as better convergence was achieved in the
-same amount of compute time. However, the structure of the actual
-solutions would indicate otherwise. Figure~\ref{fig:7_sec_1_level}
-gives the 1 level result and Figure~\ref{fig:7_sec_9_level} gives the
-9 level result, both plotted against an exact solution computed using
-a minimal residual iteration. Clearly, the estimate produced by the 9
-level computation is more representative of the exact solution than
-the single level scheme due to the well-converged results provided
-coarse level estimates.
+the solution is non-trivial and fix $N$ at 10,000 while increasing the
+number of levels. Table~\ref{tab:heterogeneous_fixed_N} gives the
+results of these calculations. Here the relative error error
+calculations were computed using an exact solution computed using a
+minimal residual iteration converged to a residual norm tolerance of
+$\sn{1}{-12}$.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllllllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} &
+      \multicolumn{1}{l}{\textbf{Samples}} &
+      \multicolumn{1}{l}{\textbf{$||\ve{r}||_{\infty}$}} &
+      \multicolumn{1}{l}{\textbf{$||\ve{r}||_{2}$}} &
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} &
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{2}$}} &
+      \multicolumn{1}{l}{\textbf{Time (s)}} &
+      \multicolumn{1}{l}{\textbf{RRESFOM}} &
+      \multicolumn{1}{l}{\textbf{RERRFOM}} \\
+      \hline
+      1 & 10,000 & 0.025 & 0.223 & 0.241 & 2.97 & 6.77 & 1.0 & 1.0 \\
+      2 & 17,596 & 0.044 & 0.339 & 0.277 & 5.49 & 7.19 & 0.31 & 0.71 \\
+      3 & 29,571 & 0.050 & 0.365 & 0.131 & 1.66 & 7.21 & 0.23 & 3.18 \\
+      4 & 53,295 & 0.062 & 0.483 & 0.081 & 0.96 & 7.29 & 0.16 & 8.28 \\
+      5 & 92,301 & 0.086 & 0.662 & 0.105 & 0.88 & 7.13 & 0.08 & 5.00 \\
+      6 & 185,266 & 0.120 & 0.891 & 0.102 & 0.72 & 7.10 & 0.04 & 5.32 \\
+      7 & 355,543 & 0.069 & 0.513 & 0.056 & 0.51 & 6.92 & 0.13 & 18.12 \\
+      8 & 742,337 & 0.093 & 0.654 & 0.081 & 0.67 & 7.07 & 0.07 & 8.47 \\
+      9 & 1,624,121 & 0.122 & 0.758 & 0.107 & 0.83 & 7.08 & 0.04 &
+      4.85 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Geometric Multilevel Monte Carlo results for
+      $\ve{b} = 1$ with $M = 2$, $t \approx 7s$ and run times reported
+      in seconds.} \textit{RRESSFOM values are the relative figure of
+      merit computed using the infinity norm of the residual. RERRFORM
+    values are the relative figure of merit computed using the
+    infinity norm of the absolute error vector.}}
+  \label{tab:heterogeneous_fixed_N}
+\end{table}
+Based on the residual norm values, it would seem as though the 1 level
+result was superior as better convergence was achieved in the same
+amount of compute time using residual norm metrics. However, the
+structure of the actual solutions would indicate
+otherwise. Figure~\ref{fig:heterogeneous_ml_solutions} gives the
+results for with 1-9 levels all plotted against the exact
+solution. Clearly, the estimate produced by the 9 level computation is
+more representative of the exact solution than the single level scheme
+as indicated by the error norms due to the well-converged results
+provided coarse level estimates. Furthermore, the error norms also
+indicate a more efficient solution scheme when multiple levels are
+used showing similar relative figure of merit metrics for the error
+norm computations.
+
 \begin{figure}[h!]
-  \begin{center}
-    \includegraphics[width=4in]{7_sec_1_level.pdf}
-  \end{center}
+  \centering
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_1_level.pdf}
+    \caption{\textbf{1 level.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_2_level.pdf}
+    \caption{\textbf{2 levels.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_3_level.pdf}
+    \caption{\textbf{3 levels.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_4_level.pdf}
+    \caption{\textbf{4 levels.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_5_level.pdf}
+    \caption{\textbf{5 levels.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_6_level.pdf}
+    \caption{\textbf{6 levels.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_7_level.pdf}
+    \caption{\textbf{7 levels.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_8_level.pdf}
+    \caption{\textbf{8 levels.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_9_level.pdf}
+    \caption{\textbf{9 levels.}}
+  \end{subfigure}
   \caption{\textbf{Solution of the Poisson problem using the geometric
-      multilevel Monte Carlo method with 1 level and 6.87 seconds of
-      compute time.}}
-  \label{fig:7_sec_1_level}
+      multilevel Monte Carlo method with 1-9 levels and approximately
+      7 seconds of compute time.}}
+  \label{fig:heterogenous_ml_solutions}
 \end{figure}
-\begin{figure}[h!]
-  \begin{center}
-    \includegraphics[width=4in]{7_sec_9_level.pdf}
-  \end{center}
-  \caption{\textbf{Solution of the Poisson problem using the geometric
-      multilevel Monte Carlo method with 9 levels and 6.78 seconds of
-      compute time.}}
-  \label{fig:7_sec_9_level}
-\end{figure}
+
 To show that the 1 level calculation will eventually converge to the
@@
 convergence is satisfactory to improve the convergence of MCSA or if
-convergence of global metrics is also required of the Monte Carlo method.
+convergence of global metrics is also required of the Monte Carlo
+method.
+
+%%---------------------------------------------------------------------------%%
+\section{Coarse Grid Monte Carlo Acceleration}
+Based on the above results we then consider using only the coarse grid
+calculations as the acceleration. Consider the following results where
+a coarse grid solution has been produced and then simply prolongated
+back to the fine grid rather than solving all levels as a simple
+coarse approximation of the solution.
 

### 2014-04-29 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding coarse grid solution figures to mlmc tech note

@@
 the solution is non-trivial and fix $N$ at 10,000 while increasing the
-number of levels. Table~\ref{tab:heterogeneous_fixed_N} gives the
+number of levels. Table~\ref{tab:heterogeneous_fixed_t} gives the
 results of these calculations. Here the relative error error
@@
     infinity norm of the absolute error vector.}}
-  \label{tab:heterogeneous_fixed_N}
+  \label{tab:heterogeneous_fixed_t}
 \end{table}
@@
       multilevel Monte Carlo method with 1-9 levels and approximately
-      7 seconds of compute time.}}
+      7 seconds of compute time.} \textit{Increasing the number of
+      levels in the computation improves convergence with respect to
+      the relative solution error but convergence with respect to the
+      linear problem residual is worse.}}
   \label{fig:heterogenous_ml_solutions}
@@
 
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cllllllll}\hline\hline
+      \multicolumn{1}{c}{\textbf{Levels}} &
+      \multicolumn{1}{l}{\textbf{Samples}} &
+      \multicolumn{1}{l}{\textbf{$||\ve{r}||_{\infty}$}} &
+      \multicolumn{1}{l}{\textbf{$||\ve{r}||_{2}$}} &
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} &
+      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{2}$}} &
+      \multicolumn{1}{l}{\textbf{Time (s)}} &
+      \multicolumn{1}{l}{\textbf{RRESFOM}} &
+      \multicolumn{1}{l}{\textbf{RERRFOM}} \\
+      \hline
+      1 & 10,000 & 0.025 & 0.223 & 0.241 & 2.97 & 6.77 & 1.0 & 1.0 \\
+      2 & 17,596 & 0.044 & 0.339 & 0.277 & 5.49 & 7.19 & 0.31 & 0.71 \\
+      3 & 29,571 & 0.050 & 0.365 & 0.131 & 1.66 & 7.21 & 0.23 & 3.18 \\
+      4 & 53,295 & 0.062 & 0.483 & 0.081 & 0.96 & 7.29 & 0.16 & 8.28 \\
+      5 & 92,301 & 0.086 & 0.662 & 0.105 & 0.88 & 7.13 & 0.08 & 5.00 \\
+      6 & 185,266 & 0.120 & 0.891 & 0.102 & 0.72 & 7.10 & 0.04 & 5.32 \\
+      7 & 355,543 & 0.069 & 0.513 & 0.056 & 0.51 & 6.92 & 0.13 & 18.12 \\
+      8 & 742,337 & 0.093 & 0.654 & 0.081 & 0.67 & 7.07 & 0.07 & 8.47 \\
+      9 & 1,624,121 & 0.122 & 0.758 & 0.107 & 0.83 & 7.08 & 0.04 &
+      4.85 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Coarse grid results for $\ve{b} = 1$ with $M = 2$,
+      $N = 10,000$ and run times reported in seconds.}
+    \textit{RRESSFOM values are the relative figure of merit computed
+      using the infinity norm of the residual. RERRFORM values are the
+      relative figure of merit computed using the infinity norm of the
+      absolute error vector.}}
+  \label{tab:heterogeneous_coarse_grids}
+\end{table}
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{7_sec_1_level.pdf}
+    \caption{\textbf{Level 1 solution.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{cg_2_level.pdf}
+    \caption{\textbf{Level 2 solution.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{cg_3_level.pdf}
+    \caption{\textbf{Level 3 solution.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{cg_4_level.pdf}
+    \caption{\textbf{Level 4 solution.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{cg_5_level.pdf}
+    \caption{\textbf{Level 5 solution.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{cg_6_level.pdf}
+    \caption{\textbf{Level 6 solution.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{cg_7_level.pdf}
+    \caption{\textbf{Level 7 solution.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{cg_8_level.pdf}
+    \caption{\textbf{Level 8 solution.}}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{cg_9_level.pdf}
+    \caption{\textbf{Level 9 solution.}}
+  \end{subfigure}
+  \caption{\textbf{Solution of the Poisson problem using the geometric
+      multilevel Monte Carlo method with 1-9 levels and approximately
+      7 seconds of compute time.} \textit{Increasing the number of
+      levels in the computation improves convergence with respect to
+      the relative solution error but convergence with respect to the
+      linear problem residual is worse.}}
+  \label{fig:heterogenous_ml_solutions}
+\end{figure}
+
 %%---------------------------------------------------------------------------%%

### 2014-04-29 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding initial coarse grid results to the mlmc tech report

@@
   \end{center}
-  \caption{\textbf{Geometric Multilevel Monte Carlo results for
-      $\ve{b} = 1$ with $M = 2$, $t \approx 7s$ and run times reported
-      in seconds.} \textit{RRESSFOM values are the relative figure of
-      merit computed using the infinity norm of the residual. RERRFORM
-    values are the relative figure of merit computed using the
-    infinity norm of the absolute error vector.}}
-  \label{tab:heterogeneous_fixed_t}
+  \caption{\textbf{Coarse grid results for $\ve{b} = 1$ with $M = 2$,
+      $t \approx 7s$ and run times reported in seconds.}
+    \textit{RRESSFOM values are the relative figure of merit computed
+      using the infinity norm of the residual. RERRFORM values are the
+      relative figure of merit computed using the infinity norm of the
+      absolute error vector.}}
+  \label{tab:heterogeneous_coarse_grids}
 \end{table}
+
 Based on the residual norm values, it would seem as though the 1 level
@@
   \centering
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_1_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_2_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_3_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_4_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_5_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_6_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_7_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_8_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{7_sec_9_level.pdf}
@@
       linear problem residual is worse.}}
-  \label{fig:heterogenous_ml_solutions}
+  \label{fig:heterogenous_fixed_t}
 \end{figure}
@@
 Based on the above results we then consider using only the coarse grid
-calculations as the acceleration. Consider the following results where
-a coarse grid solution has been produced and then simply prolongated
-back to the fine grid rather than solving all levels as a simple
-coarse approximation of the solution.
+calculations as the acceleration. Consider the following results given
+in Table~\ref{tab:heterogeneous_cg_solutions} where a coarse grid
+solution has been produced at a fixed $N=10,000$ and then simply
+prolongated back to the fine grid rather than solving all levels using
+Algorithm~\ref{alg:mlamc}. Figure~\ref{fig:hetergeneous_cg_solutions}
+plots the coarse grid solutions prolongated back to the fine grid
+along with the exact solution. We see that as we coarsen the grid
+while fixing $N$, the coarse grid statistics are better than the fine
+grid as there is a larger ratio of samples to grid points. The result
+is a very precise coarse grid result that when prolongated to the fine
+grid simply incurs the grid error associated with the linear
+interpolation.
 
@@
       1 & 10,000 & 0.025 & 0.223 & 0.241 & 2.97 & 6.77 & 1.0 & 1.0 \\
-      2 & 17,596 & 0.044 & 0.339 & 0.277 & 5.49 & 7.19 & 0.31 & 0.71 \\
-      3 & 29,571 & 0.050 & 0.365 & 0.131 & 1.66 & 7.21 & 0.23 & 3.18 \\
-      4 & 53,295 & 0.062 & 0.483 & 0.081 & 0.96 & 7.29 & 0.16 & 8.28 \\
-      5 & 92,301 & 0.086 & 0.662 & 0.105 & 0.88 & 7.13 & 0.08 & 5.00 \\
-      6 & 185,266 & 0.120 & 0.891 & 0.102 & 0.72 & 7.10 & 0.04 & 5.32 \\
-      7 & 355,543 & 0.069 & 0.513 & 0.056 & 0.51 & 6.92 & 0.13 & 18.12 \\
-      8 & 742,337 & 0.093 & 0.654 & 0.081 & 0.67 & 7.07 & 0.07 & 8.47 \\
-      9 & 1,624,121 & 0.122 & 0.758 & 0.107 & 0.83 & 7.08 & 0.04 &
-      4.85 \\
+      2 & 10,000 & 0.012 & 0.067 & 0.303 & 6.05 & 3.29 & 9.70 & 1.3 \\
+      3 & 10,000 & 0.005 & 0.028 & 0.133 & 1.70 & 1.62 & 103.6 & 13.7 \\
+      4 & 10,000 & 0.004 & 0.013 & 0.058 & 0.61 & 0.81 & 292.7 & 146.3 \\
+      5 & 10,000 & 0.023 & 0.031 & 0.141 & 3.06 & 0.40 & 20.8 & 49.4 \\
+      6 & 10,000 & 0.006 & 0.008 & 0.051 & 0.77 & 0.20 & 557.0 & 741.1 \\
+      7 & 10,000 & 0.006 & 0.006 & 0.049 & 0.77 & 0.10 & 1,103 & 1,589 \\
+      8 & 10,000 & 0.009 & 0.009 & 0.022 & 0.29 & 0.05 & 952.8 & 15,329 \\
+      9 & 10,000 & 0.013 & 0.014 & 0.032 & 0.51 & 0.03 & 997.9 &
+      14,826 \\
       \hline\hline
@@
   \end{center}
-  \caption{\textbf{Coarse grid results for $\ve{b} = 1$ with $M = 2$,
-      $N = 10,000$ and run times reported in seconds.}
-    \textit{RRESSFOM values are the relative figure of merit computed
-      using the infinity norm of the residual. RERRFORM values are the
-      relative figure of merit computed using the infinity norm of the
-      absolute error vector.}}
-  \label{tab:heterogeneous_coarse_grids}
+  \caption{\textbf{Geometric Multilevel Monte Carlo results for
+      $\ve{b} = 1$ with $M = 2$, $N = 10,000$ and run times reported
+      in seconds.} \textit{RRESSFOM values are the relative figure of
+      merit computed using the infinity norm of the residual. RERRFORM
+    values are the relative figure of merit computed using the
+    infinity norm of the absolute error vector.}}
+  \label{tab:heterogeneous_cg_solutions}
 \end{table}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{cg_2_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{cg_3_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{cg_4_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{cg_5_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{cg_6_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{cg_7_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{cg_8_level.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \includegraphics[width=\textwidth]{cg_9_level.pdf}
@@
   \caption{\textbf{Solution of the Poisson problem using the geometric
-      multilevel Monte Carlo method with 1-9 levels and approximately
-      7 seconds of compute time.} \textit{Increasing the number of
-      levels in the computation improves convergence with respect to
-      the relative solution error but convergence with respect to the
-      linear problem residual is worse.}}
-  \label{fig:heterogenous_ml_solutions}
+      multilevel Monte Carlo method with 1-9 levels and $N = 10,000$.}
+    \textit{Coarse solutions prolongated to the fine grid give better
+      results in less time than fine grid solutions.}}
+  \label{fig:heterogenous_cg_solutions}
 \end{figure}

### 2014-04-29 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
fixing broken references in multilevel mc note

@@
   \ve{r} = -\ve{A}\ve{x}^0\:,
-  \label{eq:homogenous_residual}
+  \label{eq:homogeneous_residual}
 \end{equation}
@@
 10$ given in Figure~\ref{fig:fourier_modes}.
-\begin{figure}[h!]
+\begin{figure}[t!]
   \centering
-  \begin{subfigure}[b]{0.5\textwidth}
+  \begin{subfigure}[b]{0.49\textwidth}
     \includegraphics[width=\textwidth]{mode_1.pdf}
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.5\textwidth}
+  \begin{subfigure}[b]{0.49\textwidth}
     \includegraphics[width=\textwidth]{mode_10.pdf}
@@
 approach.
-
-\clearpage
 
@@
 We next solve the same linear problem as before using $M=2$ but this
-time with a constant forcing of $\ve{b}=1.0$ such that:
+time with a constant forcing of $\ve{b}=\ve{1}$ such that:
 \begin{equation}
@@
 \end{equation}
-We will choose an initial guess of $\ve{0}$ for these calculations as
-the solution is non-trivial and fix $N$ at 10,000 while increasing the
-number of levels. Table~\ref{tab:heterogeneous_fixed_t} gives the
-results of these calculations. Here the relative error error
-calculations were computed using an exact solution computed using a
-minimal residual iteration converged to a residual norm tolerance of
-$\sn{1}{-12}$.
+We will choose an initial guess of $\ve{x}^0 = \ve{0}$ for these
+calculations as the solution is non-trivial and fix $N$ at 10,000
+while increasing the number of
+levels. Table~\ref{tab:heterogeneous_fixed_t} gives the results of
+these calculations. Here the relative error error calculations were
+computed using an exact solution computed using a minimal residual
+iteration converged to a residual norm tolerance of $\sn{1}{-12}$.
 \begin{table}[h!]
@@
       absolute error vector.}}
-  \label{tab:heterogeneous_coarse_grids}
+  \label{tab:heterogeneous_fixed_t}
 \end{table}
@@
 structure of the actual solutions would indicate
-otherwise. Figure~\ref{fig:heterogeneous_ml_solutions} gives the
+otherwise. Figure~\ref{fig:heterogeneous_fixed_t} gives the
 results for with 1-9 levels all plotted against the exact
@@
 used showing similar relative figure of merit metrics for the error
-norm computations.
+norm computations as those computed in the heterogeneous cases.
 
@@
       the relative solution error but convergence with respect to the
-      linear problem residual is worse.}}
-  \label{fig:heterogenous_fixed_t}
+      linear problem residual is worse. Numeric results from these
+      calculations are given in
+      Table~\ref{tab:heterogeneous_fixed_t}.}}
+  \label{fig:heterogeneous_fixed_t}
 \end{figure}
@@
 prolongated back to the fine grid rather than solving all levels using
-Algorithm~\ref{alg:mlamc}. Figure~\ref{fig:hetergeneous_cg_solutions}
+Algorithm~\ref{alg:mlamc}. Figure~\ref{fig:heterogeneous_cg_solutions}
 plots the coarse grid solutions prolongated back to the fine grid
@@
     \begin{tabular}{cllllllll}\hline\hline
-      \multicolumn{1}{c}{\textbf{Levels}} &
+      \multicolumn{1}{c}{\textbf{Grid}} &
       \multicolumn{1}{l}{\textbf{Samples}} &
@@
   \end{center}
-  \caption{\textbf{Geometric Multilevel Monte Carlo results for
+  \caption{\textbf{Coarse grid Monte Carlo results for
       $\ve{b} = 1$ with $M = 2$, $N = 10,000$ and run times reported
@@
     \includegraphics[width=\textwidth]{7_sec_1_level.pdf}
-    \caption{\textbf{Level 1 solution.}}
+    \caption{\textbf{Grid 1 solution.}}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{cg_2_level.pdf}
-    \caption{\textbf{Level 2 solution.}}
+    \caption{\textbf{Grid 2 solution.}}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{cg_3_level.pdf}
-    \caption{\textbf{Level 3 solution.}}
+    \caption{\textbf{Grid 3 solution.}}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{cg_4_level.pdf}
-    \caption{\textbf{Level 4 solution.}}
+    \caption{\textbf{Grid 4 solution.}}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{cg_5_level.pdf}
-    \caption{\textbf{Level 5 solution.}}
+    \caption{\textbf{Grid 5 solution.}}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{cg_6_level.pdf}
-    \caption{\textbf{Level 6 solution.}}
+    \caption{\textbf{Grid 6 solution.}}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{cg_7_level.pdf}
-    \caption{\textbf{Level 7 solution.}}
+    \caption{\textbf{Grid 7 solution.}}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{cg_8_level.pdf}
-    \caption{\textbf{Level 8 solution.}}
+    \caption{\textbf{Grid 8 solution.}}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{cg_9_level.pdf}
-    \caption{\textbf{Level 9 solution.}}
+    \caption{\textbf{Grid 9 solution.}}
   \end{subfigure}
-  \caption{\textbf{Solution of the Poisson problem using the geometric
-      multilevel Monte Carlo method with 1-9 levels and $N = 10,000$.}
-    \textit{Coarse solutions prolongated to the fine grid give better
-      results in less time than fine grid solutions.}}
-  \label{fig:heterogenous_cg_solutions}
+  \caption{\textbf{Solution of the Poisson problem using geometric
+      grids 1-9 and Monte Carlo with $N = 10,000$ for each
+      calculation.} \textit{Coarse solutions prolongated to the fine
+      grid give better results in less time than fine grid
+      solutions. Numeric results for these calculations are given in
+      Table~\ref{tab:heterogeneous_cg_solutions}.}}
+  \label{fig:heterogeneous_cg_solutions}
 \end{figure}

### 2014-05-08 doc/ttsp2013/ttsp_2013.tex
rearranging the mcsa and spn sections in the ttsp paper

@@
 %%---------------------------------------------------------------------------%%
-\author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton
-\\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}}
-}
+\author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton, Paul
+  P.H. Wilson
+  \\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}} }
 
@@
 radiaton transport systems. Of most significance, it was discovered
-that simply reducing the spectral radius of the Monte Carlo problem
-below unity is not enough to achieve convergence as outlined in
-previous work. In order to effectively solve the $SP_N$ equations for
-the model problem, an algebraic preconditioning strategy is developed
-for MCSA where we utilize incomplete factorizations as preconditioners
-to achieve convergence. Using the preconditioners, MCSA performance is
-compared with production implementations of the preconditioned GMRES
-and BiCGSTAB methods. Studies indicate that the iterative performance
-of the method is competitive with the preconditioned Krylov methods
+for these problems that simply reducing the spectral radius of the
+Monte Carlo problem below unity is not enough to efficiently achieve
+convergence as reported in previous work. In order to effectively
+solve the $SP_N$ equations for the model problem, an algebraic
+preconditioning strategy is developed for MCSA where we utilize
+incomplete factorizations as preconditioners to achieve
+convergence. Using the preconditioners, MCSA performance is compared
+with production implementations of the preconditioned GMRES and
+BiCGSTAB methods. Our studies indicate that the iterative performance
+of MCSA when used to solve the $SP_N$ equations for light water
+reactor problems is competitive with the preconditioned Krylov methods
 when operator spectrum is properly conditioned.
@@
 capability can enable tighter design tolerances for improved thermal
-performance and efficiency, higher fuel utilization, and high
-confidence in accident scenario models. Modern high fidelity
-deterministic methods for large scale problems are often variants on
-the discrete ordinates ($S_N$) method \cite{evans_denovo:_2010}. For
-fission reactor neutronics simulations, the $S_N$ method requires
+performance and efficiency, higher fuel utilization and burnup, and
+high confidence in accident scenario models. Modern high fidelity
+deterministic methods for large scale problems where the solution over
+the entire core of a reactor is desired are often variants on the
+discrete ordinates ($S_N$) method \cite{evans_denovo:_2010} or the
+method of characteristics \cite{askew_1972}. For fission reactor
+neutronics simulations, the discrete ordinates method requires
 potentially trillions of unknown angular flux moments to be computed
 to achieve good accuracy for the responses of interest
-\cite{slaybaugh_acceleration_2011}. In the 1960's, Gelbard developed
-an ad-hoc, multidimensional extension of the simple single dimension
-planar $P_N$ equations that created a system of coupled,
-diffusion-like equations known as the simplified $P_N$ ($SP_N$)
-equations (cite gelbard). The reduction in numerical complexity
-compared to current deterministic solution methods using the $S_N$
-approximation could mean significant savings in both compute time and
-memory required and permit larger simulations to be performed on the
-available computing platforms.
+\cite{slaybaugh_acceleration_2011} which in many cases can be
+intractable without sufficient computational hardware. On the other
+side of the spectrum, low-fidelity diffusion solutions are often
+leveraged as an alternative to compute a solution over the entire
+domain a reactor core. In the 1960's, Gelbard developed an ad-hoc,
+multidimensional extension of the simple single dimension planar $P_N$
+equations that created a system of coupled, diffusion-like equations
+known as the simplified $P_N$ ($SP_N$) equations \cite{gelbard_1960}
+which incorporates additional angular information into a set of
+diffusion-like equations. Further work in both aysmptotic analysis and
+algorithmic development proceeded in the following decades to serve as
+a theoretical basis for the method
+\cite{morel_1996,brantley_simplified_2000}. Recent work has expanded
+the discretization to a fully algebraic form ammenable to solution
+with modern linear algebra techniques including Krylov solvers and
+advanced precondtioners including multigrid methods and incomplete
+factorizations \cite{hamilton_2014}. Using these new techniques, the
+reduction in numerical complexity compared to current deterministic
+solution methods using the $SP_N$ approximation could mean significant
+savings in both compute time and memory required and permit larger
+simulations to be performed on the available computing platforms.
 
@@
 toward reduced energy consumption will require incredibly high levels
-of concurrency to achieve the desired computation rates. The end
-result of these hardware changes is that the larger number of
-low-powered processors will be prone to both soft failures such as bit
-errors in floating point operations and hard failures where the data
-owned by that processor cannot be recovered. Because these failures
-are predicted to be common, resilient solver technologies are required
-to overcome these events at the application level. With solutions to
-deterministic transport problems based on Monte Carlo techniques, such
-issues are potentially alleviated by statistical arguments. In the
-case of soft failures, isolated floating point errors in the Monte
-Carlo simulation are absorbed within tally statistics while a hard
-failure is treated as a high variance event where some portion of the
-Monte Carlo histories lost.
+of concurrency to achieve the desired computation rates
+\cite{kogge_using_2011}. The end result of these hardware changes is
+that the larger number of low-powered processors will be prone to both
+soft failures such as bit errors in floating point operations and hard
+failures where the data owned by that processor cannot be recovered
+\cite{u.s._department_of_energy_resilient_2012}. Because these
+failures are predicted to be common, resilient solver technologies are
+are a possible means of overcoming these events at the application
+level. With solutions to deterministic transport problems based on
+Monte Carlo techniques, such issues are potentially alleviated by
+statistical arguments. In the case of soft failures, isolated floating
+point errors in the Monte Carlo simulation are absorbed within tally
+statistics while a hard failure is treated as a high variance event
+where some portion of the Monte Carlo histories lost.
 
@@
 of the physics of interest and desire to identify areas in which
-improvements can be made. In previous work \cite{evans_monte_2012}, we
+improvements can be made. In previous work \cite{evans_monte_2014}, we
 applied the Monte Carlo synthetic acceleration (MCSA) method to the
@@
 The paper is organized as follows. First, we briefly introduce the
-model $SP_N$ problem and describe the application of the MCSA
-algorithm. Next, we demonstrate that the basic Jacobi preconditioning
-applied in our previous work is unsatisfactory to achieve convergence
-for neutron transport in light water reactors and subsquently develop
-a new preconditioning scheme. We then apply this preconditioning
-scheme to a light water reactor fuel assembly and compare the
-performance of the preconditioned MCSA method to preconditioned Krylov
-methods.
+$SP_N$ equations for eigenvalue problems along with the discretization
+used for this work and the model problem. We then describe the
+application of the MCSA algorithm. Next, we demonstrate that the basic
+Jacobi preconditioning applied in our previous work is unsatisfactory
+to achieve convergence for neutron transport in light water reactors
+and subsquently develop a new preconditioning scheme. We then apply
+this preconditioning scheme to a light water reactor fuel assembly and
+compare the performance of the preconditioned MCSA method to
+preconditioned Krylov methods followed by a summary and conclusion.
 
@@
 To complete the spatial discretization, the gradient operators are
-resolved by a finite volume scheme as in
-\cite{evans_simplified_2013}. It should be noted here that with
-respect to a general MCSA scheme, the introduction of fission in the
-system will only affect the source vector in the linear system for the
-eigenvalue schemes used in this work. The linear operator, and
-therefore overall MCSA performance will be dictated by streaming and
-scattering as defined on the left-hand side.
-
-\subsection{Monte Carlo Synthetic Acceleration}
-\label{subsec:mcsa}
+resolved by a finite volume scheme as in \cite{hamilton_2014} such
+that the flux and current are continuous at cell boundaries.. It
+should be noted here that with respect to a general MCSA scheme, the
+introduction of fission in the system will only affect the source
+vector in the linear system for the eigenvalue schemes used in this
+work. The linear operator, and therefore overall MCSA performance will
+be dictated by streaming and scattering as defined on the left-hand
+side. We refer the reader to \cite{hamilton_2014} for an extensive
+presentation of the $SP_N$ discretization used in this work.
+
+\subsection{Fuel Assembly Model Problem}
+\label{subsec:fuel_assembly}
+Fuel assembly calculations are a critical piece of nuclear engineering
+infrastructure for reactor core analysis and design. At this level,
+individual fuel pins may be resolved at fine resolution in a variety
+of configurations. As a sophisticated problem of interest to push the
+limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
+used with varying energy group structure and $SP_N$ discretization in
+a criticality calculation. A cross section along the vertical axis
+showing homogenized fuel pin materials and the associated grid is
+given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
+the materials configuration along the horizontal axis is given by
+Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
+bottom is given in Figure~\ref{fig:problem3_end}. On the top and
+bottom of the assembly, vacuum conditions are used as well as on the
+top and right boundaries in
+Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
+on the left and bottom boundaries of
+Figure~\ref{fig:problem3_radial_mat}, effectively giving a
+representation of one quarter of the assembly. For the spatial
+discretization, each fuel pin (gray regions in
+Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
+mesh with materials and cross sections homogenized over this region.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4in]{problem3_radial_mat.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
+    \textit{Reflecting boundaries are used on the left and lower
+      boundaries to create a complete $17 \times 17$ assembly
+      geometry. Gray regions are homogenized fuel and red regions are
+      homogenized moderator. Each fuel pin is resolved by a $2 \times
+      2$ mesh.}}
+  \label{fig:problem3_radial_mat}
+\end{figure}
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=6.5in]{problem3_axial_mat.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
+      geometry is subdivided along the axial direction into 50 zones
+      spaced to capture material boundaries. Important details include
+      spacer grids along the length of the fuel pins and reactor core
+      structures on the top and bottom of the assembly. Lighter purple
+      material in the center of the assembly is moderator and darker
+      purple/red material is fuel.}}
+  \label{fig:problem3_axial_mat}
+\end{figure}
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=6.0in]{problem3_end.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly geometry end detail.}
+    \textit{Reactor core structure including spacer grids and plenum
+      has been included. Lighter purple material on the right of the
+      figure is moderator and darker purple/red material is fuel.}}
+  \label{fig:problem3_end}
+\end{figure}
+
+Significant geometric details are contained in the model including
+spacer grids, fuel pins with homogenized cladding and gas gap, core
+plenum, and moderator with boron. Group cross sections and other
+discrete nuclear data are generated as needed by a cross-section
+processing module dependent on the meshing parameters used to
+discretize the geometry and single-dimension pin-cell calculations for
+initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
+gives the primary design parameters for the fuel assembly
+calculations.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{ll}\hline\hline
+      \multicolumn{1}{l}{\textbf{Parameter}} & 
+      \multicolumn{1}{l}{\textbf{Value}} \\
+      Power Level & 0 MW \\
+      Inlet Temperature & 326.85C \\
+      Fuel Temperature & 600C \\
+      Boron Concentration & 1300 ppm \\
+      Moderator Density & 0.743 g/cc \\
+      Helium Density & \sn{1.79}{-4} g/cc \\
+      Zirconium Density & 6.56 g/cc \\
+      Stainless Steel Density & 8.0 g/cc \\
+      Inconel Density & 8.19 g/cc \\
+      UO2 Density & 10.257 g/cc \\
+      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
+      assembly criticality calculation.}}
+  \label{tab:problem3_parameters}
+\end{table}
+
+\clearpage
+
+%%---------------------------------------------------------------------------%%
+\section{Monte Carlo Synthetic Acceleration}
+\label{sec:mcsa}
 In this section we briefly review the Monte Carlo method for linear
@@
 directly to discrete Monte Carlo by forming the \textit{expected value
-  estimator}\footnote{The variance for this estimator is discussed in
-  Appendix~\ref{chap:estimator_variance}.} for a random walk of $k$
-events:
+  estimator} for a random walk of $k$ events:
 \begin{equation}
@@
 again indicates the eigenvalue iteration. We refer the reader to
-\cite{evans_monte_2012} for more a more complete definition of the
+\cite{evans_monte_2014} for more a more complete definition of the
 Monte Carlo method for linear systems and MCSA.
@@
 %%---------------------------------------------------------------------------%%
-\section{Fuel Assembly Model Problem}
-\label{sec:fuel_assembly}
-Fuel assembly calculations are a critical piece of nuclear engineering
-infrastructure for reactor core analysis and design. At this level,
-individual fuel pins may be resolved at fine resolution in a variety
-of configurations. As a sophisticated problem of interest to push the
-limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
-used with varying energy group structure and $SP_N$ discretization in
-a criticality calculation. A cross section along the vertical axis
-showing homogenized fuel pin materials and the associated grid is
-given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
-the materials configuration along the horizontal axis is given by
-Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
-bottom is given in Figure~\ref{fig:problem3_end}. On the top and
-bottom of the assembly, vacuum conditions are used as well as on the
-top and right boundaries in
-Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
-on the left and bottom boundaries of
-Figure~\ref{fig:problem3_radial_mat}, effectively giving a
-representation of one quarter of the assembly. For the spatial
-discretization, each fuel pin (gray regions in
-Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
-mesh with materials and cross sections homogenized over this region.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{problem3_radial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
-    \textit{Reflecting boundaries are used on the left and lower
-      boundaries to create a complete $17 \times 17$ assembly
-      geometry. Gray regions are homogenized fuel and red regions are
-      homogenized moderator. Each fuel pin is resolved by a $2 \times
-      2$ mesh.}}
-  \label{fig:problem3_radial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{problem3_axial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
-      geometry is subdivided along the axial direction into 50 zones
-      spaced to capture material boundaries. Important details include
-      spacer grids along the length of the fuel pins and reactor core
-      structures on the top and bottom of the assembly. Lighter purple
-      material in the center of the assembly is moderator and darker
-      purple/red material is fuel.}}
-  \label{fig:problem3_axial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{problem3_end.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry end detail.}
-    \textit{Reactor core structure including spacer grids and plenum
-      has been included. Lighter purple material on the right of the
-      figure is moderator and darker purple/red material is fuel.}}
-  \label{fig:problem3_end}
-\end{figure}
-
-Significant geometric details are contained in the model including
-spacer grids, fuel pins with homogenized cladding and gas gap, core
-plenum, and moderator with boron. Group cross sections and other
-discrete nuclear data are generated as needed by a cross-section
-processing module dependent on the meshing parameters used to
-discretize the geometry and single-dimension pin-cell calculations for
-initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
-gives the primary design parameters for the fuel assembly
-calculations.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Parameter}} & 
-      \multicolumn{1}{l}{\textbf{Value}} \\
-      Power Level & 0 MW \\
-      Inlet Temperature & 326.85C \\
-      Fuel Temperature & 600C \\
-      Boron Concentration & 1300 ppm \\
-      Moderator Density & 0.743 g/cc \\
-      Helium Density & \sn{1.79}{-4} g/cc \\
-      Zirconium Density & 6.56 g/cc \\
-      Stainless Steel Density & 8.0 g/cc \\
-      Inconel Density & 8.19 g/cc \\
-      UO2 Density & 10.257 g/cc \\
-      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
-      assembly criticality calculation.}}
-  \label{tab:problem3_parameters}
-\end{table}
-
-%%---------------------------------------------------------------------------%%
 \section{Results}
@@
 equation had success in using Jacobi preconditioning
-\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
+\cite{evans_monte_2014} and therefore we apply it here to the $SP_N$
 equations given their similar diffusion-like form. A single energy
@@
   \begin{center}
-    \includegraphics[width=5in]{block_jacobi_res.pdf}
+    \includegraphics[width=3.5in]{block_jacobi_res.pdf}
   \end{center}
@@
 making the method unusable.
+
 \begin{figure}[t!]
@@
 \end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=3.5in]{breakdown_time.pdf}
+  \end{center}
+  \caption{\textbf{CPU time per history as a function of spectral
+      radius for the neutron diffusion problem.} \textit{As the
+      spectral radius grows, so do the length of the random
+      walks. Longer random walks require more CPU time to compute.}}
+  \label{fig:breakdown_time}
+\end{figure}
 
@@
 is more applicable and in which performance is improved.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.5in]{breakdown_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time per history as a function of spectral
-      radius for the neutron diffusion problem.} \textit{As the
-      spectral radius grows, so do the length of the random
-      walks. Longer random walks require more CPU time to compute.}}
-  \label{fig:breakdown_time}
-\end{figure}
 
@@
 that MCSA can be competitive with these production Krylov methods
-using CPU time as a measure \cite{evans_monte_2012}.
+using CPU time as a measure \cite{evans_monte_2014}.
 

### 2014-05-12 doc/polynomial_note/polynomial.tex
Adding note on polynomial Monte Carlo.

\documentclass[10pt]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[labelfont=bf,labelsep=period]{caption}
\usepackage{booktabs}
\usepackage{multirow}

\title{Monte Carlo as Approximate Polynomial Preconditioning within MCSA}
\author{Steven Hamilton and Tom Evans and Stuart Slattery}

\newcommand{\bx}{\ensuremath{\mathbf{x}}}
\newcommand{\bb}{\ensuremath{\mathbf{b}}}
\newcommand{\br}{\ensuremath{\mathbf{r}}}
\newcommand{\bp}{\ensuremath{\mathbf{p}}}
\newcommand{\bw}{\ensuremath{\mathbf{w}}}
\newcommand{\bA}{\ensuremath{\mathbf{A}}}
\newcommand{\bH}{\ensuremath{\mathbf{H}}}
\newcommand{\bM}{\ensuremath{\mathbf{M}}}
\newcommand{\bP}{\ensuremath{\mathbf{P}}}
\newcommand{\bW}{\ensuremath{\mathbf{W}}}
\newcommand{\bI}{\ensuremath{\mathbf{I}}}

\begin{document}
\maketitle

\section{Background}
\label{sec:background}

Monte Carlo Synthetic Acceleration (MCSA), as described in Ref.~\cite{evans_13},
is an approach for solving the linear system
\begin{equation}
\bA \bx = \bb \label{eq:lin_problem}
\end{equation}
using a Monte Carlo (stochastic) process.
The original MCSA algorithm can be written as
\begin{equation}
\begin{aligned}
\bx^{k+1/2} &= \bx^{k} + \br^{k} \\
\bx^{k+1} &= \bx^{k+1/2} + \bM^{-1}\br^{k+1/2} \:,
\end{aligned}
\end{equation}
where $\bM^{-1}$ denotes performing an approximate linear solve using the
adjoint Monte Carlo process and $k$ is the iteration index.
Thus, MCSA consists of a single unpreconditioned
Richardson iteration followed by a single preconditioned Richardson iteration.
This iteration scheme has been observed to exhibit far better convergence
behavior than the original sequential Monte Carlo scheme proposed by Halton \cite{halton_94}.

The adjoint Monte Carlo process works by approximating the Neumann
series of the inverse of \bA:
\begin{equation}
\bA^{-1} = \sum_{n=0}^{\infty} ( \bI - \bA )^n
\equiv \sum_{n=0}^{\infty} \bH^n \:, \label{eq:neumann_inv}
\end{equation}
where \bH\ is the Richardson iteration matrix.
Correspondingly, the solution to a linear system of equations can
be written as
\begin{equation}
\bx = \sum_{n=0}^{\infty} \bH^n \bb \:. \label{eq:neumann_soln}
\end{equation}
Clearly
Eqs.~\eqref{eq:neumann_inv} and \eqref{eq:neumann_soln} are only
valid if the spectral radius of
the iteration matrix, $\rho(\bH)$, is less than unity.
The Monte Carlo process is conducted by decomposing the transpose of
\bH\ into an element-wise
(Hadamard) product of a probability matrix and a weight matrix, i.e.
\begin{equation}
\bH^T = \bP \circ \bW \:. \label{eq:mc_decomp}
\end{equation}
The usual approach is to define rows of \bP\ as normalized rows of $\bH^T$
(normalized columns of \bH), i.e.
\begin{equation}
\bP_{ij} =
\frac{\vert \bH_{ji} \vert}{\displaystyle \sum_{i=1}^{N} \vert \bH_{ij} \vert} \:.
\end{equation}
With this definition for \bP, the weight matrix is given by
\begin{equation}
\bW_{ij} = \frac{\bH_{ji}}{\bP_{ij}}. \label{eq:wgt_mat}
\end{equation}
Note that both \bP\ and \bW\ should have the same sparsity pattern
as $\bH^T$ and so Eq.~\eqref{eq:wgt_mat} is only taken over the indices corresponding
to nonzero entries in $\bH^T$.
The solution to a system of equations is approximated by following
a number of particle histories (although we are considering arbitrary
linear systems here, the nomenclature of tracking ``particles'' is
carried over from the particle transport community)
and averaging the contributions
to the solution over all conducted histories.  Each history is started
in a state determined by normalizing \bb to produce a probability
distribution:
\begin{equation}
\bp_i^\text{start} =
\frac{\vert \bb_i \vert}{\displaystyle \sum_{i=1}^{N} \vert \bb_i \vert} \:.
\end{equation}
The initial weight of a history is therefore given by
\begin{equation}
\bw_i^\text{start} = \frac{\bb_i}{\bp_i^\text{start}} \:.
\end{equation}
The history proceeds by allowing the particle to transition
to a new state by sampling from the discrete probability distribution
defined by the row of \bP\ corresponding to the current state.
The weight of the particle is updated by multiplying by the
entry in \bW\ corresponding to the given transition.
Each history provides a contribution to the solution vector proportional
to the current weight of the particle.  Either collision or
expected value tallies are possible, see [cite Slattery] for details
about these tally methods.
A given history can either be terminated via a weight cutoff (i.e.
when the weight of a particle drops below a prescribed fraction
of its initial weight) or by a history length cutoff (i.e. after
a certain number of transitions have taken place).

\section{Monte Carlo as Approximate Polynomial Preconditioning}
\label{sec:polynomial_prec}

As described in the previous section, the original idea behind
the adjoint Monte Carlo approach is to approximate the
Neumann series applied to the vector \bb.
However, by applying a history length cutoff it is possible to
view the Monte Carlo random walk as an approximation of a
fixed degree polynomial in $\bA$:
\begin{equation}
p(\bA) = \sum_{n=0}^{N} c_n \bA^n \:. \label{eq:power_poly}
\end{equation}
Alternatively, through a change of basis, it is possible to write
any such polynomial in powers of $\bI - \bA$ as
\begin{equation}
p(\bA) = \sum_{n=0}^{N} d_n (\bI - \bA)^n \:. \label{eq:neumann_poly}
\end{equation}
Note that the standard adjoint Monte Carlo algorithm corresponds
to the use of Eq.~\eqref{eq:neumann_poly} with $d_n=1$.

\subsection{Chebyshev Polynomials}
\label{subsec:chebyshev}

One possible candidate for a choice in polynomials is the
Chebyshev polynomials, denoted $T_N(x)$.
The Chebyshev polynomials have the property
that out of all monic polynomials of degree $N$,
$\frac{1}{2^{N-1}} T_N(x)$ is the polynomial whose maximum
absolute value is minimized over the interval $[-1,1]$.
One approach to defining a preconditioner, $\bM$, is to
attempt to minimize the eigenvalues of $\left( \bI - \bA \bM \right)$.
To accomplish this with $\bM$ as a polynomial of $\bA$, we can
first map the eigenvalues of $\bA$ into the interval $[-1,1]$
using the transformation
\begin{equation}
\ell (\bA) = \alpha \bI + \beta \bA \:,
\end{equation}
where $\alpha  \equiv -\dfrac{\lambda_\text{max} + \lambda_\text{min}}
{\lambda_\text{max} -  \lambda_\text{min}}$,
$\beta \equiv \dfrac{2}{\lambda_\text{max} -  \lambda_\text{min}}$,
and $\lambda_\text{min}$ and $\lambda_\text{max}$ are the
minimum and maximum eigenvalues of $\bA$, respectively.
To see that $\ell(A)$ maps the eigenvalues of $\bA$ to $[-1,1]$,
observe that $\ell(x)$ maps $\lambda_\text{min}$ to -1 and
$\lambda_\text{max}$ to 1.
Then we can select $\bM$ such that
\begin{equation}
\bI - \bA \bM  =
T_N \left( \ell(\bA) \right) T_N^{-1} \left( \ell(\mathbf{0}) \right) \:, \label{eq:cheby_prec}
\end{equation}
where $T_N^{-1} \left( \ell(\mathbf{0}) \right)$ forces the leading
term of the right hand side to have a coefficient of 1.
To determine an explicit representation for $\bM$, we first
use Eq.~\eqref{eq:power_poly} to write
\begin{equation}
\bM = \sum_{n=0}^{N-1} c_n \bA^n
\end{equation}
and therefore
\begin{equation}
\bI - \bA \bM = \bI - \sum_{n=0}^{N-1} c_n \bA^{n+1} \:, \label{eq:poly_prec}
\end{equation}
where the summation ends at $N-1$ in order for $\bI - \bA \bM$ to have
degree $N$.
One possible explicit representation for the Chebyshev polynomial of degree $N$ is
\begin{equation}
T_N(x) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
\begin{pmatrix} N-n \\ n \end{pmatrix} x^{N-2n} \:,
\end{equation}
and therefore
\begin{equation}
T_N \left( \ell(\bA) \right) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
\begin{pmatrix} N-n \\ n \end{pmatrix} \left( \alpha \bI + \beta \bA \right)^{N-2n}
\end{equation}
and
\begin{equation}
T_N \left( \ell(\bA) \right) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
\begin{pmatrix} N-n \\ n \end{pmatrix} \alpha^{N-2n} \bI \:.
\end{equation}
Thus an explicit representation for the polynomial preconditioner $\bM$ can
be obtained by equating the right hand sides of Eqs.~\eqref{eq:cheby_prec}
and \eqref{eq:poly_prec}.  For example, using a Chebyshev polynomial of
degree two provides a polynomial preconditioner of degree one described by
\begin{equation}
\begin{aligned}
c_0 &= - \frac{4 \alpha \beta}{2 \alpha^2 -1} \:, \\
c_1 &= - \frac{2 \beta^2}{2 \alpha^2 - 1} \:,
\end{aligned}
\end{equation}
and selecting a Chebyshev polynomial of degree three gives a polynomial
preconditioner of degree two described by
\begin{equation}
\begin{aligned}
c_0 &= - \frac{12 \alpha^2 \beta - 3 \beta}{4 \alpha^3 - 3 \alpha} \:, \\
c_1 &= - \frac{12 \alpha \beta^2}{4 \alpha^3 - 3 \alpha} \:, \\
c_2 &= - \frac{4 \beta^3}{4 \alpha^3 - 3 \alpha} \:.
\end{aligned}
\end{equation}

\subsection{GMRES Polynomial}
\label{subsec:gmres}

A significant downside to the use of Chebyshev polynomials is the need for
accurate bounds on the spectrum of $\bA$.  For nonsymmetric matrices,
this requires not only computing smallest and largest eigenvalues as is
required in the symmetric case but requires
bounding the entire spectrum of $\bA$ in an ellipse.  The process for
computing such bounding ellipses is described in Ref.~\cite{manteuffel_77}.
An alternative class of polynomials requiring less spectral information
while handling nonsymmetric systems naturally is the polynomials generated
by GMRES \cite{saad_86}.  The GMRES algorithm computes, at iteration $m$,
an approximate solution to \eqref{eq:lin_problem}, $\bx^{(m)}$, 
such that the $L_2$ norm of the residual
is minimized over the $(m+1)$-dimensional Krylov subspace given by
\begin{equation}
K_{m+1}(\bA,\bb) = \text{span}\{ \bb, \bA \bb, \ldots, \bA^{m} \bb \}.
\end{equation}
From this definition it is clear that GMRES is constructing (implicitly)
a polynomial of degree $m$, $p_m(\bA)$, such that $p_m(\bA) \approx \bA^{-1}$.
Use of the the GMRES polynomial as a preconditioner to fixed point iterations
was studied in Refs.~\cite{nachtigal_92,joubert_94} among several others
and as a preconditioner to {GMRES} itself in Ref.~\cite{liu_14}.

\section{Results}

As a simple test problem, we choose the ``JPWH 991'' matrix from
the Harwell-Boeing Collection, available from the Matrix Market.  
The sparsity pattern of the JPWH matrix is given in Figure~\ref{fig:jpwh_sparsity}.
\begin{figure}
\includegraphics[width=4.5in]{jpwh_sparsity}
\caption{Sparsity pattern of JPWH matrix \label{fig:jpwh_sparsity}}
\end{figure}
The matrix is nonsymmetric, but its eigenvalues are all real and
contained in the interval $[-16.3,-0.12]$.  After a diagonal row
scaling, the eigenvalues are contained in the interval $[0.02,1.71]$.
All test cases use the diagonally scaled matrix and the right
hand side of the linear system is taken to be a vector of all ones.
The stopping criterion is a reduction in the L$_2$ norm of the residual
by six orders of magnitude.

Table~\ref{tab:jpwh_standard_neumann} shows the behavior of
the standard MCSA algorithm applied to the JPWH matrix for a range
of different polynomial orders and Monte Carlo histories per iteration.
As expected, increasing the polynomial order generally leads to a decrease
in the number of iterations required for convergence.
For a fixed number of histories, however, there is a threshold beyond
which an increase in the polynomial order actually leads to very slow
convergence or even divergence.  Conversely, for a fixed polynomial order
it is necessary to track ``enough'' histories to ensure stability, but 
beyond that point there is little benefit to executing additional histories.
\begin{table}
\caption{Standard MCSA with Neumann Basis, JPWH Matrix. 
\label{tab:jpwh_standard_neumann}}
\centering
\begin{tabular}{cccccccccc}
\toprule
Histories per & \multicolumn{9}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 4 & 8 & 16 & 32 & 64 & 128 & 256 \\
\midrule
100 & 234 & 175 & 121 & 183 & - & - & - & - & - \\
200 & 230 & 173 & 118 & 74 & 164 & - & - & - & - \\
400 & 227 & 171 & 115 & 71 & 42 & 91 & - & - & - \\
800 & 227 & 170 & 114 & 70 & 39 & 31 & 47 & 67 & 78 \\
1600 & 226 & 170 & 113 & 69 & 39 & 22 & 22 & 27 & 27 \\
3200 & 225 & 169 & 113 & 68 & 38 & 21 & 15 & 17 & 17 \\
6400 & 225 & 169 & 113 & 68 & 38 & 21 & 12 & 12 & 13 \\
12800 & 225 & 169 & 113 & 68 & 38 & 21 & 11 & 10 & 10 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:jpwh_cheby_power} shows the same result,
but now instead of using $\bI - \bA$ as the basis for the
polynomial expansion, $\bA$ is used instead.  
Thus, instead of Eq.~\eqref{eq:mc_decomp} defining the
matrices to be used in the Monte Carlo process, we instead
have
\begin{equation}
\bA^T = \bP \circ \bW \:.
\end{equation}
We see that
for this basis, the majority of cases result in divergence.
Interestingly, at low polynomial orders and large history
counts, the behavior is nearly identical to that with 
$\bI - \bA$ as the basis.
\begin{table}
\caption{Standard MCSA with Monomial Basis, JPWH Matrix. 
\label{tab:jpwh_standard_power}}
\centering
\begin{tabular}{cccccccccc}
\toprule
Histories per & \multicolumn{9}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 4 & 8 & 16 & 32 & 64 & 128 & 256 \\
\midrule
    100 & -     & -     & - & - & - & - & - & - & - \\
    200 & 330 & -     & - & - & - & - & - & - & - \\
    400 & 240 & -     & - & - & - & - & - & - & - \\
    800 & 235 & 290 & - & - & - & - & - & - & - \\
  1600 & 230 & 182 & - & - & - & - & - & - & - \\
  3200 & 228 & 177 & - & - & - & - & - & - & - \\
  6400 & 227 & 174 & - & - & - & - & - & - & - \\
12800 & 226 & 172 & - & - & - & - & - & - & - \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:jpwh_cheby_neumann} shows the performance
with Chebyshev polynomials rather than the Neumann
series.  Note that the polynomial order is increased
linearly in this case rather than exponentially with the
Neumann series.  In general, using Chebyshev polynomials
results in a significant reduction in iterations relative to
the equivalent order Neumann series polynomial.  The
downside is that significantly more histories must be
tracked to achieve stability for a given polynomial order.
\begin{table}
\caption{Chebyshev MCSA with Neumann Basis, JPWH Matrix. 
\label{tab:jpwh_cheby_neumann}}
\centering
\begin{tabular}{cccccc}
 \toprule
 Histories per & \multicolumn{5}{c}{Polynomial Order} \\
 Iteration & 1 & 2 & 3 & 4 & 5 \\
 \midrule
 100 & - & - & - & - & - \\
 200 & 156 & - & - & - & - \\
 400 & 131 & - & - & - & - \\
 800 & 126 & 96 & - & - & - \\
 1600 & 125 & 67 & - & - & - \\
 3200 & 124 & 65 & 85 & - & - \\
 6400 & 123 & 64 & 42 & - & - \\
 12800 & 123 & 63 & 40 & 87 & - \\
 25600 & 123 & 63 & 40 & 33 & - \\
 51200 & 123 & 63 & 39 & 28 & 100 \\
 102400 & 122 & 63 & 39 & 28 & 33 \\
 204800 & 122 & 63 & 39 & 27 & 22 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:jpwh_cheby_power} shows the result of
again using Chebyshev polynomials, but now using
$\bA$ as the basis for the polynomial expansion.
As with the Neumann series polynomial, using
$\bA$ as a basis performs considerably worse than
using $\bI - \bA$ in that considerably more histories
are required to achieve stability for a given polynomial
order.
\begin{table}
\caption{Chebyshev MCSA with Monomial Basis, JPWH Matrix. 
\label{tab:jpwh_cheby_power}}
\centering
\begin{tabular}{cccccc}
 \toprule
 Histories per & \multicolumn{5}{c}{Polynomial Order} \\
 Iteration & 1 & 2 & 3 & 4 & 5 \\
 \midrule
 100 & - & - & - & - & - \\
 200 & - & - & - & - & - \\
 400 & - & - & - & - & - \\
 800 & - & - & - & - & - \\
 1600 & 178 & - & - & - & - \\
 3200 & 130 & - & - & - & - \\
 6400 & 127 & - & - & - & - \\
 12800 & 125 & - & - & - & - \\
 25600 & 124 & 80 & - & - & - \\
 51200 & 123 & 67 & - & - & - \\
 102400 & 123 & 65 & - & - & - \\
 204800 & 123 & 64 & - & - & - \\
\bottomrule
\end{tabular}
\end{table}

\section{Questions}

\begin{enumerate}

\item Using $\bI - \bA$ as a basis for the polynomial expansion rather
than $\bA$ results in a much more stable approach with either
Neumann series or Chebyshev polynomials.  What property of $\bH$
(or the resulting $\bP$ and $\bW$ matrices) is responsible for this
and is there a basis that is superior to $\bH$?

\item Chebyshev polynomials offer significant reduction in iteration
count for a given polynomial order relative to the Neumann series polynomial, 
but require significantly more
histories for stability.  Is there another class of polynomials that
might improve on the Neumann series without requiring as many
histories as Chebyshev?
\end{enumerate}

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2014-05-13 doc/polynomial_note/polynomial.tex
Adding discussion of computing GMRES polynomial coefficients.

@@
 \usepackage{graphicx}
+\usepackage[margin=1in]{geometry}
 \usepackage[labelfont=bf,labelsep=period]{caption}
@@
 \usepackage{multirow}
+\usepackage{subcaption}
 
@@
 \newcommand{\bx}{\ensuremath{\mathbf{x}}}
+\newcommand{\by}{\ensuremath{\mathbf{y}}}
 \newcommand{\bb}{\ensuremath{\mathbf{b}}}
+\newcommand{\bc}{\ensuremath{\mathbf{c}}}
 \newcommand{\br}{\ensuremath{\mathbf{r}}}
@@
 \newcommand{\bI}{\ensuremath{\mathbf{I}}}
+\newcommand{\bK}{\ensuremath{\mathbf{K}}}
+\newcommand{\calK}{\ensuremath{\mathcal{K}}}
+
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 \begin{equation}
-K_{m+1}(\bA,\bb) = \text{span}\{ \bb, \bA \bb, \ldots, \bA^{m} \bb \}.
+\calK_{m+1}(\bA,\bb) = \text{span}\{ \bb, \bA \bb, \ldots, \bA^{m} \bb \}.
 \end{equation}
@@
 Use of the the GMRES polynomial as a preconditioner to fixed point iterations
+(so-called hybrid GMRES methods)
 was studied in Refs.~\cite{nachtigal_92,joubert_94} among several others
 and as a preconditioner to {GMRES} itself in Ref.~\cite{liu_14}.
+
+In the context of hybrid GMRES methods, the coefficients of the GMRES polynomial
+are generally not computed explicitly, rather the polynomial is applied
+to a vector using relationships between the Krylov basis vectors arising from
+the Arnoldi process.  In order to use a polynomial within the Monte Carlo
+process, however, an explicit representation of the polynomial coefficients
+is required.  Multiple approaches for computation of these coefficients
+are possible.  One possibility is to observe that the eigenvalues of the
+upper Hessenberg matrix generated by the Arnoldi process are the roots of the
+GMRES polynomial.  Therefore the polynomial can be written as
+\begin{equation}
+p_m(x) = \prod_{i=1}^{m} (x - \lambda_i) \:, \label{eq:prod_gmres_poly}
+\end{equation}
+where $\lambda_i$ are the eigenvalues of the Hessenberg matrix.
+Computing the coefficients in the form of Eq.~\eqref{eq:power_poly}
+is then simply a matter of expanding the terms in Eq.~\eqref{eq:prod_gmres_poly}.
+A second method for computing the GMRES polynomial coefficients arises
+directly from the GMRES minimization problem:
+\begin{align}
+\bx^{(m)} = \argmin_x \Vert \bb - \bA \bx \Vert_2 \\
+\text{s.t.} \quad \bx \in \calK_{m+1}(\bA,\bb) \:.
+\end{align}
+By choosing the ``na\"ive'' basis for the Krylov subspace,
+\begin{equation}
+\bK_{m+1} = \left[ \bb, \bA \bb, \ldots, \bA^m \bb \right] \:,
+\end{equation}
+the coeffcients of the GMRES polynomial are given by the solution to
+the least squares problem
+\begin{equation}
+\min_{\bc} \Vert \bA \bK_{m+1} \bc - \bb \Vert_2 \:. \label{eq:gmres_min}
+\end{equation}
+In Ref.~\cite{liu_14}, the surprising observation was made that solving
+Eq.~\eqref{eq:gmres_min} using the normal equations actually produced
+a more accurate result than using a QR factorization to solve
+Eq.~\eqref{eq:gmres_min} or using Eq.~\eqref{eq:prod_gmres_poly}.
+In our experience, the selection of the preferred method is less clear
+cut.  Figure~\ref{fig:gmres_poly} shows the GMRES polynomial with coefficients
+computed using either Eq.~\eqref{eq:prod_gmres_poly} or by
+Eq.~\eqref{eq:gmres_min} with two different least squares solvers.
+The matrix \bA is a $25 \times 25$ 1-D Laplacian with $0.05$ added
+to each diagonal entry and the initial vector is taken to be a
+normally-distributed random vector.  At a polynomial order of 5,
+the two least squares methods are equivalent and appear to perform
+significantly better than the Hessenberg eigenvalue approach
+(i.e. the magnitude of the polynomials are smaller over the majority
+of the region of interest).  At a polynomial order of 10, the behavior
+is similar to the $m=5$ case except that a slight difference between the
+two least square solutions is now evident.  All methods perform fairly
+well for $m=15$, although the normal equations approach is noticably
+larger over portions of the domain.  Finally, for $m=20$ the
+eigenvalue-based approach appears the best with the least squares QR
+now appearing to suffer.  Table~\ref{tab:gmres_poly} shows the value
+of the spectral radius of $\bI - p_m(\bA) \bA$ for each
+of these cases.  Although the eigenvalue-based approach
+appears to perform worse at small polynomial orders, it actually
+produces the smallest spectral radius for all four cases and is the
+only approach to produce a spectral radius that is monotonically decreasing
+with polynomial order.
+The dramatic variability in the different methods for computing the
+coefficients of the GMRES polynomial are indicative of the ill-conditioned
+nature of computations involving polynomial coefficients.
+\begin{figure}
+\centering
+\begin{subfigure}[b]{0.49\linewidth}
+\includegraphics[width=\textwidth]{gmres_poly5}
+\caption{$m = 5$}
+\end{subfigure}
+\begin{subfigure}[b]{0.49\linewidth}
+\includegraphics[width=\textwidth]{gmres_poly10}
+\caption{$m = 10$}
+\end{subfigure}
+\begin{subfigure}[b]{0.49\linewidth}
+\includegraphics[width=\textwidth]{gmres_poly15}
+\caption{$m = 15$}
+\end{subfigure}
+\begin{subfigure}[b]{0.49\linewidth}
+\includegraphics[width=\textwidth]{gmres_poly20}
+\caption{$m = 20$}
+\end{subfigure}
+\caption{Plot of $q_{m+1}(x)=1-xp_{m}(x)$, where $p_m(x)$ is the GMRES
+ polynomial with different methods for computing the coefficients.
+ \bA\ is a shifted Laplacian with $N=25$.  Markers
+ indicate the polynomials evaluated at the eigenvalues of \bA.}
+\label{fig:gmres_poly}
+\end{figure}
+
+\begin{table}
+\centering
+\caption{Spectral Radius of $\left( \bI - p_m(\bA) \bA \right)$.}
+\label{tab:gmres_poly}
+\begin{tabular}{ccccc}
+\toprule
+& \multicolumn{4}{c}{Polynomial Order} \\
+\cmidrule(r){2-5}
+Method & 5 & 10 & 15 & 20 \\
+\midrule
+Hessenberg Eigenvalues     & 0.569 & 0.231 & 0.109 & 0.088 \\
+Least Squares Normal Eqns. & 0.632 & 0.259 & 0.184 & 0.247 \\
+Least Squares QR           & 0.632 & 0.273 & 0.176 & 0.586 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 \begin{figure}
+\centering
 \includegraphics[width=4.5in]{jpwh_sparsity}
@@
 
-\section{Questions}
-
-\begin{enumerate}
-
-\item Using $\bI - \bA$ as a basis for the polynomial expansion rather
-than $\bA$ results in a much more stable approach with either
-Neumann series or Chebyshev polynomials.  What property of $\bH$
-(or the resulting $\bP$ and $\bW$ matrices) is responsible for this
-and is there a basis that is superior to $\bH$?
-
-\item Chebyshev polynomials offer significant reduction in iteration
-count for a given polynomial order relative to the Neumann series polynomial, 
-but require significantly more
-histories for stability.  Is there another class of polynomials that
-might improve on the Neumann series without requiring as many
-histories as Chebyshev?
-\end{enumerate}
-
 \bibliographystyle{plain}

### 2014-05-13 doc/polynomial_note/polynomial.tex
Cleaning up Chebyshev polynomial section.

@@
 view the Monte Carlo random walk as an approximation of a
-fixed degree polynomial in $\bA$:
+fixed degree polynomial in $\bH$:
 \begin{equation}
-p(\bA) = \sum_{n=0}^{N} c_n \bA^n \:. \label{eq:power_poly}
+p(\bH) = \sum_{n=0}^{N} c_n \bH^n \:, \label{eq:neumann_poly}
 \end{equation}
-Alternatively, through a change of basis, it is possible to write
-any such polynomial in powers of $\bI - \bA$ as
+where in the standard algorithm, $c_n=1$.
+The Neumann series is not the only polynomial that can approximate
+$\bA^{-1}$, however.  By varying the selection of $c_n$, it is
+possible to use other polynomials.  Commonly, polynomials used
+for preconditioning are defined in terms of powers of \bA\ rather
+than powers of \bH, giving the expansion
 \begin{equation}
-p(\bA) = \sum_{n=0}^{N} d_n (\bI - \bA)^n \:. \label{eq:neumann_poly}
+p(\bA) = \sum_{n=0}^{N} \hat{c}_n \bA^n \:. \label{eq:power_poly}
 \end{equation}
-Note that the standard adjoint Monte Carlo algorithm corresponds
-to the use of Eq.~\eqref{eq:neumann_poly} with $d_n=1$.
+Converting between coefficients defined by Eq.~\eqref{eq:power_poly}
+and those defined by Eq.~\eqref{eq:neumann_poly} can be
+accomplished by means of a change of basis.  The impact of the
+choice of the basis will be evaluated in Section~\ref{sec:results}.
+We now consider several candidates for the matrix polynomial.
 
@@
 
-One possible candidate for a choice in polynomials is the
-Chebyshev polynomials, denoted $T_N(x)$.
-The Chebyshev polynomials have the property
-that out of all monic polynomials of degree $N$,
-$\frac{1}{2^{N-1}} T_N(x)$ is the polynomial whose maximum
+One approach to defining a preconditioner, $\bM$, is to
+attempt to minimize the eigenvalues of $\bI - p(\bA) \bA$.
+The Chebyshev polynomials, denoted $T_m(x)$ have the property
+that out of all monic polynomials of degree $m$,
+$\frac{1}{2^{m-1}} T_m(x)$ is the polynomial whose maximum
 absolute value is minimized over the interval $[-1,1]$.
-One approach to defining a preconditioner, $\bM$, is to
-attempt to minimize the eigenvalues of $\left( \bI - \bA \bM \right)$.
-To accomplish this with $\bM$ as a polynomial of $\bA$, we can
-first map the eigenvalues of $\bA$ into the interval $[-1,1]$
-using the transformation
+Thus, if the spectrum of \bA is real, then an optimal polynomial
+preconditioner can be achieved by mapping the eigenvalues of
+\bA\ into the interval $[-1,1]$ using the transformation
 \begin{equation}
@@
 minimum and maximum eigenvalues of $\bA$, respectively.
-To see that $\ell(A)$ maps the eigenvalues of $\bA$ to $[-1,1]$,
+To see that $\ell(A)$ maps the eigenvalues of $\bA$ into $[-1,1]$,
 observe that $\ell(x)$ maps $\lambda_\text{min}$ to -1 and
 $\lambda_\text{max}$ to 1.
-Then we can select $\bM$ such that
+Then we can select $p(\bA)$ such that
 \begin{equation}
-\bI - \bA \bM  =
-T_N \left( \ell(\bA) \right) T_N^{-1} \left( \ell(\mathbf{0}) \right) \:, \label{eq:cheby_prec}
+\bI - p(\bA) \bA  = \frac{1}{T_{m+1}(\alpha)}
+T_{m+1} \left( \ell(\bA) \right) \:, \label{eq:cheby_prec}
 \end{equation}
-where $T_N^{-1} \left( \ell(\mathbf{0}) \right)$ forces the leading
+where $\frac{1}{T_m ( \alpha )}$ forces the leading
 term of the right hand side to have a coefficient of 1.
-To determine an explicit representation for $\bM$, we first
+To determine an explicit representation for $p(\bA)$, we first
 use Eq.~\eqref{eq:power_poly} to write
 \begin{equation}
-\bM = \sum_{n=0}^{N-1} c_n \bA^n
+p(\bA) = \sum_{n=0}^{m} c_n \bA^n
 \end{equation}
@@
 \begin{equation}
-\bI - \bA \bM = \bI - \sum_{n=0}^{N-1} c_n \bA^{n+1} \:, \label{eq:poly_prec}
+\bI - p(\bA) \bA = \bI - \sum_{n=0}^{m} c_n \bA^{n+1} \:. \label{eq:poly_prec}
 \end{equation}
-where the summation ends at $N-1$ in order for $\bI - \bA \bM$ to have
-degree $N$.
-One possible explicit representation for the Chebyshev polynomial of degree $N$ is
+One possible explicit representation for the Chebyshev polynomial of degree $m$ is
 \begin{equation}
-T_N(x) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
-\begin{pmatrix} N-n \\ n \end{pmatrix} x^{N-2n} \:,
+T_m(x) = \sum_{n=0}^{\lfloor \frac{m}{2} \rfloor} \left( -1 \right)^n 2^{m - 2n -1} \frac{m}{m-n}
+\begin{pmatrix} m-n \\ n \end{pmatrix} x^{m-2n} \:,
 \end{equation}
@@
 \begin{equation}
-T_N \left( \ell(\bA) \right) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
-\begin{pmatrix} N-n \\ n \end{pmatrix} \left( \alpha \bI + \beta \bA \right)^{N-2n}
+T_m \left( \ell(\bA) \right) = \sum_{n=0}^{\lfloor \frac{m}{2} \rfloor} \left( -1 \right)^n 2^{m - 2n -1} \frac{m}{m-n}
+\begin{pmatrix} m-n \\ n \end{pmatrix} \left( \alpha \bI + \beta \bA \right)^{m-2n} \:.
 \end{equation}
-and
-\begin{equation}
-T_N \left( \ell(\bA) \right) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
-\begin{pmatrix} N-n \\ n \end{pmatrix} \alpha^{N-2n} \bI \:.
-\end{equation}
-Thus an explicit representation for the polynomial preconditioner $\bM$ can
+Thus an explicit representation for the polynomial preconditioner can
 be obtained by equating the right hand sides of Eqs.~\eqref{eq:cheby_prec}
@@
 
-\subsection{GMRES Polynomial}
-\label{subsec:gmres}
-
 A significant downside to the use of Chebyshev polynomials is the need for
@@
 computing such bounding ellipses is described in Ref.~\cite{manteuffel_77}.
+
+\subsection{GMRES Polynomial}
+\label{subsec:gmres}
+
 An alternative class of polynomials requiring less spectral information
+than Chebyshev polynomials
 while handling nonsymmetric systems naturally is the polynomials generated
@@
 \section{Results}
+\label{sec:results}
 

### 2014-05-14 doc/polynomial_note/polynomial.tex
Updating GMRES polynomial figures and discussion.

@@
 Eq.~\eqref{eq:gmres_min} or using Eq.~\eqref{eq:prod_gmres_poly}.
-In our experience, the selection of the preferred method is less clear
-cut.  Figure~\ref{fig:gmres_poly} shows the GMRES polynomial with coefficients
+Our experience, however, differs from this result.
+Figure~\ref{fig:gmres_poly} shows the GMRES polynomial with coefficients
 computed using either Eq.~\eqref{eq:prod_gmres_poly} or by
@@
 is similar to the $m=5$ case except that a slight difference between the
-two least square solutions is now evident.  All methods perform fairly
-well for $m=15$, although the normal equations approach is noticably
-larger over portions of the domain.  Finally, for $m=20$ the
-eigenvalue-based approach appears the best with the least squares QR
-now appearing to suffer.  Table~\ref{tab:gmres_poly} shows the value
+two least square solutions is now evident.
+For orders $15$ and $20$, however, both the eigenvalue-based approach
+and the least squares QR perform very well while the normal equations
+solution begins to rapidly degrade.
+Table~\ref{tab:gmres_poly} shows the value
 of the spectral radius of $\bI - p_m(\bA) \bA$ for each
-of these cases.  Although the eigenvalue-based approach
-appears to perform worse at small polynomial orders, it actually
-produces the smallest spectral radius for all four cases and is the
-only approach to produce a spectral radius that is monotonically decreasing
-with polynomial order.
-The dramatic variability in the different methods for computing the
+of these cases.  Both the eigenvalue approach and the least squares QR
+produce monotonically decreasing spectral radii whereas the spectral
+radius produced by the normal equations increases at high polynomial
+orders.  The least squares QR method seems to produce the most consistent
+behavior over the full range of polynomial orders.
+The significant variability in the different methods for computing the
 coefficients of the GMRES polynomial are indicative of the ill-conditioned
@@
 \midrule
-Hessenberg Eigenvalues     & 0.569 & 0.231 & 0.109 & 0.088 \\
-Least Squares Normal Eqns. & 0.632 & 0.259 & 0.184 & 0.247 \\
-Least Squares QR           & 0.632 & 0.273 & 0.176 & 0.586 \\
+Hessenberg Eigenvalues     & 1.027 & 0.630 & 0.081 & 0.034 \\
+Least Squares Normal Eqns. & 0.686 & 0.268 & 0.162 & 0.575 \\
+Least Squares QR           & 0.686 & 0.281 & 0.085 & 0.033 \\
 \bottomrule

### 2014-05-19 doc/polynomial_note/polynomial.tex
Adding results on shifted Laplacian.

@@
 
-As a simple test problem, we choose the ``JPWH 991'' matrix from
-the Harwell-Boeing Collection, available from the Matrix Market.
-The sparsity pattern of the JPWH matrix is given in Figure~\ref{fig:jpwh_sparsity}.
-\begin{figure}
-\centering
-\includegraphics[width=4.5in]{jpwh_sparsity}
-\caption{Sparsity pattern of JPWH matrix \label{fig:jpwh_sparsity}}
-\end{figure}
-The matrix is nonsymmetric, but its eigenvalues are all real and
-contained in the interval $[-16.3,-0.12]$.  After a diagonal row
-scaling, the eigenvalues are contained in the interval $[0.02,1.71]$.
 All test cases use the diagonally scaled matrix and the right
@@
 \begin{table}
-\caption{Standard MCSA with Neumann Basis, JPWH Matrix.
-\label{tab:jpwh_standard_neumann}}
+\caption{Forward MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+Values are MCSA iteration counts (timing in milliseconds)
+\label{tab:lap_forward_neumann}}
 \centering
-\begin{tabular}{cccccccccc}
+\begin{tabular}{ccccccc}
 \toprule
-Histories per & \multicolumn{9}{c}{Polynomial Order} \\
-Iteration & 1 & 2 & 4 & 8 & 16 & 32 & 64 & 128 & 256 \\
+Histories per & \multicolumn{6}{c}{Polynomial Order} \\
+Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
 \midrule
-100 & 234 & 175 & 121 & 183 & - & - & - & - & - \\
-200 & 230 & 173 & 118 & 74 & 164 & - & - & - & - \\
-400 & 227 & 171 & 115 & 71 & 42 & 91 & - & - & - \\
-800 & 227 & 170 & 114 & 70 & 39 & 31 & 47 & 67 & 78 \\
-1600 & 226 & 170 & 113 & 69 & 39 & 22 & 22 & 27 & 27 \\
-3200 & 225 & 169 & 113 & 68 & 38 & 21 & 15 & 17 & 17 \\
-6400 & 225 & 169 & 113 & 68 & 38 & 21 & 12 & 12 & 13 \\
-12800 & 225 & 169 & 113 & 68 & 38 & 21 & 11 & 10 & 10 \\
+250 & - & - & - & - & - & - \\
+500 & - & - & - & - & - & - \\
+1000 & - & - & - & - & - & - \\
+2000 & - & - & - & - & - & - \\
+4000 & - & 765(249) & 605(285) & 947(560) & 396(302) & 560(458) \\
+8000 & 923(325) & 698(413) & 554(485) & 471(560) & 396(530) & 354(561) \\
+16000 & 923(616) & 695(809) & 554(946) & 463(1015) & 396(1068) & 348(1079) \\
 \bottomrule
@@
 
-Table~\ref{tab:jpwh_cheby_power} shows the same result,
-but now instead of using $\bI - \bA$ as the basis for the
-polynomial expansion, $\bA$ is used instead.
-Thus, instead of Eq.~\eqref{eq:mc_decomp} defining the
-matrices to be used in the Monte Carlo process, we instead
-have
-\begin{equation}
-\bA^T = \bP \circ \bW \:.
-\end{equation}
-We see that
-for this basis, the majority of cases result in divergence.
-Interestingly, at low polynomial orders and large history
-counts, the behavior is nearly identical to that with
-$\bI - \bA$ as the basis.
 \begin{table}
-\caption{Standard MCSA with Monomial Basis, JPWH Matrix.
-\label{tab:jpwh_standard_power}}
+\caption{Adjoint MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+Values are MCSA iteration counts (timing in milliseconds)
+\label{tab:lap_adjoint_neumann}}
 \centering
-\begin{tabular}{cccccccccc}
+\begin{tabular}{ccccccc}
 \toprule
-Histories per & \multicolumn{9}{c}{Polynomial Order} \\
-Iteration & 1 & 2 & 4 & 8 & 16 & 32 & 64 & 128 & 256 \\
+Histories per & \multicolumn{6}{c}{Polynomial Order} \\
+Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
 \midrule
-    100 & -     & -     & - & - & - & - & - & - & - \\
-    200 & 330 & -     & - & - & - & - & - & - & - \\
-    400 & 240 & -     & - & - & - & - & - & - & - \\
-    800 & 235 & 290 & - & - & - & - & - & - & - \\
-  1600 & 230 & 182 & - & - & - & - & - & - & - \\
-  3200 & 228 & 177 & - & - & - & - & - & - & - \\
-  6400 & 227 & 174 & - & - & - & - & - & - & - \\
-12800 & 226 & 172 & - & - & - & - & - & - & - \\
+250 & - & 775(100) & - & 651(96) & - & 664(110) \\
+500 & - & 725(122) & - & 502(104) & - & 394(95) \\
+1000 & - & 707(174) & - & 482(179) & - & 366(144) \\
+2000 & 2096(521) & 703(280) & 1991(855) & 471(259) & 2668(1514) & 356(251) \\
+4000 & 1694(821) & 698(494) & 1701(1194) & 467(497) & 864(921) & 350(458) \\
+8000 & 1384(1142) & 697(923) & 1256(1761) & 464(905) & 628(1290) & 350(873) \\
+16000 & 1220(1969) & 695(1796) & 859(2493) & 462(1768) & 588(2384) & 347(1711) \\
 \bottomrule
@@
 
-Table~\ref{tab:jpwh_cheby_neumann} shows the performance
-with Chebyshev polynomials rather than the Neumann
-series.  Note that the polynomial order is increased
-linearly in this case rather than exponentially with the
-Neumann series.  In general, using Chebyshev polynomials
-results in a significant reduction in iterations relative to
-the equivalent order Neumann series polynomial.  The
-downside is that significantly more histories must be
-tracked to achieve stability for a given polynomial order.
 \begin{table}
-\caption{Chebyshev MCSA with Neumann Basis, JPWH Matrix.
-\label{tab:jpwh_cheby_neumann}}
+\caption{Forward MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+Values are MCSA iteration counts (timing in milliseconds)
+\label{tab:lap_forward_cheby}}
 \centering
-\begin{tabular}{cccccc}
+\begin{tabular}{ccccccc}
 \toprule
- Histories per & \multicolumn{5}{c}{Polynomial Order} \\
- Iteration & 1 & 2 & 3 & 4 & 5 \\
+Histories per & \multicolumn{6}{c}{Polynomial Order} \\
+Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
 \midrule
- 100 & - & - & - & - & - \\
- 200 & 156 & - & - & - & - \\
- 400 & 131 & - & - & - & - \\
- 800 & 126 & 96 & - & - & - \\
- 1600 & 125 & 67 & - & - & - \\
- 3200 & 124 & 65 & 85 & - & - \\
- 6400 & 123 & 64 & 42 & - & - \\
- 12800 & 123 & 63 & 40 & 87 & - \\
- 25600 & 123 & 63 & 40 & 33 & - \\
- 51200 & 123 & 63 & 39 & 28 & 100 \\
- 102400 & 122 & 63 & 39 & 28 & 33 \\
- 204800 & 122 & 63 & 39 & 27 & 22 \\
+250 & - & - & - & - & - & - \\
+500 & - & - & - & - & - & - \\
+1000 & - & - & - & - & - & - \\
+2000 & - & - & - & - & - & - \\
+4000 & - & - & - & - & - & - \\
+8000 & - & - & - & - & - & - \\
+16000 & 556(369) & - & - & - & - & - \\
 \bottomrule
@@
 
-Table~\ref{tab:jpwh_cheby_power} shows the result of
-again using Chebyshev polynomials, but now using
-$\bA$ as the basis for the polynomial expansion.
-As with the Neumann series polynomial, using
-$\bA$ as a basis performs considerably worse than
-using $\bI - \bA$ in that considerably more histories
-are required to achieve stability for a given polynomial
-order.
 \begin{table}
-\caption{Chebyshev MCSA with Monomial Basis, JPWH Matrix.
-\label{tab:jpwh_cheby_power}}
+\caption{Adjoint MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+Values are MCSA iteration counts (timing in milliseconds)
+\label{tab:lap_adjoint_cheby}}
 \centering
-\begin{tabular}{cccccc}
+\begin{tabular}{ccccccc}
 \toprule
- Histories per & \multicolumn{5}{c}{Polynomial Order} \\
- Iteration & 1 & 2 & 3 & 4 & 5 \\
+Histories per & \multicolumn{6}{c}{Polynomial Order} \\
+Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
 \midrule
- 100 & - & - & - & - & - \\
- 200 & - & - & - & - & - \\
- 400 & - & - & - & - & - \\
- 800 & - & - & - & - & - \\
- 1600 & 178 & - & - & - & - \\
- 3200 & 130 & - & - & - & - \\
- 6400 & 127 & - & - & - & - \\
- 12800 & 125 & - & - & - & - \\
- 25600 & 124 & 80 & - & - & - \\
- 51200 & 123 & 67 & - & - & - \\
- 102400 & 123 & 65 & - & - & - \\
- 204800 & 123 & 64 & - & - & - \\
+250 & - & - & - & - & - & - \\
+500 & - & - & - & - & - & - \\
+1000 & - & - & - & - & - & - \\
+2000 & - & 325(130) & - & - & - & - \\
+4000 & - & 294(210) & - & - & - & - \\
+8000 & - & 288(380) & 201(335) & - & - & - \\
+16000 & - & 285(736) & 175(554) & - & - & - \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+\caption{Forward MCSA with GMRES Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+Values are MCSA iteration counts (timing in milliseconds)
+\label{tab:lap_forward_gmres}}
+\centering
+\begin{tabular}{ccccccc}
+\toprule
+Histories per & \multicolumn{6}{c}{Polynomial Order} \\
+Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
+\midrule
+250 & - & - & - & - & - & - \\
+500 & - & - & - & - & - & - \\
+1000 & 923(89) & - & - & - & - & - \\
+2000 & 923(127) & 547(110) & - & - & - & - \\
+4000 & 923(195) & 547(204) & - & - & - & - \\
+8000 & 923(330) & 547(322) & 357(312) & - & - & - \\
+16000 & 923(607) & 547(640) & 357(607) & 246(536) & - & - \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+\caption{Adjoint MCSA with GMRES Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+Values are MCSA iteration counts (timing in milliseconds)
+\label{tab:lap_adjoint_gmres}}
+\centering
+\begin{tabular}{ccccccc}
+\toprule
+Histories per & \multicolumn{6}{c}{Polynomial Order} \\
+Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
+\midrule
+250 & 1153(136) & 2572(337) & - & - & - & - \\
+500 & 1003(151) & 616(105) & 462(87) & - & - & - \\
+1000 & 946(226) & 566(142) & 385(110) & 496(161) & - & - \\
+2000 & 933(306) & 557(222) & 368(175) & 260(146) & - & - \\
+4000 & 928(519) & 554(412) & 362(311) & 254(259) & 203(236) & - \\
+8000 & 924(971) & 549(731) & 362(585) & 251(490) & 184(423) & - \\
+16000 & 923(1816) & 548(1426) & 360(1135) & 247(965) & 179(797) & 146(744) \\
 \bottomrule

### 2014-05-21 doc/polynomial_note/polynomial.tex
Edits to polynomial note.

@@
 \usepackage{graphicx}
+\usepackage{tabularx}
 \usepackage[margin=1in]{geometry}
@@
 \title{Monte Carlo as Approximate Polynomial Preconditioning within MCSA}
-\author{Steven Hamilton and Tom Evans and Stuart Slattery}
+\author{Steven Hamilton}
 
@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+\newcolumntype{Y}{>{\centering\arraybackslash}X}
 
@@
 \end{equation}
-where $\bM^{-1}$ denotes performing an approximate linear solve using the
-adjoint Monte Carlo process and $k$ is the iteration index.
+where $\bM^{-1}$ denotes performing an approximate linear solve using a
+Monte Carlo process and $k$ is the iteration index.
 Thus, MCSA consists of a single unpreconditioned
@@
 This iteration scheme has been observed to exhibit far better convergence
-behavior than the original sequential Monte Carlo scheme proposed by Halton \cite{halton_94}.
+behavior than the original sequential Monte Carlo scheme proposed by Halton
+\cite{halton_94}.
 
-The adjoint Monte Carlo process works by approximating the Neumann
+Both forward (direct) and adjoint Monte Carlo processes have been considered
+\cite{slattery_13,evans_13}.
+Both approaches work by approximating the Neumann
 series of the inverse of \bA:
@@
 the iteration matrix, $\rho(\bH)$, is less than unity.
-The Monte Carlo process is conducted by decomposing the transpose of
+
+The forward Monte Carlo process is conducted by decomposing the transpose of
 \bH\ into an element-wise
@@
 \begin{equation}
-\bH^T = \bP \circ \bW \:. \label{eq:mc_decomp}
+\bH = \bP \circ \bW \:. \label{eq:forward_mc_decomp}
 \end{equation}
-The usual approach is to define rows of \bP\ as normalized rows of $\bH^T$
+The usual approach is to define rows of \bP\ as normalized rows of \bH\
 (normalized columns of \bH), i.e.
@@
 \bP_{ij} =
-\frac{\vert \bH_{ji} \vert}{\displaystyle \sum_{i=1}^{N} \vert \bH_{ij} \vert} \:.
+\frac{\vert \bH_{ij} \vert}{\displaystyle \sum_{j=1}^{N} \vert \bH_{ij} \vert} \:.
 \end{equation}
@@
 \begin{equation}
-\bW_{ij} = \frac{\bH_{ji}}{\bP_{ij}}. \label{eq:wgt_mat}
+\bW_{ij} = \frac{\bH_{ij}}{\bP_{ij}}. \label{eq:wgt_mat}
 \end{equation}
@@
 to nonzero entries in $\bH^T$.
+A similar process is conducted in the adjoint Monte Carlo process,
+except rather than Eq.~\eqref{eq:forward_mc_decomp}, the {\em transpose}
+of \bH\ is decomposed, i.e.
+\begin{equation}
+\bH^T = \bP \circ \bW \:. \label{eq:adjoint_mc_decomp}
+\end{equation}
 The solution to a system of equations is approximated by following
-a number of particle histories (although we are considering arbitrary
-linear systems here, the nomenclature of tracking ``particles'' is
-carried over from the particle transport community)
-and averaging the contributions
-to the solution over all conducted histories.  Each history is started
-in a state determined by normalizing \bb to produce a probability
-distribution:
-\begin{equation}
-\bp_i^\text{start} =
-\frac{\vert \bb_i \vert}{\displaystyle \sum_{i=1}^{N} \vert \bb_i \vert} \:.
-\end{equation}
-The initial weight of a history is therefore given by
-\begin{equation}
-\bw_i^\text{start} = \frac{\bb_i}{\bp_i^\text{start}} \:.
-\end{equation}
-The history proceeds by allowing the particle to transition
-to a new state by sampling from the discrete probability distribution
-defined by the row of \bP\ corresponding to the current state.
-The weight of the particle is updated by multiplying by the
-entry in \bW\ corresponding to the given transition.
-Each history provides a contribution to the solution vector proportional
-to the current weight of the particle.  Either collision or
-expected value tallies are possible, see [cite Slattery] for details
-about these tally methods.
-A given history can either be terminated via a weight cutoff (i.e.
-when the weight of a particle drops below a prescribed fraction
-of its initial weight) or by a history length cutoff (i.e. after
-a certain number of transitions have taken place).
+a number of histories and averaging the contributions
+to the solution over all conducted histories.  Histories can be terminated
+when the weight decreases by a prescribed fraction (denoted a weight cutoff)
+or, alternatively, a history can be terminated after a fixed number of
+transitions have taken place (denoted a history length cutoff).
+
+The primary difference
+between the forward and adjoint methods is that in the forward method
+a contribution is made to the element of the solution vector where the
+history originated based on the value of the right hand side in the state
+where the history currently resides.  In the adjoint method, however,
+a contribution is made to the element of the solution vector where the
+history currently resides based on the value of the right hand side
+where the history originated.  In the forward method, therefore, it is
+necessary to start at least one history in every state in order to
+estimate every element of the solution vector whereas in the adjoint
+method an estimate of all entries in the solution vector can be
+obtained with far fewer than one history per state.
+Further details about the Monte Carlo process, including selection
+of starting probabilities and different solution tally approaches,
+can be found in Ref.~\cite{slattery_13}.
 
@@
 view the Monte Carlo random walk as an approximation of a
-fixed degree polynomial in $\bH$:
+fixed degree polynomial:
 \begin{equation}
-p(\bH) = \sum_{n=0}^{N} c_n \bH^n \:, \label{eq:neumann_poly}
+p(\bA) = \sum_{n=0}^{m} c_n \left( \bI - \bA \right)^n \:, \label{eq:neumann_poly}
 \end{equation}
@@
 for preconditioning are defined in terms of powers of \bA\ rather
-than powers of \bH, giving the expansion
+than powers of $\bI-\bA$, giving the expansion
 \begin{equation}
-p(\bA) = \sum_{n=0}^{N} \hat{c}_n \bA^n \:. \label{eq:power_poly}
+p(\bA) = \sum_{n=0}^{m} \hat{c}_n \bA^n \:. \label{eq:power_poly}
 \end{equation}
@@
 and those defined by Eq.~\eqref{eq:neumann_poly} can be
-accomplished by means of a change of basis.  The impact of the
-choice of the basis will be evaluated in Section~\ref{sec:results}.
+accomplished by means of a change of basis.  It should be denoted
+that using $\bI - \bA$ as a basis is generally preferred over using
+\bA\ itself.  The reason for this is that if a diagonal scaling
+is first applied to \bA, then $\bI-\bA$ will have all zeros
+on the main diagonal which eliminates Monte Carlo events in which
+a history does not change states.
 We now consider several candidates for the matrix polynomial.
@@
 required in the symmetric case but requires
-bounding the entire spectrum of $\bA$ in an ellipse.  The process for
+bounding the entire spectrum of $\bA$ in an ellipse.  A process for
 computing such bounding ellipses is described in Ref.~\cite{manteuffel_77}.
@@
 \subsection{GMRES Polynomial}
-\label{subsec:gmres}
+\label{subsec:gmres_poly}
 
@@
 by GMRES \cite{saad_86}.  The GMRES algorithm computes, at iteration $m$,
-an approximate solution to \eqref{eq:lin_problem}, $\bx^{(m)}$,
+an approximate solution to Eq.~\eqref{eq:lin_problem}, $\bx^{(m)}$,
 such that the $L_2$ norm of the residual
@@
 a polynomial of degree $m$, $p_m(\bA)$, such that $p_m(\bA) \approx \bA^{-1}$.
-Use of the the GMRES polynomial as a preconditioner to fixed point iterations
-(so-called hybrid GMRES methods)
-was studied in Refs.~\cite{nachtigal_92,joubert_94} among several others
-and as a preconditioner to {GMRES} itself in Ref.~\cite{liu_14}.
+Use of the the GMRES polynomial has been studied as a preconditioner
+to fixed point iterations (so-called hybrid GMRES methods)
+\cite{elman_86,nachtigal_92,joubert_94}
+and as a preconditioner to {GMRES} itself \cite{liu_14}.
+Additionally, using the GMRES/Arnoldi process to obtain information about
+the matrix for use in a subsequent iterative process has also been
+studied \cite{elman_86_2,saylor_90,saylor_91}.
 
@@
 \end{equation}
-the coeffcients of the GMRES polynomial are given by the solution to
+the coefficients of the GMRES polynomial are given by the solution to
 the least squares problem
@@
 Eq.~\eqref{eq:gmres_min} with two different least squares solvers.
-The matrix \bA is a $25 \times 25$ 1-D Laplacian with $0.05$ added
+The matrix \bA\ is a $25 \times 25$ 1-D Laplacian with $0.05$ added
 to each diagonal entry and the initial vector is taken to be a
@@
 \end{figure}
-
 \begin{table}
@@
 
+A significant advantage of GMRES polynomials over Chebyshev is that
+GMRES polynomials handle nonsymmetric matrices with potentially complex
+eigenvalue naturally, whereas computing the coefficients of
+Chebyshev polynomials requires the costly task of bounding the spectrum
+of \bA\ in an ellipse.  Additionally, accurate knowledge of the bounding eigenvalues
+of \bA\ are required to compute the Chebyshev coefficients of any order
+(which may incur a cost of tens or hundreds of matrix-vector products),
+while the computation of GMRES polynomial coefficients requires only
+$m$ matrix-vector products for an order $m$ polynomial.
+
 \section{Results}
@@
 
-All test cases use the diagonally scaled matrix and the right
-hand side of the linear system is taken to be a vector of all ones.
-The stopping criterion is a reduction in the L$_2$ norm of the residual
+Tables~\ref{tab:lap_forward_neumann} and \ref{tab:lap_adjoint_neumann}
+show the behavior of
+the standard MCSA algorithm (with forward and adjoint Monte Carlo, respectively)
+applied to a $1000 \times 1000$ shifted Laplacian matrix with a shift
+of $0.01$.  The right hand side is given by $\bb_i = \sin \left( \pi \frac{i-1}{N-1} \right)$
+and the stopping criterion is a reduction in the L$_2$ norm of the residual
 by six orders of magnitude.
-
-Table~\ref{tab:jpwh_standard_neumann} shows the behavior of
-the standard MCSA algorithm applied to the JPWH matrix for a range
-of different polynomial orders and Monte Carlo histories per iteration.
+The effect of varying the polynomial order and the number of Monte Carlo
+histories per iteration is shown.
 As expected, increasing the polynomial order generally leads to a decrease
@@
 beyond that point there is little benefit to executing additional histories.
+Comparing the results for the forward and adjoint methods, the adjoint method
+generally requires fewer histories to achieve a stable result.  When both
+methods converge, however, the forward method generally achieves convergence
+in fewer MCSA iterations.  At low history counts, the adjoint method displays
+greater stability for even order polynomials than for add orders, the reason
+for this behavior is not clear.
 \begin{table}
@@
 \centering
-\begin{tabular}{ccccccc}
+\begin{tabularx}{0.8\textwidth}{c *{6}{Y}}
 \toprule
 Histories per & \multicolumn{6}{c}{Polynomial Order} \\
+\cmidrule(lr){2-7}
 Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
@@
 \bottomrule
-\end{tabular}
+\end{tabularx}
 \end{table}
-
 \begin{table}
@@
 \bottomrule
-\end{tabular}
+\end{tabularx}
 \end{table}
 
+Tables~\ref{tab:lap_forward_cheby} and \ref{tab:lap_adjoint_cheby}
+show the same problem using Chebyshev polynomials instead of the Neumann
+series polynomial.  Stability seems to be a much greater concern when using
+Chebyshev polynomials, as significantly more histories must be executed to
+achieve convergence as compared to the corresponding Neumann series results.
+One possible explanation for this behavior is that the Chebyshev coefficients
+increase exponentially in magnitude, making approximation of higher order terms
+far more difficult.
+When convergence is achieved, however, significantly smaller iteration counts
+are observed.  As with the Neumann series polynomial, greater stability is
+achieved when using the adjoint method.
 \begin{table}
@@
 4000 & - & - & - & - & - & - \\
-8000 & - & - & - & - & - & - \\
-16000 & 556(369) & - & - & - & - & - \\
+8000 & 556(198) & - & - & - & - & - \\
+16000 & 556(363) & - & - & - & - & - \\
 \bottomrule
-\end{tabular}
+\end{tabularx}
 \end{table}
-
 \begin{table}
@@
 1000 & - & - & - & - & - & - \\
-2000 & - & 325(130) & - & - & - & - \\
-4000 & - & 294(210) & - & - & - & - \\
-8000 & - & 288(380) & 201(335) & - & - & - \\
-16000 & - & 285(736) & 175(554) & - & - & - \\
+2000 & - & 328(134) & - & - & - & - \\
+4000 & - & 296(210) & - & - & - & - \\
+8000 & - & 288(380) & 262(423) & - & - & - \\
+16000 & 1132(1866) & 283(721) & 175(550) & - & - & - \\
 \bottomrule
-\end{tabular}
+\end{tabularx}
 \end{table}
 
+Tables~\ref{tab:lap_forward_gmres} and \ref{tab:lap_adjoint_gmres}
+again show the results of MCSA applied to the same problem,
+now using the GMRES polynomial (computing using the least squares
+QR solution described in Section~\ref{subsec:gmres_poly} with a random
+initial vector).  The GMRES polynomial seems to offer a favorable
+middle ground between the Neumann series and Chebyshev polynomials:
+iteration counts for a given polynomial order are smaller than with
+the Neumann series and yet far fewer histories are required to achieve
+stability than with the Chebyshev polynomial.
 \begin{table}
@@
 \bottomrule
-\end{tabular}
+\end{tabularx}
 \end{table}

### 2014-06-27 notebooks/stuart/mc_tools.py
adding batch monte carlo solver

@@ -99,6 +99,22 @@ def multilevelMonteCarloSolve( A, b, w_c, np ):
     return (x, sigma)
 
 ##---------------------------------------------------------------------------##
+## Solve a linear problem using Monte Carlo and return the result in batches.
+##---------------------------------------------------------------------------##
+def batchMonteCarloSolve( A, b, w_c, np ):
+    H, C, W = mc_data.makeMonteCarloHCW( A )
+    source_c, starting_weight = mc_data.makeSourceCDF( b )
+    w_f = w_c * starting_weight
+    x = []
+    tally = numpy.zeros( len(b) )
+    sigma = numpy.zeros( len(b) )
+    for i in xrange(np):
+        dummy = numpy.zeros( len(b) )
+        tally, sigma = doOneHistory( b, source_c, starting_weight, C, W, dummy, sigma, w_f )
+        x.append(tally)
+    return (x, sigma)
+
+##---------------------------------------------------------------------------##
 ## end mc_tools.py
 ##---------------------------------------------------------------------------##
 

### 2014-06-27 notebooks/stuart/array_tools.py
adding mcsa notebooks

@@ -99,6 +99,15 @@ def mmMultiply( A, B ):
     return C
 
 ##---------------------------------------------------------------------------##
+## Vector update x = x+y
+##---------------------------------------------------------------------------##
+def updateVector( x, y ):
+    size = len(x)
+    for i in xrange(size):
+        x[i] = x[i] + y[i]
+    return x
+
+##---------------------------------------------------------------------------##
 ## end array_tools.py
 ##---------------------------------------------------------------------------##
 

### 2014-06-27 notebooks/stuart/array_tools.py
updating batch tools to solve least squares problem with QR decomposition

@@ -108,6 +108,18 @@ def updateVector( x, y ):
     return x
 
 ##---------------------------------------------------------------------------##
+## Compute the transpose of a matrix.
+##---------------------------------------------------------------------------##
+def matrixTranspose(A):
+    num_rows = len(A)
+    num_cols = len(A[0])
+    A_T = numpy.zeros((num_cols,num_rows))
+    for i in xrange(num_rows):
+        for j in xrange(num_cols):
+            A_T[j][i] = A[i][j]
+    return A_T
+
+##---------------------------------------------------------------------------##
 ## end array_tools.py
 ##---------------------------------------------------------------------------##
 

### 2014-06-30 doc/mrb_mc/kmc_presentation/krylov_mc.tex
adding krylov monte carlo notebook

\documentclass{beamer}
\usetheme{Warsaw}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{longtable}
\usepackage{listings}
\usepackage{color}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} 
\usepackage{amsmath} 
\usepackage{tmadd,tmath}
\usepackage[mathcal]{euscript} 
\usepackage{color}
\usepackage{textcomp}
\usepackage{algorithm,algorithmic}
\definecolor{listinggray}{gray}{0.9}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{
  backgroundcolor=\color{lbcolor},
  tabsize=4,
  rulecolor=,
  language=c++,
  basicstyle=\scriptsize,
  upquote=true,
  aboveskip={1.5\baselineskip},
  columns=fixed,
  showstringspaces=false,
  extendedchars=true,
  breaklines=true,
  prebreak =
  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
  frame=single,
  showtabs=false,
  showspaces=false,
  showstringspaces=false,
  identifierstyle=\ttfamily,
  keywordstyle=\color[rgb]{0,0,1},
  commentstyle=\color[rgb]{0.133,0.545,0.133},
  stringstyle=\color[rgb]{0.627,0.126,0.941},
}

%% colors
\setbeamercolor{boxheadcolor}{fg=white,bg=black}
\setbeamercolor{boxbodycolor}{fg=black,bg=white}

%% slide numbers

%%---------------------------------------------------------------------------%%
\author[Stuart Slattery]{Stuart Slattery\\
  \bigskip
  \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}} \\
  \bigskip
  Oak Ridge National Laboratory}

\date{\today} 
\title[Krylov Monte Carlo \hspace{1mm}
  \insertframenumber/\inserttotalframenumber]{Stochastic Krylov
  Subspace Approximation}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\begin{frame}{Projection Methods}

  \medskip \medskip
  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Search Subspace $\mathcal{K}$} Extract the solution to
    $\ve{A}\ve{x}=\ve{b}$ from the search subspace:
    \[
    \tilde{\ve{x}} = \ve{x}_0 +
    \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}
    \]
  \end{beamerboxesrounded}

  \medskip \medskip
  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Constraint Subspace $\mathcal{L}$} Constrain the extraction with
    the constraint subspace by asserting orthogonality with the
    residual ($\ve{r} = \ve{b}-\ve{A}\ve{x}$):
    \[
    \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
    \mathcal{L}
    \]  
  \end{beamerboxesrounded}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{The Orthogonality Constraint}

  \[
  \tilde{\ve{r}} = \ve{r}_0 - \ve{A}\boldsymbol{\delta}
  \]

  \begin{figure}[htpb!]
    \begin{center}
      \scalebox{1.1}{ \input{orthogonal_residual.pdftex_t} }
    \end{center}
    \caption{\textbf{Orthogonality constraint of the new residual with
        respect to $\mathcal{L}$.}}
  \end{figure}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Minimization Property}

    The residual of the system is \textit{minimized} with respect to
    the constraints
    \[
    ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
    \mathbb{R}^N
    \]
  \end{beamerboxesrounded}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Polynomials}

  The Krylov subspace looks a lot like our Monte Carlo process which
  estimates successive matrix-vector products:
  \[
  \ve{x} = \sum_k \ve{H}^k\ve{b} = \ve{b} + \ve{H}{b} + \ve{H}^2{b} +
  \dots + \ve{H}^k{b}
  \]
  Better yet, we are just constructing a general matrix polynomial:
  \[
  \ve{x} = \sum_k \alpha_k \ve{v}_k
  \]
  with $\ve{v}_k = \ve{H}^k\ve{b}$ and all $\alpha_k=1$. Each step of
  the random walk estimates a component of the polynomial basis:
  \[
  \ve{b} \rightarrow \ve{H}{b} \rightarrow \ve{H}^2{b} \rightarrow
  \dots \rightarrow \ve{H}^k{b}
  \]

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Stochastic Krylov Subspace Approximation}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Krylov Subspace}

    \[
    \mathcal{K}_m(\ve{A},\ve{r}_0) = span\{\ve{r}_0, \ve{A}\ve{r}_0,
    \ve{A}^2\ve{r}_0, \dots, \ve{A}^{m-1}\ve{r}_0\}
    \]
    For GMRES: $\mathcal{L} = \ve{A} \mathcal{K}_m(\ve{A},\ve{r}_0)$
  \end{beamerboxesrounded}

  We typically extract a correction from this subspace:
  \[
  \boldsymbol{\delta} = \sum_k \alpha_k^m \ve{A}^k \ve{r}_0
  \]
  which is just a polynomial in $\ve{A}$. Use the Monte Carlo process
  to stochastically construct $\mathcal{K}_m$. Apply linearly
  constrained least squares to extract $\boldsymbol{\delta}$ via
  orthogonality:
  \[
  minimize\ ||\ve{b} - \ve{A}(\ve{x_0}+\boldsymbol{\delta})||_2
  \]
  such that:
  \[
  \sum_{b=1}^B \alpha_b = 1
  \]

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}[fragile]{The Algorithm}

  \begin{algorithm}[H]
    \caption{Krylov Method}
    \begin{algorithmic}[1]
      \WHILE{$||(\ve{r})||/||(\ve{b})|| > \epsilon$}
      \STATE $\ve{r} = \ve{b} - \ve{A}\ve{x}$
      \STATE $\ve{V}$ = monteCarloPolynomial($\ve{A}$,$\ve{r}$,$N_p$,$m$)
      \STATE $\boldsymbol{\alpha}$ = constrainedLeastSquaresSolve($\ve{A}$,$\ve{x}$,$\ve{b}$,$\ve{V}$)
      \STATE $\boldsymbol{\delta} = \sum_k \alpha_k^m \ve{v}_k$
      \STATE $\ve{x} = \ve{x} + \boldsymbol{\delta}$
      \ENDWHILE
    \end{algorithmic}
  \end{algorithm}

  \begin{itemize}
  \item Expect most of the work to happen in polynomial construction
    in order to get good statistics
  \item Finding $\boldsymbol{\alpha}$ should be cheap except for dot
    products
  \item Parallel Monte Carlo strategies including threading and domain
    decomposition apply to polynomial construction
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Deterministic Results}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Results}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Summary}

  \begin{itemize}
  \item Stochastic computation of Krylov subspace - replace sequence
    of matrix-vector multiplies with a random walk of fixed length
    ($\mathcal{K}$ is defined stochastically)
  \item Petrov-Galerkin condition applied (correction extraction from
    $\mathcal{L}$ is defined deterministically
  \item Works for $\rho(\ve{H}) > 1$!
  \item Left/right preconditioning is now tractable given row-wise
    access to preconditioning matrix entries
  \item If Krylov subspace is computed exactly there are some nice
    optimality properties
  \item The bigger the subspace the better the convergence behavior
    (to a point)
  \item Right now subspace approximation is slow to converge with only
    diagonal preconditioning - more histories gives better convergence
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%

\end{document}

### 2014-06-30 notebooks/stuart/batch_tools.py

@@ -72,11 +72,12 @@ def computeMRB( A, x_sample, f, num_batch ):
     return x_bmr
 
 ##---------------------------------------------------------------------------##
-## Given a set of basis vectors, find their linear combination that
-## minimizes the linear problem residual by solving an linearly
-## constrained linear least squares problem.
+## Given a set of basis vectors from which to extract a correction,
+## find their linear combination that minimizes the linear problem
+## residual by solving an linearly constrained linear least squares
+## problem.
 ## ---------------------------------------------------------------------------##
-def computeConstrainedLeastSquares( A, basis, f ):
+def computeConstrainedLeastSquares( A, basis, f, x_0 ):
 
     num_basis = len(basis)
     grid_size = len(basis[0])
@@ -106,8 +107,9 @@ def computeConstrainedLeastSquares( A, basis, f ):
 
     # Make the least-squares problem RHS
     b = numpy.zeros(grid_size)
+    Ax_0 = numpy.dot(A,x_0)
     for i in xrange(grid_size):
-        b[i] = f[i] - V[i][num_basis-1]
+        b[i] = f[i] - V[i][num_basis-1] - Ax_0[i]
     ZTb = numpy.dot(Z_T,b)
 
     # Solve the least-squares problem with QR decomposition and back
@@ -133,7 +135,7 @@ def computeConstrainedLeastSquares( A, basis, f ):
 ## minimizes the linear problem residual by solving an unconstrained
 ## linear least squares problem.
 ## ---------------------------------------------------------------------------##
-def computeLeastSquares( A, basis, f ):
+def computeLeastSquares( A, basis, f, x_0 ):
 
     num_basis = len(basis)
     grid_size = len(basis[0])
@@ -152,8 +154,9 @@ def computeLeastSquares( A, basis, f ):
 
     # Make the least-squares problem RHS
     b = numpy.zeros(grid_size)
+    Ax_0 = numpy.dot(A,x_0)
     for i in xrange(grid_size):
-        b[i] = f[i] - V[i][num_basis-1]
+        b[i] = f[i] - V[i][num_basis-1] - Ax_0[i]
     VTb = numpy.dot(V_T,b)
 
     # Solve the least-squares problem with QR decomposition and back

### 2014-06-30 notebooks/stuart/mc_data.py

@@ -5,7 +5,7 @@
 import numpy
 
 ##---------------------------------------------------------------------------##
-## Make an iteration matrix H = I-A for a given operator
+## Make an iteration matrix H = I-A for a given operator for the adjoint
 ##---------------------------------------------------------------------------##
 def makeIterationMatrix( A ):
     grid_size = len(A)
@@ -48,8 +48,8 @@ def makeCDFMatrix( P ):
     return C
 
 ##---------------------------------------------------------------------------##
-## Make a weight matrix from an iteration matrix and a probability
-## matrix.
+## Make an adjoint weight matrix from an iteration matrix and an
+## adjoint probability matrix.
 ## ---------------------------------------------------------------------------##
 def makeWeightMatrix( H, P ):
     grid_size = len(H)

### 2014-07-11 MAX_PROJECT/current_code/utils/model_problems/model_problems.tex
code reorganized

\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}

\title{Test Matrices for Monte Carlo Synthetic Acceleration}

\begin{document}
\maketitle

\section{Introduction}

This document is intended to define a standard suite of linear
systems that can be used as test cases for Monte Carlo and
MCSA solvers.  The suite of matrices is intended to span a 
range of different matrix properties and problem difficulty.
This is intended to avoid potential pitfalls when performing
numerical experiments involving drawing conclusions based on
observations that may be specific to a particular linear system.

\section{Shifted 1-D Laplacian}

The most basic test matrix we propose is essentially a shifted
1-D discrete (negative) Laplacian operator.  In order to achieve a matrix
that is strongly diagonally dominant, we introduce a shift of
$2.0$ added to the main diagonal, resulting in the matrix
\begin{equation}
 \mathbf{A} = \begin{bmatrix} 4 & -1 \\ -1 & \ddots & \ddots \\ 
     & \ddots & \ddots & -1 \\  & & -1 & 4 \end{bmatrix} \:,
\end{equation}
which can be generated in MATLAB with the command
\begin{verbatim}
  A = 4*eye(N) - diag(ones(N-1),1) - diag(ones(N-1),-1);
\end{verbatim}
where $N$ is the size of the matrix.
The right hand side is taken to be
\begin{equation}
 \mathbf{b}_i = i, \quad i=0,\ldots,N-1 \:.
\end{equation}
A matrix of any size can be generated in this way and will have
similar numerical properties, but the sample matrices
provided here take $N=50$.
The spectral radius of $\mathbf{I - D^{-1}A}$ is equal to approximately
$0.499$.

\section{2-D Laplacian}

The second test problem is a finite difference discretization of the 2-D
Laplacian with homogeneous Dirichlet boundaries.  This matrix can be
generated in MATLAB by the command
\begin{verbatim}
  A = delsq(numgrid('S',M));
\end{verbatim}
where $M$ is the number of grid points in each direction.
As a right hand size, we take a sinusoid distribution in both the
$x$ and $y$ directions, i.e.
\begin{equation}
 \mathbf{b}(x_i,y_j) = \sin \left( \frac{\pi x_i}{M-1} \right) \;
                       \sin \left( \frac{\pi y_j}{M-1} \right), \quad i=0,\ldots,M-1 \:,
\end{equation}
where $x_i=\frac{i}{M}$ and $y_j=\frac{j}{M}$.  As with the previous example,
any size problem can be selected; however, for this matrix the problem becomes
more difficult as $M$ is increased.  We choose a value of $M=32$, for which
the size of the resulting matrix is 900 (the unknowns on Dirichlet boundaries
are eliminated from the matrix) and spectral radius of the iteration
matrix for this problem is approximately $0.9949$.

\section{IFISS Convection-Diffusion}

The next test problem is a finite element discretization of a 
2-D convection-diffusion problem using the MATLAB IFISS package.
Continuous bilinear elements on quadrilaterals (Q1 elements) are used.
A $32 \times 32$ mesh is used on the unit square with a viscosity of
$\nu=0.1$.  Streamline upwind Petrov--Galerkin (SUPG) stabilization
is used in the construction of the matrix.  Homogeneous Dirichlet
boundary conditions are applied on the top, bottom, and left sides of
the domain and a nonhomogeneous Dirichlet boundary with a constant value
of 1 is applied on the right side.  A constant forcing term with a value
of 1 is present throughout the domain.  The dimension of the resulting
matrix is 1089 and the spectral radius of $\mathbf{I - D^{-1}A}$
is 0.9835.

\section{JPWH Matrix}

The next test problem is the matrix ``JPWH\_991'' matrix available from
the Matrix Market.  This matrix
is described as a ``computer random simulation of a circuit physics model.''
We select the right hand side to be the constant vector of all ones.  The
matrix has dimension 991 and the
spectral radius of $\mathbf{I - D^{-1}A}$ is 0.9797.

\section{FS 680 1}

As another matrix market example we choose the matrix ``FS\_680\_1,''
which is described as a mixed kinetics diffusion problem from radiation chemistry.
The dimension of the problem is 680 and the spectral radius of the iteration matrix
is $0.9697$.  The primary reason for including this matrix is that approximately
half of the columns of the matrix are columns of the identity matrix.  This
results in the iteration matrix having a large number of zero columns and
therefore the transpose of the iteration matrix having a large number of 
zero rows.  This poses no problem for forward Monte Carlo calculations but
seems to result in very poor convergence behavior for the adjoint Monte Carlo
approach.  This is in contrast to typical behavior where the performance of
the adjoint Monte Carlo is typically significantly better than the forward
method.

\section{Marshak Wave}

The final test problem (so far) that we propose is from a Marshak wave problem
in thermal radiation diffusion.  The problem selected is a single time step
from a time dependent calculation.  An initial guess is provided which is the
solution at the previous time step.  The interesting feature for this problem
is that the solution only changes in the immediate vicinity of the wave front
and therefore the residual is very large within a small portion of the problem
domain and nearly zero everywhere else.  This presents a significant advantage
for the adjoint Monte Carlo method, which preferentially starts Monte Carlo
histories in locations where the residual is large.

\end{document}

### 2014-07-11 MAX_PROJECT/current_code/core/prob_adjoint2.m

@@ -1,0 +1,47 @@
+function [Pb, cdfb, P, cdf]=prob_adjoint2(A, b, p1, p2)
+
+    display('Building of transition matrix');
+
+
+    %initial probability distribution
+    Pb=zeros(size(A,1),1);
+    for i=1:size(A,1)
+        Pb(i)=(abs(b(i)).^p1)/(sum(abs(b)).^p1);
+    end
+
+
+    P=zeros(size(A));
+    for i=1:size(P,1)
+        for j=1:size(P,2)
+            if sum(abs(A(:,i)))>0
+                P(i,j)=abs(A(j,i)).^p2/(sum(abs(A(:,i)).^p2));
+            end
+        end
+    end
+
+    
+    
+    %computation of the cumulative initial probability: 
+    cdfb=Pb;
+    for i=2:size(cdfb)
+        aux=max(find(cdfb(1:i-1)))+0;
+        if(cdfb(i)~=0 && ~isempty(aux))
+            cdfb(i)=cdfb(i)+cdfb(aux);
+        end
+    end
+    
+    cdf=P;
+    %computation of the cumulative probability
+    for i=1:size(cdf,1)
+        for j=2:size(cdf,2)
+            aux=max(find(cdf(i,1:j-1)))+0;
+            if (cdf(i,j)~=0 && ~isempty(aux))
+                cdf(i,j)=cdf(i,j)+cdf(i,aux);
+            end
+        end
+    end
+    
+    P=sparse(P);
+    cdf=sparse(cdf);
+    
+end

### 2014-07-11 MAX_PROJECT/current_code/tests/MCSA_adjoint_test.m

@@ -1,32 +1,54 @@
 addpath('../core')
 addpath('../utils')
 
-dimen=50;
-A=4*diag(ones(dimen,1)) - diag(ones(dimen-1,1),1) - diag(ones(dimen-1,1),-1);
-rhs=[1:50]';
-u=A\rhs;
+% 'jpwh_991'; 'fs_680_1'; 'ifiss_convdiff'; 'shifted_laplacian_1d'; 'thermal_eq_diff'
+matrix='thermal_eq_diff';
+
+addpath(strcat('../utils/model_problems/', matrix))
+
+if strcmp(matrix, 'simple')
+    dimen=50;
+    A=4*diag(ones(dimen,1)) - diag(ones(dimen-1,1),1) - diag(ones(dimen-1,1),-1);
+    rhs=ones(dimen,1);
+    u=A\rhs;
+    
+else
+     [A, dimen, ~, ~] = mmread('A.mtx');
+     rhs=mmread('b.mtx');
+     u=mmread('x.mtx');
+end
 
 Prec=diag(diag(A));
 
 H=eye(size(A))-Prec\A;
 rhs=Prec\rhs;
 
-dist='MAO';
+%% Numerical setting
 
+numer.eps=10^(-3);
+numer.rich_it=300;
+
+%% Statistical setting
+stat.nwalks=600;
+stat.max_step=20;
+stat.adapt=1;
+stat.varcut=0.1;
+dist=1;
+
+%% Definition of initial and transitional probabilities
 [Pb, cdfb, P, cdf]=prob_adjoint(H, rhs, dist);
 
-eps=10^(-3);
-n_walks=1000;
-max_step=20;
-rich_it=10;
+%% Preconditioning setting
 
-fp.u=u;
-fp.H=H;
+fp.u=u; %reference solution
+fp.H=H; %iteration matrix
 fp.rhs=rhs;
 fp.precond='diag';
 %% Monte Carlo Adjoint Method resolution
 
-[sol, rel_error, var, NWALKS, iterations]=MCSA_adjoint(fp, dist, P, cdf, rich_it, n_walks, max_step, eps);
+start=cputime;
+[sol, rel_residual, var, VAR, DX, NWALKS, iterations]=MCSA_adjoint(fp, dist, P, cdf, numer, stat);
+finish=cputime;
 
 conf=0.05;
 
@@ -36,4 +58,7 @@ plot(sol-var*norminv(1-conf/2, 0, 1), 'g*');
[…]

### 2014-07-11 MAX_PROJECT/current_code/tests/MCSA_forward_test.m

@@ -1,32 +1,55 @@
 addpath('../core')
 addpath('../utils')
 
-dimen=50;
-A=4*diag(ones(dimen,1)) - diag(ones(dimen-1,1),1) - diag(ones(dimen-1,1),-1);
-rhs=[1:50]';
-u=A\rhs;
+% 'jpwh_991'; 'fs_680_1', 'ifiss_convdiff'; 'shifted_laplacian_1d'; 'thermal_eq_diff'
+matrix='jpwh_991';
 
+addpath(strcat('../utils/model_problems/', matrix))
+
+if strcmp(matrix, 'simple')
+    dimen=50;
+    A=4*diag(ones(dimen,1)) - diag(ones(dimen-1,1),1) - diag(ones(dimen-1,1),-1);
+    rhs=ones(dimen,1);
+    u=A\rhs;
+    
+else
+     [A, dimen, ~, ~] = mmread('A.mtx');
+     rhs=mmread('b.mtx');
+     u=mmread('x.mtx');
+end
+    
 Prec=diag(diag(A));
 
 H=eye(size(A))-Prec\A;
 rhs=Prec\rhs;
 
-eps=10^(-3);
-n_walks=100;
-max_step=10;
-rich_it=10;
+%% NUmerical setting
 
-dist='MAO';
+numer.eps=10^(-3);
+numer.rich_it=300;
 
+%% Statistical setting
+
+stat.nwalks=2;
+stat.max_step=20;
+stat.adapt=1;
+stat.varcut=0.5;
+dist=1;
+
+%% Definition of the transitional probability
 [P, cdf]=prob_forward(H, dist);
 
-fp.u=u;
-fp.H=H;
+%% Preconditioning setting
+
+fp.u=u; %reference solution
+fp.H=H; %iteration matrix
 fp.rhs=rhs;
 fp.precond='diag';
 %% MCSA forward method resolution
 
-[sol, rel_err, var, NWALKS, iterations]=MCSA_forward(fp, P, cdf, rich_it, n_walks, max_step, eps);
+start=cputime;
+[sol, rel_residual, var, VAR, DX, NWALKS, iterations]=MCSA_forward(fp, P, cdf, numer, stat);
+finish=cputime;
 
 conf=0.05;
 
@@ -44,4 +67,7 @@ plot(sol-var*norminv(1-conf/2, 0, 1), 'g*');
 plot(sol+var*norminv(1-conf/2, 0, 1), 'g*');
 plot(u, 'r*')
 
[…]

### 2014-07-11 MAX_PROJECT/current_code/tests/MC_adjoint_plain.m

@@ -1,9 +1,9 @@
 addpath('../core')
 addpath('../utils')
 
-dimen=500;
+dimen=50;
 A=4*diag(ones(dimen,1)) - diag(ones(dimen-1,1),1) - diag(ones(dimen-1,1),-1);
-rhs=[1:500]';
+rhs=[1:50]';
 u=A\rhs;
 
 Prec=diag(diag(A));
@@ -12,20 +12,22 @@ H=eye(size(A))-Prec\A;
 rhs=Prec\rhs;
 
 eps=10^(-3);
-dist='MAO';
+dist1=1;
+dist2=1;
 
-[Pb, cdfb, P, cdf]=prob_adjoint(H, rhs, dist);
+[Pb, cdfb, P, cdf]=prob_adjoint2(H, rhs, dist1, dist2);
 %%
 rel_errs=[];
 NWALKS=[];
 VAR=[];
 SOL=[];
 
-for i=1:6
+for i=1:2
+    X=[];
     n_walks=10^i;
     NWALKS=[NWALKS n_walks];
     max_step=100;
-    [u_approx, var]=MC_adjoint(H,rhs,P,cdf,Pb,cdfb, n_walks, max_step);
+    [u_approx, var, X, tally]=MC_adjoint(X,H,rhs,P,cdf,Pb,cdfb, n_walks, max_step);
     rel_error=sqrt(sum((u-u_approx).^2))/sqrt(sum((u).^2));
     rel_errs=[rel_errs, rel_error];
     SOL=[SOL u_approx];
@@ -60,5 +62,5 @@ plot(u_approx+var*norminv(1-conf/2, 0,1), 'g*')
 plot(u, 'r*')
 
 
-%save(strcat('../results/MC_adjoint_plain/MC_adjoint_plain_', dist));
-   save(strcat('MC_adjoint_plain_', dist)); 
+save(strcat('../results/MC_adjoint_plain/MC_adjoint_plain_p1=', num2str(dist1), '_p2=', num2str(dist2)));
+%   save(strcat('MC_adjoint_plain_', dist)); 

### 2014-07-11 MAX_PROJECT/current_code/tests/SEQ_forward_test.m

@@ -1,31 +1,53 @@
 addpath('../core')
 addpath('../utils')
 
-dimen=50;
-A=4*diag(ones(dimen,1)) - diag(ones(dimen-1,1),1) - diag(ones(dimen-1,1),-1);
-rhs=[1:50]';
-u=A\rhs;
+% 'jpwh_991'; 'fs_680_1', 'ifiss_convdiff'; 'shifted_laplacian_1d'; 'thermal_eq_diff'
+matrix='fs_680_1';
+
+addpath(strcat('../utils/model_problems/', matrix))
+
+if strcmp(matrix, 'simple')
+    dimen=50;
+    A=4*diag(ones(dimen,1)) - diag(ones(dimen-1,1),1) - diag(ones(dimen-1,1),-1);
+    rhs=ones(dimen,1);
+    u=A\rhs;
+    
+else
+     [A, dimen, ~, ~] = mmread('A.mtx');
+     rhs=mmread('b.mtx');
+     u=mmread('x.mtx');
+end
 
 Prec=diag(diag(A));
 
 H=eye(size(A))-Prec\A;
 rhs=Prec\rhs;
 
-eps=10^(-3);
-n_walks=1000;
-max_step=20;
-rich_it=10;
-dist='MAO';
+%% Numerical setting
 
+numer.eps=10^(-3);
+numer.rich_it=300;
+
+%% Statistical setting
+
+stat.nwalks=2;
+stat.max_step=20;
+stat.adapt=1;
+stat.varcut=0.5;
+dist=1;
+
+%% Definition of the transitional probability
 [P, cdf]=prob_forward(H, dist);
 
-fp.u=u;
-fp.H=H;
+%% Preconditioning setting
+
+fp.u=u; %reference solution
+fp.H=H; %iteration matrix
 fp.rhs=rhs;
 fp.precond='diag';
 %% SEQ forward method resolution
 
-[sol, rel_err, var, NWALKS, iterations]=SEQ_forward(fp, P, cdf, rich_it, n_walks, max_step, eps);
+[sol, rel_residual, var, VAR, DX, NWALKS, iterations]=SEQ_forward(fp, P, cdf, numer, stat);
 
 conf=0.05;
 var=var*norminv(1-conf/2, 0, 1);
@@ -43,4 +65,4 @@ plot(sol+var, 'g*');
 plot(sol-var, 'g*');
 plot(u, 'r*');
 
[…]

### 2014-07-11 MAX_PROJECT/current_code/tests/laplace_adjoint.m

@@ -19,21 +19,27 @@ precond='diag';
 
 %Sequential Monte Carlo or Monte Carlo Synthetic Acceleration
 %possible choices: 'SEQ', 'MCSA'
-scheme='SEQ';
+scheme='MCSA';
 method='adjoint';
 
 [fixed_point]=iteration_matrix(precond, D, u, rhs, G);
 
 spy_matrices(fixed_point);
 
-%% Building of the iteration matrix
+%% Numerical setting
 
-eps=10^(-3);
-n_walks=(10^1)*(n-2)^2;
-max_step=20;
-rich_it=200;%maximal number of Richardson iteration
-dist='MAO';
-p=2;
+numer.eps=10^(-3);
+numer.rich_it=300;%maximal number of Richardson iterations
+
+%% Statistical setting
+
+stat.nwalks=size(u,1);
+stat.max_step=20;
+stat.varcut=0.1;
+stat.adapt=1;
+dist=1;
+
+%% Definition of initial and transitional prbabilities
 
 if ~ strcmp(precond, 'alternating')
     [Pb, cdfb, P, cdf]=prob_adjoint(fixed_point.H, fixed_point.rhs, dist);
@@ -43,11 +49,15 @@ else
     [Pb.Pb2, cdfb.cdfb2, P.P2, cdf.cdf2]=prob_adjoint(fixed_point.H2, fixed_point.rhs2, dist);
 end
 
-[sol, rel_err, var, NWALKS, iterations, time]=system_solver(scheme, method, fixed_point, dist, P, cdf, rich_it, n_walks, max_step, eps);
+%% Solver call
+
+[sol, rel_residual, var, VAR, DX, NWALKS, iterations, time]=system_solver(scheme, method, fixed_point, dist, P, cdf, numer, stat);
 
 %delete(poolobj)
 
-conf=0.025;
+%% Post-processing rendering
+
+conf=0.025;% error of type 1
 
 plot_sol(G, u, 'ex');
 plot_sol(G, sol, scheme);
@@ -72,6 +82,9 @@ for i=1:size(u,1)
     amplitude=amplitude+2*var(i)*norminv(1-conf/2, 0, 1);
 end
 amplitude=amplitude/norm(u,2);
+
+hold off
+bar(NWALKS)
 
 if reac==0
     if strcmp(scheme, 'SEQ')

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/batchfiles/T-NS3_batch.m

@@ -1,0 +1,13 @@
+3     % problem
+3     % cavity type
+6     % grid parameter
+2     % stretched grid
+3     % discretisation
+0.002 % viscosity parameter
+100   % target time
+5e-5  % time accuracy tolerance
+10    % time averaging frequency
+1     % 0 solution plot switch
+13    % timestep evolution figure number
+
+%% Data file for test problem T-NS3 

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/boussinesq_flow/stabtrBouss.m

@@ -1,0 +1,315 @@
+function [DT,U,Udot,T,Tdot,time] = stabtrBouss(nonlin,qmethod,grid,spmat,...
+                                       initdata,tfinal,tol,nstar,nmax,tout)
+%STABTRBOUSS Boussinesq integrator based on stabilized TR (robust version)
+%   [DT,U,Udot,T,Tdot,time] = ...
+%   stabtrBouss(nonlin,qmethod,grid,spmat,initdata,tfinal,tol,nstar,2,1);
+%   input
+%          nonlin    number of nonlinear corrections (0 for linearized TR)
+%          qmethod   approximation method
+%          grid      grid data structure (see unpack_boussdata)
+%          spmat     Galerkin matrix data structure (see unpack_boussdata)
+%          initdata  initial/restart data structure 
+%          tfinal    final time 
+%          tol       local accuracy error tolerance   
+%          nstar     averaging frequency  
+%          nmax      number of itegration steps (*nnt defined below) 
+%          tout      output data switch
+%   output
+%          DT        timestep history
+%          U         velocity solution history
+%          Udot      solution time derivative history
+%          T         temperature solution history
+%          Tdot      Temperature time derivative history
+%          time      discrete time evolution
+%
+% calls functions:  dtboussbc, unsteadyflowbc, resetflowbc, xxheatbc, resetheatbc
+% calls scriptfiles: unpack_boussdata, bousshistory
+% hot restart datafile: stabtrBouss_restart
+%   IFISS function: DJS; 10 May 2012; 24 June 2012; 30 September 2013
+% Copyright (c) 2012 D.J. Silvester, M.D. Mihajlovic.
+global Pr Ra H L 
+viscosity =sqrt(Pr/Ra);  viscosityT =1/sqrt(Pr*Ra);
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/convection/ref_cd.m

@@ -1,0 +1,19 @@
+%REF_CD set up problem in reference square domain 
+%   IFISS scriptfile: DJS; 27 May 2012. 
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+
+%% define geometry
+pde=2; domain=1;
+ref_domain(sn)
+load ref_grid.mat
+%
+%% set up matrices
+qmethod=1;
+[ev,ebound] = q1grid(xy,mv,bound,mbound);
+[A,N,Q,epe,eph,epw] = femq1_cd(xy,ev);
+%
+%% save resulting system
+fprintf('system matrices saved in square_cd_nobc.mat ...\n')
+gohome
+cd datafiles
+save square_cd_nobc.mat qmethod grid_type A N Q xy ev epe eph epw ebound bound outbc x y

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/convection/square_cd.m

@@ -1,0 +1,19 @@
+%SQUARE_CD set up problem in unit square domain 
+%   IFISS scriptfile: DJS; 26 May 2012
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+clear variables
+%% define geometry
+pde=2; domain=1;
+square_domain
+load square_grid.mat
+%
+%% set up matrices
+qmethod=1;
+[ev,ebound] = q1grid(xy,mv,bound,mbound);
+[A,N,Q,epe,eph,epw] = femq1_cd(xy,ev);
+%
+%% save resulting system
+fprintf('system matrices saved in square_cd_nobc.mat ...\n')
+gohome
+cd datafiles
+save square_cd_nobc.mat qmethod grid_type A N Q xy ev epe eph epw ebound bound outbc x y 

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/convection/test_problems/cd_testproblem.m

@@ -1,0 +1,32 @@
+%cd_testproblem    sets up Reference Examples 3.1 to 3.4 
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference convection-diffusion problem.\n')
+fprintf('\nchoose specific example');
+fprintf('\n     1  Constant vertical wind')
+fprintf('\n     2  Vertical wind, characteristic layers')
+fprintf('\n     3  Constant wind @ 30 degree angle')
+fprintf('\n     4  Recirculating wind\n');
+sn = default('',1);
+
+if sn==3 
+system('/bin/cp ./convection/test_problems/constant_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/test_bc.m ./diffusion/specific_bc.m');
+   square_cd, solve_cd
+elseif sn==4
+system('/bin/cp ./convection/test_problems/circular_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/hotwall_bc.m ./diffusion/specific_bc.m');
+   square_cd, solve_cd
+elseif sn==1
+system('/bin/cp ./convection/test_problems/vertical_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/solution1_bc.m ./diffusion/specific_bc.m');
+   ref_cd, solve_cd
+elseif sn==2
+system('/bin/cp ./convection/test_problems/ref_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/ref_bc.m ./diffusion/specific_bc.m');
+   ref_cd, solve_cd
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/diffusion/test_problems/diff_testproblem.m

@@ -1,0 +1,32 @@
+%diff_testproblem    sets up Reference Examples 1.1 to 1.4
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference Poisson problem.')
+fprintf('\nchoose specific example');
+fprintf('\n     1  Square domain, constant source')
+fprintf('\n     2  L-shaped domain, constant source')
+fprintf('\n     3  Square domain, analytic solution')
+fprintf('\n     4  L-shaped domain, analytic solution\n');
+sn = default('',1);
+%
+if sn==1
+system('/bin/cp ./diffusion/test_problems/unit_rhs.m ./diffusion/specific_rhs.m');
+system('/bin/cp ./diffusion/test_problems/zero_bc.m ./diffusion/specific_bc.m');
+   square_diff
+elseif sn==2
+system('/bin/cp ./diffusion/test_problems/unit_rhs.m ./diffusion/specific_rhs.m');
+system('/bin/cp ./diffusion/test_problems/zero_bc.m ./diffusion/specific_bc.m');
+   ell_diff
+elseif sn==3
+system('/bin/cp ./diffusion/test_problems/zero_rhs.m ./diffusion/specific_rhs.m');
+system('/bin/cp ./diffusion/test_problems/solution3_bc.m ./diffusion/specific_bc.m');
+   square_diff
+elseif sn==4
+system('/bin/cp ./diffusion/test_problems/zero_rhs.m ./diffusion/specific_rhs.m');
+system('/bin/cp ./diffusion/test_problems/solution4_bc.m ./diffusion/specific_bc.m');
+   ell_diff
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/gohome.m

@@ -1,0 +1,7 @@
+%GOHOME positions command prompt at top level directory
+%   IFISS scriptfile: DJS; 25 Septmber 2013.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage (see readme.m)
+cd('/home/max/MAX_PROJECT/current_code/utils/ifiss3.3')
+%cd('/home/caas63/ifiss3.3');
+%cd('/fs/helios/elman/ifiss/ifiss3.3');
+%cd('c:\Documents and Settings\elman\My Documents\MATLAB\ifiss3.3');

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/graphs/Contents.m

@@ -1,0 +1,65 @@
+% GRAPHS
+%
+% Files
+%   bookplot                  - saves IFISS figure as postscript file
+%   box_heatplot              - evolves temperature data on cavity domain
+%   dttplot                   - plots timestep data 
+%   ell_heatplot              - evolves temperature data on L-shaped domain
+%   ellx                      - outlines L-shaped domain
+%   eplot                     - plots Q1 element data on square-shaped domain
+%   eplotl                    - plots Q1 element data on L-shaped domain
+%   errplot                   - plots solution and error on square domain
+%   errplotl                  - plots solution and error on L-shaped domain
+%   exportfig                 - Export a figure to Encapsulated Postscript.
+%   flowplot                  - IFISS2.2 version
+%   flowplot09                - plots flow data on general domain
+%   flowplotl                 - plots flow data on standard step domain
+%   flowplotp                 - plots flow data on slit shaped domain
+%   flowplotz                 - plots flow data on extended step domain
+%   flowvolume                - plots flow solution on vertical X-section 
+%   htmlplot                  - saves IFISS figure as html file
+%   logoplot                  - plots IFISS logo
+%   mplot                     - plots 2x2 macro data on square domain
+%   mplotl                    - plots 2x2 macro data on L-shaped domain
+%   obstacle_flowmovie        - movie of flow around square obstacle
+%   obstacle_unsteadyflowplot - plots flow data in obstructed channel 
+%   obstacle_vorticityplot    - plots vorticity data in obstructed channel
+%   outflow                   - plots/explores tangential flow at outflow
+%   solplot                   - plots nodal data on square-shaped domain
+%   solplotl                  - plots nodal data on L-shaped domain
+%   solsurf                   - plots solution surface on square domain
+%   solsurfl                  - plots solution surface on L-shaped domain
+%   square_flowmovie          - generates flow movie on square  domain
+%   square_heatplot           - evolves temperature data on square domain
+%   square_pressureplot       - compares pressure data on square domain
+%   square_unsteadyflowplot   - plots flow data on square domain
+%   square_vorticityplot      - plots vorticity data on square domain
+%   squarex                   - outlines square shaped domain
+%   step_flowmovie            - flow movie on extended step domain
+%   step_unsteadyflowplot     - evolves flow data on extended step domain
+%   step_vorticityplot        - plots vorticity data on step domain
+%   stepx                     - outlines step domain
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/graphs/eplot.m

@@ -1,0 +1,38 @@
+function eplot(eldata,ev,xy,x,y,fig,titlestr)
+%EPLOT plots Q1 element data on square-shaped domain
+%   eplot(eldata,ev,xy,x,y,fig,'Element data');
+%   input
+%          eldata     element data vector
+%          ev         element mapping matrix
+%          xy         nodal coordinate vector  
+%          x          vector of x-axis interpolation points
+%          y          vector of y-axis interpolation points
+%          fig        figure number
+%          titlestr   character string
+%
+%   IFISS function: DJS; 29 September 2013.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+fprintf('plotting element data... ')
+figure(fig)
+xx=xy(:,1); yy=xy(:,2);
+nel=length(eldata);
+xc = zeros(nel,2);
+% loop over elements    
+for ielem = 1:nel
+   xl = xx(ev(ielem,:));
+   yl = yy(ev(ielem,:)); 
+   xc(ielem,1) = 0.25*sum(xl);
+   xc(ielem,2) = 0.25*sum(yl);
+end
+%
+% interpolate to a cartesian product mesh
+x=0.5*(x(1:end-1)+x(2:end));
+y=0.5*(y(1:end-1)+y(2:end));
+[X,Y]=meshgrid(x,y);
+xysol = griddata(xc(:,1),xc(:,2),eldata,X,Y);
+subplot(121),contour(X,Y,xysol,15),axis('square')
+title(titlestr,'FontSize',12)
+subplot(122),mesh(X,Y,xysol),axis('square')
+view(330,30)
+fprintf('done\n')
+return

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/graphs/flowplot09.m

@@ -1,0 +1,77 @@
+function flowplot09(qmethod,sol,By,Bx,A,xy,xyp,x,y,bound,bndxy,bnde,obs,contourn,spc,fig)
+%FLOWPLOT09 plots flow data on general domain
+%   flowplot09(qmethod,xns,By,Bx,A,xy,xyp,x,y,bound,bndxy,bnde,obs,contourn,1,69)
+%   input
+%          qmethod    mixed method 
+%          sol        flow solution vector
+%          By         velocity  y-derivative matrix    
+%          Bx         velocity x-derivative matrix    
+%          A          vector diffusion matrix
+%          xy         velocity nodal coordinate vector  
+%          xyp        pressure nodal coordinate vector  
+%          x          vector of x-axis interpolation points
+%          y          vector of y-axis interpolation points
+%          bound      boundary vertex vector
+%          spc        uniform/nonuniform streamline switch 
+%          fig        figure number
+%
+% calls function streambc.m to set boundary values
+%   IFISS function: HCE; DJS; 30 April 2012.
+%   Adapted from code written by M. Wu, 2009
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage;
+%
+if ishandle(fig), clf(fig); end
+nvtx=length(xy); nu=2*nvtx;   %np=length(xyp);
+Asv=A(1:nvtx,1:nvtx);
+%
+%% compute auxiliary quantities
+u=sol(1:nu);p=sol(nu+1:end);
+f=[By,-Bx]*u;
+[Asv,fsv] = streambc(Asv,f,xy,bound);
+phi=Asv\fsv;
+%
+%% plot pressure
+if qmethod==2
+   xx=x(1:2:end); yy=y(1:2:end);
+elseif qmethod==3
+   p=p(1:3:end); xx=x(1:end); yy=y(1:end);
+else
+   xx=x(1:end); yy=y(1:end);
+end
+% interpolate to a cartesian product mesh
+[X,Y]=meshgrid(xx,yy);
+xysol = griddata(xyp(:,1),xyp(:,2),p,X,Y);
+if size(obs,1)~=0
+   [II,JJ] = findobsXY(obs,X,Y,bndxy);  xysol(II,JJ)=nan;
+end
+figure(fig)
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/graphs/logoplot.m

@@ -1,0 +1,60 @@
+function logoplot(sol,xy,x,y)
+%LOGOPLOT plots IFISS logo
+%   logoplot(sol,xy,x,y,fig);
+%   input
+%          sol        nodal solution vector
+%          xy         nodal coordinate vector  
+%          x          vector of x-axis interpolation points
+%          y          vector of y-axis interpolation points
+%
+%   IFISS function: AR; 26 January 2011.
+% Copyright (c) 2011 D.J. Silvester, H.C. Elman, A. Ramage 
+fprintf('plotting the logo ... ')
+% interpolate to a cartesian product mesh
+%figure
+[X,Y]=meshgrid(x,y);
+L = 40*griddata(xy(:,1),xy(:,2),sol,X,Y);
+%fprintf('original ... ')
+%surf(X,Y,L),axis('square')
+%view(-70,22), axis('off')
+shading interp
+fprintf('and refined ... ')
+%% following code taken from the Mathworks function logo ...
+%logoFig = figure('Color','w');
+ logoFig = figure('Color',[0 0 0]);
+logoax = axes('CameraPosition', [-393.4013 -565.1546  260.4819],...
+    'CameraTarget',[33 33 10], ...
+    'CameraUpVector',[0 0 1], ...
+    'CameraViewAngle',9.5, ...
+    'DataAspectRatio', [1 1 0.5],...
+    'Position',[0 0 1 1], ...
+    'Visible','off', ...
+    'parent',logoFig);
+s=surface(L, ...
+    'EdgeColor','none', ...
+    'FaceColor', 'red', ... %[0.9 0.2 0.2] , ...
+    'FaceLighting','phong', ...
+    'AmbientStrength',0.3, ...
+    'DiffuseStrength',0.6, ... 
+    'Clipping','off',...
+    'BackFaceLighting','lit', ...
+    'SpecularStrength',1.1, ...
+    'SpecularColorReflectance',1, ...
+    'SpecularExponent',7, ...
+    'Tag','IFISSLogo', ...
+    'parent',logoax);
+%   'FaceColor', [0.9 0.2 0.2] , ...
+l1 = light('Position',[80 100 20], ...
+    'Style','local', ...
+    'Color','yellow', ...
+    'parent',logoax);
+%   'Color',[0 0.8 0.8], ...
+l2 = light('Position',[.5 -1 .4], ...
+    'Color','w', ...
+    'parent',logoax);
+%   'Color',[0.8 0.8 0], ...
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/graphs/square_flowmovie.m

@@ -1,0 +1,65 @@
+function square_flowmovie(qmethod,sol,tt,By,Bx,A,xy,x,y,bound,ftime,fign,avi)
+%SQUARE_FLOWMOVIE generates flow movie on square  domain
+%   square_flowmovie(qmethod,U,time,By,Bx,A,xy,x,y,bound,.1,998,0);
+%   input
+%          qmethod    mixed method 
+%          U        velocity solution vector
+%          time         solution time vector  
+%          By         velocity  y-derivative matrix    
+%          Bx         velocity x-derivative matrix    
+%          A          vector diffusion matrix
+%          xy         velocity nodal coordinate vector   
+%          x          vector of x-axis interpolation points
+%          y          vector of y-axis interpolation points
+%          bound      boundary vertex vector
+%          ftime      controls speed of animation
+%          fign       figure number
+%          avi        0/1 switch to generate .avi file
+%
+% pressure solution is assumed to be essentially zero at outflow 
+% so streamfunction satisfies zero Neumann condition there
+% calls function xxstreambc.m to set boundary values
+%   IFISS function: DJS; 27 April 2012.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+fprintf('\nGenerating flow field movie ... ')
+fprintf('\ndo not adjust the figure ...')
+
+L=max(x); nstep=length(tt);
+nvtx=length(xy); nu=2*nvtx; 
+Asv=A(1:nvtx,1:nvtx); fzero=zeros(nvtx,1);
+%
+%% apply boundary conditions to the diffusion matrix
+[Abc,fzero]=streambc(Asv,fzero,xy,bound);
+%[LA,UA]= lu(Abc); 
+fprintf('\n   step      min_phi    max_phi\n')
+for k=1:nstep
+u=sol(:,k); ttk=tt(k);
+f=[By,-Bx]*u;
+[fsv]=xxstreambc(Asv,f,xy,bound,ttk);
+phi=Abc\fsv; %%UA\(LA\fsv);
+fprintf('  %4i   %12.5f   %9.3e\n', k, min(phi),max(phi));
+%
+%% plot velocity
+figure(fign)
+set(gcf,'Position',[5,5,600,600]);
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/graphs/square_unsteadyflowplot.m

@@ -1,0 +1,68 @@
+function square_unsteadyflowplot(qmethod,ev,sol,tt,By,Bx,G,xy,xyp,x,y,bound,ftime,fig)
+%SQUARE_UNSTEADYFLOWPLOT plots flow data on square domain
+%   square_unsteadyflowplot(qmethod,mv,sol,By,Bx,A,xy,xyp,x,y,bound,ftime,fig);
+%   input
+%          qmethod    mixed method 
+%          ev        mv/ev  Q2/Q1 element mapping matrix
+%          sol        flow solution vector
+%          tt         solution time vector  
+%          By         velocity  y-derivative matrix    
+%          Bx         velocity x-derivative matrix    
+%          G          veclocity mass matrix
+%          xy         velocity nodal coordinate vector  
+%          xyp        pressure nodal coordinate vector  
+%          x          vector of x-axis interpolation points
+%          y          vector of y-axis interpolation points
+%          bound      boundary vertex vector
+%          ftime      controls speed of animation
+%          fig        figure number
+%
+%   IFISS function: DJS; 3 May 2012.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+fprintf('\nrunning flow field animation ... ')
+L=max(x); nstep=length(tt);
+nvtx=length(xy); nu=2*nvtx; np=length(xyp);
+[LG,UG]= lu(G(1:nvtx,1:nvtx)); 
+fprintf('\n   step   mean_vorticity\n')
+for k=1:nstep
+% compute auxilliary quantites
+u=sol(:,k);
+ux=u(1:nvtx); uy=u(nvtx+1:nu);  utotal=sqrt(ux.*ux+ uy.*uy);
+fsv=-[By,-Bx]*u;
+omega=UG\(LG\fsv);
+if qmethod > 1,
+wev = vorticity_q2(xy,ev,omega,0);
+else
+wev = vorticity_q1(xy,ev,omega,0);
+end
+fprintf('  %4i    %11.3e \n', k, sum(wev));
+%
+% interpolate to a cartesian product mesh
+[X,Y]=meshgrid(x,y);
+figure(fig)
+%
+%% plot velocity magnitude
+xysol = griddata(xy(:,1),xy(:,2),utotal,X,Y);
+solheight = max(max(xysol))-min(min(xysol));
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/graphs/square_unsteadyflowref.m

@@ -1,0 +1,83 @@
+function square_unsteadyflowref(qmethod,ev,sol,tt,A,By,Bx,G,xy,xyp,x,y,bound,snaptime)
+%SQUARE_UNSTEADYFLOWREF plots square domain data at snapshot times
+%   square_unsteadyflowref(qmethod,mv,sol,tt,A,By,Bx,G,xy,xyp,x,y,bound,snaptime);
+%   input
+%          qmethod    mixed method 
+%          ev         mv/ev  Q2/Q1 element mapping matrix
+%          sol        flow solution vector
+%          tt         snapshot time vector
+%          A          vector diffusion matrix
+%          By         velocity  y-derivative matrix    
+%          Bx         velocity x-derivative matrix    
+%          G          veclocity mass matrix
+%          xy         velocity nodal coordinate vector  
+%          xyp        pressure nodal coordinate vector  
+%          x          vector of x-axis interpolation points
+%          y          vector of y-axis interpolation points
+%          bound      boundary vertex vector
+%          snaptime   vector of snapshot time step levels
+%
+% calls function xxstreambc.m to set boundary values
+%   IFISS function: DJS; 3 May 2012.
+% Copyright (c) 2012 D.J. Silvester, H.C. Elman, A. Ramage 
+fprintf('\n   Plotting flow field snapshots ... ')
+L=max(x); nstep=length(snaptime);
+if nstep>9, error('Too many snapshots!'), end
+nvtx=length(xy); nu=2*nvtx; np=length(xyp);
+[LG,UG]= lu(G(1:nvtx,1:nvtx)); 
+Asv=A(1:nvtx,1:nvtx); fzero=zeros(nvtx,1);
+[Abc,fzero]=streambc(Asv,fzero,xy,bound);
+[LA,UA]=lu(Abc); 
+%
+fprintf('\n   step   time    mean_vorticity    min_phi  max_phi\n')
+%
+% ------------------ loop over snapshots
+for k=1:nstep
+kk=snaptime(k); ttk=tt(kk);
+% compute derived quantites
+u=sol(:,kk);
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/graphs/step_flowmovie.m

@@ -1,0 +1,73 @@
+function step_flowmovie(qmethod,sol,tt,By,Bx,A,xy,x,y,bound,ftime,fign,avi)
+%STEP_FLOWMOVIE flow movie on extended step domain
+%   step_flowmovie(qmethod,U,time,By,Bx,A,xy,x,y,bound,0.1,999,0);
+%   input
+%          qmethod    mixed method 
+%          sol        velocity solution vector
+%          tt         solution time vector  
+%          By         velocity  y-derivative matrix    
+%          Bx         velocity x-derivative matrix    
+%          A          vector diffusion matrix
+%          xy         velocity nodal coordinate vector   
+%          x          vector of x-axis interpolation points
+%          y          vector of y-axis interpolation points
+%          bound      boundary vertex vector
+%          ftime      controls speed of animation
+%          fign       figure number
+%          avi        0/1 switch to generate .avi file
+%
+% pressure solution is assumed to be essentially zero at outflow 
+% so streamfunction satisfies zero Neumann condition there
+% calls function xxstreambc.m to set boundary values
+%   IFISS function: DJS; 27 April 2012.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+fprintf('\nGenerating flow field movie ... ')
+fprintf('\ndo not adjust the figure ...')
+L=max(x); nstep=length(tt);
+nvtx=length(xy); nu=2*nvtx; 
+Asv=A(1:nvtx,1:nvtx); fzero=zeros(nvtx,1);
+%
+%% apply boundary conditions to the diffusion matrix
+[Abc,fzero]=streambc(Asv,fzero,xy,bound);
+%[LA,UA]= lu(Abc); 
+fprintf('\n   step      min_phi    max_phi\n')
+for k=1:nstep
+u=sol(:,k); ttk=tt(k);
+f=[By,-Bx]*u;
+[fsv]=xxstreambc(Asv,f,xy,bound,ttk);   
+phi=Abc\fsv;%%UA\(LA\fsv);
+fprintf('  %4i   %12.5f   %9.3e\n', k, min(phi),max(phi));
+%
+%% plot velocity
+figure(fign)
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/grids/grid_mergeleftright.m

@@ -1,0 +1,59 @@
+function [xy,mv,left,bottom,top,right,mbound,x,y] = ...
+   grid_mergeleftright(xyl,mvl,leftl,rightl,bottoml,topl,mboundl,xl,yl,...
+   xyr,mvr,leftr,rightr,bottomr,topr,mboundr,xr,yr);
+%GRID_MERGELEFTRIGHT  merger of inlet/outlet channel domain 
+% 
+%   IFISS function: DJS; 6 November 2011. 
+% Copyright (c) 2012 D.J. Silvester, H.C. Elman, A. Ramage 
+fprintf('   Merger of two x-channel grids \n')
+nleft=length(rightl); nright=length(leftr);
+% right domain numbering is modified
+% identify the end nodes on the left
+startl=rightl(1); endl=rightl(end);
+if endl~=startl+nleft-1, 
+    error('left domain is not y-numbered, try again.'), end
+% identify the end nodes on the right
+startr=leftr(1); endr=leftr(end);
+if endr~=startr+nright-1, 
+    error('right domain is not y-numbered, try again.'), end
+xloop=leftr(2)-leftr(1); %fprintf('right increment is %g\n', xloop)
+% find y-values of zipped nodes on left
+yzip=xyl(rightl,2);
+% find bottom of zip on the right
+kb=find(xyr(leftr,2)==yzip(1));
+refleftr=leftr(kb:end);
+% find y-values of zipped nodes on right
+yzipz=xyr(refleftr,2);
+fprintf('   zip distance is %9.4e ... it should be close to zero!\n',...
+          norm(yzip-yzipz,inf))
+%
+% merge xy data
+nvtxl=size(xyl,1); nell=length(mvl(:,1));
+xyl(rightl,:)=[];
+xy=[xyl;xyr];
+x=unique([xl,xr]); y=unique([yl,yr]);
+% merge mv data
+extra=nvtxl-nleft;
+mvr=mvr+extra; refleftr=refleftr+extra;
+kk=find(mboundl(:,2)==2); nkk=length(kk);
+list=mboundl(kk,1); mboundl(kk,:)=[];
+      mvl(list,2) = refleftr(1:2:end-1);  
+      mvl(list,3) = refleftr(3:2:end);   
+      mvl(list,6) = refleftr(2:2:end-1); mvl(list,:);     
+mv=[mvl;mvr];
+%
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/grids/square_domain.m

@@ -1,0 +1,93 @@
+function square_domain
+%SQUARE_DOMAIN square domain Q2 grid generator
+%   square_domain;
+% 
+% grid defining data is saved to the file: square_grid.mat
+%   IFISS function: DJS; 7 May 2006.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+fprintf('\n\nGrid generation for unit square  domain.\n')
+nc=default('grid parameter: 3 for underlying 8x8 grid (default is 16x16)',4);
+if nc<2, error('illegal parameter choice, try again.'), end
+grid_type=default('uniform/stretched grid (1/2) (default is uniform)',1);
+n=2^nc; np=n/2; nq=n/4;
+%
+%% compute (x,y) coordinates of vertices
+% y-direction
+if grid_type==2
+   hmax=nc/(2^(nc+1));
+   x1=-1;x2=-2*hmax;x3=2*hmax;x4=1;nx1=2^(nc-1)-1;nx2=2;nx3=2^(nc-1)-1;
+   y1=-1;y2=-2*hmax;y3=2*hmax;y4=1;ny1=2^(nc-1)-1;ny2=2;ny3=2^(nc-1)-1;
+   y=subint(y1,y2,y3,y4,ny1,ny2,ny3);
+   stretch=(y(3)-y(2))/(y(2)-y(1));
+   left=-1;
+   x=y;
+else
+   square_type=default('[0,1] or [-1,1] square (enter 1/2) (default is [-1,1])',2);
+   yy=[1/np:1/np:1];
+   ypos=[0,yy];
+   yneg=-yy(length(yy):-1:1);
+   y=[yneg,ypos]'; 
+   left=-1;
+   if square_type==1
+      y=[0:1/(2*np):1]'; left=0;
+   end
+   x=y; 
+   end
+%
+%% compute biquadratic element coordinates
+nvtx=(n+1)*(n+1);
+[X,Y]=meshgrid(x,y);
+xx=reshape(X',nvtx,1);
+yy=reshape(Y',nvtx,1);
+xy=[xx(:),yy(:)];
+%
+kx = 1;
+ky = 1;
+mel=0;
+for j=1:np
+   for i=1:np
+      mref=(n+1)*(ky-1)+kx;
+      mel=mel+1;
+      nvv(1) = mref;
+      nvv(2) = mref+2;
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/ifissdemos/stokesflowdemo.m

@@ -1,0 +1,19 @@
+%ifissdemo  EST_MINRES demonstration 
+close all, batchmode('S3'), 
+fprintf('\nCHECK OUT the Stokes driven cavity flow solution\n'),
+%%% positioning is [left,bottom,width,height]
+figure(30), set(gcf,'Position',[0,450,450,450],'Visible','off');
+figure(33), set(gcf,'Position',[0,450,450,450]); 
+figure(34), set(gcf,'Position',[450,450,450,450]); pause(15) 
+figure(34), set(gcf,'Position',[0,450,450,450],'Visible','off');
+figure(33), set(gcf,'Position',[0,350,350,350],'Visible','on'); drawnow
+fprintf('\n\nCHECK the iterative solver convergence ...\n'),
+batchmode('itsolve_stokesx1'), load batchrun_itsolve.mat
+[np,nu]=size(Bst);
+xdiff=norm(xest(1:nu)-xst(1:nu),inf);
+fprintf('velocity solution difference is %7.3e\n',xdiff),
+figure(1),title('ESTMINRES | convergence comparison','FontSize',12),
+figure(30), set(gcf,'Visible','on');
+figure(34), set(gcf,'Visible','on');
+figure(33), set(gcf,'Position',[0,450,450,450]);
+fprintf('\nEnd of Stokes flow demo. Voila!\n'),

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/matlab704/fcnchk.m

@@ -1,0 +1,141 @@
+function [f,msg] = fcnchk(fun,varargin)
+%FCNCHK Check FUNFUN function argument.
+%   FCNCHK(FUN,...) returns an inline object based on FUN if FUN
+%   is a string containing parentheses, variables, and math
+%   operators.  FCNCHK simply returns FUN if FUN is a function handle, 
+%   or a matlab object with an feval method (such as an inline object). 
+%   If FUN is a string name of a function (e.g. 'sin'), FCNCHK returns a
+%   function handle to that function.
+%
+%   FCNCHK is a helper function for FMINBND, FMINSEARCH, FZERO, etc. so they
+%   can compute with string expressions in addition to m-file functions.
+%
+%   FCNCHK(FUN,...,'vectorized') processes the string (e.g., replacing
+													   %   '*' with '.*') to produce a vectorized function.
+%
+%   When FUN contains an expression then FCNCHK(FUN,...) is the same as
+%   INLINE(FUN,...) except that the optional trailing argument 'vectorized'
+%   can be used to produce a vectorized function.
+%
+%   [F,MSG] = FCNCHK(...) returns an empty structure in MSG if successful
+%   or an error message structure if not.
+%
+%   See also ERROR, INLINE, @.
+
+%   Copyright 1984-2008 The MathWorks, Inc.
+%   $Revision: 1.29.4.12 $  $Date: 2008/09/13 06:56:42 $
+
+message = '';
+msgident = '';
+
+nin = nargin;
+if nin > 1 && strcmp(varargin{end},'vectorized')
+vectorizing = true;
+nin = nin - 1;
+else
+vectorizing = false;
+end
+
+if ischar(fun)
+fun = strtrim_local_function(fun);
+% Check for non-alphanumeric characters that must be part of an
+% expression.
+if isempty(fun),
+f = inline('[]');
+elseif ~vectorizing && isidentifier_local_function(fun)
+f = str2func(fun); % Must be a function name only
+% Note that we avoid collision of f = str2func(fun) with any local
+% function named fun, by uglifying the local function's name
+if isequal('x',fun)
+warning('MATLAB:fcnchk:AmbiguousX', ...
+['Ambiguous expression or function input.\n The string ''x'' will be ',...
+'interpreted as the name of a ',...
+'function called ''x'' \n (e.g., x.m) and not as the mathematical expression ''x'' (i.e., f(x)=x). \n ',...
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/matlab704/iterapp_704.m

@@ -1,0 +1,70 @@
+function y = iterapp_704(op,afun,atype,afcnstr,x,varargin)
+%ITERAPP   Apply matrix operator to vector and error gracefully.
+%   ITERAPP(OP,AFUN,ATYPE,AFCNSTR,X) applies matrix operator AFUN to vector
+%   X. If ATYPE is 'matrix, then AFUN is a matrix and the OP is applied
+%   directly. OP is either 'mtimes' or 'mldivide'.
+%   ATYPE and AFCNSTR are used in case of error.
+%   ITERAPP(OP,AFUN,ATYPE,AFCNSTR,X,P1,P2,...) allows extra arguments to
+%   AFUN(X,P1,P2,...) although this usage is now discouraged in favor of
+%   using anonymous functions.
+%   AFUN(X,P1,P2,...,PN,TFLAG) should accept a TFLAG as its final input if
+%   the calling function is BICG, LSQR or QMR. TFLAG is either 'transp' or
+%   'notransp' depending on whether A' OP X or A OP X is required.
+%   ITERAPP is designed for use by iterative methods like PCG which
+%   require matrix operators AFUN representing matrices A to operate on
+%   vectors X and return A*X and may also have operators MFUN representing
+%   preconditioning matrices M operate on vectors X and return M\X.
+%
+%   See also BICG, BICGSTAB, CGS, GMRES, LSQR, MINRES, PCG, QMR, SYMMLQ.
+
+%   Copyright 1984-2004 The MathWorks, Inc.
+%   $Revision: 1.7.4.2 $ $Date: 2004/12/06 16:35:55 $
+
+if isequal(atype,'matrix')
+    switch lower(op)
+        case 'mtimes'
+            if (nargin >= 6) & isequal(varargin{end},'transp')
+                y = afun' * x;
+            else
+                y = afun * x;
+            end
+        case 'mldivide'
+            if (nargin >= 6) & isequal(varargin{end},'transp')
+                y = afun' \ x;
+            else
+                y = afun \ x;
+            end
+        otherwise
+            error('MATLAB:iterapp:InvalidOp', 'Invalid operation.')
+    end
+else
+    try
+        if (nargin >= 6) & isequal(varargin{end},'notransp')
+            % A request for A*x coming from BICG, LSQR and QMR
+            try
+                % New syntax: we now request afun(x,P1,P2,...,PN,'notransp')
+                y = afun(x,varargin{:});
+            catch
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/navier_flow/Contents.m

@@ -1,0 +1,35 @@
+% NAVIER_FLOW
+%
+% Files
+%   Cpre_q1p0             - generate stabilization matrices for Q1-P0 
+%   Cpre_q1q1             - generate stabilization matrices for Q1-Q1 
+%   fpsetup_q0            - Q0 pressure convection-diffusion matrix 
+%   fpsetup_q1            - Q1 pressure convection-diffusion matrix 
+%   fpsetup_q2p1          - Q2-P1 pressure convection-diffusion matrix 
+%   fpszsetup_q1          - modified Q1 PCD matrix with stabilization 
+%   fpzsetup_q0           - modified Q0 PCD matrix 
+%   fpzsetup_q1           - modified Q1 PCD matrix 
+%   fpzsetup_q2p1         - modified P1 PCD matrix 
+%   helpme_navier         - Navier-Stokes flow problem interactive help 
+%   localbc_xycd          - imposes vector BC for Poisson error estimator 
+%   navier_q1             - Q1 convection matrix 
+%   navier_q2             - Q2 convection matrix 
+%   navierpost_q1p0_bc    - postprocesses Poisson error estimator 
+%   navierpost_q1p0_p     - computes Poisson error estimator for Q1-P0 
+%   newton_q1             - Q1 convection derivative matrices 
+%   newton_q2             - Q2 convection derivative matrices 
+%   newtonbc              - imposes Dirichlet BC on Jacobian
+%   null_pressure_index   - index associated with the pressure nullspace
+%   pressurebc            - fixes singularity in Laplacian matrix
+%   solve_navier          - solve singular Navier-Stokes problem
+%   solve_obstacle_navier - solve Navier-Stokes problem in obstacle domain
+%   solve_plate_navier    - solve Navier-Stokes problem in slit domain
+%   solve_step_navier     - solve Navier-Stokes problem in step domain
+%   navierpost            - estimates Q2-P1 or Q2-Q1 NS error distribution 
+%   navierpost_q2p1       - vectorized Q2-P1 error estimator 
+%   navierpost_q2q1       - vectorized Q2-Q1 error estimator 
+%   fpxsetup_q1           - modified Q1 PCD matrix (ESW2005 version)
+%   it_nstokes            - iterative solver for flow linear system
+%   it_solve_navier       - iterative solution of N-S in square domain
+%   it_solve_step_navier  - iterative solution of N-S in step domain
+%   xsolve_navier         - solve (possibly singular) Navier-Stokes problem 

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/navier_flow/navierpost.m

@@ -1,0 +1,27 @@
+%NAVIERPOST estimates Q2-P1 or Q2-Q1 NS error distribution 
+%   IFISS scriptfile: DJS; 30 September 2013.
+% Copyright (c) 2010 D.J. Silvester, Qifeng Liao
+if qmethod <2,
+error('This code is designed for inf-sup stable approximation!'), end  
+% code for outflow boundary
+if domain == 3 || domain == 10 || domain==4,  % step, channel, obstacle
+      fprintf('Natural outflow on right hand vertical boundary ..\n')
+      neumannb=out_Neumann_bound(mv,xy);
+else, fprintf('Enclosed flow ..\n'),  neumannb=[];
+end
+if qmethod==3,
+fprintf('FAST Q2-P1 NS a posteriori error estimation \n')
+[hx,hy,eex] = eexgen(xy,xyp,mv,ee); % check element numbering
+[error_x,error_y,fex,fey,ae]=navierpost_q2p1(viscosity,xy,mv,mbound,neumannb,xns,eex,hx,hy);
+elseif qmethod==2,
+fprintf('FAST Q2-Q1 NS a posteriori error estimation \n')
+[x_q2p1,y_q2p1,xy_q2p1,xyp_q2p1,ee] = q2p1grid(x,y,xy,mv,bound);
+[hx,hy,eex] = eexgen(xy_q2p1,xyp_q2p1,mv,ee); % check element numbering
+[error_x,error_y,fex,fey,ae]=navierpost_q2q1(viscosity,xy,mv,mp,mbound,neumannb,xns,eex,hx,hy);
+else, error('undefined mixed approximation!'), end
+error_div = q2div(xy,mv,xns);
+error_tot = sqrt(error_x+error_y+error_div.^2);
+errorest=norm(error_tot,2);
+fprintf('estimated energy error is %10.4e \n',errorest)
+if domain ==3, eplotl2(error_tot,mv,xy,x,y,67,'Estimated energy error');
+else, eplot2(error_tot,mv,xy,x,y,67,'Estimated energy error'); end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/navier_flow/solve_obstacle_navier.m

@@ -1,0 +1,165 @@
+%SOLVE_OBSTACLE_NAVIER solve Navier-Stokes problem in obstacle domain
+%   IFISS scriptfile: HCE; DJS;  29 September 2013.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+% Code written by M. Wu, 2009
+
+clear variables
+fprintf('Incompressible flow problem on obstacle domain ...\n')
+viscosity=default('viscosity parameter (default 1/50)',1/50);
+nlmethod=default('Picard/Newton/hybrid linearization 1/2/3 (default hybrid)',3);
+nlmethod=nlmethod-1;
+if nlmethod==0,
+   maxit_p=default('number of Picard iterations (default 9)',9);
+   maxit_n=0;
+elseif nlmethod==1,
+   maxit_p=0;
+   maxit_n=default('number of Newton iterations (default 6)',6);
+else
+   maxit_p=default('number of Picard iterations (default 6)',6);
+   maxit_n=default('number of Newton iterations (default 5)',5);
+end
+tol_nl=default('nonlinear tolerance (default 1.d-8)',1.d-8);
+%
+%
+%% initialize for nonlinear iteration: compute Stokes solution
+%% load assembled matrices
+gohome
+cd datafiles
+load obstacle_stokes_nobc.mat
+%
+fprintf('stokes system ...\n') 
+%% boundary conditions
+[Ast,Bst,fst,gst] = flowbc(A,B,f,g,xy,bound);
+nlres0_norm = norm([fst;gst]);
+%
+nv=length(fst)/2; np=length(gst); 
+if qmethod>1,
+   beta=0;
+   xst=[Ast,Bst';Bst,sparse(np,np)]\[fst;gst];
+elseif qmethod==1
+		beta=1/4;     % default parameter
+   xst=[Ast,Bst';Bst,-beta*C]\[fst;gst];
+elseif qmethod==0
+   fprintf('computing pressure stabilized solution...\n')
+   xst=[Ast,Bst';Bst,-C]\[fst;gst]; beta=1;
+end
+% compute residual of Stokes solution
+if qmethod>1
+   N = navier_q2(xy,mv,xst);
+elseif qmethod<=1,
+   nubeta=beta/viscosity;
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/navier_flow/solve_step_navier.m

@@ -1,0 +1,167 @@
+%SOLVE_STEP_NAVIER solve Navier-Stokes problem in step domain
+%   IFISS scriptfile: DJS; 29 September 2013.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+clear variables
+fprintf('Incompressible flow problem on step domain ...\n')
+viscosity=default('viscosity parameter (default 1/50)',1/50);
+nlmethod=default('Picard/Newton/hybrid linearization 1/2/3 (default hybrid)',3);
+nlmethod=nlmethod-1;
+if nlmethod==0,
+   maxit_p=default('number of Picard iterations (default 9)',9);
+   maxit_n=0;
+elseif nlmethod==1,
+   maxit_p=0;
+   maxit_n=default('number of Newton iterations (default 6)',6);
+else
+   maxit_p=default('number of Picard iterations (default 2)',2);
+   maxit_n=default('number of Newton iterations (default 4)',4);
+end
+tol_nl=default('nonlinear tolerance (default 1.e-8)',1.e-8);
+%
+%
+%% initialize for nonlinear iteration: compute Stokes solution
+%% load assembled matrices
+gohome
+cd datafiles
+load step_stokes_nobc.mat
+%
+fprintf('stokes system ...\n') 
+%% boundary conditions
+[Ast,Bst,fst,gst] = flowbc(A,B,f,g,xy,bound);
+nlres0_norm = norm([fst;gst]);
+%
+nv=length(fst)/2; np=length(gst); 
+if qmethod>1,
+   beta=0;
+   xst=[Ast,Bst';Bst,sparse(np,np)]\[fst;gst];
+elseif qmethod==1
+		beta=1/4;     % default parameter
+   xst=[Ast,Bst';Bst,-beta*C]\[fst;gst];
+elseif qmethod==0
+   fprintf('computing pressure stabilized solution...\n')
+   xst=[Ast,Bst';Bst,-C]\[fst;gst]; beta=1;
+end
+% compute residual of Stokes solution
+if qmethod>1
+   N = navier_q2(xy,mv,xst);
+elseif qmethod<=1,
+   nubeta=beta/viscosity;
+fprintf('\n nubeta is set to %e \n',nubeta)
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/navier_flow/test_problems/navier_testproblem.m

@@ -1,0 +1,43 @@
+%navier_testproblem    sets up reference Examples 7.1 to 7.4
+%   IFISS scriptfile: DJS; 29 September 2013.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference Navier-Stokes problem.\n');
+fprintf('\nchoose specific example (default is cavity)');
+fprintf('\n     1  Channel domain');
+fprintf('\n     2  Flow over a backward facing step');
+fprintf('\n     3  Lid driven cavity');
+fprintf('\n     4  Flow over a plate');
+fprintf('\n     5  Flow over an obstacle\n');
+sn = default('',3);
+if sn==1,
+system('/bin/cp ./stokes_flow/test_problems/poiseuille_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/poiseuille_bc.m ./stokes_flow/stream_bc.m');
+   channel_stokes; xsolve_navier;
+elseif sn==2,
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_bc.m ./stokes_flow/stream_bc.m');
+   longstep_stokes; solve_step_navier;
+elseif sn==3,
+   lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3); 
+   if lid_model ==3,
+system('/bin/cp ./stokes_flow/test_problems/regcavity_flow.m ./stokes_flow/specific_flow.m');
+   elseif lid_model ==2,
+system('/bin/cp ./stokes_flow/test_problems/tightcavity_flow.m ./stokes_flow/specific_flow.m');
+   else
+system('/bin/cp ./stokes_flow/test_problems/leakycavity_flow.m ./stokes_flow/specific_flow.m');
+   end
+system('/bin/cp ./stokes_flow/test_problems/zero_bc.m ./stokes_flow/stream_bc.m');
+   square_stokes; solve_navier;
+elseif sn==4,
+system('/bin/cp ./stokes_flow/test_problems/plate_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/plate_bc.m ./stokes_flow/stream_bc.m');
+   plate_stokes; solve_plate_navier;
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/navier_flow/xsolve_navier.m

@@ -1,0 +1,191 @@
+%XSOLVE_NAVIER solve (possibly singular) Navier-Stokes problem 
+%   IFISS scriptfile: DJS; 29 September 2013.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+clear variables
+fprintf('(incompressible flow problem on square domain ...\n')
+viscosity=default('viscosity parameter (default 1/100)',1/100);
+nlmethod=default('Picard/Newton/hybrid linearization 1/2/3 (default hybrid)',3);
+nlmethod=nlmethod-1;
+if nlmethod==0,
+   maxit_p=default('number of Picard iterations (default 9)',9);
+   maxit_n=0;
+elseif nlmethod==1,
+   maxit_p=0;
+   maxit_n=default('number of Newton iterations (default 6)',6);
+else
+   maxit_p=default('number of Picard iterations (default 2)',2);
+   maxit_n=default('number of Newton iterations (default 4)',4);
+end
+tol_nl=default('nonlinear tolerance (default 1.e-8)',1.e-8);
+%
+%
+%% initialize for nonlinear iteration: compute Stokes solution
+%% load assembled matrices
+gohome
+cd datafiles
+load square_stokes_nobc.mat
+%
+fprintf('stokes system ...\n') 
+%% boundary conditions
+[Ast,Bst,fst,gst] = flowbc(A,B,f,g,xy,bound);
+nlres0_norm = norm([fst;gst]);
+%
+nv=length(fst)/2; np=length(gst); 
+if qmethod>1,
+   beta=0;
+   xst=[Ast,Bst';Bst,sparse(np,np)]\[fst;gst];
+elseif qmethod==1
+	  beta=1/4;     % default parameter
+   xst=[Ast,Bst';Bst,-beta*C]\[fst;gst];
+elseif qmethod==0
+   fprintf('computing pressure stabilized solution...\n')
+   xst=[Ast,Bst';Bst,-C]\[fst;gst]; beta=1;
+end
+[msgstr,msgid] = lastwarn;
+if length(msgid)==27,
+   if sum(msgid=='MATLAB:nearlySingularMatrix')==27,
+      fprintf('This should not cause difficulty for enclosed flow problems.\n\n');
+   end
+end
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/Contents.m

@@ -1,0 +1,31 @@
+% POISSON CONTROL:  J.W. Pearson, 30 July 2012
+%
+% Files | Poisson control problem and direct solution
+%   ell_poissoncontrol - forms and solves matrix system on L-shaped domain
+%   nonzerobc_input - imposes specific Dirichlet boundary conditions
+%   poissoncontrol_errortest - computes error estimates for Problem 3 sol'n
+%   poissoncontrol_rhs_custom - inputs specific desired state and BCs
+%   poissoncontrol_rhs_ex1 - inputs desired state and BCs for Problem 1
+%   poissoncontrol_rhs_ex2 - inputs desired state and BCs for Problem 2
+%   poissoncontrol_rhs_ex3 - inputs desired state and BCs for Problem 3
+%   poissoncontrol_rhs_ex4 - inputs desired state and BCs for Problem 4
+%   square_poissoncontrol - forms and solves matrix system on square domain
+%
+% Files | iterative solution of matrix system
+%   it_solve_poissoncontrol - solves iteratively the matrix system
+%   it_solve_poissoncontrol_reduced - solves (reduced) 2x2 block system
+%   mass_chebyshev - applies (1,1)-block preconditioner (Chebyshev)
+%   mass_chebyshev_reduced - as above for (reduced) 2x2 block system
+%   mass_diagonal - applies (1,1)-block preconditioner (diagonal approx.)
+%   mass_diagonal_reduced - as above for (reduced) 2x2 block system
+%   mass_exact - applies (1,1)-block preconditioner (exact solution)
+%   mass_exact_reduced - as above for (reduced) 2x2 block system
+%   mass_identity - identity operator for (1,1)-block
+%   mg_poissoncontrol - sets up inputs for geometric multigrid solver
+%   poissoncontrol_minres - MINRES algorithm for Poisson control problem
+%   poissoncontrol_minres_reduced - MINRES for (reduced) 2x2 block system
+%   schur_amg - applies Schur complement preconditioner (AMG)
+%   schur_diagonal - applies Schur comp. preconditioner (diagonal approx.)
+%   schur_exact - applies Schur complement preconditioner (exact solution)
+%   schur_gmg - applies Schur complement preconditioner (GMG)
+%   schur_identity - identity operator for Schur complement

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/ell_poissoncontrol.m

@@ -1,0 +1,104 @@
+%ELL_POISSONCONTROL is the driver for solving test problems
+%                      for Poisson control on the L-shaped domain
+%   
+% Within the code, the user has the following options:
+%   nc - grid parameter, so number of points in each dimension is 2^nc+1
+%   beta - regularization parameter in the problem
+%   grid_type - selects a uniform or stretched grid
+%   qmethod - selects Q1 or Q2 finite element basis functions
+%   system_reduce - chooses to reduce matrix to 2x2 block system, or not
+%   iter_solve - solves problem directly or iteratively (by calling
+%   function poissoncontrol_iter)
+%   
+%   IFISS function: JWP;  DJS; 30 July 2012.
+% Copyright (c) 2012 J.W. Pearson 
+
+% Make choices as to the problem being solved
+nc = default('grid parameter: 3 for underlying 8x8 grid (default is 16x16)',4);
+if nc<2, error('illegal parameter choice ''nc'', try again.'), end
+beta = default('regularization parameter beta (default is 0.01)',0.01);
+if beta<eps, error('illegal parameter choice ''beta'', try again.'), end
+ell_type = default('[0,1] or [-1,1] ''L'' (enter 1/2) (default is [-1,1])',2);
+stretch = default('grid stretch factor (default is 1)',1);
+qmethod = default('Q1/Q2 approximation 1/2? (default Q1)',1);
+
+% Denote that this is Problem 4
+problem = 4;
+
+% Compute matrices specifying location of nodes
+[x,y,xy,bound,mv,mbound] = ell_domain_x(nc,stretch);
+% How large is resulting stiffness or mass matrix?
+np = length(xy(:,1)); % entire matrix system is thus 3*np-by-3*np in size
+
+if ell_type==1
+    x = (x+1)/2; y = (y+1)/2; xy = (xy+1)/2;
+elseif ell_type==2
+else, error('illegal parameter choice ''ell_type'', try again.'), end
+
+% Compute connectivity, stiffness and mass matrices, for Q1 or Q2 elements
+if qmethod==1
+    [ev,ebound] = q1grid(xy,mv,bound,mbound);
+    [K,M] = femq1_diff(xy,ev);
+elseif qmethod==2
+    [x,y,xy] = q2grid(x,y,xy,mv,bound); % to generate plot of Q2 elements
+    [K,M] = femq2_diff(xy,mv);
+else
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/it_solve_poissoncontrol.m

@@ -1,0 +1,265 @@
+%IT_SOLVE_POISSONCONTROL iterative solution of Poisson control problem
+%IFISS scriptfile: JWP; DJS; 29 June 2012. 
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester
+
+% Declare global variables for scalar and vector problems
+global amg_grid amg_smoother number_vcycles
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%%% POISSON CONTROL PROBLEM
+fprintf('discrete Poisson control system ...\n')
+% use MINRES as iterative method
+fprintf('iterative solution with preconditioned MINRES\n')
+%
+% set parameters
+tol = default('tolerance? (default 1e-6)',1e-6);
+maxit = default('maximum number of iterations? (default 100)',100);
+%
+% select preconditioner and construct it
+fprintf('preconditioner:\n');
+fprintf('   0   none\n');
+fprintf('   1   diagonal preconditioner\n');
+fprintf('   2   ideal block preconditioner, Schur comp. approx. 1\n');
+fprintf('   3   ideal block preconditioner, Schur comp. approx. 2\n');
+fprintf('   4   diagonal mass matrix approx., Schur comp. approx. 1, GMG\n');
+fprintf('   5   diagonal mass matrix approx., Schur comp. approx. 1, AMG\n');
+fprintf('   6   diagonal mass matrix approx., Schur comp. approx. 2, GMG\n');
+fprintf('   7   diagonal mass matrix approx., Schur comp. approx. 2, AMG\n');
+fprintf('   8   Chebyshev mass matrix approx., Schur comp. approx. 1, GMG\n');
+fprintf('   9   Chebyshev mass matrix approx., Schur comp. approx. 1, AMG\n');
+fprintf('   10  Chebyshev mass matrix approx., Schur comp. approx. 2, GMG\n');
+fprintf('   11  Chebyshev mass matrix approx., Schur comp. approx. 2, AMG\n');
+precon = default('preconditioner (default is Chebyshev, Schur comp. approx. 2, AMG)',11);
+if precon==0,      % none
+  fprintf('no preconditioning ...\n') 
+  MA = 'mass_identity';
+  MS = 'schur_identity';
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/it_solve_poissoncontrol_reduced.m

@@ -1,0 +1,266 @@
+%IT_SOLVE_POISSONCONTROL_REDUCED iterative solution of 2x2 block matrix
+%system for the Poisson control problem
+%IFISS scriptfile: JWP; DJS; 29 June 2012. 
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester
+
+% Declare global variables for scalar and vector problems
+global amg_grid amg_smoother number_vcycles
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%%% POISSON CONTROL PROBLEM
+fprintf('discrete Poisson control system ...\n')
+% use MINRES as iterative method
+fprintf('iterative solution with preconditioned MINRES\n')
+%
+% set parameters
+tol = default('tolerance? (default 1e-6)',1e-6);
+maxit = default('maximum number of iterations? (default 100)',100);
+%
+% select preconditioner and construct it
+fprintf('preconditioner:\n');
+fprintf('   0   none\n');
+fprintf('   1   diagonal preconditioner\n');
+fprintf('   2   ideal block preconditioner, Schur comp. approx. 1\n');
+fprintf('   3   ideal block preconditioner, Schur comp. approx. 2\n');
+fprintf('   4   diagonal mass matrix approx., Schur comp. approx. 1, GMG\n');
+fprintf('   5   diagonal mass matrix approx., Schur comp. approx. 1, AMG\n');
+fprintf('   6   diagonal mass matrix approx., Schur comp. approx. 2, GMG\n');
+fprintf('   7   diagonal mass matrix approx., Schur comp. approx. 2, AMG\n');
+fprintf('   8   Chebyshev mass matrix approx., Schur comp. approx. 1, GMG\n');
+fprintf('   9   Chebyshev mass matrix approx., Schur comp. approx. 1, AMG\n');
+fprintf('   10  Chebyshev mass matrix approx., Schur comp. approx. 2, GMG\n');
+fprintf('   11  Chebyshev mass matrix approx., Schur comp. approx. 2, AMG\n');
+precon = default('preconditioner (default is Chebyshev, Schur comp. approx. 2, AMG)',11);
+if precon==0,      % none
+  fprintf('no preconditioning ...\n') 
+  MA = 'mass_identity';
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/mass_chebyshev_reduced.m

@@ -1,0 +1,14 @@
+function x_it=mass_chebyshev_reduced(x_it,massparams)
+%MASS_CHEBYSHEV_REDUCED (1,1)-block preconditioner for 2x2 block Poisson
+%control problem, using Chebyshev semi-iteration
+%   x_it = mass_chebyshev_reduced(x_it,mparams);
+%   input
+%          x_it         operand for preconditioning operator
+%          mparams      structure defining block preconditioning operator
+%   output
+%          x_it         result of preconditioning operation
+%
+%   IFISS function: HCE, JWP, DJS; 20 October 2013.
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester, H.C. Elman, A. Ramage
+
+x_it = m_masscheb(x_it,massparams);

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/mass_diagonal.m

@@ -1,0 +1,18 @@
+function x_it=mass_diagonal(x_it,massparams)
+%MASS_DIAGONAL (1,1)-block preconditioner for Poisson control, using
+%diagonal matrix inversion
+%   x_it = mass_diagonal(x_it,mparams);
+%   input
+%          x_it         operand for preconditioning operator
+%          mparams      structure defining block preconditioning operator
+%   output
+%          x_it         result of preconditioning operation
+%
+%   IFISS function: JWP, DJS, HCE; 29 June 2012.
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester, H.C. Elman, A. Ramage
+
+nm = size(massparams.M,1);
+r1 = x_it(1:nm); r2 = x_it(nm+1:2*nm);
+z1 = diag(diag(massparams.M))\r1;
+z2 = 1/massparams.beta*(diag(diag(massparams.M))\r2);
+x_it = [z1; z2];

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/mass_diagonal_reduced.m

@@ -1,0 +1,14 @@
+function x_it=mass_diagonal_reduced(x_it,massparams)
+%MASS_DIAGONAL_REDUCED (1,1)-block preconditioner for 2x2 block Poisson
+%control problem, using diagonal matrix inversion
+%   x_it = mass_diagonal_reduced(x_it,mparams);
+%   input
+%          x_it         operand for preconditioning operator
+%          mparams      structure defining block preconditioning operator
+%   output
+%          x_it         result of preconditioning operation
+%
+%   IFISS function: JWP, DJS, HCE; 29 June 2012.
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester, H.C. Elman, A. Ramage
+
+x_it = diag(diag(massparams.M))\x_it;

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/mass_exact_reduced.m

@@ -1,0 +1,14 @@
+function x_it=mass_exact_reduced(x_it,massparams)
+%MASS_EXACT_REDUCED (1,1)-block preconditioner for 2x2 block Poisson
+%control problem, using exact mass matrix inversion
+%   x_it = mass_exact_reduced(x_it,mparams);
+%   input
+%          x_it         operand for preconditioning operator
+%          mparams      structure defining block preconditioning operator
+%   output
+%          x_it         result of preconditioning operation
+%
+%   IFISS function: JWP, DJS, HCE; 29 June 2012.
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester, H.C. Elman, A. Ramage
+
+x_it = massparams.M\x_it;

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/mg_poissoncontrol.m

@@ -1,0 +1,59 @@
+%MG_POISSON CONTROL GMG preconditioner for Poisson control problem
+%IFISS scriptfile: JWP, AR, HCE, DJS; 4 July 2012.
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester, H.C. Elman, A. Ramage
+
+nc=log2(length(y)-1);
+%
+% compute new MG data
+% h=2^(1-nc);
+fprintf('Setting up MG data ...')
+% top level
+mgdata(nc).matrix=A;
+if domain==1 % square
+  mgdata(nc).prolong=mg_prolong(2^nc,2^nc,x,y);
+elseif domain==2 % L-shaped
+  x=x';y=y';
+  mgdata(nc).prolong=mg_prolong_ell(nc,x,y);
+end
+xn=x;yn=y;
+% loop over remaining levels
+for level = nc-1:-1:2;
+  xn=xn(1:2:end);yn=yn(1:2:end);
+  if domain==1 % square
+     mgdata(level).matrix=mg_diff_setup(xn,yn);
+     mgdata(level).prolong=mg_prolong(2^level,2^level,xn,yn);
+  elseif domain==2 % L-shaped
+     mgdata(level).matrix=mg_diff_setup_ell(xn,yn);
+     mgdata(level).prolong=mg_prolong_ell(level,xn,yn);
+  end
+end
+fprintf('done\n')
+% gohome, cd datafiles, save mgdata_diff.mat mgdata
+if domain~=1, x=x';y=y';, end % square domain
+%
+% MG parameters
+smooth = default('Jacobi / ILU smoother? 1/3 (default is Jacobi)',1);
+if smooth==3
+   % point ILU
+   sweeps=1;stype=1;
+elseif smooth==2
+   % Gauss-Seidel
+   stype = default('point / line Gauss-Seidel? 1/2 (default is line)',2);
+   if stype==2
+      sweeps = default('number of Gauss-Seidel sweeps? 1/2/3/4 (default is 2)',2);
+   else
+      sweeps=1;
+   end
+else
+   % point Jacobi
+   sweeps=1;stype=1;
+end
+npre = default('number of pre- and post-smoothing steps? (default is 2)',2);
+npost = npre;
+%
+% construct smoother 
+if domain==1
+   smooth_data = mg_smooth(mgdata,nc,sweeps,smooth,stype);
+elseif domain==2
+   smooth_data = mg_smooth_ell(mgdata,nc,sweeps,smooth,stype);
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/poissoncontrol_rhs_custom.m

@@ -1,0 +1,38 @@
+function [yhat_vec,bc_nodes] = ...
+    poissoncontrol_rhs_custom(xy,bound,square_type)
+%POISSONCONTROL_RHS_CUSTOM allows the user to create vectors relating to
+%desired state and Dirichlet boundary condition for a chosen test problem
+%   [yhat_vec,bc_nodes] = poissoncontrol_rhs_custom(xy,bound,square_type)
+%   input
+%          xy           vertex coordinate vector  
+%          bound        boundary vertex vector
+%          square_type  specifies whether square is [0,1]^2 or [-1,1]^2
+%   output
+%          yhat_vec     vector relating to desired state
+%          bc_nodes     vector relating to Dirichlet boundary condition
+%
+% Guide for user:
+%   For square_type=1 and square_type=2, specify vectors yhat_vec and
+%   bc_nodes. For example:
+%     - If desired state is given by x_1*x_2, then type
+%          yhat_vec = xy(:,1).*xy(:,2);
+%     - If desired state is given by 1 if x_2<=0.5 and 0 o.w., then type:
+%          yhat_vec = xy(:,2)<=0.5;
+%     - If Dirichlet BC is given by x_1+x_2, then type:
+%          bc_nodes = xy(bound,1)+xy(bound,2);
+%     - If Dirichlet BC is 1 on top and right boundaries and 0 o.w., type:
+%          bc_nodes = (xy(bound,2)==1)+(xy(bound,2)==1)- ...
+%             ((xy(bound,2)==1).*(xy(bound,2)==1));
+%       The final term is to ensure the conditions x_1=0 and x_2=0 aren't
+%       "counted twice" at node (1,1); without it BC would be 2 there.
+%   
+%   IFISS function: JWP; 27 June 2012.
+% Copyright (c) 2012 J.W. Pearson
+
+if square_type==1 % specify functions if square is [0,1]^2
+    yhat_vec = ones(length(xy),1);
+    bc_nodes = zeros(length(bound),1);
+elseif square_type==2 % do same if square is [-1,1]^2
+    yhat_vec = ones(length(xy),1);
+    bc_nodes = zeros(length(bound),1);
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/poissoncontrol_rhs_ex1.m

@@ -1,0 +1,26 @@
+function [yhat_vec,bc_nodes] = poissoncontrol_rhs_ex1(xy,bound,square_type)
+%POISSONCONTROL_RHS_EX1 creates vectors relating to desired state and
+%Dirichlet boundary condition for Problem 1 
+%   [yhat_vec,bc_nodes] = poissoncontrol_rhs_ex1(xy,bound,square_type)
+%   input
+%          xy           vertex coordinate vector  
+%          bound        boundary vertex vector
+%          square_type  specifies whether square is [0,1]^2 or [-1,1]^2
+%   output
+%          yhat_vec     vector relating to desired state
+%          bc_nodes     vector relating to Dirichlet boundary condition
+%   
+%   IFISS function: JWP; 27 June 2012.
+% Copyright (c) 2012 J.W. Pearson
+
+if square_type==1  % specify functions if square is [0,1]^2
+    yhat_vec = (xy(:,1)<=0.5).*(xy(:,2)<=0.5).* ...
+        ((2*xy(:,1)-1).^2).*((2*xy(:,2)-1).^2); % specify desired state
+    bc_nodes = (xy(bound,1)<=0.5).*(xy(bound,2)<=0.5).* ...
+        ((2*xy(bound,1)-1).^2).*((2*xy(bound,2)-1).^2); % and Dirichlet BC
+elseif square_type==2 % do same if square is [-1,1]^2
+    yhat_vec = (xy(:,1)<=0).*(xy(:,2)<=0).* ...
+        (xy(:,1).^2).*(xy(:,2).^2);
+    bc_nodes = (xy(bound,1)<=0).*(xy(bound,2)<=0).* ...
+        (xy(bound,1).^2).*(xy(bound,2).^2);
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/poissoncontrol_rhs_ex2.m

@@ -1,0 +1,22 @@
+function [yhat_vec,bc_nodes] = poissoncontrol_rhs_ex2(xy,bound,square_type)
+%POISSONCONTROL_RHS_EX2 creates vectors relating to desired state and
+%Dirichlet boundary condition for Problem 2 
+%   [yhat_vec,bc_nodes] = poissoncontrol_rhs_ex2(xy,bound,square_type)
+%   input
+%          xy           vertex coordinate vector  
+%          bound        boundary vertex vector
+%          square_type  specifies whether square is [0,1]^2 or [-1,1]^2
+%   output
+%          yhat_vec     vector relating to desired state
+%          bc_nodes     vector relating to Dirichlet boundary condition
+%   
+%   IFISS function: JWP; 27 June 2012.
+% Copyright (c) 2012 J.W. Pearson
+
+if square_type==1 % specify functions if square is [0,1]^2
+    yhat_vec = (xy(:,1)<=0.5).*(xy(:,2)<=0.5); % specify desired state
+    bc_nodes = zeros(length(bound),1); % specify Dirichlet BC
+elseif square_type==2 % do same if square is [-1,1]^2
+    yhat_vec = (xy(:,1)<=0).*(xy(:,2)<=0);
+    bc_nodes = zeros(length(bound),1);
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/poissoncontrol_rhs_ex4.m

@@ -1,0 +1,34 @@
+function [yhat_vec,bc_nodes] = poissoncontrol_rhs_ex4(xy,bound,ell_type)
+%POISSONCONTROL_RHS_EX3 creates vectors relating to desired state and
+%Dirichlet boundary condition for Problem 4 
+%   [yhat_vec,bc_nodes] = poissoncontrol_rhs_ex4(xy,bound,square_type)
+%   input
+%          xy           vertex coordinate vector  
+%          bound        boundary vertex vector
+%          ell_type     whether L-shape is based around [0,1] or [-1,1]
+%   output
+%          yhat_vec     vector relating to desired state
+%          bc_nodes     vector relating to Dirichlet boundary condition
+%   
+%   IFISS function: JWP; 28 June 2012.
+% Copyright (c) 2012 J.W. Pearson
+
+% Specify coordinates of centre of Gaussian
+if ell_type==1 % if 'L' shape is based on [0,1]
+    gauss_ctr = [0.5 0.5];
+%     gauss_ctr = [0.25 0.75]; % centre of target Gaussian can be altered
+elseif ell_type==2 % and if 'L' shape is based on [-1,1]
+    gauss_ctr = [0 0];
+%     gauss_ctr = [-0.5 0.5]; % centre of target Gaussian can be altered
+end
+
+% Specify Gaussian parameter
+gauss_param = 1;
+
+% Specify desired state
+yhat_vec = exp(-gauss_param*((xy(:,1)-gauss_ctr(1)).^2+ ...
+    (xy(:,2)-gauss_ctr(2)).^2));
+% Specify Dirichlet BC
+bc_nodes = exp(-gauss_param*((xy(bound,1)-gauss_ctr(1)).^2+ ...
+    (xy(bound,2)-gauss_ctr(2)).^2));
+% bc_nodes = zeros(length(bound),1); % can also force BCs to be 0

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/schur_diagonal.m

@@ -1,0 +1,14 @@
+function x_it=schur_diagonal(x_it,schurparams)
+%SCHUR_DIAGONAL Schur complement preconditioner for Poisson control, using
+%diagonal matrix inversion
+%   x_it = schur_diagonal(x_it,mparams);
+%   input
+%          x_it         operand for preconditioning operator
+%          mparams      structure defining block preconditioning operator
+%   output
+%          x_it         result of preconditioning operation
+%
+%   IFISS function: JWP, DJS, HCE; 29 June 2012.
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester, H.C. Elman, A. Ramage
+
+x_it = diag(diag(schurparams.D))\x_it;

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/pde_control/schur_gmg.m

@@ -1,0 +1,25 @@
+function x_it=schur_gmg(x_it,schurparams)
+%SCHUR_GMG Schur complement preconditioner for Poisson control, using GMG
+%   x_it = schur_gmg(x_it,mparams);
+%   input
+%          x_it         operand for preconditioning operator
+%          mparams      structure defining block preconditioning operator
+%   output
+%          x_it         result of preconditioning operation
+%
+%   IFISS function: JWP, DJS, HCE; 29 June 2012.
+% Copyright (c) 2012 J.W. Pearson, D.J. Silvester, H.C. Elman, A. Ramage
+
+x1 = mg_iter(schurparams.mgdata,zeros(size(x_it)),x_it,schurparams.smooth_data,...
+	schurparams.nc,schurparams.npre,schurparams.npost,schurparams.sweeps);
+for j = 2:schurparams.no_gmg
+    x1 = mg_iter(schurparams.mgdata,x1,x_it,schurparams.smooth_data,...
+        schurparams.nc,schurparams.npre,schurparams.npost,schurparams.sweeps);
+end
+x2 = schurparams.M*x1;
+x_it = mg_iter(schurparams.mgdata,zeros(size(x_it)),x2,schurparams.smooth_data,...
+	schurparams.nc,schurparams.npre,schurparams.npost,schurparams.sweeps);
+for j = 2:schurparams.no_gmg
+    x_it = mg_iter(schurparams.mgdata,x_it,x2,schurparams.smooth_data,...
+        schurparams.nc,schurparams.npre,schurparams.npost,schurparams.sweeps);
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/setpath.m

@@ -1,0 +1,29 @@
+%SETPATH sets IFISS search path 
+%   IFISS scriptfile: DJS; 2 January 2011.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+
+% initilise matlab path for ./ifiss/ directories
+gohome, 
+warning('off', 'all')
+addpath(genpath(pwd),'-end')
+cd matlab704, rmpath(pwd), gohome
+cd matlabpre74, rmpath(pwd), gohome
+%fprintf('Resolving these path conflicts ... done')
+warning('on','all')
+%
+
+% fix iterapp bug (affects pcg & minres)
+if strncmp(version,'7.0.4',5), 
+cd matlab704, addpath(pwd), gohome
+end
+
+% include old version of ilu 
+if strncmp(version,'6.5',3) | strncmp(version,'7.0',3) | strncmp(version,'7.1.',4) | ...
+    strncmp(version,'7.2',3) | strncmp(version,'7.3',3), 
+cd matlabpre74, addpath(pwd), gohome
+end
+
+% fix MATLAB UMFPACK bug
+if strncmp(version,'6.5',3) | strncmp(version,'7.0',3), 
+spparms('default'),spparms('piv_tol',1), end
+

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/sirev/NS3SR_batch.m

@@ -1,0 +1,13 @@
+3     % problem
+1     % cavity type
+6     % grid parameter
+2     % uniform/stretched grid
+3     % discretisation
+0.005 % viscosity parameter
+3     % Picard/Newton/hybrid linearization
+2     % number of Picard iterations
+4     % number of Newton iterations
+1.d-8 % nonlinear tolerance
+2     % uniform/exponential streamlines
+
+%% Data file for test problem NS3

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/Contents.m

@@ -1,0 +1,100 @@
+% SOLVERS
+%
+% Files
+%   a_cdt               - matrix-vector product for scalar operator
+%   a_nst               - matrix-vector product for saddle-point operator
+%   amg_coarsen_plot    - function for visualisation of amg coarsening
+%   amg_grids_setup     - performs algebraic multigrid setup phase
+%   amg_smoother        - performs smoothing 
+%   amg_smoother_params - generates structure for AMG smoother parameters
+%   amg_smoother_setup  - generates smoother data for AMG
+%   amg_v_cycle         - performs one AMG V-cycle
+%   bicgstab_ell        - bicgstab(ell) for right-preconditioned iterates
+%   bicgstab_ell_r      - bicgstab(ell) iteration right preconditioning 
+%   bsolve              - solves Galerkin system using backslash
+%   cg_test             - CG convergence demo 
+%   givapp              - apply a sequence of Givens rotations
+%   gmres_r             - GMRES iteration with right preconditioning 
+%   helpme_it           - iterative solvers interactive help 
+%   helpme_mg           - geometric multigrid interactive help
+%   idrs_r              - Induced Dimension Reduction right preconditioning
+%   it_solve            - iterative solution of predefined steady problem
+%   m_amgzt             - AMG preconditioner for scalar operator
+%   m_amgzz             - AMG preconditioner with multiple v-cycles
+%   m_bfbt              - ideal LSC preconditioner (unscaled)
+%   m_diagt             - action of diagonal preconditioning operator
+%   m_fp                - ideal PCD preconditioner
+%   m_fp_amgz           - AMG iterated PCD preconditioner
+%   m_fp_mg             - GMG iterated PCD preconditioner
+%   m_ilut              - incomplete LU preconditioner
+%   m_masscheb          - mass matrix Chebyshev preconditioning operator
+%   m_massdiag          - mass matrix diagonal preconditioning operator
+%   m_mg                - GMG preconditioner for scalar problems
+%   m_nonet             - "no preconditioning" operator
+%   m_st_amgz           - AMG block preconditioner for Stokes equations
+%   m_st_block          - block preconditioner for Stokes equations
+%   m_st_mg             - block MG preconditioner for Stokes equations
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/it_solve.m

@@ -1,0 +1,663 @@
+%IT_SOLVE iterative solution of predefined steady problem
+%IFISS scriptfile: DJS; HCE; 13 October 2013.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+
+% Declare global variables for scalar and vector problems
+global amg_grid amg_smoother   
+global amg_gridA amg_smootherA amg_gridF amg_smootherF amg_gridAw amg_smootherAw
+
+if exist('pde','var')==0,
+   error('You need to set up a specific discrete problem first!'), 
+end
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+if pde==1,
+   %%% DIFFUSION PROBLEM
+   fprintf('discrete diffusion system ...\n')
+
+   % select Krylov subspace method
+   itmeth = default('PCG/MINRES? 1/2 (default PCG)',1);
+
+   % set parameters
+   tol = default('tolerance? (default 1e-6)',1e-6);
+   maxit = default('maximum number of iterations? (default 100)',100);
+   
+   % select preconditioner and construct it
+   fprintf('preconditioner:\n');
+   fprintf('   0  none\n');
+   fprintf('   1  diagonal\n');
+   fprintf('   2  incomplete cholesky\n');
+   fprintf('   3  geometric multigrid\n');
+   fprintf('   4  algebraic multigrid\n');
+   precon = default('default is AMG ',4);
+   if precon==0,     % none
+      M1=[]; M2=[];
+   elseif precon==1, % diagonal
+      D=diag(diag(Agal)); M1=sqrt(D); M2=M1; 
+   elseif precon==2, % incomplete Cholesky
+     M1 = ichol(Agal); M2=M1';
+%    M2 = cholinc(Agal,'0'); M1=M2'; fprintf('--- cholinc.m\n')%
+   elseif precon==3, % MG
+      if (domain==1 || domain==2), mg_diff,
+      else  error('MG for this problem/domain is not available!')
+      end
+      M1 = 'm_mg'; M2 = []; 
+      mparams=struct('mgdata',mgdata,'smooth_data',smooth_data, ...
+                     'nc',nc,'npre',npre,'npost',npost,'sweeps',sweeps);
+   elseif precon==4, % AMG
+ % uses global variables amg_grid amg_smoother
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/m_bcscale.m

@@ -1,0 +1,236 @@
+function bcweights = m_bcscale(xy,domain,qmethod)
+%M_BCSCALE   diagonal scaling for boundary-adjusted LSC preconditioner
+%   bcweights = m_bcscale(xy,domain,qmethod);
+%   input
+%          xy         grid coordinates
+%          domain     domain identifier
+%          qmethod    discretization identifier
+%   output
+%          bcweights  diagonal matrix of scalings used to deemphasize
+%                     tangential components of LSC preconditioner near
+%                     boundaries
+%
+%   IFISS function: HCE; 21 August 2012.
+% Copyright (c) 2012 D.J. Silvester, H.C. Elman, A. Ramage
+
+% Deemphasize tangential component near all boundaries
+% Identify indices at which diagonal scaling should be epsilon
+% For qmethod>1 (Q2 velocities):  nodes involved are on both element
+%                                 boundaries and interiors
+% For qmethod<1 (Q1 velocities):  nodes involved are only on element
+%                                 boundaries
+
+% For vertical boundaries, offset by nv
+nv = size(xy,1);
+
+%% Cavity
+if domain==1 || domain==7,
+    x_sorted = sort(xy(:,1)); 
+    xmin = x_sorted(1); xmax = x_sorted(length(x_sorted));
+    y_sorted = sort(xy(:,2)); 
+    ymin = y_sorted(1); ymax = y_sorted(length(y_sorted));
+% Near left boundary    
+   x1 = xmin;
+   x2_index = find(x_sorted>x1,1);
+   x2 = x_sorted(x2_index);
+   if qmethod <= 1,
+      left = nv + find( (xy(:,1)==x2) & (xy(:,2)>ymin) & (xy(:,2)<ymax) );
+   else
+      x3_index = find(x_sorted>x2,1);
+      x3 = x_sorted(x3_index);
+      left = nv + find( (xy(:,1)==x2 | xy(:,1)==x3) & ...
+                        (xy(:,2)>ymin) & (xy(:,2)<ymax) );
+   end        
+% Near top boundary
+   y1 = ymax;
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/m_st_mg.m

@@ -1,0 +1,21 @@
+function x_it=m_st_mg(x_it,mparams)
+%M_ST_MG block MG preconditioner for Stokes equations
+%   x_it = m_st_mg(x_it,mparams)
+%   input
+%          x_it         operand for preconditioning operator
+%          mparams      structure defining MG preconditioning operator
+%   output
+%          x_it         result of MG preconditioning operation
+%
+%   IFISS function: DJS, HCE; 9 April 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+nv=mparams.nv; np=size(mparams.Q,1); nu=nv/2;
+rvx=x_it(1:nu); rvy=x_it(nu+1:nv); rp=x_it(nv+1:nv+np);
+zvx=mg_iter(mparams.mgdata,zeros(size(rvx)),rvx,mparams.smooth_data,...
+            mparams.nc,mparams.npre,mparams.npost,mparams.sweeps);
+zvy=mg_iter(mparams.mgdata,zeros(size(rvy)),rvy,mparams.smooth_data, ...
+            mparams.nc,mparams.npre,mparams.npost,mparams.sweeps);
+%zp=Q\rp;
+zp=diag(mparams.Q).\rp;
+x_it = [zvx;zvy;zp];

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/m_sxbfbt_bc.m

@@ -1,0 +1,49 @@
+function y = m_sxbfbt_bc(x_it,aparams,mparams)
+%M_SXBFBT_BC  boundary adjusted ideal stabilized LSC preconditioner
+%   y = m_xbfbt_bc(x_it,aparams,mparams);
+%   input
+%          x_it         operand for preconditioning operator
+%          aparams      structure defining coefficient matrix
+%          mparams      structure defining preconditioning matrix
+%   output
+%          y            result of preconditioning operation
+%
+%   IFISS function: HCE; 26 September 2012.
+% Copyright (c) 2012 D.J. Silvester, H.C. Elman, A. Ramage
+
+nv = length(aparams.F);
+%nu = nv/2;
+np = size(aparams.B,1);
+Gdiag=spdiags(diag(mparams.G),0,nv,nv);
+
+sigma = mparams.viscosity;
+
+rv=x_it(1:nv); rp=x_it(nv+1:nv+np);
+
+%% pressure solve
+xB = (Gdiag\aparams.B')';
+BBt = aparams.B*xB';
+BBt = BBt.*(abs(BBt)>5.d-15);
+BCWeights = mparams.weights;
+weightedBBt = aparams.B*(BCWeights*xB');
+weightedBBt = weightedBBt.*(abs(weightedBBt)>5.d-15);
+weight = min(diag(mparams.weights));
+
+if mparams.domain==1,
+   n_null = mparams.n_null;
+   minor = [1:n_null-1,n_null+1:np]';
+   rp1 = zeros(np,1);
+   rp1(minor) = (weightedBBt(minor,minor)+weight*mparams.Cp1(minor,minor))\rp(minor);
+   rp2 =  xB*(aparams.F*(BCWeights*(xB'*rp1))) + weight*sigma*(mparams.Cp2*rp1);
+   zp = zeros(np,1);
+   zp(minor) = - (BBt(minor,minor)+mparams.Cp1(minor,minor))\rp2(minor);
+else
+   zp = (weightedBBt+weight*mparams.Cp1)\rp;
+   zp = -(BBt+mparams.Cp1) \ (xB*(aparams.F*(BCWeights*(xB'*zp))) + ...
+           weight*sigma*(mparams.Cp2*zp) );
+end 
+
+%% velocity solve
+rv = rv-(aparams.B')*zp;
+zv = aparams.F \ rv;
+y = [zv;zp];

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/m_xbfbt_mg.m

@@ -1,0 +1,46 @@
+function y = m_xbfbt_mg(x_it,aparams,mparams)
+%M_XBFBT_MG GMG iterated LSC preconditioner
+%   y = m_xbfbt_mg(x_it,aparams,mparams);
+%   input
+%          x_it         operand for preconditioning operator
+%          aparams      structure defining coefficient matrix
+%          mparams      structure defining preconditioning matrix
+%                       including B diag(G)\B' and mgdata
+%   output
+%          y            result of preconditioning operation
+%
+%   IFISS function: HCE; 12 April 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+nv = length(aparams.F);
+nu = nv/2;
+np = size(aparams.B,1);
+
+Gdiag=spdiags(diag(mparams.G),0,nv,nv);
+xB = (Gdiag\aparams.B')';
+%BBt = aparams.B*xB';
+
+rv=x_it(1:nv); rp=x_it(nv+1:nv+np);
+
+domain = mparams.domain;
+
+%% pressure solve
+mg_mat = 1;
+   
+%zp = -BBt \ (xB*(aparams.F*(xB'*(BBt\rp))));
+nc = size(mparams.mgdata_diff,2);
+BGiBt = mparams.mgdata_diff(nc).matrix;
+rp1 = mg_ns(BGiBt,rp, domain,mg_mat,mparams.mgdata_diff);
+rp2 =  xB*(aparams.F*(xB'*rp1));
+%zp = - BBt\rp2;
+zp = - mg_ns(BGiBt,rp2, domain,mg_mat,mparams.mgdata_diff);
+
+%% velocity solve
+mg_mat = 2;
+rv = rv-(aparams.B')*zp;
+%zvx = (aparams.F(1:nu,1:nu))\rv(1:nu);  
+zvx = mg_ns(aparams.F(1:nu,1:nu),rv(1:nu), domain,mg_mat,mparams.mgdata_cd);   
+%zvy = (aparams.F(1:nu,1:nu))\rv(nu+1:nv);
+zvy = mg_ns(aparams.F(1:nu,1:nu),rv(nu+1:nv), domain,mg_mat,mparams.mgdata_cd); 
+
+y = [zvx;zvy;zp]; 

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_apsetup_q1.m

@@ -1,0 +1,84 @@
+function Ap = mg_apsetup_q1(xy,xyp,mv,mp,domain)
+%MG_APSETUP_Q1 Q1 pressure diffusion matrix for GMG
+%   Ap = mg_apsetup_q1(xy,xyp,mv,mp,domain);
+%   input
+%          xy         Q2 nodal coordinate vector 
+%          xyp        Q1 element coordinate vector
+%          mv         Q2 element mapping matrix
+%          mp         Q1 element mapping matrix
+%          domain     domain index
+%   output
+%          Ap         Q1 scalar diffusion matrix
+%   IFISS function: HCE; 18 March 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+%   modelled after fpsetup_q1
+nngpt=4; 
+x=xy(:,1); y=xy(:,2);
+xp=xyp(:,1); yp=xyp(:,2);
+nvtx=length(x); nu=2*nvtx; np=length(xp);
+nel=length(mv(:,1));
+
+% Initialise global matrix
+Ap = sparse(np,np);
+
+% Gauss point integration rules
+if (nngpt==4)        % 2x2 Gauss points
+   gpt=1.0e0/sqrt(3.0e0);
+   s(1) = -gpt; t(1) = -gpt; wt(1)=1;
+   s(2) =  gpt; t(2) = -gpt; wt(2)=1;
+   s(3) =  gpt; t(3) =  gpt; wt(3)=1; 
+   s(4) = -gpt; t(4) =  gpt; wt(4)=1;
+elseif (nngpt==1)   % 1x1 Gauss point
+   s(1) =    0; t(1) =    0; wt(1)=4;
+else
+   error('Check Gauss point integration specification')
+end
+%
+% Inner loop over elements    
+for ivtx = 1:4
+   xl_v(:,ivtx) = x(mv(:,ivtx));
+   yl_v(:,ivtx) = y(mv(:,ivtx)); 
+end
+ae = zeros(nel,4,4);
+% 
+% Loop over Gauss points
+for igpt = 1:nngpt
+   sigpt=s(igpt);
+   tigpt=t(igpt);
+   wght=wt(igpt);
+% Evaluate derivatives etc
+   [jac,invjac,phi,dphidx,dphidy] = deriv(sigpt,tigpt,xl_v,yl_v);
+   for j = 1:4
+      for i = 1:4    
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_cavity_domain.m

@@ -1,0 +1,86 @@
+function [x,y,xy,mv,mp,bound] = mg_cavity_domain(nc,grid_type)
+%MG_CAVITY_DOMAIN square cavity Q2 grid generator for GMG
+%   [x,y,xy,mv,mp,bound] = mg_cavity_domain(nc,grid_type)
+%   input
+%          nc           grid level indicator
+%          grid_type    1 for uniform, 2 for stretched
+%   output
+%          grid-defining data
+%
+%   IFISS function: HCE; 24 January 2004.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+%   modelled after IFISS function cavity_domain, adapted for GMG
+n=2^nc; np=n/2; nq=n/4;
+% y-direction
+if grid_type==2
+   hmax=nc/(2^(nc+1));
+   x1=-1;x2=-2*hmax;x3=2*hmax;x4=1;nx1=2^(nc-1)-1;nx2=2;nx3=2^(nc-1)-1;
+   y1=-1;y2=-2*hmax;y3=2*hmax;y4=1;ny1=2^(nc-1)-1;ny2=2;ny3=2^(nc-1)-1;
+   y=subint(y1,y2,y3,y4,ny1,ny2,ny3);
+   stretch=(y(3)-y(2))/(y(2)-y(1));
+   x=y;
+else
+   yy=[1/np:1/np:1];
+   ypos=[0,yy];
+   yneg=-yy(length(yy):-1:1);
+   y=[yneg,ypos]';
+   x=y; 
+end
+%
+%% compute biquadratic element coordinates
+nvtx=(n+1)*(n+1);
+[X,Y]=meshgrid(x,y);
+xx=reshape(X',nvtx,1);
+yy=reshape(Y',nvtx,1);
+xy=[xx(:),yy(:)];
+%
+kx = 1;
+ky = 1;
+mel=0;
+for j=1:np
+   for i=1:np
+      mref=(n+1)*(ky-1)+kx;
+      pref=(np+1)*(j-1)+i;
+      mel=mel+1;
+      nvv(1) = mref;
+      nvv(2) = mref+2;
+      nvv(3) = mref+2*n+4;
+      nvv(4) = mref+2*n+2;
+      nvv(5) = mref+1;
+      nvv(6) = mref+n+3; 
+      nvv(7) = mref+2*n+3; 
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_diff_setup_step.m

@@ -1,0 +1,136 @@
+function Agal=step_mg_diff_setup(x,y)
+%MG_DIFF_SETUP_STEP GMG diffusion problem on step domain
+%   Agal=mg_diff_setup_step(x,y)
+%   input
+%          x       x coordinate vector for coarse grid
+%          y       y coordinate vector for coarse grid
+%   output
+%          Agal    discrete diffusion operator
+%
+%   IFISS function: AR; 20 January, 2003.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+n=length(y)-1; np=n/2; nq=n/4;
+nvtx= np*(np*1) + (5*np+1)*(n+1);
+% negative x-values
+xneg=x(1:n/2);yneg=y(1:n/2);
+xpos=x(n/2+1:end);ypos=y(n/2+1:end);
+[Xneg,Ypos]=meshgrid(xneg,ypos);
+xx=reshape(Xneg',np*(np+1),1);
+yy=reshape(Ypos',np*(np+1),1);
+xyleft=[xx(:),yy(:)];
+% assembly process
+kx = 1;
+ky = 1;
+mel=0;
+% loop over 2x2 macroelements
+for j=1:nq
+   for i=1:nq
+      mref=np*(ky-1)+kx;
+      pref=nq*(j-1)+i;
+      mel=mel+1;
+      nvv(1) = mref;
+      nvv(2) = mref+2;
+      nvv(3) = mref+2*np+2;
+      nvv(4) = mref+2*np;
+      nvv(5) = mref+1;
+      nvv(6) = mref+np+2; 
+      nvv(7) = mref+2*np+1; 
+      nvv(8)=  mref+np;
+      nvv(9)=  mref+np+1; 
+      npp(1) = pref;
+      npp(2) = pref+1;
+      npp(3) = pref+nq+1;
+      npp(4) = pref+nq;
+      mv(mel,1:9)=nvv(1:9);
+      mp(mel,1:4)=npp(1:4);
+      kx = kx + 2;
+   end
+   ky = ky + 2;
+   kx = 1;
+end
+%
+% correction along the internal boundary
+mref=2*np*(3*np+1)+1;
+pref=2*nq*(3*nq+1)+1;
+for mel=nq:nq:nq*nq;
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_ns.m

@@ -1,0 +1,86 @@
+function x_mg = mg_ns(Amat,f, domain,mg_mat,mgdata)
+%MG_NS GMG preconditioner for Navier-Stokes equations
+%   x_mg = mg_ns(Amat,f, domain,mg_mat,mgdata)
+%   input
+%          Amat         coefficient matrix
+%          f            right-hand side
+%          domain       physical domain, 1 for cavity, 3 for step
+%          mgmat        problem type, 1 for diffusion, 2 for
+%                       convection-diffusion 
+%          mgdata       structure containing all GMG data
+%   output
+%          x_mg            result of preconditioning operation
+%
+%   IFISS function: HCE; 17 March 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+if domain==1,   % Cavity
+% MG parameters
+   tau = 1.d-8;
+   maxit = 1;
+   npre = 1;
+   npost = 1;
+   if mg_mat==1      % Apply MG to Poisson equation
+      sweeps = 1;    % See mg_solve for explanation
+      np = sqrt(length(Amat));  
+      nc = log2(np-1);
+   elseif mg_mat==2, % Apply MG to convection-diffusion equation
+      sweeps = 4;    % See mg_solve for explanation
+      nuv = sqrt(length(Amat));
+      nc = log2(nuv-1);
+   else
+      error('Multigrid solver not implemented for this matrix')
+   end 
+elseif domain==3,  % Step
+   tau = 1.d-8;
+   maxit = 1;
+   sweeps = 1;
+   npre = 1;
+   npost = 1; 
+   nc = mgdata(size(mgdata,2)).nc;
+end
+
+% MG set-up
+A = mgdata(nc).matrix;
+x_mg = zeros(length(f),1);
+r = f;
+nr0 = norm(r);
+
+if nr0==0, return; end
+
+nr = nr0;
+its = 0;
+stats = [its,nr];
+
+% MG iterative loop
+flag=0;
+its=0;
+while (nr>tau*nr0)&(its<maxit)
+   c = mg_ns_iter(mgdata,zeros(size(r)),r,nc,npre,npost,sweeps); 
+   x_mg = x_mg+c;
+   r = f - A*x_mg;
+   nr = norm(r);
+   its = its+1;
+   stats = [stats;[its,nr]];
+end
+if its==(maxit)|isnan(nr)
+   flag=1;
+end
+
+look_at_mg = 0;
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_ns_cd.m

@@ -1,0 +1,76 @@
+function mgdata = mg_ns_cd(domain,viscosity,flowsol,outbnd)
+%MG_NS_CD GMG for CD problem  (Navier-Stokes)
+%   mgdata = mg_ns_cd(domain,viscosity,flowsol,outbnd)
+%   input
+%          domain       domain information, 1 for cavity, 3 for step
+%          viscosity    viscosity in NS equations
+%          flowsol      current velocity solution iterate
+%          outbnd       location of outflow boundary (for step)
+%   output
+%          mgdata       structure containing GMG data at all levels
+%             matrix    discrete convection-diffusion operators 
+%             prolong   grid transfer operators 
+%             smoother  structure containing smoothing operators
+%             nc        grid parameter identifying finest level
+%
+%   IFISS function: HCE; 27 May 2012.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+fprintf('Setting up MG data ...')
+
+if domain==1,   % Cavity
+   n  = sqrt(length(flowsol)/2);
+   nc = log2(n-1);
+   grid_type = 1;
+   
+% Smoothing strategy 
+   sweeps=4; smooth=2; stype=2;  % Means four (sweeps)-directional line (stype) GS (smooth)
+   
+%  Top level
+   %mgdata(nc).matrix=A;
+   [x,y,xy,mv,mp,bound] = mg_cavity_domain(nc,grid_type);        
+   mgdata(nc).matrix = mg_ns_cd_setup(x,y,viscosity,flowsol, domain,nc,nc,outbnd);
+   mgdata(nc).prolong=mg_prolong(2^nc,2^nc,x,y);
+   mgdata(nc).smoother = mg_ns_smooth(mgdata(nc).matrix,sweeps,smooth,stype);
+   mgdata(nc).nc = nc;  % Not used, saved to be consistent with domain=3
+
+%  x=x(1:2:end);y=y(1:2:end);
+   %%% loop over remaining levels
+   for level = nc-1:-1:1;
+      [x,y,xy,mv,mp,bound] = mg_cavity_domain(level,grid_type); 
+      mgdata(level).matrix=mg_ns_cd_setup(x,y,viscosity,flowsol, domain,level,nc,outbnd);
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_ns_diff.m

@@ -1,0 +1,105 @@
+function mgdata = mg_ns_diff(Ap,domain,flowsol,qmethod,outbnd)
+%MG_NS_DIFF GMG for diffusion problem (Navier-Stokes)
+%   mgdata = mg_ns_diff(Ap,domain,flowsol,qmethod,outbnd)
+%   input
+%          Ap           top level discrete diffusion operator
+%          domain       domain information, 1 for cavity, 3 for step
+%          flowsol      current velocity solution iterate
+%          qmethod      discretization method, 0 for Q1-Q1, 2 for Q2-Q1
+%          outbnd       location of outflow boundary (for step)
+%   output
+%          mgdata       structure containing GMG data at all levels
+%             matrix    discrete diffusion operators 
+%             prolong   grid transfer operators 
+%             smoother  structure containing smoothing operators
+%             nc        grid parameter identifying finest level
+%
+%   IFISS function: HCE; 18 March 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+if domain==1,      % Cavity
+   np = sqrt(length(Ap));
+   nc = log2(np-1);
+   grid_type = 1;
+   
+% Smoothing strategy 
+   sweeps=1; smooth=2; stype=2;  % Means one (sweeps)-directional line (stype) GS (smooth)
+   
+% Top level
+% Note (19 April 2004):  Fine level matrix created here is not pinned down
+   [x,y,xy,mv,mp,bound] = mg_cavity_domain(nc+1,grid_type);        
+   [x,y,xy,xyp] = mg_diff_q2q1grid(x,y,xy,mv,mp,bound); 
+   mgdata(nc).matrix = mg_apsetup_q1(xy,xyp,mv,mp,domain); 
+   if qmethod==2,
+      [x,y,xy,mv,mp,bound] = mg_cavity_domain(nc,grid_type);        
+      [x,y,xy,xyp] = mg_diff_q2q1grid(x,y,xy,mv,mp,bound);   
+      mgdata(nc).prolong = mg_prolong(2^nc,2^nc,x,y); 
+   else % qmethod=0
+      mgdata(nc).prolong=mg_prolong(2^nc,2^nc,x,y); 
+   end
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_ns_diff_lstsq.m

@@ -1,0 +1,121 @@
+function mgdata = mg_ns_diff_lstsq(BGiBt,domain,flowsol,qmethod,outbnd)
+%MG_NS_DIFF_LSTSQ GMG for diffusion (Navier-Stokes, least-squares)
+%   mgdata = mg_ns_diff_lstsq(BGiBt,domain,flowsol,qmethod,outbnd)
+%   input
+%          BGiBt        top level discrete diffusion operator
+%          domain       domain information, 1 for cavity, 3 for step
+%          flowsol      current velocity solution iterate
+%          qmethod      discretization method, 0 for Q1-Q1, 2 for Q2-Q1
+%          outbnd       location of outflow boundary (for step)
+%   output
+%          mgdata       structure containing GMG data at all levels
+%             matrix    discrete diffusion operators 
+%             prolong   grid transfer operators 
+%             smoother  structure containing smoothing operators
+%             nc        grid parameter identifying finest level
+%
+%   IFISS function: HCE; 18 March 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+if domain==1,      % Cavity
+   np = sqrt(length(BGiBt));
+   nc = log2(np-1);
+   grid_type = 1;
+   
+% Smoothing strategy 
+   sweeps=1; smooth=2; stype=2;  % Means one (sweeps)-directional line (stype) GS (smooth)
+   
+% Top level
+% 
+%  [x,y,xy,mv,mp,bound] = mg_cavity_domain(nc+1,grid_type);        
+%  [x,y,xy,xyp] = mg_diff_q2q1grid(x,y,xy,mv,mp,bound); 
+%  mgdata(nc).matrix = mg_apsetup_q1(xy,xyp,mv,mp,domain); 
+   mgdata(nc).matrix = BGiBt.*(BGiBt~=0);      % BGiBt;
+   if qmethod==2,
+      [x,y,xy,mv,mp,bound] = mg_cavity_domain(nc,grid_type);        
+      [x,y,xy,xyp] = mg_diff_q2q1grid(x,y,xy,mv,mp,bound);   
+      mgdata(nc).prolong = mg_prolong(2^nc,2^nc,x,y); 
+%  else % qmethod=0
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_ns_prolong_step.m

@@ -1,0 +1,42 @@
+function P=mg_ns_prolong_step(nc,x,y,outbnd)
+%MG_NS_PROLONG_STEP GMG prolongation for step domain (Navier-Stokes)
+%   P=mg_ns_prolong_step(nc,x,y,outbnd)
+%   input
+%          nc      grid parameter
+%          x       x coordinate vector for coarse grid
+%          y       y coordinate vector for coarse grid
+%          outbnd  location of outflow boundary
+%   output
+%          P       prolongation operator
+%
+%   IFISS function: HCE; 5 May 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+nelf=2^nc;nelc=nelf/2;nelcx=outbnd*nelc;
+nnf=(nelc+1)*((outbnd+1)*nelc+1)+nelc*(outbnd*nelc+1);
+nnc=(nelc/2+1)*(((outbnd+1)/2)*nelc+1)+(nelc/2)*(outbnd*nelc/2+1);
+
+% split coordinates 
+x1=x(1:nelc+1); x2=x(nelc+1:end);
+y1=y(nelc+1:end); y2=y;
+
+% block size
+nnf1=nelc*(nelc+1);
+nnc1=nelc/2*(nelc/2+1);
+
+start_col=(nelc/2+1)*(nelc/2+1);
+
+% Prolongation matrix has the form 
+% P = [P_1, 0, P_2]
+%     [   0 ,  P_3]
+
+% nonzero components of upper block, for upper left of step and coupling
+[P_1,P_2]=mg_ellblock(nelc,nelc,x1,y1);
+% P(1:nnf1,1:nnc1)=P_1;                % upper block
+% P(1:nnf1,nnc-start_col+1:nnc)=P_2;   % coupling
+
+% nonzero component of lower block
+P_3=mg_prolong(nelcx,nelf,x2,y2);
+% P(nnf1+1:end,nnc1+1:end)=P_3;
+
+P = [[P_1,sparse(nnf1,nnc-start_col-nnc1),P_2];[sparse(nnf-nnf1,nnc1),P_3]];

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_ns_q1cd.m

@@ -1,0 +1,166 @@
+function [a,n,r,epe,eph,epw,usol_c,vsol_c] = mg_ns_q1cd(xy,ev,flowsol,domain,lncoarse,lnfine,outbnd)
+%MG_NS_Q1CD CD matrix generator for GMG (Navier-Stokes)
+%   [a,n,r,epe,eph,epw,usol_c,vsol_c] = mg_ns_q1cd(xy,ev,flowsol,domain,lncoarse,lnfine,outbnd)  
+%   input
+%           xy         vertex coordinate vector  
+%           ev         element mapping matrix
+%           flowsol    current velocity field
+%           domain     domain parameter, 1 for square, 3 for step
+%           lncoarse   coarse grid index, log2(nc) for nc x nc square grid
+%           lnfine     finest grid index, log2(nf) for nf x nf square grid
+%           outbnd     location of outflow boundary
+%   output 
+%           a          discrete diffusion operator
+%           n          discrete convection operator
+%           r          mass matrix
+%           epe        viscosity normalised element peclet numbers 
+%           eph        flow specific element lengths 
+%           epw        centroid evaluated wind 
+%           usol_c     first convection coefficient on coarse grid nodes
+%           vsol_c     second convection coefficient on coarse grid nodes 
+%
+%   IFISS function: HCE; 18 March 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+% Modification of DJS routine mg_q1cd to allow convection coefficients to 
+% be passed via flowsol.  Analogous to femq1_cd.
+
+% Square domain
+if domain==1, 
+% Size of fine grid, obtained from underlying flow solution flowsol
+   nfine  = length(flowsol)/2;
+   nf = sqrt(nfine);
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_ns_smooth.m

@@ -1,0 +1,63 @@
+function Q = mg_ns_smooth(A,sweeps,smooth,stype)
+%MG_NS_SMOOTH smoothers for GMG iteration (Navier-Stokes)
+%   Q = mg_ns_smooth(A,sweeps,smooth,stype)
+%   input
+%          A         coefficient matrix
+%          sweeps    number of directions for multidirectional Gauss-Seidel 
+%          smooth    smoother (1, Jacobi, 2, Gauss-Seidel, 3 ILU)
+%          stype     type of smoother (1 point, 2 line)
+%   output
+%          Q         structure, smoothing operators in factored form
+%
+%   IFISS function: HCE; 18 March 2005.
+%            update: AR; 30 July 2008.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage
+
+N=size(A,1);
+n = sqrt(size(A,1));
+if stype==2
+   % line Gauss-Seidel
+   Q1 = tril(A,1);
+   [L1,U1] = lu(Q1);
+   if sweeps>=2
+      Q2 = diag(diag(A,0)) + diag(diag(A,-n),-n) + diag(diag(A,n),n) ...
+         + diag(diag(A,-n-1),-n-1) + diag( diag(A,-1),-1)        ...
+         + diag(diag(A,n-1),n-1);
+      [L2,U2] = lu(Q2);
+      if sweeps>=3
+         Q3 = triu(A,-1);
+         [L3,U3] = lu(Q3);
+      if sweeps==4
+         Q4 = diag(diag(A,0)) + diag(diag(A,-n),-n) + diag(diag(A,n),n) ...
+            + diag(diag(A,n+1),n+1) + diag( diag(A,1),1)            ...
+            + diag(diag(A,-n+1),-n+1);
+         [L4,U4] = lu(Q4);
+         else
+            L4=sparse(N,N); U4=L4;
+         end
+      else
+         L3=sparse(N,N); U3=L3; L4=L3; U4=L3;
+      end
+   else
+      L2=sparse(N,N); U2=L2; L3=L2; U3=L2; L4=L2; U4=L2;
+   end
+else
+   % point smoothers
+   if smooth==3,
+      % ILU
+      setup.type='nofill';
+      [L1,U1]=ilu(A,setup);
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_prolong_step.m

@@ -1,0 +1,41 @@
+function P=mg_prolong_step(nc,x,y)
+%MG_PROLONG_STEP GMG prolongation operator for step domain
+%   P = mg_prolong_step(nc,x,y)
+%   input
+%          nc      grid level 
+%          x       x coordinate vector for coarse grid
+%          y       y coordinate vector for coarse grid
+%   output
+%          P       prolongation operator
+%
+%   IFISS function: AR, HCE; 05 May, 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+nelf=2^nc;nelc=nelf/2;nelcx=5*nelc;
+nnf=(nelc+1)*(6*nelc+1)+nelc*(5*nelc+1);
+nnc=(nelc/2+1)*(3*nelc+1)+(nelc/2)*(5*nelc/2+1);
+
+% split coordinates 
+x1=x(1:nelc+1); x2=x(nelc+1:end);
+y1=y(nelc+1:end); y2=y;
+
+% block size
+nnf1=nelc*(nelc+1);
+nnc1=nelc/2*(nelc/2+1);
+
+start_col=(nelc/2+1)*(nelc/2+1);
+
+% Prolongation matrix has the form 
+% P = [P_1, 0, P_2]
+%     [   0 ,  P_3]
+
+% nonzero components of upper block, for upper left of step and coupling
+[P_1,P_2]=mg_ellblock(nelc,nelc,x1,y1);
+% P(1:nnf1,1:nnc1)=P_1;                % upper block
+% P(1:nnf1,nnc-start_col+1:nnc)=P_2;   % coupling
+
+% nonzero component of lower block
+P_3=mg_prolong(nelcx,nelf,x2,y2);
+% P(nnf1+1:end,nnc1+1:end)=P_3;
+
+P = [[P_1,sparse(nnf1,nnc-start_col-nnc1),P_2];[sparse(nnf-nnf1,nnc1),P_3]];

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/solvers/mg_step_domain.m

@@ -1,0 +1,151 @@
+function [x,y,xy,mv,mp,bound] = mg_step_domain(nc,outbnd)
+%MG_STEP_DOMAIN Q2 grid generator for GMG on step domain
+%   [x,y,xy,mv,mp,bound] = mg_step_domain(nc,outbnd)
+%   input
+%          nc           grid level indicator
+%          outbnd       location of outflow boundary
+%   output
+%          grid-defining data
+%
+%   IFISS function: HCE; 24 January 2004.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage
+
+%   modelled after IFISS function step_domain, adapted for GMG
+stretch=1;
+n=2^nc; np=n/2; nq=n/4;
+ny=n;nx=(outbnd+1)*n/2; 
+%
+%% compute (x,y) coordinates of vertices
+% y-direction
+if stretch >1
+   dy=(1-stretch)/(1-stretch^np);
+   yy(1)=dy;
+   for k=1:np-1, yy(k+1)=yy(k)+dy*stretch^k; end
+   yy(np)=1;
+else
+   yy=[1/np:1/np:1];
+end
+ypos=[0,yy];
+yneg=-yy(length(yy):-1:1);
+y=[yneg,ypos];
+% x-direction
+xx=yneg;
+xneg=[xx];
+xpos=[0:1/np:outbnd];
+x=[xneg,xpos];
+%                  
+%% compute biquadratic element coordinates
+nvtx= np*(np*1) + ((outbnd+1)*np+1)*(n+1);
+% negative x-values
+[Xneg,Ypos]=meshgrid(xneg,ypos);
+xx=reshape(Xneg',np*(np+1),1);
+yy=reshape(Ypos',np*(np+1),1);
+xyleft=[xx(:),yy(:)];
+kx = 1;
+ky = 1;
+mel=0;
+% symbolic assembly 
+% loop over 2x2 macroelements
+for j=1:nq
+   for i=1:nq
+      mref=np*(ky-1)+kx;
+      pref=nq*(j-1)+i;
+      mel=mel+1;
+      nvv(1) = mref;
+      nvv(2) = mref+2;
+      nvv(3) = mref+2*np+2;
+      nvv(4) = mref+2*np;
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/Contents.m

@@ -1,0 +1,23 @@
+% SPECIFIC_FILES
+%
+% Files
+%   install_pc                       - sets up IFISS on non-UNIX computer
+%   install_unix                     - sets up IFISS on UNIX computer
+%   cd_testproblem_pc                - sets up Examples 3.1 to 3.4 
+%   cd_testproblem_unix              - sets up Examples 3.1 to 3.4 
+%   diff_testproblem_pc              - sets up Examples 1.1 to 1.4
+%   diff_testproblem_unix            - sets up Examples 1.1 to 1.4
+%   heat_testproblem_pc              - sets up Examples H1 to H3
+%   heat_testproblem_unix            - sets up Examples H1 to H3
+%   navier_testproblem_pc            - sets up Examples 7.1 to 7.4
+%   navier_testproblem_unix          - sets up Examples 7.1 to 7.4
+%   setpath_pc                       - sets IFISS search path (PC)
+%   stokes_testproblem_pc            - sets up Examples 5.1 to 5.4
+%   setpath_unix                     - sets IFISS search path (Unix)
+%   stokes_testproblem_unix          - sets up Examples 5.1 to 5.4
+%   unsteady_bouss_testproblem_unix  - sets up Boussinesq flow problem
+%   unsteady_bouss_testproblem_pc    - sets up Boussinesq flow problem
+%   unsteady_cd_testproblem_pc       - sets up Examples T-CD1 to T-CD4
+%   unsteady_cd_testproblem_unix     - sets up Examples T-CD1 to T-CD4
+%   unsteady_navier_testproblem_pc   - sets up Examples T-NS1 to T-NS5
+%   unsteady_navier_testproblem_unix - sets up Examples T-NS1 to T-NS5

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/cd_testproblem_pc.m

@@ -1,0 +1,32 @@
+%CD_TESTPROBLEM_PC sets up Examples 3.1 to 3.4 
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables, clear functions
+fprintf('\nspecification of reference convection-diffusion problem.\n')
+fprintf('\nchoose specific example');
+fprintf('\n     1  Constant vertical wind')
+fprintf('\n     2  Vertical wind, characteristic layers')
+fprintf('\n     3  Constant wind @ 30 degree angle')
+fprintf('\n     4  Recirculating wind\n');
+sn = default('',1);
+
+if sn==3 
+system('copy .\convection\test_problems\constant_wind.m .\convection\specific_wind.m');
+system('copy .\convection\test_problems\test_bc.m .\diffusion\specific_bc.m');
+   square_cd, solve_cd
+elseif sn==4
+system('copy .\convection\test_problems\circular_wind.m .\convection\specific_wind.m');
+system('copy .\convection\test_problems\hotwall_bc.m .\diffusion\specific_bc.m');
+   square_cd, solve_cd
+elseif sn==1
+system('copy .\convection\test_problems\vertical_wind.m .\convection\specific_wind.m');
+system('copy .\convection\test_problems\solution1_bc.m .\diffusion\specific_bc.m');
+   ref_cd, solve_cd
+elseif sn==2
+system('copy .\convection\test_problems\ref_wind.m .\convection\specific_wind.m');
+system('copy .\convection\test_problems\ref_bc.m .\diffusion\specific_bc.m');
+   ref_cd, solve_cd
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/cd_testproblem_unix.m

@@ -1,0 +1,32 @@
+%CD_TESTPROBLEM_UNIX sets up Examples 3.1 to 3.4 
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference convection-diffusion problem.\n')
+fprintf('\nchoose specific example');
+fprintf('\n     1  Constant vertical wind')
+fprintf('\n     2  Vertical wind, characteristic layers')
+fprintf('\n     3  Constant wind @ 30 degree angle')
+fprintf('\n     4  Recirculating wind\n');
+sn = default('',1);
+
+if sn==3 
+system('/bin/cp ./convection/test_problems/constant_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/test_bc.m ./diffusion/specific_bc.m');
+   square_cd, solve_cd
+elseif sn==4
+system('/bin/cp ./convection/test_problems/circular_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/hotwall_bc.m ./diffusion/specific_bc.m');
+   square_cd, solve_cd
+elseif sn==1
+system('/bin/cp ./convection/test_problems/vertical_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/solution1_bc.m ./diffusion/specific_bc.m');
+   ref_cd, solve_cd
+elseif sn==2
+system('/bin/cp ./convection/test_problems/ref_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/ref_bc.m ./diffusion/specific_bc.m');
+   ref_cd, solve_cd
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/it_navier_testproblem_pc.m

@@ -1,0 +1,28 @@
+%it_navier_testproblem    driver for iterative solution of Examples 7.1 to 7.4
+%   IFISS scriptfile: HCE 4 June 2013, from DJS; 9 May 2012.
+% Copyright (c) 2013 D.J. Silvester, H.C. Elman, A. Ramage
+gohome
+clear variables, clear functions
+fprintf('\nspecification of reference Navier-Stokes problem.\n');
+fprintf('\nchoose specific example (default is cavity)');
+fprintf('\n     2  Flow over a backward facing step');
+fprintf('\n     3  Lid driven cavity\n');
+sn = default('',3);
+if sn==2,
+system('copy .\stokes_flow\test_problems\backwardstep_flow.m .\stokes_flow\specific_flow.m');
+system('copy .\stokes_flow\test_problems\backwardstep_bc.m .\stokes_flow\stream_bc.m');
+longstep_stokes; it_solve_step_navier;
+elseif sn==3,
+lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3);
+if lid_model ==3,
+system('copy .\stokes_flow\test_problems\regcavity_flow.m .\stokes_flow\specific_flow.m');
+elseif lid_model ==2,
+system('copy .\stokes_flow\test_problems\tightcavity_flow.m .\stokes_flow\specific_flow.m');
+else
+system('copy .\stokes_flow\test_problems\leakycavity_flow.m .\stokes_flow\specific_flow.m');
+end
+system('copy .\stokes_flow\test_problems\zero_bc.m .\stokes_flow\stream_bc.m');
+square_stokes; it_solve_navier;
+else
+error('reference problem datafile not found!');
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/navier_testproblem_pc.m

@@ -1,0 +1,43 @@
+%NAVIER_TESTPROBLEM_PC sets up Examples 7.1 to 7.4
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables, clear functions
+fprintf('\nspecification of reference Navier-Stokes problem.\n');
+fprintf('\nchoose specific example (default is cavity)');
+fprintf('\n     1  Channel domain');
+fprintf('\n     2  Flow over a backward facing step');
+fprintf('\n     3  Lid driven cavity');
+fprintf('\n     4  Flow over a plate');
+fprintf('\n     5  Flow over an obstacle\n');
+sn = default('',3);
+if sn==1,
+system('copy .\stokes_flow\test_problems\poiseuille_flow.m .\stokes_flow\specific_flow.m');
+system('copy .\stokes_flow\test_problems\poiseuille_bc.m .\stokes_flow\stream_bc.m');
+   channel_stokes; solve_navier;
+elseif sn==2,
+system('copy .\stokes_flow\test_problems\backwardstep_flow.m .\stokes_flow\specific_flow.m');
+system('copy .\stokes_flow\test_problems\backwardstep_bc.m .\stokes_flow\stream_bc.m');
+   longstep_stokes; solve_step_navier;
+elseif sn==3,
+   lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3); 
+   if lid_model ==3,
+system('copy .\stokes_flow\test_problems\regcavity_flow.m .\stokes_flow\specific_flow.m');
+   elseif lid_model ==2,
+system('copy .\stokes_flow\test_problems\tightcavity_flow.m .\stokes_flow\specific_flow.m');
+   else
+system('copy .\stokes_flow\test_problems\leakycavity_flow.m .\stokes_flow\specific_flow.m');
+   end
+system('copy .\stokes_flow\test_problems\zero_bc.m .\stokes_flow\stream_bc.m');
+   square_stokes; solve_navier;
+elseif sn==4,
+system('copy .\stokes_flow\test_problems\plate_flow.m .\stokes_flow\specific_flow.m');
+system('copy .\stokes_flow\test_problems\plate_bc.m .\stokes_flow\stream_bc.m');
+   plate_stokes; solve_plate_navier;
+elseif sn==5,
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/navier_testproblem_unix.m

@@ -1,0 +1,43 @@
+%NAVIER_TESTPROBLEM_UNIX sets up Examples 7.1 to 7.4
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference Navier-Stokes problem.\n');
+fprintf('\nchoose specific example (default is cavity)');
+fprintf('\n     1  Channel domain');
+fprintf('\n     2  Flow over a backward facing step');
+fprintf('\n     3  Lid driven cavity');
+fprintf('\n     4  Flow over a plate');
+fprintf('\n     5  Flow over an obstacle\n');
+sn = default('',3);
+if sn==1,
+system('/bin/cp ./stokes_flow/test_problems/poiseuille_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/poiseuille_bc.m ./stokes_flow/stream_bc.m');
+   channel_stokes; solve_navier;
+elseif sn==2,
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_bc.m ./stokes_flow/stream_bc.m');
+   longstep_stokes; solve_step_navier;
+elseif sn==3,
+   lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3); 
+   if lid_model ==3,
+system('/bin/cp ./stokes_flow/test_problems/regcavity_flow.m ./stokes_flow/specific_flow.m');
+   elseif lid_model ==2,
+system('/bin/cp ./stokes_flow/test_problems/tightcavity_flow.m ./stokes_flow/specific_flow.m');
+   else
+system('/bin/cp ./stokes_flow/test_problems/leakycavity_flow.m ./stokes_flow/specific_flow.m');
+   end
+system('/bin/cp ./stokes_flow/test_problems/zero_bc.m ./stokes_flow/stream_bc.m');
+   square_stokes; solve_navier;
+elseif sn==4,
+system('/bin/cp ./stokes_flow/test_problems/plate_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/plate_bc.m ./stokes_flow/stream_bc.m');
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/setpath_pc.m

@@ -1,0 +1,59 @@
+%SETPATH_PC sets IFISS search path (PC)
+%   IFISS scriptfile: DJS; 3 May 2012.
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+
+% initilises matlab path for ./ifiss/ directories
+gohome, home=pwd;
+P=path;
+path(P,[home]);
+P=path;
+path(P,[home,'\datafiles']);
+P=path;
+path(P,[home,'\grids']);
+P=path;
+path(P,[home,'\graphs']);
+P=path;
+path(P,[home,'\diffusion']);
+P=path;
+path(P,[home,'\diffusion\test_problems']);
+P=path;
+path(P,[home,'\convection']);
+P=path;
+path(P,[home,'\convection\test_problems']);
+P=path;
+path(P,[home,'\stokes_flow']);
+P=path;
+path(P,[home,'\stokes_flow\test_problems']);
+P=path;
+path(P,[home,'\navier_flow']);
+P=path;
+path(P,[home,'\navier_flow\test_problems']);
+P=path;
+path(P,[home,'\solvers']);
+P=path;
+path(P,[home,'\solvers\ch4_code']);
+P=path;
+path(P,[home,'\timestepping']);
+P=path;
+path(P,[home,'\timestepping\test_problems']);
+if exist('djs') == 7
+   P=path; path(P,[home,'\djs']);
+end
+if exist('hce') == 7
+   P=path; path(P,[home,'\hce']);
+end
+if exist('ar') == 7
+   P=path; path(P,[home,'\ar']);
+end
+if exist('demo') == 7
+   P=path; path(P,[home,'\demo']);
+end
+%
+% fix UMFPACK bug
+if strncmp(version,'6.5',3) | strncmp(version,'7.0',3), 
+   spparms('default'),spparms('piv_tol',1), end
+% fix iterapp bug (affects pcg & minres)
+if strncmp(version,'7.0.4',5), 
+   P=path; path([home,'\matlab704'],P);
+end
+clear P

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/stokes_testproblem_unix.m

@@ -1,0 +1,38 @@
+%STOKES_TESTPROBLEM_UNIX sets up Examples 5.1 to 5.4
+%   IFISS scriptfile: DJS; 9 May 2012.  
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference Stokes problem.\n')
+fprintf('\nchoose specific example (default is cavity)');
+fprintf('\n     1  Channel domain')
+fprintf('\n     2  Flow over a backward facing step')
+fprintf('\n     3  Lid driven cavity')
+fprintf('\n     4  Colliding flow\n');
+sn = default('',3);
+if sn==1, 
+system('/bin/cp ./stokes_flow/test_problems/poiseuille_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/poiseuille_bc.m ./stokes_flow/stream_bc.m');
+   channel_stokes, solve_stokes  
+elseif sn==2 
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_bc.m ./stokes_flow/stream_bc.m');
+   longstep_stokes, solve_step_stokes  
+elseif sn==3
+   lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3); 
+   if lid_model ==3    
+system('/bin/cp ./stokes_flow/test_problems/regcavity_flow.m ./stokes_flow/specific_flow.m');
+   elseif lid_model ==2    
+system('/bin/cp ./stokes_flow/test_problems/tightcavity_flow.m ./stokes_flow/specific_flow.m');
+   else
+system('/bin/cp ./stokes_flow/test_problems/leakycavity_flow.m ./stokes_flow/specific_flow.m');
+   end
+system('/bin/cp ./stokes_flow/test_problems/zero_bc.m ./stokes_flow/stream_bc.m');
+   square_stokes, solve_stokes  
+elseif sn==4
+system('/bin/cp ./stokes_flow/test_problems/collide_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/collide_bc.m ./stokes_flow/stream_bc.m');
+   square_stokes, solve_stokes  
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/unsteady_navier_testproblem_pc.m

@@ -1,0 +1,33 @@
+%UNSTEADY_NAVIER_TESTPROBLEM_PC sets up Examples T-NS1 to T-NS5
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables, clear functions
+fprintf('\nspecification of reference unsteady Navier-Stokes problem.\n')
+fprintf('\nchoose specific example (default is step)');
+fprintf('\n     2  Flow over a backward facing step')
+fprintf('\n     3  Lid driven cavity');
+fprintf('\n     5  Flow around a square obstruction\n');
+sn = default('',2);
+if  sn==2
+system('copy .\stokes_flow\test_problems\backwardstep_flow.m .\stokes_flow\specific_flow.m');
+system('copy .\stokes_flow\test_problems\backwardstep_bc.m .\stokes_flow\stream_bc.m');
+   longstep_stokes, unsteady_step_navier
+elseif sn==3
+   lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3); 
+   if lid_model ==3    
+system('copy .\stokes_flow\test_problems\regcavity_flow.m .\stokes_flow\specific_flow.m');
+   elseif lid_model ==2    
+system('copy .\stokes_flow\test_problems\tightcavity_flow.m .\stokes_flow\specific_flow.m');
+   else
+system('copy .\stokes_flow\test_problems\leakycavity_flow.m .\stokes_flow\specific_flow.m');
+   end
+system('copy .\stokes_flow\test_problems\zero_bc.m .\stokes_flow\stream_bc.m');
+   square_stokes, unsteady_navier
+   elseif sn==5
+system('copy .\stokes_flow\test_problems\obstacle_flow.m .\stokes_flow\specific_flow.m');
+system('copy .\stokes_flow\test_problems\obstacle_bc.m .\stokes_flow\stream_bc.m');
+    obstacle_stokes, unsteady_obstacle_navier
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/specific_files/unsteady_navier_testproblem_unix.m

@@ -1,0 +1,33 @@
+%UNSTEADY_NAVIER_TESTPROBLEM_UNIX sets up Examples T-NS1 to T-NS5
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference unsteady Navier-Stokes problem.\n')
+fprintf('\nchoose specific example (default is step)');
+fprintf('\n     2  Flow over a backward facing step')
+fprintf('\n     3  Lid driven cavity');
+fprintf('\n     5  Flow around a square obstruction\n');
+sn = default('',2);
+if  sn==2
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_bc.m ./stokes_flow/stream_bc.m');
+   longstep_stokes, unsteady_step_navier
+elseif sn==3
+   lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3); 
+   if lid_model ==3    
+system('/bin/cp ./stokes_flow/test_problems/regcavity_flow.m ./stokes_flow/specific_flow.m');
+   elseif lid_model ==2    
+system('/bin/cp ./stokes_flow/test_problems/tightcavity_flow.m ./stokes_flow/specific_flow.m');
+   else
+system('/bin/cp ./stokes_flow/test_problems/leakycavity_flow.m ./stokes_flow/specific_flow.m');
+   end
+system('/bin/cp ./stokes_flow/test_problems/zero_bc.m ./stokes_flow/stream_bc.m');
+   square_stokes, unsteady_navier
+   elseif sn==5
+system('/bin/cp ./stokes_flow/test_problems/obstacle_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/obstacle_bc.m ./stokes_flow/stream_bc.m');
+    obstacle_stokes, unsteady_obstacle_navier
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/stokes_flow/Contents.m

@@ -1,0 +1,44 @@
+% STOKES_FLOW
+%
+% Files
+%   altinfsup          - inf-sup eigenvalue distribution of potential flow
+%   box_stokes         - set up flow problem in box domain 
+%   channel_stokes     - set up inflow/outflow problem in square domain
+%   flowbc             - imposes inflow boundary condition 
+%   forwardstep_stokes - set up flow problem in forward step domain 
+%   helpme_stokes      - Stokes flow problem interactive help 
+%   infsup             - computes inf-sup eigenvalue distribution 
+%   localbc_xy         - imposes vector BC for Poisson error estimator 
+%   longstep_stokes    - set up flow problem in extended step domain 
+%   obstacle_stokes    - set up flow problem in domain with square obstacle
+%   out_Neumann_bound  - locates elements on the step outflow boundary
+%   plate_stokes       - set up flow problem in slit domain 
+%   q1div              - computes norm of divergence of Q1 flow solution
+%   q2div              - computes norm of divergence of Q2 flow solution
+%   q2divx             - divergence of Q2 flow solution (no screen output)
+%   quad_stokes        - set up Stokes problem in quadrilateral domain 
+%   solve_step_stokes  - solve Stokes problem in step domain
+%   solve_stokes       - solve Stokes problem in square domain
+%   specific_flow      - (current) problem flow boundary condition
+%   square_stokes      - set up flow problem in unit square domain 
+%   step_stokes        - set up flow problem in standard step domain 
+%   stokes_q1p0        - Q1-P0 matrix generator
+%   stokes_q1q1        - Q1-Q1 matrix generator
+%   stokes_q2p1        - Q2-P1 matrix generator
+%   stokes_q2q1        - Q2-Q1 matrix generator
+%   stokespost         - estimates Stokes error distribution 
+%   stokespost_q1p0_bc - postprocesses Poisson error estimator 
+%   stokespost_q1p0_p  - computes Poisson error estimator for Q1-P0 
+%   stokespost_q1q1_bc - postprocesses Poisson error estimator 
+%   stokespost_q1q1_p  - computes Poisson error estimator for Q1-Q1 
+%   stokespost_q2p1    - vectorized Q2-P1 local Poisson error estimator 
+%   stokespost_q2q1    - vectorized Q2-Q1 local Poisson error estimator 
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/stokes_flow/channel_stokes.m

@@ -1,0 +1,45 @@
+%CHANNEL_STOKES set up inflow/outflow problem in square domain
+%   IFISS scriptfile: DJS; 27 May 2012. 
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+clear variables
+%% define geometry
+pde=3; domain=10; enclosed=0;
+channel_domain,
+load channel_grid.mat
+%
+%% set up matrices
+q_in=default('Q1-Q1/Q1-P0/Q2-Q1/Q2-P1: 1/2/3/4? (default Q1-P0)',2);
+qmethod=q_in-1;
+if qmethod==2,
+   [x,y,xy,xyp,mp,map] = q2q1gridx(x,y,xy,mv,bound);
+   [A,B,Q,G,Bx,By,f,g] = stokes_q2q1(xy,xyp,mv,mp);
+elseif qmethod==3,
+   [x,y,xy,xyp,ee] = q2p1grid(x,y,xy,mv,bound);
+   [A,B,Q,G,Bx,By,f,g] = stokes_q2p1(xy,xyp,mv);
+elseif qmethod==8,
+   [x,y,xy,xyp] = q2p1grid(x,y,xy,mv,bound);
+   [A,BB,QQ,G,Bx,By,f,gg] = stokes_q2p1(xy,xyp,mv);
+   nnp=length(gg); ppk=[1:3:nnp];
+   B=BB(ppk,:); Q=QQ(ppk,ppk); g=gg(ppk);
+   fprintf('Reduction to Q2-P0 Stokes system done.\n')
+elseif qmethod==0, 
+   [ev,ee,ebound,xyp] = q1q1grid(x,y,xy,mv,bound,mbound);
+   [A,B,Q,C,G,Bx,By,f,g] = stokes_q1q1(xy,ev);
+elseif qmethod==1 
+   [ev,ee,ebound,xyp] = q1p0grid(x,y,xy,mv,bound,mbound);
+   [A,B,Q,C,G,Bx,By,f,g] = stokes_q1p0(xy,xyp,mv,ev);
+end
+gohome
+cd datafiles
+save square_stokes_nobc.mat pde domain qmethod grid_type A B Q f g xy xyp mbound bound x y 
+save square_stokes_nobc.mat Bx By bndxy bnde obs -append
+if qmethod==1 
+   save square_stokes_nobc.mat C G ev ee ebound -append
+elseif qmethod==0 
+   save square_stokes_nobc.mat C G ev ee ebound mv enclosed -append
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/stokes_flow/q1div.m

@@ -1,0 +1,53 @@
+function ediv = q1div(xy,ev,flowsol)
+%Q1DIV computes norm of divergence of Q1 flow solution
+%   ediv = q1div(xy,ev,flowsol);
+%   input
+%          xy         vertex coordinate vector  
+%          ev         element mapping matrix
+%          flowsol    Q1 flow solution vector
+%
+%   IFISS function: DJS; 8 March 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+      x=xy(:,1); y=xy(:,2); nvtx=length(x);
+      nel=length(ev(:,1)); 
+	  usol=flowsol(1:nvtx); vsol=flowsol(nvtx+1:2*nvtx); 
+      fprintf('computing divergence of discrete velocity solution ...  ')
+
+%
+% initialise global matrices
+       ediv=zeros(nel,1);
+%
+% set up 2x2 Gauss points
+      gpt=1.0e0/sqrt(3.0e0);
+      s(1) = -gpt;  t(1) = -gpt;
+      s(2) =  gpt;  t(2) = -gpt;
+      s(3) =  gpt;  t(3) =  gpt;
+      s(4) = -gpt;  t(4) =  gpt;
+%
+% inner loop over elements    
+        for ivtx = 1:4
+        xl_v(:,ivtx) = x(ev(:,ivtx));
+        yl_v(:,ivtx) = y(ev(:,ivtx)); 
+        xsl(:,ivtx) = usol(ev(:,ivtx));
+		ysl(:,ivtx) = vsol(ev(:,ivtx));
+		end
+% loop over 2x2 Gauss points
+         for igpt = 1:4
+         sigpt=s(igpt);
+         tigpt=t(igpt);
+%  evaluate derivatives 
+         [jac,invjac,phi,dphidx,dphidy] = deriv(sigpt,tigpt,xl_v,yl_v);
+		 evec=zeros(nel,1);
+            for j=1:4
+		    evec(:) = evec(:) + xsl(:,j) .* dphidx(:,j) + ysl(:,j) .* dphidy(:,j);
+		    end
+	     ediv(:)=ediv(:) + evec(:).*evec(:);
+% end of Gauss point loop
+         end
+%
+% end of element loop
+%
+      err_div = sqrt(sum(ediv)); ediv = sqrt(ediv);
+fprintf('done\n')
+fprintf('estimated velocity divergence error:  %10.6e \n',err_div) 
+return

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/stokes_flow/q2div.m

@@ -1,0 +1,63 @@
+function ediv = q2div(xy,mv,flowsol)
+%Q2DIV computes norm of divergence of Q2 flow solution
+%   ediv = q2div(xy,mv,flowsol);
+%   input
+%          xy         vertex coordinate vector  
+%          mv         Q2 element mapping matrix
+%          flowsol    Q2 solution vector
+%
+%   IFISS function: DJS; 6 March 2005.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+      x=xy(:,1); y=xy(:,2); nvtx=length(x);
+      nel=length(mv(:,1)); 
+	  usol=flowsol(1:nvtx); vsol=flowsol(nvtx+1:2*nvtx);
+      fprintf('computing divergence of discrete velocity solution ...  ')
+
+%
+% initialise global matrices
+       ediv=zeros(nel,1);
+%
+% set up 3x3 Gauss points
+      gpt=sqrt(0.6); 
+      s(1) = -gpt; t(1) = -gpt; wt(1)=25/81;
+      s(2) =  gpt; t(2) = -gpt; wt(2)=25/81;
+      s(3) =  gpt; t(3) =  gpt; wt(3)=25/81; 
+      s(4) = -gpt; t(4) =  gpt; wt(4)=25/81;
+      s(5) =  0.0; t(5) = -gpt; wt(5)=40/81;
+      s(6) =  gpt; t(6) =  0.0; wt(6)=40/81;
+      s(7) =  0.0; t(7) =  gpt; wt(7)=40/81; 
+      s(8) = -gpt; t(8) =  0.0; wt(8)=40/81;
+      s(9) =  0.0; t(9) =  0.0; wt(9)=64/81;
+%
+%
+% inner loop over elements    
+        for ivtx = 1:4
+        xl_v(:,ivtx) = x(mv(:,ivtx));
+        yl_v(:,ivtx) = y(mv(:,ivtx)); 
+        end
+        for idx = 1:9		
+		xsl(:,idx) = usol(mv(:,idx));
+		ysl(:,idx) = vsol(mv(:,idx));
+		end
+% loop over 3x3 Gauss points
+        for igpt = 1:9
+        sigpt=s(igpt);
+        tigpt=t(igpt);
+        wght=wt(igpt);
+%  evaluate derivatives etc
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/stokes_flow/stokespost.m

@@ -1,0 +1,35 @@
+%STOKESPOST estimates Stokes error distribution 
+%This implements error estimation for stable Stokes Approximation
+%for further details see
+% "Qifeng Liao and David Silvester",
+% "A simple yet effective a posteriori error estimator for classical 
+%       mixed approximation of Stokes equations.",
+% "Applied Numerical Mathematics, 2011",
+% "http://eprints.ma.man.ac.uk/1488/"
+
+%   IFISS scriptfile: DJS; 29 September 2013.
+% Copyright (c) 2010 D.J. Silvester, Qifeng Liao
+
+% code for outflow boundary
+if domain == 3 || domain == 10 || domain==4,  % step, channel, obstacle
+      fprintf('Natural outflow on right hand vertical boundary ..\n')
+      neumannb=out_Neumann_bound(mv,xy);
+else, fprintf('Enclosed flow ..\n'),  neumannb=[]; end
+%
+if qmethod==3,
+fprintf('FAST Stokes Q2-P1 a posteriori error estimation \n') 
+[hx,hy,eex] = eexgen(xy,xyp,mv,ee); % check element numbering
+[error_x,error_y,fex,fey,ae]=stokespost_q2p1(xy,mv,mbound,neumannb,xst,eex,hx,hy);
+elseif qmethod==2,
+fprintf('FAST Stokes Q2-Q1 a posteriori error estimation \n')
+[x_q2p1,y_q2p1,xy_q2p1,xyp_q2p1,ee] = q2p1grid(x,y,xy,mv,bound);
+[hx,hy,eex] = eexgen(xy_q2p1,xyp_q2p1,mv,ee); % check element numbering
+[error_x,error_y,fex,fey,ae]=stokespost_q2q1(xy,mv,mp,mbound,neumannb,xst,eex,hx,hy);
+else, error('Undefined mixed approximation!'), end
+%
+error_div = q2div(xy,mv,xst);
+error_tot = sqrt(error_x+error_y+error_div.^2);
+errorest=norm(error_tot,2);
+fprintf('estimated energy error is %10.4e \n',errorest)
+if domain ==3, eplotl2(error_tot,mv,xy,x,y,34,'Estimated energy error');
+else, eplot2(error_tot,mv,xy,x,y,34,'Estimated energy error');  end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/stokes_flow/test_problems/stokes_testproblem.m

@@ -1,0 +1,38 @@
+%stokes_testproblem    sets up reference Examples 5.1 to 5.4
+%   IFISS scriptfile: DJS; 29 September 2013.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference Stokes problem.\n')
+fprintf('\nchoose specific example (default is cavity)');
+fprintf('\n     1  Channel domain')
+fprintf('\n     2  Flow over a backward facing step')
+fprintf('\n     3  Lid driven cavity')
+fprintf('\n     4  Colliding flow\n');
+sn = default('',3);
+if sn==1, 
+system('/bin/cp ./stokes_flow/test_problems/poiseuille_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/poiseuille_bc.m ./stokes_flow/stream_bc.m');
+   channel_stokes, xsolve_stokes
+elseif sn==2 
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_bc.m ./stokes_flow/stream_bc.m');
+   longstep_stokes, solve_step_stokes  
+elseif sn==3
+   lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3); 
+   if lid_model ==3    
+system('/bin/cp ./stokes_flow/test_problems/regcavity_flow.m ./stokes_flow/specific_flow.m');
+   elseif lid_model ==2    
+system('/bin/cp ./stokes_flow/test_problems/tightcavity_flow.m ./stokes_flow/specific_flow.m');
+   else
+system('/bin/cp ./stokes_flow/test_problems/leakycavity_flow.m ./stokes_flow/specific_flow.m');
+   end
+system('/bin/cp ./stokes_flow/test_problems/zero_bc.m ./stokes_flow/stream_bc.m');
+   square_stokes, solve_stokes
+elseif sn==4
+system('/bin/cp ./stokes_flow/test_problems/collide_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/collide_bc.m ./stokes_flow/stream_bc.m');
+   square_stokes, solve_stokes
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/stokes_flow/xsolve_stokes.m

@@ -1,0 +1,65 @@
+%XSOLVE_STOKES solve (possibly singular) Stokes problem 
+%   IFISS scriptfile: DJS; 29 September 2013.
+% Copyright (c) 2005 D.J. Silvester, H.C. Elman, A. Ramage 
+clear variables
+gohome
+cd datafiles
+load square_stokes_nobc.mat
+%
+fprintf('imposing boundary conditions and solving system ...\n') 
+%% boundary conditions
+[Ast,Bst,fst,gst] = flowbc(A,B,f,g,xy,bound);
+%
+np=length(gst); nu=length(f)/2; tic, 
+%% compute solution
+if qmethod>1
+   beta=0;
+   xst=[Ast,Bst';Bst,sparse(np,np)]\[fst;gst];
+elseif qmethod==1
+   beta=default('stabilization parameter (default is 1/4)',1/4);
+   xst=[Ast,Bst';Bst,-beta*C]\[fst;gst];
+elseif qmethod==0
+   fprintf('computing pressure stabilized solution...\n')
+   xst=[Ast,Bst';Bst,-C]\[fst;gst]; beta=1;
+end
+[msgstr,msgid] = lastwarn;
+if length(msgid)==27,
+   if sum(msgid=='MATLAB:nearlySingularMatrix')==27,
+      fprintf('This should not cause difficulty for enclosed flow problems.\n\n');
+   end
+end
+lastwarn('');
+etoc=toc; fprintf('Stokes system solved in %8.3e seconds\n',etoc) 
+%
+%%% plot solution
+spc=default('uniform/nonuniform streamlines 1/2 (default uniform)',1);
+flowplot(qmethod,xst,By,Bx,A,xy,xyp,x,y,bound,spc,33);
+fprintf('\n') 
+%
+%%% estimate errors
+if qmethod==1
+   [jmpx,jmpy,els] = stressjmps_q1p0(1,xst,xy,ev,ebound);
+   [error_x, error_y, fex, fey, ae] = stokespost_q1p0_p(jmpx,jmpy,els,xy,ev);
+   [error_x,error_y] = stokespost_q1p0_bc(ae,fex,fey,...
+                                  error_x,error_y,xy,ev,ebound);
+   error_div = q1div(xy,ev,xst);
+   errorest=sqrt(sum(error_x.^2 + error_y.^2 + error_div.^2));
+   fprintf('estimated overall error is %10.6e \n',errorest)
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/timestepping/Contents.m

@@ -1,0 +1,26 @@
+% TIMESTEPPING
+%
+% Files
+%   box_heat                 - solve Heat Equation in box-shaped domain 
+%   dtflowbc                 - imposes updated BC at the current timestep
+%   dtheatbc                 - imposes Dirichlet BC at the current timestep
+%   ell_heat                 - solve Heat Equation in L-shaped domain 
+%   helpme_heat              - heat equation interactive help 
+%   helpme_timestepping      - unsteady problem solution interactive help 
+%   helpme_unsteady_cd       - unsteady CD problem interactive help 
+%   helpme_unsteady_navier   - unsteady flow problem interactive help 
+%   resetflowbc              - resets boundary values in unsteady flow 
+%   resetheatbc              - resets boundary values in unsteady diffusion 
+%   stabtr                   - standard integrator based on stabilized TR
+%   stabtrNS                 - Navier-Stokes integrator using stabilized TR 
+%   unsteady_cd              - unsteady CD problem in square domain 
+%   unsteady_navier          - unsteady flow problem in square domain
+%   unsteady_obstacle_navier - unsteady flow problem in obstacle domain
+%   unsteady_step_navier     - solve Navier-Stokes problem in step domain
+%   unsteadyflowbc           - imposes inflow boundary condition 
+%   xxheatbc                 - hot wall Dirichlet boundary condition 
+%   colamdAxB                - refined saddlepoint system solver      
+%   defaultAxB               - standard MATLAB saddlepoint system solver      
+%   oldstabtrNS              - legacy code | replaced by stabtrNS 
+%   step_heat                - solve heat equation in backward step domain 
+%   xxstreambc               - dynamic Dirichlet BC on streamfunction

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/timestepping/oldstabtrNS.m

@@ -1,0 +1,171 @@
+function [DT,U,Udot,time] = stabtrNS(qmethod,xy,ev,bound,A,B,C,G,AxB,...
+                                uzero,dtzero,tfinal,tol,nstar,tout)
+%OLDSTABTRNS legacy code | replaced by stabtrNS 
+%   [DT,U,Udot,time] = stabtrNS(qmethod,xy,ev,bound,A,B,C,G,AxB,...
+%                          uzero,dtzero,tfinal,tol,nstar,0);
+%   input
+%          qmethod   approximation method
+%          xy        vertex coordinate vector
+%          ev        mv/ev  Q2/Q1 element mapping matrix
+%          bound     boundary vertex vector 
+%          A, B, C   matrices defining the saddle point system
+%          G         mass matrix for velocity
+%          AxB       saddle point system solver function_name
+%          uzero     initial velocity condition
+%          dtzero    initial timestep
+%          tfinal    final time 
+%          tol       local accuracy error tolerance   
+%          nstar     averaging frequency   
+%          tout      output data switch
+%   output
+%          DT        timestep history
+%          U         solution history
+%          Udot      solution time derivative history
+%          time      discrete time evolution
+%
+% calls functions: unsteadyflowbc, dtflowbc, resetflowbc, AxBhandle
+%   IFISS function: DJS; 7 June 2010; 5 January 2011
+% Copyright (c) 2006 D.J. Silvester, D.F. Griffiths, M. Schneider 
+global viscosity 
+warning('on','MATLAB:nearlySingularMatrix'); wchar=27;
+fprintf('Solving DAE system using stabilized TR ...\n')
+AxBhandle=str2func(AxB),
+[np,nuv]=size(B); nv=nuv/2;
+ub=uzero; 
+Ndof=length(ub); T=tfinal; dt0=dtzero;
+% preallocate array dimensions
+DT = zeros(1,100); U = zeros(Ndof,100); time = zeros(1,100); 
+Udot = zeros(Ndof,100);
+%%% zeroth time step 
+f=zeros(nuv,1);gzero=zeros(np,1);              
+
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/timestepping/stabtrNS.m

@@ -1,0 +1,174 @@
+function [DT,U,Udot,time] = stabtrNS(qmethod,xy,ev,bound,A,B,C,G,AxB,...
+                                uzero,dtzero,tfinal,tol,nstar,tout)
+%STABTRNS Navier-Stokes integrator using stabilized TR 
+%   [DT,U,Udot,time] = stabtrNS(qmethod,xy,ev,bound,A,B,C,G,AxB,...
+%                          uzero,dtzero,tfinal,tol,nstar,0);
+%   input
+%          qmethod   approximation method
+%          xy        vertex coordinate vector
+%          ev        mv/ev  Q2/Q1 element mapping matrix
+%          bound     boundary vertex vector 
+%          A, B, C   matrices defining the saddle point system
+%          G         mass matrix for velocity
+%          AxB       saddle point system solver function_name
+%          uzero     initial velocity condition
+%          dtzero    initial timestep
+%          tfinal    final time 
+%          tol       local accuracy error tolerance   
+%          nstar     averaging frequency   
+%          tout      output data switch
+%   output
+%          DT        timestep history
+%          U         solution history
+%          Udot      solution time derivative history
+%          time      discrete time evolution
+%
+% calls functions: unsteadyflowbc, dtflowbc, resetflowbc, AxBhandle
+%   IFISS function: DJS; 7 June 2010; 5 January 2011; 26 April 2012; 24 June 2012
+% Copyright (c) 2006 D.J. Silvester, D.F. Griffiths, M. Schneider 
+global viscosity 
+warning('on','MATLAB:nearlySingularMatrix'); wchar=27;
+fprintf('Solving DAE system using stabilized TR ...\n')
+AxBhandle=str2func(AxB),
+[np,nuv]=size(B); nv=nuv/2;
+ub=uzero; 
+Ndof=length(ub); T=tfinal; dt0=dtzero;
+% preallocate array dimensions
+DT = zeros(1,100); U = zeros(Ndof,100); time = zeros(1,100); 
+Udot = zeros(Ndof,100);
+%%% zeroth time step 
+f=zeros(nuv,1);gzero=zeros(np,1);              
+
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/timestepping/test_problems/heat_testproblem.m

@@ -1,0 +1,27 @@
+%heat_testproblem    sets up Reference Examples H1 to H3
+%   IFISS scriptfile: DJS; 9 May 2012.  
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nSpecification of reference heat equation problem.')
+fprintf('\nchoose specific example');
+fprintf('\n     1  Rectangle domain, differently heated sides')
+fprintf('\n     2  L-shaped domain, line and point sources')
+fprintf('\n     3  Backward-facing step domain, differently heated sides\n')
+sn = default('',2);
+%
+if sn==1
+system('/bin/cp ./diffusion/test_problems/zero_rhs.m ./diffusion/specific_rhs.m');
+system('/bin/cp ./timestepping/test_problems/heat1_bc.m ./diffusion/specific_bc.m');
+   box_heat
+elseif sn==2
+system('/bin/cp ./diffusion/test_problems/zero_rhs.m ./diffusion/specific_rhs.m');
+system('/bin/cp ./timestepping/test_problems/heat2_bc.m ./diffusion/specific_bc.m');
+   ell_heat
+elseif sn==3
+system('/bin/cp ./diffusion/test_problems/zero_rhs.m ./diffusion/specific_rhs.m');
+system('/bin/cp ./timestepping/test_problems/heat3_bc.m ./diffusion/specific_bc.m');
+   step_heat
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/timestepping/test_problems/unsteady_cd_testproblem.m

@@ -1,0 +1,22 @@
+%unsteady_cd_testproblem    sets up Reference Examples T-CD1 to T-CD4
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables 
+fprintf('\nspecification of reference unsteady convection-diffusion problem.')
+fprintf('\nchoose specific example');
+fprintf('\n     2  Vertical wind, characteristic layers')
+fprintf('\n     4  Recirculating wind\n')
+sn = default('',2);
+%
+if sn==4
+system('/bin/cp ./convection/test_problems/circular_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/hotwall_bc.m ./diffusion/specific_bc.m');
+square_cd, unsteady_cd
+elseif sn==2
+system('/bin/cp ./convection/test_problems/ref_wind.m ./convection/specific_wind.m');
+system('/bin/cp ./convection/test_problems/ref_bc.m ./diffusion/specific_bc.m');
+ref_cd, unsteady_cd
+else
+error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/timestepping/test_problems/unsteady_navier_testproblem.m

@@ -1,0 +1,33 @@
+%unsteady_navier_testproblem    sets up reference Examples T-NS1 to T-NS5
+%   IFISS scriptfile: DJS; 9 May 2012. 
+% Copyright (c) 2009 D.J. Silvester, H.C. Elman, A. Ramage 
+gohome
+clear variables
+fprintf('\nspecification of reference unsteady Navier-Stokes problem.\n')
+fprintf('\nchoose specific example (default is step)');
+fprintf('\n     2  Flow over a backward facing step')
+fprintf('\n     3  Lid driven cavity');
+fprintf('\n     5  Flow around a square obstruction\n');
+sn = default('',2);
+if  sn==2
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/backwardstep_bc.m ./stokes_flow/stream_bc.m');
+   longstep_stokes, unsteady_step_navier
+elseif sn==3
+   lid_model=default('cavity type leaky/tight/regularised 1/2/3 (regularised)',3); 
+   if lid_model ==3    
+system('/bin/cp ./stokes_flow/test_problems/regcavity_flow.m ./stokes_flow/specific_flow.m');
+   elseif lid_model ==2    
+system('/bin/cp ./stokes_flow/test_problems/tightcavity_flow.m ./stokes_flow/specific_flow.m');
+   else
+system('/bin/cp ./stokes_flow/test_problems/leakycavity_flow.m ./stokes_flow/specific_flow.m');
+   end
+system('/bin/cp ./stokes_flow/test_problems/zero_bc.m ./stokes_flow/stream_bc.m');
+   square_stokes, unsteady_navier
+   elseif sn==5
+system('/bin/cp ./stokes_flow/test_problems/obstacle_flow.m ./stokes_flow/specific_flow.m');
+system('/bin/cp ./stokes_flow/test_problems/obstacle_bc.m ./stokes_flow/stream_bc.m');
+    obstacle_stokes, unsteady_obstacle_navier
+else
+   error('reference problem datafile not found!')
+end

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/timestepping/unsteady_navier.m

@@ -1,0 +1,71 @@
+%UNSTEADY_NAVIER unsteady flow problem in square domain
+%   IFISS scriptfile: DJS; 27 May 2012. 
+% Copyright (c) 2009 D.J. Silvester
+clear variables
+global pde 
+fprintf('Unsteady flow in a square domain ...\n')
+global viscosity
+viscosity=default('viscosity parameter (default 1/200)',1/200);
+%
+%% initialize for time stepping iteration: compute Stokes solution
+%% load assembled matrices
+gohome
+cd datafiles
+load square_stokes_nobc.mat
+%
+%
+pde=14; domain=1;
+%% set parameters
+fprintf('Discrete Saddle-Point DAE system ...\n')
+tfinal = default('target time? (default 100)',100);
+tol = default('accuracy tolerance? (default 1e-4)',1e-4);
+nstar = default('averaging frequency? (default 10)',10);
+vswitch = default('plot vorticity evolution? 1/0',0);
+dtzero=1e-9; uzero=zeros(size(f)); gzero=zeros(size(g));
+%% compute solution
+tstart = tic;
+if qmethod>1, 
+   np=length(g); 
+   AxB='defaultAxB'; 
+   [DT,U,Udot,time] = stabtrNS(qmethod,xy,mv,bound,A,B,sparse(np,np),G,AxB,...
+                                uzero,dtzero,tfinal,tol,nstar,1);
+else  
+% call the specialized saddle point system solver 
+   AxB='colamdAxB'; 
+% fixed beta (independent of the viscosity)    
+   beta=1;   
+   if qmethod==1,   
+	  beta=1/4;     % default parameter
+   end  
+% scaled beta (mutiplied by the viscosity)   
+   beta=beta*viscosity;
+   [DT,U,Udot,time] = stabtrNS(qmethod,xy,ev,bound,A,B,beta*C,G,AxB,...
+                                uzero,dtzero,tfinal,tol,nstar,1);   
+end
+etoc=toc(tstart); fprintf('Integration took  %8.3e seconds\n\n',etoc) 
+
+save square_unsteadyflow.mat DT U Udot time viscosity
+%
+% visualise solution and hold the timestep sequence plot
+marker = 'bo';
+dttplot(DT,99,marker)
+if vswitch==1
+pause(3)
+   if qmethod>1,
[…]

### 2014-07-11 MAX_PROJECT/current_code/utils/ifiss3.3/timestepping/unsteady_step_navier.m

@@ -1,0 +1,70 @@
+%UNSTEADY_STEP_NAVIER solve Navier-Stokes problem in step domain
+%   IFISS scriptfile: DJS; 27 May 2012
+% Copyright (c) 2009 D.J. Silvester
+clear variables
+global pde 
+fprintf('Unsteady flow in a step domain ...\n')
+global viscosity
+viscosity=default('viscosity parameter (default 1/50)',1/50);
+%
+%% initialize for time stepping iteration: compute Stokes solution
+%% load assembled matrices
+gohome
+cd datafiles
+load step_stokes_nobc.mat
+%
+%
+pde=14; domain=3;
+%% set parameters
+fprintf('Discrete Saddle-Point DAE system ...\n')
+tfinal = default('target time? (default 200)',200);
+tol = default('accuracy tolerance? (default 1e-4)',1e-4);
+nstar = default('averaging frequency? (default 10)',10);
+vswitch = default('plot vorticity evolution? 1/0',0);
+dtzero=1e-9; uzero=zeros(size(f));gzero=zeros(size(g));
+%% compute solution
+tstart = tic;
+if qmethod>1, 
+np=length(g); 
+AxB='defaultAxB'; 
+[DT,U,Udot,time] = stabtrNS(qmethod,xy,mv,bound,A,B,sparse(np,np),G,AxB,...
+							uzero,dtzero,tfinal,tol,nstar,1);
+else  
+% call the specialized saddle point system solver 
+AxB='colamdAxB';  
+beta=1;   
+if qmethod==1, 
+	  beta=1/4;     % default parameter
+end  
+% scaled beta (mutiplied by the viscosity)   
+beta=beta*viscosity;
+[DT,U,Udot,time] = stabtrNS(qmethod,xy,ev,bound,A,B,beta*C,G,AxB,...
+							uzero,dtzero,tfinal,tol,nstar,1);   
+end
+etoc=toc(tstart); fprintf('Integration took  %8.3e seconds\n\n',etoc)
+
+save step_unsteadyflow.mat DT U Udot time viscosity
+%
+% visualise solution and hold the timestep sequence plot
+marker = 'bo';
+dttplot(DT,13,marker)
+if vswitch==1
+pause(3)
+   if qmethod>1,
+   step_unsteadyflowplot(qmethod,mv,U,time,By,Bx,G,xy,xyp,x,y,bound,0.1,167);
+   else
[…]

### 2014-08-15 doc/ttsp2013/paper/ttsp_2013.tex
removing old preconditioning results from ttsp paper

\documentclass[letterpaper,11pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{tmadd,tmath}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{tmath,tmadd}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[mathcal]{euscript} 
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery, Steven P. Hamilton, Thomas M. Evans, and Paul
  P.H. Wilson
  \\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}} }

\date{\today} \title{A Parallel and Preconditioned Monte Carlo
  Synthetic-Acceleration Method for the Simplified $P_N$
  Approximation}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\abstract

We present Monte Carlo Synthetic Acceleration (MCSA) as a linear
solver technique for the simplified $P_N$ ($SP_N$) discretization of
the Boltzmann neutron transport equation. Using the fully-formed
linear operator for the transport problem, we explore solutions to the
$SP_N$ equations with MCSA using criticality calculations driving
problem. Due to neutron scattering in the light water moderator,
several difficulties arise when applying MCSA to these problems that
were not observed in previous work where the method was applied to
more opaque thermal radiaton transport systems. Of most significance,
it was discovered for these problems that simply reducing the spectral
radius of the Monte Carlo problem below unity is not enough to
efficiently achieve convergence as reported in previous work. In order
to effectively solve the $SP_N$ equations for the model problem, an
algebraic preconditioning strategy is developed for MCSA where we
utilize sparse approximate inverse representations of the operator as
preconditioners to achieve convergence. Using the preconditioners,
MCSA performance is compared with production implementations of
preconditioned Krylov methods. Our studies indicate that
the iterative performance of MCSA when used to solve the $SP_N$
equations for light water reactor problems is competitive with the
preconditioned Krylov methods when operator spectrum is properly
conditioned. To solve large problems, we have parallelized the method
by constructing a variant of the multiple-set overlapping-domain
(MSOD) algorithm for Monte Carlo calculations. We discuss the method
and then present results indicating its scalability.

%%---------------------------------------------------------------------------%%
\section{Introduction}

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization and burnup, and
high confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems where the solution over
the entire core of a reactor is desired are often variants on the
discrete ordinates ($S_N$) method \cite{evans_denovo:_2010} or the
method of characteristics \cite{askew_1972}. For fission reactor
neutronics simulations, the discrete ordinates method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011} which in many cases can be
intractable without sufficient computational hardware. On the other
side of the spectrum, low-fidelity diffusion solutions are often
leveraged as an alternative to compute a solution over the entire
domain a reactor core. In the 1960's, Gelbard developed an ad-hoc,
multidimensional extension of the simple single dimension planar $P_N$
equations that created a system of coupled, diffusion-like equations
known as the simplified $P_N$ ($SP_N$) equations \cite{gelbard_1960}
which incorporates additional angular information into a set of
diffusion-like equations. Further work in both aysmptotic analysis and
algorithmic development proceeded in the following decades to serve as
a theoretical basis for the method
\cite{morel_1996,brantley_simplified_2000}. Recent work has expanded
the discretization to a fully algebraic form ammenable to solution
with modern linear algebra techniques including Krylov solvers and
advanced precondtioners including multigrid methods and incomplete
factorizations \cite{hamilton_2014}. Using these new techniques, the
reduction in numerical complexity compared to current deterministic
solution methods using the $SP_N$ approximation could mean significant
savings in both compute time and memory required and permit larger
simulations to be performed on the available computing platforms.

Looking towards future computing platforms, as machines begin to
operate at hundreds of petaflops peak performance and beyond, trends
toward reduced energy consumption will require incredibly high levels
of concurrency to achieve the desired computation rates
\cite{kogge_using_2011}. The end result of these hardware changes is
that the larger number of low-powered processors will be prone to both
soft failures such as bit errors in floating point operations and hard
failures where the data owned by that processor cannot be recovered
\cite{u.s._department_of_energy_resilient_2012}. Because these
failures are predicted to be common, resilient solver technologies are
are a possible means of overcoming these events at the application
level. With solutions to deterministic transport problems based on
Monte Carlo techniques, such issues are potentially alleviated by
statistical arguments. In the case of soft failures, isolated floating
point errors in the Monte Carlo simulation are absorbed within tally
statistics while a hard failure is treated as a high variance event
where some portion of the Monte Carlo histories lost.

Considering these physics and hardware issues, the goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. In previous work \cite{evans_monte_2014}, we
applied the Monte Carlo synthetic acceleration (MCSA) method to the
thermal radiation diffusion equation. For several transient problems
we observed that MCSA performed competetively with preconditioned
Krylov methods. In a continuation of these studies, we apply the
method to light water reactor problems discretized by the $SP_N$
equations. This system has no time dependence, a large scattering
component, and assymetry in the linear operator, creating a set of
problems complementary to the transient, opaque, and symmetric systems
of our previous analyis. In addition, we apply a new fixed point
iteration in the MCSA sequence as well as a new estimator in order to
study their effects on convergence.

The paper is organized as follows. First, we briefly introduce the
$SP_N$ equations for eigenvalue problems along with the discretization
used for this work and the model problem. We then describe the
application of the MCSA algorithm to the solution of the
equations. Next, we demonstrate that the basic Jacobi preconditioning
applied in our previous work is unsatisfactory to achieve convergence
for neutron transport in light water reactors and subsquently develop
a new preconditioning scheme. We then apply this preconditioning
scheme to a set of criticality problems and compare the
performance of the preconditioned MCSA method to preconditioned Krylov
methods. We then parallelize the MCSA algorithm leveraging the MSOD
scheme to enable very large problems with a demonstration of
scalability using the test problems. We then finish with conclusions
and indication of future work.

%%---------------------------------------------------------------------------%%
\section{Simplified $P_N$ Approximation}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}):
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
fission matrix, $\mathbf{F}$. We apply the following change of
variables to yield the set of multigroup pseudo-moments
$\mathbb{U}_n$:
\begin{subequations}
  \begin{gather}
    \mathbb{U}_1 = \mathbf{\Phi}_0 + 2\mathbf{\Phi}_2 \\
    \mathbb{U}_2 = 3\mathbf{\Phi}_2 + 4\mathbf{\Phi}_4 \\
    \mathbb{U}_3 = 5\mathbf{\Phi}_4 + 6\mathbf{\Phi}_6 \\
    \mathbb{U}_4 = 7\mathbf{\Phi}_6 \:.
  \end{gather}
  \label{eq:spn7_subs}
\end{subequations}
where now the application of the fission matrix to the moment vectors
will be expanded into a set of block matrices in identical fashion to
the scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
To complete the spatial discretization, the gradient operators are
resolved by a finite volume scheme as in \cite{hamilton_2014} such
that the flux and current are continuous at cell boundaries.. It
should be noted here that with respect to a general MCSA scheme, the
introduction of fission in the system will only affect the source
vector in the linear system for the eigenvalue schemes used in this
work. The linear operator, and therefore overall MCSA performance will
be dictated by streaming and scattering as defined on the left-hand
side. We refer the reader to \cite{hamilton_2014} for an extensive
presentation of the $SP_N$ discretization used in this work.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
In this section we briefly review the Monte Carlo method for linear
systems and Monte Carlo Synthetic Acceleration. Given a linear
problem:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in
\mathbb{R}^N$. To form the Monte Carlo method we define the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$ which we hten split to get:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}^T$ is the adjoint iteration matrix. The spectral radius
of $\ve{H}^T$ must remain less than 1 for convergence. By defining the
following adjoint inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
  \ve{x} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
We can acquire the adjoint solution by forming the Neumann series by
writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence by forming
the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

In terms of radiation transport, this adjoint method is equivalent to
a traditional forward method where the initial state $i_0$ of the
random walk is determined by sampling the source vector $\ve{b}$ with
probabilities:
\begin{equation}
  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
  \label{eq:adjoint_source_probability}
\end{equation}
with a random walk starting weight of:
\begin{equation}
  W_0 = ||\ve{b}||_1 \frac{b_i}{|b_i|}\:,
  \label{eq:adjoint_starting_weight}
\end{equation}
which gives the additional useful relation:
\begin{equation}
  b_{i_0} = W_0 P_{(i_0=i)}\:.
  \label{eq:adjoint_source_definition}
\end{equation}
We build weights and probabilities for the random walk using the
\textit{adjoint Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where we are forming the decomposition with respect to the transpose
of $\ve{H}$. $\ve{P}$ is defined element-wise as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state, $j$, from the current state
in the random walk, $i$, by sampling column-wise (or row-wise if an
adjoint probability matrix is formed). Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. The contribution to the solution in state $j$ from
a particular random walk permutation of $k$ events is given by
\textit{collision estimator}:
\begin{equation}
  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state, $i_m$, of the random walk. The expectation value
using all permutations is then:
\begin{equation}
  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:,
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded utilizing Eq~(\ref{eq:adjoint_source_definition})
to insert the source term, directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore, also providing an unbiased Monte Carlo estimate of the
solution. It should be noted here that
Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
component of our desired solution vector when really what we desire is
the entire solution vector. In an adjoint Monte Carlo simulation using
this estimator, the $w_{ij}$ elements that are added into the tally
for each state are only selected if/when the random walk currently
resides in that state. Much like a mesh tally in a particle transport
simulation, we have $N$ simultaneous tallies for $\ve{A} \in
\mathbb{R}^{N \times N}$ that will yield the entire solution vector.

In addition to the collision estimator, an additional estimator is
available due to the work of Okten \cite{okten_solving_2005} that uses
the method of expected values as a means to improve the Monte Carlo
estimate. As outlined by Spanier and Gelbard
\cite{spanier_monte_1969}, the method of expected values is a
deterministic averaging of events that may potentially occur in the
Monte Carlo random walk sequence. Okten applied this principle
directly to discrete Monte Carlo by forming the \textit{expected value
  estimator} for a random walk of $k$ events:
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
  \label{eq:expected_value_estimator}
\end{equation}
where now the contribution of the iteration matrix is
deterministically averaged at step $m$ over all potential states $j$
that may be reached from the current state $i_m$. Via Okten, the
estimator can be shown to be unbiased through a comparison to the
collision estimator. We can first rewrite the summation in
Eq~(\ref{eq:expected_value_estimator}):
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
  \delta_{i_m,i} h_{ji}\,
  \label{eq:unbiased_eval_1}
\end{equation}
where $N$ is the number of states in the system. Immediately, we see
the collision estimator as defined by
Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
the expectation value as:
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
which is equivalently is the $j^{th}$ component of
Eq~(\ref{eq:richardson_split}):
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
and is therefore an unbiased estimate. Compared to the collision
estimator, the expected value estimator provides additional
information at every step of the random walk, yielding potentially
better statistics with the same amount of transport
work. Conveniently, even if no Monte Carlo histories are computed, the
expected value estimator still deterministically computes the first
term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
estimator will provide no information.

\subsection{MCSA Solution of the $SP_N$ Equations}
\label{subsec:mcsa_solution}
To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
computation while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor while not
affecting Monte Carlo transport.

At each eigenvalue iteration, we then solve the linear system with the
MCSA:
\begin{subequations}
  \begin{gather}
    \ve{r}_{l} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1/2} = \mathbf{\Phi}^{n+1}_l + \ve{r}_l\:,\\
    \ve{r}_{l+1/2} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l+1/2}\:,\\
    \ve{M}\delta\mathbf{\Phi}^{n+1}_{l+1/2} = \ve{r}_{l+1/2}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1} = \mathbf{\Phi}^{n+1}_{l+1/2} + \delta
    \mathbf{\Phi}_{l+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where the $l$ subscript indicates the MCSA iteration number and $n$
again indicates the eigenvalue iteration. We refer the reader to
\cite{evans_monte_2014} for more a more complete definition of the
Monte Carlo method for linear systems and MCSA.

\subsection{Preconditioned MCSA}
\label{subsec:preconditioning}
In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. Preconditioning serves as a
means of achieving this by altering the eigenvalue spectrum of the
iteration matrix. It is possible to use general left, right, and left/right
preconditioning with MCSA by carefully considering the underlying
Monte Carlo problem that will be solved with the Neumann-Ulam
method. We consider here the general left/right preconditioned method
as the left or right preconditioned methods can be inferred from its
formulation.

We consider a left preconditioner $\ve{M_L}$ and a right
preconditioner $\ve{M_R}$. The left/right preconditioned linear
problem is then:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
  \label{eq:left_right_linear_problem}
\end{equation}
To handle the right preconditioning, the system is written with a
substitution of variables:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
  \label{eq:left_right_subs_problem}
\end{equation}
with
\begin{equation}
  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
  \label{eq:left_right_recover}
\end{equation}
To apply such a method to MCSA, we solve for the substituted variable
$\ve{u}$ during the iteration sequence:
\begin{subequations}
  \begin{gather}
    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
  \end{gather}
  \label{eq:left_right_mcsa}
\end{subequations}
and then recover the original solution vector with
Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
isolate the generation of the correction:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
  \label{eq:left_right_correction}
\end{equation}
and note that the preconditioned residual of the substituted variable
is now serving as the source and the new iteration matrix is:
\begin{equation}
  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
  \label{eq:left_right_iteration_matrix}
\end{equation}
As we require $(i,j)$ element-wise access to the iteration matrix in
order to construct probabilities and weights for the Monte Carlo
procedure from the Neumann-Ulam decomposition, the \textit{composite
  operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
matrix-matrix multiplication.

Several possible shortcomings of this preconditioning approach are
readily observed. First, the matrix-matrix multiplication operation
for sparse, parallel distributed matrices is significantly more
expensive than a matrix-vector multiplication operation. Second, each
preconditioner must be explicitly inverted, an operation in itself
that may be expensive and which prohibits the use of any
preconditioners which provide no mechanism to extract their
inverse. Third, for many modern preconditioning methods, this
inversion may yield dense matrices, destroying sparsity and further
impeding the performance of a matrix-matrix multiplication
operation. It is also interesting to note that the Monte Carlo problem
in the general left/right preconditioned scheme given by
Eq~(\ref{eq:left_right_correction}) is not fully left/right
preconditioned (meaning that we do not recover $\ve{x}$), but instead
part of a sequence for finding the substituted variable $\ve{u}$. We
do, however, gain the benefits of this general preconditioning by
building the iteration matrix in
Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
preconditioned linear operator.

\subsubsection{Block Jacobi Preconditioning}

\subsubsection{Sparse Approximate Inverse Preconditioning}

%%---------------------------------------------------------------------------%%
\section{Criticality Calculations}

\subsection{C5G7 Benchmark}

\subsection{Light Water Reactor Fuel Assemblies}

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA}
\label{sec:parallel_mcsa}
To date, parallel Neumann-Ulam methods have been limited to full
domain replication with parallelism exploited through individual
histories \cite{alexandrov_efficient_1998} and in this work we will
exploit particle transport algorithms to alleviate this. In this
section we briefly review particle transport methods for domain
decomposed Monte Carlo. We then devise a parallel algorithm for the
Neumann-Ulam method based on the multiple-set overlapping-domain
decomposition algorithm and a parallel algorithm for the MCSA
iteration that leverages the parallel Monte Carlo algorithm and
general parallel matrix-vector operations. We then perform parallel
scaling studies to test its performance on a leadership class machine
and compare this performance to the production Krylov methods.

%%---------------------------------------------------------------------------%%
\subsection{Domain Decomposed Monte Carlo}
\label{subsec:msod}
Large-scale problems will for reasons typically related to memory
restrictions or performance have their data partitioned such that each
parallel process owns a subset of the equations in the linear
system. Given this convention, the Neumann-Ulam Monte Carlo algorithm
must perform random walks over a domain that is decomposed and must
remain decomposed to avoid the same performance and memory
restrictions. To parallelize this algorithm, we then seek parallel
Monte Carlo algorithms that handle domain decomposition.

To motivate this problem, consider the square domain presented in
Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
Carlo particle transport problem. If the domain were decomposed into 9
subdomains as shown, each of those subdomains and their associated
data (i.e. cross subsections) could be owned by a different parallel
process in the computation. In Figure~\ref{fig:ddmc_example}, the
tracks of 3 particles born in the center subdomain are shown. In this
example, particle A is first transported from the center domain to the
domain directly to the left. Before the scattering event in the new
domain may be processed, particle A must be communicated between the
two parallel processes that own those domains. For any given particle
in the transport simulation, this communication event may hardly occur
during the lifetime of the particle (particle B), or may occur many
times (particle C) depending on the problem parameters and the random
path in phase space taken by the particle. Scalable parallel
algorithms for domain decomposed Monte Carlo are those that handle
this domain-to-domain communication of particles effectively and
balance it with the on-process computations for particle interactions,
tallies, response calculations, and other simulation requirements.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Monte Carlo transport example
      illustrating how domain decomposition requires parallel
      communication of particles.} \textit{If a particle crosses a
      domain boundary it must be communicated to the next domain in
      its path.}}
  \label{fig:ddmc_example}
\end{figure}

For a domain decomposed Neumann-Ulam method, the situation is very
much the same where instead the transport process is now discrete and
the "physics" of the transport process is the Monte Carlo game with
probabilities and weights described by
Eq~(\ref{eq:neumann_ulam_decomposition}) for the given linear
problem. Figure~\ref{fig:ddnu_example} gives the domain decomposed
Neumann-Ulam analog of particle transport problem in
Figure~\ref{fig:ddmc_example}. Imagine in this problem that we are
solving for the monoenergetic scalar neutron flux as in the model
diffusion problem presented in Appendix~\ref{chap:diffusion_problem}.
In this problem, the solution has only a spatial dependence and
therefore each point in the mesh corresponds to an equation in the
resulting linear system. In Figure~\ref{fig:ddnu_example}, the domain
has now been discretized by this mesh and again decomposed into 9
subdomains, each owning a piece of the mesh and therefore the
corresponding equations in the linear system. As we play the Monte
Carlo game presented in Chapter~\ref{ch:stochastic_methods}, the
discretization and resulting Neumann-Ulam decomposition of the problem
describes how each state in the system is coupled to the other states
in the system. This coupling is then responsible for the discrete
random walk that each history takes as shown in this example. As
discrete states (or rows in the linear system) that do not exist in
the local domain are reached during the random walk by stochastic
histories, the histories must be communicated to the subdomain that
does own the discrete state in an analogous fashion to the particles
in the previous example.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
      how domain decomposition requires parallel communication of
      histories.} \textit{Each mesh point corresponds to an equation
      in the linear system and the coupling among equations is
      described by the discretization of the problem.}}
  \label{fig:ddnu_example}
\end{figure}

To further motivate using a parallel algorithm similar to particle
transport methods, again consider history A in
Figure~\ref{fig:ddnu_example}. This history begins in the center
subdomain and through the course of transitioning through states in
the system arrives at a state that exists in the subdomain directly to
the left of the center subdomain. History A must then be communicated
from the center subdomain to the subdomain immediately to the left to
continue the random walk. As with the particle transport system, we
may find that histories in a domain decomposed Neumann-Ulam method are
only communicated a few times (history B) if at all, or they may be
communicated many times (history C) depending on the discretization
and the outcome of the Monte Carlo game. In identical fashion to the
particle transport problem, a domain decomposed Neumann-Ulam method
has the same communication requirements with all histories in both
examples requiring the same parallel operations. The amount of
communication of histories from domain to domain will be the primary
factor in the parallel scalability of the algorithm. In
Appendix~\ref{chap:parallel_theory}, the number of histories
communicated in a given problem are quantified analytically for a
simple model problem.

%%---------------------------------------------------------------------------%%
\subsection{Domain Decomposed Neumann-Ulam Algorithm}
\label{subsec:asynchronous_algorithm}
In the context of radiation transport, in 2009 Brunner and colleagues
provided a fully asynchronous domain decomposed parallel algorithm as
implemented in production implicit Monte Carlo codes
\cite{brunner_efficient_2009}. We will adapt their algorithm and
directly apply it to construct a parallel formulation of the
Neumann-Ulam method. Direct analogs can be derived from their works by
noting that the primary difference between solving a linear transport
system with Monte Carlo methods and traditional fixed source Monte
Carlo transport problems is the content of the Markov chains that are
generated.

In this subsection Brunner and Brantley's fully asynchronous algorithm,
which was effectively implemented verbatim for this work, is presented
along with its application to the Neumann-Ulam method. In their work
they identify two data sets that are required to be communicated: the
sharing of particles that are transported from one domain to another
and therefore from one processor to another and a global communication
that signals if particle transport has been completed on all
processors. Both of these communication sequences will be addressed at
a high level along with how the Monte Carlo data structures they
require are constructed in parallel.

\subsubsection{Parallel Transport Domain and Source}
\label{subsubsec:domain_generation}
To utilize a parallel transport algorithm, we must generate the
required data with the correct parallel decomposition. For the
Neumann-Ulam method, the transport domain consists of all states in
the system that are local, and the probabilities and weights for all
state transitions possible in the local domain. At the core of this
representation is the Neumann-Ulam decomposition of the linear
operator as given by Eq~\ref{eq:neumann_ulam_decomposition}. Given an
input parallel decomposition for the linear operator, by definition
the Neumann-Ulam decomposition will have the same parallel
decomposition. In addition, the source vector will have the same
parallel decomposition as the input matrix and therefore all of the
birth states for the histories will exist for at least a single
transition event within the local domain.

Compared to the serial construction of the Neumann-Ulam decomposition,
data for all states that are required to be on process must be
collected in parallel so that all possible transitions that leave the
local domain may be correctly communicated. As an example of this
state collection, Figure~\ref{fig:diffusion_graph} gives the adjacency
graph of a single state in the neutron diffusion matrix for the model
problem presented in Appendix~\ref{chap:diffusion_problem}. The
discretization stencil of the diffusion problem dictates the structure
of this graph and the states to which a history may transition.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
  \end{center}
  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
    \textit{The structure of the graph comes from the discretization
      of the Laplacian operator describing the diffusion
      physics. Adjacent boundary states are collected by traversing
      the graph one step for all local boundary states and collecting
      the equations for those states that are not local. For a given
      transition in the random walk for the diffusion problem, a
      history at mesh point $(i,j)$ may transition to all adjacent
      mesh points including itself, corresponding to a global state
      transition from $m$ to $n$ in the linear system.}}
  \label{fig:diffusion_graph}
\end{figure}

In this example, we currently reside in state $m$ of the system,
directly correlating to physical location $(i,j)$ in the mesh (the
center node). If we play the Monte Carlo game to move to a new state
$n$, then we may move to any of the nodes in this graph, including the
node at which we started. If grid point $(i,j)$ in this example is on
the right side of the local domain boundary and we transition to the
state $n$ corresponding to mesh point $(i+1,j)$, we are now in a state
that is owned by the adjacent domain. We must then gather the
identification number for the neighboring process that owns grid point
$(i+1,j)$ such that we have an address to send all histories that
leave the local domain boundary with this state as their
destination. For the parallel source, these adjacent states are not
required as histories may only be born in states in the local
domain. Once the local Neumann-Ulam decomposition has been generated
with the proper communication data collected from adjacent domains,
Monte Carlo transport may proceed in parallel.

\subsubsection{Domain Decomposed Algorithm}
\label{subsubsec:parallel_mc_algorithm}

We next present Brunner and Brantley's 2009 algorithm in detail and
discuss how it is adapted to parallelize the Neumann Ulam
method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
top level sequence performs the task of managing history transport
through the local domain, communication of histories to adjacent
domains, and the completion of transport. For each of these specific
tasks, additional algorithms, shown in bold in
Algorithm~\ref{alg:parallel_mc_algorithm}
(e.g. \textbf{LocalHistoryTransport()}), are presented for additional
detail in the same manner as Brunner and Brantley.

\begin{algorithm}[h!]
  \caption{Parallel Neumann-Ulam Algorithm}
  \label{alg:parallel_mc_algorithm}
  \begin{algorithmic}[1]
    \State get list of neighbor processors 
    \Comment{Each neighbor owns an adjacent subdomain}
    \ForAll{neighbors}
    \State post non-blocking receive for maximum history buffer size
    \State allocate history buffer
    \EndFor
    \State \textit{historiesCompleted} = 0
    \Comment{local+child finished histories}
    \State \textit{localProcessed} = 0
    \Comment{local trajectories computed (not necessarily finished)}
    \State calculate parent and children processor ID numbers in
    binary tree
    \ForAll{child processes}
    \State post non-blocking receive for \textit{historiesCompleted} tally
    \EndFor
    \State post non-blocking receive for stop message from parent
    \While{stop flag not set}
    \If{any local histories in source or stack}
    \State \textbf{LocalHistoryTransport()}
    \State ++\textit{localProcessed}
    \EndIf
    \If{message check period == \textit{localProcessed} || no local histories}
    \State \textbf{ProcessMessages()}
    \State \textit{localProcessed} = 0
    \EndIf
    \If{no local histories}
    \State \textbf{ControlTermination()}
    \EndIf
    \EndWhile
    \State cancel outstanding non-blocking requests
    \State free all history buffers
  \end{algorithmic}
\end{algorithm}

Successful execution of this algorithm requires construction of the
parallel Neumann-Ulam decomposition as described in the preceding
subsection. This data is used to begin
Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
collected list of neighboring processors generated while building the
Neumann-Ulam decomposition to setup the set of asynchronous messages
required for domain-to-domain communication. Again, consider this
communication pattern for the 9 subdomain example given by
Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each boundary
domain has two neighbors and the center domain four neighbors with
which they will communicate parallel histories and this communication
goes both ways as represented by the adjacent arrows in the
figure. For each set of neighbors, a non-blocking send and receive is
required with data buffers allocated with a user-defined size for
incoming and outgoing histories. This non-blocking structure is
critical to the performance of the algorithm in that it permits local
history transport to continue while new histories to transport are
being collected in the buffers. When a given process is ready to do
more work, it can check these data buffers for incoming histories
requiring further transport. In this way there is a maximum amount of
overlap between communication and transport of histories.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
  \end{center}
  \caption{\textbf{Nearest neighbor history communication sequence.}
    \textit{Each subdomain in the system has a set of nearest
      neighbors determined by the parallel adjacency graph of the
      input matrix. The subdomains are indexed by an integer.}}
  \label{fig:nearest_neighbor_comm}
\end{figure}

Once the nearest-neighbor communication sequence has been prepared,
the completion of transport sequence is readied in lines 6 and 8-12 in
Algorithm~\ref{alg:parallel_mc_algorithm}. In these lines we are
setting up an asynchronous binary communication tree as presented for
the same 9 subdomain example in Figure~\ref{fig:binary_comm_tree}. In
this communication pattern, each process has one parent process
(except for the MASTER process 0) to which it will non-blocking send
the number of histories that terminated their transport procedure by
weight cutoff in the local domain. Equivalently, each process has up
to two child processes from which it will receive their completed
number of histories with a non-blocking receive operation. Setting up a
tree in this manner lets the completed history tally
(\textit{historiesCompleted} in the algorithm) be updated
incrementally and funneled to the root process. Because we are solving
fixed source problems with the Neumann-Ulam algorithm without any type
of variance reduction that may generate more histories that we started
with, once the root process completion sum tallies to the number of
histories in the source, transport is complete. Once this occurs, the
root process non-blocking sends the stop message to its children and
each process non-blocking receives a stop message from its parent as
shown in Figure~\ref{fig:binary_comm_tree}. The stop message is then
propagated up the tree in the same manner.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.0}{
      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
  \end{center}
  \caption{\textbf{Binary communication tree for coordinating the end
      of a parallel Neumann-Ulam solve.} \textit{Each child process
      reports to its parents how many histories completed within its
      domain. When the root process sums all complete histories, it
      forwards the stop signal to its children which in turn forward
      the message up the tree. The subdomains are indexed by an
      integer.}}
  \label{fig:binary_comm_tree}
\end{figure}

With these communication structures prepared, we can now enter the
main transport loop at line 13 of
Algorithm~\ref{alg:parallel_mc_algorithm}. In this loop, a set of
mutually exclusive tasks enabled by the fully asynchronous
communication patterns are executed until the stop signal is received
from the parent process in the binary tree. In this case, mutually
exclusivity permits all local operations to occur independently of
other processes in the system and their current state. For each
process, there are two data structures from which histories may be
obtained for the transport procedure. The first is the local source
and the second is a LIFO (last in first out) stack of histories
transported to the local domain from the adjacent domains. In lines
14-17, if histories exist in either of these data structures, then
they are transported through the local domain using
Algorithm~\ref{alg:local_history_transport}. If the history is
terminated by weight cutoff the \textit{historiesCompleted} tally is
updated. If it hits the boundary of the domain it is added to the
outgoing history buffer for the neighboring domain and if the buffer
is full, it is sent to the neighboring domain with a non-blocking
operation and the memory associated with that buffer reallocated. In
all instances that Algorithm~\ref{alg:local_history_transport} is
executed, \textit{localProcessed} is incremented to account for
histories that have been processed locally.

\begin{algorithm}[h!]
  \caption{\textbf{LocalHistoryTransport()}}
  \label{alg:local_history_transport}
  \begin{algorithmic}[1]
    \State transport history through the domain until termination
    \If{history is in neighbor boundary state}
    \State add history to neighbor buffer
    \If{neighbor buffer is full}
    \State non-blocking send history buffer to neighbor
    \State allocate new history buffer for neighbor
    \EndIf
    \Else
    \If{history terminated by weight cutoff}
    \State post-process history
    \State ++\textit{historiesCompleted}
    \EndIf
    \EndIf
  \end{algorithmic}
\end{algorithm}

Continuing in the transport loop, if there are no local histories to
transport or the \textit{localProcessed} count has reached some
user-defined check frequency, lines 18-21 in
Algorithm~\ref{alg:local_history_transport} check for more histories
to transport in the incoming data buffers by calling
Algorithm~\ref{alg:process_messages}. For each neighboring domain in
the problem, if there are histories in those data buffers then they
are added to the stack for processing and the non-blocking receive
operation re-instantiated. In addition, the terminated histories count
is updated with those values received from the child processes.

\begin{algorithm}[h!]
  \caption{\textbf{ProcessMessages()}}
  \label{alg:process_messages}
  \begin{algorithmic}[1]
    \ForAll{received history buffers from neighbor}
    \State unpack number of histories in buffer
    \State add the histories to the running stack
    \State re-post non-blocking receive with neighbor
    \EndFor
    \ForAll{\textit{historiesCompleted} messages from children}
    \State \textit{historiesCompleted} += message value
    \State re-post non-blocking receive with child
    \EndFor
  \end{algorithmic}
\end{algorithm}

Finally, the transport loop finishes in lines 22-24 of
Algorithm~\ref{alg:local_history_transport} by calling
Algorithm~\ref{alg:control_termination} if there are no local
histories to transport in the stack or the source. In
Algorithm~\ref{alg:control_termination}, we continue to forward the
terminated history tally down to the parent in the binary tree and
send all history buffers to our neighbors, even if they are not
full. If Algorithm~\ref{alg:control_termination} is called by the
MASTER process, it checks for completion of the problem and if
complete, forwards the stop flag onto its children as in
Figure~\ref{fig:binary_comm_tree}. If the process is not the MASTER,
then we check for a stop signal from the parent and forward it to the
child processes.

\begin{algorithm}[h!]
  \caption{\textbf{ControlTermination()}}
  \label{alg:control_termination}
  \begin{algorithmic}[1]
    \State non-blocking send all partially full history buffers
    \ForAll{\textit{historiesCompleted} messages from children} \State
    \textit{historiesCompleted} += message \textit{historiesCompleted}
    \State re-post non-blocking receive with child \EndFor \If{MASTER
      processor} \If{\textit{historiesCompleted} == global \# of
      source histories} \State set stop flag \ForAll{children} \State
    non-blocking send stop message to child \EndFor \EndIf \Else \State
    non-blocking send \textit{historiesCompleted} to parent \State
    \textit{historiesCompleted} = 0 \State check for stop signal from
    parent \If{stop signal from parent} \ForAll{children} \State
    non-blocking send stop message to child \EndFor \EndIf \EndIf
  \end{algorithmic}
\end{algorithm}

The transport loop continues in this manner until all processes have
received the stop signal from their parents. Older versions of this
algorithm, particularly some of those presented in
\cite{brunner_comparison_2006}, use a master/slave approach as given
by Figure~\ref{fig:master_comm_tree} for the completion of a transport
stage instead of the binary tree scheme. In this approach, the root
process still manages the final summation of the completion tally, but
it directly receives completed tally results from all other processes
in the problem instead of from just its children. Although this
implementation is potentially simpler to understand than the binary
tree approach, the authors of \cite{brunner_comparison_2006} observed
that this type of termination pattern, even when implemented in a
fully asynchronous manner, was a significant scalability
bottleneck. This bottleneck is due to the fact that the master process
falls behind the others, creating a load imbalance even with the
addition of a few integer addition operations. We will explicitly
demonstrate in scaling studies how this bottleneck also occurs within
a Neumann-Ulam implementation of this algorithm thus requiring the
implementation of a binary communication tree.

\begin{figure}[t!]
  \begin{center}
    \scalebox{0.75}{
      \input{chapters/parallel_mc/master_comm_tree.pdftex_t} }
  \end{center}
    \caption{\textbf{Master/slave scheme for coordinating the end of a
        parallel Neumann-Ulam solve.} \textit{Each slave process
        reports to the master how many histories completed within its
        domain. When the master process sums all complete histories,
        it sends the stop signal to all slave processes. The
        subdomains are indexed by an integer.}}
  \label{fig:master_comm_tree}
\end{figure}

There is an additional advantage to Brunner and Brantley's work that
although not immediately applicable to this work has the potential to
provide value in the future. In addition to a robust fully
asynchronous communication pattern, this algorithm may also be
modified to account for situations where the total number of histories
in a given stage are not known before starting transport. From a
physics perspective, we might expect this for situations where perhaps
an (n,2n) interaction occurs in a neutronics problem. In this case,
the algorithm is modified to account for both histories terminated and
created and several mechanisms are introduced to determine completion
of the transport stage. For future variations of this work, certain
variance reduction techniques that create histories, such as
splitting, have the potential to be successfully employed as a means
of accelerating the time to solution for a given problem using
MCSA. The parallel Neumann-Ulam algorithm presented here may be
adapted to account for these additional techniques.

%%---------------------------------------------------------------------------%%
\subsection{Multiple-Set Overlapping-Domain Algorithm}
\label{subsubsec:msod}
Although the implementation presented in the previous subsection was
observed by Brunner and Brantley to be robust and allowed for scaling
to large numbers of processors, performance issues were still noted
with parallel efficiency improvements needed in both the weak and
strong scaling cases for unbalanced problems. These results led them
to conclude that a combination of domain decomposition and domain
replication could be used to solve some of these issues. In 2010,
Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\cite{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times potentially achieves efficient simulation
on $O(100,000)$ processors, thus providing both spatial and particle
parallelism.

Each collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its sub-domains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication during particle transport only occurs within a set,
limiting communications during the transport procedure to a group of
$O(100)$ processors, a number that was shown to have excellent
parallel efficiencies in Brunner's work and therefore will scale well
in this algorithm.

To demonstrate this, consider the example adapted from Mervin's work
with Wagner and others in the same area \cite{mervin_variance_2012}
and presented in Figure~\ref{fig:msod_example}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/msod_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Overlapping domain example illustrating how domain
      overlap can reduce communication costs.}
    \textit{All particles start in the blue region of interest. The
      dashed line represents 0.5 domain overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented emanating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as that represented by the path of particle
C. In this case, particle C is scattering between two adjacent
domains, incurring a large latency cost for a single
particle. Finally, with particle B we observe that 0.5 domain overlap
will still not eliminate all communications. However, if 1 domain
overlap were used, the entire geometry shown in
Figure~\ref{fig:msod_example} would be contained on the source
processor and therefore transport of all 3 particles without
communication would occur.

Wagner and colleagues used this methodology for a 2-dimensional
calculation of a pressurized water reactor core and varied the domain
overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
of our example) where a domain contained an entire fuel assembly. For
the fully domain decomposed case, they observed that 76.12\% of all
source particles leave the domain. At 1.5 domain overlap, the
percentage of source particles born in the center assembly leaving the
processor domain dropped to 1.05\% and even further for 0.02\% for the
3 domain overlap. Based on their results, we hypothesize that the
overlap approach coupled with the multiple sets paradigm that will
enhance the scalability of the pure domain-decomposition algorithm for
Neumann-Ulam presented in the previous subsection.

\subsubsection{Generating the MSOD Decomposition}
\label{subsubsec:msod_generation}
In \S~\ref{subsubsec:domain_generation} we discussed how to generate the
parallel transport domain from the Neumann-Ulam decomposition of a
domain decomposed linear operator. We can readily adapt those data
structures to account for the extra information required by an
implementation of the MSOD algorithm. First, we consider the
generation of overlap. Conveniently, this is identical to the
neighboring states discovery problem discussed in conjunction with
Figure~\ref{fig:diffusion_graph}. To generate the boundary for the
local transport domain, we needed to gather the identification numbers
of all owning processes from the immediately adjacent states in the
system. We solved this problem by traversing the graph of the matrix
for each of the boundary states and determining which processes owned
those adjacent states.

To generate overlap, we use the identical algorithm but in this case
we perform as many graph level traversals as the specified amount of
overlap and collect additional information. For the overlapping domain
decomposition case, given an input parallel decomposition for the
Neumann-Ulam decomposition, for all local states $m$, including those
in the overlap, we require access to $P_{mn}$ and $W_{mn}$ (or
$P^T_{mn}$ and $W^T_{mn}$ for the adjoint method) for all possible
states $n$. Doing this provides us with the data required by the
estimators permitting $P_{mn}$ and $W_{mn}$ to be computed locally for
any boundary transition.

As an example, consider the analytic relations derived in
Appendix~\ref{chap:parallel_theory} for the length of random walks in
a Neumann-Ulam method and the amount of leakage from a domain
considering a neutron diffusion problem. From these relations we might
determine that, for our particular problem, if we can grow the domain
by 10 additional discrete state transitions, we can reduce the number
of histories that must be communicated by a significant fraction. We
determine the information we must gather from neighboring domains to
build the overlap by traversing the graph of the matrix outward from
the boundary 10 levels. At each step, we find the data in the adjacent
domain that we require and make that data local, incrementing the size
of the overlap. Once the overlap has been gathered, one extra graph
traversal on the boundary states is performed to get the new
neighboring states and their owning processes.

Once overlap has been generated, the transport domain for a single set
is complete. However, if multiple sets are to be used, an additional
set of parallel procedures is required to generate the necessary data
structures. Figure~\ref{fig:msod_construction} gives a schematic
representation of the MSOD construction process using 4 sets. For a
problem where $P$ parallel processors are available and $S$ sets are
to be used, each set is allocated $P/S$ processors for computation. On
the first $P/S$ processors, the linear problem is generated. From the
linear problem, the linear operator is used to construct the transport
domain with overlap, the solution vector used to construct the
tallies, and the forcing term vector used to construct the fixed
source. Once these Monte Carlo data structures have been generated,
they are broadcast to the remaining sets in the problem as shown in
Figure~\ref{fig:msod_construction} such that we have $S$ replications
of the Monte Carlo problem for a single instance of the original
linear problem.

\clearpage

\begin{figure}[t!]
  \begin{center}
    \scalebox{0.55}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
  \end{center}
  \caption{\textbf{MSOD construction for 4 sets with overlap.}
    \textit{The linear system is used to construct the Monte Carlo
      transport domain on the primary set. The Monte Carlo data
      structures are then broadcast among the blocks to all sets in
      the system.}}
  \label{fig:msod_construction}
\end{figure}

\subsubsection{MSOD Neumann Ulam Algorithm}
\label{subsubsec:msod_algorithm}

With the ability to generate the transport domain within an MSOD
decomposition as well as the source and tally structures, we can now
define how to perform a Neumann-Ulam solve using MSOD. Given by
Algorithm~\ref{alg:msod_transport}, we begin by constructing the Monte
Carlo data structures for each set including the transport domain, the
source, and the tallies. Next, in line 4 we perform the asynchronous
transport algorithm presented in \S~\ref{subsec:asynchronous_algorithm}
for each of the individual sets. Adding overlap to
Algorithm~\ref{alg:parallel_mc_algorithm} in this case simply modifies
the set of local states and the set of neighboring states with the
additional states gathered during overlap generation.

\begin{algorithm}[h!]
  \caption{\textbf{MSOD Transport Sequence}}
  \label{alg:msod_transport}
  \begin{algorithmic}[1]
    \State build MSOD domain
    \State build Monte Carlo source
    \State build Monte Carlo tally
    \State perform parallel Neumann-Ulam transport in each set
    \State combine set tallies
    \State combine block tallies
  \end{algorithmic}
\end{algorithm}

Once Monte Carlo transport is complete, the tallies for each
individual computation must be gathered to the original set in order
to build the correction vector within an MCSA sequence. To do this, we
perform two parallel vector operations. First, overlap not only
creates additional states in the local transport domain but also
additional states in the local tally vector such that any events that
occur in the local domain may be tallied in a local vector. Because of
this, the tally vector in a single set will share local states with
other domains and a parallel sum reduction operation is required to
transform the tally into the original Neumann-Ulam parallel
decomposition. Second, the tallies in individual sets must be combined
and applied to the primary set in the problem that shares the
processor space with the original linear problem. For this operation,
we employ the concept of blocks as a subdivision of parallel space
complementary to the concept of sets.

Consider the schematic representation of the final parallel reduction
presented in Figure~\ref{fig:msod_tally_reduction}. For this
operation, the tally vectors in each set must be summed together and
applied to the primary set. As each set is an exact replica of the
others, the parallel decomposition of the tally vector is the same
within all sets. We can take advantage of this by building a processor
space that encapsulates each identical domain across all sets and then
use this processor space for the reduction. For the schematic in
Figure~\ref{fig:msod_tally_reduction}, the combined upper-left domains
in each set create a block, giving 9 total blocks for this example
problem. The processor space that encapsulates each of the upper-left
domains is then used for the parallel reduction operation. This
reduction will happen in a mutually exclusive manner for each of the
blocks in the system.

\begin{figure}[t!]
  \begin{center}
    \scalebox{0.55}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
  \end{center}
  \caption{\textbf{MSOD tally reduction over blocks.} \textit{The
      tally computed independently in each set is reduced across the
      blocks to the primary set. The linear system vector is updated
      on the primary set. In this example the tally vector in the
      block containing upper-left subdomains is combined. Identical
      reduction operations occur for all other blocks
      simultaneously.}}
  \label{fig:msod_tally_reduction}
\end{figure}

%%---------------------------------------------------------------------------%%
\subsection{Parallel MCSA}
\label{subsec:parallel_mcsa}
With the parallel adjoint Neumann-Ulam solver implementation described
above, the parallel implementation of an MCSA iteration is
trivial. Recall the MCSA iteration procedure presented again here for
clarity:
\begin{subequations}
  \begin{gather}
    \label{eq:mcsa_r1}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \label{eq:mcsa_r2}
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \label{eq:mcsa_r3}
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \label{eq:mcsa_r4}
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \label{eq:mcsa_r5}
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
  \end{gather}
\end{subequations}
In \S\ref{subsec:parallel_krylov_methods} we discuss parallel matrix and
vector operations as utilized in conventional projection methods. We
utilize these here for the parallel MCSA implementation for the
computations required in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
(\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}). In the first step, a
parallel vector update is used to apply the residual to the previous
iterate's solution in a Richardson iteration. In the next step and
Eq~(\ref{eq:mcsa_r5}), the residual is computed by a parallel
matrix-vector multiply and vector update. Once the correction is
computed with a parallel adjoint Neumann-Ulam solve, this correction
is applied to the solution with a parallel vector
update. Additionally, as given by
Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
will be required to check the stopping criteria: one initially to
compute the infinity norm of the source vector, and another at every
iteration to compute the infinity norm of the residual vector.

To parallelize the solution of Eq~(\ref{eq:mcsa_r3}), we apply the
MSOD Neumann-Ulam algorithm presented in the previous subsection. The
problem is potentially replicated with overlap and the Monte Carlo
procedure returns a correction vector, $\delta \ve{x}$, with the same
parallel decomposition as the input linear problem. As was the case
for generating the replications required for a multiple set
implementation, the explicit matrix and vector operations required for
the rest of an MCSA iteration occur only on the subset of processors
containing the primary set\footnote{For applications where resiliency
  is potentially of concern, the linear problem and subsequently the
  operations in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
  (\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}) may be replicated with
  the penalty of broadcasting the data. Doing so may alleviate issues
  where any part of the linear problem or residual is disturbed as now
  there will be multiple copies of these data structures.}. Only the
Monte Carlo data structures exist on the rest of the processors in the
problem as replication of the other operations does not generate any
more information that could be used to accelerate the time to
solution.

%%---------------------------------------------------------------------------%%
\section{Conclusion}
\label{sec:conclusion}

In this paper, the Monte Carlo Synthetic Acceleration method has been
applied to the $SP_N$ discretization of the neutron transport equation
and explored in the context of a difficult nuclear fuel assembly
criticality problem. We observed that MCSA can in fact solve the
asymmetric system generated by the $SP_N$ equations. In contrast to
the thermal radiation diffusion problem addressed in our previous
work, light water reactor problems are difficult to solve with MCSA as
they have large spectral radii due to the neutron scattering in the
moderator. Because of this, simple Jacobi-based preconditioning
reduces the spectral radius of the $SP_N$ system below unity, however,
this alone was discovered not to be sufficient criteria alone for MCSA
convergence. As the spectral radius approaches unity, MCSA was
demonstrated to break down with more stochastic histories required for
convergence and more CPU time required per history. To alleviate these
effects, a left-right preconditioned form of the MCSA method was
developed and applied to the $SP_N$ equations to obtain
convergence. For the nuclear fuel assembly problem, MCSA was observed
to converge in fewer iterations per eigenvalue iteration than GMRES
for the fuel assembly criticality problem and more than Bi-CGStab
using the same preconditioning. However, the explicit preconditioning
strategy required to overcome the MCSA spectral radius restriction
creates run times $O(100)$ larger than the production Krylov methods
for the fuel assembly criticality problem, indicating that significant
research in the area of Monte Carlo preconditioning implementations is
required in order for the method to be competetive in terms of time to
solution.

%%---------------------------------------------------------------------------%%
\pagebreak
\bibliographystyle{ieeetr} 
\bibliography{references}
\end{document}

### 2014-09-17 doc/ttsp2013/paper/ttsp_2013.tex
adding c5g7 7-group cross sections

@@
 %%---------------------------------------------------------------------------%%
-\author{Stuart R. Slattery, Steven P. Hamilton, Thomas M. Evans, and Paul
-  P.H. Wilson
+\author{Stuart R. Slattery, Steven P. Hamilton, and Thomas M. Evans
   \\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}} }
@@
 the Boltzmann neutron transport equation. Using the fully-formed
-linear operator for the transport problem, we explore solutions to the
-$SP_N$ equations with MCSA using criticality calculations driving
-problem. Due to neutron scattering in the light water moderator,
-several difficulties arise when applying MCSA to these problems that
-were not observed in previous work where the method was applied to
-more opaque thermal radiaton transport systems. Of most significance,
-it was discovered for these problems that simply reducing the spectral
-radius of the Monte Carlo problem below unity is not enough to
-efficiently achieve convergence as reported in previous work. In order
-to effectively solve the $SP_N$ equations for the model problem, an
-algebraic preconditioning strategy is developed for MCSA where we
-utilize sparse approximate inverse representations of the operator as
-preconditioners to achieve convergence. Using the preconditioners,
-MCSA performance is compared with production implementations of
-preconditioned Krylov methods. Our studies indicate that
-the iterative performance of MCSA when used to solve the $SP_N$
+linear operators for the transport problem, we explore solutions to
+the $SP_N$ equations with MCSA using criticality calculations as the
+driving problems. Due to neutron scattering in the light water
+moderator, several difficulties arise when applying MCSA to these
+problems that were not observed in previous work where the method was
+applied to more opaque thermal radiaton transport systems. Of most
+significance, it was discovered for these problems that simply
+reducing the spectral radius of the Monte Carlo problem below unity is
+not enough to efficiently achieve convergence as reported in previous
+work. In order to effectively solve the $SP_N$ equations for the model
+problem, an algebraic preconditioning strategy is developed for MCSA
+where we utilize sparse approximate inverse representations of the
+operator as preconditioners to achieve convergence. Using the
+preconditioners, MCSA performance is compared with production
+implementations of preconditioned Krylov methods. Our studies indicate
+that the iterative performance of MCSA when used to solve the $SP_N$
 equations for light water reactor problems is competitive with the
@@
 %%---------------------------------------------------------------------------%%
-\section{Criticality Calculations}
-
-\subsection{C5G7 Benchmark}
-
-\subsection{Light Water Reactor Fuel Assemblies}
-
-%%---------------------------------------------------------------------------%%
 \section{Parallel MCSA}
@@
 %%---------------------------------------------------------------------------%%
+\section{Criticality Calculations}
+
+\subsection{C5G7 Benchmark}
+
+\subsection{Light Water Reactor Fuel Assemblies}
+
+%%---------------------------------------------------------------------------%%
 \section{Conclusion}

### 2014-10-17 doc/emory_seminar_2014/emory_talk.tex
Adding Emory talk.

%%---------------------------------------------------------------------------%%
%% Shift Roadmap Presentation
%%---------------------------------------------------------------------------%%

\documentclass{beamer}
\usefonttheme[onlymath]{serif}
\usepackage{colortbl}
\usepackage{longtable,ltcaption,array}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{caption}
\usepackage[labelformat=empty]{subcaption}
\usepackage{tabularx}

%%---------------------------------------------------------------------------%%
%% THEME SETUP

\usetheme{CambridgeUS}
\usecolortheme{spruce}

%%---------------------------------------------------------------------------%%
%% SETUP STUFF
%%---------------------------------------------------------------------------%%

\logo{\includegraphics[width=2cm]{new_logo}}

\setbeamercolor{item}{fg=MSUgreen}
\setbeamertemplate{headline}{}
\setbeamertemplate{navigation symbols}{}

\newlength{\DUtablewidth} % internal use in tables

%%---------------------------------------------------------------------------%%
%% MATH STUFF
%%---------------------------------------------------------------------------%%

\newcommand{\Pn}{\textit{P}$\negthinspace_N$}
\newcommand{\SPn}{\textit{SP}$\negthinspace_N$}

\newcommand{\vOmega}{\ensuremath{\ve{\Omega}}}
\newcommand{\hOmega}{\ensuremath{\hat{\ve{\Omega}}}}

\newcommand{\sigs}{\ensuremath{\sigma_{\text{s}}}}
\newcommand{\sigf}{\ensuremath{\sigma_{\text{f}}}}
\newcommand{\sigsm}{\ensuremath{\sigma_{\text{s}m}}}
\newcommand{\sigsn}{\ensuremath{\sigma_{\text{s}n}}}

\newcommand{\phig}{\ensuremath{\Phi}}
\newcommand{\Sigmag}{\ensuremath{\mathbf{\Sigma}}}
\newcommand{\Dg}{\ensuremath{\mathbb{D}}}
\newcommand{\Fg}{\ensuremath{\mathbb{F}}}
\newcommand{\Qg}{\ensuremath{\mathbb{Q}}}
\newcommand{\ug}{\ensuremath{\mathbb{U}}}
\newcommand{\Ag}{\ensuremath{\mathbb{A}}}
\newcommand{\Bg}{\ensuremath{\mathbb{B}}}
\newcommand{\Cg}{\ensuremath{\mathbb{C}}}
\newcommand{\Jg}{\ensuremath{\mathbb{J}}}
\newcommand{\sg}{\ensuremath{\mathbb{S}}}

\newcommand{\aphig}{\ensuremath{\Phi^{\dagger}}}
\newcommand{\aSigmag}{\ensuremath{\mathbf{\Sigma}^{\dagger}}}

\newcommand{\apsi}[1]{\ensuremath{\psi^{\dagger\,#1}}}
\newcommand{\aphi}[1]{\ensuremath{\phi^{\dagger\,#1}}}
\newcommand{\aq}[1]{\ensuremath{q^{\dagger\,#1}}}
\newcommand{\ak}{\ensuremath{k^{\dagger}}}

\newcommand{\normal}{\ensuremath{\hat{\ve{n}}}}

% Phantom minus sign for helping with alignment of matrices
\newcommand{\phmin}{\ensuremath{\phantom{-}}}

\DeclareMathOperator{\diag}{diag}

%%---------------------------------------------------------------------------%%
%% TITLE

\title[Monte Carlo Linear Solvers]{Monte Carlo Linear Solvers}
\author[Hamilton, Evans, Slattery]{Steven Hamilton \\ Tom Evans \\ Stuart Slattery}
\date[Emory Seminar]{Emory Seminar \\ September 19, 2014}
\institute[ORNL]{Oak Ridge National Lab}
\titlegraphic{\includegraphics[width=1.5in]{ORNL_Balls.pdf}}

\defbeamertemplate{footline}{titlefoot}{
    \vspace{-0.5cm}
    \hspace*{0.25cm}
    \includegraphics[width=2cm]{doe_logo}
    \hfill
    \includegraphics[width=3.25cm]{WordMarkLeaf}
    \hspace*{0.5cm}
}

\setbeamertemplate{title page}
{
  \begin{tabular}{cr}
    \begin{minipage}{4.7cm}
      {\bf \Large\textcolor{MSUgreen}{\inserttitle}}\\

      \vspace{2\baselineskip}
      \insertauthor\\

      \insertinstitute\\

      \insertdate
    \end{minipage}
    &
    \begin{minipage}{5cm}
      \raggedright
      \vspace{-0.75cm}
      \includegraphics[height=9.8cm]{combined_background}
    \end{minipage}
  \end{tabular}
}

%%---------------------------------------------------------------------------%%
\begin{document}

%%---------------------------------------------------------------------------%%

{
\setbeamertemplate{logo}{}
\setbeamertemplate{footline}[titlefoot]
\begin{frame}
\titlepage
\end{frame}
}

%%---------------------------------------------------------------------------%%
\begin{frame}{Motivation}
  \vfill
\begin{itemize}
  \item As we move towards exascale computing, the rate of errors is expected
  to increase dramatically
  \vfil
  \begin{itemize}
    \item The probability that a compute node will fail during the course
      of a large scale calculation may be near 1
  \end{itemize}
  \vfill
  \item Algorithms need to not only have increased concurrency/scalability
    but have the ability to recover from hardware faults
\end{itemize}
  \vfill
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Towards Exascale Concurrency and Resiliency}
  \begin{itemize}
    \item Two basic strategies:
      \vfill
      \begin{enumerate}
        \item State with current ``state of the art'' methods and make
          incremental modifications to improve scalability and fault
          tolerance
          \begin{itemize}
            \item Many efforts are heading in this direction, attempting
              to find additional concurrency to exploit
          \end{itemize}
        \vfill
        \item Start with methods having natural scalability and resiliency
          aspects and work at improving performance
          \begin{itemize}
            \item One possibility: Monte Carlo methods
          \end{itemize}
      \end{enumerate}
  \end{itemize}
  \vfill
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo for Linear Systems}
  \begin{itemize}
    \item Suppose we want to solve $\mathbf{Ax}=\mathbf{b}$
    \vfill
    \item If $\rho(\mathbf{I-A})<1$, we can write the solution using the
      Neumann series
      \begin{equation*}
        \mathbf{x} = \sum_{n=0}^{\infty} (\mathbf{I-A})^n \mathbf{b}
         = \sum_{n=0}^{\infty} \mathbf{H}^n \mathbf{b}
      \end{equation*}
      where $\mathbf{H} \equiv ( \mathbf{I-A} )$ is Richardson iteration matrix
    \vfill
    \item Traverse the Neumann series stochastically
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Forward Monte Carlo}
\begin{itemize}
  \item Choose a row-stochastic matrix $\mathbf{P}$ and weight matrix
    $\mathbf{W}$ such that $\mathbf{H} = \mathbf{P} \circ \mathbf{W}$
  \item Typical choice (Monte Carlo Almost-Optimal):
    \begin{equation*}
      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ij}| }
      {\sum_{j=1}^{N} | \mathbf{H}_{ij} | }
    \end{equation*}
  \item To compute solution component $\mathbf{x}_i$:
    \begin{itemize}
      \item Start a history in state $i$ (with initial weight of 1)
      \item Transition to new state $j$ based probabilities determined by
        $\mathbf{P}_i$
      \item Modify history weight based on corresponding entry in
        $\mathbf{W}_{ij}$
      \item Add contribution to $\mathbf{x}_i$ based on current history weight
        and value of $\mathbf{b}_j$
    \end{itemize}
  \item A given random walk can only contribute to a single component of
    the solution vector
\end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Adjoint Monte Carlo}
\begin{itemize}
  \item Choose $\mathbf{P}$ and $\mathbf{W}$ such that
    $\mathbf{H}^{T} = \mathbf{P} \circ \mathbf{W}$
  \item Typical choice (Monte Carlo Almost-Optimal):
    \begin{equation*}
      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ji}| }
      {\sum_{i=1}^{N} | \mathbf{H}_{ji} |}
    \end{equation*}
  \item To estimate solution:
    \begin{itemize}
      \item Start a history in random state $i$ by sampling from distribution
        determined by $\mathbf{b}$
      \item Transition to new state $j$ based probabilities determined by
        $\mathbf{P}_i$
      \item Modify history weight based on corresponding entry in
        $\mathbf{W}_{ij}$
      \item Add contribution to $\mathbf{x}_j$ based on current history weight
    \end{itemize}
  \item A given random walk can contribute to many different components of
    the solution vector
\end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Example (Forward Monte Carlo)}
  \begin{itemize}
    \item Suppose
  \begin{equation*}
    \mathbf{A} = \begin{bmatrix}
      \phmin 1.0 & -0.2 & -0.6 \\
      -0.4 & \phmin 1.0 & -0.4 \\
      -0.1 & -0.4 & \phmin 1.0 \end{bmatrix} \to
    \mathbf{H} \equiv (\mathbf{I-A}) = \begin{bmatrix}
       0.0 &  0.2 &  0.6 \\
       0.4 &  0.0 &  0.4 \\
       0.1 &  0.4 &  0.0 \end{bmatrix}
  \end{equation*}
    then
  \begin{equation*}
    \mathbf{P} = \begin{bmatrix}
       0.0 & 0.25 & 0.75 \\
       0.5 &  0.0 & 0.5 \\
       0.2 &  0.8 & 0.0 \end{bmatrix}, \quad
    \mathbf{W} = \begin{bmatrix}
       0.0 &  0.8 &  0.8 \\
       0.8 &  0.0 &  0.8 \\
       0.5 &  0.5 &  0.0 \end{bmatrix}
  \end{equation*}
    \vfill
    \item If a history is started in state $3$, there is a $20\%$ chance of
      it transitioning to state $1$ and an $80\%$ chance of moving to state
      $2$
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Solving Heat Equation: Forward Method}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics<1>[width=4in]{direct_1.png}
      \includegraphics<2>[width=4in]{direct_10.png}
      \includegraphics<3>[width=4in]{direct_100.png}
      \includegraphics<4>[width=4in]{direct_1000.png}
    \end{center}
    \caption{
      \only<1>{\textbf{Forward solution.}
        \textit{\sn{2.5}{3} total histories.} }
      \only<2>{\textbf{Forward solution.}
        \textit{\sn{2.5}{4} total histories.} }
      \only<3>{\textbf{Forward solution.}
        \textit{\sn{2.5}{5} total histories.} }
      \only<4>{\textbf{Forward solution.}
        \textit{\sn{2.5}{6} total histories.} }
    }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Solving Heat Equation: Adjoint Method}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics<1>[width=4in]{adjoint_1.png}
      \includegraphics<2>[width=4in]{adjoint_10.png}
      \includegraphics<3>[width=4in]{adjoint_100.png}
      \includegraphics<4>[width=4in]{adjoint_1000.png}
      \includegraphics<5>[width=4in]{adjoint_10000.png}
      \includegraphics<6>[width=4in]{adjoint_100000.png}
      \includegraphics<7>[width=4in]{adjoint_1000000.png}
      \includegraphics<8>[width=4in]{adjoint_10000000.png}
    \end{center}
    \caption{
      \only<1>{\textbf{Adjoint solution.}
        \textit{\sn{1}{0} total histories.} }
      \only<2>{\textbf{Adjoint solution.}
        \textit{\sn{1}{1} total histories.} }
      \only<3>{\textbf{Adjoint solution.}
        \textit{\sn{1}{2} total histories.} }
      \only<4>{\textbf{Adjoint solution.}
        \textit{\sn{1}{3} total histories.} }
      \only<5>{\textbf{Adjoint solution.}
        \textit{\sn{1}{4} total histories.} }
      \only<6>{\textbf{Adjoint solution.}
        \textit{\sn{1}{5} total histories.} }
      \only<7>{\textbf{Adjoint solution.}
        \textit{\sn{1}{6} total histories.} }
      \only<8>{\textbf{Adjoint solution.}
        \textit{\sn{1}{7} total histories.} }
    }
  \end{figure}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Limitations of Monte Carlo}
\begin{itemize}
  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
    is generally intractable
    \begin{itemize}
      \item Central limit theorem is barrier to accurate solutions
    \end{itemize}
\end{itemize}
\begin{figure}
  \centering
  \includegraphics[width=3.5in]{Ifiss_ConvDiff}
\end{figure}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Beyond the Central Limit Theorem}
  \begin{itemize}
    \item Instead of directly solving $\mathbf{Ax}=\mathbf{b}$ with Monte
      Carlo, apply Monte Carlo to residual equation
      $\mathbf{A}\delta = \mathbf{r}$
    \vfill
    \item Preconditioned Richardson iteration using Monte carlo as
      preconditioner:
      \begin{align*}
        \mathbf{r}^{k} &= \mathbf{b} - \mathbf{Ax}^{k} \\
        \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k} \\
        \mathbf{x}^{k+1} &= \mathbf{x}^k + \delta
      \end{align*}
    \vfill
    \item This is the sequential Monte Carlo method devised by Halton
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Further Beyond the Central Limit Theorem}
  \begin{itemize}
    \item Combine with Richardson iteration as a ``smoother'' in between
      Monte Carlo steps:
      \begin{align*}
        \mathbf{r}^k &= \mathbf{b} - \mathbf{Ax}^k \\
        \mathbf{x}^{k+1/2} &= \mathbf{x}^k + \mathbf{r}^k \\
        \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} \\
        \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k+1/2} \\
        \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta
      \end{align*}
    \vfill
    \item We call this algorithm Monte Carlo Synthetic Acceleration (MCSA)
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Algorithm Comparison}
  \begin{figure}
    \includegraphics[width=3.5in]{seq_conv_100}
  \end{figure}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Recent Success}
\begin{itemize}
  \item T. Evans, S. Mosher, S. Slattery, S. Hamilton, ``A Monte Carlo
    synthetic-acceleration method for solving the thermal radiation diffusion
    equation,'' Journal of Computational Physics \textbf{258},
    pp.~338--358 (2014).
  \item Applied MCSA to time-dependent thermal radiation diffusion equation
\end{itemize}
\begin{table}
  \centering
  \begin{tabular}{cccc}
    \toprule
    \multirow{2}{*}{\bfseries Solver} &
    \bfseries Total &
    \bfseries Relative Time &
    \bfseries Relative Total \\
    & \bfseries Iterations &
    \bfseries Per Iteration &
    \bfseries Time \\
    \midrule
    CG (ML) & 12150 & 74.7 & 2.7 \\
    GMRES (ILU) & 40294 & 103.0 & 9.2 \\
    MCSA & 98863 & 1.0 & 1.0 \\
    \bottomrule
  \end{tabular}
\end{table}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Path to Harder Problems}
  \begin{itemize}
    \item Experience with radiation diffusion suggests that for problems
      where spectral radius is low ($\lesssim 0.8$),
      MCSA can be competitive with leading methods
      \begin{itemize}
        \item Very low cost per iteration to approximate Neumann series
      \end{itemize}
    \vfill
    \item What about more challenging problems?
      \begin{itemize}
        \item Neumann series converges more slowly, so histories last longer
        \item More histories may be required
        \item What if Neumann series is not convergent?
      \end{itemize}
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Stopping Criteria -- Weight Cutoff}
  \begin{itemize}
    \item Histories are usually terminated when the relative weight drops
      below a specified cutoff (i.e. $\frac{w}{w_0} < \tau$)
    \vfill
    \item Adjoint MCSA on JPWH\_991 matrix with diagonal preconditioning
      ($\rho ( \mathbf{I-D^{-1}A} ) \approx 0.98$):
  \end{itemize}
\begin{figure}
  \centering
  \includegraphics[width=2.5in]{WtCutoff}
\end{figure}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Stopping Criteria -- History Length Cutoff}
  \begin{itemize}
    \item Somewhat counterintuitive result: using larger weight cutoff often
      results in better MCSA convergence
      \begin{itemize}
        \item A good approximation to a few terms in the Neumann series
          performs better than a statistically noisy approximation to
          full series
      \end{itemize}
    \vfill
    \item If statistical performance is better for large weight cutoff,
      what happens if we simply terminate histories after prescribed number
      of steps
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Weight vs. History Length Cutoff}
\begin{figure}
  \centering
  \includegraphics[width=3in]{WtVsLength}
\end{figure}
\begin{itemize}
  \item Unlike weight cutoff, history length cutoff ``saturates'' at some
    number of histories -- fixed length Neumann series has been perfectly
    reproduced
  \item Saturation point is higher for longer history lengths
\end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo as Approximate Polynomial Preconditioning}
  \begin{itemize}
    \item Using history length cutoff, Monte Carlo process is approximating
      a fixed length Neumann series polynomial
    \vfill
    \item As number of histories grows, iteration count becomes identical to
      using ``true'' polynomial
    \vfill
    \item Why limit ourselves to the Neumann series polynomial?
      \begin{itemize}
        \item Chebyshev or GMRES polynomials are viable alternatives
      \end{itemize}
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Neumann Series Polynomial}
\begin{table}
\caption{Adjoint MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_adjoint_neumann}}
\centering
\begin{tabular}{cccc}
\toprule
Histories per & \multicolumn{3}{c}{Polynomial Order} \\
\cmidrule(lr){2-4}
Iteration & 2 & 4 & 6 \\
\midrule
250 &  775(100) & 651(96) & 664(110) \\
500 &  725(122) & 502(104) & 394(95) \\
1000 & 707(174) & 482(179) & 366(144) \\
2000 & 703(280) & 471(259) & 356(251) \\
4000 & 698(494) & 467(497) & 350(458) \\
8000 & 697(923) & 464(905) & 350(873) \\
16000 & 695(1796) & 462(1768) & 347(1711) \\
\bottomrule
\end{tabular}
\end{table}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Chebyshev Polynomial}
\begin{table}
\caption{Adjoint MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_adjoint_cheby}}
\centering
\begin{tabular}{cccc}
\toprule
Histories per & \multicolumn{3}{c}{Polynomial Order} \\
\cmidrule(lr){2-4}
Iteration & 1 & 2 & 3 \\
\midrule
250 & - & - & - \\
500 & - & - & - \\
1000 & - & - & - \\
2000 & - & 328(134) & - \\
4000 & - & 296(210) & - \\
8000 & - & 288(380) & 262(423) \\
16000 & 1132(1866) & 283(721) & 175(550) \\
\bottomrule
\end{tabular}
\end{table}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{GMRES Polynomial}
\begin{table}
\caption{Adjoint MCSA with GMRES Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_adjoint_gmres}}
\centering
\begin{tabular}{cccc}
\toprule
Histories per & \multicolumn{3}{c}{Polynomial Order} \\
\cmidrule(lr){2-4}
Iteration & 2 & 4 & 6 \\
\midrule
250 &  2572(337) & - & - \\
500 &  616(105)  & - & - \\
1000 & 566(142)  &  - & - \\
2000 & 557(222)  &  - & - \\
4000 & 554(412)  &  203(236) & - \\
8000 & 549(731)  &  184(423) & - \\
16000 & 548(1426) & 247(965) & 146(744) \\
\bottomrule
\end{tabular}
\end{table}

\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Polynomial Methods -- Summary}
  \begin{itemize}
    \item Using history length cutoff rather than weight cutoff can lead to
      large reductions in iteration counts
    \vfill
    \item Significant reduction in iteration counts are possible using
      alternate polynomials, but generally outweighed by increase in number
      of histories required
      \begin{itemize}
        \item May be very beneficial from resiliency and parallel efficiency
          standpoints
        \item Will be re-evaluated in future efforts
      \end{itemize}
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Preconditioning}
  \begin{itemize}
    \item If a non-trivial preconditioner is used, the Monte Carlo problem
      must be formed based on preconditioned iteration matrix,
      $\mathbf{H} = \mathbf{I-M^{-1}A}$
    \vfill
    \item Requires explicit construction of preconditioned matrix
      \begin{itemize}
        \item This has generally limited preconditioner selection to diagonal,
          block diagonal, sparse approximate inverse approaches
      \end{itemize}
    \vfill
    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
      in $\mathbf{M^{-1}A}$
    \vfill
    \item Need to investigate preconditoning strategies that lead to sparse
      preconditioned systems (Michele/Max)
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}[fragile]{Multiple-Set Overlapping-Domain Decomposition}

  \vspace{-1cm}
  \begin{columns}

    \begin{column}{0.52\textwidth}

      \begin{figure}[t!]
        \begin{center}
          \scalebox{0.275}{ \input{msod_construction.pdftex_t} }
        \end{center}
        \caption{\small MSOD construction.}
      \end{figure}

    \end{column}

    \begin{column}{0.48\textwidth}

      \begin{figure}[t!]
        \begin{center}
          \scalebox{0.2}{ \input{msod_tally.pdftex_t} }
        \end{center}
        \caption{\small MSOD tally reduction.}
      \end{figure}

      \begin{itemize}
        {\small
        \item Multiple sets replicate the domain
        \item Domains overlap within a set
        \item Each set contains the full domain
        }
      \end{itemize}

    \end{column}

  \end{columns}

%  \let\thefootnote\relax\footnote{\tiny{Wagner et. al., "Hybrid and
%      parallel domain-decomposition methods development to enable
%      Monte Carlo for reactor analysis", Joint International
%      Conference on Supercomputing in Nuclear Applications and Monte
%      Carlo (SNA+MC 2010), 2010.}}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Alternative Parallelism -- Additive Schwarz}
  \begin{itemize}
    \item Instead of performing Monte Carlo on full problem, another
      possibility is to apply Monte Carlo as an additive Schwarz approach
    \vfill
    \item Decompose problem into (possibly overlapping) domains
    \vfill
    \item Perform Monte Carlo on individual subdomains
      \begin{itemize}
        \item No communication costs in Monte Carlo problem!
      \end{itemize}
    \vfill
    \item With domain decomposed Monte Carlo, iteration counts are effectively
      independent of the number of processors
    \vfill
    \item In an additive Schwarz approach, the preconditioner will become
      less effective as processor counts grow -- algorithmic scalability
      may be an issue
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{More Parallelism -- Threading}
  \begin{itemize}
    \item Within a Monte Carlo solve, every history is independent of other
      histories -- great potential for highly concurrent hardware
      (GPU, Xeon Phi)
    \vfill
    \item Early experiments using the Trilinos Kokkos library show promising
      performance for multicore CPUs
    \vfill
    \item Will be taking part in OLCF ``Hackathon'' in late October to begin
      implementing computation kernels in OpenACC to allow GPU capability
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Application -- Nuclear Reactor Analysis}
  \begin{itemize}
    \item The simplified $P_N$ ($SP_N$) equations are an approximation
      to the Boltzmann neutron transport equation used to simulate nuclear
      reactors
      \begin{itemize}
        \item We are interested in solving generalized eigenvalue problem,
          for this study we use Arnoldi as the eigensolver and compare
          different methods for solving linear systems
      \end{itemize}
    \vfill
    \item Test problem -- $3 \times 3$ array of fuel assemblies with
      control rod in center location
      \begin{itemize}
        \item 23 energy groups, 2 angular moments, 25M degrees of freedom
        \item 1000 computational cores
      \end{itemize}
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{SPN Solution}
  \begin{figure}
    \centering
    \includegraphics[width=3in]{prob4}
  \end{figure}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{SPN Results}
\begin{table}
\centering
\begin{tabular}{cccc}
\toprule
%Method & Total GMRES Iterations & Setup Time (s) & Solve Time (s) \\
\multirow{2}{*}{\bfseries Method} &
\bfseries Total GMRES & \bfseries Setup & \bfseries Solve \\
& \bfseries Iteration & \bfseries Time (s) & \bfseries Time (s) \\
\midrule
GMRES-ILUT      & 1675 & 0.7  & 18.4 \\
GMRES-AMG       & 626  & 0.7  & 46.0 \\
GMRES-MGE       & 498  & 1.5  & 33.7 \\
Richardson-AINV & 5208 & 20.6 & 52.0 \\
MCSA-AINV       & 1268 & 25.5 & 46.6 \\
\bottomrule
\end{tabular}
\end{table}
\begin{itemize}
  \item ILUT preconditioning is winner here, but known to have issues
    with parallel scaling on large core counts
  \item Solve times for MCSA are competitive, but setup times are very large
    due to types of preconditioning necessary
\end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Additional Thoughts}
  \begin{itemize}
    \item Our current implementations rely on performing fixed number of
      histories per Monte Carlo solve
      \begin{itemize}
        \item How can we dynamically select/adapt the number of histories
          that should be performed?  Use variance-based stopping criteria?
      \end{itemize}
    \vfill
    \item The ``almost optimal'' selection of the probability and weight
      matrices is arbitrary (as long as
      $\mathbf{H}/\mathbf{H}^T=\mathbf{P} \circ \mathbf{W}$)
      \begin{itemize}
        \item If $\mathbf{P}$ is taken to have a uniform distribution within
          each row, then sampling from the distribution can be done in
          constant (rather than logarithmic) time
      \end{itemize}
    \vfill
    \item Anderson acceleration of MCSA (Tim Kelly @ NCSU)
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Conclusions}
  \begin{itemize}
    \item Monte Carlo methods offer great potential for both resilient and
      highly parallel solvers
    \vfill
    \item For certain classes of problems, Monte Carlo methods can be
      competitive with leading modern solvers
    \vfill
    \item Extending methods to broader problem areas is significant challenge
      and an attractive area for continued research
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\end{document}
%%---------------------------------------------------------------------------%%

%%---------------------------------------------------------------------------%%
%% end of pres.tex
%%---------------------------------------------------------------------------%%

### 2014-12-04 doc/ttsp2013/paper/ttsp_2013.tex
cleaning up abstract and introduction in ttsp paper

@@
 linear operators for the transport problem, we explore solutions to
-the $SP_N$ equations with MCSA using criticality calculations as the
-driving problems. Due to neutron scattering in the light water
-moderator, several difficulties arise when applying MCSA to these
-problems that were not observed in previous work where the method was
-applied to more opaque thermal radiaton transport systems. Of most
-significance, it was discovered for these problems that simply
-reducing the spectral radius of the Monte Carlo problem below unity is
-not enough to efficiently achieve convergence as reported in previous
-work. In order to effectively solve the $SP_N$ equations for the model
-problem, an algebraic preconditioning strategy is developed for MCSA
-where we utilize sparse approximate inverse representations of the
-operator as preconditioners to achieve convergence. Using the
-preconditioners, MCSA performance is compared with production
-implementations of preconditioned Krylov methods. Our studies indicate
-that the iterative performance of MCSA when used to solve the $SP_N$
-equations for light water reactor problems is competitive with the
-preconditioned Krylov methods when operator spectrum is properly
-conditioned. To solve large problems, we have parallelized the method
-by constructing a variant of the multiple-set overlapping-domain
-(MSOD) algorithm for Monte Carlo calculations. We discuss the method
-and then present results indicating its scalability.
+the $SP_N$ equations with MCSA using light water criticality
+calculations as the driving problems. Due to neutron scattering in the
+light water moderator, several difficulties arise when applying MCSA
+to these problems that were not observed in previous work where the
+method was applied to more opaque thermal radiaton transport
+systems. In addition to numerical difficulties, the level of fidelity
+needed for the accurate solution of these problems creates enough
+degrees of freedom to warrant domain decomposition on hundreds to
+thousands of cores in order to obtain both a time-efficient solution
+and enough local memory to solve the problem. To effectively solve the
+$SP_N$ equations for these systems, we take two steps. First, an
+algebraic preconditioning strategy is developed for MCSA to address
+convergence issues. We utilize a sparse approximate inverse
+preconditioner and a sparse matrix-matrix mutliplication kernel to
+maintain element-level access to the preconditioned linear operator
+and to improve iterative performance. Second, we have parallelized the
+Monte Carlo solver by constructing a variant of fully asynchronous
+domain decomposition algorithms for Monte Carlo particle transport to
+enable solutions on a large number of processors. Using the
+preconditioning strategy and domain decomposition algorithm, MCSA
+performance is compared with production implementations of Krylov
+methods and state-of-the-art algebraic preconditioners for two
+eigenvalue problems modeled in three dimensions: a light water reactor
+assembly and the C5G7 criticality benchmark. Our studies indicate that
+the performance of MCSA when used to solve the $SP_N$ equations for
+light water reactor problems in terms of both iterations to converge
+and time to solution can be competitive with the preconditioned Krylov
+methods when the operator spectrum is properly conditioned using our
+sparse approximate inverse strategy. In addition, scalability of the
+new domain decomposition algorithm is compared to traditional methods
+with the scheme demonstrating good performance on thousands of
+processors.
 
@@
 side of the spectrum, low-fidelity diffusion solutions are often
-leveraged as an alternative to compute a solution over the entire
-domain a reactor core. In the 1960's, Gelbard developed an ad-hoc,
-multidimensional extension of the simple single dimension planar $P_N$
-equations that created a system of coupled, diffusion-like equations
-known as the simplified $P_N$ ($SP_N$) equations \cite{gelbard_1960}
-which incorporates additional angular information into a set of
-diffusion-like equations. Further work in both aysmptotic analysis and
-algorithmic development proceeded in the following decades to serve as
-a theoretical basis for the method
+leveraged as an alternative to computing a solution over the entire
+domain a reactor core where diffusion theory is often valid.
+
+In the 1960's, Gelbard developed an ad-hoc, multidimensional extension
+of the simple one-dimensional planar $P_N$ equations that created a
+system of coupled, diffusion-like equations known as the simplified
+$P_N$ ($SP_N$) equations \cite{gelbard_1960} which incorporate
+additional angular information. Further work in both aysmptotic
+analysis and algorithmic development proceeded in the following
+decades to serve as a theoretical basis for the method
 \cite{morel_1996,brantley_simplified_2000}. Recent work has expanded
@@
 with modern linear algebra techniques including Krylov solvers and
-advanced precondtioners including multigrid methods and incomplete
+advanced preconditioners including multigrid methods and incomplete
 factorizations \cite{hamilton_2014}. Using these new techniques, the
@@
 savings in both compute time and memory required and permit larger
-simulations to be performed on the available computing platforms.
+simulations to be performed on the available computing platforms using
+standard tools.
 
@@
 \cite{kogge_using_2011}. The end result of these hardware changes is
-that the larger number of low-powered processors will be prone to both
-soft failures such as bit errors in floating point operations and hard
-failures where the data owned by that processor cannot be recovered
+that the large number of low-powered processors populating the machine
+will be prone to both soft failures such as bit errors in floating
+point operations and hard failures where the data owned by that
+processor cannot be recovered
 \cite{u.s._department_of_energy_resilient_2012}. Because these
 failures are predicted to be common, resilient solver technologies are
-are a possible means of overcoming these events at the application
-level. With solutions to deterministic transport problems based on
-Monte Carlo techniques, such issues are potentially alleviated by
-statistical arguments. In the case of soft failures, isolated floating
-point errors in the Monte Carlo simulation are absorbed within tally
-statistics while a hard failure is treated as a high variance event
-where some portion of the Monte Carlo histories lost.
+a possible means of mitigating the impact of these events at the
+application level. With solutions to deterministic transport problems
+based on Monte Carlo techniques, such issues are potentially
+alleviated by statistical arguments. In the case of soft failures,
+isolated floating point errors in the Monte Carlo simulation are
+absorbed within tally statistics while a hard failure is treated as a
+high variance event where some portion of the Monte Carlo histories
+are lost.
 
-Considering these physics and hardware issues, the goal of this work
-is to research and develop Monte Carlo Synthetic Acceleration methods
-for neutron transport problems. These methods are based on using the
-Monte Carlo method to directly accelerate the convergence of
-traditional fixed-point iterative schemes for these problems. We aim
-to address both their benefits and their shortcomings in the context
-of the physics of interest and desire to identify areas in which
-improvements can be made. In previous work \cite{evans_monte_2014}, we
-applied the Monte Carlo synthetic acceleration (MCSA) method to the
-thermal radiation diffusion equation. For several transient problems
-we observed that MCSA performed competetively with preconditioned
-Krylov methods. In a continuation of these studies, we apply the
-method to light water reactor problems discretized by the $SP_N$
-equations. This system has no time dependence, a large scattering
-component, and assymetry in the linear operator, creating a set of
-problems complementary to the transient, opaque, and symmetric systems
-of our previous analyis. In addition, we apply a new fixed point
-iteration in the MCSA sequence as well as a new estimator in order to
-study their effects on convergence.
+The goal of this work is to research and develop Monte Carlo Synthetic
+Acceleration (MCSA) methods for neutron transport problems at
+scale. These methods are based on using the Monte Carlo method to
+directly accelerate the convergence of traditional fixed-point
+iterative schemes for these problems and have the potential to improve
+the resiliency of large-scale solutions to the neutron transport
+problem. We aim to address both the benefits and shortcomings of MCSA
+in the context of the physics of interest and desire to identify areas
+in which improvements can be made to the algorithm so that the
+tradeoff between algorithmic performance and the potential for
+resilience at scale is not too great. In previous work
+\cite{evans_monte_2014}, we applied MCSA to the thermal radiation
+diffusion equation where, for several transient problems, we observed
+that MCSA performed competetively with preconditioned Krylov
+methods. In a continuation of these studies, we apply the method to
+light water reactor problems discretized by the $SP_N$ equations. This
+system has no time dependence, a large scattering component, and
+assymetry in the linear operator, creating a set of problems
+complementary to the transient, opaque, and symmetric systems of our
+previous analysis. For problems of this type, our results indicate
+that the primitive diagonal preconditioning used in our previous work
+is not suitable to achieve convergence with the method. In addition,
+the problems studied in this work are large enough that domain
+decomposition is required in order to obtain enough memory as well as
+time-efficient solutions.
+
+To enable solutions of the $SP_N$ equations for these problems, this
+work develops two new techniques. First, we define an algebraic
+preconditioning strategy for MCSA that leverages an explicitly formed
+preconditioner as well as a matrix-matrix product to construct the
+matrix elements of the preconditioned system. As a preconditioner,
+this work uses a sparse approximate inverse technique which results in
+a sparse matrix with element-wise access, a necessary requirement for
+implementation of the Neumann-Ulam Monte Carlo algorithm. Second, we
+develop a new domain decomposition scheme for MCSA with a new fully
+asynchronous domain decomposed Monte Carlo kernel for the Neumann-Ulam
+algorithm. In tandem, these two developments permit solutions of the
+$SP_N$ equations for large-scale light water criticality problems. We
+believe both of these contributions in preconditioning and
+parallelizing the Neumann-Ulam algorithm are new and unique.
 
@@
 $SP_N$ equations for eigenvalue problems along with the discretization
-used for this work and the model problem. We then describe the
-application of the MCSA algorithm to the solution of the
-equations. Next, we demonstrate that the basic Jacobi preconditioning
+used for this work. We then describe the application of the MCSA
+algorithm to the solution of the equations and introduce the
+criticality problems used for the numerical studies in the this
+work. Next, we demonstrate that the basic Jacobi preconditioning
 applied in our previous work is unsatisfactory to achieve convergence
-for neutron transport in light water reactors and subsquently develop
-a new preconditioning scheme. We then apply this preconditioning
-scheme to a set of criticality problems and compare the
-performance of the preconditioned MCSA method to preconditioned Krylov
-methods. We then parallelize the MCSA algorithm leveraging the MSOD
-scheme to enable very large problems with a demonstration of
-scalability using the test problems. We then finish with conclusions
-and indication of future work.
+for these problems and subsquently present our new preconditioning
+scheme. We then parallelize the MCSA algorithm and describe our new
+domain decomposition scheme deisgned to enable very large problems. We
+then apply this preconditioning scheme and parallel algorithm to the
+criticality problems and compare the performance of the preconditioned
+MCSA method to preconditioned Krylov methods. We then finish with
+conclusions and indication of future work.
 
@@
 \section{Simplified $P_N$ Approximation}
-For criticality problems, the space, angle, and energy discretizations
-for the $SP_N$ equations can be applied to the general eigenvalue form
-of the transport equation given by
-Eq~(\ref{eq:eigenvalue_transport}):
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
-  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_eigenvalue}
-\end{multline}
-with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
-fission matrix, $\mathbf{F}$. We apply the following change of
-variables to yield the set of multigroup pseudo-moments
-$\mathbb{U}_n$:
-\begin{subequations}
-  \begin{gather}
-    \mathbb{U}_1 = \mathbf{\Phi}_0 + 2\mathbf{\Phi}_2 \\
-    \mathbb{U}_2 = 3\mathbf{\Phi}_2 + 4\mathbf{\Phi}_4 \\
-    \mathbb{U}_3 = 5\mathbf{\Phi}_4 + 6\mathbf{\Phi}_6 \\
-    \mathbb{U}_4 = 7\mathbf{\Phi}_6 \:.
-  \end{gather}
-  \label{eq:spn7_subs}
-\end{subequations}
-where now the application of the fission matrix to the moment vectors
-will be expanded into a set of block matrices in identical fashion to
-the scattering matrices:
-\begin{equation}
-  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
-  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
-  \label{eq:spn_fission_matrix}
-\end{equation}
-To complete the spatial discretization, the gradient operators are
-resolved by a finite volume scheme as in \cite{hamilton_2014} such
-that the flux and current are continuous at cell boundaries.. It
-should be noted here that with respect to a general MCSA scheme, the
-introduction of fission in the system will only affect the source
-vector in the linear system for the eigenvalue schemes used in this
-work. The linear operator, and therefore overall MCSA performance will
-be dictated by streaming and scattering as defined on the left-hand
-side. We refer the reader to \cite{hamilton_2014} for an extensive
-presentation of the $SP_N$ discretization used in this work.
+
+Steven or Tom....
 
@@
 \label{sec:mcsa}
-In this section we briefly review the Monte Carlo method for linear
-systems and Monte Carlo Synthetic Acceleration. Given a linear
-problem:
+In this section we briefly review an adjoint variant the Monte Carlo
+method of Neumann and Ulam for linear systems and Monte Carlo
+Synthetic Acceleration.
+
+\subsection{Neumann-Ulam Monte Carlo Algorithm}
+An alternative approach to approximate matrix inversion is to employ
+Monte Carlo methods that sample a distribution with an expectation
+value equivalent to that of the inverted operator. Such methods have
+been in existence for decades with the earliest reference noted here a
+manuscript published in 1950 by Forsythe and Leibler
+\cite{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
+in fact credit the creation of this technique to J. Von Neumann and
+S.M. Ulam some years earlier than its publication. In 1952 Wasow
+provided a more formal explanation of Von Neumann and Ulam's method
+\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
+\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
+\cite{spanier_monte_1969} present additional detail on this topic
+using a collection of references from the 1950's and early
+1960's which we use here in our discussion.
+
+To derive the algoritm, consider a linear problem:
 \begin{equation}
@@
 \end{equation}
-where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
-that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
-\mathbb{R}^N$ is the solution vector, and $\ve{b} \in
-\mathbb{R}^N$. To form the Monte Carlo method we define the linear
-system adjoint to Eq~(\ref{eq:linear_problem}):
+where $\ve{A} \in \mathbb{R}^{N \times N}$, $\ve{x} \in \mathbb{R}^N$,
+and $\ve{b} \in \mathbb{R}^N$. We define the linear system adjoint to
+Eq~(\ref{eq:linear_problem}):
 \begin{equation}
@@
 vectors respectively and $\ve{A}^T$ is the adjoint operator for
-$\ve{A} \in \mathbb{R}^{N \times N}$ which we hten split to get:
+$\ve{A} \in \mathbb{R}^{N \times N}$ which we then split to get:
 \begin{equation}
@@
 \end{equation}
-where $\ve{H}^T$ is the adjoint iteration matrix. The spectral radius
-of $\ve{H}^T$ must remain less than 1 for convergence. By defining the
-following adjoint inner product equivalence \cite{spanier_monte_1969}:
+where $\ve{H}^T = \ve{I} - \ve{A}^T$ is the adjoint \textit{iteration
+  matrix}.  We then solve for $\ve{y}$ by writing
+Eq~(\ref{eq:adjoint_split_system}) as:
+\begin{equation}
+  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
+  \label{eq:adjoint_split_system_2}
+\end{equation}
+which in turn yields the Neumann series using the adjoint operators:
+\begin{equation}
+  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
+  \label{eq:adjoint_neumann_series}
+\end{equation}
+
+The Monte Carlo algorithm is generated by expanding
+Eq~(\ref{eq:adjoint_neumann_series}) into individual terms:
+\begin{equation}
+  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}h^T_{i,i_1} h^T_{i_1,i_2}\ldots h^T_{i_{k-1},i_k}
+  d_{i_k}\:,
+  \label{eq:adjoint_neumann_solution}
+\end{equation}
+which can interpreted as a series of transitions between states of the 
+adjoint iteration matrix,
+\begin{equation}
+ \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
+ \rightarrow i_{k}\:,
+  \label{eq:mc_walk_permutation}
+\end{equation}
+where $\nu$ is interpreted as a particular sequence instance. If we
+consider $\nu$ to be a sequence created by a Markov chain, we can
+build probabilities, weights, and a stochastic estimator to play a
+Monte Carlo game whos expectation value is the solution to
+Eq~(\ref{eq:linear_problem}). To do this, in
+Eqs~(\ref{eq:linear_problem}) and (\ref{eq:adjoint_linear_problem}) we
+have 3 unknowns: $\ve{x}$, $\ve{y}$ and $\ve{d}$. To eliminate
+$\ve{y}$ and $\ve{d}$ and give a system to solve for $\ve{x}$ we
+require two constraints. By defining the following adjoint inner
+product equivalence \cite{spanier_monte_1969}:
 \begin{equation}
@@
 \end{equation}
-it follows that:
+the first constraint follows as:
 \begin{equation}
@@
 \end{equation}
-We can acquire the adjoint solution by forming the Neumann series by
-writing Eq~(\ref{eq:adjoint_split_system}) as:
-\begin{equation}
-  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
-  \label{eq:adjoint_split_system_2}
-\end{equation}
-which in turn yields the Neumann series using the adjoint operator:
-\begin{equation}
-  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
-  \label{eq:adjoint_neumann_series}
-\end{equation}
-We expand this summation to again yield a series of transitions that
-can be approximated by a Monte Carlo random walk sequence by forming
-the Neumann series in reverse order:
-\begin{equation}
-  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
-  \label{eq:adjoint_neumann_solution}
-\end{equation}
-We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the solution to
-Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
-$\ve{d}$, and therefore we require two constraints to close the
-system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
-constraint and as a second constraint we select:
+As a second constraint we select:
 \begin{equation}
@@
 \end{equation}
-meaning that if we compute the inner product of the original source
-and the adjoint solution using a delta function source, we recover one
-component of the original solution.
+meaning that if we compute the inner product of the original
+right-hand-side of the linear system and the adjoint solution using a
+delta function source, we recover one component of the original
+solution.
 
-In terms of radiation transport, this adjoint method is equivalent to
-a traditional forward method where the initial state $i_0$ of the
-random walk is determined by sampling the source vector $\ve{b}$ with
-probabilities:
+We can build probabilities and weights by thinking in terms of
+radiation transport. This adjoint variation of the Neumann-Ulam
+algorithm is equivalent to a traditional forward method where the
+initial state, $i_0$, of the random walk is determined by sampling
+$\ve{b}$ with probabilities:
 \begin{equation}
@@
 We build weights and probabilities for the random walk using the
-\textit{adjoint Neumann-Ulam decomposition} of $\ve{H}$:
+adjoint \textit{Neumann-Ulam decomposition} of $\ve{H}$:
 \begin{equation}
@@
 \end{equation}
-where we are forming the decomposition with respect to the transpose
-of $\ve{H}$. $\ve{P}$ is defined element-wise as:
+where $P$ is a row-stochastic matrix with each row being a probability
+distribution function and $W$ a corresponding weight matrix.  The
+probability matrix is defined element-wise as:
 \begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
+  p_{ij} = \frac{|h^T_{ij}|}{\sum_j |h^T_{ij}|}\:,
   \label{eq:adjoint_probability}
@@
 such that we expect to select a new state, $j$, from the current state
-in the random walk, $i$, by sampling column-wise (or row-wise if an
-adjoint probability matrix is formed). Per
-Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
-defined as:
+in the random walk, $i$, by sampling from the distribution function in
+row $i$ of the proability matrix. Per
+Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight from state
+$i$ to state $j$ is then defined as:
 \begin{equation}
-  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
+  w_{ij} = \frac{h^T_{ij}}{p_{ij}}\:.
   \label{eq:adjoint_weight}
 \end{equation}
-Using the decomposition we can then define an expectation value for
-the adjoint method. The contribution to the solution in state $j$ from
-a particular random walk permutation of $k$ events is given by
-\textit{collision estimator}:
+
+Next we define an expectation value for the random walks. The
+contribution to the solution in state $j$ from a particular random
+walk permutation of length $k$ is given by the \textit{collision
+  estimator}:
 \begin{equation}
@@
 \end{equation}
-where the Kronecker delta indicates that the tally contributes only in
-the current state, $i_m$, of the random walk. The expectation value
-using all permutations is then:
+where the Kronecker delta, defined as the right-hand-side of the
+adjoint linear problem in Eq~(\ref{eq:adjoint_second_constraint}),
+indicates that the tally contributes only in the current state, $i_m$,
+of the random walk. The expectation value using all random walk
+permutations is then:
 \begin{equation}
@@
 \end{equation}
-which, if expanded utilizing Eq~(\ref{eq:adjoint_source_definition})
-to insert the source term, directly recovers the exact solution:
+with $P_{\nu}$ the probability of a given random walk, $\nu$.  If
+Eq~(\ref{eq:adjoint_expectation_value}) is expanded utilizing
+Eq~(\ref{eq:adjoint_source_definition}) to insert the source term, we
+directly recover a component of the solution to the linear problem:
 \begin{equation}
@@
 
-In addition to the collision estimator, an additional estimator is
-available due to the work of Okten \cite{okten_solving_2005} that uses
-the method of expected values as a means to improve the Monte Carlo
-estimate. As outlined by Spanier and Gelbard
-\cite{spanier_monte_1969}, the method of expected values is a
-deterministic averaging of events that may potentially occur in the
-Monte Carlo random walk sequence. Okten applied this principle
-directly to discrete Monte Carlo by forming the \textit{expected value
-  estimator} for a random walk of $k$ events:
+Convergence of the Monte Carlo method, previously assumed by many to
+be entirely bound by the spectral radius of $\ve{H}^T$
+(e.g. \cite{spanier_monte_1969}), was recently identified as a only a
+subset of the convergence criteria. In particular, Ji, Mascagni, and
+Li demonstrate in \cite{ji_2013} that a necessary and sufficient
+condition for convergence of this algorithm is $\rho(\ve{H^*}) < 1$
+with $\ve{H^*}$ defined element-wise as:
 \begin{equation}
-  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
-  \label{eq:expected_value_estimator}
+  h^*_{ij} = \frac{(h^T_{ij})^2}{p_{ij}}\:,
+  \label{eq:h_star}
 \end{equation}
-where now the contribution of the iteration matrix is
-deterministically averaged at step $m$ over all potential states $j$
-that may be reached from the current state $i_m$. Via Okten, the
-estimator can be shown to be unbiased through a comparison to the
-collision estimator. We can first rewrite the summation in
-Eq~(\ref{eq:expected_value_estimator}):
+which is indicative that both the structure of $\ve{H}^T$ and the
+selection of transition probabilities affect the convergence of the
+algorithm.
+
+\subsection{Monte Carlo Synthetic Acceleration}
+Halton presented a residual Monte Carlo scheme that dramatically
+improved performance by constructing an iterative refinement scheme
+\cite{halton_sequential_1962}. Using the ideas of Halton, Evans and
+Mosher recently developed a Monte Carlo solution method that was not
+prohibited severely by the quality of the initial guess for the system
+\cite{evans_monte_2009} and later applied it more rigorously as a
+solution mechanism for the transient radiation diffusion equation
+\cite{evans_monte_2014}. Their approach was instead to use residual
+Monte Carlo as a synthetic acceleration for a stationary method. To
+derive this method, we begin by splitting the operator in
+Eq~(\ref{eq:linear_problem})
 \begin{equation}
-  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
-  \delta_{i_m,i} h_{ji}\,
-  \label{eq:unbiased_eval_1}
+  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
+  \label{eq:linear_split}
 \end{equation}
-where $N$ is the number of states in the system. Immediately, we see
-the collision estimator as defined by
-Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
-the expectation value as:
+With this we can then define the stationary method
+\textit{Richardson's iteration} as:
 \begin{equation}
-  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
-  \label{eq:unbiased_eval_2}
+  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
+  \label{eq:richardsons_iteration}
 \end{equation}
-which is equivalently is the $j^{th}$ component of
-Eq~(\ref{eq:richardson_split}):
+which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
+solution error at the $k^{th}$ iterate relative to the true solution:
 \begin{equation}
-  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
-  \label{eq:unbiased_eval_2}
+  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
+  \label{eq:mcsa_error}
 \end{equation}
-and is therefore an unbiased estimate. Compared to the collision
-estimator, the expected value estimator provides additional
-information at every step of the random walk, yielding potentially
-better statistics with the same amount of transport
-work. Conveniently, even if no Monte Carlo histories are computed, the
-expected value estimator still deterministically computes the first
-term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
-estimator will provide no information.
-
-\subsection{MCSA Solution of the $SP_N$ Equations}
-\label{subsec:mcsa_solution}
-To generate the multiplication factor and steady-state flux
-distribution for this problem, at every eigenvalue iteration MCSA is
-used to solve the resulting $SP_N$ problem using the provided fission
-source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
-within a power iteration strategy to find the multiplication factor.
-\begin{algorithm}[h!]
-  \caption{Power Iteration MCSA Scheme}
-  \label{alg:power_iteration}
-  \begin{algorithmic}
-    \State $k_0 =$ initial guess
-    \State $\mathbf{\Phi}_0 =$ initial guess
-    \State $n = 0$
-    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
-    \Comment{Iterate until convergence of the eigenvalue}
-    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
-    \Comment{Solve for the new flux state with MCSA}
-    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
-      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
-    \Comment{Update the multiplication factor}
-    \State $n = n+1$
-    \EndWhile
-  \end{algorithmic}
-\end{algorithm}
-Here, $\mathbf{M}$ is the transport operator generated on the
-left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
-fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
-problem is significantly more complicated than the simple test problem
-used for the previous spectral analysis. Fission has been introduced
-into the set of equations and the addition of moderator into the
-system will increase the amount of scattering, creating a
-significantly more difficult problem manifesting itself in an
-iteration matrix with a larger spectral radius. When using MCSA, the
-linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
-iteration will dictate convergence and remain unchanged throughout the
-computation while the addition of fission to the system will only
-modify the source of neutrons and the multiplication factor while not
-affecting Monte Carlo transport.
-
-At each eigenvalue iteration, we then solve the linear system with the
-MCSA:
+Subtracting Eq~(\ref{eq:richardsons_iteration}) from
+Eq~(\ref{eq:linear_split}) we get:
+\begin{equation}
+  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
+  \label{eq:mcsa_setup_1}
+\end{equation}
+Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
+\begin{equation}
+  \begin{split}
+    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
+    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
+    \label{eq:mcsa_setup_2}
+  \end{split}
+\end{equation}
+Using this, we define the following scheme that will converge in one
+iteration if $\ve{A}$ is inverted exactly:
 \begin{subequations}
   \begin{gather}
-    \ve{r}_{l} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
-    \ve{M}\mathbf{\Phi}^{n+1}_{l}\:,\\
-    \mathbf{\Phi}^{n+1}_{l+1/2} = \mathbf{\Phi}^{n+1}_l + \ve{r}_l\:,\\
-    \ve{r}_{l+1/2} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
-    \ve{M}\mathbf{\Phi}^{n+1}_{l+1/2}\:,\\
-    \ve{M}\delta\mathbf{\Phi}^{n+1}_{l+1/2} = \ve{r}_{l+1/2}\:,\\
-    \mathbf{\Phi}^{n+1}_{l+1} = \mathbf{\Phi}^{n+1}_{l+1/2} + \delta
-    \mathbf{\Phi}_{l+1/2}\:,
+    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
+    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
+    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
+  \end{gather}
+  \label{eq:mcsa_setup_3}
+\end{subequations}
+However, $\ve{A}$ is only approximately inverted by our numerical
+methods and therefore we instead pose an iterative scheme in which the
+Monte Carlo solvers are used to invert the operator. The fixed-point
+\textit{Monte Carlo Synthetic Acceleration} (MCSA) method is defined
+as:
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
+    \label{eq:mc_mcsa_solve}
+    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
   \end{gather}
@@
 \end{subequations}
-where the $l$ subscript indicates the MCSA iteration number and $n$
-again indicates the eigenvalue iteration. We refer the reader to
-\cite{evans_monte_2014} for more a more complete definition of the
-Monte Carlo method for linear systems and MCSA.
+where a Neumann-Ulam Monte Carlo method is used to approximately solve
+Eq~(\ref{eq:mc_mcsa_solve}).
+
+Using Monte Carlo in this way achieves the same effect as Halton's
+method, decoupling the convergence rate of the Monte Carlo from the
+overall convergence rate of the method. Here, the approximate Monte
+Carlo solution is not driven to a particular convergence as it merely
+supplies a correction for the initial guess generated by Richardson's
+iteration. Rather, only a set number of histories are required using
+the Neumann-Ulam method to generate the correction. In addition, the
+fact that the scheme in Eq~(\ref{eq:mcsa_setup_3}) will converge in
+one iteration if $\ve{A}$ is inverted exactly means that as the number
+of random walks in a given iteration are increased, stochastic error
+in $\delta \ve{x}$ is reduced towards zero and the number of
+iterations required for MCSA to converge should decrease accordingly,
+thus accelerating the solution.
+
+\subsection{Arnoldi Iteration with MCSA}
+
+Steven.....
+
+%%---------------------------------------------------------------------------%%
+\section{Criticality Model Problems}
+
+Steven or Tom....
+
+\subsection{Light Water Reactor Assembly}
+
+Steven or Tom....
+
+\subsection{C5G7}
+
+Steven or Tom....
+
+%%---------------------------------------------------------------------------%%
+\section{Preconditioning Algorithm}
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Parallel MCSA}
+\section{Parallel Algorithm}
 \label{sec:parallel_mcsa}
@@
 
-%%---------------------------------------------------------------------------%%
-\subsection{Domain Decomposed Monte Carlo}
-\label{subsec:msod}
-Large-scale problems will for reasons typically related to memory
-restrictions or performance have their data partitioned such that each
-parallel process owns a subset of the equations in the linear
-system. Given this convention, the Neumann-Ulam Monte Carlo algorithm
-must perform random walks over a domain that is decomposed and must
-remain decomposed to avoid the same performance and memory
-restrictions. To parallelize this algorithm, we then seek parallel
-Monte Carlo algorithms that handle domain decomposition.
-
-To motivate this problem, consider the square domain presented in
-Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
-Carlo particle transport problem. If the domain were decomposed into 9
-subdomains as shown, each of those subdomains and their associated
-data (i.e. cross subsections) could be owned by a different parallel
-process in the computation. In Figure~\ref{fig:ddmc_example}, the
-tracks of 3 particles born in the center subdomain are shown. In this
-example, particle A is first transported from the center domain to the
-domain directly to the left. Before the scattering event in the new
-domain may be processed, particle A must be communicated between the
-two parallel processes that own those domains. For any given particle
-in the transport simulation, this communication event may hardly occur
-during the lifetime of the particle (particle B), or may occur many
-times (particle C) depending on the problem parameters and the random
-path in phase space taken by the particle. Scalable parallel
-algorithms for domain decomposed Monte Carlo are those that handle
-this domain-to-domain communication of particles effectively and
-balance it with the on-process computations for particle interactions,
-tallies, response calculations, and other simulation requirements.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Domain decomposed Monte Carlo transport example
-      illustrating how domain decomposition requires parallel
-      communication of particles.} \textit{If a particle crosses a
-      domain boundary it must be communicated to the next domain in
-      its path.}}
-  \label{fig:ddmc_example}
-\end{figure}
-
-For a domain decomposed Neumann-Ulam method, the situation is very
-much the same where instead the transport process is now discrete and
-the "physics" of the transport process is the Monte Carlo game with
-probabilities and weights described by
-Eq~(\ref{eq:neumann_ulam_decomposition}) for the given linear
-problem. Figure~\ref{fig:ddnu_example} gives the domain decomposed
-Neumann-Ulam analog of particle transport problem in
-Figure~\ref{fig:ddmc_example}. Imagine in this problem that we are
-solving for the monoenergetic scalar neutron flux as in the model
-diffusion problem presented in Appendix~\ref{chap:diffusion_problem}.
-In this problem, the solution has only a spatial dependence and
-therefore each point in the mesh corresponds to an equation in the
-resulting linear system. In Figure~\ref{fig:ddnu_example}, the domain
-has now been discretized by this mesh and again decomposed into 9
-subdomains, each owning a piece of the mesh and therefore the
-corresponding equations in the linear system. As we play the Monte
-Carlo game presented in Chapter~\ref{ch:stochastic_methods}, the
-discretization and resulting Neumann-Ulam decomposition of the problem
-describes how each state in the system is coupled to the other states
-in the system. This coupling is then responsible for the discrete
-random walk that each history takes as shown in this example. As
-discrete states (or rows in the linear system) that do not exist in
-the local domain are reached during the random walk by stochastic
-histories, the histories must be communicated to the subdomain that
-does own the discrete state in an analogous fashion to the particles
-in the previous example.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
-      how domain decomposition requires parallel communication of
-      histories.} \textit{Each mesh point corresponds to an equation
-      in the linear system and the coupling among equations is
-      described by the discretization of the problem.}}
-  \label{fig:ddnu_example}
-\end{figure}
-
-To further motivate using a parallel algorithm similar to particle
-transport methods, again consider history A in
-Figure~\ref{fig:ddnu_example}. This history begins in the center
-subdomain and through the course of transitioning through states in
-the system arrives at a state that exists in the subdomain directly to
-the left of the center subdomain. History A must then be communicated
-from the center subdomain to the subdomain immediately to the left to
-continue the random walk. As with the particle transport system, we
-may find that histories in a domain decomposed Neumann-Ulam method are
-only communicated a few times (history B) if at all, or they may be
-communicated many times (history C) depending on the discretization
-and the outcome of the Monte Carlo game. In identical fashion to the
-particle transport problem, a domain decomposed Neumann-Ulam method
-has the same communication requirements with all histories in both
-examples requiring the same parallel operations. The amount of
-communication of histories from domain to domain will be the primary
-factor in the parallel scalability of the algorithm. In
-Appendix~\ref{chap:parallel_theory}, the number of histories
-communicated in a given problem are quantified analytically for a
-simple model problem.
-
-%%---------------------------------------------------------------------------%%
 \subsection{Domain Decomposed Neumann-Ulam Algorithm}
@@
 
-In this subsection Brunner and Brantley's fully asynchronous algorithm,
-which was effectively implemented verbatim for this work, is presented
-along with its application to the Neumann-Ulam method. In their work
-they identify two data sets that are required to be communicated: the
-sharing of particles that are transported from one domain to another
-and therefore from one processor to another and a global communication
-that signals if particle transport has been completed on all
-processors. Both of these communication sequences will be addressed at
-a high level along with how the Monte Carlo data structures they
-require are constructed in parallel.
+%%---------------------------------------------------------------------------%%
+\section{Results}
 
-\subsubsection{Parallel Transport Domain and Source}
-\label{subsubsec:domain_generation}
-To utilize a parallel transport algorithm, we must generate the
-required data with the correct parallel decomposition. For the
-Neumann-Ulam method, the transport domain consists of all states in
-the system that are local, and the probabilities and weights for all
-state transitions possible in the local domain. At the core of this
-representation is the Neumann-Ulam decomposition of the linear
-operator as given by Eq~\ref{eq:neumann_ulam_decomposition}. Given an
-input parallel decomposition for the linear operator, by definition
-the Neumann-Ulam decomposition will have the same parallel
-decomposition. In addition, the source vector will have the same
-parallel decomposition as the input matrix and therefore all of the
-birth states for the histories will exist for at least a single
-transition event within the local domain.
-
-Compared to the serial construction of the Neumann-Ulam decomposition,
-data for all states that are required to be on process must be
-collected in parallel so that all possible transitions that leave the
-local domain may be correctly communicated. As an example of this
-state collection, Figure~\ref{fig:diffusion_graph} gives the adjacency
-graph of a single state in the neutron diffusion matrix for the model
-problem presented in Appendix~\ref{chap:diffusion_problem}. The
-discretization stencil of the diffusion problem dictates the structure
-of this graph and the states to which a history may transition.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
-    \textit{The structure of the graph comes from the discretization
-      of the Laplacian operator describing the diffusion
-      physics. Adjacent boundary states are collected by traversing
-      the graph one step for all local boundary states and collecting
-      the equations for those states that are not local. For a given
-      transition in the random walk for the diffusion problem, a
-      history at mesh point $(i,j)$ may transition to all adjacent
-      mesh points including itself, corresponding to a global state
-      transition from $m$ to $n$ in the linear system.}}
-  \label{fig:diffusion_graph}
-\end{figure}
-
-In this example, we currently reside in state $m$ of the system,
-directly correlating to physical location $(i,j)$ in the mesh (the
-center node). If we play the Monte Carlo game to move to a new state
-$n$, then we may move to any of the nodes in this graph, including the
-node at which we started. If grid point $(i,j)$ in this example is on
-the right side of the local domain boundary and we transition to the
-state $n$ corresponding to mesh point $(i+1,j)$, we are now in a state
-that is owned by the adjacent domain. We must then gather the
-identification number for the neighboring process that owns grid point
-$(i+1,j)$ such that we have an address to send all histories that
-leave the local domain boundary with this state as their
-destination. For the parallel source, these adjacent states are not
-required as histories may only be born in states in the local
-domain. Once the local Neumann-Ulam decomposition has been generated
-with the proper communication data collected from adjacent domains,
-Monte Carlo transport may proceed in parallel.
-
-\subsubsection{Domain Decomposed Algorithm}
-\label{subsubsec:parallel_mc_algorithm}
-
-We next present Brunner and Brantley's 2009 algorithm in detail and
-discuss how it is adapted to parallelize the Neumann Ulam
-method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
-top level sequence performs the task of managing history transport
-through the local domain, communication of histories to adjacent
-domains, and the completion of transport. For each of these specific
-tasks, additional algorithms, shown in bold in
-Algorithm~\ref{alg:parallel_mc_algorithm}
-(e.g. \textbf{LocalHistoryTransport()}), are presented for additional
-detail in the same manner as Brunner and Brantley.
-
-\begin{algorithm}[h!]
-  \caption{Parallel Neumann-Ulam Algorithm}
-  \label{alg:parallel_mc_algorithm}
-  \begin{algorithmic}[1]
-    \State get list of neighbor processors 
-    \Comment{Each neighbor owns an adjacent subdomain}
-    \ForAll{neighbors}
-    \State post non-blocking receive for maximum history buffer size
-    \State allocate history buffer
-    \EndFor
-    \State \textit{historiesCompleted} = 0
-    \Comment{local+child finished histories}
-    \State \textit{localProcessed} = 0
-    \Comment{local trajectories computed (not necessarily finished)}
-    \State calculate parent and children processor ID numbers in
-    binary tree
-    \ForAll{child processes}
-    \State post non-blocking receive for \textit{historiesCompleted} tally
-    \EndFor
-    \State post non-blocking receive for stop message from parent
-    \While{stop flag not set}
-    \If{any local histories in source or stack}
-    \State \textbf{LocalHistoryTransport()}
-    \State ++\textit{localProcessed}
-    \EndIf
-    \If{message check period == \textit{localProcessed} || no local histories}
-    \State \textbf{ProcessMessages()}
-    \State \textit{localProcessed} = 0
-    \EndIf
-    \If{no local histories}
-    \State \textbf{ControlTermination()}
-    \EndIf
-    \EndWhile
-    \State cancel outstanding non-blocking requests
-    \State free all history buffers
-  \end{algorithmic}
-\end{algorithm}
-
-Successful execution of this algorithm requires construction of the
-parallel Neumann-Ulam decomposition as described in the preceding
-subsection. This data is used to begin
-Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
-collected list of neighboring processors generated while building the
-Neumann-Ulam decomposition to setup the set of asynchronous messages
-required for domain-to-domain communication. Again, consider this
-communication pattern for the 9 subdomain example given by
-Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each boundary
-domain has two neighbors and the center domain four neighbors with
-which they will communicate parallel histories and this communication
-goes both ways as represented by the adjacent arrows in the
-figure. For each set of neighbors, a non-blocking send and receive is
-required with data buffers allocated with a user-defined size for
-incoming and outgoing histories. This non-blocking structure is
-critical to the performance of the algorithm in that it permits local
-history transport to continue while new histories to transport are
-being collected in the buffers. When a given process is ready to do
-more work, it can check these data buffers for incoming histories
-requiring further transport. In this way there is a maximum amount of
-overlap between communication and transport of histories.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Nearest neighbor history communication sequence.}
-    \textit{Each subdomain in the system has a set of nearest
-      neighbors determined by the parallel adjacency graph of the
-      input matrix. The subdomains are indexed by an integer.}}
-  \label{fig:nearest_neighbor_comm}
-\end{figure}
-
-Once the nearest-neighbor communication sequence has been prepared,
-the completion of transport sequence is readied in lines 6 and 8-12 in
-Algorithm~\ref{alg:parallel_mc_algorithm}. In these lines we are
-setting up an asynchronous binary communication tree as presented for
-the same 9 subdomain example in Figure~\ref{fig:binary_comm_tree}. In
-this communication pattern, each process has one parent process
-(except for the MASTER process 0) to which it will non-blocking send
-the number of histories that terminated their transport procedure by
-weight cutoff in the local domain. Equivalently, each process has up
-to two child processes from which it will receive their completed
-number of histories with a non-blocking receive operation. Setting up a
-tree in this manner lets the completed history tally
-(\textit{historiesCompleted} in the algorithm) be updated
-incrementally and funneled to the root process. Because we are solving
-fixed source problems with the Neumann-Ulam algorithm without any type
-of variance reduction that may generate more histories that we started
-with, once the root process completion sum tallies to the number of
-histories in the source, transport is complete. Once this occurs, the
-root process non-blocking sends the stop message to its children and
-each process non-blocking receives a stop message from its parent as
-shown in Figure~\ref{fig:binary_comm_tree}. The stop message is then
-propagated up the tree in the same manner.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.0}{
-      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Binary communication tree for coordinating the end
-      of a parallel Neumann-Ulam solve.} \textit{Each child process
-      reports to its parents how many histories completed within its
-      domain. When the root process sums all complete histories, it
-      forwards the stop signal to its children which in turn forward
-      the message up the tree. The subdomains are indexed by an
-      integer.}}
-  \label{fig:binary_comm_tree}
-\end{figure}
-
-With these communication structures prepared, we can now enter the
-main transport loop at line 13 of
-Algorithm~\ref{alg:parallel_mc_algorithm}. In this loop, a set of
-mutually exclusive tasks enabled by the fully asynchronous
-communication patterns are executed until the stop signal is received
-from the parent process in the binary tree. In this case, mutually
-exclusivity permits all local operations to occur independently of
-other processes in the system and their current state. For each
-process, there are two data structures from which histories may be
-obtained for the transport procedure. The first is the local source
-and the second is a LIFO (last in first out) stack of histories
-transported to the local domain from the adjacent domains. In lines
-14-17, if histories exist in either of these data structures, then
-they are transported through the local domain using
-Algorithm~\ref{alg:local_history_transport}. If the history is
-terminated by weight cutoff the \textit{historiesCompleted} tally is
-updated. If it hits the boundary of the domain it is added to the
-outgoing history buffer for the neighboring domain and if the buffer
-is full, it is sent to the neighboring domain with a non-blocking
-operation and the memory associated with that buffer reallocated. In
-all instances that Algorithm~\ref{alg:local_history_transport} is
-executed, \textit{localProcessed} is incremented to account for
-histories that have been processed locally.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{LocalHistoryTransport()}}
-  \label{alg:local_history_transport}
-  \begin{algorithmic}[1]
-    \State transport history through the domain until termination
-    \If{history is in neighbor boundary state}
-    \State add history to neighbor buffer
-    \If{neighbor buffer is full}
-    \State non-blocking send history buffer to neighbor
-    \State allocate new history buffer for neighbor
-    \EndIf
-    \Else
-    \If{history terminated by weight cutoff}
-    \State post-process history
-    \State ++\textit{historiesCompleted}
-    \EndIf
-    \EndIf
-  \end{algorithmic}
-\end{algorithm}
-
-Continuing in the transport loop, if there are no local histories to
-transport or the \textit{localProcessed} count has reached some
-user-defined check frequency, lines 18-21 in
-Algorithm~\ref{alg:local_history_transport} check for more histories
-to transport in the incoming data buffers by calling
-Algorithm~\ref{alg:process_messages}. For each neighboring domain in
-the problem, if there are histories in those data buffers then they
-are added to the stack for processing and the non-blocking receive
-operation re-instantiated. In addition, the terminated histories count
-is updated with those values received from the child processes.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{ProcessMessages()}}
-  \label{alg:process_messages}
-  \begin{algorithmic}[1]
-    \ForAll{received history buffers from neighbor}
-    \State unpack number of histories in buffer
-    \State add the histories to the running stack
-    \State re-post non-blocking receive with neighbor
-    \EndFor
-    \ForAll{\textit{historiesCompleted} messages from children}
-    \State \textit{historiesCompleted} += message value
-    \State re-post non-blocking receive with child
-    \EndFor
-  \end{algorithmic}
-\end{algorithm}
-
-Finally, the transport loop finishes in lines 22-24 of
-Algorithm~\ref{alg:local_history_transport} by calling
-Algorithm~\ref{alg:control_termination} if there are no local
-histories to transport in the stack or the source. In
-Algorithm~\ref{alg:control_termination}, we continue to forward the
-terminated history tally down to the parent in the binary tree and
-send all history buffers to our neighbors, even if they are not
-full. If Algorithm~\ref{alg:control_termination} is called by the
-MASTER process, it checks for completion of the problem and if
-complete, forwards the stop flag onto its children as in
-Figure~\ref{fig:binary_comm_tree}. If the process is not the MASTER,
-then we check for a stop signal from the parent and forward it to the
-child processes.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{ControlTermination()}}
-  \label{alg:control_termination}
-  \begin{algorithmic}[1]
-    \State non-blocking send all partially full history buffers
-    \ForAll{\textit{historiesCompleted} messages from children} \State
-    \textit{historiesCompleted} += message \textit{historiesCompleted}
-    \State re-post non-blocking receive with child \EndFor \If{MASTER
-      processor} \If{\textit{historiesCompleted} == global \# of
-      source histories} \State set stop flag \ForAll{children} \State
-    non-blocking send stop message to child \EndFor \EndIf \Else \State
-    non-blocking send \textit{historiesCompleted} to parent \State
-    \textit{historiesCompleted} = 0 \State check for stop signal from
-    parent \If{stop signal from parent} \ForAll{children} \State
-    non-blocking send stop message to child \EndFor \EndIf \EndIf
-  \end{algorithmic}
-\end{algorithm}
-
-The transport loop continues in this manner until all processes have
-received the stop signal from their parents. Older versions of this
-algorithm, particularly some of those presented in
-\cite{brunner_comparison_2006}, use a master/slave approach as given
-by Figure~\ref{fig:master_comm_tree} for the completion of a transport
-stage instead of the binary tree scheme. In this approach, the root
-process still manages the final summation of the completion tally, but
-it directly receives completed tally results from all other processes
-in the problem instead of from just its children. Although this
-implementation is potentially simpler to understand than the binary
-tree approach, the authors of \cite{brunner_comparison_2006} observed
-that this type of termination pattern, even when implemented in a
-fully asynchronous manner, was a significant scalability
-bottleneck. This bottleneck is due to the fact that the master process
-falls behind the others, creating a load imbalance even with the
-addition of a few integer addition operations. We will explicitly
-demonstrate in scaling studies how this bottleneck also occurs within
-a Neumann-Ulam implementation of this algorithm thus requiring the
-implementation of a binary communication tree.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.75}{
-      \input{chapters/parallel_mc/master_comm_tree.pdftex_t} }
-  \end{center}
-    \caption{\textbf{Master/slave scheme for coordinating the end of a
-        parallel Neumann-Ulam solve.} \textit{Each slave process
-        reports to the master how many histories completed within its
-        domain. When the master process sums all complete histories,
-        it sends the stop signal to all slave processes. The
-        subdomains are indexed by an integer.}}
-  \label{fig:master_comm_tree}
-\end{figure}
-
-There is an additional advantage to Brunner and Brantley's work that
-although not immediately applicable to this work has the potential to
-provide value in the future. In addition to a robust fully
-asynchronous communication pattern, this algorithm may also be
-modified to account for situations where the total number of histories
-in a given stage are not known before starting transport. From a
-physics perspective, we might expect this for situations where perhaps
-an (n,2n) interaction occurs in a neutronics problem. In this case,
-the algorithm is modified to account for both histories terminated and
-created and several mechanisms are introduced to determine completion
-of the transport stage. For future variations of this work, certain
-variance reduction techniques that create histories, such as
-splitting, have the potential to be successfully employed as a means
-of accelerating the time to solution for a given problem using
-MCSA. The parallel Neumann-Ulam algorithm presented here may be
-adapted to account for these additional techniques.
-
-%%---------------------------------------------------------------------------%%
-\subsection{Multiple-Set Overlapping-Domain Algorithm}
-\label{subsubsec:msod}
-Although the implementation presented in the previous subsection was
-observed by Brunner and Brantley to be robust and allowed for scaling
-to large numbers of processors, performance issues were still noted
-with parallel efficiency improvements needed in both the weak and
-strong scaling cases for unbalanced problems. These results led them
-to conclude that a combination of domain decomposition and domain
-replication could be used to solve some of these issues. In 2010,
-Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\cite{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times potentially achieves efficient simulation
-on $O(100,000)$ processors, thus providing both spatial and particle
-parallelism.
-
-Each collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm.
-
-To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \cite{mervin_variance_2012}
-and presented in Figure~\ref{fig:msod_example}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/msod_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
-  \label{fig:msod_example}
-\end{figure}
-In this example, 3 particle histories are presented emanating from the
-blue region of interest. Starting with particle A, if no domain
-overlap is used then the only the blue domain exists on the starting
-processor. Particle A is then transported through 3 other domains
-before the history ends, therefore requiring three communications to
-occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
-shown by the dashed line, then the starting process owns enough of the
-domain such that no communications must occur in order to complete the
-particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as that represented by the path of particle
-C. In this case, particle C is scattering between two adjacent
-domains, incurring a large latency cost for a single
-particle. Finally, with particle B we observe that 0.5 domain overlap
-will still not eliminate all communications. However, if 1 domain
-overlap were used, the entire geometry shown in
-Figure~\ref{fig:msod_example} would be contained on the source
-processor and therefore transport of all 3 particles without
-communication would occur.
-
-Wagner and colleagues used this methodology for a 2-dimensional
-calculation of a pressurized water reactor core and varied the domain
-overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain contained an entire fuel assembly. For
-the fully domain decomposed case, they observed that 76.12\% of all
-source particles leave the domain. At 1.5 domain overlap, the
-percentage of source particles born in the center assembly leaving the
-processor domain dropped to 1.05\% and even further for 0.02\% for the
-3 domain overlap. Based on their results, we hypothesize that the
-overlap approach coupled with the multiple sets paradigm that will
-enhance the scalability of the pure domain-decomposition algorithm for
-Neumann-Ulam presented in the previous subsection.
-
-\subsubsection{Generating the MSOD Decomposition}
-\label{subsubsec:msod_generation}
-In \S~\ref{subsubsec:domain_generation} we discussed how to generate the
-parallel transport domain from the Neumann-Ulam decomposition of a
-domain decomposed linear operator. We can readily adapt those data
-structures to account for the extra information required by an
-implementation of the MSOD algorithm. First, we consider the
-generation of overlap. Conveniently, this is identical to the
-neighboring states discovery problem discussed in conjunction with
-Figure~\ref{fig:diffusion_graph}. To generate the boundary for the
-local transport domain, we needed to gather the identification numbers
-of all owning processes from the immediately adjacent states in the
-system. We solved this problem by traversing the graph of the matrix
-for each of the boundary states and determining which processes owned
-those adjacent states.
-
-To generate overlap, we use the identical algorithm but in this case
-we perform as many graph level traversals as the specified amount of
-overlap and collect additional information. For the overlapping domain
-decomposition case, given an input parallel decomposition for the
-Neumann-Ulam decomposition, for all local states $m$, including those
-in the overlap, we require access to $P_{mn}$ and $W_{mn}$ (or
-$P^T_{mn}$ and $W^T_{mn}$ for the adjoint method) for all possible
-states $n$. Doing this provides us with the data required by the
-estimators permitting $P_{mn}$ and $W_{mn}$ to be computed locally for
-any boundary transition.
-
-As an example, consider the analytic relations derived in
-Appendix~\ref{chap:parallel_theory} for the length of random walks in
-a Neumann-Ulam method and the amount of leakage from a domain
-considering a neutron diffusion problem. From these relations we might
-determine that, for our particular problem, if we can grow the domain
-by 10 additional discrete state transitions, we can reduce the number
-of histories that must be communicated by a significant fraction. We
-determine the information we must gather from neighboring domains to
-build the overlap by traversing the graph of the matrix outward from
-the boundary 10 levels. At each step, we find the data in the adjacent
-domain that we require and make that data local, incrementing the size
-of the overlap. Once the overlap has been gathered, one extra graph
-traversal on the boundary states is performed to get the new
-neighboring states and their owning processes.
-
-Once overlap has been generated, the transport domain for a single set
-is complete. However, if multiple sets are to be used, an additional
-set of parallel procedures is required to generate the necessary data
-structures. Figure~\ref{fig:msod_construction} gives a schematic
-representation of the MSOD construction process using 4 sets. For a
-problem where $P$ parallel processors are available and $S$ sets are
-to be used, each set is allocated $P/S$ processors for computation. On
-the first $P/S$ processors, the linear problem is generated. From the
-linear problem, the linear operator is used to construct the transport
-domain with overlap, the solution vector used to construct the
-tallies, and the forcing term vector used to construct the fixed
-source. Once these Monte Carlo data structures have been generated,
-they are broadcast to the remaining sets in the problem as shown in
-Figure~\ref{fig:msod_construction} such that we have $S$ replications
-of the Monte Carlo problem for a single instance of the original
-linear problem.
-
-\clearpage
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.55}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
-  \end{center}
-  \caption{\textbf{MSOD construction for 4 sets with overlap.}
-    \textit{The linear system is used to construct the Monte Carlo
-      transport domain on the primary set. The Monte Carlo data
-      structures are then broadcast among the blocks to all sets in
-      the system.}}
-  \label{fig:msod_construction}
-\end{figure}
-
-\subsubsection{MSOD Neumann Ulam Algorithm}
-\label{subsubsec:msod_algorithm}
-
-With the ability to generate the transport domain within an MSOD
-decomposition as well as the source and tally structures, we can now
-define how to perform a Neumann-Ulam solve using MSOD. Given by
-Algorithm~\ref{alg:msod_transport}, we begin by constructing the Monte
-Carlo data structures for each set including the transport domain, the
-source, and the tallies. Next, in line 4 we perform the asynchronous
-transport algorithm presented in \S~\ref{subsec:asynchronous_algorithm}
-for each of the individual sets. Adding overlap to
-Algorithm~\ref{alg:parallel_mc_algorithm} in this case simply modifies
-the set of local states and the set of neighboring states with the
-additional states gathered during overlap generation.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{MSOD Transport Sequence}}
-  \label{alg:msod_transport}
-  \begin{algorithmic}[1]
-    \State build MSOD domain
-    \State build Monte Carlo source
-    \State build Monte Carlo tally
-    \State perform parallel Neumann-Ulam transport in each set
-    \State combine set tallies
-    \State combine block tallies
-  \end{algorithmic}
-\end{algorithm}
-
-Once Monte Carlo transport is complete, the tallies for each
-individual computation must be gathered to the original set in order
-to build the correction vector within an MCSA sequence. To do this, we
-perform two parallel vector operations. First, overlap not only
-creates additional states in the local transport domain but also
-additional states in the local tally vector such that any events that
-occur in the local domain may be tallied in a local vector. Because of
-this, the tally vector in a single set will share local states with
-other domains and a parallel sum reduction operation is required to
-transform the tally into the original Neumann-Ulam parallel
-decomposition. Second, the tallies in individual sets must be combined
-and applied to the primary set in the problem that shares the
-processor space with the original linear problem. For this operation,
-we employ the concept of blocks as a subdivision of parallel space
-complementary to the concept of sets.
-
-Consider the schematic representation of the final parallel reduction
-presented in Figure~\ref{fig:msod_tally_reduction}. For this
-operation, the tally vectors in each set must be summed together and
-applied to the primary set. As each set is an exact replica of the
-others, the parallel decomposition of the tally vector is the same
-within all sets. We can take advantage of this by building a processor
-space that encapsulates each identical domain across all sets and then
-use this processor space for the reduction. For the schematic in
-Figure~\ref{fig:msod_tally_reduction}, the combined upper-left domains
-in each set create a block, giving 9 total blocks for this example
-problem. The processor space that encapsulates each of the upper-left
-domains is then used for the parallel reduction operation. This
-reduction will happen in a mutually exclusive manner for each of the
-blocks in the system.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.55}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
-  \end{center}
-  \caption{\textbf{MSOD tally reduction over blocks.} \textit{The
-      tally computed independently in each set is reduced across the
-      blocks to the primary set. The linear system vector is updated
-      on the primary set. In this example the tally vector in the
-      block containing upper-left subdomains is combined. Identical
-      reduction operations occur for all other blocks
-      simultaneously.}}
-  \label{fig:msod_tally_reduction}
-\end{figure}
-
-%%---------------------------------------------------------------------------%%
-\subsection{Parallel MCSA}
-\label{subsec:parallel_mcsa}
-With the parallel adjoint Neumann-Ulam solver implementation described
-above, the parallel implementation of an MCSA iteration is
-trivial. Recall the MCSA iteration procedure presented again here for
-clarity:
-\begin{subequations}
-  \begin{gather}
-    \label{eq:mcsa_r1}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \label{eq:mcsa_r2}
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_r3}
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \label{eq:mcsa_r4}
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_r5}
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
-  \end{gather}
-\end{subequations}
-In \S\ref{subsec:parallel_krylov_methods} we discuss parallel matrix and
-vector operations as utilized in conventional projection methods. We
-utilize these here for the parallel MCSA implementation for the
-computations required in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
-(\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}). In the first step, a
-parallel vector update is used to apply the residual to the previous
-iterate's solution in a Richardson iteration. In the next step and
-Eq~(\ref{eq:mcsa_r5}), the residual is computed by a parallel
-matrix-vector multiply and vector update. Once the correction is
-computed with a parallel adjoint Neumann-Ulam solve, this correction
-is applied to the solution with a parallel vector
-update. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector.
-
-To parallelize the solution of Eq~(\ref{eq:mcsa_r3}), we apply the
-MSOD Neumann-Ulam algorithm presented in the previous subsection. The
-problem is potentially replicated with overlap and the Monte Carlo
-procedure returns a correction vector, $\delta \ve{x}$, with the same
-parallel decomposition as the input linear problem. As was the case
-for generating the replications required for a multiple set
-implementation, the explicit matrix and vector operations required for
-the rest of an MCSA iteration occur only on the subset of processors
-containing the primary set\footnote{For applications where resiliency
-  is potentially of concern, the linear problem and subsequently the
-  operations in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
-  (\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}) may be replicated with
-  the penalty of broadcasting the data. Doing so may alleviate issues
-  where any part of the linear problem or residual is disturbed as now
-  there will be multiple copies of these data structures.}. Only the
-Monte Carlo data structures exist on the rest of the processors in the
-problem as replication of the other operations does not generate any
-more information that could be used to accelerate the time to
-solution.
-
-%%---------------------------------------------------------------------------%%
-\section{Criticality Calculations}
+\subsection{Light Water Reactor Fuel Assemblies}
 
 \subsection{C5G7 Benchmark}
-
-\subsection{Light Water Reactor Fuel Assemblies}
 

### 2015-02-05 doc/ttsp2013/paper/ttsp_2013.tex
updates to ttsp data

@@
 Predictive modeling and simulation capability requires the combination
-of high fidelity models, high performance computing hardware that can
-handle the intense computational loads required by these models, and
-modern algorithms for solving these problems that leverage this high
-performance hardware. For nuclear reactor analysis, this predictive
+of high fidelity models, high performance computing hardware, and
+modern algorithms. For nuclear reactor analysis, this predictive
 capability can enable tighter design tolerances for improved thermal
@@
 leveraged as an alternative to computing a solution over the entire
-domain a reactor core where diffusion theory is often valid.
+domain of a reactor core where diffusion theory is often valid.
 
@@
 believe both of these contributions in preconditioning and
-parallelizing the Neumann-Ulam algorithm are new and unique.
+parallelizing the Neumann-Ulam algorithm are new and unique. Although
+we have motivated this work by looking forward to exascale computing,
+resiliency of this solution technique in the presence of faults will
+not be considered here but instead be addressed in future work.
 
@@
 criticality problems used for the numerical studies in the this
-work. Next, we demonstrate that the basic Jacobi preconditioning
-applied in our previous work is unsatisfactory to achieve convergence
-for these problems and subsquently present our new preconditioning
-scheme. We then parallelize the MCSA algorithm and describe our new
-domain decomposition scheme deisgned to enable very large problems. We
-then apply this preconditioning scheme and parallel algorithm to the
-criticality problems and compare the performance of the preconditioned
-MCSA method to preconditioned Krylov methods. We then finish with
-conclusions and indication of future work.
+work. Next, we present our new preconditioning scheme consisting of an
+approximate inverse algorithm combined with solving an approximate
+Monte Carlo problem for the acceleration. We then parallelize the MCSA
+algorithm and describe our new domain decomposition scheme for the
+Neumann-Ulam Monte Carlo algorithm designed to enable very large
+problems. We then apply the preconditioning scheme and the parallel
+algorithm to the criticality problems and compare the performance of
+the preconditioned MCSA method to preconditioned Krylov methods. We
+then finish with conclusions and indication of future work.
 
@@
 Li demonstrate in \cite{ji_2013} that a necessary and sufficient
-condition for convergence of this algorithm is $\rho(\ve{H^*}) < 1$
+condition for convergence of this algorithm is 
+\begin{equation}
+  \rho(\ve{H^*}) < 1
+  \label{eq:necessary_condition}
+\end{equation} 
 with $\ve{H^*}$ defined element-wise as:
@@
   h^*_{ij} = \frac{(h^T_{ij})^2}{p_{ij}}\:,
+  \label{eq:condition_values}
+\end{equation}
+or for our choice of probabilities via
+Eq~(\ref{eq:adjoint_neumann_ulam})
+\begin{equation}
+  h^*_{ij} = h_{ij} w_{ij}\:,
   \label{eq:h_star}
@@
 \section{Preconditioning Algorithm}
+In this section we present our preconditioning algorithm.
 
@@
 \label{subsec:preconditioning}
-In most cases, at least a minimal amount of \textit{preconditioning}
-of the linear system will be required in order to use the class of
-stochastic methods described. Although these methods have no symmetry
-requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. Preconditioning serves as a
-means of achieving this by altering the eigenvalue spectrum of the
-iteration matrix. It is possible to use general left, right, and left/right
-preconditioning with MCSA by carefully considering the underlying
-Monte Carlo problem that will be solved with the Neumann-Ulam
-method. We consider here the general left/right preconditioned method
-as the left or right preconditioned methods can be inferred from its
-formulation.
-
-We consider a left preconditioner $\ve{M_L}$ and a right
-preconditioner $\ve{M_R}$. The left/right preconditioned linear
-problem is then:
+In most cases, at least a minimal amount of preconditioning of the
+linear system will be required in order to satisfy the necessary
+convergence condition given in Eq~(\ref{eq:necessary_condition}) for
+the Neumann-Ulam algorithm. Preconditioning serves as a means of
+achieving this by altering the eigenvalue spectrum of the adjoint
+iteration matrix and subsequently $\ve{H}^*$. It is possible to use
+general left and right preconditioning with MCSA by carefully
+considering the underlying Monte Carlo problem that will be solved
+with the Neumann-Ulam algorithm. We consider a left preconditioner,
+$\ve{M_L}$, and a right preconditioner, $\ve{M_R}$. The left/right
+preconditioned linear problem is then:
 \begin{equation}
@@
 \end{equation}
-To apply such a method to MCSA, we solve for the substituted variable
-$\ve{u}$ during the iteration sequence:
+To apply this preconditioning to MCSA, we solve for the substituted
+variable $\ve{u}$ during the iteration sequence:
 \begin{subequations}
   \begin{gather}
-    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
-    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
+    \ve{u}^{k+1/2} = \ve{u}^k + \ve{M}_L^{-1}\ve{r}^k\:,\\
+    \ve{r}^{k+1/2} = \ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2}\:,\\ 
+    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{M}_L^{-1}\ve{r}^{k+1/2}\:,\\ 
     \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
+    \ve{r}^{k+1} = \ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1}\:,
   \end{gather}
@@
 \begin{equation}
-  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
+  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} =
+  \ve{M}_L^{-1}\ve{r}^{k+1/2}\:,
   \label{eq:left_right_correction}
@@
 and note that the preconditioned residual of the substituted variable
-is now serving as the source and the new iteration matrix is:
+is now serving as the source and the new adjoint iteration matrix is:
 \begin{equation}
-  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
+  \ve{H}^T = \big(\ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\big)^T\:.
   \label{eq:left_right_iteration_matrix}
 \end{equation}
-As we require $(i,j)$ element-wise access to the iteration matrix in
-order to construct probabilities and weights for the Monte Carlo
-procedure from the Neumann-Ulam decomposition, the \textit{composite
-  operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
-matrix-matrix multiplication.
+As we require element-wise access to the iteration matrix in order to
+construct probabilities and weights from the Neumann-Ulam
+decomposition for the Monte Carlo procedure, the composite operator,
+$\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via matrix-matrix
+multiplication. Provided that $\ve{M}_L^{-1}$, $\ve{A}$, and
+$\ve{M}_R^{-1}$ are relatively sparse, then we can readily leverage
+sparse matrix-matrix multiplication algorithms that are in wide use in
+scalable algebraic multigrid implementations.
 
@@
 preconditioned (meaning that we do not recover $\ve{x}$), but instead
-part of a sequence for finding the substituted variable $\ve{u}$. We
-do, however, gain the benefits of this general preconditioning by
-building the iteration matrix in
+is part of a sequence for finding the substituted variable,
+$\ve{u}$. We do, however, gain the benefits of this general
+preconditioning by building the iteration matrix in
 Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
@@
 
-\subsubsection{Block Jacobi Preconditioning}
+\subsection{Sparse Approximate Inverse Preconditioning}
 
-\subsubsection{Sparse Approximate Inverse Preconditioning}
+\subsection{Approximate Monte Carlo Problem}
 

### 2015-02-05 doc/ned2015/slattery_wilson_evans.tex
udpating ned paper for formating

\documentclass[preprint,11pt]{elsarticle}
\usepackage[top=1.0in,bottom=1.0in,left=1.5in,right=1.5in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{longtable}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb,amsthm}
\usepackage[mathcal]{euscript}
\usepackage{tabularx}
\usepackage{tmadd,tmath}
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

\usepackage{booktabs}
\biboptions{sort&compress}
\journal{Nuclear Engineering Design}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%   BEGIN DOCUMENT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\begin{document}

\begin{frontmatter}

\title{A Spectral Analysis of the Domain Decomposed Monte Carlo Method for
  Linear Systems}

  \tnotetext[ornl-cr]{Notice: This manuscript has been authored by
    UT-Battelle, LLC, under contract DE-AC05-00OR22725 with the
    U.S. Department of Energy. The United States Government retains
    and the publisher, by accepting the article for publication,
    acknowledges that the United States Government retains a
    non-exclusive, paid-up, irrevocable, world-wide license to publish
    or reproduce the published form of this manuscript, or allow
    others to do so, for United States Government purposes.}

  \author[ornl]{Stuart R. Slattery\fnref{fn1}}
  \ead{slatterysr@ornl.gov}

  \fntext[fn1]{Computational Engineering and Energy Sciences Group,
    Computer Science and Mathematics Division}

  \author[ornl]{Thomas M. Evans\fnref{fn1}}
  \ead{evanstm@ornl.gov}

  \fntext[fn1]{Computational Engineering and Energy Sciences Group,
    Computer Science and Mathematics Division}

  \author[wisc]{Paul P.H. Wilson\fnref{fn2}}
  \ead{wilsonp@engr.wisc.edu}

  \fntext[fn2]{Engineering Physics Department, College of Engineering}

  \address[ornl]{Oak Ridge National Laboratory, 1 Bethel Valley Rd., Oak
    Ridge, TN 37831 U.S.A.}

  \address[wisc]{University of Wisconsin - Madison, 1500 Engineering Dr. ,
    Madison, WI 53706 U.S.A.}

%%---------------------------------------------------------------------------%%
  \begin{abstract}
    The domain decomposed behavior of the adjoint Neumann-Ulam Monte Carlo
    method for solving linear systems is analyzed using the spectral
    properties of the linear operator. Relationships for the average
    length of the adjoint random walks, a measure of convergence speed and
    serial performance, are made with respect to the eigenvalues of the
    linear operator. In addition, relationships for the effective optical
    thickness of a domain in the decomposition are presented based on the
    spectral analysis and diffusion theory. Using the effective optical
    thickness, the Wigner rational approximation and the mean chord
    approximation are applied to estimate the leakage fraction of
    stochastic histories from a domain in the decomposition as a measure
    of parallel performance and potential communication costs. The
    one-speed, two-dimensional neutron diffusion equation is used as a
    model problem to test the models for symmetric operators. In general,
    the derived approximations show good agreement with measured
    computational results.
\end{abstract}

\begin{keyword}
Monte Carlo \sep domain decomposition \sep MCSA \sep parallel \sep
computing \sep linear solvers
\end{keyword}

\end{frontmatter}

%%---------------------------------------------------------------------------%%
\section{Introduction}
\label{sec:intro}

An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable manuscript published in 1950 by Forsythe and Leibler
\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler in
fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method
adjoint to that described by Neumann and Ulam. In recent years, the
adjoint Neuman-Ulam method has been leveraged in Monte Carlo Synthetic
Acceleration (MCSA) methods that are competitive with conventional
methods including Krylov methods and are applicable to a wide variety
of linear systems \cite{evans_monte_2009,evans_monte_2012}. To advance
the applicability of these new MCSA methods, we therefore aim to apply
the underlying Monte Carlo methods to large-scale problems in parallel
computing environments.

To date, parallel Neumann-Ulam methods have been limited to full
domain replication with parallelism exploited through individual
histories \cite{alexandrov_efficient_1998}. In reactor physics Monte
Carlo applications, however, domain decomposition has been identified
as a key principle in moving forward in high performance computing to
enable higher fidelity simulations
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because these nearest-neighbor
communication sequences, specifically, have poor algorithmic strong
scaling \cite{gropp_high-performance_2001}.

The purpose of this study is to provide a simple, analytic theory
based on the properties of the linear system that will allow for
estimates of the domain decomposed behavior of the adjoint
Neumann-Ulam method. When solving problems where the linear operator
is symmetric, a host of analytic theories exist based on the
eigenvalue spectrum of the operator that characterize their behavior
in the context of deterministic linear solvers. Using past work, these
theories are adapted to the domain decomposed adjoint Neumann-Ulam
method using the one-speed, two-dimensional neutron diffusion
equation. In this paper we describe the adjoint Neumann-Ulam Monte
Carlo method followed by a presentation of the model problem. Using
the linear system generated by the discretization of the model
problem, we use a spectral analysis to generate analytic relations for
the eigenvalues of the operator based on system parameters. Using the
eigenvalue spectra, we then build relationships to characterize the
transport of stochastic histories in a decomposed domain and the
fraction of histories that leak from a domain and will therefore have
to be communicated. Finally, we compare these analytic results to
numerical experiments conducted with the model problem and draw
conclusions looking towards future work.

%%---------------------------------------------------------------------------%%
\section{Adjoint Neumann-Ulam Method}
\label{sec:adjoint_nu}

We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:.
  \label{eq:linear_problem}
\end{equation}
Choosing the adjoint method Monte Carlo method to invert the linear
operator, we define the linear system adjoint to
Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator. We can derive a
Monte Carlo estimator from the adjoint method that will also give the
solution vector, $\ve{x}$. We first rearrange
Eq~(\ref{eq:adjoint_linear_problem}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
where $\ve{H} = \ve{I} - \ve{A}$ is the \textit{iteration matrix}.
This yields the adjoint Neumann series:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
For this series to converge, we require that the spectral radius of
$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
eigenvalues and therefore has the same spectral radius. We expand this
summation to yield a series of transitions that can be approximated by
a Monte Carlo random walk sequence:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
To find the solution to Eq~(\ref{eq:linear_problem}) we build an
estimator for the adjoint solution from this series expansion. The
adjoint estimator can be related to the solution by defining the
following inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
require two constraints to close the system. We use
Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

In order to perform the random walks required to approximate the
summation in Eq~(\ref{eq:adjoint_neumann_solution}), we build the
column-scaled transition probability matrix and associated transition
weights as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
  \frac{|h_{ji}|}{p_{ij}},
  \label{eq:adjoint_probability}
\end{equation}
such that for each step, $m$, in the random walk, we expect to select
a new state, $j$, from the current state, $i_m$, by sampling the
$i_m^{th}$ row of the probability matrix as a discrete distribution,
and updating the weight, $W_m$, accordingly.  Using our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the Monte Carlo estimator for the solution in
state $i$ for a particular random walk permutation, $\nu$, is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $W_m$ is the history weight on the $m^{th}$
step. Finally, the expectation value using all permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which if expanded directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{m=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_m}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
    h_{i_{m-1},i_m} \delta_{i_m,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
where $b_{i_0}$ will be the sampled source starting weight, therefore
also providing an unbiased Monte Carlo estimate of the solution.

We also desire a criteria for random walk termination for problems
where only an approximate solution is necessary. For the adjoint
method, we utilize a relative weight cutoff parameter:
\begin{equation}
  W_f = W_c b_{i_0}\:,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is the \textit{weight cutoff}. The adjoint random walk
will then be terminated after $m$ steps if $W_m < W_f$ as tally
contributions become increasingly small, effectively establishing a
measure for the degree of approximation of the solution.

%%---------------------------------------------------------------------------%%
\section{Model Problem}
\label{sec:model_problem}

For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil over a grid of size $h$ \cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis}
\label{sec:spectral_analysis}

The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the eigenvalues of the iteration matrix. We
will compute these eigenvalues by assuming eigenfunctions of the form
\cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these eigenfunctions should
give the eigenvalues of the matrix as they exist on the unit circle in
the complex plane.

\subsection{Iteration Matrix Spectrum}
\label{subsec:iteration_spectrum}

For the model problem, we first compute the eigenvalues for the
diffusion operator $\ve{D}$ by applying the operator to the
eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
preconditioning is sufficient to reduce the spectral radius of the
iteration matrix below unity and therefore ensure convergence of the
Neumann series. The preconditioner in this case is then $\ve{M} =
diag(\ve{D})$ such that we are solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogeneous domain, the scaling factor, $\alpha$, is the
same for all rows in the operator and defined as the $\phi_{i,j}$
coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

The spectral radius of the iteration matrix is obtained by seeking its
largest eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the eigenfunctions defined by
Eq~(\ref{eq:eigenfunction_form}) we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. We find that the maximum eigenvalue exists when $p=q=0$,
giving the following for the spectral radius of the Jacobi
preconditioned iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

\Subsection{Neumann Series Convergence}
\label{subsec:neumann_convergence}

The adjoint Monte Carlo method is effectively an approximation to a
stationary method. Stationary methods for linear systems arise from
splitting the operator in Eq~(\ref{eq:linear_problem}) and iterating:
\begin{equation}
  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
  index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
error at the $k^{th}$ iterate, the error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we
then have:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}

In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a random
walk of average length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution})
\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
length, or the number of transitions before the termination of a
history (either by the weight cutoff, absorption, or exiting the
global domain) is therefore approximately the number of stationary
iterations required to converge to the specified tolerance. In the
case of the adjoint Neumann-Ulam method, no such tolerance exists,
however, we have specified a weight cutoff, $W_c$, that determines
when low-weight histories will be prematurely terminated as their
contributions are deemed minute. After $k$ iterations, a stationary
method is terminated as the error has reached some fraction,
$\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
Neumann-Ulam method, if we take this fraction to be the weight cutoff,
a measure of how accurately the contributions of a particular history
to the solution are tallied, we then have the following relationship
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method.

\subsection{Domain Leakage Approximations}
\label{subsec:domain_leak_approx}

In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
          \label{eq:leakage_fraction}
\end{equation}
For their studies, Siegel and colleagues assumed that the value of
$\Lambda$ was dependent on the total cross section of the system via
the Wigner rational approximation. Outlined more thoroughly by Hwang's
chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
rational approximation and the mean chord approximation as a means to
estimate the leakage fraction.

In the case of domain decomposed linear operator equations, we can use
diffusion theory to estimate the optical thickness of a domain in the
decomposition and the corresponding leakage fraction in terms of
properties of the linear operator and the discretization. To begin we
must first calculate the mean distance a Monte Carlo history will move
in the grid by computing the mean squared distance of its movement
along the chord of length $l$ defined across the domain. After a
single transition a history will have moved a mean squared distance
of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the discrete grid elements along the chord
and $n_s$ is the number of grid elements a history will move on
average every transition. For our diffusion model problem, $n_s$ would
equate to the expected number of states in the $i$ (or $j$ as the problem is
symmetric) direction that a history will move in a single
transition and is dependent on the stencil used for the
discretization. After $k$ transitions in the random walk, the history
will have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our chord is of length $l$ and there are $n_i$ grid elements (or
states to which a history may transition) along that chord, then $h =
l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimensionality of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo history in the domain. We can readily interpret $\tau$ to be the
\textit{effective optical thickness} of a domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

For optically thin domains, we expect that most histories will be
communicated, while optically thick domains will leak the fraction of
histories that did not interact within. Using the optical thickness
defined in Eq~(\ref{eq:optical_thickness_3}), we can then complete the
leakage approximations by defining the bounds of $\tau \rightarrow 0,
\Lambda \rightarrow 1$ and $\tau \rightarrow \infty, \Lambda
\rightarrow \tau^{-1}$.  With these bounds we can then define the
leakage fraction out of a domain for the adjoint Neumann-Ulam method
using the Wigner rational approximation:
\begin{equation}
  \Lambda = \frac{1}{1+\tau}\:,
  \label{eq:wigner_domain_leakage}
\end{equation}
and using the mean-chord approximation:
\begin{equation}
  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
  \label{eq:mean_chord_domain_leakage}
\end{equation}
Here, the leakage fraction is explicitly bound to the eigenvalues of
the iteration matrix, the size of the domain, the content of the
discretization stencil, and the weight cutoff selected to terminate
low weight histories.

%%---------------------------------------------------------------------------%%
\section{Numerical Experiments}
\label{sec:numerical_experiments}

To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4in,clip]{measured_spec_rad.png}
    \end{center}
    \caption{Measured and analytic preconditioned diffusion operator
      spectral radius as a function of the absorption cross section to
      scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
      $h=1.0$ were used. The red data was computed numerically by an
      eigensolver while the black dashed data was generated by
      Eq~(\ref{eq:iteration_radius}).}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

\Subsection{Random Walk Length}
\label{subsec:walk_length}

With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and the absorption cross varied from 0
to 100 while the scattering cross section was fixed at unity. Three
weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
used with 10,000 histories generated by a point source of strength 1
in the center of the domain. For each of the histories, the number of
transitions made was tallied to provide an effective value of $k$ for
each history. This value was then averaged over all histories to get a
measured value of $k$ for the particular operator. On the left,
Figure~\ref{fig:measured_length} presents these measurements as well
as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
the right, Figure~\ref{fig:measured_length} gives the relative error
between the predicted and observed results. We note good qualitative
agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=6.5in,clip]{measured_length.png}
    \end{center}
    \caption{Measured and analytic random walk length as a function of
      the iteration matrix spectral radius. The weight cutoff was
      varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
      plot, the red data was computed numerically by an adjoint
      Neumann-Ulam implementation while the black dashed data was
      generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
      relative error between the predicted and measured results is
      presented for each weight cutoff.}
    \label{fig:measured_length}
  \end{spacing}
\end{figure}

\subsection{Domain Leakage}
\label{subsec:domain_leakage}

Finally, we seek to measure the leakage from a domain in a domain
decomposed Monte Carlo calculation and assess the quality of our
analytic relation for the optical thickness of domain and the
associated leakage approximations. For this experiment, a square grid
with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
direction with measurements occurring in the central domain without
boundary grid points. For the cross sections, the absorption cross
section was varied from 1 to 100 while the scattering cross section
was set to zero to create a purely absorbing environment with weight
cutoff of \sn{1}{-4}. The optical thickness of these domains will vary
as a function of the absorption cross section if the other parameters
are fixed. To compute the optical thickness, along with the spectral
radius as given by Eq~(\ref{eq:iteration_radius}), we also need the
parameters $n_i$ and $n_s$ which respectively describe the typical
domain length and the average number of states moved along that
typical length per history transition. For our grid above, the domains
are varied in size with $50 \times 50$, $100 \times 100$, and $200
\times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$ grid
points or states along the typical length of the domain
respectively. Looking at the Laplacian stencil in
Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
will only move a single state in either the $i$ or $j$ directions due
to the symmetry of the problem. Furthermore, if we choose the $i$
direction, not all states we will transition to will move the history
in that direction. Therefore, we look to the definition of the
iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
to estimate the $n_s$ parameter. For a particular transition starting
at state $(i,j)$, 6 of the 8 possible new states in the stencil move
the history in $i$ direction with relative coefficients of 4 for
moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
i,\pm j)$. These coefficients dictate the frequency those states are
visited relative to the others. For those 6 states we can visit along
the typical length, their sum is 12 out of the total 20 for the
coefficients for all possible states with their ratio giving $n_s =
\frac{3}{5}$.

To compute the leakage fraction numerically, \sn{3}{5} histories were
sampled from a uniform source of strength unity over the global
domain. At the start of a stage of histories, the number of histories
starting in the center domain was computed and as the stage
progressed, the number of histories that exited that domain was
tallied with the ratio of the two numbers providing a numerical
measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
gives the domain leakage measurements for the domain in the center of
the global grid as well as the analytic result computed by
Eqs~(\ref{eq:wigner_domain_leakage}) and
(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
matrix spectral radius.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4.5in,clip]{leakage_variation.png}
    \end{center}
    \caption{Measured and analytic domain leakage as a function of the
      iteration matrix spectral radius. To test the behavior with
      respect to domain size, $n_i=50$, $n_i=100$,and $n_i=200$ were
      used. The red data was computed numerically by a
      domain-decomposed adjoint Neumann-Ulam implementation, the black
      dashed data was generated by
      Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
      approximation, and the dashed-dotted black data was generated by
      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
      approximation.}
    \label{fig:measured_leakage}
  \end{spacing}
\end{figure}
Again, we note good qualitative agreement between the measured and
analytic quantities but we begin to see the limits of the leakage
approximations. To compare the quality of the two approximations, the
absolute error between the computed leakage fraction and that
generated by the Wigner rational and mean chord approximations is
plotted in Figure~\ref{fig:leakage_error} for all domain sizes
tested. 
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4.5in,clip]{leakage_error.png}
    \end{center}
    \caption{Measured and analytic domain leakage absolute error as a
      function of the iteration matrix spectral radius.  To test the
      behavior with respect to domain size, $n_i=50$ (green),
      $n_i=100$ (blue), and $n_i=200$ (red) were used. The dashed
      lines represent the error using the Wigner rational
      approximation while the solid lines represent the error using
      the mean-chord approximation.}
    \label{fig:leakage_error}
  \end{spacing}
\end{figure}
From these error results, the mean chord approximation is shown to
have a lower error for ill-conditioned systems as compared to the
Wigner approximation while the Wigner approximation produces less
error for more well-conditioned systems. We also note that for the
optically thick domains, the error is likely corresponded to that
observed in Figure~\ref{fig:measured_length} for the $k$ parameter
while the large relative error in $k$ for optically thin domains does
not affect the approximation significantly. In general, the mean chord
approximation is a better choice to estimate the leakage fraction in a
domain from the adjoint Neumann-Ulam method and except for a single
data point with $n_i=50$, the mean chord approximation yielded leakage
fractions within 0.05 of the measured results. As the domain becomes
more optically thick (with both increasing $n_i$ and decreasing
$\rho(\ve{H})$), the approximations are more accurate.

%%---------------------------------------------------------------------------%%
\section{Conclusion}

We have presented an analytic analysis of the domain decomposed
behavior of the adjoint Neumann Ulam method for linear systems. Good
agreement was observed for the derived analytic relationships for
random walk length and leakage fraction when compared to numerical
results generated by a domain decomposed implementation of the adjoint
method.

In future work, these relationships will serve as guidelines for
selecting an appropriate parallel algorithm strategy and provide a
basis for performance models for future parallel implementations of
the adjoint Neumann-Ulam method. To be applicable to a broader range
of problems, these models could potentially be extended to
non-symmetric systems. In addition, these relations will be used to
analyze parallel implementations of Monte Carlo synthetic acceleration
methods that leverage the adjoint method and serve as an initial
grounds for assessing their feasibility for large-scale problems.

%%---------------------------------------------------------------------------%%
\section{Acknowledgements}
This material is based upon work supported by the U.S. Department of
Energy, Office of Science, Advanced Scientific Computing Research
program.

This work was performed under appointment to the Nuclear Regulatory
Commission Fellowship program at the University of Wisconsin - Madison
Engineering Physics Department.

%%---------------------------------------------------------------------------%%
\bibliographystyle{ieeetr} \bibliography{references}

\end{document}

### 2015-02-05 doc/ned2015/slattery_wilson_evans.tex
adding references to ned paper and doing spell check

@@
 
-\begin{document}
-
 \begin{frontmatter}
@@
 
-  \author[ornl]{Thomas M. Evans\fnref{fn1}}
+  \author[ornl]{Thomas M. Evans\fnref{fn2}}
   \ead{evanstm@ornl.gov}
 
-  \fntext[fn1]{Computational Engineering and Energy Sciences Group,
-    Computer Science and Mathematics Division}
+  \fntext[fn1]{Radiation Transport Group, Reactor and Nuclear Systems
+    Division}
 
-  \author[wisc]{Paul P.H. Wilson\fnref{fn2}}
+  \author[wisc]{Paul P.H. Wilson\fnref{fn3}}
   \ead{wilsonp@engr.wisc.edu}
 
-  \fntext[fn2]{Engineering Physics Department, College of Engineering}
+  \fntext[fn3]{Engineering Physics Department, College of Engineering}
 
   \address[ornl]{Oak Ridge National Laboratory, 1 Bethel Valley Rd., Oak
-    Ridge, TN 37831 U.S.A.}
+    Ridge, TN 37831, U.S.A.}
 
-  \address[wisc]{University of Wisconsin - Madison, 1500 Engineering Dr. ,
-    Madison, WI 53706 U.S.A.}
+  \address[wisc]{University of Wisconsin - Madison, 1500 Engineering Dr.,
+    Madison, WI 53706, U.S.A.}
 
@@
 
-An alternative approach to approximate matrix inversion is to employ
-Monte Carlo methods that sample a distribution with an expectation
-value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here
-an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler in
-fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. In 1952 Wasow
-provided a more formal explanation of Von Neumann and Ulam's method
-\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
-\cite{spanier_monte_1969} present additional detail on a method
-adjoint to that described by Neumann and Ulam. In recent years, the
-adjoint Neuman-Ulam method has been leveraged in Monte Carlo Synthetic
-Acceleration (MCSA) methods that are competitive with conventional
-methods including Krylov methods and are applicable to a wide variety
-of linear systems \cite{evans_monte_2009,evans_monte_2012}. To advance
-the applicability of these new MCSA methods, we therefore aim to apply
-the underlying Monte Carlo methods to large-scale problems in parallel
-computing environments.
+An alternative approach to approximate matrix inversion is to employ Monte
+Carlo methods that sample a distribution with an expectation value equivalent
+to that of the inverted operator. Such methods have been in existence for
+decades with the earliest reference noted here an enjoyable manuscript
+published in 1950 by Forsythe and Leibler \cite{forsythe_matrix_1950}. In
+their outline, Forsythe and Leibler in fact credit the creation of this
+technique to J. Von Neumann and S.M. Ulam some years earlier than its
+publication. In 1952 Wasow provided a more formal explanation of Von Neumann
+and Ulam's method \cite{wasow_note_1952} and Hammersley and Handscomb's 1964
+monograph \cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
+\cite{spanier_monte_1969} present additional detail on a method adjoint to
+that described by Neumann and Ulam. To improve their convergence, Halton
+leveraged the Monte Carlo algorithm in an iterative refinement scheme where a
+residual Monte Carlo solution serves as a correction to a fixed point
+iteration \cite{halton_sequential_1962}. In recent years, the adjoint
+Neumann-Ulam method has been leveraged in Monte Carlo Synthetic Acceleration
+(MCSA) methods that are competitive with conventional methods including Krylov
+methods and are applicable to a wide variety of linear systems
+\cite{evans_monte_2009,evans_monte_2014}. To advance the applicability of
+these new MCSA methods, we therefore aim to apply the underlying Monte Carlo
+methods to large-scale problems in parallel computing environments.
 
-To date, parallel Neumann-Ulam methods have been limited to full
-domain replication with parallelism exploited through individual
-histories \cite{alexandrov_efficient_1998}. In reactor physics Monte
-Carlo applications, however, domain decomposition has been identified
-as a key principle in moving forward in high performance computing to
-enable higher fidelity simulations
-\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
-this, we recognize from the literature that stochastic histories must
-be transported from domain to domain as the simulation progresses and
-they transition to states that are not in the local domain. Because we
-have chosen a domain decomposition strategy in a parallel environment,
-this means that communication of these histories must occur between
-compute nodes owning neighboring pieces of the global domain. We wish
-to characterize this communication not only because communication is
-in general expensive, but also because these nearest-neighbor
-communication sequences, specifically, have poor algorithmic strong
-scaling \cite{gropp_high-performance_2001}.
+To date, parallel Neumann-Ulam methods have been limited to full domain
+replication with parallelism exploited through individual histories
+\cite{alexandrov_efficient_1998}. In reactor physics Monte Carlo applications,
+however, domain decomposition has been identified as a key principle in moving
+forward in high performance computing to enable higher fidelity simulations
+\cite{brunner_comparison_2006,brunner_efficient_2009,siegel_analysis_2012}. To
+accomplish this, we recognize from the literature that stochastic histories
+must be transported from domain to domain as the simulation progresses and
+they transition to states that are not in the local domain. Because we have
+chosen a domain decomposition strategy in a parallel environment, this means
+that communication of these histories must occur between compute nodes owning
+neighboring pieces of the global domain. We wish to characterize this
+communication not only because communication is in general expensive, but also
+because these nearest-neighbor communication sequences, specifically, have
+poor algorithmic strong scaling \cite{gropp_high-performance_2001}.
 
-The purpose of this study is to provide a simple, analytic theory
-based on the properties of the linear system that will allow for
-estimates of the domain decomposed behavior of the adjoint
-Neumann-Ulam method. When solving problems where the linear operator
-is symmetric, a host of analytic theories exist based on the
-eigenvalue spectrum of the operator that characterize their behavior
-in the context of deterministic linear solvers. Using past work, these
-theories are adapted to the domain decomposed adjoint Neumann-Ulam
-method using the one-speed, two-dimensional neutron diffusion
-equation. In this paper we describe the adjoint Neumann-Ulam Monte
-Carlo method followed by a presentation of the model problem. Using
-the linear system generated by the discretization of the model
-problem, we use a spectral analysis to generate analytic relations for
-the eigenvalues of the operator based on system parameters. Using the
-eigenvalue spectra, we then build relationships to characterize the
-transport of stochastic histories in a decomposed domain and the
-fraction of histories that leak from a domain and will therefore have
-to be communicated. Finally, we compare these analytic results to
-numerical experiments conducted with the model problem and draw
-conclusions looking towards future work.
+The purpose of this study is to provide a simple, analytic theory based on the
+properties of the linear system that will allow for estimates of the domain
+decomposed behavior of the adjoint Neumann-Ulam method. When solving problems
+where the linear operator is symmetric, a host of analytic theories exist
+based on the eigenvalue spectrum of the operator that characterize their
+behavior in the context of deterministic linear solvers. Using past work,
+these theories are adapted to the domain decomposed adjoint Neumann-Ulam
+method using the one-speed, two-dimensional neutron diffusion equation. In
+this paper we describe the adjoint Neumann-Ulam Monte Carlo method followed by
+a presentation of the model problem. Using the linear system generated by the
+discretization of the model problem, we use a spectral analysis to generate
+analytic relations for the eigenvalues of the operator based on system
+parameters. Using the eigenvalue spectra, we then build relationships to
+characterize the transport of stochastic histories in a decomposed domain and
+the fraction of histories that leak from a domain and will therefore have to
+be communicated. Finally, we compare these analytic results to numerical
+experiments conducted with the model problem and draw conclusions looking
+towards future work.
 
@@
 
-\Subsection{Neumann Series Convergence}
+\subsection{Neumann Series Convergence}
 \label{subsec:neumann_convergence}
@@
 \begin{figure}[ht!]
-  \begin{spacing}{1.0}
   \begin{center}
@@
   \label{fig:measured_spec_rad}
-  \end{spacing}
 \end{figure}
 
-\Subsection{Random Walk Length}
+\subsection{Random Walk Length}
 \label{subsec:walk_length}
@@
   \label{fig:measured_length}
-  \end{spacing}
 \end{figure}
@@
   \label{fig:measured_leakage}
-  \end{spacing}
 \end{figure}
@@
   \label{fig:leakage_error}
-  \end{spacing}
 \end{figure}
@@
 %%---------------------------------------------------------------------------%%
-\section{Acknowledgements}
+\section{Acknowledgments}
 This material is based upon work supported by the U.S. Department of

### 2015-02-05 doc/ned2015/slattery_wilson_evans.tex
formatting figures in ned paper

@@
 Neumann-Ulam method has been leveraged in Monte Carlo Synthetic Acceleration
-(MCSA) methods that are competitive with conventional methods including Krylov
-methods and are applicable to a wide variety of linear systems
-\cite{evans_monte_2009,evans_monte_2014}. To advance the applicability of
-these new MCSA methods, we therefore aim to apply the underlying Monte Carlo
-methods to large-scale problems in parallel computing environments.
+(MCSA) methods that, for certain classes of problems, are competitive with
+conventional methods including Krylov methods and are applicable to a wide
+variety of linear systems \cite{evans_monte_2009,evans_monte_2014}. To advance
+the applicability of these new MCSA methods, we therefore aim to apply the
+underlying Monte Carlo methods to large-scale problems in parallel computing
+environments.
 
@@
   \begin{center}
-    \includegraphics[width=4in,clip]{measured_spec_rad.png}
+    \includegraphics[width=\textwidth]{measured_spec_rad.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6.5in,clip]{measured_length.png}
+    \includegraphics[width=\textwidth]{measured_length.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.5in,clip]{leakage_variation.png}
+    \includegraphics[width=\textwidth]{leakage_variation.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.5in,clip]{leakage_error.png}
+    \includegraphics[width=\textwidth]{leakage_error.png}
   \end{center}

### 2015-02-09 doc/ned2015/slattery_wilson_evans.tex
updating ned draft

@@
 \section{Acknowledgments}
+This work was performed under appointment to the Nuclear Regulatory
+Commission Fellowship program at the University of Wisconsin - Madison
+Engineering Physics Department.
+
 This material is based upon work supported by the U.S. Department of
@@
 program.
-
-This work was performed under appointment to the Nuclear Regulatory
-Commission Fellowship program at the University of Wisconsin - Madison
-Engineering Physics Department.
 

### 2015-03-10 doc/siam_cse_2015/siam_cse_2015.tex
adding siam cse 2015 talk

%%---------------------------------------------------------------------------%%
%% SIAM CSE 2015 Presentation
%%---------------------------------------------------------------------------%%

\documentclass{beamer}
\usefonttheme[onlymath]{serif}
\usepackage{colortbl}
\usepackage{longtable,ltcaption,array}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{caption}
\usepackage[labelformat=empty]{subcaption}
\usepackage{tabularx}

%%---------------------------------------------------------------------------%%
%% THEME SETUP

\usetheme{CambridgeUS}
\usecolortheme{spruce}

%%---------------------------------------------------------------------------%%
%% SETUP STUFF
%%---------------------------------------------------------------------------%%

\logo{\includegraphics[width=2cm]{new_logo}}

\setbeamercolor{item}{fg=MSUgreen}
\setbeamertemplate{headline}{}
\setbeamertemplate{navigation symbols}{}

\newlength{\DUtablewidth} % internal use in tables

%%---------------------------------------------------------------------------%%
%% MATH STUFF
%%---------------------------------------------------------------------------%%

\newcommand{\Pn}{\textit{P}$\negthinspace_N$}
\newcommand{\SPn}{\textit{SP}$\negthinspace_N$}

\newcommand{\vOmega}{\ensuremath{\ve{\Omega}}}
\newcommand{\hOmega}{\ensuremath{\hat{\ve{\Omega}}}}

\newcommand{\sigs}{\ensuremath{\sigma_{\text{s}}}}
\newcommand{\sigf}{\ensuremath{\sigma_{\text{f}}}}
\newcommand{\sigsm}{\ensuremath{\sigma_{\text{s}m}}}
\newcommand{\sigsn}{\ensuremath{\sigma_{\text{s}n}}}

\newcommand{\phig}{\ensuremath{\Phi}}
\newcommand{\Sigmag}{\ensuremath{\mathbf{\Sigma}}}
\newcommand{\Dg}{\ensuremath{\mathbb{D}}}
\newcommand{\Fg}{\ensuremath{\mathbb{F}}}
\newcommand{\Qg}{\ensuremath{\mathbb{Q}}}
\newcommand{\ug}{\ensuremath{\mathbb{U}}}
\newcommand{\Ag}{\ensuremath{\mathbb{A}}}
\newcommand{\Bg}{\ensuremath{\mathbb{B}}}
\newcommand{\Cg}{\ensuremath{\mathbb{C}}}
\newcommand{\Jg}{\ensuremath{\mathbb{J}}}
\newcommand{\sg}{\ensuremath{\mathbb{S}}}

\newcommand{\aphig}{\ensuremath{\Phi^{\dagger}}}
\newcommand{\aSigmag}{\ensuremath{\mathbf{\Sigma}^{\dagger}}}

\newcommand{\apsi}[1]{\ensuremath{\psi^{\dagger\,#1}}}
\newcommand{\aphi}[1]{\ensuremath{\phi^{\dagger\,#1}}}
\newcommand{\aq}[1]{\ensuremath{q^{\dagger\,#1}}}
\newcommand{\ak}{\ensuremath{k^{\dagger}}}

\newcommand{\normal}{\ensuremath{\hat{\ve{n}}}}

% Phantom minus sign for helping with alignment of matrices
\newcommand{\phmin}{\ensuremath{\phantom{-}}}

\DeclareMathOperator{\diag}{diag}

%%---------------------------------------------------------------------------%%
%% TITLE

\title[Parallel Monte Carlo Solvers]{Parallel Algorithms for Monte Carlo
  Linear Solvers}
\author[Slattery, Hamilton, Evans]{Stuart Slattery
  \\ Steven Hamilton \\ Tom Evans}
\date[3/17/15]{March 17, 2015}
\institute[ORNL]{\small Oak Ridge National Laboratory}
\titlegraphic{\includegraphics[width=1.5in]{ORNL_Balls.pdf}}

\defbeamertemplate{footline}{titlefoot}{
    \vspace{-0.5cm}
    \hspace*{0.25cm}
    \includegraphics[width=2cm]{doe_logo}
    \hfill
    \includegraphics[width=3.25cm]{WordMarkLeaf}
    \hspace*{0.5cm}
}

\setbeamertemplate{title page}
{
  \begin{tabular}{cr}
    \begin{minipage}{4.7cm}
      {\bf \Large\textcolor{MSUgreen}{\inserttitle}}\\

      \vspace{2\baselineskip}
      \insertauthor\\

      \insertinstitute\\

      \insertdate
    \end{minipage}
    &
    \begin{minipage}{5cm}
      \raggedright
      \vspace{-0.75cm}
      \includegraphics[height=9.8cm]{combined_background}
    \end{minipage}
  \end{tabular}
}

%%---------------------------------------------------------------------------%%
\begin{document}

%%---------------------------------------------------------------------------%%

{
\setbeamertemplate{logo}{}
\setbeamertemplate{footline}[titlefoot]
\begin{frame}
\titlepage
\end{frame}
}

%%---------------------------------------------------------------------------%%
\begin{frame}{Acknowledgements}

  This material is based upon work supported by the U.S. Department of
  Energy, Office of Science, Advanced Scientific Computing Research
  program.

  \vfill

  This research used resources of the Oak Ridge Leadership Computing
  Facility at the Oak Ridge National Laboratory, which is supported by the
  Office of Science of the U.S. Department of Energy under Contract No.
  DE-AC05-00OR22725.

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Motivation}
  \vfill
\begin{itemize}
  \item As we move towards exascale computing, the rate of errors is expected
  to increase dramatically
  \vfill
  \begin{itemize}
  \item The probability that a compute node will fail during the course
    of a large scale calculation may be near 1
  \end{itemize}
  \vfill
\item Algorithms need to not only have increased concurrency/scalability
  but have the ability to recover from hardware faults
  \begin{itemize}
  \item Lightweight machines
  \item Heterogeneous machines
  \item Both characterized by low power and high concurrency
  \end{itemize}
  \vfill
\end{itemize}
\vfill
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Towards Exascale Concurrency and Resiliency}
  \begin{itemize}
    \item Two basic strategies:
      \vfill
      \begin{enumerate}
        \item State with current ``state of the art'' methods and make
          incremental modifications to improve scalability and fault
          tolerance
          \begin{itemize}
            \item Many efforts are heading in this direction, attempting
              to find additional concurrency to exploit
          \end{itemize}
        \vfill
        \item Start with methods having natural scalability and resiliency
          aspects and work at improving performance (e.g. Monte Carlo)
          \begin{itemize}
            \item Soft failures a component of the tally variance
            \item Hard failures potentially mitigated by replication
            \item Concurrency enabled by several levels of parallelism
          \end{itemize}
      \end{enumerate}
  \end{itemize}
  \vfill
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Outline}

  \begin{itemize}
  \item Monte Carlo Linear Solvers
    \vfill
  \item Domain Decomposition and Replication
    \vfill
  \item Scaling Studies
    \vfill
  \item Matrix-Free and Stochastic Approximate Inverse Algorithms
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}

  \center Monte Carlo Methods

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo for Linear Systems}
  \begin{itemize}
    \item Suppose we want to solve $\mathbf{Ax}=\mathbf{b}$
    \vfill
    \item If $\rho(\mathbf{I-A})<1$, we can write the solution using the
      Neumann series
      \begin{equation*}
        \mathbf{x} = \sum_{n=0}^{\infty} (\mathbf{I-A})^n \mathbf{b}
         = \sum_{n=0}^{\infty} \mathbf{H}^n \mathbf{b}
      \end{equation*}
      where $\mathbf{H} \equiv ( \mathbf{I-A} )$ is the Richardson
      iteration matrix 
      \vfill
    \item Build the Neumann series stochastically
  \end{itemize}

  \[
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}
  \]

  \begin{itemize}
  \item Define a sequence of state transitions
  \end{itemize}
  \vspace*{-0.1in}
  \[
  \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
  \rightarrow i_{k}
  \]

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Forward Monte Carlo}
\begin{itemize}
  \item Choose a row-stochastic matrix $\mathbf{P}$ and weight matrix
    $\mathbf{W}$ such that $\mathbf{H} = \mathbf{P} \circ \mathbf{W}$
  \item Typical choice (Monte Carlo Almost-Optimal):
    \begin{equation*}
      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ij}| }
      {\sum_{j=1}^{N} | \mathbf{H}_{ij} | }
    \end{equation*}
  \item To compute solution component $\mathbf{x}_i$:
    \begin{itemize}
      \item Start a history in state $i$ (with initial weight of 1)
      \item Transition to new state $j$ based probabilities determined by
        $\mathbf{P}_i$
      \item Modify history weight based on corresponding entry in
        $\mathbf{W}_{ij}$
      \item Add contribution to $\mathbf{x}_i$ based on current history weight
        and value of $\mathbf{b}_j$
    \end{itemize}
  \item A given random walk can only contribute to a single component of
    the solution vector
\end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Sampling Example (Forward Monte Carlo)}
  \begin{itemize}
    \item Suppose
  \begin{equation*}
    \mathbf{A} = \begin{bmatrix}
      \phmin 1.0 & -0.2 & -0.6 \\
      -0.4 & \phmin 1.0 & -0.4 \\
      -0.1 & -0.4 & \phmin 1.0 \end{bmatrix} \to
    \mathbf{H} \equiv (\mathbf{I-A}) = \begin{bmatrix}
       0.0 &  0.2 &  0.6 \\
       0.4 &  0.0 &  0.4 \\
       0.1 &  0.4 &  0.0 \end{bmatrix}
  \end{equation*}
    then
  \begin{equation*}
    \mathbf{P} = \begin{bmatrix}
       0.0 & 0.25 & 0.75 \\
       0.5 &  0.0 & 0.5 \\
       0.2 &  0.8 & 0.0 \end{bmatrix}, \quad
    \mathbf{W} = \begin{bmatrix}
       0.0 &  0.8 &  0.8 \\
       0.8 &  0.0 &  0.8 \\
       0.5 &  0.5 &  0.0 \end{bmatrix}
  \end{equation*}
    \vfill
    \item If a history is started in state $3$, there is a $20\%$ chance of
      it transitioning to state $1$ and an $80\%$ chance of moving to state
      $2$
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Solving the Heat Equation: Forward Method}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics<1>[width=4in]{direct_1.png}
      \includegraphics<2>[width=4in]{direct_10.png}
      \includegraphics<3>[width=4in]{direct_100.png}
      \includegraphics<4>[width=4in]{direct_1000.png}
    \end{center}
    \caption{
      \only<1>{\textbf{Forward solution.}
        \textit{\sn{2.5}{3} total histories.} }
      \only<2>{\textbf{Forward solution.}
        \textit{\sn{2.5}{4} total histories.} }
      \only<3>{\textbf{Forward solution.}
        \textit{\sn{2.5}{5} total histories.} }
      \only<4>{\textbf{Forward solution.}
        \textit{\sn{2.5}{6} total histories.} }
    }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Synthetic Acceleration}
  \begin{itemize}
  \item Devised by Evans and Mosher in the 2000's as an acceleration
    scheme for radiation diffusion problems (LANL)
    \vfill
  \item Can be abstracted as a general linear solver with Monte Carlo as a
    preconditioner, $M_{MC}$
    \vfill
  \item Combine with Richardson iteration as a ``smoother'' in between
    Monte Carlo steps:
    \begin{align*}
      \mathbf{r}^k &= \mathbf{b} - \mathbf{Ax}^k \\
      \mathbf{x}^{k+1/2} &= \mathbf{x}^k + \mathbf{r}^k \\
      \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} \\
      \mathbf{\delta} &= \mathbf{M_{MC}} \mathbf{r}^{k+1/2} \\
      \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta
    \end{align*}
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}

  \center Domain Decomposition and Replication

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Domain Decomposed Monte Carlo}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Each parallel process owns a piece of the domain (linear
        system)
        \bigskip
      \item Random walks must be transported between adjacent domains
        through parallel communication
        \bigskip
      \item Domain decomposition determined by the input system
        \bigskip
      \item Load balancing straightforward for the forward method
      \end{itemize}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.75}{ \input{ddnu_example.pdftex_t} }
        \end{center}
        \caption{\small Domain decomposition example illustrating
          how domain-to-domain transport creates communication costs.}
      \end{figure}
    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Asynchronous Monte Carlo Transport Kernel}

  \begin{columns}

    \begin{column}{0.5\textwidth}

      \small
      \begin{itemize}
      \item General extension of the Milagro algorithm (LANL)
      \item Asynchronous nearest neighbor communication of histories
      \item System-tunable communciation parameters of buffer size and check
        frequency (performance impact)
      \item Need an asynchronous strategy for exiting the transport loop
        without a collective
      \end{itemize}

      \vspace{-0.25in}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.45}{ \input{domain_to_domain.pdftex_t} }
        \end{center}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \vspace{-0.3in}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.85}{ \input{transport_loop.pdftex_t} }
        \end{center}
      \end{figure}

    \end{column}

  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Stopping the Asynchronous Kernel}

  \vspace{-0.25in}

  \begin{figure}[htpb!]
    \begin{center}
      \scalebox{0.4}{ \input{master_comm_tree.pdftex_t} }
    \end{center}
    \caption{Master/Slave}
  \end{figure}

  \begin{columns}

    \begin{column}{0.5\textwidth}
      \vspace{-0.75in}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.4}{ \input{binary_comm_tree.pdftex_t} }
        \end{center}
        \caption{Binary Tree}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \vspace{-0.5in}

      \begin{figure}[t!]
        \begin{center}
          \includegraphics[width=0.99\textwidth]{titan_weak_bvsm.pdf}
        \end{center}
        \caption{Weak scaling absolute efficiency}
        \label{fig:titan_weak_bvsm}
      \end{figure}

    \end{column}

  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Replication}

  Different batches of Monte Carlo samplescan be combined in summation via
  superposition if they have different random number streams. For two
  different batches:

  \[
  \mathbf{M_{MC}} \mathbf{x} =
  \frac{1}{2}(\mathbf{M_1}+\mathbf{M_2})\mathbf{x} 
  \]

  Consider each of these batches independent \textit{subsets} of a Monte Carlo
  operator where now the operator can be formed as a general additive
  decomposition of $N_S$ subsets:

  \[
  \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
  \]

  We replicate the linear problem and form each subset on a different group of
  parallel processes. Applying the subsets to a vector requires an AllReduce
  to form the sum. Each subset is domain decomposed.

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}

  \center Scaling Studies

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Parallel Test Application -- Nuclear Reactor Analysis}
  \small
  The simplified $P_N$ ($SP_N$) equations are an approximation
  to the Boltzmann neutron transport equation used to simulate nuclear
  reactors

  \vspace*{-0.1in}

  \begin{multline}
    \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
    \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
    \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
    \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
    q(\vec{r},\hat{\Omega},E)
  \end{multline}

  \begin{multline}
    -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
      \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
      \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
      \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
      \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
    0,2,4,\cdots,N
  \end{multline}

  \begin{equation}
  \hspace*{-0.75in}
    -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
    \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
    \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4
  \end{equation}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{$SP_N$ Assembly Problem}
\small
  \begin{columns}
    \begin{column}{0.4\textwidth}
      Test problem -- $3 \times 3$ array of fuel assemblies with
      control rod in center location (Profugus)
      \begin{itemize}
      \item 23 energy groups, 2 angular moments, 25M degrees of freedom
      \item 1,000 computational cores via domain decomposition
      \item We are interested in solving generalized eigenvalue problem,
        for this study we use Arnoldi as the eigensolver and compare
        different methods for solving linear systems
      \end{itemize}
    \end{column}
    \begin{column}{0.6\textwidth}
      \vspace*{-0.5in}
      \begin{figure}
        \centering
        \includegraphics[width=3in]{prob4}
      \end{figure}
    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Parallel $SP_N$ Results}
\begin{table}
\centering
\begin{tabular}{cccc}
\toprule
%Method & Total GMRES Iterations & Setup Time (s) & Solve Time (s) \\
\multirow{2}{*}{\bfseries Method} &
\bfseries Total Linear & \bfseries Setup & \bfseries Solve \\
& \bfseries Iteration & \bfseries Time (s) & \bfseries Time (s) \\
\midrule
GMRES-ILUT      & 1675 & 0.7  & 18.4 \\
GMRES-AMG       & 626  & 0.7  & 46.0 \\
GMRES-MGE       & 498  & 1.5  & 33.7 \\
Richardson-AINV & 5208 & 20.6 & 52.0 \\
MCSA-AINV       & 1268 & 25.5 & 46.6 \\
\bottomrule
\end{tabular}
\end{table}
\begin{itemize}
  \item ILUT preconditioning is winner here, but known to have issues
    with parallel scaling on large core counts
  \item Solve times for MCSA are competitive, but setup times are very large
    due to construction of sparse approximate inverse
\end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}

  \center Matrix-Free and Stochastic Approximate Inverse Algorithms

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Alternative Parallelism -- Additive Schwarz}
  \begin{itemize}
    \item Instead of performing Monte Carlo on full problem, another
      possibility is to apply Monte Carlo as an additive Schwarz approach
    \vfill
    \item Decompose problem into (possibly overlapping) domains
    \vfill
    \item Perform Monte Carlo on individual subdomains
      \begin{itemize}
        \item No communication costs in Monte Carlo problem!
      \end{itemize}
    \vfill
    \item With domain decomposed Monte Carlo, iteration counts are effectively
      independent of the number of processors
    \vfill
    \item In an additive Schwarz approach, the preconditioner will become
      less effective as processor counts grow -- algorithmic scalability
      may be an issue
      \vfill
    \item \textbf{Replication for resiliency and performance}
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{More Parallelism -- Threading}
  \begin{itemize}
    \item Within a Monte Carlo solve, every history is independent of other
      histories -- great potential for highly concurrent hardware
      (GPU, Xeon Phi)
    \vfill
    \item Polynomial formulation enables a priori determination of
      operation counts per thread
      \vfill
    \item Memory locality an issue due to random access via random
      walks (block formulation?)
      \vfill
    \item Early experiments using the Trilinos Kokkos library show promising
      performance for multi-core CPUs
    \vfill
    \item Team members recently took part in OLCF ``Hackathon'' in
      late October to begin implementing computation kernels in
      OpenACC to allow for GPU capability on Titan with early results
      indicating \textbf{1.3-9.4x} speedup for MCSA (largely dependent
      on random number generation)
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Conclusions}
  \begin{itemize}
  \item Monte Carlo methods offer great potential for both resilient and
    highly parallel solvers
    \vfill
  \item For certain classes of problems, Monte Carlo methods can be
    competitive with leading modern solvers
    \vfill
    \item Extending methods to broader problem areas is significant challenge
      and an attractive area for continued research
      \vfill
    \item Performance modeling and resiliency simulations this FY
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\end{document}
%%---------------------------------------------------------------------------%%

%%---------------------------------------------------------------------------%%
%% end of pres.tex
%%---------------------------------------------------------------------------%%

### 2015-03-10 doc/siam_cse_2015/siam_cse_2015.tex
adding siam cse 2015 talk
updating first half of cse talk

@@
   \item Can be abstracted as a general linear solver with Monte Carlo as a
-    preconditioner, $M_{MC}$
+    preconditioner, $\mathbf{M_{MC}}$
     \vfill
@@
         \bigskip
-      \item Load balancing straightforward for the forward method
+      \item Load balancing not yet addressed
       \end{itemize}
@@
     \begin{column}{0.5\textwidth}
-
+      \vspace{-0.1in}
       \small
@@
       \item Need an asynchronous strategy for exiting the transport loop
-        without a collective
+        without a collective (running sum)
       \end{itemize}
 
-      \vspace{-0.25in}
+      \vspace{-0.1in}
       \begin{figure}[htpb!]
@@
 
-  Different batches of Monte Carlo samplescan be combined in summation via
+  Different batches of Monte Carlo samples can be combined in summation via
   superposition if they have different random number streams. For two

### 2015-04-13 doc/ned2015/slattery_wilson_evans.tex
working on reviewer comments

@@
     method for solving linear systems is analyzed using the spectral
-    properties of the linear operator. Relationships for the average
-    length of the adjoint random walks, a measure of convergence speed and
-    serial performance, are made with respect to the eigenvalues of the
-    linear operator. In addition, relationships for the effective optical
-    thickness of a domain in the decomposition are presented based on the
-    spectral analysis and diffusion theory. Using the effective optical
-    thickness, the Wigner rational approximation and the mean chord
-    approximation are applied to estimate the leakage fraction of
-    stochastic histories from a domain in the decomposition as a measure
-    of parallel performance and potential communication costs. The
-    one-speed, two-dimensional neutron diffusion equation is used as a
-    model problem to test the models for symmetric operators. In general,
-    the derived approximations show good agreement with measured
-    computational results.
+    properties of the linear operator. Relationships for the average length of
+    the adjoint random walks, a measure of convergence speed and serial
+    performance, are made with respect to the eigenvalues of the linear
+    operator. In addition, relationships for the effective optical thickness
+    of a domain in the decomposition are presented based on the spectral
+    analysis and diffusion theory. Using the effective optical thickness, the
+    Wigner rational approximation and the mean chord approximation are applied
+    to estimate the leakage fraction of random walks from a domain in the
+    decomposition as a measure of parallel performance and potential
+    communication costs. The one-speed, two-dimensional neutron diffusion
+    equation is used as a model problem in numerical experiments to test the
+    models for symmetric operators with spectral qualities similar to light
+    water reactor problems. In general, the derived approximations show good
+    agreement with effective optical thicknesses and leakage fractions
+    computed by numerical experiments.
 \end{abstract}
@@
 (MCSA) methods that, for certain classes of problems, are competitive with
-conventional methods including Krylov methods and are applicable to a wide
-variety of linear systems \cite{evans_monte_2009,evans_monte_2014}. To advance
-the applicability of these new MCSA methods, we therefore aim to apply the
-underlying Monte Carlo methods to large-scale problems in parallel computing
-environments.
+conventional methods methods and are applicable to a wide variety of linear
+systems \cite{evans_monte_2009,evans_monte_2014}. To advance the applicability
+of these new MCSA methods, we therefore aim to apply the underlying Monte
+Carlo methods to large-scale problems in parallel computing environments.
 
 To date, parallel Neumann-Ulam methods have been limited to full domain
-replication with parallelism exploited through individual histories
+replication with parallelism exploited through individual random walks
 \cite{alexandrov_efficient_1998}. In reactor physics Monte Carlo applications,
@@
 \cite{brunner_comparison_2006,brunner_efficient_2009,siegel_analysis_2012}. To
-accomplish this, we recognize from the literature that stochastic histories
-must be transported from domain to domain as the simulation progresses and
-they transition to states that are not in the local domain. Because we have
-chosen a domain decomposition strategy in a parallel environment, this means
-that communication of these histories must occur between compute nodes owning
+accomplish this, we recognize from the literature that random walks must be
+transported from domain to domain as the simulation progresses and they
+transition to states that are not in the local domain. Because we have chosen
+a domain decomposition strategy in a parallel environment, this means that
+communication of these random walks must occur between compute nodes owning
 neighboring pieces of the global domain. We wish to characterize this
@@
 because these nearest-neighbor communication sequences, specifically, have
-poor algorithmic strong scaling \cite{gropp_high-performance_2001}.
+poor algorithmic strong scaling due to the simultaneous reduction of
+on-process work and increase in communication as processor counts grow and
+local problem sizes shrink \cite{gropp_high-performance_2001}.
 
@@
 parameters. Using the eigenvalue spectra, we then build relationships to
-characterize the transport of stochastic histories in a decomposed domain and
-the fraction of histories that leak from a domain and will therefore have to
+characterize the transport of random walks in a decomposed domain and
+the fraction of random walks that leak from a domain and will therefore have to
 be communicated. Finally, we compare these analytic results to numerical
@@
 \end{equation}
-Choosing the adjoint method Monte Carlo method to invert the linear
-operator, we define the linear system adjoint to
-Eq~(\ref{eq:linear_problem}):
+Choosing the adjoint Monte Carlo method to invert the linear operator, we
+define the linear system adjoint to Eq~(\ref{eq:linear_problem}):
 \begin{equation}
@@
 \end{equation}
-where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively and $\ve{A}^T$ is the adjoint operator. We can derive a
-Monte Carlo estimator from the adjoint method that will also give the
-solution vector, $\ve{x}$. We first rearrange
-Eq~(\ref{eq:adjoint_linear_problem}) as:
+where $\ve{y}$ and $\ve{d}$ are the yet to be determined adjoint solution and
+source respectively and $\ve{A}^T$ is the adjoint operator. We can derive a
+Monte Carlo estimator from the adjoint method that will also give the solution
+vector, $\ve{x}$. We first rearrange Eq~(\ref{eq:adjoint_linear_problem}) as:
 \begin{equation}
@@
 \end{equation}
-such that for each step, $m$, in the random walk, we expect to select
-a new state, $j$, from the current state, $i_m$, by sampling the
-$i_m^{th}$ row of the probability matrix as a discrete distribution,
-and updating the weight, $W_m$, accordingly.  Using our result from
-Eq~(\ref{eq:inner_product_constraint}) generated by applying the
-adjoint constraints, the Monte Carlo estimator for the solution in
-state $i$ for a particular random walk permutation, $\nu$, is then:
+such that for each step, $m$, in the random walk, we expect to select a new
+state, $j$, from the current state, $i_m$, by sampling the $i_m^{th}$ row of
+the probability matrix as a discrete distribution, and updating the weight of
+the random walk at the $m^{th}$ step, $W_m$, accordingly.  Using our result
+from Eq~(\ref{eq:inner_product_constraint}) generated by applying the adjoint
+constraints, the Monte Carlo estimator for the solution in state $i$ for a
+particular random walk permutation, $\nu$, is then:
 \begin{equation}
@@
 \end{equation}
-where the Kronecker delta indicates that the tally contributes only in
-the current state and $W_m$ is the history weight on the $m^{th}$
+where the Kronecker delta indicates that the tally contributes only in the
+current state and $W_m$ is the random walk weight on the $m^{th}$
 step. Finally, the expectation value using all permutations is:
@@
 
-We also desire a criteria for random walk termination for problems
-where only an approximate solution is necessary. For the adjoint
-method, we utilize a relative weight cutoff parameter:
+We also desire a criteria for random walk termination for problems where only
+an approximate solution is necessary. We utilize a relative weight cutoff
+parameter:
 \begin{equation}
-  W_f = W_c b_{i_0}\:,
+  W_f = W_c W_0:\,
   \label{eq:relative_weight_cutoff}
 \end{equation}
-where $W_c$ is the \textit{weight cutoff}. The adjoint random walk
-will then be terminated after $m$ steps if $W_m < W_f$ as tally
-contributions become increasingly small, effectively establishing a
-measure for the degree of approximation of the solution.
+where $W_c$ is the \textit{weight cutoff} and $W_0$ is the initial weight of
+the random walk. The adjoint random walk will then be terminated after $m$
+steps if $W_m < W_f$ as tally contributions become increasingly small,
+effectively establishing a measure for the degree of approximation of the
+solution.
 
@@
 
-We choose a finite difference scheme on a square Cartesian grid to
-discretize the problem. For the Laplacian, we choose the 9-point
-stencil over a grid of size $h$ \cite{leveque_finite_2007}:
+We choose a finite difference scheme on a square Cartesian grid to discretize
+the problem. We choose the 9-point stencil over a grid of size $h$
+\cite{leveque_finite_2007} for the Laplacian:
 \begin{multline}
@@
 \end{equation}
-where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
-vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
+where $\ve{D}$ is the diffusion operator resulting from the finite difference
+discretization, $\ve{s}$ is the source in vector form and $\boldsymbol{\phi}$
+is the vector of unknown fluxes.
+
+%%---------------------------------------------------------------------------%%
+\subsection{Domain Decomposition of the Model Problem}
+\label{subsec:model_problem_decomp}
+To consider domain-to-domain communication in the parallel Monte Carlo
+algorithm, we must consider the required data with the correct parallel
+decomposition. In the adjoint Neumann-Ulam method, the domain decomposition
+consists of all states in the system that are local, and the probabilities and
+weights for all state transitions possible in the local domain. Given an input
+parallel decomposition for the linear operator, by definition in
+Eq~(\ref{eq:adjoint_probability}) probabilities and weights will have the same
+parallel decomposition and underlying sparse graph as the input matrix.
+
+Compared to the serial construction of the probabilities and weights, data for
+all states that are required to be on-process must be collected in parallel so
+that all possible random walks that leave the local domain may be correctly
+communicated. As an example of walking through the sparse graph of the matrix
+and parallel communication, Figure~\ref{fig:diffusion_graph} gives the
+adjacency graph of a single state in the neutron diffusion matrix for the
+model problem presented in Eq~(\ref{eq:fd_system}). The discretization stencil
+of the diffusion problem dictates the structure of this graph and the states
+to which a random walk may transition. Consider a random walk starting at node
+$(i,j)$ in the mesh (the center node). If we play the Monte Carlo game to move
+to a new state, then we may move to any of the nodes in this graph, including
+the node at which we started with a probability proportional to the
+coefficients of each node in Eq~(\ref{eq:fd_system}). If any grid point
+adjacent to node $(i,j)$ is on another processor, the random walk must be
+communicated to that processor to continue the walk.
+
+\begin{figure}[ht!]
+  \begin{center}
+    \scalebox{1.25}{ \input{stencil_graph.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Neutron diffusion equation stencil adjacency graph.}
+    \textit{The structure of the graph comes from the discretization of the
+      Laplacian operator describing the diffusion physics in
+      Eq~(\ref{eq:nine_point_stencil}). A random walk at mesh point $(i,j)$
+      may transition to all adjacent mesh points including itself. If a
+      random walk transitions to a node that is not owned by the local
+      processor, it must be communicated to that processor to continue the
+      walk.}}
+  \label{fig:diffusion_graph}
+\end{figure}
 
@@
 
-For the model problem, we first compute the eigenvalues for the
-diffusion operator $\ve{D}$ by applying the operator to the
-eigenfunctions and noting that $x=ih$ and $y=jh$:
+We first compute the eigenvalues for the diffusion operator $\ve{D}$ by
+applying the operator to the eigenfunctions and noting that $x=ih$ and $y=jh$:
 \begin{multline}
@@
 
-As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
-preconditioning is sufficient to reduce the spectral radius of the
-iteration matrix below unity and therefore ensure convergence of the
-Neumann series. The preconditioner in this case is then $\ve{M} =
-diag(\ve{D})$ such that we are solving the following linear system:
+As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi preconditioning
+is sufficient to reduce the spectral radius of the iteration matrix below
+unity and therefore ensure convergence of the Neumann series. The
+preconditioner in this case is then $\ve{M} = diag(\ve{D})$ such that we are
+solving the following linear system:
 \begin{equation}
@@
 \end{equation}
-The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
-operator with each row scaled by the diagonal component. As we have
-defined a homogeneous domain, the scaling factor, $\alpha$, is the
-same for all rows in the operator and defined as the $\phi_{i,j}$
-coefficient from Eq~(\ref{eq:fd_system}):
+The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion operator
+with each row scaled by the diagonal component. As we have defined a
+homogeneous domain, the scaling factor, $\alpha$, is the same for all rows in
+the operator and defined as the $\phi_{i,j}$ coefficient from
+Eq~(\ref{eq:fd_system}):
 \begin{equation}
@@
 
-The spectral radius of the iteration matrix is obtained by seeking its
-largest eigenvalue. As with the diffusion operator, we can use the
-same analysis techniques to find the eigenvalues for the iteration
-matrix. We use a few simplifications by noting that if the Jacobi
-preconditioned iteration matrix is $\ve{H} = \ve{I} -
-\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
-iteration matrix to be zero such that we have the following stencil:
+The spectral radius of the iteration matrix is obtained by seeking its largest
+eigenvalue. As with the diffusion operator, we can use the same analysis
+techniques to find the eigenvalues for the iteration matrix. We use a few
+simplifications by noting that if the Jacobi preconditioned iteration matrix
+is $\ve{H} = \ve{I} - \ve{M}^{-1}\ve{D}$, the we except all terms on the
+diagonal of the iteration matrix to be zero such that we have the following
+stencil:
 \begin{equation}
@@
 \end{equation}
-Inserting the eigenfunctions defined by
-Eq~(\ref{eq:eigenfunction_form}) we get:
+Inserting the eigenfunctions defined by Eq~(\ref{eq:eigenfunction_form}) we
+get:
 \begin{multline}
@@
 \end{equation}
-giving the eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. We find that the maximum eigenvalue exists when $p=q=0$,
-giving the following for the spectral radius of the Jacobi
-preconditioned iteration matrix:
+giving the eigenvalue spectrum of the Jacobi preconditioned iteration
+matrix. We find that the maximum eigenvalue exists when $p=q=0$, giving the
+following spectral radius of the Jacobi preconditioned iteration matrix:
 \begin{equation}
@@
 In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
-applications of the iteration matrix, are approximated by a random
-walk of average length $k$ to yield the summation in
+applications of the iteration matrix, are approximated by a random walk of
+average length $k$ to yield the summation in
 Eq~(\ref{eq:adjoint_neumann_solution})
-\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
-length, or the number of transitions before the termination of a
-history (either by the weight cutoff, absorption, or exiting the
-global domain) is therefore approximately the number of stationary
-iterations required to converge to the specified tolerance. In the
-case of the adjoint Neumann-Ulam method, no such tolerance exists,
-however, we have specified a weight cutoff, $W_c$, that determines
-when low-weight histories will be prematurely terminated as their
-contributions are deemed minute. After $k$ iterations, a stationary
-method is terminated as the error has reached some fraction,
-$\epsilon$, of the initial error:
+\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk length, or the
+number of transitions before the termination of a random walk (either by the
+weight cutoff, absorption, or exiting the global domain) is therefore
+approximately the number of stationary iterations required to converge to the
+specified tolerance. In the case of the adjoint Neumann-Ulam method, no such
+tolerance exists, however, we have specified a weight cutoff, $W_c$, that
+determines when low-weight random walks will be prematurely terminated as
+their contributions are deemed minute. After $k$ iterations, a stationary
+method is terminated as the error has reached some fraction, $\epsilon$, of
+the initial error:
 \begin{equation}
@@
 \end{equation}
-Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
-equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
-Neumann-Ulam method, if we take this fraction to be the weight cutoff,
-a measure of how accurately the contributions of a particular history
-to the solution are tallied, we then have the following relationship
-for $k$:
+Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is equivalent
+to $\epsilon = \rho(\ve{H})^k$. In adjoint method, if we take this fraction to
+be the weight cutoff, a measure of how accurately the contributions of a
+particular random walk to the solution are tallied, we then have the following
+relationship for $k$:
 \begin{equation}
@@
 
-In a domain decomposed situation, not all histories will remain within
-the domain they started in and must instead be communicated. This
-communication, expected to be expensive, was analyzed by Siegel and
-colleagues for idealized, load balanced situations for full nuclear
-reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
-quantify the number of particles that leak out of the local domain
-they define a leakage fraction, $\Lambda$, as:
+In a domain decomposed situation, not all random walk will remain within the
+domain they started in and must instead be communicated. This communication,
+expected to be expensive, was analyzed by Siegel and colleagues for idealized,
+load balanced situations for full nuclear reactor core Monte Carlo simulations
+\cite{siegel_analysis_2012}.  To quantify the number of particles that leak
+out of the local domain they define a leakage fraction, $\Lambda$, as:
 \begin{equation}
@@
 diffusion theory to estimate the optical thickness of a domain in the
-decomposition and the corresponding leakage fraction in terms of
-properties of the linear operator and the discretization. To begin we
-must first calculate the mean distance a Monte Carlo history will move
-in the grid by computing the mean squared distance of its movement
-along the chord of length $l$ defined across the domain. After a
-single transition a history will have moved a mean squared distance
-of:
+decomposition and the corresponding leakage fraction in terms of properties of
+the linear operator and the discretization. To begin we must first calculate
+the mean distance a Monte Carlo random walk will move in the grid by computing
+the mean squared distance of its movement along the chord of length $l$
+defined across the domain. After a single transition a random walk will have
+moved a mean squared distance of:
 \begin{equation}
@@
 \end{equation}
-where $h$ is the size of the discrete grid elements along the chord
-and $n_s$ is the number of grid elements a history will move on
-average every transition. For our diffusion model problem, $n_s$ would
-equate to the expected number of states in the $i$ (or $j$ as the problem is
-symmetric) direction that a history will move in a single
-transition and is dependent on the stencil used for the
-discretization. After $k$ transitions in the random walk, the history
-will have moved a mean squared distance of:
+where $h$ is the size of the discrete grid elements along the chord and $n_s$
+is the number of grid elements a random walk will move on average every
+transition. For our diffusion model problem, $n_s$ would equate to the
+expected number of states in the $i$ (or $j$ as the problem is symmetric)
+direction that a random walk will move in a single transition and is dependent on
+the stencil used for the discretization. After $k$ transitions in the random
+walk, the random walk will have moved a mean squared distance of:
 \begin{equation}
@@
 If our chord is of length $l$ and there are $n_i$ grid elements (or
-states to which a history may transition) along that chord, then $h =
+states to which a random walk may transition) along that chord, then $h =
 l / n_i$ giving:
@@
   \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
-Carlo history in the domain. We can readily interpret $\tau$ to be the
+Carlo random walk in the domain. We can readily interpret $\tau$ to be the
 \textit{effective optical thickness} of a domain of length
@@
 
-For optically thin domains, we expect that most histories will be
+For optically thin domains, we expect that most random walks will be
 communicated, while optically thick domains will leak the fraction of
-histories that did not interact within. Using the optical thickness
-defined in Eq~(\ref{eq:optical_thickness_3}), we can then complete the
-leakage approximations by defining the bounds of $\tau \rightarrow 0,
-\Lambda \rightarrow 1$ and $\tau \rightarrow \infty, \Lambda
-\rightarrow \tau^{-1}$.  With these bounds we can then define the
-leakage fraction out of a domain for the adjoint Neumann-Ulam method
-using the Wigner rational approximation:
+random walks that did not interact within. Using the optical thickness defined in
+Eq~(\ref{eq:optical_thickness_3}), we can then complete the leakage
+approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
+\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow \tau^{-1}$.
+With these bounds we then define the leakage fraction out of a domain using
+the Wigner rational approximation:
 \begin{equation}
@@
 discretization stencil, and the weight cutoff selected to terminate
-low weight histories.
+low weight random walks.
 
@@
 
-With the spectral data in hand, we can go about setting up an
-experiment to measure the length of the random walks generated by the
-adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
-100$ square grid with $h=0.1$ and the absorption cross varied from 0
-to 100 while the scattering cross section was fixed at unity. Three
-weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
-used with 10,000 histories generated by a point source of strength 1
-in the center of the domain. For each of the histories, the number of
-transitions made was tallied to provide an effective value of $k$ for
-each history. This value was then averaged over all histories to get a
-measured value of $k$ for the particular operator. On the left,
-Figure~\ref{fig:measured_length} presents these measurements as well
-as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
-function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
-the right, Figure~\ref{fig:measured_length} gives the relative error
-between the predicted and observed results. We note good qualitative
-agreement between the measured and analytic results. However, we
-observe a larger relative error for both long and short random walks.
+With the spectral data in hand, we can go about setting up an experiment to
+measure the length of the random walks generated by the adjoint Neumann-Ulam
+solver. To do this, we again use a $100 \times 100$ square grid with $h=0.1$
+and the absorption cross varied from 0 to 100 while the scattering cross
+section was fixed at unity. Three weight cutoff values of \sn{1}{-2},
+\sn{1}{-4}, and \sn{1}{-8} were used with 10,000 random walks generated by a
+point source of strength 1 in the center of the domain. For each of the
+random walks, the number of transitions made was tallied to provide an effective
+value of $k$ for each random walk. This value was then averaged over all random walks
+to get a measured value of $k$ for the particular operator. On the left,
+Figure~\ref{fig:measured_length} presents these measurements as well as the
+analytic result computed by Eq~(\ref{eq:analytic_k}) as a function of the
+iteration matrix spectral radius, $\rho(\ve{H})$. On the right,
+Figure~\ref{fig:measured_length} gives the relative error between the
+predicted and observed results. We note good qualitative agreement between the
+measured and analytic results. However, we observe a larger relative error for
+both long and short random walks.
 \begin{figure}[ht!]
@@
 
-Finally, we seek to measure the leakage from a domain in a domain
-decomposed Monte Carlo calculation and assess the quality of our
-analytic relation for the optical thickness of domain and the
-associated leakage approximations. For this experiment, a square grid
-with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
-direction with measurements occurring in the central domain without
-boundary grid points. For the cross sections, the absorption cross
-section was varied from 1 to 100 while the scattering cross section
-was set to zero to create a purely absorbing environment with weight
-cutoff of \sn{1}{-4}. The optical thickness of these domains will vary
-as a function of the absorption cross section if the other parameters
-are fixed. To compute the optical thickness, along with the spectral
-radius as given by Eq~(\ref{eq:iteration_radius}), we also need the
-parameters $n_i$ and $n_s$ which respectively describe the typical
-domain length and the average number of states moved along that
-typical length per history transition. For our grid above, the domains
-are varied in size with $50 \times 50$, $100 \times 100$, and $200
-\times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$ grid
-points or states along the typical length of the domain
+Finally, we seek to measure the leakage from a domain in a domain decomposed
+Monte Carlo calculation and assess the quality of our analytic relation for
+the optical thickness of domain and the associated leakage approximations. For
+this experiment, a square grid with $h=0.1$ was decomposed into 9 square
+domains, 3 in each cardinal direction with measurements occurring in the
+central domain without boundary grid points. For cross sections, the
+absorption cross section was varied from 1 to 100 while the scattering cross
+section was set to zero to create a purely absorbing environment with weight
+cutoff of \sn{1}{-4}. The optical thickness of these domains will vary as a
+function of the absorption cross section if the other parameters are fixed. To
+compute the optical thickness, along with the spectral radius as given by
+Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and $n_s$
+which respectively describe the typical domain length and the average number
+of states moved along that typical length per random walk transition. For our
+grid above, the domains are varied in size with $50 \times 50$, $100 \times
+100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$
+grid points or states along the typical length of the domain
 respectively. Looking at the Laplacian stencil in
-Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
-will only move a single state in either the $i$ or $j$ directions due
-to the symmetry of the problem. Furthermore, if we choose the $i$
-direction, not all states we will transition to will move the history
-in that direction. Therefore, we look to the definition of the
-iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
-of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
-to estimate the $n_s$ parameter. For a particular transition starting
-at state $(i,j)$, 6 of the 8 possible new states in the stencil move
-the history in $i$ direction with relative coefficients of 4 for
-moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
-i,\pm j)$. These coefficients dictate the frequency those states are
-visited relative to the others. For those 6 states we can visit along
-the typical length, their sum is 12 out of the total 20 for the
+Eq~(\ref{eq:nine_point_stencil}), we see that all random walk transitions will
+only move a single state in either the $i$ or $j$ directions due to the
+symmetry of the problem. Furthermore, if we choose the $i$ direction, not all
+states we will transition to will move the random walk in that
+direction. Therefore, we look to the definition of the iteration matrix in
+Eq~(\ref{eq:iteration_stencil}) and the definition of the adjoint probability
+matrix in Eq~(\ref{eq:adjoint_probability}) to estimate the $n_s$
+parameter. For a particular transition starting at state $(i,j)$, 6 of the 8
+possible new states in the stencil move the random walk in $i$ direction with
+relative coefficients of 4 for moving in the $(\pm i,0)$ direction and of 1
+for moving in the $(\pm i,\pm j)$. These coefficients dictate the frequency
+those states are visited relative to the others. For those 6 states we can
+visit along the typical length, their sum is 12 out of the total 20 for the
 coefficients for all possible states with their ratio giving $n_s =
@@
 
-To compute the leakage fraction numerically, \sn{3}{5} histories were
-sampled from a uniform source of strength unity over the global
-domain. At the start of a stage of histories, the number of histories
-starting in the center domain was computed and as the stage
-progressed, the number of histories that exited that domain was
-tallied with the ratio of the two numbers providing a numerical
-measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
-gives the domain leakage measurements for the domain in the center of
-the global grid as well as the analytic result computed by
-Eqs~(\ref{eq:wigner_domain_leakage}) and
-(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
-matrix spectral radius.
+To compute the leakage fraction numerically, \sn{3}{5} random walks were
+sampled from a uniform source of strength unity over the global domain. At the
+start of a stage of random walks, the number of random walks starting in the
+center domain was computed and as the stage progressed, the number of random
+walks that exited that domain was tallied with the ratio of the two numbers
+providing a numerical measure of the leakage
+fraction. Figure~\ref{fig:measured_leakage} gives the domain leakage
+measurements for the domain in the center of the global grid as well as the
+analytic result computed by Eqs~(\ref{eq:wigner_domain_leakage}) and
+(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration matrix
+spectral radius.
 \begin{figure}[ht!]
@@
 \end{figure}
-From these error results, the mean chord approximation is shown to
-have a lower error for ill-conditioned systems as compared to the
-Wigner approximation while the Wigner approximation produces less
-error for more well-conditioned systems. We also note that for the
-optically thick domains, the error is likely corresponded to that
-observed in Figure~\ref{fig:measured_length} for the $k$ parameter
-while the large relative error in $k$ for optically thin domains does
-not affect the approximation significantly. In general, the mean chord
-approximation is a better choice to estimate the leakage fraction in a
-domain from the adjoint Neumann-Ulam method and except for a single
-data point with $n_i=50$, the mean chord approximation yielded leakage
-fractions within 0.05 of the measured results. As the domain becomes
-more optically thick (with both increasing $n_i$ and decreasing
-$\rho(\ve{H})$), the approximations are more accurate.
+From these error results, the mean chord approximation is shown to have a
+lower error for ill-conditioned systems as compared to the Wigner
+approximation while the Wigner approximation produces less error for more
+well-conditioned systems. We also note that for optically thick domains, the
+error is likely corresponded to that observed in
+Figure~\ref{fig:measured_length} for the $k$ parameter while the large
+relative error in $k$ for optically thin domains does not affect the
+approximation significantly. In general, the mean chord approximation is a
+better choice to estimate the leakage fraction in a domain from the adjoint
+Neumann-Ulam method and except for a single data point with $n_i=50$, the mean
+chord approximation yielded leakage fractions within 0.05 of the measured
+results. As the domain becomes more optically thick (with both increasing
+$n_i$ and decreasing $\rho(\ve{H})$), the approximations are more accurate.
 

### 2015-04-13 doc/ned2015/response_to_reviewer.txt
updating ned paper for reviewer comments

Response to reviewer's comments:

------------------------------------------------------------------------------
1. Since the underlying analysis seems to be graphical, they should use as
many graphs to explain what is happening.  Following words and moving
information is not so easy for a reader.

We agree that without visual aids the discussion is complex and is very much
based on the underlying graph of the matrix. We have added a figure (Figure 1)
in section 3.1 and discussion the random walk in terms of the sparse graph
underneath the matrices that should help clarify the explanation, particularly
for the n_s and n_i parameters in our diffusion analysis. In addition, we have
expanded our description of the algorithm in section 2 and have provided more
definitions for clarity.

------------------------------------------------------------------------------
2. Abstract: The term "measured computational results" is not so clear at
first.  The authors may want to find another name for this quantity.

We agree and have clarified that leakage fractions computed via numerical
experiments show good agreement with the equations we have derived.

------------------------------------------------------------------------------
3. Page 2: Last sentence of the second paragraph, it is not clear what is
meant by "specifically, have poor algorithmic strong scaling."

This statement has been clarified to indicate that when an algorithm is strong
scaled, the amount of on-process work decreases while the amount of
communication increases due to increasing processor count and decreasing local
problem size.

------------------------------------------------------------------------------
4. Top of Page 3: "adjoint method Monte Carlo method", is it necessary to use
'method' twice?  Seems a bit awkward.

This is a typo and it has been corrected.

------------------------------------------------------------------------------
5. Should indicate after Eq(2) .. yet to be determined adjoint source…

We agree and have indicated that these quantities will be determined in the
derivation.

------------------------------------------------------------------------------
4.  I assume Wm is a weight vector.

Wm is a scalar quantity equal to the weight of a random walk on its m^th
step. We have clarified the definition of Wm by indicating this.

------------------------------------------------------------------------------
6. D is the diffusion FD operator.

We have clarified that D is the diffusion operator resulting from our choice
of finite difference scheme.

------------------------------------------------------------------------------
7. Why not diagonalize D when it first appears?

We agree that we could diagonalize D initially. However, for more difficult
problems diagonal preconditioning may not be sufficient to satisfy the
convergence criteria and therefore it is instructive to separate it in the
analysis to show how a preconditioner would affect both iterative performance
as well as indicate how our analysis could be extended for more complicated
preconditioners. In addition, for our model problem the preconditioner simply
results in a scaling of the operator spectrum computed in section 4.1 and
therefore lets us analyze the operator independently from the preconditioner.

------------------------------------------------------------------------------
8. I believe D can be written as a block tridiagonal matrix for the stencil
used.  Is this the case?  If not see Burlish and Stoerer.  This would simplify
the analysis greatly.

Although we agree other matrix orderings could have been used, the results of
the paper are fundamentally based on the local stencil and independent the
overall global structure of the matrix. A block tridiagonal ordering could
reduce the bandwidth of the matrix and improve computational efficiency for a
matrix-vector multiply. However, the Monte Carlo algorithm would perform
identically in both cases given the same parallel decomposition of the
problem. The addition of Figure 1 in response to the first comment should also
help with this part of the analysis.

For example, given a banded ordering like that used in this work or the
suggested block tridiagonal ordering, the probability distribution functions
would be of the same size and have the same nearest-neighbor connections
because both use the same local stencil. Therefore, both cases will yield
random walks of the same time and space complexity. In addition, if both cases
had the same domain decomposition, random walks would be communicated from
domain to domain in an identical matter because the same local stencil was
used, therefore also resulting in the same parallel analysis.

------------------------------------------------------------------------------
9. Finally, there are no flux results with estimated errors.  We need results
since, after all, they are the bottom-line.

The purpose of the paper is to provide analytic relationships for the
domain-to-domain communication of random walks, and therefore a measure of
potential bandwidth and latency issues, when the Neumann-Ulam algorithm is
subject to domain decomposition. The domain decomposition itself does not
perturb the results of the stochastic algorithm (e.g. the flux in this
case). Rather, it simply provides a means to perform random walks via the
rules of the Neumann-Ulam game where the data for the random walks (weights
and probablities) has been domain-decomposed.

Variance estimation, however, is a different story and domain decomposition
schemes to date trade inexact variance estimates for scalablity. This is due
to the fact that the order of events must be maintained to acheive a correct
variance estimate and fully asynchronous algorithms do not, in general, permit
this. The literature indicates that this order of operations is less
significant for the correctness of the flux estimate with differences in
computations induced by round-off error in floating point
computations. Furthermore, the accuracy and convergence properties of the
Monte Carlo method have been well documented in the references cited by the
paper.

Therefore, because the purpose of this work is to analyze communication costs
of the algorithm using an operator with spectral properties similar to large
scale neutronics problems and not computing the global flux and variance
estimate, we do not feel it is necessary in this paper to report flux results.

We will, however, in an upcoming paper present our full domain decomposition
algorithm for parallelizing the solver and give a thorough analysis of its
correctness and scalability.

------------------------------------------------------------------------------

### 2015-04-13 doc/ned2015/slattery_wilson_evans.tex
updating ned paper for reviewer comments

@@
     water reactor problems. In general, the derived approximations show good
-    agreement with effective optical thicknesses and leakage fractions
-    computed by numerical experiments.
+    agreement with random walk lengths and leakage fractions computed by the
+    numerical experiments.
   \end{abstract}
@@
 
-An alternative approach to approximate matrix inversion is to employ Monte
-Carlo methods that sample a distribution with an expectation value equivalent
-to that of the inverted operator. Such methods have been in existence for
+An alternative approach to solving linear systems is to employ Monte Carlo
+methods that sample a distribution with an expectation value equivalent to
+that of the inverted linear operator. Such methods have been in existence for
 decades with the earliest reference noted here an enjoyable manuscript
@@
 conventional methods methods and are applicable to a wide variety of linear
-systems \cite{evans_monte_2009,evans_monte_2014}. To advance the applicability
-of these new MCSA methods, we therefore aim to apply the underlying Monte
-Carlo methods to large-scale problems in parallel computing environments.
+systems \cite{evans_monte_2009,evans_monte_2014}. To apply these new MCSA
+methods to engineering problems, our research aims to improve the underlying
+Monte Carlo algorithms to handle large-scale problems in parallel computing
+environments.
 
@@
 \end{equation}
-For this series to converge, we require that the spectral radius of
-$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
-eigenvalues and therefore has the same spectral radius. We expand this
-summation to yield a series of transitions that can be approximated by
-a Monte Carlo random walk sequence:
+For this series to converge, we require that the spectral radius of $\ve{H}$
+must remain less than unity\footnote{$\ve{H}^T$ contains the same eigenvalues
+  as $\ve{H}$ and therefore has the same spectral radius.}. We expand the
+matrix-vector multiply element-wise in this summation to yield a sum of sums:
 \begin{equation}
@@
 \end{equation}
-Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
-require two constraints to close the system. We use
-Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
-second constraint we select:
+We have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we require two
+constraints to close the system. We use Eq~(\ref{eq:adjoint_vector_relation})
+as the first constraint and as a second constraint we select:
 \begin{equation}
@@
 \end{equation}
-where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
-$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
-we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
-constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
-convenient outcome:
+where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the $j^{th}$
+component is the Kronecker delta function, $\delta_{i,j}$. If we apply
+Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint in
+Eq~(\ref{eq:adjoint_vector_relation}) we get the following convenient
+outcome:
 \begin{equation}
@@
 \end{equation}
-meaning that if we compute the inner product of the original source
-and the adjoint solution using a delta function source, we recover one
-component of the original solution.
+meaning that if we compute the inner product of the original source and the
+adjoint solution using a delta function source, we recover one component of
+the original solution.
 
-In order to perform the random walks required to approximate the
-summation in Eq~(\ref{eq:adjoint_neumann_solution}), we build the
-column-scaled transition probability matrix and associated transition
-weights as:
+In the Neumann-Ulam method, a Monte Carlo game is played to construct the
+terms of the Neumann series in Eq~(\ref{eq:adjoint_neumann_solution}). We use
+random walks of length $k$ to build the series of state transitions in the
+summation:
+\begin{equation}
+  \nu = i_0 \rightarrow i_1 \rightarrow i_2 \rightarrow \cdots \rightarrow
+  i_{k-1}\:,
+  \label{eq:permutation}
+\end{equation}
+where $\nu$ is a particular random walk permutation. Starting a random walk
+permutation requires the computation of an initial starting state, $i_0$, and
+initial weight, $W_0$. The initial state, $i_0$, of the random walk is
+determined by sampling the source vector $\ve{b}$ with probabilities:
+\begin{equation}
+  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
+  \label{eq:adjoint_source_probability}
+\end{equation}
+and with a corresponding starting weight of:
+\begin{equation}
+  W_0 = ||\ve{b}||_1 \frac{b_{i_0}}{|b_{i_0}|}\:,
+  \label{eq:adjoint_starting_weight}
+\end{equation}
+which gives the additional useful relation:
+\begin{equation}
+  b_{i_0} = W_0 P_{(i_0=i)}\:.
+  \label{eq:adjoint_source_definition}
+\end{equation}
+After a random walk is started, we construct the remaining terms in
+Eq~(\ref{eq:adjoint_neumann_solution}) using a row-scaled transition probability
+matrix and corresponding transition weights:
 \begin{equation}
@@
 \end{equation}
-such that for each step, $m$, in the random walk, we expect to select a new
-state, $j$, from the current state, $i_m$, by sampling the $i_m^{th}$ row of
-the probability matrix as a discrete distribution, and updating the weight of
-the random walk at the $m^{th}$ step, $W_m$, accordingly.  Using our result
-from Eq~(\ref{eq:inner_product_constraint}) generated by applying the adjoint
-constraints, the Monte Carlo estimator for the solution in state $i$ for a
-particular random walk permutation, $\nu$, is then:
+where each row of the probability matrix, $\ve{P}_i$, gives a probability
+distribution function for moving from row $i$ to a new row, $j$. As a random
+walk is computed by sampling the probability matrix, every stochastic
+transition from a state $i$ to a state $j$ has a corresponding weight,
+$w_{ij}$, that is used to update the current weight of the random walk. At the
+$m^{th}$ step of a random walk, its weight is:
+\begin{equation}
+  W_{m} = W_0 w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
+  \label{eq:direct_permutation_weight}
+\end{equation}
+with $W_0$ given by Eq~(\ref{eq:adjoint_starting_weight}). To stop a random
+walk, we utilize a relative weight cutoff parameter:
+\begin{equation}
+  W_f = W_c W_0:\,
+  \label{eq:relative_weight_cutoff}
+\end{equation}
+where $W_c$ is the \textit{weight cutoff}. The random walk will be terminated
+after $m$ steps if $W_m < W_f$.
+
+Using our result from Eq~(\ref{eq:inner_product_constraint}) the Monte Carlo
+estimator for the solution in state $i$ for a particular random walk
+permutation of length $k$ is:
 \begin{equation}
@@
 where the Kronecker delta indicates that the tally contributes only in the
-current state and $W_m$ is the random walk weight on the $m^{th}$
-step. Finally, the expectation value using all permutations is:
+current state, $i_m$. Finally, the expectation value using all random walk
+permutations is:
 \begin{equation}
@@
 \end{equation}
-which if expanded directly recovers the exact solution:
+which if expanded with Eqs~(\ref{eq:adjoint_probability}) and
+(\ref{eq:adjoint_source_definition}) directly recovers the exact solution:
 \begin{equation}
   \begin{split}
-    E\{X_j\} &=\sum_{m=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_m}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
-    h_{i_{m-1},i_m} \delta_{i_m,j} \\ &= x_{j}\:,
+    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
+    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
+    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:.
   \end{split}
@@
 \end{equation}
-where $b_{i_0}$ will be the sampled source starting weight, therefore
-also providing an unbiased Monte Carlo estimate of the solution.
-
-We also desire a criteria for random walk termination for problems where only
-an approximate solution is necessary. We utilize a relative weight cutoff
-parameter:
-\begin{equation}
-  W_f = W_c W_0:\,
-  \label{eq:relative_weight_cutoff}
-\end{equation}
-where $W_c$ is the \textit{weight cutoff} and $W_0$ is the initial weight of
-the random walk. The adjoint random walk will then be terminated after $m$
-steps if $W_m < W_f$ as tally contributions become increasingly small,
-effectively establishing a measure for the degree of approximation of the
-solution.
 
@@
 We choose a finite difference scheme on a square Cartesian grid to discretize
-the problem. We choose the 9-point stencil over a grid of size $h$
-\cite{leveque_finite_2007} for the Laplacian:
+the problem. We choose the 9-point stencil over a grid of size $h$ for the
+Laplacian \cite{leveque_finite_2007}:
 \begin{multline}
@@
 \label{subsec:model_problem_decomp}
-To consider domain-to-domain communication in the parallel Monte Carlo
+To investigate domain-to-domain communication in a parallel Monte Carlo
 algorithm, we must consider the required data with the correct parallel
@@
 parallel decomposition for the linear operator, by definition in
-Eq~(\ref{eq:adjoint_probability}) probabilities and weights will have the same
-parallel decomposition and underlying sparse graph as the input matrix.
+Eq~(\ref{eq:adjoint_probability}) the probabilities and weights will have the
+same parallel decomposition and underlying sparse graph as the input
+matrix. In the case of the model problem, parallel decomposition of states and
+subsequently the input matrix is determined by a spatial partitioning of the
+grid.
 
@@
 We first compute the eigenvalues for the diffusion operator $\ve{D}$ by
-applying the operator to the eigenfunctions and noting that $x=ih$ and $y=jh$:
+applying the operator to the eigenfunctions and noting that $x=ih$ and $y=jh$
+on a rectilinear Cartesian grid:
 \begin{multline}
@@
 unity and therefore ensure convergence of the Neumann series. The
-preconditioner in this case is then $\ve{M} = diag(\ve{D})$ such that we are
+preconditioner in this case is $\ve{M} = diag(\ve{D})$ such that we are
 solving the following linear system:
@@
 \end{equation}
-Using this coefficient, we then have the following spectrum of
-preconditioned eigenvalues:
+Using this coefficient, we have the following spectrum of preconditioned
+eigenvalues:
 \begin{equation}
@@
 simplifications by noting that if the Jacobi preconditioned iteration matrix
-is $\ve{H} = \ve{I} - \ve{M}^{-1}\ve{D}$, the we except all terms on the
+is $\ve{H} = \ve{I} - \ve{M}^{-1}\ve{D}$, then we expect all terms on the
 diagonal of the iteration matrix to be zero such that we have the following
@@
 giving the eigenvalue spectrum of the Jacobi preconditioned iteration
-matrix. We find that the maximum eigenvalue exists when $p=q=0$, giving the
+matrix. We find that the maximum eigenvalue exists when $p=q=0$ giving the
 following spectral radius of the Jacobi preconditioned iteration matrix:
@@
   index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
-error at the $k^{th}$ iterate, the error after $k$ iterations is then:
+error at the $k^{th}$ iterate, the error after $k$ iterations is:
 \begin{equation}
@@
 \end{equation}
-By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we
-then have:
+By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we have:
 \begin{equation}
@@
 \cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk length, or the
-number of transitions before the termination of a random walk (either by the
-weight cutoff, absorption, or exiting the global domain) is therefore
+number of transitions before the termination of a random walk is therefore
 approximately the number of stationary iterations required to converge to the
 specified tolerance. In the case of the adjoint Neumann-Ulam method, no such
-tolerance exists, however, we have specified a weight cutoff, $W_c$, that
-determines when low-weight random walks will be prematurely terminated as
-their contributions are deemed minute. After $k$ iterations, a stationary
-method is terminated as the error has reached some fraction, $\epsilon$, of
-the initial error:
+tolerance exists. However, we have specified a weight cutoff, $W_c$, that
+determines when low-weight random walks will be terminated as their
+contributions are deemed minute. After $k$ iterations, a stationary method is
+terminated as the error has reached some fraction, $\epsilon$, of the initial
+error:
 \begin{equation}
@@
 Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is equivalent
-to $\epsilon = \rho(\ve{H})^k$. In adjoint method, if we take this fraction to
-be the weight cutoff, a measure of how accurately the contributions of a
-particular random walk to the solution are tallied, we then have the following
-relationship for $k$:
+to $\epsilon = \rho(\ve{H})^k$. In the adjoint method, if we take this
+fraction to be the weight cutoff, a measure of how accurately the
+contributions of a particular random walk to the solution are tallied, we have
+the following relationship for $k$:
 \begin{equation}
@@
 \end{equation}
-This then gives us a means to estimate the length of the random walks
-that will be generated from a particular linear operator based on the
-eigenvalues of its iteration matrix (independent of the linear
-operator splitting chosen) and based on the weight cutoff parameter
-used in the Neumann-Ulam method.
+This gives us a means to estimate the length of the random walks that will be
+generated from a particular linear operator based on the eigenvalues of its
+iteration matrix (independent of the linear operator splitting chosen) and
+based on the weight cutoff parameter used in the Neumann-Ulam method.
 
@@
 
-In a domain decomposed situation, not all random walk will remain within the
+In a domain decomposed situation, not all random walks will remain within the
 domain they started in and must instead be communicated. This communication,
@@
 \begin{equation}
-  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
-          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
+  \Lambda = \frac{\text{average \# of particles leaving local domain}}
+          {\text{total \# of particles starting in local domain}}\:.
           \label{eq:leakage_fraction}
 \end{equation}
-For their studies, Siegel and colleagues assumed that the value of
-$\Lambda$ was dependent on the total cross section of the system via
-the Wigner rational approximation. Outlined more thoroughly by Hwang's
-chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
-rational approximation and the mean chord approximation as a means to
-estimate the leakage fraction.
+For their studies, it was assumed that the value of $\Lambda$ was dependent on
+the total cross section of the system via the Wigner rational
+approximation. Outlined more thoroughly by Hwang's chapter in
+\cite{azmy_nuclear_2010}, we will use both the Wigner rational approximation
+and the mean chord approximation as a means to estimate the leakage fraction.
 
-In the case of domain decomposed linear operator equations, we can use
-diffusion theory to estimate the optical thickness of a domain in the
-decomposition and the corresponding leakage fraction in terms of properties of
-the linear operator and the discretization. To begin we must first calculate
-the mean distance a Monte Carlo random walk will move in the grid by computing
-the mean squared distance of its movement along the chord of length $l$
-defined across the domain. After a single transition a random walk will have
-moved a mean squared distance of:
+In the case of parallel linear systems, we can use diffusion theory to
+estimate the optical thickness of a domain in the decomposition and the
+corresponding leakage fraction in terms of properties of the linear operator
+and the discretization. We first calculate the mean distance a random walk
+will move in the grid by computing the mean squared distance of its movement
+along the chord of length $l$ defined along a cardinal direction of the
+domain. After a single transition a random walk will have moved a mean squared
+distance of:
 \begin{equation}
@@
 is the number of grid elements a random walk will move on average every
-transition. For our diffusion model problem, $n_s$ would equate to the
-expected number of states in the $i$ (or $j$ as the problem is symmetric)
-direction that a random walk will move in a single transition and is dependent on
-the stencil used for the discretization. After $k$ transitions in the random
-walk, the random walk will have moved a mean squared distance of:
+transition in the direction of the chord. For our diffusion model problem,
+$n_s$ would equate to the expected number of states in the $i$ (or $j$ as the
+problem is symmetric) direction that a random walk will move in a single
+transition and is dependent on the stencil given in
+Figure~\ref{fig:diffusion_graph}. After $k$ transitions, the random walk will
+have moved a mean squared distance of:
 \begin{equation}
@@
 \textit{effective optical thickness} of a domain of length
-$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
+$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we get:
 \begin{equation}
@@
 For optically thin domains, we expect that most random walks will be
-communicated, while optically thick domains will leak the fraction of
-random walks that did not interact within. Using the optical thickness defined in
-Eq~(\ref{eq:optical_thickness_3}), we can then complete the leakage
-approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
-\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow \tau^{-1}$.
-With these bounds we then define the leakage fraction out of a domain using
-the Wigner rational approximation:
+communicated, while optically thick domains will leak the fraction of random
+walks that did not interact within. Using the optical thickness defined in
+Eq~(\ref{eq:optical_thickness_3}), we can complete the leakage approximations
+by defining the bounds of $\tau \rightarrow 0, \Lambda \rightarrow 1$ and
+$\tau \rightarrow \infty, \Lambda \rightarrow \tau^{-1}$.  With these bounds
+we define the leakage fraction out of a domain using the Wigner rational
+approximation:
 \begin{equation}
@@
 
-To test the relationships developed by the spectral analysis, we form
-two simple numerical experiments using the diffusion model problem:
-one to measure the length of the random walks as a function of the
-iteration matrix eigenvalues, and one to measure the domain leakage
-fraction as a function of the iteration matrix eigenvalues and the
-discretization properties. Before doing this, we verify our
-computation of the spectral radius of the iteration matrix by
-numerically computing the largest eigenvalue of the diffusion operator
-using an iterative eigenvalue solver. For this verification, a $100
-\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
-absorption cross varied from 0 to 100 while the scattering cross
-section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
-the measured spectral radius of the iteration matrix and the computed
-spectral radius for the preconditioned diffusion operator using
-Eq~(\ref{eq:iteration_radius}) as function of the absorption to
-scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
-observed between the analytic and numerical results with all data
-points computed within the tolerance of the iterative eigenvalue
-solver.
+To test the relationships developed by the spectral analysis, we form two
+simple numerical experiments using the diffusion model problem: one to measure
+the length of the random walks as a function of the iteration matrix
+eigenvalues, and one to measure the domain leakage fraction as a function of
+the iteration matrix eigenvalues and the discretization properties. Before
+doing this, we verify our computation of the spectral radius of the iteration
+matrix by numerically computing the largest eigenvalue of the diffusion
+operator using an iterative eigenvalue solver. For this verification, a $100
+\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ was used and the
+absorption cross varied from 0 to 100 while the scattering cross section was
+fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured spectral
+radius of the iteration matrix and the computed spectral radius for the
+preconditioned diffusion operator using Eq~(\ref{eq:iteration_radius}) as
+function of the absorption to scattering ratio $(\Sigma_a /
+\Sigma_s)$. Excellent agreement was observed between the analytic and
+numerical results with all data points computed within the tolerance of the
+iterative eigenvalue solver.
 \begin{figure}[ht!]
   \begin{center}
-    \includegraphics[width=\textwidth]{measured_spec_rad.png}
+    \includegraphics[width=0.8\textwidth]{measured_spec_rad.png}
   \end{center}
@@
 
-With the spectral data in hand, we can go about setting up an experiment to
-measure the length of the random walks generated by the adjoint Neumann-Ulam
-solver. To do this, we again use a $100 \times 100$ square grid with $h=0.1$
-and the absorption cross varied from 0 to 100 while the scattering cross
-section was fixed at unity. Three weight cutoff values of \sn{1}{-2},
-\sn{1}{-4}, and \sn{1}{-8} were used with 10,000 random walks generated by a
-point source of strength 1 in the center of the domain. For each of the
-random walks, the number of transitions made was tallied to provide an effective
-value of $k$ for each random walk. This value was then averaged over all random walks
-to get a measured value of $k$ for the particular operator. On the left,
-Figure~\ref{fig:measured_length} presents these measurements as well as the
-analytic result computed by Eq~(\ref{eq:analytic_k}) as a function of the
-iteration matrix spectral radius, $\rho(\ve{H})$. On the right,
-Figure~\ref{fig:measured_length} gives the relative error between the
-predicted and observed results. We note good qualitative agreement between the
-measured and analytic results. However, we observe a larger relative error for
-both long and short random walks.
+We next create an experiment to measure the length of the random walks
+generated by the adjoint Neumann-Ulam solver. To do this, we again use a $100
+\times 100$ square grid with $h=0.1$ and the absorption cross varied from 0 to
+100 while the scattering cross section was fixed at unity. Three weight cutoff
+values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with 10,000 random
+walks generated by a point source of strength 1 in the center of the
+domain. For each of the random walks, the number of transitions made was
+tallied to provide an effective value of $k$ for each random walk. This value
+was then averaged over all random walks to get a measured value of $k$ for the
+particular operator. On the left, Figure~\ref{fig:measured_length} presents
+these measurements as well as the analytic result computed by
+Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix spectral
+radius, $\rho(\ve{H})$. On the right, Figure~\ref{fig:measured_length} gives
+the relative error between the predicted and observed results. We note good
+qualitative agreement between the measured and analytic results. However, we
+observe a larger relative error for both long and short random walks.
 \begin{figure}[ht!]
@@
 
-Finally, we seek to measure the leakage from a domain in a domain decomposed
-Monte Carlo calculation and assess the quality of our analytic relation for
-the optical thickness of domain and the associated leakage approximations. For
-this experiment, a square grid with $h=0.1$ was decomposed into 9 square
-domains, 3 in each cardinal direction with measurements occurring in the
-central domain without boundary grid points. For cross sections, the
-absorption cross section was varied from 1 to 100 while the scattering cross
-section was set to zero to create a purely absorbing environment with weight
-cutoff of \sn{1}{-4}. The optical thickness of these domains will vary as a
-function of the absorption cross section if the other parameters are fixed. To
-compute the optical thickness, along with the spectral radius as given by
+Finally, we measure the leakage of random walks from a domain in a Monte Carlo
+calculation and assess the quality of our analytic relation for the optical
+thickness of the domain and the associated leakage approximations. For this
+experiment, a square grid with $h=0.1$ was decomposed into 9 square domains, 3
+in each cardinal direction with measurements occurring in the central domain
+without boundary grid points. For cross sections, the absorption cross section
+was varied from 1 to 100 while the scattering cross section was set to zero to
+create a purely absorbing environment with weight cutoff of \sn{1}{-4}. The
+optical thickness of these domains will vary as a function of the absorption
+cross section if the other parameters are fixed.
+
+To compute the optical thickness, along with the spectral radius as given by
 Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and $n_s$
@@
 of states moved along that typical length per random walk transition. For our
-grid above, the domains are varied in size with $50 \times 50$, $100 \times
-100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$
-grid points or states along the typical length of the domain
-respectively. Looking at the Laplacian stencil in
-Eq~(\ref{eq:nine_point_stencil}), we see that all random walk transitions will
-only move a single state in either the $i$ or $j$ directions due to the
-symmetry of the problem. Furthermore, if we choose the $i$ direction, not all
-states we will transition to will move the random walk in that
-direction. Therefore, we look to the definition of the iteration matrix in
-Eq~(\ref{eq:iteration_stencil}) and the definition of the adjoint probability
-matrix in Eq~(\ref{eq:adjoint_probability}) to estimate the $n_s$
-parameter. For a particular transition starting at state $(i,j)$, 6 of the 8
-possible new states in the stencil move the random walk in $i$ direction with
-relative coefficients of 4 for moving in the $(\pm i,0)$ direction and of 1
-for moving in the $(\pm i,\pm j)$. These coefficients dictate the frequency
-those states are visited relative to the others. For those 6 states we can
-visit along the typical length, their sum is 12 out of the total 20 for the
-coefficients for all possible states with their ratio giving $n_s =
-\frac{3}{5}$.
+two-dimensional Cartesian grid, the domains are varied in size with $50 \times
+50$, $100 \times 100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$,
+and $n_i=200$ grid points or states along the typical length of the domain
+respectively.
+
+To compute $n_s$, again consider the graph of the diffusion matrix stencil
+given in Figure~\ref{fig:diffusion_graph}. We see that all random walk
+transitions will only move a single state in either the $i$ or $j$ directions
+due to the symmetry of the problem. Furthermore, if we choose a chord in the
+$i$ direction, not all states to which we will transition move the random walk
+in that direction. Instead, $n_s$ will be representative of the average
+movement in the $i$ direction based on probabilities defined by the definition
+of the iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
+of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability}). For a
+particular transition starting at state $(i,j)$, 6 of the 8 possible new
+states in the stencil move the random walk in $i$ direction with relative
+coefficients of 4 for moving in the $(\pm i,0)$ direction and of 1 for moving
+in the $(\pm i,\pm j)$. These coefficients dictate the frequency those states
+are visited relative to the others. For those 6 states we can visit along the
+typical length, their sum is 12 out of the total 20 for the coefficients for
+all possible states with their ratio giving $n_s = \frac{3}{5}$.
 
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{leakage_variation.png}
+    \includegraphics[width=0.9\textwidth]{leakage_variation.png}
   \end{center}
-  \caption{Measured and analytic domain leakage as a function of the
-    iteration matrix spectral radius. To test the behavior with
-    respect to domain size, $n_i=50$, $n_i=100$,and $n_i=200$ were
-    used. The red data was computed numerically by a
-    domain-decomposed adjoint Neumann-Ulam implementation, the black
-    dashed data was generated by
+  \caption{Measured and analytic domain leakage as a function of the iteration
+    matrix spectral radius. To test the behavior with respect to domain size,
+    $n_i=50$, $n_i=100$,and $n_i=200$ were used. The red data was computed
+    numerically by a domain-decomposed adjoint Neumann-Ulam implementation,
+    the black dashed data was generated by
     Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{leakage_error.png}
+    \includegraphics[width=0.9\textwidth]{leakage_error.png}
   \end{center}

### 2015-04-14 doc/ned2015/slattery_wilson_evans.tex
adding improved data plots

@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{measured_spec_rad.png}
+    \includegraphics[width=0.8\textwidth]{spec_rad_results.eps}
   \end{center}
@@
     scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
-    $h=1.0$ were used. The red data was computed numerically by an
+    $h=1.0$ were used. The colored data was computed numerically by an
     eigensolver while the black dashed data was generated by
@@
 was then averaged over all random walks to get a measured value of $k$ for the
-particular operator. On the left, Figure~\ref{fig:measured_length} presents
+particular operator. Figure~\ref{fig:measured_length} presents
 these measurements as well as the analytic result computed by
 Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix spectral
-radius, $\rho(\ve{H})$. On the right, Figure~\ref{fig:measured_length} gives
+radius, $\rho(\ve{H})$. Figure~\ref{fig:measured_length_error} gives
 the relative error between the predicted and observed results. We note good
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{measured_length.png}
+    \includegraphics[width=0.8\textwidth]{random_walk_length.eps}
   \end{center}
-  \caption{Measured and analytic random walk length as a function of
-    the iteration matrix spectral radius. The weight cutoff was
-    varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
-    plot, the red data was computed numerically by an adjoint
-    Neumann-Ulam implementation while the black dashed data was
-    generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
-    relative error between the predicted and measured results is
-    presented for each weight cutoff.}
+  \caption{Measured and analytic random walk length as a function of the
+    iteration matrix spectral radius. The weight cutoff was varied with
+    \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. The colored data was computed
+    numerically by an adjoint Neumann-Ulam implementation while the black
+    dashed data was generated by Eq~(\ref{eq:analytic_k}).}
   \label{fig:measured_length}
+\end{figure}
+\begin{figure}[ht!]
+  \begin{center}
+    \includegraphics[width=0.8\textwidth]{random_walk_error.eps}
+  \end{center}
+  \caption{Relative error between computed random walk length analytic length
+    computed by Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
+    spectral radius. The weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4},
+    and \sn{1}{-8}.}
+  \label{fig:measured_length_error}
 \end{figure}
@@
   \begin{center}
-    \includegraphics[width=0.9\textwidth]{leakage_variation.png}
+    \includegraphics[width=0.8\textwidth]{leakage_fraction.eps}
   \end{center}
@@
     matrix spectral radius. To test the behavior with respect to domain size,
-    $n_i=50$, $n_i=100$,and $n_i=200$ were used. The red data was computed
+    $n_i=50$, $n_i=100$,and $n_i=200$ were used. The colored data was computed
     numerically by a domain-decomposed adjoint Neumann-Ulam implementation,
@@
   \begin{center}
-    \includegraphics[width=0.9\textwidth]{leakage_error.png}
+    \includegraphics[width=0.8\textwidth]{leakage_error.eps}
   \end{center}
@@
     function of the iteration matrix spectral radius.  To test the
-    behavior with respect to domain size, $n_i=50$ (green),
-    $n_i=100$ (blue), and $n_i=200$ (red) were used. The dashed
+    behavior with respect to domain size, $n_i=50$,
+    $n_i=100$, and $n_i=200$ were used. The dashed
     lines represent the error using the Wigner rational

### 2015-04-14 doc/ned2015/response_to_reviewer.txt
adding improved data plots
finishing ned edits

@@
 definitions for clarity.
+
+To improve clarity of the results, we also have replotted our results using a
+different plotting tool to create higher quality images that should aid the
+reader. Note that the y-axis in figure 2 has been changed to a logarithmic
+scale and the random walk length and error measurements have been broken into
+two different plots.
 

### 2015-04-14 doc/ned2015/slattery_wilson_evans.tex
finishing ned edits

@@
   \begin{keyword}
-    Monte Carlo \sep domain decomposition \sep MCSA \sep parallel \sep
-    computing \sep linear solvers
+    Monte Carlo \sep domain decomposition \sep MCSA \sep parallel computing
+    \sep linear solvers
   \end{keyword}
@@
 (MCSA) methods that, for certain classes of problems, are competitive with
-conventional methods methods and are applicable to a wide variety of linear
-systems \cite{evans_monte_2009,evans_monte_2014}. To apply these new MCSA
-methods to engineering problems, our research aims to improve the underlying
-Monte Carlo algorithms to handle large-scale problems in parallel computing
-environments.
+conventional methods methods and are applicable to a variety of linear systems
+\cite{evans_monte_2009,evans_monte_2014}. To apply these new MCSA methods to
+engineering problems, our research aims to improve the underlying Monte Carlo
+algorithms to handle large-scale problems in parallel computing environments.
 
@@
 permutation requires the computation of an initial starting state, $i_0$, and
-initial weight, $W_0$. The initial state, $i_0$, of the random walk is
-determined by sampling the source vector $\ve{b}$ with probabilities:
+initial weight, $W_0$. The initial state is determined by sampling the source
+vector $\ve{b}$ with probabilities:
 \begin{equation}
@@
 Using our result from Eq~(\ref{eq:inner_product_constraint}) the Monte Carlo
-estimator for the solution in state $i$ for a particular random walk
-permutation of length $k$ is:
+estimator for the solution from a particular random walk permutation of length
+$k$ is:
 \begin{equation}
@@
 \end{equation}
-where the Kronecker delta indicates that the tally contributes only in the
-current state, $i_m$. Finally, the expectation value using all random walk
-permutations is:
+where the Kronecker delta indicates that the tally contribution for each step
+of the random walk is only in its current state, $i_m$. Finally, the
+expectation value using all random walk permutations is:
 \begin{equation}
@@
 
-For our numerical experiments, we choose the one-speed,
+For our derivations and numerical experiments, we choose the one-speed,
 two-dimensional neutron diffusion equation as a model problem
@@
 
-Compared to the serial construction of the probabilities and weights, data for
-all states that are required to be on-process must be collected in parallel so
-that all possible random walks that leave the local domain may be correctly
+Compared to the serial construction of the probabilities and weights, domain
+decomposition requires that data for all states be on-process so that all
+possible random walks that leave the local domain may be correctly
 communicated. As an example of walking through the sparse graph of the matrix
@@
 the node at which we started with a probability proportional to the
-coefficients of each node in Eq~(\ref{eq:fd_system}). If any grid point
-adjacent to node $(i,j)$ is on another processor, the random walk must be
-communicated to that processor to continue the walk.
+coefficients of each node in Eq~(\ref{eq:fd_system}). If any node adjacent to
+node $(i,j)$ is on another processor and the walk moves to that node, the
+random walk must be communicated to that processor to continue the walk.
 
@@
 applying the operator to the eigenfunctions and noting that $x=ih$ and $y=jh$
-on a rectilinear Cartesian grid:
+on a square Cartesian grid:
 \begin{multline}
@@
 is sufficient to reduce the spectral radius of the iteration matrix below
-unity and therefore ensure convergence of the Neumann series. The
-preconditioner in this case is $\ve{M} = diag(\ve{D})$ such that we are
-solving the following linear system:
+unity and therefore ensure convergence of the Neumann series\footnote{Although
+  $\rho(\ve{H}) < 1$ is a sufficient criteria for convergence of the Neumann
+  series, recent work by Ji, Mascagni, and Li provides an additional necessary
+  and sufficient condition for convergence of the Neumann-Ulam Monte Carlo
+  algorithm \cite{ji_2013}. A diagonal scaling of the model problem also
+  satisfies this condition.}. The preconditioner in this case is $\ve{M}
+= diag(\ve{D})$ such that we are solving the following linear system:
 \begin{equation}
@@
 \end{equation}
-with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
-  index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
-error at the $k^{th}$ iterate, the error after $k$ iterations is:
+with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index} and $\ve{c}$
+a modified source term. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the
+solution error at the $k^{th}$ iterate, the error after $k$ iterations is:
 \begin{equation}
@@
 
-In a domain decomposed situation, not all random walks will remain within the
+In a domain decomposed algorithm, not all random walks will remain within the
 domain they started in and must instead be communicated. This communication,
@@
 will move in the grid by computing the mean squared distance of its movement
-along the chord of length $l$ defined along a cardinal direction of the
-domain. After a single transition a random walk will have moved a mean squared
-distance of:
+along the chord of length $l$ defined along a cardinal direction of the domain
+(i.e. $x$ or $y$). After a single transition a random walk will have moved a
+mean squared distance of:
 \begin{equation}
@@
 \end{equation}
-Here, the leakage fraction is explicitly bound to the eigenvalues of
-the iteration matrix, the size of the domain, the content of the
-discretization stencil, and the weight cutoff selected to terminate
-low weight random walks.
+Our derived leakage fraction is dependent on the eigenvalues of the iteration
+matrix, the size of the domain, the content of the discretization stencil, and
+the weight cutoff selected to terminate low weight random walks.
 
@@
   \end{center}
-  \caption{Measured and analytic preconditioned diffusion operator
+  \caption{\textbf{Measured and analytic preconditioned diffusion operator
       spectral radius as a function of the absorption cross section to
-    scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
-    $h=1.0$ were used. The colored data was computed numerically by an
+      scattering cross section ratio.} \textit{Values of $h=0.01$, $h=0.1$,
+      and $h=1.0$ were used. The colored data was computed numerically by an
       eigensolver while the black dashed data was generated by
-    Eq~(\ref{eq:iteration_radius}).}
+      Eq~(\ref{eq:iteration_radius}).}}
   \label{fig:measured_spec_rad}
@@
 was then averaged over all random walks to get a measured value of $k$ for the
-particular operator. Figure~\ref{fig:measured_length} presents
-these measurements as well as the analytic result computed by
+particular operator. Figure~\ref{fig:measured_length} presents these
+measurements as well as the analytic result computed by
 Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix spectral
-radius, $\rho(\ve{H})$. Figure~\ref{fig:measured_length_error} gives
-the relative error between the predicted and observed results. We note good
+radius, $\rho(\ve{H})$. Figure~\ref{fig:measured_length_error} gives the
+relative error between the predicted and observed results. We note good
 qualitative agreement between the measured and analytic results. However, we
@@
   \end{center}
-  \caption{Measured and analytic random walk length as a function of the
-    iteration matrix spectral radius. The weight cutoff was varied with
-    \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. The colored data was computed
-    numerically by an adjoint Neumann-Ulam implementation while the black
-    dashed data was generated by Eq~(\ref{eq:analytic_k}).}
+  \caption{\textbf{Measured and analytic random walk length as a function of
+      the iteration matrix spectral radius.} \textit{The weight cutoff was
+      varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. The colored data was
+      computed numerically by an adjoint Neumann-Ulam implementation while the
+      black dashed data was generated by Eq~(\ref{eq:analytic_k}).}}
   \label{fig:measured_length}
@@
   \end{center}
-  \caption{Relative error between computed random walk length analytic length
-    computed by Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
-    spectral radius. The weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4},
-    and \sn{1}{-8}.}
+  \caption{\textbf{Relative error between computed random walk length analytic
+      length computed by Eq~(\ref{eq:analytic_k}) as a function of the
+      iteration matrix spectral radius.} \textit{The weight cutoff was varied
+      with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}.}}
   \label{fig:measured_length_error}
@@
   \end{center}
-  \caption{Measured and analytic domain leakage as a function of the iteration
-    matrix spectral radius. To test the behavior with respect to domain size,
-    $n_i=50$, $n_i=100$,and $n_i=200$ were used. The colored data was computed
-    numerically by a domain-decomposed adjoint Neumann-Ulam implementation,
-    the black dashed data was generated by
+  \caption{\textbf{Measured and analytic domain leakage as a function of the
+      iteration matrix spectral radius.} \textit{To test the behavior with
+      respect to domain size, $n_i=50$, $n_i=100$,and $n_i=200$ were used. The
+      colored data was computed numerically by a domain-decomposed adjoint
+      Neumann-Ulam implementation, the black dashed data was generated by
       Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
@@
       Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
-    approximation.}
+      approximation.}}
   \label{fig:measured_leakage}
 \end{figure}
+
 Again, we note good qualitative agreement between the measured and
@@
   \end{center}
-  \caption{Measured and analytic domain leakage absolute error as a
-    function of the iteration matrix spectral radius.  To test the
-    behavior with respect to domain size, $n_i=50$,
-    $n_i=100$, and $n_i=200$ were used. The dashed
-    lines represent the error using the Wigner rational
-    approximation while the solid lines represent the error using
-    the mean-chord approximation.}
+  \caption{\textbf{Measured and analytic domain leakage absolute error as a
+      function of the iteration matrix spectral radius.} \textit{To test the
+      behavior with respect to domain size, $n_i=50$, $n_i=100$, and $n_i=200$
+      were used. The dashed lines represent the error using the Wigner
+      rational approximation while the solid lines represent the error using
+      the mean-chord approximation.}}
   \label{fig:leakage_error}
@@
 
-In future work, these relationships will serve as guidelines for
-selecting an appropriate parallel algorithm strategy and provide a
-basis for performance models for future parallel implementations of
-the adjoint Neumann-Ulam method. To be applicable to a broader range
-of problems, these models could potentially be extended to
-non-symmetric systems. In addition, these relations will be used to
-analyze parallel implementations of Monte Carlo synthetic acceleration
-methods that leverage the adjoint method and serve as an initial
-grounds for assessing their feasibility for large-scale problems.
+In future work, these relationships will serve as guidelines for selecting an
+appropriate parallel algorithm strategy and provide a basis for performance
+models for future parallel implementations of the adjoint Neumann-Ulam
+method. To be applicable to a broader range of problems, these models could
+potentially be extended to non-symmetric and three-dimensional systems. In
+addition, these relations will be used to analyze parallel implementations of
+Monte Carlo synthetic acceleration methods that leverage the adjoint method
+and serve as an initial grounds for assessing their feasibility for
+large-scale problems.
 

### 2015-04-29 doc/ned2015/response_to_reviewer.txt

@@
 based on the underlying graph of the matrix. We have added a figure (Figure 1)
-in section 3.1 and discussion the random walk in terms of the sparse graph
+in section 3.1 and discussion of the random walk in terms of the sparse graph
 underneath the matrices that should help clarify the explanation, particularly
@@
 
-To improve clarity of the results, we also have replotted our results using a
+To improve clarity of the results, we also have re-plotted our results using a
 different plotting tool to create higher quality images that should aid the
@@
 rules of the Neumann-Ulam game where the data for the random walks (weights
-and probablities) has been domain-decomposed.
+and probabilities) has been domain-decomposed.
 
 Variance estimation, however, is a different story and domain decomposition
-schemes to date trade inexact variance estimates for scalablity. This is due
-to the fact that the order of events must be maintained to acheive a correct
+schemes to date trade inexact variance estimates for scalability. This is due
+to the fact that the order of events must be maintained to achieve a correct
 variance estimate and fully asynchronous algorithms do not, in general, permit

### 2015-07-10 doc/ned2015/build_figs.sh
adding ned final changes

@@ -1,0 +1,2 @@
+fig2dev -L pdftex   -p stencil_graph.pdf -m 0.4 stencil_graph.fig stencil_graph.pdf
+fig2dev -L pdftex_t -p stencil_graph.pdf -m 0.4 stencil_graph.fig stencil_graph.tex

### 2015-08-21 MAX_reports/CUDA_report/CUDA_report.tex
tex files included for the reports

\documentclass[a4paper,10pt]{article}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{mathtools}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{geometry}
\usepackage{subfig}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabu}
\usepackage{multirow}
\usepackage{color}
\usepackage{listings}
\usepackage[]{algorithm2e}
\usepackage[autostyle,english=british]{csquotes}
\usepackage{minitoc}

\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem{prop}{Proposition}

%opening
\title{Technical Report about implementation in CUDA of Monte Carlo 
Linear Solvers (MCLS)}
\author{Steven Hamilton, Massimiliano Lupo Pasini}
\date{}

\begin{document}

\maketitle

\begin{abstract}
In this report timings are presented as concerns Monte 
Carlo Linear Solvers (MCLS) with different configurations. Different test 
cases are used to get quite a general overview of the behavior. 
In particular 
various options to generate random numbers as well as handling of data storage 
are 
tested for the sake of efficiency. All the test cases have been run three times 
to filter the fluctuations due to hardware issues. 
\end{abstract}

\section*{Introduction}
In this report different approaches are tested in order to maximize the 
performance of Monte Carlo linear solvers algorithms in GPU environment. 
Both Forward and Adjoint method are analyzed, pointing out the different 
implementation issues and pinpointing the different viable options to improve 
timings.
Four different numerical test cases are employed with different properties in 
terms of spectral radius of the iteration matrix, sparsity pattern, degrees 
of freedom and number of histories employed. The study is accomplished both by 
using a weight cutoff to kill the 
histories and by using a fixed number of steps for each of those, in order to 
discriminate the actual effectiveness of the different modifications 
introduced. An attempt to justify the improvements sometimes detected is 
provided, however a deeper understanding of the behaviors still has to be 
carried out.

\section{Adjoint method - Different random number generation techniques}
The generation of random numbers plays an important role in the calculation of 
the solution to 
a linear system through a Monte Carlo procedure, since a new random number must 
be generated 
to either step ahead in a random walk or to kink off a new one.
Since this operation must be repeated very frequently in the code, it is 
recommendable to 
decrease its cost as much as possible, not to affect the efficiency of the 
overall performance.
Two viable options to initialize random number generators are:

\begin{enumerate}
 \item Having the same seed but different sequence number generates a number 
guaranteed 
to be $2^{67}$ away from each other, 
but the downside is the heavy computations to advance the $2^{67}$ position

\hspace*{-2cm}
\begin{lstlisting}
__global__ void initialize_rng(curandState *state, int seed, int offset)
{
    int tid = threadIdx.x + blockIdx.x * blockDim.x;

    curand_init(seed,tid,offset,&state[tid]);

}
\end{lstlisting}

\item Giving different seeds, and just keeping the sequence number at 0, 
it is a lot faster but there might be correlation between threads, since there 
is 
no 
guarantee on the separation between each threads. In this case the different 
seeds can be picked accordingly to different criteria. In this report we 
explore two approaches:
\begin{itemize}
 \item picking the seeds with consecutive indices
 \item picking the seeds with indices provided by a generator of integer random 
numbers
\end{itemize}

In case seeds with consecutive indices are chosen the set of instructions 
employed to generate the seed vector is the following:

\hspace*{-2cm}
\begin{lstlisting}
else if ( d_seed_type==SEED_TYPE::DIFF )
{
    std::cout<<"Different adjacent seeds instantiated"<<std::endl;

    thrust::device_vector<int> seeds( BLOCK_SIZE*num_blocks);
    thrust::sequence(seeds.begin(), seeds.end(), d_rng_seed);
    int* seed_ptr = thrust::raw_pointer_cast(seeds.data());

    initialize_rng<<<num_blocks, BLOCK_SIZE>>>(rng_states, seed_ptr,
        d_num_curand_calls, d_seed_type);
}

\end{lstlisting}

Instead in the case of seeds chosen with random indices the set of instructions 
executed is:

\hspace*{-2cm}
\begin{lstlisting}
else if ( d_seed_type==SEED_TYPE::RAND )
{
    std::cout<<"Different random seeds instantiated from 0 to "<<
    RAND_MAX<<std::endl;

    thrust::device_vector<int> dev_seeds( BLOCK_SIZE*num_blocks);
    thrust::host_vector<int> host_seeds( BLOCK_SIZE*num_blocks );
    std::srand(std::time(0));
    thrust::generate(host_seeds.begin(), host_seeds.end(), std::rand);
    dev_seeds=host_seeds;
    int* seed_ptr = thrust::raw_pointer_cast(dev_seeds.data());

    initialize_rng<<<num_blocks, BLOCK_SIZE>>>(rng_states, seed_ptr,
       d_num_curand_calls, d_seed_type);
}

\end{lstlisting}

The macro variable \texttt{RAND\_MAX} is set to 2147483647. Therefore 
we have the guarantee that a wide range of integers is spanned for the choice 
of the seeds.

In case as many seeds as the number of threads are 
employed, the kernel remains the same independently of the policy adopted for 
the selection of the seeds.

\hspace*{-2cm}
\begin{lstlisting}
__global__ void initialize_rng(curandState *state, int*seed, int offset)
{
    int tid = threadIdx.x + blockIdx.x * blockDim.x;

    curand_init(seed[tid], 0, offset, &state[tid]);
}
\end{lstlisting}
\end{enumerate}

\begin{table}[!h]
\hspace*{-2cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & \textbf{\% time rng} 
& \textbf{\% time kernel} & tot. time (ms)& rel. residual norm\\
\hline
2D Laplacian& 900 & $10^7$ & 36.0 & 64.0 & 299,563 & 0.0931241 \\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.05 & 99.5 & 50,016 & 12.3509\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 18.3 & 81.7 & 640,563 & 0.391686 \\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 18.0 & 82.0 & 349,096 & 0.450792 \\
\hline
\end{tabular}
\caption{Adjoint method - Timings using same seed and different sequences.} 
\label{tab1}
\end{table}

\begin{table}[!h]
\hspace*{-2.5cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & \textbf{\% time rng} 
& \textbf{\% time kernel} & tot. time (ms)& rel. res. norm & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.0 & 100.0 & 191,818 & 0.119588 & 0.64 \\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ & 0.0 & 100.0 & 35,351 & 12.3914 & 0.707\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 0.0 & 100.0 & 517,788 & 0.408134 & 
0.808\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.0 & 100.0 & 241,671 & 0.397887 & 0.692\\
\hline
\end{tabular}
\caption{Adjoint method - Timings using different consecutive seeds and same 
sequence.} 
\label{tab2}
\end{table}

\begin{table}[!h]
\hspace*{-2.5cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & \textbf{\% time rng} 
& \textbf{\% time kernel} & tot. time (ms)& rel. res. norm & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.0 & 100.0 & 193,381 & 0.112068 & 0.64\\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ & 0.0 & 100.0 & 36,295 & 12.3404 & 0.726\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 0.0 & 100.0 & 549,992 & 0.408296 & 
0.859\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.0 & 100.0 & 241,817 & 0.383815 & 0.693\\
\hline
\end{tabular}
\caption{Adjoint method - Timings using different seeds picked randomly and 
same 
sequence.} 
\label{tab3}
\end{table}

As it can be noticed from the previous Tables \ref{tab1}, \ref{tab2} and 
\ref{tab3}, the second and third options 
provides better timings. However the payoff 
is an increase of the final relative residual, likely due to the correlation 
between the sequences generated by different seeds.
The quantities "time rng" and "time kernel" are expressed in terms of percentage 
of the total time.

\section{Adjoint method - Use of texture memory}

The LDG instruction (exposed via the \texttt{\_\_ldg} intrinsic) is a memory 
load that uses the texture path. 
It has the advantage that it does not require the explicit use of textures, 
since it does not 
explicitly bind one. 
Therefore \texttt{\_\_ldg()} reads data through the texture path, without 
requiring a texture itself. 
It is an overloaded function with the prototype \texttt{\_\_ldg(const *T)} 
where T is one of CUDA's built-in types. 
The perk of using LDG instruction is that explicit uses of textures cause a 
certain 
amount of code clutter and overhead (e.g. for API calls to bind textures).
Classical textures also use the texture load path, but in addition can transform 
both index (e.g. clamping modes) 
and data returned (e.g. interpolation) in various ways; the necessary control 
information 
is provided to the hardware during texture binding.
Because the texture cache is non-coherent with 
respect to writes in the same kernel, use of the texture load path requires that 
the underlying data 
is read-only across the entire kernel.

\section{Adjoint method - LDG calls into the code}

LDG instructions have been introduced in the following functions of Profugus 
code:

\begin{enumerate}
 \item \texttt{lower\_bound}
 \item \texttt{initialize\_history}
 \item \texttt{getNewState}
 \item \texttt{tallyContribution}
\end{enumerate}

All the instructions employed above provide a runtime decrease. However, some of 
them are more
significant than others. In particular the most effective calls are the ones 
located in
1) and 3). This is due to the fact that these functions are the ones called most 
frequently.
The option of using LDG instruction at point 4) actually plays a role just when 
the expected value
estimator is employed.

In the examples represented below the length of the history is set to 10000 
steps. The simulations have been accomplished both with and without a weight 
cut off equal to $10^{-9}$.

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 191,818 \\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 35,351\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 517,788\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 241,671\\
\hline
\end{tabular}
\caption{Adjoint method - Timings without LDG instructions and constant history 
length equal to 
10,000.}
\label{tab4}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)& Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 139,882 & 0.729\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 37,610 & 1.064\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 500,092 & 0.966\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 157,545 & 0.652\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 1) and constant 
history length equal 
to 
10,000.}
\label{tab5}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 193,336 & 1.008\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 37,759 & 1.068\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 548,909 & 1.060\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 241,490 & 1\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 2) and constant 
history length equal 
to 
10,000.}
\label{tab6}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 105,010 & 0.547\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 36,340 & 1.028\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 519,408 & 1.003\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 200,559 & 0.830\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 3) and constant 
history length equal 
to 
10,000.}
\label{tab7}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 140,347 & 0.732\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 36,344 & 1.028\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 499,007 & 0.964\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 157,458 & 0.652\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 1), 2) and constant 
history length 
equal 
to 
10,000.}
\label{tab8}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)& Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 105,166 & 0.548\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 36,141 & 1.022\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 547,777 & 1.052\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 200,325 & 0.829\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 2), 3) and constant 
history length 
equal 
to 
10,000.}
\label{tab9}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 105,331 & 0.549\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 34,616 & 0.979\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 520,762 & 1.006\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 138,091 & 0.571\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 1), 2), 3) and 
constant history 
length 
equal 
to 
10,000.}
\label{tab10}
\end{table}

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 27,686\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 35,108\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 40,720\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 4,919\\
\hline
\end{tabular}
\caption{Adjoint method - Timings without LDG instructions and with weight 
cutoff.}
\label{tab11}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 26,344 & 0.952\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 35,911 & 1.023\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 41,092 & 1.009\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 4,759 & 0.967\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 1) and with weight 
cutoff.}
\label{tab12}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)& Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 27,722 & 1.001\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 36,005 & 1.026\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 40,001 & 0.982\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 5,585 & 1.135\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 2) and with weight 
cutoff.}
\label{tab13}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 24,814 & 0.896\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 36,604 & 1.043\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 40,305 & 0.99\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 5,065 & 1.03\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 3) and with weight 
cutoff.}
\label{tab14}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 26,248 & 0.948\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 35,741 & 1.018\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 40,017 & 0.983\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 4,701 & 0.956\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 1) and 2) and with 
weight cutoff.}
\label{tab15}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 27,184 & 0.982\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 36,141 & 1.029\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 40,727 & 1\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 5,086 & 1.034\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 2), 3) and with 
weight cutoff.}
\label{tab16}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)& Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 24,307 & 0.878\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 36,980 & 1.053\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 40,596 & 0.997\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 4,921 & 1\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with LDG instructions at 1), 2) and 3) and 
with weight cutoff.}
\label{tab17}
\end{table}

In Tables \ref{tab5}-\ref{tab10} the relative timing is compared with the 
reference values in Table \ref{tab4}. Tables \ref{tab12}-\ref{tab17} contain 
relative timings with respect to the reference values in Table \ref{tab11}.

By comparing the values of Tables \ref{tab4}-\ref{tab11} it is pointed out that 
a benefit in terms of timings is achieved 
when \texttt{ldg} instructions are employed in all the three subroutines at 
issue.
Moreover the improvement is more 
evident for the 2D laplacian and the $SP_1$ matrix. The reason of this might be 
associated with the sparsity pattern. 
Indeed a higher number of nonzero entries induces the histories to run for 
longer, enhancing the utility of the 
texture memory might increase as well. For the 2D laplacian and the $SP_1$ 
matrix the employment of \texttt{\_\_ldg} 
instructions almost halves the time for the computation. 
In particular the instruction which seems to be more effective is the one 
located in the \texttt{getNewState} subroutine.
The time reduction gets 
weaker for the other test cases. In fact for the 1D laplacian it seems there is 
no benefit in terms of timings coming from the texture memory.

By looking at the results of the same test cases when a weight cutoff is used, 
we can see that the utility of \texttt{\_\_ldg} instructions is almost vanished 
(see 
Tables \ref{tab12}-\ref{tab17}). It still plays a significant role for the 2D 
Laplacian, plausibly because of the spectral radius $\rho(H)\approx 1$ which 
induces the histories to run for many steps even when the cutoff is employed. 
For the other cases instead the spectral radius is so small that the weight 
cutoff kills the histories too son for the texture memory to be effective.

\section{Adjoint method - Generation of random numbers}

Random numbers must be employed for each random walk to:
\begin{itemize}
 \item determine what is the initial state
 \item determine the following state in the path accordingly to the current one
\end{itemize}

Essentially the generation of random numbers is located in the subroutines:
\begin{enumerate}
\item \texttt{initializeHistory}
\item \texttt{getNewState}
\end{enumerate}

These two operations have to be repeated for all the random walks employed in 
the computations. 
In order to minimize the time spent in the generation of random numbers, it 
might be useful
to gather the generation of many of these at the same time. 
Therefore a gathering of the random number generator's calls has been 
accomplished, partially modifying 
the subroutines "\texttt{initializeHistory}" and "\texttt{getNewState}".
The relationship one-to-one between the generation of a random number and a call 
to one of these two 
subroutines is broken. A group of random numbers is generated in advance before 
the call to the
aforementioned functions. The size of the batch for this grouped random numbers 
is a parameter that can be tuned to find the optimal configuration, 
reducing the total time 
of execution.

The viable options that might be adopted are:
\begin{enumerate}
 \item to call separately the random number generator once for the 
initializeHistory and then
   employ the batch for the successive steps
  \item to start employing the batch even for the initial step of the history 
\end{enumerate}

\begin{table}[!h]
\hspace*{-2.5cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist} & \textbf{\% time rng} 
& \textbf{\% time kernel} & tot. time (ms)& rel. res. norm & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.0 & 100.0 & 192,325 & 0.119588 & 1.003\\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ & 0.0 & 100.0 & 36,602 & 12.3914 & 1.035\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 0.0 & 100.0 & 517,536 & 0.408134 & 1\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.0 & 100.0 & 241,928 & 0.397887 & 1.001\\
\hline
\end{tabular}
\caption{Adjoint method - Results for a single call of rng for both 
\texttt{initializeHistory} 
and \texttt{getNewState}.} 
\label{tab18}
\end{table}

\begin{table}[!h]
\hspace*{-2.5cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & \textbf{\% time rng} 
& \textbf{\% time kernel} & tot. time (ms)& rel. res. norm & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.0 & 100.0 & 22,634 & 0.119588 & 0.818\\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ & 0.0 & 100.0 & 37,002 & 12.3914 & 1.054\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 0.0 & 100.0 & 40,473 & 0.408134 & 0.994\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.0 & 100.0 & 4,806 & 0.397887 & 0.977\\
\hline
\end{tabular}
\caption{Adjoint method - Results for a single call of rng for both 
\texttt{initializeHistory} 
and \texttt{getNewState} with weight cut-off and use of ldg.} 
\label{tab19}
\end{table}

Overall, for a configuration where the length of the random walk is fixed, the 
two options seem to produce similar results (compare Tables \ref{tab4} 
and \ref{tab18}). Slightly better results are obtained instead with the second 
approach (compare values in Tables \ref{tab11} and \ref{tab19}) when a weight 
cutoff is used to terminate the histories.

The time reduction accomplished by grouping the generation of several random 
numbers has not produced significant effects though. Therefore
from now on \texttt{initializeHistory} and \texttt{getNewState} will be 
employed by making an explicit call to the random number generator every time 
that is 
necessary to produce a new random number.

\section{Adjoint method - Reorganization of data through \texttt{struct}}

In order to attempt the reduction of GPU timings, a 
reorganization of the matrices used by the Monte Carlo 
linear solver is accomplished.

In particular the pursue is to increase the vicinity of data that are going to 
be consulted in adjacent time steps by the computer.
Because of this a C++ \texttt{struct} has been introduced:
\hspace*{-2cm}
\begin{lstlisting}
struct device_row_data{
        double H;
        double P;
        double W;
        int inds;
};
\end{lstlisting}

When a new state 
has to be taken by a random walk, corresponding values of the 
transition probability, iteration matrix and weight are picked from the memory 
storage, providing the motivation for this attempt. The integer \texttt{inds} 
is used to store the index of the state, 
while the double values \texttt{H}, \texttt{P} and \texttt{W} are employed to 
store the entries of the iteration matrix, probability and weight respectively. 
This way, values of H, P, W associated with a particular entry are stored in 
contiguous cells of the memory.

This allows to reorganize the data necessary for the Monte Carlo linear solver 
in an array whose entries are \texttt{struct} elements. The length of the array 
corresponds to the number of nonzero entries of the iteration matrix.

In the following tables results associated with the employment of such data 
structure are presented, both by using a fixed length for the histories and by 
resorting to a weight cutoff of $10^{-9}$.

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 0.994869 & 0.99447 & 105,093 & 0.998\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 27,404 & 0.792\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 422,197 & 0.811\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 0.977674 & 0.999836 & 138,921& 1.006\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with \texttt{struct} data used and without 
weight cutoff. LDG 
instructions at 1), 2) and 
3)}
\label{tab20}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^7$ & 0.994869 & 0.99447 & 24,669 & 1.015\\
\hline 
1D shifted Laplacian& $10^6$ & $10^4$ & 0.799972 & 0.639977 & 26,907 & 0.728\\
\hline
1D shifted Laplacian& $10^6$ & $10^7$ & 0.799972 & 0.639977 & 30,773 & 0.758\\
\hline
$\text{SP}_1$ & $25,568$ & $10^7$ & 0.977674 & 0.999836 & 5,005 & 1.017\\
\hline
\end{tabular}
\caption{Adjoint method - Timings with \texttt{struct} data used. LDG 
instructions at 1), 2) and 
3) and with weight cutoff.}
\label{tab21}
\end{table}

Looking at both results in Tables \ref{tab20} and \ref{tab21} it is pointed out 
that the reorganization of data does not introduce any benefits as regards the 
2D Laplacian or the $SP_1$ matrix. However significant reduction of timings are 
obtained for the 1D shifted Laplacian.

\section{Forward method - Assignment of tasks to threads}

In the Adjoint method a single history contributes in the 
evaluation of different entries of the solution vector accordingly to the 
actual state visited. The Forward method instead is defined such that a single 
history always contributes on the same entry of the solution 
vector, depending on the state from which it has been initiated.

This typical property of the Forward method opens the way to different 
techniques to distribute the tasks between threads. In the problems at issue 
the concept of task coincide with a single history. Therefore the intent is to 
discover what is the best way to map histories to different threads accordingly 
to their index.

Regardless of the amount o histories employed for a single entry (when this 
number is bigger than one), two viable options to execute the assignment are:

\begin{itemize}
 \item to employ different threads for the estimation of an entry of the 
solution vector
\item create a one-to-one relationship between the entries of the 
solution vector and the threads initialized during the process.
\end{itemize}

The first option induces to deal with the issue of different threads trying to 
modify the value of the same entry (data concurrency). From the 
point of view of implementation different approaches can be use to cope with 
it. In the following dissertation we proceed storing the solution 
vector in 
global memory and executing an atomization of all those operation that
access and modify values contained in the same memory cell.
The second approach, instead, does not entail any data concurrency since each 
thread always accesses a memory cell which is never consulted by any other 
thread. However we have a restriction to the number of threads which 
actually do a job, since it is necessarily  
equal to the size of the linear system.

The goal is to compare the timings between these two settings, focusing on 
determining the effects of the serialization occurring in the execution 
of the kernel for the first option.\newline

In the case when different threads are allowed to work on the same entry, a 
possible mapping might be the following one:

\hspace*{-2cm}
\begin{lstlisting}
__global__ void run_forward_monte_carlo(...)
{

    ...

    int tid = threadIdx.x + blockIdx.x * blockDim.x;
    int entry = tid / entry_histories;

    ...

}   
\end{lstlisting}

As already mentioned previously in the dissertation, this requires a 
serialization of the access by different threads to the same memory cell. This 
is accomplished by the following instruction:

\hspace*{-2cm}
\begin{lstlisting}
__device__ void tallyContribution(int state, double wt, double * const x)
{
        // Collision estimator just adds weight
        atomicAdd(x+state,wt);
}
\end{lstlisting}

The one-to-one relationship between entries and threads instead induces 
instead the following mapping:

\begin{lstlisting}
__global__ void run_forward_monte_carlo(...)
{

    ...

    int tid = threadIdx.x + blockIdx.x * blockDim.x;
    int entry = tid / entry_histories;

    ...

}

\hspace*{-2cm}
\begin{lstlisting}
__global__ void run_forward_monte_carlo(...)
{

    ...

    int tid = threadIdx.x + blockIdx.x * blockDim.x; 
    if( tid < N )
    {
       ...
    }

    ...

}   
\end{lstlisting}

In the last case obviously it is necessary to instantiate a number of threads 
bigger than the actual size of the problem to solve, since in general the size 
of the blocks is not a divisor of the number of degrees of freedom. However, 
the higher is the size of the problem, the less this is an issue.\newline

For both the configurations the test cases are run both by using a fixed number 
of steps equal to 10,000 and by employing a weight cutoff equal to $10^{-9}$. 
No texture memory instructions are employed.

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 437,444 \\
\hline 
1D shifted Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 78,767\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 241,573\\
\hline
\end{tabular}
\caption{Forward method - mapping of multiple threads to a single entry. Fixed 
number of steps equal to 10,000.}
\label{tab22}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 79,306 & 0.181\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 74,621 & 0.947\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 122,773 & 0.508\\
\hline
\end{tabular}
\caption{Forward method - one-to-one mapping between threads and entries. Fixed 
number of steps equal to 10,000.}
\label{tab23}
\end{table}

Even if the case of the $SP_1$ matrix does not converge, we are going to 
consider the behavior of the code for this test case anyway. Indeed the 
priority is to
focus on the performance of the algorithm with respect to different 
instantiations of the threads rather than the accuracy of the final 
numerical result 
itself.

Comparing values in Tables \ref{tab22} and \ref{tab23}, we can see that the 
atomization of the contribution coming from each tally affect the 
performance of the algorithm for all the test cases analyzed, in particular for 
the 2D Laplacian and the $SP_1$ matrix. The outstanding improvement for these 
cases can be explained with the small size of the problems. 
In the first approach for the distribution of tasks, indeed, the tasks are 
attributed to the same entry according to a rotation of the blocks. Therefore 
threads working on the same entry of the solution vector will always belong to 
different blocks. In fact for a 
fixed block size, the higher is the size of the problem the lower is the 
contention of memory resources between threads belonging to different blocks.
This explains why results are pretty similar for the 1D shifted Laplacian, 
since its size is order of magnitudes bigger than for the other test cases 
analyzed.

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Histories per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 206,439 \\
\hline 
1D shifted Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 39,295\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 33,169\\
\hline
\end{tabular}
\caption{Forward method - mapping of multiple threads to a single entry. 
Weight cutoff equal to $10^{-9}$.}
\label{tab24}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 11,572 & 0.056\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 38,484 & 0.976\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 6,505 & 0.196\\
\hline
\end{tabular}
\caption{Forward method - one-to-one mapping between threads and entries. 
Weight cutoff equal to $10^{-9}$.}
\label{tab25}
\end{table}

Looking at Tables \ref{tab24} and \ref{tab25}, the employment of 
the one-to-one map still has a huge impact in terms of performance as concerns 
that 2D Laplacian and the $SP_1$ matrix. As regards the 1D shifted Laplacian, 
instead, the weight cutoff seems to have e dominant effect on the timing, 
eliding the utility of the different techniques for the assignment of the tasks 
to the threads instantiated. A plausible explanation for this phenomenon might 
be 
the sparsity pattern of the matrices analyzed. In fact the number of nonzero 
entries per row is much higher for the first and third problem than for the 
second. Moreover the spectral radii of $H$ and $\hat{H}$ are much smaller for 
the shifted 1D Laplacian. Therefore the average 
number of steps per history is lower for the 1D shifted Laplacian than for the 
other cases, explaining why this particular problem is transparent to the 
employment of different mappings for the task distribution.

\section{Forward method - Use of texture memory}
We now focus on comparing the change of performance of the Forward method 
depending on whether texture memory instructions are used or not. 
The collocation of the texture memory instructions is the same as for the 
Adjoint method.

By default the one-to-one 
mapping for the execution of the tasks by different threads is employed. 
Results 
are shown both by resorting to a fixed length of the histories and also by 
using an adaptive weight cutoff.

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms)& Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 52,077 & 0.657\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 63,353 & 0.849\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 72,198 & 0.588\\
\hline
\end{tabular}
\caption{Forward method - one-to-one mapping between threads and entries. Fixed 
number of steps equal to 10,000. LDG instructions employed.}
\label{tab26}
\end{table}

Comparing results in Table \ref{tab26} with the ones in Table \ref{tab23}, it 
is detected a significant improvement of the timings thanks to the employment 
of texture memory. Especially for the $SP_1$ matrix, supposedly for its 
sparsity pattern.

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timings\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 9,316 & 0.805\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 40,005 & 1.04\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 4,269 & 0.656\\
\hline
\end{tabular}
\caption{Forward method - one-to-one mapping between threads and entries. 
Weight cutoff equal to $10^{-9}$. LDG instructions employed.}
\label{tab27}
\end{table}

By making a comparison between Tables \ref{tab27} and \ref{tab25} we discover a 
similar behavior to the one verified for the Adjoint method. In fact once the 
weight cutoff is employed, the use of \texttt{\_\_ldg} instructions seem not to 
be effective for the reduction of the computational time.

\section{Forward method - Reorganization of data through \texttt{struct}}
In this section we focus on testing the efficiency of the employment of C++ 
\texttt{struct} objects to handle data, on the same lead as for the Adjoint 
method.
The experiments are repeated both for a fixed length of histories equal to 
10,000 and for the use of a weight cutoff equal to $10^{-9}$, with and without 
\texttt{ldg} instructions as well. The one-to-one mapping between 
tasks and threads is employed.

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 80,059 & 1.009\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 79,807 & 1.069\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 122,861 & 1\\
\hline
\end{tabular}
\caption{Forward method - use of struct for data handling. One-to-one mapping 
between threads and entries. 
Fixed number of steps equal to 10,000. No LDG instructions.}
\label{tab28}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 11,534 & 1.003\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 29,636 & 0.770\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 6,247 & 0.960\\
\hline
\end{tabular}
\caption{Forward method - use of struct for data handling. One-to-one mapping 
between threads and entries. 
Weight cutoff equal to $10^{-9}$. No LDG instructions.}
\label{tab29}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 50,159 & 0.627\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 67,993 & 0.852\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 75,305 & 0.598\\
\hline
\end{tabular}
\caption{Forward method - use of struct for data handling. One-to-one mapping 
between threads and entries. 
Fixed number of steps equal to 10,000. LDG instructions employed.}
\label{tab30}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & $\rho(H)$ 
& $\rho(\hat{H})$ & tot. time (ms) & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 0.994869 & 0.99447 & 8,950 & 1.041\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 0.799972 & 0.639977 & 30,314 & 0.758\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 0.977674 & 1.99619 & 4,040 & 0.946\\
\hline
\end{tabular}
\caption{Forward method - use of struct for data handling. One-to-one mapping 
between threads and entries. 
Weight cutoff equal to $10^{-9}$. LDG instructions employed.}
\label{tab31}
\end{table}

Results in Tables \ref{tab28} (compared with Table \ref{tab23}) and \ref{tab29} 
show (compared with Table \ref{tab25}) show
that actually the employment of \texttt{struct} for the data storage does not 
provide 
improvements without using texture memory. However an 
interesting progress is detected for all the test cases at hand when use of 
\texttt{struct} and texture memory are combined for a fixed length of the 
histories. Moreover, on the same lead as for the Adjoin method, the 
effectiveness of all the modifications shades with a weight cutoff to shut down 
the random walks adaptively (compare results in Tables \ref{tab31} with the 
ones in Table \ref{tab27}).

\section{Use of L1 global cache}
Primary versions of NVIDIA graphic board were already garnered with L1 local 
cache memories. However they were not endowed with a similar memory to cache 
accesses to global memory which are known to cause the highest latency timings.
Using a caching system for such accesses can provide evident advantages, 
relatively to the particular application at hand and t its implementation.

Recent versions of NVIDIA GPUs enable the use of L1 cache memories.
Tesla K40m graphic boards, the ones installed in Titan, belong to this set of 
processors.

Even on Kepler GPUs that do support caching global memory in L1, the default 
behavior is to not cache global memory in L1. You have to opt-in to enable 
caching by passing \texttt{-Xptxas=-dlcm=ca} as an argument to NVCC when 
compiling your kernels. Kernels compiled in this fashion will still function on 
earlier Kepler devices though global memory operations will bypass the L1 cache.

You can programmatically query whether a GPU supports caching global 
memory operations using \texttt{cudaGetDeviceProperties} and examining the 
\texttt{globalL1CacheSupported} property.

In our code we have included the following commands in the main file in order 
to display the technical properties of GPUs at hand:

\hspace*{-2cm}
\begin{lstlisting}
  int nDevices;

  cudaGetDeviceCount(&nDevices);
  printf("The number of GPU devices is: %d\n", nDevices);
  for (int i = 0; i < nDevices; i++) {
    cudaDeviceProp prop;
    cudaGetDeviceProperties(&prop, i);
    printf("Device Number: %d\n", i);
    printf("  Device name: %s\n", prop.name);
    printf("  Memory Clock Rate (KHz): %d\n",
           prop.memoryClockRate);
    printf("  Memory Bus Width (bits): %d\n",
           prop.memoryBusWidth);
    printf("  Peak Memory Bandwidth (GB/s): %f\n\n",
           2.0*prop.memoryClockRate*(prop.memoryBusWidth/8)/1.0e6);
    printf("  globalL1CacheSupported:  %d\n", prop.globalL1CacheSupported);
    printf("  localL1CacheSupported:  %d\n", prop.localL1CacheSupported);
  }
\end{lstlisting}

One we have verified the presence of L1 Global caches in the NVDIA GPUs 
available, we can still customize even more the use for the resources at hand. 
In fact the L1 cache and the shared memory use a single memory for 
their purposes, splitting a common physical resources in two equal parts. In 
the case when either the shared memory or the L1 global cache are not used, it 
is pointless to keep occupied a part of memory for something tat will never be 
employed. Therefore it is possible to recalibrate the amount of Kbytes 
dedicated to both the kind of memories. 
This is possible through the use of two functions named 
\texttt{cudaDeviceSetCacheConfig} and \texttt{cudaFuncSetCacheConfig}. Their 
definition is included in the \texttt{cuda\_runtime\_api.h} header file.

The subroutine \texttt{cudaDeviceSetCacheConfig} requires only an argument, 
since its action affect the way all the subsequent kernels are executed. 
Instead \texttt{cudaFuncSetCacheConfig} requires two arguments. The first one 
is the name of the kernel to which it has to be applied, since it changes the 
handling of hardware resources between L1 cache and shared memory just for the 
specific kernel declared. 
The argument of \texttt{cudaDeviceSetCacheConfig} and the second one of 
\texttt{cudaFuncSetCacheConfig} can attain one of the following values:

\begin{itemize}
 \item \texttt{cudaFuncCachePreferShared}
 \item \texttt{cudaFuncCachePreferL1}
 \item \texttt{cudaFuncCachePreferEqual}
\end{itemize}

The third one corresponds to the default setting. The first one dedicate more 
memory cells to the shared memory, the second one instead shifts mos of it to 
the L1 cache use.\newline

It is possible to monitor the amount of resources used at runtime by passing 
some parameter to the NVIDIA profiler (NVPROF or NVVP). 
In the case of \texttt{nvprof} the command line looks like the one presented 
below here:
\hspace*{-2cm}
\begin{lstlisting}
 nvprof --metrics <type_of_counter> <executable_name> 
<arguments_for_executable>
\end{lstlisting}

The kind of counters that can be used as metric tools are:
\begin{itemize}
 \item \texttt{l1\_cache\_global\_hit\_rate} - percentage of reads that hit in 
L1 (Opt-In, K40/K80 only)
\item \texttt{tex\_cache\_hit\_rate} - percentage of reads that hit in texture 
cache (\_\_ldg() or texture access only) 
\item \texttt{l2\_l1\_read\_hit\_rate} - percentage of l1 misses which hit in L2
\item \texttt{l2\_texture\_read\_hit\_rate} - percentage of texture misses 
which hit in L2
\end{itemize}

All these counters provide you information regarding the use of hardware 
resources for each separate kernel employed in the code.

For our current interest we use \texttt{l1\_cache\_global\_hit\_rate}. In the 
following treatise we present timings for the most important setting described 
before (for Adjoint and Forward MC), enabling the use of L1 global cache. 
Percentage of such resource used by the kernels are provided too. However, 
since 
the profiling affect the performance of the code, each configuration has been 
run twice for the sake of validity of the results presented: once without and 
once with the profiler. \newline

As regards the Adjoint method, enabling L1 global cache does not provide 
evident improvements, as it can be verified with results in Tables 
\ref{tab32}-\ref{tab37}. Sporadic time reductions occur for all the test cases 
considered but it is hard to find a correlation between the role played by the 
global cache memory and the other factors of the setting. In particular it is 
not possible to declare that such reductions are due to the global cache, since 
in most of the cases the profiler prompts out that no cache resources 
are actually employed.

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & tot. time (ms)& \% rng - 
\% mc & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 167,270 & 0 - 0 & 0.874 \\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ &  36,854 & 0 - 0 & 1.021\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 573,288 & 0 - 0 & 1.075\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 306,248 & 0 - 0 & 1.268\\
\hline
\end{tabular}
\caption{Adjoint method - Use of L1 global cache. Fixed number of steps equal 
to 10,000. No LDG instructions. Relative Timings compared with Table 
\ref{tab4}.} 
\label{tab32}
\end{table}

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & tot. time (ms)& \% rng - 
\% mc & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 32,460 & 0 - 0 & 1.172 \\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ &  38,002 & 0 - 0 & 1.082\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 40,910 & 0 - 0 & 1.005\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 6,781 & 0 - 0 & 1.379\\
\hline
\end{tabular}
\caption{Adjoint method - Use of L1 global cache. Weight cutoff equal to 
$10^{-9}$. No LDG instructions. Relative Timings compared with Table 
\ref{tab11}.} 
\label{tab33}
\end{table}

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & tot. time (ms)& \% rng - 
\% mc & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 101,068 & 0 - 0 & 0.960 \\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ &  36,776 & 0 - 0 & 1.062\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 574,863 & 0 - 0 & 1.104\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 118,604 & 0 - 0 & 0.859\\
\hline
\end{tabular}
\caption{Adjoint method - Use of L1 global cache. Fixed number of steps equal 
to 10,000. LDG instructions employed. Relative Timings compared with Table 
\ref{tab10}.} 
\label{tab34}
\end{table}

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & tot. time (ms)& \% rng - 
\% mc & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 28,222 & 0 - 0 & 1.161 \\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ &  36,042 & 0 - 0 & 0.975\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 41,473 & 0 - 0 & 1.022\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 4,644 & 0 - 0 & 0.944\\
\hline
\end{tabular}
\caption{Adjoint method - Use of L1 global cache. Weight cutoff equal to 
$10^{-9}$. LDG instructions employed. Relative Timings compared with Table 
\ref{tab17}.} 
\label{tab35}
\end{table}

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & tot. time (ms)& \% rng - 
\% mc & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 101,004 & 0 - 0 & 0.961 \\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ &  27,351 & 0 - 21.37 & 0.998\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 413,696 & 0 - 0 & 0.980\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 133,346 & 0 - 0 & 0.96\\
\hline
\end{tabular}
\caption{Adjoint method - Use of \texttt{struct} for data handling. Use of L1 
global cache. Fixed number of steps equal 
to 10,000. LDG instructions employed. Relative Timings compared with Table 
\ref{tab20}.} 
\label{tab36}
\end{table}

\begin{table}[!h]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist.} & tot. time (ms)& \% rng - 
\% mc & Rel. Timing\\
\hline
2D Laplacian& 900 & $10^7$ & 28,986 & 0 - 74.29 & 1.175 \\
\hline 
1D shift Laplacian& $10^6$ & $10^4$ &  26,838 & 0 - 0 & 0.997\\
\hline
1D shift Laplacian& $10^6$ & $10^7$ & 30,671 & 0 - 0 & 0.997\\
\hline
$\text{SP}_1$ & 25,568 & $10^7$ & 4,886 & 0 - 0 & 0.976\\
\hline
\end{tabular}
\caption{Adjoint method - Use of \texttt{struct} for data handling. Use 
of L1 global cache. Weight 
cutoff equal to $10^{-9}$. Relative Timings compared with Table 
\ref{tab21}.} 
\label{tab37}
\end{table}

Better results are obtained by activating L1 cache for the Forward method. In 
particular the combination of the use of \texttt{struct} and L1 global cache 
provides significant improvement of timings and cache resources are employed 
almost to the full for all the test cases considered. This phenomenon can be 
explained with the fact that the \texttt{struct} enables to reorder data so 
that values accessed in consecutive moments in time are physically close in the 
memory storage. Therefore caching information before using them helps in 
reducing the runtime effectively.

A possible explanation for such discrepancy between 
Adjoint and Forward method is associated with the policy to estimate single 
entries of the solution vector. In fact the Forward method is characterized by 
a one-to-one relationship between the row of the iteration matrix and 
the entry for which such row provides contributions. It entails that in the 
Forward method consecutive states of the same random walk visit entries that 
are likely close in the compressed storage of the matrix.

The same thing does not necessarily apply for the Adjoint method, since from 
one state to the consecutive one the row visited by the algorithm changes as 
well. Therefore, it is very likely that consecutive steps visit entries far one 
from the other in the matrix storage. L1 cache hardly provides benefits for 
time reduction in situations like these.

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & tot. time 
(ms) & \% rng - 
\% mc  & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 74,274 & 0 - 0 & 0.937\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 82,440 & 0 - 0 & 1.105\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 197,493 & 0 - 0 & 1.609\\
\hline
\end{tabular}
\caption{Forward method. Fixed number of steps equal to 10,000. No LDG 
instructions. Relative Timings compared with Table 
\ref{tab23}.}
\label{tab38}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & tot. time 
(ms) & \% rng - 
\% mc  & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 9,716 & 0 - 0 & 0.840\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 36,040 & 0 - 0 & 0.936\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 8,189 & 0 - 0 & 1.259\\
\hline
\end{tabular}
\caption{Forward method. Weight cutoff equal to $10^{-9}$. No LDG 
instructions. Relative Timings compared with Table 
\ref{tab25}.}
\label{tab39}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & tot. time 
(ms) & \% rng - 
\% mc  & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 44,134 & 0 - 0 & 0.847\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 52,816 & 0 - 0 & 0.834\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 74,393 & 0 - 0 & 1.030\\
\hline
\end{tabular}
\caption{Forward method. Fixed number of steps equal to 10,000. LDG 
instructions employed. Relative Timings compared with Table 
\ref{tab26}.}
\label{tab40}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & tot. time 
(ms) & \% rng - 
\% mc  & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 44,134 & 0 - 0 & 0.847\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 52,816 & 0 - 0 & 0.834\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 74,393 & 0 - 0 & 1.030\\
\hline
\end{tabular}
\caption{Forward method. Weight cutoff equal to $10^{-9}$. LDG instructions 
employed. Relative Timings compared with Table 
\ref{tab27}.}
\label{tab41}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & tot. time 
(ms) & \% rng - 
\% mc  & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 42,610 & 0 - 0 & 0.849\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 58,473 & 0 - 0 & 0.860\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 83,258 & 0 - 0 & 1.106\\
\hline
\end{tabular}
\caption{Forward method - Use of \texttt{struct} for data handling. Fixed 
number of steps equal 
to 10,000. LDG instructions employed. Relative Timings compared with Table 
\ref{tab30}.}
\label{tab42}
\end{table}

\begin{table}[!h]
\hspace*{-1cm}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Matrix} & \textbf{Size} &\textbf{Nb. Hist. per entry} & tot. time 
(ms) & \% rng - 
\% mc  & Rel. Timing\\
\hline
2D Laplacian& $900$ & $10^3$ & 7,713 & 0 - 94.16 & 0.862\\
\hline 
1D shift Laplacian& $10^6$ & $5$ & 29,512 & 0 - 82.14 & 0.974\\
\hline
$\text{SP}_1$ & $25,568$ & $200$ & 3,782 & 0 - 42.40 & 0.936\\
\hline
\end{tabular}
\caption{Forward method - Use of \texttt{struct} for data handling. Weight 
cutoff equal to $10^{-9}$. LDG instructions employed. Relative Timings compared 
with Table 
\ref{tab31}.}
\label{tab43}
\end{table}

\section{Adjoint method - Precomputation and sorting of the initial states}
In this section we try to reschedule in a different fashion the computation of 
the initial states to kick off the random walks. In particular we adopt a 
precomputing approach. In algorithms, precomputation is the act of performing an 
initial computation before run time to generate a lookup table that can be used 
by an algorithm to avoid repeated computation each time it is executed.

For this we just consider the Adjoint method, since in the Forward the initial 
state is determined automatically by the index of the entry estimated. \newline

In particular, instead of generating the random numbers one at a time to 
figure out the initial steps, we compute all of them through a separate GPU 
kernel. Once this done, we reorganize the random numbers in an increasing order 
so that the histories are reordered accordingly to the index of their 
initial state.

This approach is attempted to maximize the locality of data stored temporarily 
on local memories (e.g. L1 cahes, texture memory), with a related reduction of 
the runtime hopefully.

In order to manage the two different policies to estimate the initial states we 
introduced a templatization on the kernel with respect to a class which handles 
such task.

The class \texttt{OnTheFly} is employed to execute the standard algorithm.

\hspace*{-2cm}
\begin{lstlisting}
class OnTheFly
{

public: 
       OnTheFly(curandState*, unsigned int, unsigned int){};
        __device__ inline double get(curandState* rng_state)
       {
              double rand = curand_uniform_double(rng_state);
              return rand;
       };
       inline void free_data(){};
};

\end{lstlisting}

The \texttt{Precomputed} class instead manages the computation of all the 
initial states in advance, before computing actual histories.

\hspace*{-2cm}
\begin{lstlisting}
class Precomputed
{

private:
        bool computed = false;
        double* starting_states;
public:
        inline Precomputed(curandState*, unsigned int, unsigned int);
        __device__ inline double get(curandState*)
        {
          if( computed == false )
            return -1;
          unsigned int tid = threadIdx.x + blockIdx.x * blockDim.x;
          return starting_states[tid];
        };
        inline void free_data();
};

inline Precomputed::Precomputed(curandState* state,
        unsigned int num_blocks, unsigned int block_size):computed(true)
{
        cudaError e = cudaMalloc( (void **)&starting_states,
           block_size * num_blocks * sizeof(double) );
        if( cudaSuccess != e )
           std::cout << "Cuda Error: " << cudaGetErrorString(e) << std::endl;

        thrust::device_ptr< double > dev_ptr( starting_states );

        VALIDATE(cudaSuccess==e,"Failed to allocate memory");
        initial_state<<<num_blocks, block_size>>>(state, starting_states);

        //cudaDeviceSynchronize();
        thrust::sort( dev_ptr, dev_ptr + (block_size * num_blocks) );
}

\end{lstlisting}

We accomplished a comparison between the times employed by the standard 
generation of random numbers and their scheduled precomputation. All 
the simulations have been run by using a fixed number of steps, setting such 
parameter to different values (10 and 1000).

\begin{table}[!h]
\hspace*{1.5cm}
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Approach} & \textbf{\# histories} &\textbf{\# steps} & \textbf{batch 
size} 
& \textbf{Time (ms)} \\
\hline
OnTheFly& $10^8$ & $10^3$ & - & 434,003 \\
\hline 
Precomputed& $10^8$ & $10^3$ & $10^8$ & 244,099\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^7$ & 64,822\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^6$ & 75,976\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^5$ & 261,381\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^4$ & 354,505\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^3$ & 1,191,359\\
\hline
\end{tabular}
\caption{Adjoint method - OnTheFly vs. Precomputed approach. 1D shifted 
Laplacian with $10^3$ steps per history.} 
\label{tab44}
\end{table}

\begin{table}[!h]
\hspace*{1.5cm}
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Approach} & \textbf{\# histories} &\textbf{\# steps} & \textbf{batch 
size} 
& \textbf{Time (ms)} \\
\hline
OnTheFly& $10^8$ & $10$ & - & 32,686 \\
\hline 
Precomputed& $10^8$ & $10$ & $10^8$ & 40,918\\
\hline
Precomputed& $10^8$ & $10$ & $10^7$ & 28,670\\
\hline
Precomputed& $10^8$ & $10$ & $10^6$ & 28,237\\
\hline
Precomputed& $10^8$ & $10$ & $10^5$ & 31,478\\
\hline
Precomputed& $10^8$ & $10$ & $10^4$ & 41,876\\
\hline
Precomputed& $10^8$ & $10$ & $10^3$ & 226,996\\
\hline
\end{tabular}
\caption{Adjoint method - OnTheFly vs. Precomputed approach. 1D shifted 
Laplacian with $10$ steps per history.} 
\label{tab45}
\end{table}

\begin{table}[!h]
\hspace*{1.5cm}
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Approach} & \textbf{\# histories} &\textbf{\# steps} & \textbf{batch 
size} 
& \textbf{Time (ms)} \\
\hline
OnTheFly& $10^8$ & $10^3$ & - & 183,837 \\
\hline 
Precomputed& $10^8$ & $10^3$ & $10^8$ & 674,372\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^7$ & 186,705\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^6$ & 138,801\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^5$ & 146,899\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^4$ & 168,424\\
\hline
Precomputed& $10^8$ & $10^3$ & $10^3$ & 1,368,421\\
\hline
\end{tabular}
\caption{Adjoint method - OnTheFly vs. Precomputed approach. $SP_1$ problem 
with $10^3$ steps per history.} 
\label{tab46}
\end{table}

\begin{table}[!h]
\hspace*{1.5cm}
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Approach} & \textbf{\# histories} &\textbf{\# steps} & \textbf{batch 
size} 
& \textbf{Time (ms)} \\
\hline
OnTheFly& $10^8$ & $10$ & - & 3,435 \\
\hline 
Precomputed& $10^8$ & $10$ & $10^8$ & 1,092,207\\
\hline
Precomputed& $10^8$ & $10$ & $10^7$ & 123,578\\
\hline
Precomputed& $10^8$ & $10$ & $10^6$ & 13,546\\
\hline
Precomputed& $10^8$ & $10$ & $10^5$ & 7,404\\
\hline
Precomputed& $10^8$ & $10$ & $10^4$ & 10,803\\
\hline
Precomputed& $10^8$ & $10$ & $10^3$ & 156,745\\
\hline
\end{tabular}
\caption{Adjoint method - OnTheFly vs. Precomputed approach. $SP_1$ problem 
with $10$ steps per history.} 
\label{tab47}
\end{table}

A graphic representation of the results in Tables (\ref{tab44}-\ref{tab47}) is 
represented in Figures \textcolor{red}{reference to the pictures}.

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{1d_lap_onthefly_vs_precompute.png}
    \caption{Adjoint method - OnTheFly vs. Precomputed approach. 1D shifted 
Laplacian.}
\label{1d_lap}
\end{figure}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{c5g7_onthefly_vs_precompute.png}
    \caption{Adjoint method - OnTheFly vs. Precomputed approach. $SP_1$ 
problem.}
\label{1d_lap}
\end{figure}

As you can notice, for the 1D shifted laplacian (Figure 
\textcolor{red}{figura}) a significant drop of the timing is achieved when the 
size of the batch assumes intermediate values with respect to the size of the 
problem ($10^6$). However, when these two values assume a comparable order of 
magnitude, a worsening of the performance is detected. Such behavior can be 
explained with the significant augmentation of atomic operation occurring. In 
fact the decrease of the batch size increases also the number of random walks 
starting from the same state, implying their request to modify the same value in 
the memory.

As regards the $SP_1$ problem the situation is more problematic. In fact, 
regardless of the batch size, the Precomputed approach never succeeds in 
outperforming the standard way of proceeding. Moreover, an increase of the time 
along with the increase of the batch size is verified in this case as well. 
\newline

We employed the \texttt{NVIDIA} profiler to validate the theory that attributes 
the increase of the time to the enhancement of atomic operations.
A standard employment of such tool already provided us with useful information 
regarding the portion of the algorithm employing most of the time. In 
particular the 99\% of the time is spent to execute the 
\texttt{run\_adjoint\_monte\_carlo} kernel. This is a positive feedback, since 
it shows that almost the entire runtime is spent for the actual 
computation rather than for preprocessing or postprocessing operations.

Once this verified, a more detailed analysis of the performance is required to 
reach the goal of the validation. In particular, since we want to focus on 
profiling just the \texttt{run\_adjoint\_monte\_carlo} kernel, we can employ 
some \texttt{CUDA} subroutine to restrict the area profiled.

To limit profiling to a region of your application, CUDA provides functions to 
start and stop profile data collection. \texttt{cudaProfilerStart()} is used to 
start profiling and \texttt{cudaProfilerStop()} is used to stop profiling. In 
order to make such calls effective, it is necessary to employ the metric 
\texttt{--profile-from-start off} when the command line profiler 
\texttt{nvprof} is called. If instead the visual profiler \texttt{nvvp} is 
employed, the previous calls must be combined by disabling the profiling from 
the start during the creation of the new session. \newline

The previous subroutines have been combined with the use of the compiling flag 
\texttt{-lineinfo} to generate line number information for applications 
without affecting the optimization level of the output. Differently from the 
standard debugging flag \texttt{-G}, \texttt{-lineinfo} does not force the 
compiler to generate debug information for the CUDA application.

\end{document}

### 2015-08-21 MAX_reports/MC_solvers_article/MC_solvers.tex
tex files included for the reports

\documentclass[a4paper,10pt]{article}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{mathtools}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{geometry}
\usepackage{subfig}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabu}
\usepackage{multirow}
\usepackage{color}
\usepackage{listings}
\usepackage[]{algorithm2e}
\usepackage[autostyle,english=british]{csquotes}
\usepackage{minitoc}

\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem{prop}{Proposition}

%opening
\title{Monte Carlo solvers for sparse linear systems}
\author{Michele Benzi, Massimiliano Lupo Pasini}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We consider linear systems of equations characterized by a sparsity pattern, 
focusing in particular on problems derived from the discretization of partial 
differential equations. We propose a treatise of the current stochastic 
techniques present in literature that might be employed to solve the problems at 
hand. New theoretical results are introduced as sufficient criteria for the 
guarantee of the convergence of the methods analyzed. An analysis about viable 
preconditioners is accomplished as well. 
\end{abstract}

\section{Introduction}
For a long period the  
pursuit of a reliable solver for linear and nonlinear algebraic systems has 
been one of the most important issues in Numerical Analaysis. The 
mathematical properties of the model (e.g. likely sparsity pattern of the 
matrix) have urged scientists to look for efficient and pattern-driven 
solvers, capable to cope with curse of dimensionality and high scalability 
requests.

As scalability computing move towards exascale facilities (which stands for 
machines computing up to $10^{18}$ flops), new numerical schemes suitable for 
this 
kind of architectures are strongly demanded. In fact the purpose is to combine 
high 
fidelity of the results with an optimized use of hardware resources at hand.

Recently new algorithms that combine statistical and numerical techniques have 
been developed. Even though \textit{numerical efficiency} has been sacrificed 
sometimes, the advantages in terms of \textit{robustness} of these methods make 
it worth carrying on studies in this direction.
A class of methods that may be employed for this purpose is represented by 
Monte Carlo linear solvers.

These algorithms are the main topic covered in the work presented here, 
starting from what was developed so far by some of 
the authors who gave significant contributions (see \cite{Hal1962}, 
\cite{Hal1994}, 
\cite{DA1998}, \cite{DVA2001}, \cite{AADBTW2005},\cite{ESW2013} and 
\cite{EMSH2014}). \newline

As underlined in \cite{DA1998}, Monte Carlo methods may be split into two 
classes: \textit{direct methods} (\cite{DA1998}, \cite{DVA2001}) and 
\textit{iterative methods} (\cite{Hal1962}, 
\cite{Hal1994}, \cite{ESW2013} 
and \cite{EMSH2014}). The first are characterized by a merely stochastic 
scheme, 
therefore the provided error with respect to the exact solution is made of 
just a stochastic component. The second ones, instead, combine numerical and 
statistical schemes generating two types of error: a 
\textit{stochastic} one and a \textit{systematic} one. It does not 
mean 
that 
it will be always simple to recognize them separately. However it is important 
to 
be 
aware of this intrinsic structure, in order to target what is the part of the 
scheme that requires a refinement (e.g. increase of numerical iterations 
rather than random walks).\newline

The paper is organized as follows. In Section 2 we describe all the 
viable stochastic settings for the definition of the Markovian process. In 
Section 3 we reinterpret probabilistic facts in terms of spectral properties of 
the linear system to be solved. Section 4 is dedicated to necessary conditions 
and sufficient conditions for the convergence of the algorithms. In Section 5 
different adaptive approaches are presented, in order to truncate the random 
walks and to decide how many samplings to use. Section 6 regards the study 
of matrices for which the convergence of the stochastic schemes is facilitated. 
Then in Section 7 a hybrid scheme combining numerical and probabilistic 
techniques is described: the \textit{Monte Carlo Synthetic Acceleration}. 
Numerical results are presented in Section 8 and 9, while Section 10 is 
dedicated to 
conclusive remarks and goals for the future.

\section{Stochastic setting}

We start our treatise by introducing the direct methods.\newline

Assume to start from a linear system such as 
\begin{equation}
A \mathbf{x}=\mathbf{b},
\label{linsys}
\end{equation}
where $A\in \mathbb{R}^{n\times n}$, $\mathbf{x}$, $\mathbf{b} \in 
\mathbb{R}^n$.

By applying a left preconditioning, the system (\ref{linsys}) gets the 
following shape:
\begin{equation}
P^{-1}A \mathbf{x}=P^{-1}\mathbf{b},
\label{linsys_prec}
\end{equation}
(\ref{linsys_prec}) can be 
reinterpreted as a fixed point scheme
\begin{equation}
 \mathbf{x}=H\mathbf{x}+\mathbf{f}.
 \label{fixedpoint}
\end{equation}
where $H=I-P^{-1}A$ and $\mathbf{f}=P^{-1}\mathbf{b}$.

Assuming that the spectral radius $\rho(H)<1$, the solution to 
(\ref{fixedpoint}) can be written in terms of a power series of 
$H$ (Neumann series)
\[
\mathbf{x}=\sum_{i=0}^\infty H^i\mathbf{f}.
\]
Therefore the fixed point scheme generates a sequence of approximate solutions
$\{\mathbf{x}^{(k)}\}_{k=0}^{\infty}$ which converges to the exact solution 
regardless of the initial guess $\mathbf{x}_0$.

By restricting the attention to a single component of $\mathbf{x}$ we 
have
\begin{equation}
x_i=\sum_{\ell=0}^\infty \sum_{k_1}^n\sum_{k_2}^n\cdots \sum_{k_{\ell}=1}^n 
H_{k_0,k_1}H_{k_1,k_2}\cdots H_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{forward}
\end{equation}
The last equation can be reinterpreted as the realization of an estimator 
defined on a random walk.

Let us start considering a random walk whose 
state space $S$ is characterized by the set of indices of the forcing term 
$\mathbf{f}$: 
\[
S=\{1,2,\cdots, n\} \subset \mathbb{N}.
\]
Each $i$-th step of the random walk has a random variable 
$k_i$ associated with it. The realization of $k_i$ represents the index of the 
component of $\mathbf{f}$ 
which is visited in the current step of the random walk.

The way the transition probability is built and the way the initial state of 
the random walk is chosen give birth to two different approaches we are 
going to use: the \textit{Forward} and \textit{Adjoint} methods.

\subsection{Forward Method}
The goal is to evaluate a functional such as 
\[
J(\mathbf{x})=(\mathbf{h},\mathbf{x})=\sum_{i=1}^n h_i x_i.
\]
where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in 
$\mathbb{R}^n$ of the functional $J$.
We can use it to build the initial probability $\tilde{p}: 
S\rightarrow [0,1]$ of the random walk such that
\[
\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n \lvert 
h_i\rvert}.
\]
It is important to stress out that the role of vector $\mathbf{h}$ is 
restricted to the construction of the initial probability. Once that is done, 
it is 
not used anymore in the definition of the stochastic process. 
A possible choice for the transition probability $P$ can be
\[
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert H_{i,j}\rvert}{\sum_{k=1}^n 
\lvert H_{i,k}\rvert}.
\]
where $\tilde{p}(\cdot,i):S\rightarrow [0,1]$ $\forall i\in S$.
A related sequence of random variables $w_{i,j}$ can be defined 
such that 
\[
w_{i,j}=\frac{H_{i,j}}{P_{i,j}}.
\]
The probability distribution of the random variables $w_{i,j}$ is represented 
by the transition matrix that rules the stochastic process. The $w_{i,j}$'s 
just introduced can be used to build one more sequence 
of random variables. 
At first we introduce quantities $W_j$ 
\[
W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_j=W_{j-1} w_{i,j}, \quad 
j=1,\cdots, i.
\]
By defining 
\[
X(\nu)=\sum_{m=0}^k W_m f_{i_m}
\]
as the random variable associated with a specific permutation $\nu$, we can 
define the estimator $\theta_i$ such as
\[
\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu).
\]
The integer $n$ represents the size of the solution vector to (\ref{linsys}) 
and 
the
index 
$i$ 
is referred to the component of the solution vector we want to compute. 
$P_{\nu}$ is the probability associated with a specific permutation of the 
random walk.  
It can be proved that 
\[
E[W_i f_{k_i}]=(\mathbf{h},H^i\mathbf{f}), \quad i=0,1,2,\cdots
\]
and 
\[
\theta_i=E\bigg[\sum_{i=0}^\infty W_i f_{k_i}\bigg]=(\mathbf{h},\mathbf{x}).
\]

A possible choice for $\mathbf{h}$ is a vector of the standard basis. This 
would correspond in setting manually the initial state of the random walk, 
which turns the related initial probability into a Kronecker delta
\[
\tilde{p}(k_0=i)=\delta_{i,j}.
\]
By doing so, we have 
\begin{equation}
\theta_i=E\bigg[\sum_{\ell=0}^\infty W_{\ell} 
f_{k_{\ell}}\bigg]=x_i=\sum_{l=0}^\infty 
\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n 
P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1}, 
k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots 
w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{dir_mean}
\end{equation}

As regards the variance, we remember that the following relation holds:
\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell} 
f_{k_{\ell}}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2 
f_{k_{\ell}}^2\bigg] - \bigg (E\bigg[\sum_{\ell=0}^\infty W_{\ell} 
f_{k_{\ell}}\bigg]\bigg )^2.
\label{dir_var}
\end{equation}

Hence the variance can be computed as the difference between the second 
moment of the random variable and the square of the first moment.\newline

In order to apply the Central Limit Theorem (CLT) to the estimators defined 
above, we must require that 
the estimators have both finite expected value and variance. This is 
equivalent in 
checking the finiteness of the expected value and of the second moment.
Therefore we have to impose the following conditions:

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell} f_{k_{\ell}}\bigg]<\infty
\end{equation}

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2 
f_{k_{\ell}}^2\bigg]<\infty
\end{equation}

The Forward method presented above, however, has the limitation of employing an 
entire set of permutations to estimate just a single entry of 
the solution at a time. Therefore, in order to estimate the entire solution 
vector for \ref{linsys_prec}, we have to employ as many sets of permutation as 
the number of entries composing the vector. Of course this might turn to be 
extremely expensive both in terms of time and resources for most of the 
challenging problems coming from applications. This is essentially the reason 
why usually it is preferred to adopt another stochastic algorithm which is 
explained here following: the \textit{Adjoint} Monte Carlo method.

\subsection{Adjoint Method}
A second Monte Carlo method can be derived by considering the linear system 
adjoint to (\ref{linsys_prec})
\begin{equation}
(P^{-1}A)^T\mathbf{y}=\mathbf{d},
\label{adj}
\end{equation}
where $\mathbf{y}$ and $\mathbf{d}$ are the adjoint solution and source term.

By reformulating the fixed point scheme, introducing initial probability, 
transition probability and weights 
sequences in the same fashion as done before, the expected value for the 
estimator becomes
\begin{equation}
\theta_j=E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell}, 
j}\bigg]=\sum_{\ell=0}^{\infty}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n 
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots 
w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}.
\label{adj_mean}
\end{equation}
This estimator is known in literature as \textit{collision} estimator.

We can notice now that the component of the source vector, in each 
permutation of the random walk, is associated with the initial step. The 
Kronecker delta 
at the end of the series stands for a filter. It means that if the goal is to 
estimate the $j$-th component of the solution vector, just a subset of all the 
permutations gives a contribution to this. The only permutations considered are 
those ones that currently reside on the index associated with the entry to be 
estimated.

The variance is

\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell} 
f_{k_0}\delta_{k_{\ell},j}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2 
f_{k_0}^2\delta_{k_{\ell},j}\bigg ] - \bigg (E\bigg[\sum_{\ell=0}^\infty 
W_{\ell} 
f_{k_0}\delta_{k_{\ell},j}\bigg]\bigg )^2\quad j=1,\cdots,n
\label{adj_var}.
\end{equation}

On the same lead of what done for the Forward method, there is need to impose 
finiteness of the expected value and second moment. Therefore the following 
conditions must be verified:

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell}, 
j}\bigg]<\infty \quad j=1,\cdots,n
\end{equation}
and
\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2 
f_{k_0}^2\delta_{k_{\ell},j}\bigg]<\infty \quad j=1,\cdots,n.
\end{equation}

The main advantage of this method, compared to the Forward one, consists in the 
fact that a single set of permutation is used to estimate the entire solution 
vector. Therefore, in general, this algorithm has to be preferred to the 
previous one since it enables to reduce dramatically the computational burden.

In literature another estimator is employed along with the Adjoint Monte Carlo
method, the so called \textit{expected value} estimator. Its 
formulations is as follows:

\begin{equation}
\theta_j=E\bigg[f_j + \sum_{\ell=0}^\infty 
W_{\ell}H_{k_{\ell}, j}^T\bigg]=f_j 
+ \sum_{\ell=0}^{\infty}\sum_{k_1}^n\sum_{k_2} ^n\cdots\sum_ { k_ { \ell}}^n 
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots 
w_{k_{\ell-1},k_{\ell}}H_{k_{\ell},j}^T.
\label{adj_mean1}
\end{equation}

Differently from the collision estimator, the expected value estimator averages 
the deterministic contribution of the iteration matrix over all the potential 
states $j$ that might be reached from the current state $\ell$. The variance 
in this case becomes:

\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell} 
f_{k_0}H_{k_{\ell},j}^T\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2 
f_{k_0}^2 {H_{k_{\ell},j}^T}^T\bigg ] - \bigg (E\bigg[\sum_{\ell=0}^\infty 
W_{\ell} 
f_{k_0}H_{k_{\ell},j}^T\bigg]\bigg )^2\quad j=1,\cdots,n
\label{adj_var}.
\end{equation}

In the sections dedicated to the numerical tests, both the collision estimator 
and the expected value are employed, in order to test their effectiveness.

\section{Numerical interpretation of the statistical requirements}
The requests imposed on the expected value and second moment of the estimators 
can be reformulated in a purely deterministic setting by looking at the 
spectral radii of some properly defined matrices.

For instance the condition of finiteness of the expected value can be 
reformulated by requiring 
\begin{equation}
 \rho(H)<1
 \label{rho_h},
\end{equation}

where $H$ is the iteration matrix of the fixed point scheme.
In fact both equations (\ref{dir_mean}) and (\ref{adj_mean}) can be considered 
as 
power series in terms of $H$ and the spectral radius of $H$ smaller 
than 1 is a necessary and sufficient condition for the Neumann series to 
converge.

By analogy we aim at reproducing a similar reasoning for the second moment of 
the estimator, since we want it to be finite as well. 
By looking at the equations of the variance (\ref{dir_var}) and (\ref{adj_var}) 
for 
the Forward and the Adjoint method respectively, we notice that the second 
moment can be reinterpreted as a power series with respect to the matrices
defined as follows:

\[
 \hat{H}_{i,j}=\frac{H_{i,j}^2}{P_{i,j}} \quad \text{- Forward Method}.
\]

or

\[
 \hat{H}_{i,j}=\frac{H_{j,i}^2}{P_{i,j}} \quad \text{- Adjoint Method}.
\]

Accordingly to the necessary and sufficient condition for the power series to 
converge, we require 
\begin{equation}
 \rho(\hat{H})<1.
 \label{rho_star}
\end{equation}

Condition (\ref{rho_h}) must be required for a generic fixed point scheme to 
reach convergence, whereas the extra condition (\ref{rho_star}) is typical of 
the stochastic schemes presented in this work. Having two conditions to 
verify makes the numerical treatment of the linear system via MC very 
difficult. We will discuss this in details later in the treatise. \newline

\section{Necessary conditions, sufficient conditions and choice of the 
transition matrix}

Here below some results presented in 
\cite{MASC2013} about necessary conditions and sufficient 
conditions for 
convergence are enunciated. In particular a suitable choice for constructing a 
transition probability is discussed to guarantee and speed-up convergence.

We remember that the construction of the transition probability must satisfy 
some constraints such as

\[
\begin{cases}
  P_{i,j}\ge 0 \\
 \sum_{j=1}^N P_{i,j}=1
\end{cases}
\]

plus one more restriction which varies depending on the particular Monte Carlo 
approach chosen

\begin{align*}
& \text{Forward Method:} \quad H_{i,j}\ne 0 \Rightarrow P_{i,j}\ne 0 \\ &
\text{Adjoint Method:} \quad H_{j,i}\ne 0 \Rightarrow P_{i,j}\ne 0 
\end{align*}.

The previous restrictions on the attainable values for the entries of the 
transition probability are called \textit{transition conditions}.

At first an introductory theoretical result is shown, necessary to motivate the
choices made afterwards for the choice of the transition probability.

\begin{thm}
 Consider a vector $a=(a_1,a_2,\cdots,a_N)^T$ where at least one element is 
nonzero, $a_k\ne0$ for some $k\in\{1,\cdots,N\}$.
\begin{itemize}
 \item For a probability vector $p=(p_1,p_2,\cdots,p_N)^T$ satisfying the 
transition conditions, the lower bound of $\displaystyle 
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is $\bigg(\sum_{k=1}^N \lvert a_k\rvert\bigg)^2$
\item There always exists a probability vector such that $\displaystyle 
\sum_{k=0}^N\frac{a_k^2}{p_k}\ge c\ge 1$ for all $c>1$.
\end{itemize}
\label{lemma}
\end{thm}

It might be observed that the minimum of the quantity $\displaystyle 
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is attained when the probability vector is 
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert 
a_k\rvert}$. It implies that the $\infty$-norm of $\hat{H}$ is minimized as 
much 
as possible. \newline

For the sake of simplicity let us introduce now a generic random walk
truncated at a certain $k$-th step
\[
 \gamma_k:\; r_0\rightarrow r_1 \rightarrow r_2 \rightarrow \cdots \rightarrow 
r_k
\]
and the statistical estimator
\[
 X(\gamma_k)=\frac{H_{r_0,r_1}H_{r_1,r_2}\cdots 
H_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots P_{r_{k-1},r_k}}f_{r_k}
\]
which is associated with the Forward Method.

Then the following theorem holds
\begin{thm}\textit{(Suited to the Forward Method)}

Let $H\in \mathbb{R}^{n\times n}$ such that $\lVert H\rVert_{\infty}<1$. 
Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the 
$k$-th step. Then, 
there always exists a 
transition matrix $P$ such that 
$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and 
$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
\label{for_thm}
\end{thm}

If we introduce the estimator
\[
 X(\gamma_k)=\frac{H^T_{r_0,r_1}H^T_{r_1,r_2}\cdots 
H^T_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots 
P_{r_{k-1},r_k}}sign(f_{r_0})\lVert \mathbf{f}\rVert_1
\]
which is associated with the Adjoint Method, then we can state a 
theorem specular to \ref{for_thm}.

\begin{thm}\textit{(Suited to the Adjoint Method)}

 Let $H\in \mathbb{R}^{n\times n}$ with $\lVert H\rVert_{1}<1$. 
Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the 
$k$-th step. Then, 
there always exists a 
transition matrix $P$ such that 
$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and 
$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
\label{adj_thm}
\end{thm}

In particular, relying on what guaranteed by Theorem \ref{lemma}, in both 
Theorems \ref{for_thm} and \ref{adj_thm} the thesis holds by picking the MAO 
transition probability.

These theoretical results represent sufficient conditions for the convergence 
of the  
Forward and Adjoint Monte Carlo and they can be easily checked. 
However requiring $\lVert 
H\rVert_{\infty}<1$ or $\lVert H\rVert_1<1$ may be too demanding. 
Indeed, even if there exist some preconditioners able to guarantee this 
condition (e.g. approximate inverse preconditioners with a proper value of 
dropping 
tolerance), it may entail to compromise the sparsity of the 
iteration 
matrix.

Thusly, in situations where the hypotheses of Theorems \ref{for_thm} and 
\ref{adj_thm} are too demanding to be verified, the necessary and sufficient 
condition 
on the 
spectral radius of $\hat{H}$ is the only criterion to take as reference.
Nonetheless this constraint consists in 
computing the spectral radii of two matrices and the calculation 
might be very costly. Moreover the computation of the spectral radii is 
affected by resilience issues as well as 
a deterministic solver for the resolution of 
a linear system. Therefore the original difficulty is transferred without being 
actually tackled.

\section{Construction of the transition probability}
The way the transition probability is defined is extremely important. In fact 
in some circumstances it might determine whether the stochastic scheme 
converges or not. Two very simple ways of defining it are listed here 
below. These are the approaches adopted typically in literature.

\subsection*{Uniform probabilities}
With this approach transition matrix $P$ is such that all the non-zero elements 
in each row have equal probability of occurring. 
\[
\text{Forward}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{i,j}=0 \\
\frac{1}{\#(\text{non-zeros in the row)}} \quad \text{if} \quad H_{i,j}\ne 0
\end{cases}\quad 
\]
\[
\text{Adjoint}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{j,i}=0 \\
\frac{1}{\#(\text{non-zeros in the column)}} \quad \text{if} \quad H_{j,i}\ne 0
\end{cases}
\]
The Monte Carlo approach resorting to this definition of the transition matrix, 
in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).

\subsection*{Weighted probabilities}
An alternative way of defining the transition matrix consists in attributing 
probabilities to all the entries of the transition probability accordingly 
to the magnitude of the elements. For instance we may employ the following 
definition
\[
\text{Forward}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert 
H_{i,j}\rvert^p}{\sum_{k=1}^n 
\lvert H_{i,k}\rvert^p}
\]

\[
\text{Adjoint}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert 
H_{j,i}\rvert^p}{\sum_{k=1}^n 
\lvert H_{k,i}\rvert^p}
\]

where $p\in \mathbb{N}$.
This way of proceeding, for $p=1$, still basing on the work \cite{AADBTW2005}, 
is called 
\textit{Monte Carlo Almost Optimal} (MAO).
The reason why the almost optimal probability is called this way is associated 
with a notice from Theorem \ref{lemma}. Indeed from 
the mentioned lemma we know that the quantity $\displaystyle 
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is minimized when the probability vector is 
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert 
a_k\rvert}$. This implies that the almost optimal probability minimizes the 
$\infty$-norm of $\hat{H}$ in the Forward method and the $1$-norm of $\hat{H}$ 
in the 
Adjoint method. This is very important to be pointed out since these two norms 
are so far the best practical upper bounds for $\rho(\hat{H})$ that we can 
easily 
handle. In the Section dedicated to the numerical results it will be proved 
that the Almost Optimal Probability eases the convergence of the MC solver, 
even in cases when the employment of the Uniform Probability does not provide 
convergence at all.

\subsection*{Examples of matrices for which $\rho(H)<1$ and $\rho(\hat{H})>1$}
By focusing on the iteration matrix related to the fixed point scheme, we 
provide some examples of matrices for which the traditional condition 
$\rho(H)<1$ is not mirrored by the equally necessary condition 
$\rho(\hat{H})<1$.
The Almost Optimal transition probability is employed in order to achieve 
optimal upper bounds for the spectral radius of $\hat{H}$. The iteration matrix 
is 
obtained by applying a left diagonal preconditioner to the original system 
\ref{linsys}.

The matrices we consider are taken from the Matrix Market repository. The first 
one is "JPWH\_991". As it can be noticed from Table \ref{table_nonconv}, the 
Forward MC guarantees convergence since both the the conditions on $H$ and 
$\hat{H}$ 
are verified. The same does not hold for the Adjoint method, since the condition 
on $\hat{H}$ does not hold anymore.

Another example provided is the matrix "FS\_680\_1". In this case both the 
Forward and the Adjoint methods do not converge, since the spectral radius of 
$\hat{H}$ is bigger than 1 for both the methods.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{d.o.f.'s} &
$\rho(H)$&Forward $\rho(\hat{H})$&Adjoint $\rho(\hat{H})$\\
\hline
JPWH\_991 & 991 & $0.9797$ & $0.9797$ & $1.0505$\\ 
\hline
FS\_680\_1 & 680 & $0.9697$ & $1.2554$ & $3.5984$\\
\hline
\end{tabular}
\caption{Examples of matrices for which $\rho(\hat{H})<1$ is violated.} 
\label{table_nonconv}
\end{table}

\section{Adaptive approaches}

In formulas (\ref{dir_mean}) and (\ref{adj_mean}) the estimation of the 
solution to
the linear system (\ref{linsys_prec}) involves infinite sums. Of course this is 
not viable for 
computational issues, therefore it is necessary to resort to truncation of 
the sums through proper criteria. In details, the truncation implies the 
employment of a finite number of permutations and a finite number of 
steps for each one of the permutations computed. Therefore it is necessary to 
truncate at two different levels to decide both how many permutations to employ 
and how many steps each one of them should be kept running.

\subsection{An adaptive cut-off for the length of random walks}

At first we focus on the definition of a weight cut-off 
in order to decide where to terminate the histories. It may be reinterpreted as 
a quantity guiding us to decide how many terms of the Neumann 
series to consider. \newline

The goal is to apply the cutoff via an automatic procedure, without requiring 
the intervention of the user directly.
Therefore we are looking for an integer $m$ such that  
\[
\tilde{\theta}_i=E\bigg[\sum_{\ell=0}^m W_{\ell} 
f_{k_{\ell}}\bigg]=x_i=\sum_{\ell=0}^m 
\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n 
P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1}, 
k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots 
w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}
\]

and

\[
\tilde{\theta}_j=E\bigg[\sum_{\ell=0}^m W_{\ell}\delta_{k_{\ell}, 
j}\bigg]=\sum_{\ell=0}^{m}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n 
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots 
w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}
\]

are good approximations of (\ref{dir_mean}) and (\ref{adj_mean}) respectively.

In \cite{Slattery2013} there is a criterion for the cutoff of the random walk 
applicable to both the Forward and the Adjoint method. It requires the set up 
of a relative cutoff threshold $W_c$ and to look for a step $m$ such that
\begin{equation}
W_m \le W_c W_0.
\label{cutoff}
\end{equation}
$W_0$ is the value of the weight at the initial step of the random walk and 
$W_m$ is the value of the weight after $m$ steps.

\subsection{An a posteriori variance-based adaptivity}

We now focus on quantities that might lead to a proper 
number of permutations to compute, in order to get a prescribed order of 
accuracy.
As regards the Forward method, we know that the expression for the variance 
is defined in formula (\ref{dir_var}).

In this context, a reasonable criterion to determine the number $\tilde{N}_i$ 
of random walks to be run is setting a threshold $\varepsilon_1$ and determine
\begin{equation}
\tilde{N}_i \quad s.t.\quad \frac{\sqrt{Var[\theta_i]}}{\lvert 
E[\theta_i]\rvert}<\varepsilon_1, \quad i=1,\cdots,n.
\label{forward_adapt}
\end{equation}
The dependence of $Var[\theta_i]$ and $E[\theta_i]$ on $\tilde{N}_i$, which 
seems 
to be absent in the previous formula, is highlighted by the fact that 
$\theta_i$ 
is estimated by fixing a finite number of histories.
Therefore we are controlling the relative standard deviation requiring it not 
to be too large. In other words we are pursuing a statistical setting where the 
uncertainty factor is not dominating over the expected value.
This simple adaptive approach can be applied for the estimation of each 
component $x_i$. Hence different number of histories may be employed to 
compute different entries of the solution vector. \newline

As concerns the Adjoint method, the estimation of the variance for each entry 
is reproduced by Formula (\ref{adj_var}).

A possible adaptive selection of $\tilde{N}$, in this situation, is 
\begin{equation}
\tilde{N} \quad s.t. \quad \frac{\lVert 
\boldsymbol{\sigma}_{\tilde{N}}\rVert_1}{\lVert 
\mathbf{x}\rVert_1}<\varepsilon_1,
\label{adjoint_adapt}
\end{equation}
where $\boldsymbol{\sigma}$ is a vector whose entries are 
$\boldsymbol{\sigma}_{\tilde{N},i}=Var[\theta_i]$.

What introduced above can be exploited in order to build 
an a posteriori adaptive algorithm, capable to identify the minimal value of 
$\tilde{N}$ that verifies (\ref{forward_adapt}) or (\ref{adjoint_adapt}) 
respectively.

Below there are the pseudo-codes associated with both Forward and Adjoint Monte 
Carlo.

\begin{algorithm}[H]
 \KwData{$N$, $\varepsilon_1$}
 \KwResult{$\tilde{N}_i$, $\sigma_i$, $x_i$}
 For each entry of the solution vector (for $i=1,\cdots,n$) \;
 $\tilde{N}_i=N$\;
 compute $\theta_i$\;
 \While{$\frac{\sigma_i}{\lvert 
E[\theta_i]\rvert}<\varepsilon_1$}{
\vspace{0.1cm}
  $\tilde{N}=\tilde{N}+N$\;
  compute $E[\theta_i]=x_i$\;
 }
 return $\theta_i$, $x_i$, $\sigma_i$\; 
 \caption{A posteriori adaptive Forward Monte Carlo}
\end{algorithm}

\begin{algorithm}[H]
 \KwData{$N$, $\varepsilon_1$}
 \KwResult{$\tilde{N}$, $\boldsymbol{\sigma_i}$, $\mathbf{x}$}
 $\tilde{N}=N$\;
 compute $\theta$\;
 \While{$\frac{\lVert\boldsymbol{\sigma} \rVert}{\lVert 
E[\theta]\rVert}<\varepsilon_1$}{
\vspace{0.1cm}
  $\tilde{N}=\tilde{N}+N$\;
  compute $E[\theta]=\mathbf{x}$\;
 }
 return $\theta$, $\mathbf{x}$, $\boldsymbol{\sigma}$\; 
 \caption{A posteriori adaptive Adjoint Monte Carlo}
\end{algorithm}

\section{Sets of matrices for which the convergence of the MC linear solver is 
guaranteed}

As we have already seen, sufficient conditions for 
the convergence of the Monte Carlo linear solvers are very restrictive (see 
\cite{MASC2013}), whereas the necessary and sufficient condition requires the 
computation of 
$\rho(H)$and $\rho(\hat{H})$, which is very expensive in general. Once a 
proper 
preconditioner is picked (e.g. Approximate Inverse), $\rho(H)<1$ may 
be always guaranteed, whilst $\rho(\hat{H})<1$ is more 
problematic to be verified.

In order to bypass the cost of these checks, it is useful to pursue for sets of 
matrices for 
which the Monte Carlo linear solver always converges, maybe with the use of an 
ad-hoc preconditioner.

 \subsection{Strictly diagonally dominant matrices}
 One of the these sets is 
represented by strictly diagonally dominant matrices.

\begin{defn}
 A matrix $A\in\mathbb{R}^{n\times n}$is strictly diagonally dominant (s.d.d) 
by rows if
 \begin{equation}
    \lvert a_{ij}\rvert>\sum_{\substack{i=1\\ i\ne j}}^{i=n}\lvert 
a_{ij}\rvert
\label{sddr}
 \end{equation}
\end{defn}

\begin{defn}
 A matrix $A\in\mathbb{R}^{n\times n}$is strictly diagonally dominant (s.d.d) 
by columns if
 \begin{equation}
    \lvert a_{ij}\rvert>\sum_{\substack{j=1\\ j\ne i}}^{j=n}\lvert 
a_{ij}\rvert
\label{sddc}
 \end{equation}
\end{defn}

When (\ref{sddr}) holds we can resort to a left diagonal preconditioning and we 
get an iteration matrix $H=I-D^{-1}A$ such that $\lVert H \rVert_{\infty}<1$.

Introducing a MAO transition probability for the Forward Method we get
\[
 P_{ij}=\frac{\lvert H_{ij}\rvert}{\sum_{k=1}^n\lvert H_{ik}\rvert}
\]
therefore the entries of $\hat{H}$ are defined as follows
\[
 \hat{H}_{ij}=\frac{H^2_{ij}}{P_{ij}}=\lvert H_{ij}\rvert\bigg 
(\sum_{k=1}^n\lvert 
H_{ik}\rvert\bigg ).
\]
Therefore
\[
 \sum_{j=1}^n\lvert \hat{H}_{ij} \rvert = \sum_{j=1}^n \hat{H}_{ij} =\bigg 
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg )\bigg (\sum_{k=1}^n\lvert 
H_{ik}\rvert\bigg ) =\bigg 
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg ) ^2 <1 \quad \forall i=1,\cdots, n.
\]
This automatically implies that $\rho(\hat{H})<=\lVert 
\hat{H}\rVert_{\infty}<1$. Thus the Forward Monte Carlo converges.
However nothing a priori can be 
stated about the behavior of the Adjoint method.\newline

Instead if (\ref{sddc}) holds we can resort to a right diagonal preconditioning 
and we 
get an iteration matrix $H=I-AD^{-1}$ such that $\lVert H \rVert_{1}<1$.
In this case, by following a similar reasoning to the one made before, we 
conclude that the Adjoint Method convergences thanks to $\lVert 
\hat{H}\rVert_1<1$, 
whereas nothing can be said in advance as concerns the Forward method.

 \subsection{Generalized diagonally dominant matrices (GDDM)}

 Another set of matrices for which the convergence of 
the MC solvers is guaranteed is made of \textit{generalized diagonally 
dominant matrices}.

\begin{defn}
A square matrix $A\in\mathbb{R}^{n\times n}$ is said to be generalized 
diagonally dominant if 
\[
 \lvert a_{ii}\rvert x_i \ge \sum_{\substack{j=1\\j\ne i}}^n \lvert 
a_{ij}\rvert 
x_j, \quad i=1,\cdots,n
\]
for some positive vector $\mathbf{x}=(x_1,\cdots,x_n)^T$.
\end{defn}

A proper subset of the generalized diagonally dominant matrices is represented 
by $M$- matrices (see \cite{Ax1996}).
There are many definitions of $M$-matrices that may be found on books and all 
of them are equivalent. The one presented in \cite{Saad} states the 
following

\begin{defn}
A matrix $A\in\mathbb{R}^{n\times n}$ is said to be an $M$-matrix if it 
satisfies the following four properties:
\begin{itemize}
 \item $a_{i,i}>0$ for $i=1,\cdots,n$
 \item $a_{i,j}\le 0$ for $i\ne j$, $i,j=1,\cdots,n$
 \item $A$ is nonsingular
 \item $A^{-1}\ge 0$
\end{itemize}
\end{defn}

A very important result about $M$-matrices enables to come up with a convergent 
splitting and, therefore, a convergent fixed point scheme.

In particular Jacobi, Block Jacobi and Gauss Seidel  are 
convergent 
splitting for an $M$-matrix.
However, the construction of a splitting like this is not enough to ensure the 
convergence of 
the Monte Carlo linear solver. In fact none of these conditions is sufficient 
to guarantee $\rho(\hat{H})<1$. Because of this obstacle, other options must 
be considered.\newline

For this purpose we 
appeal to a result in \cite{Li2002}. In this paper the author presents an 
algorithm to transform a generalized diagonally dominant matrix with 
nonzero diagonal entries into strictly diagonally dominant. The 
algorithm works for a generic complex matrix $A\in\mathbb{C}^{n\times n}$.

Here below we report the algorithm at hand\newline

\begin{algorithm}[H]
 For a given complex matrix $A$, $a_{ii}\ne 0$, $i=1,\cdots, n$;\newline
 \begin{enumerate}
  \item Compute $S_{i}=\sum_{\substack{j=1 \\ j\ne i}^{n}}\lvert 
a_{ij}\rvert$, 
$i=1,2,\cdots,n$
\item Set $t=0$. For $i=1,2,\cdots, n$, if $\lvert a_{ii}\rvert>S_i$, then set 
$t=t+1$
\item If $t=0$, then print "A is not a GDDM": END
\item If $t=n$, then print "A is a GDDM": END
\item \For{i=1,n}{$d_{i}=\frac{S_i+\varepsilon}{\lvert 
a_{ii}\rvert+\varepsilon} \quad 
\varepsilon>0,\quad j=1,2,\cdots,n$\;
$a_{ji}=a_{ji}\cdot d_i$}
\item Go to step 1.
 \end{enumerate}
 \caption{Algorithm to turn a GDDM matrix into a s.d.d by rows.}
\end{algorithm}

This approach turns a generalized diagonally dominant matrix into a 
strictly diagonally dominant matrix by rows. By substituting 
$\displaystyle S_{i}=\sum^{n}_{\substack{j=1 \\ j\ne i}}\lvert a_{ij}\rvert$ 
at step 1 with 
$\displaystyle S_{i}=\sum^{n}_{\substack{i=1 \\ i\ne j}}\lvert a_{ij}\rvert$ 
and by replacing 
$a_{ji}=a_{ji}\cdot d_i$ with $a_{ji}=a_{ji}\cdot d_j$ we obtain the algorithm 
that turns a GDDM matrix into a s.d.d. by columns.\newline

Once we have applied this transformation to the matrix at hand into a s.d.d., 
we can use the Monte Carlo linear solver which is ensured to converge.

 \subsection{"Block-diagonally dominant"-like matrices}
 In this section we wonder in which situations a block diagonal preconditioning 
might come up with a convergent Monte Carlo linear solver. 
By mimicking the computations we already showed previously, the iteration 
matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal 
preconditioning takes the form

\[
 H=I-D^{-1}A=\begin{bmatrix}0_{n_1\times n_1} & -[A_{11}]^{-1}A_{12} & \cdots &
\cdots & -[A_{11}]^{-1}A_{1p} \\
-[A_{22}]^{-1}A_{21} & 0_{n_2\times n_2} & -[A_{22}]^{-1}A_{23} & 
\cdots & -[A_{22}]^{-1}A_{2p}\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
-[A_{pp}]^{-1}A_{p1} &  \cdots & \cdots& 
-[A_{pp}]^{-1}A_{p,p-1} & 0_{n_p \times n_p}
\end{bmatrix}.
\]

For the sake of clarity, we notify that the symbol "$\%$" used in the following 
subsections represents the modulus function between two integer numbers, while 
the floor function is represented with pretty standard symbol "$\lfloor 
\cdot \rfloor$".

\subsubsection*{Forward method}
By assuming that all $n_i$'s have the same value we may define
\[
 m=n_i=\text{size of a block}=\frac{n}{p}.
\]

The MAO transition probability matrix becomes:
\[
 P_{i,j}=\frac{\lvert H_{i,j} \rvert}{\sum_{k=1}^n\lvert H_{i,k} 
\rvert}=\frac{\bigg \lvert \bigg ([A_{\lfloor 
\frac{i}{m}\rfloor \lfloor] 
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor 
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg 
\rvert}{\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg 
([A_{\lfloor 
\frac{i}{m}\rfloor \lfloor 
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor 
\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert}
\]

Consequently, the $\hat{H}$ matrix is defined such that
\[
\hat{H}_{i,j} = \lvert H_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert H_{ik}\rvert\bigg)=
\bigg \lvert \bigg ([A_{\lfloor \frac{i}{m}\rfloor \lfloor 
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor 
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg 
\rvert
\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor 
\frac{i}{m}\rfloor \lfloor 
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor 
\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert
\]

By computing the sum over a generic row of $\hat{H}$ we get:
\[
 \sum_{j=1}^n \lvert \hat{H}_{i,j}\rvert=\sum_{j=1}^n \hat{H}_{i,j} = 
 \bigg ( \sum_{\substack{j=1\\j\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor 
\frac{i}{m}\rfloor \lfloor 
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor 
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert \bigg ) ^2.
\]

If we focus on the norm $\lVert \hat{H}\rVert_{\infty}$, then the following 
equivalence condition holds:
\[
 \lVert \hat{H}\rVert_{\infty}<1 \Leftrightarrow \sum_{\substack{j=1\\j\ne 
i}}^n\bigg \lvert \bigg ([A_{\lfloor 
\frac{i}{m}\rfloor \lfloor 
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor 
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert <1 \quad \forall 
i=1,\cdots,n
\]
 A sufficient condition for this to happen is
 \begin{equation}
  \sum_{\substack{j=1\\j\ne i}}^p \lVert [A_{ii}]^{-1}A_{ij}\rVert_{\infty}<1.
    \label{block_cs}\quad \forall i=1,\ldots,p.
 \end{equation}

By defining a matrix $\tilde{H}\in \mathbb{R}^{p\times p}$ such that
\[
 \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{\infty} & 
\cdots &
\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{\infty} \\
\lVert [A_{22}]^{-1}A_{21}\rVert_{\infty} & 0 & \lVert 
[A_{22}]^{-1}A_{23}\rVert_{\infty} & 
\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{\infty} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert [A_{pp}]^{-1}A_{p1}\rVert_{\infty} &  \cdots & \cdots& 
\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{\infty} & 0
\end{bmatrix}
\]

We can use the $\tilde{H}$ matrix just defined in order to introduce a 
sufficient condition for the convergence of the Forward Monet Carlo method with 
a Block Diagonal preconditioning.

\begin{equation}
 \lVert \tilde{H} \rVert_{\infty}<1 \Rightarrow \lVert \hat{H} 
\rVert_{\infty}<1.
\end{equation}

\subsubsection*{Adjoint method} 
Analogously to the Forward method, if we define 
\[
(\hat{H})^T_{i,j} = \lvert H^T_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert 
H^T_{ik}\rvert\bigg)
\]
we can formulate a sufficient condition for the convergence of the Adjoint 
Monte 
Carlo method by introducing a matrix $\tilde{H}$ which in this case is such that
\[
  \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{1} & \cdots 
&
\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{1} \\
\lVert [A_{22}]^{-1}A_{21}\rVert_{1} & 0 & \lVert 
[A_{22}]^{-1}A_{23}\rVert_{1} & 
\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{1} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert [A_{pp}]^{-1}A_{p1}\rVert_{1} &  \cdots & \cdots& 
\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{1} & 0
\end{bmatrix}.
\]
The sufficient condition assumes the form
\begin{equation}
 \lVert \tilde{H} \rVert_{1}<1 \Rightarrow \lVert \hat{H} \rVert_{1}<1.
\end{equation}

\section{Linear solvers with hybrid schemes}

Even if the standard Monte Carlo, in both its Forward and Adjoint formulation, 
has the advantage of being embarrassingly parallelizable (because of the 
independence of the samplings employed), it has a drawback as 
well. In fact its slow rate of convergence, associated with the 
$\frac{1}{\sqrt{N}}$ behavior predicted but the CLT, prevents its use for 
problems of actual interest in applied linear algebra (with many d.o.f's).

Moreover, when the spectral radius of the iteration matrix is close to the 
unit 
value, Monte Carlo direct methods reach convergence after an excessive number 
of steps for each 
permutation. In particular this aspect will be pointed out more by some 
numerical tests analyzed in the upcoming sections of this work.
This motivation makes the standard Mote Carlo paradigm even more 
unpractical for solving linear systems associated with realistic applications. 
\newline

To cope with this, in the early 2000's Evans et al. introduced a new method: 
the 
\textit{Monte Carlo Synthetic Acceleration (MCSA)} 
(\cite{ESW2013} and \cite{EMSH2014}). Earlier contributions on these direction 
are attributed also to other scientists (e.g. \textit{Sequential Method} by 
J.H. Halton). However the MCSA is the algorithm we are going to focus on more 
hereafter. If 
we think about a fixed point formulation of the problem
such as $\mathbf{x}=H\mathbf{x}+\mathbf{b}$, then the Monte Carlo Synthetic 
Acceleration assumes this form

\begin{algorithm}[H]
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$x_{num}$}
 $\mathbf{x}^{l}=\mathbf{x}_0$\;
 \While{not reached convergence}{
  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
  $\delta \mathbf{x}^{l+\frac{1}{2}}=(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$\;
  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta 
\mathbf{x}^{l+\frac{1}{2}}$\;
 }
 $x_{num}=x^{l+1}$\; 
 \caption{Monte Carlo Synthetic Acceleration}
\end{algorithm}
The Monte Carlo method is used to compute the updating contribution $\delta 
\mathbf{x}^{l+\frac{1}{2}}$.

The role of the first line of the iterative process has the goal to compress 
the residual of the previous iteration. This is the actual difference 
between the MCSA and the Sequential Monte Carlo. Once this done, the corrected 
residual 
is employed to compute an update $\delta x^{1+\frac{1}{2}}$ to the solution via 
one of the Monte Carlo techniques introduced before. The utility of compressing 
the residual at the very beginning of the iteration will be shown in the two 
last test cases for numerical validations.
The intent of MCSA is to designate most of the time employed by the 
computation to the accomplishment of the stochastic part of the algorithm. 
This 
is useful to decrease the probability of fault occurrence in the non-resilient 
steps, which is represented by the non-stochastic component of the numerical 
scheme presented above. In particular, the higher is the percentage of time 
spent on the computation of $\delta x^{1+\frac{1}{2}}$ via Monte Carlo, the 
better it should be in terms of resilience.

\subsection{Preconditioning techniques}
In this section we focus on the pursuit of preconditioners that facilitate the 
convergence of the stochastic schemes introduced so far. A priori, all of the 
following techniques are good candidates to improve the numerical setting. 
However issues about the fill-in effect and the verification of convergence 
conditions limit their employment for our purposes.

\subsubsection*{Block Preconditioning}
Consider a matrix $A\in\mathbb{R}^{n\times n}$ of the form
\[
 A=\begin{bmatrix}A_{11} & A_{12} & \cdots & A_{1p}\\
    A_{21} & A_{22} & \cdots & A_{2p}\\
    \vdots & \vdots & \ddots & \vdots \\
    A_{p1} & \cdots & \cdots & A_{pp}
   \end{bmatrix}
\]
where $p$ is a divisor of $n$ and such that $A_{ii} \in \mathbb{R}^{n_i 
\times n_i}$ are nonsingular square matrices.
By defining a block diagonal matrix such as 
\[
 D=\begin{bmatrix}A_{11} & 0_{12} & \cdots & 0_{1p}\\
    0_{21} & A_{22} & \cdots & 0_{2p}\\
    \vdots & \vdots & \ddots & \vdots \\
    0_{p1} & \cdots & \cdots & A_{pp}
   \end{bmatrix}
\]
it is possible to compute its inverse. 
$D^{-1}$  can be used as a preconditioner for the original matrix $A$.
In particular 
\[
 D^{-1}A=\begin{bmatrix}I_{n_1\times n_1} & A_{11}^{-1}A_{12} & \cdots &
\cdots & A_{11}^{-1}A_{1p} \\
A_{22}^{-1}A_{21} & I_{n_2\times n_2} & A_{22}^{-1}A_{23} & 
\cdots & A_{22}^{-1}A_{2p}\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
A_{pp}^{-1}A_{p1} &  \cdots & \cdots& 
A_{pp}^{-1}A_{p,p-1} & I_{n_p \times n_p}
\end{bmatrix}.
\]
The ultimate goal is to resort to this preconditioning technique in order to 
achieve the condition $\rho(D^{-1}A)<1$. In fact this is necessary in order to 
reformulate \ref{linsys} into \ref{fixedpoint}.

Convergence of \ref{fixedpoint} with block diagonal preconditioning, from the 
deterministic point of view, is guaranteed when $A$ is an irreducibly 
diagonally dominant matrix or when $A$ is an $M$-matrix. In fact in situation 
such as these, theoretical results guarantee that $A=D-N$ is a regular 
splitting (see \cite{Ax1996}). However the same theoretical results cannot be 
exploited to guarantee the convergence of the MC solvers, because of the 
additional check on $\rho(\hat{H})<1$ to be considered.

\subsubsection*{Approximate Inverse of a Sparse Matrix (AINV)}

This stimulated our interests into other types of preconditioners. At first we 
concentrated on Approximate Inverse preconditioners (see \cite{Saad} and 
\cite{Benzi2002}).

There are many practical techniques that may be applied in order to build the 
AINV preconditioner.
One of these is described in \cite{Tuma1996} and \cite{Tuma1998}.
This is the actual approach we applied to the construction of the 
preconditioner. The technique constructs the preconditioner via a 
bi-conjugation fashion.
The idea is to compute an approximate factorization of the form $W^T A Z=D$. 
$W$ and $Z$ are unit upper triangular matrices and $D$ is a diagonal matrix.

In order to compute the AINV preconditioner, a \texttt{FORTRAN} code 
provided by Miroslav Tuma (Academy of Sciences, Prague) is used. 
We report as follows the algorithm employed for the construction of the so 
called \textit{Right-looking Factored AINV}. The algorithm provide the user 
with a dropping tolerance $\tau$ which can be used in order to tune the fill-in 
effect 
of the factored preconditioner. \newline

\begin{algorithm}[H]
 \KwData{$p=q=(0,\cdots,0)\in \mathbb{R}^n$, $Z=[z_1,\cdots,z_n]=I_n$, 
$W=[w_1,\cdots,w_n]=I_n$}
 \KwResult{$Z$, $W$}
 \For{$k=1,\cdots, n$}{ 
 $p_k=w_k^T A e_k$, $q_k=e_k^T A z_k$ \;
     \For{$i=k+1,\cdots,n$}{
         $p_i=(w_k^T A e_k)/p_k$, $q_i=(e_k^T A z_i)/q_k$\;
         Apply a dropping rule to $p_i$, $q_i$\;
         $w_i=w_i - w_k p_i$, $z_i=z_i - z_k q_i$\\
         Apply a dropping rule to $w_{j,i}$ and $z_{j,i}$ for $j=1,\cdots,i$\;
     }\;
  }\;
 Choose diagonal entries of $D$ as the component of $p$ or $q$\;
 return $Z$, $W$\; 
 \caption{Right-looking Factored AINV}
\end{algorithm}

\subsubsection*{Incomplete LU Factorization Preconditioners (ILU)}
This set of preconditioners has the goal of computing a sparse lower triangular 
matrix $L$ and a sparse upper triangular matrix $U$ such that the residual 
matrix $R=LU-A$ respects some constraints. One of these constraints may be the 
fact that some entries must be set to zero.
For instance we can introduce a zero pattern set 
\[
 P\subset \{(i,j)\lvert i\ne j; 1\le i,j\le n\}
\]
and use this set in order to impose some constraints on the shape of the 
preconditioner built.\newline

Different variants of the 
Incomplete LU Factorization algorithm are available (see \cite{Saad}). 
However it is 
possible to prove that if all the formulation are well defined, then they are 
all equivalent. 
One of these variants is the following

\begin{algorithm}[H]
 \KwData{$P$}
 \KwResult{$A$}
 For each $(i,j)\in P$ set $a_{ij}=0$\;
 \For{$k=1,\cdot,n-1$}{
 \For{$i=k+1,n$}{
 \If{$(i,k)\notin P$}{
 $a_{ik}=\frac{a_{ik}}{a_{kk}}$
 \For{$j=k+1,\cdots, n$ and for $(i,j)\notin P$}{
  $a_{ij}=a_{ij}-a_{ik}*a_{kj}$
 }
 }
 }
 }
 return $Z$, $W$\; 
 \caption{General Static Pattern ILU}
\end{algorithm}

This kind of algorithm is guaranteed to terminate without any breakdown just 
for $M$-matrices.

Assuming that the zero pattern $P$ coincides with the zero pattern of $A$ leads 
to the $ILU(0)$. \newline

\begin{algorithm}[H]
 \KwData{$P$}
 \KwResult{$A$}
 \For{$i=2,\cdot,n$}{
 \For{$k=i,i-1$ and for $(i,k)\in NZ(A)$}{
  $a_{ik}=\frac{a_{ik}}{a_{kk}}$
  \For{$j=k+1,\cdots,n$ and for $(i,j)\in NZ(A)$}{
  $a_{ij}=a_{ij}-a_{ik}a_{kj}$}
 }
 }
 \caption{ILU(0).}
\end{algorithm}

More sophisticated variants of the algorithm enable to enlighten the constraint 
of the sparsity pattern, finding a compromise between sparsity and accuracy in 
approximating the inverse of $A$. We refer to \cite{Saad} and 
\cite{Benzi2002} for further information about it.

\section{Numerical results}

All the numerical results shown in the following sections are computed with 
\texttt{Matlab} on a standard laptop. The simulations are accomplished in a 
serial mode.
In the 
first subsection the standard Monte Carlo solver is employed, while in the 
subsequents ones results are obtained with the MCSA scheme.

\subsection{Check on the effectiveness of the adaptive criteria}
In this subsection we show some numerical tests to validate the utility of the 
adaptive approaches for the selection of the number of permutations to employ. 
For this purpose, we restrict our attention to standard Monte Carlo linear 
solvers.

At first we take into account the Forward Monte Carlo linear solver.
In particular, the adaptive threshold associated with formulas 
(\ref{forward_adapt}) and 
(\ref{adjoint_adapt}) is set to $\varepsilon_1=0.1$ and to 
$\varepsilon_1=0.01$. \newline

A set of matrices associated with problems of various nature has been collected.
All the problems treated have a small number of d.o.f's to control the 
massive computational cost of MC solvers on a standard laptop. All the analyzed 
matrices are derived from the 
discretization of other differential problems.
One of these matrices is produced by the finite volume discretization of 
a thermal equation (Marshak problem), another one is a 1D diffusion reaction 
problem discretized with finite differences, a 
2D laplacian discretized with finite differences and an advection 
diffusion 
problem discretized with quadrilateral linear finite elements (the 
\texttt{ifiss} package is used in this case). 
For all the problems at hand a left diagonal preconditioning is applied. 
Details about all these matrices are gathered in Table 
\ref{table_data}.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{d.o.f.'s} &
$\rho(H)$&Forward $\rho(\hat{H})$&Adjoint $\rho(\hat{H})$\\
\hline
1d shifted Laplacian & 50 & $0.4991$ & $0.9827$ & $0.9827$\\ 
\hline
2d Laplacian & 900 & $0.9949$ & $0.994$ & $0.9945$\\
\hline
ifiss & 1089 &$0.9836$ & $0.983$ & $0.983$ \\
\hline
Marshak Equation & 1600 & $0.6009$ & $0.3758$ & $ 0.3815$ \\
\hline
\end{tabular}
\caption{Properties of the matrices employed for the checks on adaptivity.} 
\label{table_data}
\end{table}

For all the numerical tests results are presented both by using the Forward 
and the Adjoint Monte Carlo.

As concerns the Forward method, a safety threshold for maximal number of 
histories per 
entry is set to $10^7$. Results are presented for 
$\varepsilon_1=0.1$ and $\varepsilon_1=0.01$ in Tables \ref{tab:For_adapt} and 
\ref{tab:For_adapt2} respectively. A batch size of two is used at each 
adaptive check to verify the magnitude of the apparent relative standard 
deviation. As expected, results are aligned with the convergence rate predicted 
by the Central Limit Theorem. Indeed, decreasing of a factor of 10 the 
tolerance $\varepsilon_1$, the relative error decreases of the same order, 
employing in average one hundred times more histories.\newline

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.003$ & $5,220$\\
\hline 
2d Laplacian & $0.1051$ & $262,350$\\
\hline
ifiss advection diffusion & $0.022$  & $1,060,770$\\
\hline
Marshak equation & $0.0012$ & $1,456,000$\\
\hline
\end{tabular}
\caption{Forward Monte Carlo. Adaptive selection of histories. 
$\varepsilon_1=0.1$} 
\label{tab:For_adapt}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $4\cdot 10^{-4}$ & $512,094$\\
\hline 
2d Laplacian & $0.0032$ & $108,551,850$\\
\hline
ifiss advection diffusion & $0.0023$  & $105,476,650$\\
\hline
Marshak equation & $5.8 \cdot 10^{-4}$ & $144,018,700$\\
\hline
\end{tabular}
\caption{Forward Monte Carlo. Adaptive selection of histories. 
$\varepsilon_1=0.01$} 
\label{tab:For_adapt2}
\end{table}

As regards the Adjoint Monte Carlo, at each adaptive check the number of 
random walks employed is increased by ten.

A maximal number of histories equal to $10^10$ is employed as a safety 
threshold.

Here below Tables \ref{tab:Adj_adapt}, \ref{tab:Adj_adapt2} and 
\ref{tab:Adj_adapt3} show results 
for the different test cases 
considered using $\varepsilon_1=0.1$, $\varepsilon_1=0.01$ and 
$\varepsilon_1=10^{-3}$ respectively.
By comparing the two tables, it is noticed that a decrease of the threshold 
induces a reduction of the relative error of the same order of magnitude. 
Therefore this validates 
the effectiveness of the adaptive selection of histories with an error 
reduction goal. The total number of histories employed increases on average by 
two orders of magnitude and this aspect is in line too with what expected by 
theoretical results.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.08$ & $1820$\\
\hline 
2d Laplacian & $0.136$ & $570$\\
\hline
ifiss advection diffusion & $0.08$  & $2400$\\
\hline
Marshak equation & $0.288$ & $880$\\
\hline
\end{tabular}
\caption{Adjoint Monte Carlo. Adaptive selection of histories. 
$\varepsilon_1=0.1$} 
\label{tab:Adj_adapt}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.0082$ & $185,400$\\
\hline 
2d Laplacian & $0.0122$ & $126,800$\\
\hline
ifiss advection diffusion & $0.0093$  & $219,293$\\
\hline
Marshak equation & $0.0105$ & $650,040$\\
\hline
\end{tabular}
\caption{Adjoint Monte Carlo. Adaptive selection of histories. 
$\varepsilon_1=0.01$} 
\label{tab:Adj_adapt2}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $9.56 \cdot 10^{-4}$ & 15,268,560 \\
\hline 
2d Laplacian & $0.001$ & $12,600,000$\\
\hline
ifiss advection diffusion & $0.0011$  & $23,952,000$\\
\hline
Marshak equation & $0.0011$ &  $80,236,000$\\
\hline
\end{tabular}
\caption{Adjoint Monte Carlo. Adaptive selection of histories. 
$\varepsilon_1=10^{-3}$} 
\label{tab:Adj_adapt3}
\end{table}

The same results can be reproduced by resorting to the expected value 
estimator. Results are shown in the Tables \ref{tab:Adj_adapt4}, 
\ref{tab:Adj_adapt5} and \ref{tab:Adj_adapt6} for the threshold values of 
$\varepsilon_1=0.1$, $\varepsilon_1=0.01$ and $\varepsilon_1=10^{-3}$ 
respectively. As it can be noticed, in terms of error scaling the results are 
pretty similar to the ones obtained with the collision estimator. As regards 
the number of histories as well, used to reach a prescribed accuracy, the 
orders of magnitude are the same for both the collision estimator and the 
expected value one. However the expected value estimator requires most of the 
time a smaller number permutations. This behavior is enhances as the value of 
the threshold decreases, making the computation heavier and heavier.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.0463$ & $1000$\\
\hline 
2d Laplacian & $0.1004$ & $900$\\
\hline
ifiss advection diffusion & $0.0661$  & $1300$\\
\hline
Marshak equation & $0.0526$ & $2000$\\
\hline
\end{tabular}
\caption{Expected value estimator - Adjoint Monte Carlo. Adaptive selection of 
histories. 
$\varepsilon_1=0.1$} 
\label{tab:Adj_adapt4}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.004$ & $100,600$\\
\hline 
2d Laplacian & $0.0094$ & $83,700$\\
\hline
ifiss advection diffusion & $0.0088$  & $124,400$\\
\hline
Marshak equation & $0.0056$ & $166,000$\\
\hline
\end{tabular}
\caption{Expected value - Adjoint Monte Carlo. Adaptive selection of histories. 
$\varepsilon_1=0.01$} 
\label{tab:Adj_adapt5}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.004$ & $10,063,300$\\
\hline 
2d Laplacian & $ 9.31 \cdot 10^{-4} $ & $ 8,377,500 $\\
\hline
ifiss advection diffusion & $ 0.0013 $  & $ 12,435,000 $\\
\hline
Marshak equation & $7.79 \cdot 10^{-4}$ & $16,537,000$\\
\hline
\end{tabular}
\caption{Expected value - Adjoint Monte Carlo. Adaptive selection of histories. 
$\varepsilon_1=10^{-3}$} 
\label{tab:Adj_adapt6}
\end{table}

\subsection{A comparison between different preconditioners and their effect 
on $\rho(H)$ and $\rho(\hat{H})$}

This subsection is focused on showing the effect of different preconditioners 
for the values attained by the spectral radii $\rho(H)$ and $\rho(\hat{H})$.
For this purpose, we consider the two dimensional laplacian 
discretized with centered finite differences and the "ifiss advection 
diffusion" matrix. The properties of both the matrices have already been 
explained in previous sections.
Results in terms of spectral radii when the diagonal preconditioning is applied 
have been shown earlier as well.

Here the analysis is focused on the evaluating the effects of the 
block diagonal preconditioner and the factorized approximate sparse inverse. 
As concerns the matrix $\hat{H}$, it has been built accordingly to te 
procedure for the Adjoint Monte Carlo method. In 
the first case the size of the blocks is a tuning parameter, whilst in the 
second case the tuning parameter is the drop tolerance to manage the fill-in 
effect of the factors.

As regards the employment for the block diagonal preconditiner, in Tables 
\ref{tab:lap2d_block} and \ref{tab:ifiss_block} there is a counterintuitive 
common trend. Indeed in both the cases an increase of $\rho(H)$ and 
$\rho(\hat{H})$  occurs as the size of the blocks diminishes.

In Tables \ref{tab:lap2d_ainv} and \ref{tab:ifiss_ainv} the values of 
the spectral radii are shown for the case of the factorized AINV 
preconditioner. It is interesting to point out that for the 
two-dimensional 
laplacian, a drop tolerance $\tau=0.05$ entails $\rho(H)<1$ but the same does 
not hold for $\hat{H}$. Both the convergence conditions are restored by 
decreasing even more the drop tolerance, causing a massive fill-in effect 
though.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$ 
&$\rho(\hat{H})$\\
\hline
 5 & 2.6164 & $0.9915$ & $0.9837$\\
\hline 
 10 & 5.6027 & $0.9907$ & $0.9861$\\
\hline
 30 & 16.3356 & $0.9898$  & $0.9886$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the two dimensional 
laplacian with block diagonal preconditioning.} 
\label{tab:lap2d_block}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$ 
&$\rho(\hat{H})$\\
\hline
 3 & 1.4352 & $0.9804$ & $0.9531$\\
\hline 
 9 & 3.0220 & $0.9790$ & $0.9674$\\
\hline
 33 & 9.8164 & $0.9783$  & $0.9774$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the ifiss advection 
diffusion matrix with block diagonal preconditioning.} 
\label{tab:ifiss_block}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$ 
&$\rho(\hat{H})$\\
\hline
 0.05 & 8.2797 & $0.9610$ & $1.0231$\\
\hline 
 0.01 & 33.1390 & $0.8668$ & $0.8279$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the two dimensional 
laplacian with factorized AINV preconditioning.} 
\label{tab:lap2d_ainv}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$ 
&$\rho(\hat{H})$\\
\hline
 0.05 & 3.8392 & $0.9396$ & $0.9069$\\
\hline 
 0.01 & 15.1798 & $0.7964$ & $0.6361$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the ifiss advection 
diffusion matrix with factorized AINV preconditioning.} 
\label{tab:ifiss_ainv}
\end{table}

\subsection{The Poisson equation solved with MCSA}
In the current subsection a Poisson problem is taken into account:
\begin{equation}
\begin{cases}
 -\Delta u = f \quad \text{in}\quad \Omega \\
 u=0\quad \text{on} \quad \partial\Omega
 \end{cases}
\end{equation}

where $\Omega=(0,1)\times (0,1)$, $\sigma=1$. 
A finite difference scheme is employed to discretize the problem.
As a right hand side, a 
sinusoidal distribution in both $x$ and $y$ directions is considered, so that 
\[
 f(x_i, y_i)=sin\bigg(\frac{\pi x}{N-1}\bigg)sin\bigg(\frac{\pi y}{N-1}\bigg).
\]
N is the number of node considered on each direction, in our case $N=32$. The 
resulting linear system has 900 d.o.f.'s and $h=\frac{1}{29}$.
By applying a left diagonal preconditioning on the matrix, it is possible to 
reformulate the problem as a fixed point scheme. 
The iteration matrix $H$ is characterized by the fact that $\lVert 
H\rVert_1=1$, which is not sufficient to guarantee the convergence of 
any 
stochastic scheme analyzed in this work. However, by theoretical results, it is 
well known that applying a left diagonal preconditioning to the problem at hand 
produces an iteration matrix $H$ such that
\[
 \lim_{h\rightarrow 0}\rho(H)=1^{-}
\]
where $h$ is the spatial discretization step.
This is enough to have the guarantee that the deterministic fixed point scheme 
converges without the explicit computation of the spectral radius $\rho(H)$. 
However, the smaller is the discretization step $h$, the higher is the number of 
iterations required to reach a prescribed convergence criterion. For our 
discretization setting, the iteration matrix $H$ has a 
spectral radius of $\rho(H)\approx 0.9949$.

In order to have the guarantee of the convergence of the Adjoint Monte Carlo 
method, it is necessary to look at the spectral radius of $\hat{H}$ too. If an 
almost optimal probability is used, this implies that 
the Adjoint Monte Carlo method has a $\hat{H}$ matrix such that 
\[
 \lVert \hat{H}\rVert_1\le (\rVert H\rVert_1)^2=1.
\]
This condition by itself is not enough to guarantee that 
$\rho(\hat{H})<1$.

However, it must be recalled that the iteration matrix $H$ has zero 
valued entries on the main diagonal and it has: 
\begin{itemize}
 \item four entries equal to $\frac{-1}{4}$ on the rows associated 
with a node which is not adjacent to the boundary
\item two entries equal to $\frac{-1}{4}$ on the rows associated with nodes 
adjacent to the corner of the square domain
\item three entries equal to $\frac{-1}{4}$ on the rows associated with nodes 
adjacent to the boundary on an edge of the square domain.
\end{itemize}

Recalling the definition of $\hat{H}$ entrywise in terms of the 
iteration matrix $H$ and the transition probability $P$, we get that

\[
 \hat{H}=\tilde{D}H
\]

where $\tilde{D}$ is a diagonal matrix with diagonal entries equal to
1, $\frac{1}{2}$ or $\frac{3}{4}$. 
In particular $\tilde{d}_i=diag(\tilde D)_i=1$ if the $i$-th row of the 
discretized laplacian is associated with a node which is not adjacent to the 
boundary, $\tilde{d}_i=diag(\tilde D)_i=\frac{1}{2}$ if the row is associated 
with a node of the grid adjacent to the corner of the boundary, 
$\tilde{d}_i=diag(\tilde D)_i=\frac{3}{4}$ if the associated node of the grid 
is adjacent to the boundary of the domain far from a corner.

Since $H$ is an irreducible nonnegative matrix and $\tilde{D}$ is a 
positive definite diagonal matrix, we resort to a result presented in 
\cite{FK1975} to say that
\[
 \rho(\hat{H})=\rho(\tilde{D}H)\le \rho(\tilde D)\rho(H).
\]

Since $\rho(\tilde{D})=1$, then $\rho(\hat{H})\le \rho(H)<1$. 
Therefore the diagonal preconditioning always guarantees a convergent 
setting for the Monte Carlo linear solver, once it is applied on a 
two-dimensional Laplace 
operator discretized with finite differences. By a similar reasoning, the same 
results can be obtained for a $n$-dimensional laplace operator.

The computation of the solution is realized both with the deterministic 
Synthetic Acceleration and with Adjoint MCSA to compare the effectiveness of 
the 
stochastic technique. Results are reported in Table  
\ref{Poisson_results}. For the sake of completeness, results obtained with 
the Sequential Monte Carlo are shown as well. An evident discrepancy in terms 
of numerical iterations 
is already detected with the Sequential Monte Carlo outperforming the 
preconditioned fixed point. Results are improved even more 
when 
switching to the MCSA, which employs less histories than the Sequential MC at 
each iteration to 
reach convergence. In fact the average number of histories employed by the 
Sequential 
Monte Carlo per iteration is almost one order of magnitude higher than for 
MCSA. This 
might be explained by the fact that the first row of the iterative algorithm in 
the MCSA refines the accuracy of the solution coming from the previous 
iteration, 
before using it as the input for the MC solver at the current iterate. 
In particular, since the first line of the algorithm is a deterministic step of 
the Synthetic Acceleration, its effect is to dampen the statistical noise 
produced by the Monte Carlo linear solver to estimate the update $\delta 
x^{n + \frac{1}{2}}$ at the previous iteration. Therefore
refinement provided by the Synthetic Acceleration step decreases the number of 
random walks 
to employ for a prescribed accuracy. This is validated even more by the fact 
that for the first iteration the Sequential Monte Carlo and the MCSA resort to 
the same number of histories, while their behaviours start differing from the 
second iteration on, when the statistical noise is introduced in the estimation 
of the solution to \ref{linsys_prec} (see Figures \ref{SEQ_poisson} and 
\ref{MCSA_poisson}).

\begin{table}[!h]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
Synthetic Acceleration & $9.8081\cdot 10^{-8}$ & 3133 & - \\ 
\hline
Sequential MC & $7.9037 \cdot 10^{-8}$ & 9 & 8.264,900\\
\hline
 Adjoint MCSA & $8.0872\cdot 10^{-8}$ & 8 & 1,738,250\\
\hline
\end{tabular}
\caption{Numerical results for the Poisson problem.}
\label{Poisson_results}
\end{table}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{SEQ_poisson.jpg}
    \caption{Sequential MC - Poisson problem. Number of random walks employed 
at each numerical iteration.}
\label{SEQ_poisson}
\end{figure}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{MCSA_poisson.jpg}
      \caption{MCSA - Poisson problem. Number of random walks employed at each 
numerical iteration.}
\label{MCSA_poisson}
\end{figure}

\subsection{A diffusion reaction problem solved with MCSA}
By modifying the equation of the previous test case, we consider now 
an advection reaction problem

\begin{equation}
\begin{cases}
 -\Delta u +\sigma u= f \quad \text{in}\quad \Omega \\
 u=0\quad \text{on} \quad \partial\Omega
 \end{cases}
\end{equation}
where $\Omega=(0,1)\times (0,1)$, $\sigma=0.1$ and $f=1$. 
A finite difference scheme is applied to discretize the problem.
The number of nodes 
on each direction of the domain is 100, so that $h\approx 0.01$. The 
discretized problem has 9604 d.o.f's. A left 
diagonal preconditioning is still applied to 
the stiffness matrix obtained from the discretization. The 1-norm of the 
iteration matrix is $\lVert H\rVert_1\approx 0.9756$. This automatically 
guarantees the convergence of the Adjoint Monte Carlo linear solver. The 
computation of the solution to the partial differential problem is still 
accomplished with the MCSA algorithm. A threshold of $\varepsilon =10^{-8}$ is 
used as a stop criterion for the relative residual. The threshold 
for the adaptive selection of the random walks instead is set 
to $\varepsilon_1=0.1$.
In Table \ref{DR_results} a comparison between the deterministic Synthetic 
Acceleration 
scheme, Sequential MC and the MCSA is provided. 
The employment of the MCSA provides a significant reduction of the number of 
histories necessary to satisfy the convergence criterion, in line with results 
coming from the previous test case.

\begin{table}[!h]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
 Synthetic acceleration & $9.073\cdot 10^{-8}$ & 634 & - \\ 
\hline
 Sequential MC & $8.415 \cdot 10^{-8}$ &  8 & 12,391,375\\
 \hline
 Adjoint MCSA & $6.633 \cdot 10^{-8}$ &  7 & 3,163,700\\
\hline
\end{tabular}
\caption{Numerical results for the diffusion reaction problem.}
\label{DR_results}
\end{table}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{SEQ_diffreac.jpg}
      \caption{Sequential MC - Diffusion reaction problem. Number of random 
walks 
employed at each 
numerical iteration.}
\label{SEQ_diffreac}
\end{figure}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{MCSA_diffreac.jpg}
      \caption{MCSA - Diffusion reaction problem. Number of random walks 
employed at each 
numerical iteration.}
\label{MCSA_diffreac}
\end{figure}

\subsection{A parabolic problem solved with MCSA}
In this subsection we restrict our treatise to the analysis of parabolic 
partial 
differential problems of the form 
\begin{equation}
\begin{cases}
 \frac{\partial u }{\partial t} + \mathcal{L}(u)=f \quad \text{in} \Omega \\
\text{B.C.} \quad \text{on}\quad \delta \Omega.
\end{cases}
\label{parabolic}
\end{equation}

The partial derivative in time is discretized with finite differences, whereas 
quadrilateral linear finite elements are employed for the space discretization. 
Naming $N_h$ the number of d.o.f.'s associated with the finite element 
discretization, the problem \ref{parabolic} turn into a fully discretized 
problem such as:

\[
 \bigg (\frac{1}{\Delta t}M+A\bigg )(\mathbf{u}^{k+1}-\mathbf{u}^k)+A[\theta 
\mathbf{u}^{k+1}+(1-\theta)\mathbf{u}^k]=\theta 
\mathbf{f}^{k+1}+(1-\theta)\mathbf{f}^k,
\]

where $k$ is an index that goes with the time dimension.

By tuning the value of the parameter $\theta$ different numerical schemes can 
be obtained with different properties of numerical stability. 
However our discussion is not involved in this kind of issues.
In particular we restrict our attention to a single generic time step 
associated with an Implicit Euler time discretization (corresponding to 
$\theta=1$). The vector of the right hand side is chosen so that the exact 
solution to the linear system for the specific time step chosen is a unit 
vector $\tilde{\mathbf{u}}=\mathbbm{1}^{N_h}$.

By referring to $h$ as the space discretization step, we pick
\[
 \Delta t \le C h.
\]

Give an open and bounded set $\Omega=(0,1)\times(0,1)$, we consider the 
following problem
\begin{equation}
 \begin{cases}\frac{\partial u}{\partial t} -\mu \Delta u 
+\boldsymbol{\beta}(\mathbf{x})\cdot \nabla u=0, \;\; \quad \quad 
\mathbf{x}\in \Omega,\quad t\in(0,T] \\
u(\mathbf{x},0) = u_0, \qquad \qquad \qquad \qquad \quad \mathbf{x}\in 
\Omega \\
u(\mathbf{x},t)=u_D(\mathbf{x}),  \; \, \, \qquad \qquad \qquad \quad  
\mathbf{x}\in \partial 
\Omega, \quad t\in(0,T],
 \end{cases}
\end{equation}

where $\mu=\frac{3}{200}$, $\boldsymbol{\beta}(\mathbf{x})=[2y(1-x^2),\; 
-2x(1-y^2)]^T$, \\$u_D=0$ on $\{\{x=0\}\times (0,1)\}$, $\{(0,1)\times 
\{y=0\}\}$, $\{(0,1)\times \{y=1\}\}$.\newline

The discretization of the problem is realized with the \texttt{IFISS} toolbox.
The value of the discretization step is $h=2^{-8}$, generating a discretized 
linear system with 66,049 d.o.f.'s.
As concerns the time discretization, the time step is set to $\Delta t = 
10h$.\newline

The sparse approximate inverse is employed for 
the linear system at hand as a right preconditioner, with a drop tolerance of 
$\tau=0.05$ for both the factors.

In this situation it is not possible to 
resort to any of the sufficient conditions to verify the convergence. Therefore 
the only viable option is the explicit computation of the spectral radii of $H$ 
and $\hat{H}$.
The spectral radius of the iteration matrix is $\rho(H)\approx 
0.9218$ and the spectral radius of $\hat{H}$ for the Adjoint Monte Carlo is 
$\rho(\hat{H})\approx 0.9148$. The transition probability employed is the 
Almost 
Optimal one. Resorting to a Uniform Probability in this case would have impeded 
the 
convergence, since in this case $\rho(\hat{H})\approx 1.8401$. 
Therefore 
this is an example demonstrating that the Almost Optimal Probability 
facilitates the performance of the stochastic algorithm, outperforming the 
Uniform one.

The fill-in effect is 
\[
 \frac{nnz(H)}{nnz(A)}=4.26,
\]
therefore the relative number of nonzero elements in $H$ is still acceptable 
in terms of sparsity pattern and memory storage.

The threshold for the check on the relative residual is set to $\varepsilon 
=10^{-8}$ is, whilst the threshold 
for the adaptive selection of the random walks is set 
to $\varepsilon_1=0.1$.
In order to test the effectiveness of employing the MCSA with respect to a 
purely deterministic scheme, a comparison with the Richardson algorithm and 
the Sequential Monte Carlo is 
accomplished. The results are shown in Table \ref{parabolic_results}. As you 
can see, the adoption of both Sequential Monte Carlo and MCSA reduces 
dramatically the number of iterations with 
respect to the fixed point method. Of course this 
advantage has the 
increase of the computational time as a payoff. Nevertheless we put off this 
discussion which is of 
interest for a parallelized implementation of the method. Even more the number 
of random walks employed by the Sequential MC is higher than the one associated 
with the MCSA, in accordance with the previous results.

\begin{table}[!h]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
 Richardson & $6.277\cdot 10^{-7}$ & 178 & - \\
 \hline
 Sequential MC & $1.918 \cdot 10^{-9}$ & 8 & 51,535,375\\ 
\hline
 Adjoint MCSA & $1.341\cdot 10^{-9}$ & 6 & 16,244,000\\
\hline
\end{tabular}
\caption{Numerical results for the parabolic problem.}
\label{parabolic_results}
\end{table}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{SEQ_parabolic.jpg}
      \caption{Sequential Monte Carlo - Parabolic problem. Number of random 
walks 
employed at each 
numerical iteration.}
\label{MCSA_parabolic}
\end{figure}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{MCSA_parabolic.jpg}
      \caption{MCSA - Parabolic problem. Number of random walks 
employed at each 
numerical iteration.}
\label{MCSA_parabolic}
\end{figure}

Other preconditioners 
have 
been 
tested besides the sparse approximate inverse preconditioner, such as block 
diagonal 
and incomplete LU factorization. However the fill-in effect introduced by the 
application of 
these preconditioners compromises the sparsity of the linear system.

\section{Conclusions and future developments}
The work presented in this paper has been accomplished to provide an overview 
about MC linear solvers. In particular the goal was to combine the main 
theoretical results found in literature with additional novel contributions on 
both the theoretical and empirical sides. This made it possible to understand 
the actual limitations of the presented algorithms.

Necessary and sufficient conditions to ensure convergence are too costly for 
most of the applicative problems coming fro engineering and Computational 
Physics. On the other hand, sufficient criteria are currently viable for a
restricted set of matrices (generalized diagonally dominant).

An empirical analysis has been carries out on parabolic problems and it has 
been shown that a proper selection of the time discretization step enables the 
employment of stochastic solvers.

We reserve for future works the analysis of the plausible connections between 
$\rho(H)$, $\rho(\hat{H})$ and the number of numerical iterations necessary to 
reach convergence. Moreover a study about the resilience of the problem will be 
accomplished, resorting to fault injection techniques.

\addcontentsline{toc}{chapter}{\refname}
\bibliographystyle{alpha}
\bibliography{library}
%\nocite{*}

\end{document}

### 2015-09-16 MAX_reports/MC_solvers_article/MC_solvers.tex
Restructuring paper into smaller number of sections.

@@
 
-\section{Stochastic setting}
-
-We start our treatise by introducing the direct methods.\newline
-
-Assume to start from a linear system such as 
+\section{Stochastic Linear Solvers}
+
+We consider the solution of systems of linear equations of the form
 \begin{equation}
@@
 \subsection{Forward Method}
+
 The goal is to evaluate a functional such as
@@
 \subsection{Adjoint Method}
+
 A second Monte Carlo method can be derived by considering the linear system
@@
 
-\section{Numerical interpretation of the statistical requirements}
+\subsection{Hybrid Stochastic/Deterministic Methods}
+
+Even if the standard Monte Carlo, in both its Forward and Adjoint formulation,
+has the advantage of being embarrassingly parallelizable (because of the
+independence of the samplings employed), it has a drawback as
+well. In fact its slow rate of convergence, associated with the
+$\frac{1}{\sqrt{N}}$ behavior predicted but the CLT, prevents its use for
+problems of actual interest in applied linear algebra (with many d.o.f's).
+
+Moreover, when the spectral radius of the iteration matrix is close to the
+unit
+value, Monte Carlo direct methods reach convergence after an excessive number
+of steps for each
+permutation. In particular this aspect will be pointed out more by some
+numerical tests analyzed in the upcoming sections of this work.
+This motivation makes the standard Mote Carlo paradigm even more
+unpractical for solving linear systems associated with realistic applications.
+\newline
+
+To cope with this, in the early 2000's Evans et al. introduced a new method:
+the
+\textit{Monte Carlo Synthetic Acceleration (MCSA)}
+(\cite{ESW2013} and \cite{EMSH2014}). Earlier contributions on these direction
+are attributed also to other scientists (e.g. \textit{Sequential Method} by
+J.H. Halton). However the MCSA is the algorithm we are going to focus on more
+hereafter. If
+we think about a fixed point formulation of the problem
+such as $\mathbf{x}=H\mathbf{x}+\mathbf{b}$, then the Monte Carlo Synthetic
+Acceleration assumes this form
+
+\begin{algorithm}[H]
+ \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
+ \KwResult{$x_{num}$}
+ $\mathbf{x}^{l}=\mathbf{x}_0$\;
+ \While{not reached convergence}{
+  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
+  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
+  $\delta \mathbf{x}^{l+\frac{1}{2}}=(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$\;
+  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta
+\mathbf{x}^{l+\frac{1}{2}}$\;
+ }
+ $x_{num}=x^{l+1}$\;
+ \caption{Monte Carlo Synthetic Acceleration}
+\end{algorithm}
+The Monte Carlo method is used to compute the updating contribution $\delta
+\mathbf{x}^{l+\frac{1}{2}}$.
+
+The role of the first line of the iterative process has the goal to compress
+the residual of the previous iteration. This is the actual difference
+between the MCSA and the Sequential Monte Carlo. Once this done, the corrected
+residual
+is employed to compute an update $\delta x^{1+\frac{1}{2}}$ to the solution via
+one of the Monte Carlo techniques introduced before. The utility of compressing
+the residual at the very beginning of the iteration will be shown in the two
+last test cases for numerical validations.
+The intent of MCSA is to designate most of the time employed by the
+computation to the accomplishment of the stochastic part of the algorithm.
+This
+is useful to decrease the probability of fault occurrence in the non-resilient
+steps, which is represented by the non-stochastic component of the numerical
+scheme presented above. In particular, the higher is the percentage of time
+spent on the computation of $\delta x^{1+\frac{1}{2}}$ via Monte Carlo, the
+better it should be in terms of resilience.
+
+\subsubsection{Sequential Monte Carlo}
+
+\subsubsection{Monte Carlo Synthetic Acceleration}
+
+\section{Convergence Behavior of Stochastic Methods}
+
 The requests imposed on the expected value and second moment of the estimators
@@
 
-\section{Necessary conditions, sufficient conditions and choice of the 
-transition matrix}
+\subsection{Necessary and Sufficient Conditions}
 
@@
 
-\section{Construction of the transition probability}
+\subsubsection{Forward MC}
+
+\subsubsection{Adjoint MC with Collision Estimator}
+
+\subsubsection{Adjoint MC with Expected-Value Estimator}
+
+\subsection{Construction of Transition Probabilities}
+
 The way the transition probability is defined is extremely important. In fact
@@
 
-\subsection*{Uniform probabilities}
+\subsubsection{Uniform probabilities}
 With this approach transition matrix $P$ is such that all the non-zero elements
@@
 
-\subsection*{Weighted probabilities}
+\subsubsection{Weighted probabilities}
 An alternative way of defining the transition matrix consists in attributing
@@
 
-\subsection*{Examples of matrices for which $\rho(H)<1$ and $\rho(\hat{H})>1$}
-By focusing on the iteration matrix related to the fixed point scheme, we 
-provide some examples of matrices for which the traditional condition 
-$\rho(H)<1$ is not mirrored by the equally necessary condition 
-$\rho(\hat{H})<1$.
-The Almost Optimal transition probability is employed in order to achieve 
-optimal upper bounds for the spectral radius of $\hat{H}$. The iteration matrix 
-is 
-obtained by applying a left diagonal preconditioner to the original system 
-\ref{linsys}.
-
-The matrices we consider are taken from the Matrix Market repository. The first 
-one is "JPWH\_991". As it can be noticed from Table \ref{table_nonconv}, the 
-Forward MC guarantees convergence since both the the conditions on $H$ and 
-$\hat{H}$ 
-are verified. The same does not hold for the Adjoint method, since the condition 
-on $\hat{H}$ does not hold anymore.
-
-Another example provided is the matrix "FS\_680\_1". In this case both the 
-Forward and the Adjoint methods do not converge, since the spectral radius of 
-$\hat{H}$ is bigger than 1 for both the methods.
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{d.o.f.'s} &
-$\rho(H)$&Forward $\rho(\hat{H})$&Adjoint $\rho(\hat{H})$\\
-\hline
-JPWH\_991 & 991 & $0.9797$ & $0.9797$ & $1.0505$\\ 
-\hline
-FS\_680\_1 & 680 & $0.9697$ & $1.2554$ & $3.5984$\\
-\hline
-\end{tabular}
-\caption{Examples of matrices for which $\rho(\hat{H})<1$ is violated.} 
-\label{table_nonconv}
-\end{table}
-
-\section{Adaptive approaches}
-
-In formulas (\ref{dir_mean}) and (\ref{adj_mean}) the estimation of the 
-solution to
-the linear system (\ref{linsys_prec}) involves infinite sums. Of course this is 
-not viable for 
-computational issues, therefore it is necessary to resort to truncation of 
-the sums through proper criteria. In details, the truncation implies the 
-employment of a finite number of permutations and a finite number of 
-steps for each one of the permutations computed. Therefore it is necessary to 
-truncate at two different levels to decide both how many permutations to employ 
-and how many steps each one of them should be kept running.
-
-\subsection{An adaptive cut-off for the length of random walks}
-
-At first we focus on the definition of a weight cut-off 
-in order to decide where to terminate the histories. It may be reinterpreted as 
-a quantity guiding us to decide how many terms of the Neumann 
-series to consider. \newline
-
-The goal is to apply the cutoff via an automatic procedure, without requiring 
-the intervention of the user directly.
-Therefore we are looking for an integer $m$ such that  
-\[
-\tilde{\theta}_i=E\bigg[\sum_{\ell=0}^m W_{\ell} 
-f_{k_{\ell}}\bigg]=x_i=\sum_{\ell=0}^m 
-\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n 
-P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1}, 
-k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots 
-w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}
-\]
-
-and
-
-\[
-\tilde{\theta}_j=E\bigg[\sum_{\ell=0}^m W_{\ell}\delta_{k_{\ell}, 
-j}\bigg]=\sum_{\ell=0}^{m}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n 
-f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots 
-w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}
-\]
-
-are good approximations of (\ref{dir_mean}) and (\ref{adj_mean}) respectively.
-
-In \cite{Slattery2013} there is a criterion for the cutoff of the random walk 
-applicable to both the Forward and the Adjoint method. It requires the set up 
-of a relative cutoff threshold $W_c$ and to look for a step $m$ such that
-\begin{equation}
-W_m \le W_c W_0.
-\label{cutoff}
-\end{equation}
-$W_0$ is the value of the weight at the initial step of the random walk and 
-$W_m$ is the value of the weight after $m$ steps.
-
-\subsection{An a posteriori variance-based adaptivity}
-
-We now focus on quantities that might lead to a proper 
-number of permutations to compute, in order to get a prescribed order of 
-accuracy.
-As regards the Forward method, we know that the expression for the variance 
-is defined in formula (\ref{dir_var}).
-
-In this context, a reasonable criterion to determine the number $\tilde{N}_i$ 
-of random walks to be run is setting a threshold $\varepsilon_1$ and determine
-\begin{equation}
-\tilde{N}_i \quad s.t.\quad \frac{\sqrt{Var[\theta_i]}}{\lvert 
-E[\theta_i]\rvert}<\varepsilon_1, \quad i=1,\cdots,n.
-\label{forward_adapt}
-\end{equation}
-The dependence of $Var[\theta_i]$ and $E[\theta_i]$ on $\tilde{N}_i$, which 
-seems 
-to be absent in the previous formula, is highlighted by the fact that 
-$\theta_i$ 
-is estimated by fixing a finite number of histories.
-Therefore we are controlling the relative standard deviation requiring it not 
-to be too large. In other words we are pursuing a statistical setting where the 
-uncertainty factor is not dominating over the expected value.
-This simple adaptive approach can be applied for the estimation of each 
-component $x_i$. Hence different number of histories may be employed to 
-compute different entries of the solution vector. \newline
-
-As concerns the Adjoint method, the estimation of the variance for each entry 
-is reproduced by Formula (\ref{adj_var}).
-
-A possible adaptive selection of $\tilde{N}$, in this situation, is 
-\begin{equation}
-\tilde{N} \quad s.t. \quad \frac{\lVert 
-\boldsymbol{\sigma}_{\tilde{N}}\rVert_1}{\lVert 
-\mathbf{x}\rVert_1}<\varepsilon_1,
-\label{adjoint_adapt}
-\end{equation}
-where $\boldsymbol{\sigma}$ is a vector whose entries are 
-$\boldsymbol{\sigma}_{\tilde{N},i}=Var[\theta_i]$.
-
-What introduced above can be exploited in order to build 
-an a posteriori adaptive algorithm, capable to identify the minimal value of 
-$\tilde{N}$ that verifies (\ref{forward_adapt}) or (\ref{adjoint_adapt}) 
-respectively.
-
-Below there are the pseudo-codes associated with both Forward and Adjoint Monte 
-Carlo.
-
-\begin{algorithm}[H]
- \KwData{$N$, $\varepsilon_1$}
- \KwResult{$\tilde{N}_i$, $\sigma_i$, $x_i$}
- For each entry of the solution vector (for $i=1,\cdots,n$) \;
- $\tilde{N}_i=N$\;
- compute $\theta_i$\;
- \While{$\frac{\sigma_i}{\lvert 
-E[\theta_i]\rvert}<\varepsilon_1$}{
-\vspace{0.1cm}
-  $\tilde{N}=\tilde{N}+N$\;
-  compute $E[\theta_i]=x_i$\;
- }
- return $\theta_i$, $x_i$, $\sigma_i$\; 
- \caption{A posteriori adaptive Forward Monte Carlo}
-\end{algorithm}
-
-\begin{algorithm}[H]
- \KwData{$N$, $\varepsilon_1$}
- \KwResult{$\tilde{N}$, $\boldsymbol{\sigma_i}$, $\mathbf{x}$}
- $\tilde{N}=N$\;
- compute $\theta$\;
- \While{$\frac{\lVert\boldsymbol{\sigma} \rVert}{\lVert 
-E[\theta]\rVert}<\varepsilon_1$}{
-\vspace{0.1cm}
-  $\tilde{N}=\tilde{N}+N$\;
-  compute $E[\theta]=\mathbf{x}$\;
- }
- return $\theta$, $\mathbf{x}$, $\boldsymbol{\sigma}$\; 
- \caption{A posteriori adaptive Adjoint Monte Carlo}
-\end{algorithm}
-
-\section{Sets of matrices for which the convergence of the MC linear solver is 
-guaranteed}
+\subsection{Classes of Matrices with Guaranteed Convergence}
 
@@
 
- \subsection{Strictly diagonally dominant matrices}
+\subsubsection{Strictly Diagonally Dominant}
+
  One of the these sets is
@@
 
- \subsection{Generalized diagonally dominant matrices (GDDM)}
+\subsubsection{Generalized Diagonally Dominant}
 
@@
 
- \subsection{"Block-diagonally dominant"-like matrices}
+\subsubsection{Block Diagonally Dominant}
+
  In this section we wonder in which situations a block diagonal preconditioning
@@
 
-\subsubsection*{Forward method}
+\paragraph{Forward method}
 By assuming that all $n_i$'s have the same value we may define
@@
 
-\subsubsection*{Adjoint method} 
+\paragraph{Adjoint method}
 Analogously to the Forward method, if we define
@@
 
-\section{Linear solvers with hybrid schemes}
-
-Even if the standard Monte Carlo, in both its Forward and Adjoint formulation, 
-has the advantage of being embarrassingly parallelizable (because of the 
-independence of the samplings employed), it has a drawback as 
-well. In fact its slow rate of convergence, associated with the 
-$\frac{1}{\sqrt{N}}$ behavior predicted but the CLT, prevents its use for 
-problems of actual interest in applied linear algebra (with many d.o.f's).
-
-Moreover, when the spectral radius of the iteration matrix is close to the 
-unit 
-value, Monte Carlo direct methods reach convergence after an excessive number 
-of steps for each 
-permutation. In particular this aspect will be pointed out more by some 
-numerical tests analyzed in the upcoming sections of this work.
-This motivation makes the standard Mote Carlo paradigm even more 
-unpractical for solving linear systems associated with realistic applications. 
-\newline
-
-To cope with this, in the early 2000's Evans et al. introduced a new method: 
-the 
-\textit{Monte Carlo Synthetic Acceleration (MCSA)} 
-(\cite{ESW2013} and \cite{EMSH2014}). Earlier contributions on these direction 
-are attributed also to other scientists (e.g. \textit{Sequential Method} by 
-J.H. Halton). However the MCSA is the algorithm we are going to focus on more 
-hereafter. If 
-we think about a fixed point formulation of the problem
-such as $\mathbf{x}=H\mathbf{x}+\mathbf{b}$, then the Monte Carlo Synthetic 
-Acceleration assumes this form
+\subsection{Adaptive Methods}
+
+In formulas (\ref{dir_mean}) and (\ref{adj_mean}) the estimation of the
+solution to
+the linear system (\ref{linsys_prec}) involves infinite sums. Of course this is
+not viable for
+computational issues, therefore it is necessary to resort to truncation of
+the sums through proper criteria. In details, the truncation implies the
+employment of a finite number of permutations and a finite number of
+steps for each one of the permutations computed. Therefore it is necessary to
+truncate at two different levels to decide both how many permutations to employ
+and how many steps each one of them should be kept running.
+
+\subsubsection{History Length}
+
+At first we focus on the definition of a weight cut-off
+in order to decide where to terminate the histories. It may be reinterpreted as
+a quantity guiding us to decide how many terms of the Neumann
+series to consider. \newline
+
+The goal is to apply the cutoff via an automatic procedure, without requiring
+the intervention of the user directly.
+Therefore we are looking for an integer $m$ such that
+\[
+\tilde{\theta}_i=E\bigg[\sum_{\ell=0}^m W_{\ell}
+f_{k_{\ell}}\bigg]=x_i=\sum_{\ell=0}^m
+\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n
+P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},
+k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots
+w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}
+\]
+
+and
+
+\[
+\tilde{\theta}_j=E\bigg[\sum_{\ell=0}^m W_{\ell}\delta_{k_{\ell},
+j}\bigg]=\sum_{\ell=0}^{m}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n
+f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
+w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}
+\]
+
+are good approximations of (\ref{dir_mean}) and (\ref{adj_mean}) respectively.
+
+In \cite{Slattery2013} there is a criterion for the cutoff of the random walk
+applicable to both the Forward and the Adjoint method. It requires the set up
+of a relative cutoff threshold $W_c$ and to look for a step $m$ such that
+\begin{equation}
+W_m \le W_c W_0.
+\label{cutoff}
+\end{equation}
+$W_0$ is the value of the weight at the initial step of the random walk and
+$W_m$ is the value of the weight after $m$ steps.
+
+\subsubsection{Number of Random Walks}
+
+We now focus on quantities that might lead to a proper
+number of permutations to compute, in order to get a prescribed order of
+accuracy.
+As regards the Forward method, we know that the expression for the variance
+is defined in formula (\ref{dir_var}).
+
+In this context, a reasonable criterion to determine the number $\tilde{N}_i$
+of random walks to be run is setting a threshold $\varepsilon_1$ and determine
+\begin{equation}
+\tilde{N}_i \quad s.t.\quad \frac{\sqrt{Var[\theta_i]}}{\lvert
+E[\theta_i]\rvert}<\varepsilon_1, \quad i=1,\cdots,n.
+\label{forward_adapt}
+\end{equation}
+The dependence of $Var[\theta_i]$ and $E[\theta_i]$ on $\tilde{N}_i$, which
+seems
+to be absent in the previous formula, is highlighted by the fact that
+$\theta_i$
+is estimated by fixing a finite number of histories.
+Therefore we are controlling the relative standard deviation requiring it not
+to be too large. In other words we are pursuing a statistical setting where the
+uncertainty factor is not dominating over the expected value.
+This simple adaptive approach can be applied for the estimation of each
+component $x_i$. Hence different number of histories may be employed to
+compute different entries of the solution vector. \newline
+
+As concerns the Adjoint method, the estimation of the variance for each entry
+is reproduced by Formula (\ref{adj_var}).
+
+A possible adaptive selection of $\tilde{N}$, in this situation, is
+\begin{equation}
+\tilde{N} \quad s.t. \quad \frac{\lVert
+\boldsymbol{\sigma}_{\tilde{N}}\rVert_1}{\lVert
+\mathbf{x}\rVert_1}<\varepsilon_1,
+\label{adjoint_adapt}
+\end{equation}
+where $\boldsymbol{\sigma}$ is a vector whose entries are
+$\boldsymbol{\sigma}_{\tilde{N},i}=Var[\theta_i]$.
+
+What introduced above can be exploited in order to build
+an a posteriori adaptive algorithm, capable to identify the minimal value of
+$\tilde{N}$ that verifies (\ref{forward_adapt}) or (\ref{adjoint_adapt})
+respectively.
+
+Below there are the pseudo-codes associated with both Forward and Adjoint Monte
+Carlo.
 
 \begin{algorithm}[H]
- \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
- \KwResult{$x_{num}$}
- $\mathbf{x}^{l}=\mathbf{x}_0$\;
- \While{not reached convergence}{
-  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
-  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
-  $\delta \mathbf{x}^{l+\frac{1}{2}}=(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$\;
-  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta 
-\mathbf{x}^{l+\frac{1}{2}}$\;
+ \KwData{$N$, $\varepsilon_1$}
+ \KwResult{$\tilde{N}_i$, $\sigma_i$, $x_i$}
+ For each entry of the solution vector (for $i=1,\cdots,n$) \;
+ $\tilde{N}_i=N$\;
+ compute $\theta_i$\;
+ \While{$\frac{\sigma_i}{\lvert
+E[\theta_i]\rvert}<\varepsilon_1$}{
+\vspace{0.1cm}
+  $\tilde{N}=\tilde{N}+N$\;
+  compute $E[\theta_i]=x_i$\;
  }
- $x_{num}=x^{l+1}$\; 
- \caption{Monte Carlo Synthetic Acceleration}
+ return $\theta_i$, $x_i$, $\sigma_i$\;
+ \caption{A posteriori adaptive Forward Monte Carlo}
 \end{algorithm}
-The Monte Carlo method is used to compute the updating contribution $\delta 
-\mathbf{x}^{l+\frac{1}{2}}$.
-
-The role of the first line of the iterative process has the goal to compress 
-the residual of the previous iteration. This is the actual difference 
-between the MCSA and the Sequential Monte Carlo. Once this done, the corrected 
-residual 
-is employed to compute an update $\delta x^{1+\frac{1}{2}}$ to the solution via 
-one of the Monte Carlo techniques introduced before. The utility of compressing 
-the residual at the very beginning of the iteration will be shown in the two 
-last test cases for numerical validations.
-The intent of MCSA is to designate most of the time employed by the 
-computation to the accomplishment of the stochastic part of the algorithm. 
-This 
-is useful to decrease the probability of fault occurrence in the non-resilient 
-steps, which is represented by the non-stochastic component of the numerical 
-scheme presented above. In particular, the higher is the percentage of time 
-spent on the computation of $\delta x^{1+\frac{1}{2}}$ via Monte Carlo, the 
-better it should be in terms of resilience.
-
-\subsection{Preconditioning techniques}
+
+\begin{algorithm}[H]
+ \KwData{$N$, $\varepsilon_1$}
+ \KwResult{$\tilde{N}$, $\boldsymbol{\sigma_i}$, $\mathbf{x}$}
+ $\tilde{N}=N$\;
+ compute $\theta$\;
+ \While{$\frac{\lVert\boldsymbol{\sigma} \rVert}{\lVert
+E[\theta]\rVert}<\varepsilon_1$}{
+\vspace{0.1cm}
+  $\tilde{N}=\tilde{N}+N$\;
+  compute $E[\theta]=\mathbf{x}$\;
+ }
+ return $\theta$, $\mathbf{x}$, $\boldsymbol{\sigma}$\;
+ \caption{A posteriori adaptive Adjoint Monte Carlo}
+\end{algorithm}
+
+\subsection{Preconditioning}
+
 In this section we focus on the pursuit of preconditioners that facilitate the
@@
 
-\section{Numerical results}
-
-All the numerical results shown in the following sections are computed with 
-\texttt{Matlab} on a standard laptop. The simulations are accomplished in a 
-serial mode.
-In the 
-first subsection the standard Monte Carlo solver is employed, while in the 
-subsequents ones results are obtained with the MCSA scheme.
-
-\subsection{Check on the effectiveness of the adaptive criteria}
+\section{Numerical Results}
+
+\subsection{Adaptive Methods}
+
 In this subsection we show some numerical tests to validate the utility of the
@@
 
-\subsection{A comparison between different preconditioners and their effect 
-on $\rho(H)$ and $\rho(\hat{H})$}
+\subsection{Preconditioning Approaches}
 
@@
 
-\subsection{The Poisson equation solved with MCSA}
+\subsection{Hybrid Methods}
+
+\subsubsection{Poisson Problem}
+
 In the current subsection a Poisson problem is taken into account:
@@
 
-\subsection{A diffusion reaction problem solved with MCSA}
+\subsubsection{Reaction-Diffusion Problem}
+
 By modifying the equation of the previous test case, we consider now
@@
 
-\subsection{A parabolic problem solved with MCSA}
+\subsubsection{Parabolic Problem}
+
 In this subsection we restrict our treatise to the analysis of parabolic
@@
 
-Other preconditioners 
-have 
-been 
-tested besides the sparse approximate inverse preconditioner, such as block 
-diagonal 
-and incomplete LU factorization. However the fill-in effect introduced by the 
-application of 
-these preconditioners compromises the sparsity of the linear system.
-
-\section{Conclusions and future developments}
+\section{Conclusions and Future Work}
+
 The work presented in this paper has been accomplished to provide an overview

### 2015-09-16 MAX_reports/MC_solvers_article/MC_solvers.tex
Breaking paper into multiple files for easier editing/collaborating.

@@
 
-\section{Introduction}
-For a long period the
-pursuit of a reliable solver for linear and nonlinear algebraic systems has
-been one of the most important issues in Numerical Analaysis. The
-mathematical properties of the model (e.g. likely sparsity pattern of the
-matrix) have urged scientists to look for efficient and pattern-driven
-solvers, capable to cope with curse of dimensionality and high scalability
-requests.
-
-As scalability computing move towards exascale facilities (which stands for
-machines computing up to $10^{18}$ flops), new numerical schemes suitable for
-this
-kind of architectures are strongly demanded. In fact the purpose is to combine
-high
-fidelity of the results with an optimized use of hardware resources at hand.
-
-Recently new algorithms that combine statistical and numerical techniques have
-been developed. Even though \textit{numerical efficiency} has been sacrificed
-sometimes, the advantages in terms of \textit{robustness} of these methods make
-it worth carrying on studies in this direction.
-A class of methods that may be employed for this purpose is represented by
-Monte Carlo linear solvers.
-
-These algorithms are the main topic covered in the work presented here,
-starting from what was developed so far by some of
-the authors who gave significant contributions (see \cite{Hal1962},
-\cite{Hal1994},
-\cite{DA1998}, \cite{DVA2001}, \cite{AADBTW2005},\cite{ESW2013} and
-\cite{EMSH2014}). \newline
-
-As underlined in \cite{DA1998}, Monte Carlo methods may be split into two
-classes: \textit{direct methods} (\cite{DA1998}, \cite{DVA2001}) and
-\textit{iterative methods} (\cite{Hal1962},
-\cite{Hal1994}, \cite{ESW2013}
-and \cite{EMSH2014}). The first are characterized by a merely stochastic
-scheme,
-therefore the provided error with respect to the exact solution is made of
-just a stochastic component. The second ones, instead, combine numerical and
-statistical schemes generating two types of error: a
-\textit{stochastic} one and a \textit{systematic} one. It does not
-mean
-that
-it will be always simple to recognize them separately. However it is important
-to
-be
-aware of this intrinsic structure, in order to target what is the part of the
-scheme that requires a refinement (e.g. increase of numerical iterations
-rather than random walks).\newline
-
-The paper is organized as follows. In Section 2 we describe all the
-viable stochastic settings for the definition of the Markovian process. In
-Section 3 we reinterpret probabilistic facts in terms of spectral properties of
-the linear system to be solved. Section 4 is dedicated to necessary conditions
-and sufficient conditions for the convergence of the algorithms. In Section 5
-different adaptive approaches are presented, in order to truncate the random
-walks and to decide how many samplings to use. Section 6 regards the study
-of matrices for which the convergence of the stochastic schemes is facilitated.
-Then in Section 7 a hybrid scheme combining numerical and probabilistic
-techniques is described: the \textit{Monte Carlo Synthetic Acceleration}.
-Numerical results are presented in Section 8 and 9, while Section 10 is
-dedicated to
-conclusive remarks and goals for the future.
-
-\section{Stochastic Linear Solvers}
-
-We consider the solution of systems of linear equations of the form
-\begin{equation}
-A \mathbf{x}=\mathbf{b},
-\label{linsys}
-\end{equation}
-where $A\in \mathbb{R}^{n\times n}$, $\mathbf{x}$, $\mathbf{b} \in
-\mathbb{R}^n$.
-
-By applying a left preconditioning, the system (\ref{linsys}) gets the
-following shape:
-\begin{equation}
-P^{-1}A \mathbf{x}=P^{-1}\mathbf{b},
-\label{linsys_prec}
-\end{equation}
-(\ref{linsys_prec}) can be
-reinterpreted as a fixed point scheme
-\begin{equation}
- \mathbf{x}=H\mathbf{x}+\mathbf{f}.
- \label{fixedpoint}
-\end{equation}
-where $H=I-P^{-1}A$ and $\mathbf{f}=P^{-1}\mathbf{b}$.
-
-Assuming that the spectral radius $\rho(H)<1$, the solution to
-(\ref{fixedpoint}) can be written in terms of a power series of
-$H$ (Neumann series)
-\[
-\mathbf{x}=\sum_{i=0}^\infty H^i\mathbf{f}.
-\]
-Therefore the fixed point scheme generates a sequence of approximate solutions
-$\{\mathbf{x}^{(k)}\}_{k=0}^{\infty}$ which converges to the exact solution
-regardless of the initial guess $\mathbf{x}_0$.
-
-By restricting the attention to a single component of $\mathbf{x}$ we
-have
-\begin{equation}
-x_i=\sum_{\ell=0}^\infty \sum_{k_1}^n\sum_{k_2}^n\cdots \sum_{k_{\ell}=1}^n
-H_{k_0,k_1}H_{k_1,k_2}\cdots H_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
-\label{forward}
-\end{equation}
-The last equation can be reinterpreted as the realization of an estimator
-defined on a random walk.
-
-Let us start considering a random walk whose
-state space $S$ is characterized by the set of indices of the forcing term
-$\mathbf{f}$:
-\[
-S=\{1,2,\cdots, n\} \subset \mathbb{N}.
-\]
-Each $i$-th step of the random walk has a random variable
-$k_i$ associated with it. The realization of $k_i$ represents the index of the
-component of $\mathbf{f}$
-which is visited in the current step of the random walk.
-
-The way the transition probability is built and the way the initial state of
-the random walk is chosen give birth to two different approaches we are
-going to use: the \textit{Forward} and \textit{Adjoint} methods.
-
-\subsection{Forward Method}
-
-The goal is to evaluate a functional such as
-\[
-J(\mathbf{x})=(\mathbf{h},\mathbf{x})=\sum_{i=1}^n h_i x_i.
-\]
-where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in
-$\mathbb{R}^n$ of the functional $J$.
-We can use it to build the initial probability $\tilde{p}:
-S\rightarrow [0,1]$ of the random walk such that
-\[
-\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n \lvert
-h_i\rvert}.
-\]
-It is important to stress out that the role of vector $\mathbf{h}$ is
-restricted to the construction of the initial probability. Once that is done,
-it is
-not used anymore in the definition of the stochastic process.
-A possible choice for the transition probability $P$ can be
-\[
-p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert H_{i,j}\rvert}{\sum_{k=1}^n
-\lvert H_{i,k}\rvert}.
-\]
-where $\tilde{p}(\cdot,i):S\rightarrow [0,1]$ $\forall i\in S$.
-A related sequence of random variables $w_{i,j}$ can be defined
-such that
-\[
-w_{i,j}=\frac{H_{i,j}}{P_{i,j}}.
-\]
-The probability distribution of the random variables $w_{i,j}$ is represented
-by the transition matrix that rules the stochastic process. The $w_{i,j}$'s
-just introduced can be used to build one more sequence
-of random variables.
-At first we introduce quantities $W_j$
-\[
-W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_j=W_{j-1} w_{i,j}, \quad
-j=1,\cdots, i.
-\]
-By defining
-\[
-X(\nu)=\sum_{m=0}^k W_m f_{i_m}
-\]
-as the random variable associated with a specific permutation $\nu$, we can
-define the estimator $\theta_i$ such as
-\[
-\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu).
-\]
-The integer $n$ represents the size of the solution vector to (\ref{linsys})
-and
-the
-index
-$i$
-is referred to the component of the solution vector we want to compute.
-$P_{\nu}$ is the probability associated with a specific permutation of the
-random walk.
-It can be proved that
-\[
-E[W_i f_{k_i}]=(\mathbf{h},H^i\mathbf{f}), \quad i=0,1,2,\cdots
-\]
-and
-\[
-\theta_i=E\bigg[\sum_{i=0}^\infty W_i f_{k_i}\bigg]=(\mathbf{h},\mathbf{x}).
-\]
-
-A possible choice for $\mathbf{h}$ is a vector of the standard basis. This
-would correspond in setting manually the initial state of the random walk,
-which turns the related initial probability into a Kronecker delta
-\[
-\tilde{p}(k_0=i)=\delta_{i,j}.
-\]
-By doing so, we have
-\begin{equation}
-\theta_i=E\bigg[\sum_{\ell=0}^\infty W_{\ell}
-f_{k_{\ell}}\bigg]=x_i=\sum_{l=0}^\infty
-\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n
-P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},
-k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots
-w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
-\label{dir_mean}
-\end{equation}
-
-As regards the variance, we remember that the following relation holds:
-\begin{equation}
-Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
-f_{k_{\ell}}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
-f_{k_{\ell}}^2\bigg] - \bigg (E\bigg[\sum_{\ell=0}^\infty W_{\ell}
-f_{k_{\ell}}\bigg]\bigg )^2.
-\label{dir_var}
-\end{equation}
-
-Hence the variance can be computed as the difference between the second
-moment of the random variable and the square of the first moment.\newline
-
-In order to apply the Central Limit Theorem (CLT) to the estimators defined
-above, we must require that
-the estimators have both finite expected value and variance. This is
-equivalent in
-checking the finiteness of the expected value and of the second moment.
-Therefore we have to impose the following conditions:
-
-\begin{equation}
- E\bigg[\sum_{\ell=0}^\infty W_{\ell} f_{k_{\ell}}\bigg]<\infty
-\end{equation}
-
-\begin{equation}
- E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
-f_{k_{\ell}}^2\bigg]<\infty
-\end{equation}
-
-The Forward method presented above, however, has the limitation of employing an
-entire set of permutations to estimate just a single entry of
-the solution at a time. Therefore, in order to estimate the entire solution
-vector for \ref{linsys_prec}, we have to employ as many sets of permutation as
-the number of entries composing the vector. Of course this might turn to be
-extremely expensive both in terms of time and resources for most of the
-challenging problems coming from applications. This is essentially the reason
-why usually it is preferred to adopt another stochastic algorithm which is
-explained here following: the \textit{Adjoint} Monte Carlo method.
-
-\subsection{Adjoint Method}
-
-A second Monte Carlo method can be derived by considering the linear system
-adjoint to (\ref{linsys_prec})
-\begin{equation}
-(P^{-1}A)^T\mathbf{y}=\mathbf{d},
-\label{adj}
-\end{equation}
-where $\mathbf{y}$ and $\mathbf{d}$ are the adjoint solution and source term.
-
-By reformulating the fixed point scheme, introducing initial probability,
-transition probability and weights
-sequences in the same fashion as done before, the expected value for the
-estimator becomes
-\begin{equation}
-\theta_j=E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell},
-j}\bigg]=\sum_{\ell=0}^{\infty}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n
-f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
-w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}.
-\label{adj_mean}
-\end{equation}
-This estimator is known in literature as \textit{collision} estimator.
-
-We can notice now that the component of the source vector, in each
-permutation of the random walk, is associated with the initial step. The
-Kronecker delta
-at the end of the series stands for a filter. It means that if the goal is to
-estimate the $j$-th component of the solution vector, just a subset of all the
-permutations gives a contribution to this. The only permutations considered are
-those ones that currently reside on the index associated with the entry to be
-estimated.
-
-The variance is
-
-\begin{equation}
-Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
-f_{k_0}\delta_{k_{\ell},j}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
-f_{k_0}^2\delta_{k_{\ell},j}\bigg ] - \bigg (E\bigg[\sum_{\ell=0}^\infty
-W_{\ell}
-f_{k_0}\delta_{k_{\ell},j}\bigg]\bigg )^2\quad j=1,\cdots,n
-\label{adj_var}.
-\end{equation}
-
-On the same lead of what done for the Forward method, there is need to impose
-finiteness of the expected value and second moment. Therefore the following
-conditions must be verified:
-
-\begin{equation}
- E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell},
-j}\bigg]<\infty \quad j=1,\cdots,n
-\end{equation}
-and
-\begin{equation}
- E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
-f_{k_0}^2\delta_{k_{\ell},j}\bigg]<\infty \quad j=1,\cdots,n.
-\end{equation}
-
-The main advantage of this method, compared to the Forward one, consists in the
-fact that a single set of permutation is used to estimate the entire solution
-vector. Therefore, in general, this algorithm has to be preferred to the
-previous one since it enables to reduce dramatically the computational burden.
-
-In literature another estimator is employed along with the Adjoint Monte Carlo
-method, the so called \textit{expected value} estimator. Its
-formulations is as follows:
-
-\begin{equation}
-\theta_j=E\bigg[f_j + \sum_{\ell=0}^\infty
-W_{\ell}H_{k_{\ell}, j}^T\bigg]=f_j
-+ \sum_{\ell=0}^{\infty}\sum_{k_1}^n\sum_{k_2} ^n\cdots\sum_ { k_ { \ell}}^n
-f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
-w_{k_{\ell-1},k_{\ell}}H_{k_{\ell},j}^T.
-\label{adj_mean1}
-\end{equation}
-
-Differently from the collision estimator, the expected value estimator averages
-the deterministic contribution of the iteration matrix over all the potential
-states $j$ that might be reached from the current state $\ell$. The variance
-in this case becomes:
-
-\begin{equation}
-Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
-f_{k_0}H_{k_{\ell},j}^T\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
-f_{k_0}^2 {H_{k_{\ell},j}^T}^T\bigg ] - \bigg (E\bigg[\sum_{\ell=0}^\infty
-W_{\ell}
-f_{k_0}H_{k_{\ell},j}^T\bigg]\bigg )^2\quad j=1,\cdots,n
-\label{adj_var}.
-\end{equation}
-
-In the sections dedicated to the numerical tests, both the collision estimator
-and the expected value are employed, in order to test their effectiveness.
-
-\subsection{Hybrid Stochastic/Deterministic Methods}
-
-Even if the standard Monte Carlo, in both its Forward and Adjoint formulation,
-has the advantage of being embarrassingly parallelizable (because of the
-independence of the samplings employed), it has a drawback as
-well. In fact its slow rate of convergence, associated with the
-$\frac{1}{\sqrt{N}}$ behavior predicted but the CLT, prevents its use for
-problems of actual interest in applied linear algebra (with many d.o.f's).
-
-Moreover, when the spectral radius of the iteration matrix is close to the
-unit
-value, Monte Carlo direct methods reach convergence after an excessive number
-of steps for each
-permutation. In particular this aspect will be pointed out more by some
-numerical tests analyzed in the upcoming sections of this work.
-This motivation makes the standard Mote Carlo paradigm even more
-unpractical for solving linear systems associated with realistic applications.
-\newline
-
-To cope with this, in the early 2000's Evans et al. introduced a new method:
-the
-\textit{Monte Carlo Synthetic Acceleration (MCSA)}
-(\cite{ESW2013} and \cite{EMSH2014}). Earlier contributions on these direction
-are attributed also to other scientists (e.g. \textit{Sequential Method} by
-J.H. Halton). However the MCSA is the algorithm we are going to focus on more
-hereafter. If
-we think about a fixed point formulation of the problem
-such as $\mathbf{x}=H\mathbf{x}+\mathbf{b}$, then the Monte Carlo Synthetic
-Acceleration assumes this form
-
-\begin{algorithm}[H]
- \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
- \KwResult{$x_{num}$}
- $\mathbf{x}^{l}=\mathbf{x}_0$\;
- \While{not reached convergence}{
-  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
-  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
-  $\delta \mathbf{x}^{l+\frac{1}{2}}=(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$\;
-  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta
-\mathbf{x}^{l+\frac{1}{2}}$\;
- }
- $x_{num}=x^{l+1}$\;
- \caption{Monte Carlo Synthetic Acceleration}
-\end{algorithm}
-The Monte Carlo method is used to compute the updating contribution $\delta
-\mathbf{x}^{l+\frac{1}{2}}$.
-
-The role of the first line of the iterative process has the goal to compress
-the residual of the previous iteration. This is the actual difference
-between the MCSA and the Sequential Monte Carlo. Once this done, the corrected
-residual
-is employed to compute an update $\delta x^{1+\frac{1}{2}}$ to the solution via
-one of the Monte Carlo techniques introduced before. The utility of compressing
-the residual at the very beginning of the iteration will be shown in the two
-last test cases for numerical validations.
-The intent of MCSA is to designate most of the time employed by the
-computation to the accomplishment of the stochastic part of the algorithm.
-This
-is useful to decrease the probability of fault occurrence in the non-resilient
-steps, which is represented by the non-stochastic component of the numerical
-scheme presented above. In particular, the higher is the percentage of time
-spent on the computation of $\delta x^{1+\frac{1}{2}}$ via Monte Carlo, the
-better it should be in terms of resilience.
-
-\subsubsection{Sequential Monte Carlo}
-
-\subsubsection{Monte Carlo Synthetic Acceleration}
-
-\section{Convergence Behavior of Stochastic Methods}
-
-The requests imposed on the expected value and second moment of the estimators
-can be reformulated in a purely deterministic setting by looking at the
-spectral radii of some properly defined matrices.
-
-For instance the condition of finiteness of the expected value can be
-reformulated by requiring
-\begin{equation}
- \rho(H)<1
- \label{rho_h},
-\end{equation}
-
-where $H$ is the iteration matrix of the fixed point scheme.
-In fact both equations (\ref{dir_mean}) and (\ref{adj_mean}) can be considered
-as
-power series in terms of $H$ and the spectral radius of $H$ smaller
-than 1 is a necessary and sufficient condition for the Neumann series to
-converge.
-
-By analogy we aim at reproducing a similar reasoning for the second moment of
-the estimator, since we want it to be finite as well.
-By looking at the equations of the variance (\ref{dir_var}) and (\ref{adj_var})
-for
-the Forward and the Adjoint method respectively, we notice that the second
-moment can be reinterpreted as a power series with respect to the matrices
-defined as follows:
-
-\[
- \hat{H}_{i,j}=\frac{H_{i,j}^2}{P_{i,j}} \quad \text{- Forward Method}.
-\]
-
-or
-
-\[
- \hat{H}_{i,j}=\frac{H_{j,i}^2}{P_{i,j}} \quad \text{- Adjoint Method}.
-\]
-
-Accordingly to the necessary and sufficient condition for the power series to
-converge, we require
-\begin{equation}
- \rho(\hat{H})<1.
- \label{rho_star}
-\end{equation}
-
-Condition (\ref{rho_h}) must be required for a generic fixed point scheme to
-reach convergence, whereas the extra condition (\ref{rho_star}) is typical of
-the stochastic schemes presented in this work. Having two conditions to
-verify makes the numerical treatment of the linear system via MC very
-difficult. We will discuss this in details later in the treatise. \newline
-
-\subsection{Necessary and Sufficient Conditions}
-
-Here below some results presented in
-\cite{MASC2013} about necessary conditions and sufficient
-conditions for
-convergence are enunciated. In particular a suitable choice for constructing a
-transition probability is discussed to guarantee and speed-up convergence.
-
-We remember that the construction of the transition probability must satisfy
-some constraints such as
-
-\[
-\begin{cases}
-  P_{i,j}\ge 0 \\
- \sum_{j=1}^N P_{i,j}=1
-\end{cases}
-\]
-
-plus one more restriction which varies depending on the particular Monte Carlo
-approach chosen
-
-\begin{align*}
-& \text{Forward Method:} \quad H_{i,j}\ne 0 \Rightarrow P_{i,j}\ne 0 \\ &
-\text{Adjoint Method:} \quad H_{j,i}\ne 0 \Rightarrow P_{i,j}\ne 0
-\end{align*}.
-
-The previous restrictions on the attainable values for the entries of the
-transition probability are called \textit{transition conditions}.
-
-At first an introductory theoretical result is shown, necessary to motivate the
-choices made afterwards for the choice of the transition probability.
-
-\begin{thm}
- Consider a vector $a=(a_1,a_2,\cdots,a_N)^T$ where at least one element is
-nonzero, $a_k\ne0$ for some $k\in\{1,\cdots,N\}$.
-\begin{itemize}
- \item For a probability vector $p=(p_1,p_2,\cdots,p_N)^T$ satisfying the
-transition conditions, the lower bound of $\displaystyle
-\sum_{k=0}^N\frac{a_k^2}{p_k}$ is $\bigg(\sum_{k=1}^N \lvert a_k\rvert\bigg)^2$
-\item There always exists a probability vector such that $\displaystyle
-\sum_{k=0}^N\frac{a_k^2}{p_k}\ge c\ge 1$ for all $c>1$.
-\end{itemize}
-\label{lemma}
-\end{thm}
-
-It might be observed that the minimum of the quantity $\displaystyle
-\sum_{k=0}^N\frac{a_k^2}{p_k}$ is attained when the probability vector is
-defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
-a_k\rvert}$. It implies that the $\infty$-norm of $\hat{H}$ is minimized as
-much
-as possible. \newline
-
-For the sake of simplicity let us introduce now a generic random walk
-truncated at a certain $k$-th step
-\[
- \gamma_k:\; r_0\rightarrow r_1 \rightarrow r_2 \rightarrow \cdots \rightarrow
-r_k
-\]
-and the statistical estimator
-\[
- X(\gamma_k)=\frac{H_{r_0,r_1}H_{r_1,r_2}\cdots
-H_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots P_{r_{k-1},r_k}}f_{r_k}
-\]
-which is associated with the Forward Method.
-
-Then the following theorem holds
-\begin{thm}\textit{(Suited to the Forward Method)}
-
-Let $H\in \mathbb{R}^{n\times n}$ such that $\lVert H\rVert_{\infty}<1$.
-Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the
-$k$-th step. Then,
-there always exists a
-transition matrix $P$ such that
-$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and
-$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
-\label{for_thm}
-\end{thm}
-
-If we introduce the estimator
-\[
- X(\gamma_k)=\frac{H^T_{r_0,r_1}H^T_{r_1,r_2}\cdots
-H^T_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots
-P_{r_{k-1},r_k}}sign(f_{r_0})\lVert \mathbf{f}\rVert_1
-\]
-which is associated with the Adjoint Method, then we can state a
-theorem specular to \ref{for_thm}.
-
-\begin{thm}\textit{(Suited to the Adjoint Method)}
-
- Let $H\in \mathbb{R}^{n\times n}$ with $\lVert H\rVert_{1}<1$.
-Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the
-$k$-th step. Then,
-there always exists a
-transition matrix $P$ such that
-$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and
-$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
-\label{adj_thm}
-\end{thm}
-
-In particular, relying on what guaranteed by Theorem \ref{lemma}, in both
-Theorems \ref{for_thm} and \ref{adj_thm} the thesis holds by picking the MAO
-transition probability.
-
-These theoretical results represent sufficient conditions for the convergence
-of the
-Forward and Adjoint Monte Carlo and they can be easily checked.
-However requiring $\lVert
-H\rVert_{\infty}<1$ or $\lVert H\rVert_1<1$ may be too demanding.
-Indeed, even if there exist some preconditioners able to guarantee this
-condition (e.g. approximate inverse preconditioners with a proper value of
-dropping
-tolerance), it may entail to compromise the sparsity of the
-iteration
-matrix.
-
-Thusly, in situations where the hypotheses of Theorems \ref{for_thm} and
-\ref{adj_thm} are too demanding to be verified, the necessary and sufficient
-condition
-on the
-spectral radius of $\hat{H}$ is the only criterion to take as reference.
-Nonetheless this constraint consists in
-computing the spectral radii of two matrices and the calculation
-might be very costly. Moreover the computation of the spectral radii is
-affected by resilience issues as well as
-a deterministic solver for the resolution of
-a linear system. Therefore the original difficulty is transferred without being
-actually tackled.
-
-\subsubsection{Forward MC}
-
-\subsubsection{Adjoint MC with Collision Estimator}
-
-\subsubsection{Adjoint MC with Expected-Value Estimator}
-
-\subsection{Construction of Transition Probabilities}
-
-The way the transition probability is defined is extremely important. In fact
-in some circumstances it might determine whether the stochastic scheme
-converges or not. Two very simple ways of defining it are listed here
-below. These are the approaches adopted typically in literature.
-
-\subsubsection{Uniform probabilities}
-With this approach transition matrix $P$ is such that all the non-zero elements
-in each row have equal probability of occurring.
-\[
-\text{Forward}:\;P_{i,j}=
-\begin{cases}
-0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{i,j}=0 \\
-\frac{1}{\#(\text{non-zeros in the row)}} \quad \text{if} \quad H_{i,j}\ne 0
-\end{cases}\quad
-\]
-\[
-\text{Adjoint}:\;P_{i,j}=
-\begin{cases}
-0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{j,i}=0 \\
-\frac{1}{\#(\text{non-zeros in the column)}} \quad \text{if} \quad H_{j,i}\ne 0
-\end{cases}
-\]
-The Monte Carlo approach resorting to this definition of the transition matrix,
-in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).
-
-\subsubsection{Weighted probabilities}
-An alternative way of defining the transition matrix consists in attributing
-probabilities to all the entries of the transition probability accordingly
-to the magnitude of the elements. For instance we may employ the following
-definition
-\[
-\text{Forward}: \;
-p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
-H_{i,j}\rvert^p}{\sum_{k=1}^n
-\lvert H_{i,k}\rvert^p}
-\]
-
-\[
-\text{Adjoint}: \;
-p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
-H_{j,i}\rvert^p}{\sum_{k=1}^n
-\lvert H_{k,i}\rvert^p}
-\]
-
-where $p\in \mathbb{N}$.
-This way of proceeding, for $p=1$, still basing on the work \cite{AADBTW2005},
-is called
-\textit{Monte Carlo Almost Optimal} (MAO).
-The reason why the almost optimal probability is called this way is associated
-with a notice from Theorem \ref{lemma}. Indeed from
-the mentioned lemma we know that the quantity $\displaystyle
-\sum_{k=0}^N\frac{a_k^2}{p_k}$ is minimized when the probability vector is
-defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
-a_k\rvert}$. This implies that the almost optimal probability minimizes the
-$\infty$-norm of $\hat{H}$ in the Forward method and the $1$-norm of $\hat{H}$
-in the
-Adjoint method. This is very important to be pointed out since these two norms
-are so far the best practical upper bounds for $\rho(\hat{H})$ that we can
-easily
-handle. In the Section dedicated to the numerical results it will be proved
-that the Almost Optimal Probability eases the convergence of the MC solver,
-even in cases when the employment of the Uniform Probability does not provide
-convergence at all.
-
-\subsection{Classes of Matrices with Guaranteed Convergence}
-
-As we have already seen, sufficient conditions for
-the convergence of the Monte Carlo linear solvers are very restrictive (see
-\cite{MASC2013}), whereas the necessary and sufficient condition requires the
-computation of
-$\rho(H)$and $\rho(\hat{H})$, which is very expensive in general. Once a
-proper
-preconditioner is picked (e.g. Approximate Inverse), $\rho(H)<1$ may
-be always guaranteed, whilst $\rho(\hat{H})<1$ is more
-problematic to be verified.
-
-In order to bypass the cost of these checks, it is useful to pursue for sets of
-matrices for
-which the Monte Carlo linear solver always converges, maybe with the use of an
-ad-hoc preconditioner.
-
-\subsubsection{Strictly Diagonally Dominant}
-
- One of the these sets is
-represented by strictly diagonally dominant matrices.
-
-\begin{defn}
- A matrix $A\in\mathbb{R}^{n\times n}$is strictly diagonally dominant (s.d.d)
-by rows if
- \begin{equation}
-    \lvert a_{ij}\rvert>\sum_{\substack{i=1\\ i\ne j}}^{i=n}\lvert
-a_{ij}\rvert
-\label{sddr}
- \end{equation}
-\end{defn}
-
-\begin{defn}
- A matrix $A\in\mathbb{R}^{n\times n}$is strictly diagonally dominant (s.d.d)
-by columns if
- \begin{equation}
-    \lvert a_{ij}\rvert>\sum_{\substack{j=1\\ j\ne i}}^{j=n}\lvert
-a_{ij}\rvert
-\label{sddc}
- \end{equation}
-\end{defn}
-
-When (\ref{sddr}) holds we can resort to a left diagonal preconditioning and we
-get an iteration matrix $H=I-D^{-1}A$ such that $\lVert H \rVert_{\infty}<1$.
-
-Introducing a MAO transition probability for the Forward Method we get
-\[
- P_{ij}=\frac{\lvert H_{ij}\rvert}{\sum_{k=1}^n\lvert H_{ik}\rvert}
-\]
-therefore the entries of $\hat{H}$ are defined as follows
-\[
- \hat{H}_{ij}=\frac{H^2_{ij}}{P_{ij}}=\lvert H_{ij}\rvert\bigg
-(\sum_{k=1}^n\lvert
-H_{ik}\rvert\bigg ).
-\]
-Therefore
-\[
- \sum_{j=1}^n\lvert \hat{H}_{ij} \rvert = \sum_{j=1}^n \hat{H}_{ij} =\bigg
-(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg )\bigg (\sum_{k=1}^n\lvert
-H_{ik}\rvert\bigg ) =\bigg
-(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg ) ^2 <1 \quad \forall i=1,\cdots, n.
-\]
-This automatically implies that $\rho(\hat{H})<=\lVert
-\hat{H}\rVert_{\infty}<1$. Thus the Forward Monte Carlo converges.
-However nothing a priori can be
-stated about the behavior of the Adjoint method.\newline
-
-Instead if (\ref{sddc}) holds we can resort to a right diagonal preconditioning
-and we
-get an iteration matrix $H=I-AD^{-1}$ such that $\lVert H \rVert_{1}<1$.
-In this case, by following a similar reasoning to the one made before, we
-conclude that the Adjoint Method convergences thanks to $\lVert
-\hat{H}\rVert_1<1$,
-whereas nothing can be said in advance as concerns the Forward method.
-
-\subsubsection{Generalized Diagonally Dominant}
-
- Another set of matrices for which the convergence of
-the MC solvers is guaranteed is made of \textit{generalized diagonally
-dominant matrices}.
-
-\begin{defn}
-A square matrix $A\in\mathbb{R}^{n\times n}$ is said to be generalized
-diagonally dominant if
-\[
- \lvert a_{ii}\rvert x_i \ge \sum_{\substack{j=1\\j\ne i}}^n \lvert
-a_{ij}\rvert
-x_j, \quad i=1,\cdots,n
-\]
-for some positive vector $\mathbf{x}=(x_1,\cdots,x_n)^T$.
-\end{defn}
-
-A proper subset of the generalized diagonally dominant matrices is represented
-by $M$- matrices (see \cite{Ax1996}).
-There are many definitions of $M$-matrices that may be found on books and all
-of them are equivalent. The one presented in \cite{Saad} states the
-following
-
-\begin{defn}
-A matrix $A\in\mathbb{R}^{n\times n}$ is said to be an $M$-matrix if it
-satisfies the following four properties:
-\begin{itemize}
- \item $a_{i,i}>0$ for $i=1,\cdots,n$
- \item $a_{i,j}\le 0$ for $i\ne j$, $i,j=1,\cdots,n$
- \item $A$ is nonsingular
- \item $A^{-1}\ge 0$
-\end{itemize}
-\end{defn}
-
-A very important result about $M$-matrices enables to come up with a convergent
-splitting and, therefore, a convergent fixed point scheme.
-
-In particular Jacobi, Block Jacobi and Gauss Seidel  are
-convergent
-splitting for an $M$-matrix.
-However, the construction of a splitting like this is not enough to ensure the
-convergence of
-the Monte Carlo linear solver. In fact none of these conditions is sufficient
-to guarantee $\rho(\hat{H})<1$. Because of this obstacle, other options must
-be considered.\newline
-
-For this purpose we
-appeal to a result in \cite{Li2002}. In this paper the author presents an
-algorithm to transform a generalized diagonally dominant matrix with
-nonzero diagonal entries into strictly diagonally dominant. The
-algorithm works for a generic complex matrix $A\in\mathbb{C}^{n\times n}$.
-
-Here below we report the algorithm at hand\newline
-
-\begin{algorithm}[H]
- For a given complex matrix $A$, $a_{ii}\ne 0$, $i=1,\cdots, n$;\newline
- \begin{enumerate}
-  \item Compute $S_{i}=\sum_{\substack{j=1 \\ j\ne i}^{n}}\lvert
-a_{ij}\rvert$,
-$i=1,2,\cdots,n$
-\item Set $t=0$. For $i=1,2,\cdots, n$, if $\lvert a_{ii}\rvert>S_i$, then set
-$t=t+1$
-\item If $t=0$, then print "A is not a GDDM": END
-\item If $t=n$, then print "A is a GDDM": END
-\item \For{i=1,n}{$d_{i}=\frac{S_i+\varepsilon}{\lvert
-a_{ii}\rvert+\varepsilon} \quad
-\varepsilon>0,\quad j=1,2,\cdots,n$\;
-$a_{ji}=a_{ji}\cdot d_i$}
-\item Go to step 1.
- \end{enumerate}
- \caption{Algorithm to turn a GDDM matrix into a s.d.d by rows.}
-\end{algorithm}
-
-This approach turns a generalized diagonally dominant matrix into a
-strictly diagonally dominant matrix by rows. By substituting
-$\displaystyle S_{i}=\sum^{n}_{\substack{j=1 \\ j\ne i}}\lvert a_{ij}\rvert$
-at step 1 with
-$\displaystyle S_{i}=\sum^{n}_{\substack{i=1 \\ i\ne j}}\lvert a_{ij}\rvert$
-and by replacing
-$a_{ji}=a_{ji}\cdot d_i$ with $a_{ji}=a_{ji}\cdot d_j$ we obtain the algorithm
-that turns a GDDM matrix into a s.d.d. by columns.\newline
-
-Once we have applied this transformation to the matrix at hand into a s.d.d.,
-we can use the Monte Carlo linear solver which is ensured to converge.
-
-\subsubsection{Block Diagonally Dominant}
-
- In this section we wonder in which situations a block diagonal preconditioning
-might come up with a convergent Monte Carlo linear solver.
-By mimicking the computations we already showed previously, the iteration
-matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal
-preconditioning takes the form
-
-\[
- H=I-D^{-1}A=\begin{bmatrix}0_{n_1\times n_1} & -[A_{11}]^{-1}A_{12} & \cdots &
-\cdots & -[A_{11}]^{-1}A_{1p} \\
--[A_{22}]^{-1}A_{21} & 0_{n_2\times n_2} & -[A_{22}]^{-1}A_{23} &
-\cdots & -[A_{22}]^{-1}A_{2p}\\
-\vdots & \vdots & \ddots & \vdots & \vdots\\
-\vdots & \vdots & \vdots &\ddots & \vdots \\
--[A_{pp}]^{-1}A_{p1} &  \cdots & \cdots&
--[A_{pp}]^{-1}A_{p,p-1} & 0_{n_p \times n_p}
-\end{bmatrix}.
-\]
-
-For the sake of clarity, we notify that the symbol "$\%$" used in the following
-subsections represents the modulus function between two integer numbers, while
-the floor function is represented with pretty standard symbol "$\lfloor
-\cdot \rfloor$".
-
-\paragraph{Forward method}
-By assuming that all $n_i$'s have the same value we may define
-\[
- m=n_i=\text{size of a block}=\frac{n}{p}.
-\]
-
-The MAO transition probability matrix becomes:
-\[
- P_{i,j}=\frac{\lvert H_{i,j} \rvert}{\sum_{k=1}^n\lvert H_{i,k}
-\rvert}=\frac{\bigg \lvert \bigg ([A_{\lfloor
-\frac{i}{m}\rfloor \lfloor]
-\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
-\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg
-\rvert}{\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg
-([A_{\lfloor
-\frac{i}{m}\rfloor \lfloor
-\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
-\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert}
-\]
-
-Consequently, the $\hat{H}$ matrix is defined such that
-\[
-\hat{H}_{i,j} = \lvert H_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert H_{ik}\rvert\bigg)=
-\bigg \lvert \bigg ([A_{\lfloor \frac{i}{m}\rfloor \lfloor
-\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
-\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg
-\rvert
-\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
-\frac{i}{m}\rfloor \lfloor
-\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
-\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert
-\]
-
-By computing the sum over a generic row of $\hat{H}$ we get:
-\[
- \sum_{j=1}^n \lvert \hat{H}_{i,j}\rvert=\sum_{j=1}^n \hat{H}_{i,j} =
- \bigg ( \sum_{\substack{j=1\\j\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
-\frac{i}{m}\rfloor \lfloor
-\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
-\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert \bigg ) ^2.
-\]
-
-If we focus on the norm $\lVert \hat{H}\rVert_{\infty}$, then the following
-equivalence condition holds:
-\[
- \lVert \hat{H}\rVert_{\infty}<1 \Leftrightarrow \sum_{\substack{j=1\\j\ne
-i}}^n\bigg \lvert \bigg ([A_{\lfloor
-\frac{i}{m}\rfloor \lfloor
-\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
-\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert <1 \quad \forall
-i=1,\cdots,n
-\]
- A sufficient condition for this to happen is
- \begin{equation}
-  \sum_{\substack{j=1\\j\ne i}}^p \lVert [A_{ii}]^{-1}A_{ij}\rVert_{\infty}<1.
-    \label{block_cs}\quad \forall i=1,\ldots,p.
- \end{equation}
-
-By defining a matrix $\tilde{H}\in \mathbb{R}^{p\times p}$ such that
-\[
- \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{\infty} &
-\cdots &
-\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{\infty} \\
-\lVert [A_{22}]^{-1}A_{21}\rVert_{\infty} & 0 & \lVert
-[A_{22}]^{-1}A_{23}\rVert_{\infty} &
-\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{\infty} \\
-\vdots & \vdots & \ddots & \vdots & \vdots\\
-\vdots & \vdots & \vdots &\ddots & \vdots \\
-\lVert [A_{pp}]^{-1}A_{p1}\rVert_{\infty} &  \cdots & \cdots&
-\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{\infty} & 0
-\end{bmatrix}
-\]
-
-We can use the $\tilde{H}$ matrix just defined in order to introduce a
-sufficient condition for the convergence of the Forward Monet Carlo method with
-a Block Diagonal preconditioning.
-
-\begin{equation}
- \lVert \tilde{H} \rVert_{\infty}<1 \Rightarrow \lVert \hat{H}
-\rVert_{\infty}<1.
-\end{equation}
-
-\paragraph{Adjoint method}
-Analogously to the Forward method, if we define
-\[
-(\hat{H})^T_{i,j} = \lvert H^T_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert
-H^T_{ik}\rvert\bigg)
-\]
-we can formulate a sufficient condition for the convergence of the Adjoint
-Monte
-Carlo method by introducing a matrix $\tilde{H}$ which in this case is such that
-\[
-  \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{1} & \cdots
-&
-\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{1} \\
-\lVert [A_{22}]^{-1}A_{21}\rVert_{1} & 0 & \lVert
-[A_{22}]^{-1}A_{23}\rVert_{1} &
-\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{1} \\
-\vdots & \vdots & \ddots & \vdots & \vdots\\
-\vdots & \vdots & \vdots &\ddots & \vdots \\
-\lVert [A_{pp}]^{-1}A_{p1}\rVert_{1} &  \cdots & \cdots&
-\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{1} & 0
-\end{bmatrix}.
-\]
-The sufficient condition assumes the form
-\begin{equation}
- \lVert \tilde{H} \rVert_{1}<1 \Rightarrow \lVert \hat{H} \rVert_{1}<1.
-\end{equation}
-
-\subsection{Adaptive Methods}
-
-In formulas (\ref{dir_mean}) and (\ref{adj_mean}) the estimation of the
-solution to
-the linear system (\ref{linsys_prec}) involves infinite sums. Of course this is
-not viable for
-computational issues, therefore it is necessary to resort to truncation of
-the sums through proper criteria. In details, the truncation implies the
-employment of a finite number of permutations and a finite number of
-steps for each one of the permutations computed. Therefore it is necessary to
-truncate at two different levels to decide both how many permutations to employ
-and how many steps each one of them should be kept running.
-
-\subsubsection{History Length}
-
-At first we focus on the definition of a weight cut-off
-in order to decide where to terminate the histories. It may be reinterpreted as
-a quantity guiding us to decide how many terms of the Neumann
-series to consider. \newline
-
-The goal is to apply the cutoff via an automatic procedure, without requiring
-the intervention of the user directly.
-Therefore we are looking for an integer $m$ such that
-\[
-\tilde{\theta}_i=E\bigg[\sum_{\ell=0}^m W_{\ell}
-f_{k_{\ell}}\bigg]=x_i=\sum_{\ell=0}^m
-\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n
-P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},
-k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots
-w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}
-\]
-
-and
-
-\[
-\tilde{\theta}_j=E\bigg[\sum_{\ell=0}^m W_{\ell}\delta_{k_{\ell},
-j}\bigg]=\sum_{\ell=0}^{m}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n
-f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
-w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}
-\]
-
-are good approximations of (\ref{dir_mean}) and (\ref{adj_mean}) respectively.
-
-In \cite{Slattery2013} there is a criterion for the cutoff of the random walk
-applicable to both the Forward and the Adjoint method. It requires the set up
-of a relative cutoff threshold $W_c$ and to look for a step $m$ such that
-\begin{equation}
-W_m \le W_c W_0.
-\label{cutoff}
-\end{equation}
-$W_0$ is the value of the weight at the initial step of the random walk and
-$W_m$ is the value of the weight after $m$ steps.
-
-\subsubsection{Number of Random Walks}
-
-We now focus on quantities that might lead to a proper
-number of permutations to compute, in order to get a prescribed order of
-accuracy.
-As regards the Forward method, we know that the expression for the variance
-is defined in formula (\ref{dir_var}).
-
-In this context, a reasonable criterion to determine the number $\tilde{N}_i$
-of random walks to be run is setting a threshold $\varepsilon_1$ and determine
-\begin{equation}
-\tilde{N}_i \quad s.t.\quad \frac{\sqrt{Var[\theta_i]}}{\lvert
-E[\theta_i]\rvert}<\varepsilon_1, \quad i=1,\cdots,n.
-\label{forward_adapt}
-\end{equation}
-The dependence of $Var[\theta_i]$ and $E[\theta_i]$ on $\tilde{N}_i$, which
-seems
-to be absent in the previous formula, is highlighted by the fact that
-$\theta_i$
-is estimated by fixing a finite number of histories.
-Therefore we are controlling the relative standard deviation requiring it not
-to be too large. In other words we are pursuing a statistical setting where the
-uncertainty factor is not dominating over the expected value.
-This simple adaptive approach can be applied for the estimation of each
-component $x_i$. Hence different number of histories may be employed to
-compute different entries of the solution vector. \newline
-
-As concerns the Adjoint method, the estimation of the variance for each entry
-is reproduced by Formula (\ref{adj_var}).
-
-A possible adaptive selection of $\tilde{N}$, in this situation, is
-\begin{equation}
-\tilde{N} \quad s.t. \quad \frac{\lVert
-\boldsymbol{\sigma}_{\tilde{N}}\rVert_1}{\lVert
-\mathbf{x}\rVert_1}<\varepsilon_1,
-\label{adjoint_adapt}
-\end{equation}
-where $\boldsymbol{\sigma}$ is a vector whose entries are
-$\boldsymbol{\sigma}_{\tilde{N},i}=Var[\theta_i]$.
-
-What introduced above can be exploited in order to build
-an a posteriori adaptive algorithm, capable to identify the minimal value of
-$\tilde{N}$ that verifies (\ref{forward_adapt}) or (\ref{adjoint_adapt})
-respectively.
-
-Below there are the pseudo-codes associated with both Forward and Adjoint Monte
-Carlo.
-
-\begin{algorithm}[H]
- \KwData{$N$, $\varepsilon_1$}
- \KwResult{$\tilde{N}_i$, $\sigma_i$, $x_i$}
- For each entry of the solution vector (for $i=1,\cdots,n$) \;
- $\tilde{N}_i=N$\;
- compute $\theta_i$\;
- \While{$\frac{\sigma_i}{\lvert
-E[\theta_i]\rvert}<\varepsilon_1$}{
-\vspace{0.1cm}
-  $\tilde{N}=\tilde{N}+N$\;
-  compute $E[\theta_i]=x_i$\;
- }
- return $\theta_i$, $x_i$, $\sigma_i$\;
- \caption{A posteriori adaptive Forward Monte Carlo}
-\end{algorithm}
-
-\begin{algorithm}[H]
- \KwData{$N$, $\varepsilon_1$}
- \KwResult{$\tilde{N}$, $\boldsymbol{\sigma_i}$, $\mathbf{x}$}
- $\tilde{N}=N$\;
- compute $\theta$\;
- \While{$\frac{\lVert\boldsymbol{\sigma} \rVert}{\lVert
-E[\theta]\rVert}<\varepsilon_1$}{
-\vspace{0.1cm}
-  $\tilde{N}=\tilde{N}+N$\;
-  compute $E[\theta]=\mathbf{x}$\;
- }
- return $\theta$, $\mathbf{x}$, $\boldsymbol{\sigma}$\;
- \caption{A posteriori adaptive Adjoint Monte Carlo}
-\end{algorithm}
-
-\subsection{Preconditioning}
-
-In this section we focus on the pursuit of preconditioners that facilitate the
-convergence of the stochastic schemes introduced so far. A priori, all of the
-following techniques are good candidates to improve the numerical setting.
-However issues about the fill-in effect and the verification of convergence
-conditions limit their employment for our purposes.
-
-\subsubsection*{Block Preconditioning}
-Consider a matrix $A\in\mathbb{R}^{n\times n}$ of the form
-\[
- A=\begin{bmatrix}A_{11} & A_{12} & \cdots & A_{1p}\\
-    A_{21} & A_{22} & \cdots & A_{2p}\\
-    \vdots & \vdots & \ddots & \vdots \\
-    A_{p1} & \cdots & \cdots & A_{pp}
-   \end{bmatrix}
-\]
-where $p$ is a divisor of $n$ and such that $A_{ii} \in \mathbb{R}^{n_i
-\times n_i}$ are nonsingular square matrices.
-By defining a block diagonal matrix such as
-\[
- D=\begin{bmatrix}A_{11} & 0_{12} & \cdots & 0_{1p}\\
-    0_{21} & A_{22} & \cdots & 0_{2p}\\
-    \vdots & \vdots & \ddots & \vdots \\
-    0_{p1} & \cdots & \cdots & A_{pp}
-   \end{bmatrix}
-\]
-it is possible to compute its inverse.
-$D^{-1}$  can be used as a preconditioner for the original matrix $A$.
-In particular
-\[
- D^{-1}A=\begin{bmatrix}I_{n_1\times n_1} & A_{11}^{-1}A_{12} & \cdots &
-\cdots & A_{11}^{-1}A_{1p} \\
-A_{22}^{-1}A_{21} & I_{n_2\times n_2} & A_{22}^{-1}A_{23} &
-\cdots & A_{22}^{-1}A_{2p}\\
-\vdots & \vdots & \ddots & \vdots & \vdots\\
-\vdots & \vdots & \vdots &\ddots & \vdots \\
-A_{pp}^{-1}A_{p1} &  \cdots & \cdots&
-A_{pp}^{-1}A_{p,p-1} & I_{n_p \times n_p}
-\end{bmatrix}.
-\]
-The ultimate goal is to resort to this preconditioning technique in order to
-achieve the condition $\rho(D^{-1}A)<1$. In fact this is necessary in order to
-reformulate \ref{linsys} into \ref{fixedpoint}.
-
-Convergence of \ref{fixedpoint} with block diagonal preconditioning, from the
-deterministic point of view, is guaranteed when $A$ is an irreducibly
-diagonally dominant matrix or when $A$ is an $M$-matrix. In fact in situation
-such as these, theoretical results guarantee that $A=D-N$ is a regular
-splitting (see \cite{Ax1996}). However the same theoretical results cannot be
-exploited to guarantee the convergence of the MC solvers, because of the
-additional check on $\rho(\hat{H})<1$ to be considered.
-
-\subsubsection*{Approximate Inverse of a Sparse Matrix (AINV)}
-
-This stimulated our interests into other types of preconditioners. At first we
-concentrated on Approximate Inverse preconditioners (see \cite{Saad} and
-\cite{Benzi2002}).
-
-There are many practical techniques that may be applied in order to build the
-AINV preconditioner.
-One of these is described in \cite{Tuma1996} and \cite{Tuma1998}.
-This is the actual approach we applied to the construction of the
-preconditioner. The technique constructs the preconditioner via a
-bi-conjugation fashion.
-The idea is to compute an approximate factorization of the form $W^T A Z=D$.
-$W$ and $Z$ are unit upper triangular matrices and $D$ is a diagonal matrix.
-
-In order to compute the AINV preconditioner, a \texttt{FORTRAN} code
-provided by Miroslav Tuma (Academy of Sciences, Prague) is used.
-We report as follows the algorithm employed for the construction of the so
-called \textit{Right-looking Factored AINV}. The algorithm provide the user
-with a dropping tolerance $\tau$ which can be used in order to tune the fill-in
-effect
-of the factored preconditioner. \newline
-
-\begin{algorithm}[H]
- \KwData{$p=q=(0,\cdots,0)\in \mathbb{R}^n$, $Z=[z_1,\cdots,z_n]=I_n$,
-$W=[w_1,\cdots,w_n]=I_n$}
- \KwResult{$Z$, $W$}
- \For{$k=1,\cdots, n$}{
- $p_k=w_k^T A e_k$, $q_k=e_k^T A z_k$ \;
-     \For{$i=k+1,\cdots,n$}{
-         $p_i=(w_k^T A e_k)/p_k$, $q_i=(e_k^T A z_i)/q_k$\;
-         Apply a dropping rule to $p_i$, $q_i$\;
-         $w_i=w_i - w_k p_i$, $z_i=z_i - z_k q_i$\\
-         Apply a dropping rule to $w_{j,i}$ and $z_{j,i}$ for $j=1,\cdots,i$\;
-     }\;
-  }\;
- Choose diagonal entries of $D$ as the component of $p$ or $q$\;
- return $Z$, $W$\;
- \caption{Right-looking Factored AINV}
-\end{algorithm}
-
-\subsubsection*{Incomplete LU Factorization Preconditioners (ILU)}
-This set of preconditioners has the goal of computing a sparse lower triangular
-matrix $L$ and a sparse upper triangular matrix $U$ such that the residual
-matrix $R=LU-A$ respects some constraints. One of these constraints may be the
-fact that some entries must be set to zero.
-For instance we can introduce a zero pattern set
-\[
- P\subset \{(i,j)\lvert i\ne j; 1\le i,j\le n\}
-\]
-and use this set in order to impose some constraints on the shape of the
-preconditioner built.\newline
-
-Different variants of the
-Incomplete LU Factorization algorithm are available (see \cite{Saad}).
-However it is
-possible to prove that if all the formulation are well defined, then they are
-all equivalent.
-One of these variants is the following
-
-\begin{algorithm}[H]
- \KwData{$P$}
- \KwResult{$A$}
- For each $(i,j)\in P$ set $a_{ij}=0$\;
- \For{$k=1,\cdot,n-1$}{
- \For{$i=k+1,n$}{
- \If{$(i,k)\notin P$}{
- $a_{ik}=\frac{a_{ik}}{a_{kk}}$
- \For{$j=k+1,\cdots, n$ and for $(i,j)\notin P$}{
-  $a_{ij}=a_{ij}-a_{ik}*a_{kj}$
- }
- }
- }
- }
- return $Z$, $W$\;
- \caption{General Static Pattern ILU}
-\end{algorithm}
-
-This kind of algorithm is guaranteed to terminate without any breakdown just
-for $M$-matrices.
-
-Assuming that the zero pattern $P$ coincides with the zero pattern of $A$ leads
-to the $ILU(0)$. \newline
-
-\begin{algorithm}[H]
- \KwData{$P$}
- \KwResult{$A$}
- \For{$i=2,\cdot,n$}{
- \For{$k=i,i-1$ and for $(i,k)\in NZ(A)$}{
-  $a_{ik}=\frac{a_{ik}}{a_{kk}}$
-  \For{$j=k+1,\cdots,n$ and for $(i,j)\in NZ(A)$}{
-  $a_{ij}=a_{ij}-a_{ik}a_{kj}$}
- }
- }
- \caption{ILU(0).}
-\end{algorithm}
-
-More sophisticated variants of the algorithm enable to enlighten the constraint
-of the sparsity pattern, finding a compromise between sparsity and accuracy in
-approximating the inverse of $A$. We refer to \cite{Saad} and
-\cite{Benzi2002} for further information about it.
-
-\section{Numerical Results}
-
-\subsection{Adaptive Methods}
-
-In this subsection we show some numerical tests to validate the utility of the
-adaptive approaches for the selection of the number of permutations to employ.
-For this purpose, we restrict our attention to standard Monte Carlo linear
-solvers.
-
-At first we take into account the Forward Monte Carlo linear solver.
-In particular, the adaptive threshold associated with formulas
-(\ref{forward_adapt}) and
-(\ref{adjoint_adapt}) is set to $\varepsilon_1=0.1$ and to
-$\varepsilon_1=0.01$. \newline
-
-A set of matrices associated with problems of various nature has been collected.
-All the problems treated have a small number of d.o.f's to control the
-massive computational cost of MC solvers on a standard laptop. All the analyzed
-matrices are derived from the
-discretization of other differential problems.
-One of these matrices is produced by the finite volume discretization of
-a thermal equation (Marshak problem), another one is a 1D diffusion reaction
-problem discretized with finite differences, a
-2D laplacian discretized with finite differences and an advection
-diffusion
-problem discretized with quadrilateral linear finite elements (the
-\texttt{ifiss} package is used in this case).
-For all the problems at hand a left diagonal preconditioning is applied.
-Details about all these matrices are gathered in Table
-\ref{table_data}.
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{d.o.f.'s} &
-$\rho(H)$&Forward $\rho(\hat{H})$&Adjoint $\rho(\hat{H})$\\
-\hline
-1d shifted Laplacian & 50 & $0.4991$ & $0.9827$ & $0.9827$\\
-\hline
-2d Laplacian & 900 & $0.9949$ & $0.994$ & $0.9945$\\
-\hline
-ifiss & 1089 &$0.9836$ & $0.983$ & $0.983$ \\
-\hline
-Marshak Equation & 1600 & $0.6009$ & $0.3758$ & $ 0.3815$ \\
-\hline
-\end{tabular}
-\caption{Properties of the matrices employed for the checks on adaptivity.}
-\label{table_data}
-\end{table}
-
-For all the numerical tests results are presented both by using the Forward
-and the Adjoint Monte Carlo.
-
-As concerns the Forward method, a safety threshold for maximal number of
-histories per
-entry is set to $10^7$. Results are presented for
-$\varepsilon_1=0.1$ and $\varepsilon_1=0.01$ in Tables \ref{tab:For_adapt} and
-\ref{tab:For_adapt2} respectively. A batch size of two is used at each
-adaptive check to verify the magnitude of the apparent relative standard
-deviation. As expected, results are aligned with the convergence rate predicted
-by the Central Limit Theorem. Indeed, decreasing of a factor of 10 the
-tolerance $\varepsilon_1$, the relative error decreases of the same order,
-employing in average one hundred times more histories.\newline
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
-\hline
-1d shifted Laplacian & $0.003$ & $5,220$\\
-\hline
-2d Laplacian & $0.1051$ & $262,350$\\
-\hline
-ifiss advection diffusion & $0.022$  & $1,060,770$\\
-\hline
-Marshak equation & $0.0012$ & $1,456,000$\\
-\hline
-\end{tabular}
-\caption{Forward Monte Carlo. Adaptive selection of histories.
-$\varepsilon_1=0.1$}
-\label{tab:For_adapt}
-\end{table}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
-\hline
-1d shifted Laplacian & $4\cdot 10^{-4}$ & $512,094$\\
-\hline
-2d Laplacian & $0.0032$ & $108,551,850$\\
-\hline
-ifiss advection diffusion & $0.0023$  & $105,476,650$\\
-\hline
-Marshak equation & $5.8 \cdot 10^{-4}$ & $144,018,700$\\
-\hline
-\end{tabular}
-\caption{Forward Monte Carlo. Adaptive selection of histories.
-$\varepsilon_1=0.01$}
-\label{tab:For_adapt2}
-\end{table}
-
-As regards the Adjoint Monte Carlo, at each adaptive check the number of
-random walks employed is increased by ten.
-
-A maximal number of histories equal to $10^10$ is employed as a safety
-threshold.
-
-Here below Tables \ref{tab:Adj_adapt}, \ref{tab:Adj_adapt2} and
-\ref{tab:Adj_adapt3} show results
-for the different test cases
-considered using $\varepsilon_1=0.1$, $\varepsilon_1=0.01$ and
-$\varepsilon_1=10^{-3}$ respectively.
-By comparing the two tables, it is noticed that a decrease of the threshold
-induces a reduction of the relative error of the same order of magnitude.
-Therefore this validates
-the effectiveness of the adaptive selection of histories with an error
-reduction goal. The total number of histories employed increases on average by
-two orders of magnitude and this aspect is in line too with what expected by
-theoretical results.
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
-\hline
-1d shifted Laplacian & $0.08$ & $1820$\\
-\hline
-2d Laplacian & $0.136$ & $570$\\
-\hline
-ifiss advection diffusion & $0.08$  & $2400$\\
-\hline
-Marshak equation & $0.288$ & $880$\\
-\hline
-\end{tabular}
-\caption{Adjoint Monte Carlo. Adaptive selection of histories.
-$\varepsilon_1=0.1$}
-\label{tab:Adj_adapt}
-\end{table}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
-\hline
-1d shifted Laplacian & $0.0082$ & $185,400$\\
-\hline
-2d Laplacian & $0.0122$ & $126,800$\\
-\hline
-ifiss advection diffusion & $0.0093$  & $219,293$\\
-\hline
-Marshak equation & $0.0105$ & $650,040$\\
-\hline
-\end{tabular}
-\caption{Adjoint Monte Carlo. Adaptive selection of histories.
-$\varepsilon_1=0.01$}
-\label{tab:Adj_adapt2}
-\end{table}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
-\hline
-1d shifted Laplacian & $9.56 \cdot 10^{-4}$ & 15,268,560 \\
-\hline
-2d Laplacian & $0.001$ & $12,600,000$\\
-\hline
-ifiss advection diffusion & $0.0011$  & $23,952,000$\\
-\hline
-Marshak equation & $0.0011$ &  $80,236,000$\\
-\hline
-\end{tabular}
-\caption{Adjoint Monte Carlo. Adaptive selection of histories.
-$\varepsilon_1=10^{-3}$}
-\label{tab:Adj_adapt3}
-\end{table}
-
-The same results can be reproduced by resorting to the expected value
-estimator. Results are shown in the Tables \ref{tab:Adj_adapt4},
-\ref{tab:Adj_adapt5} and \ref{tab:Adj_adapt6} for the threshold values of
-$\varepsilon_1=0.1$, $\varepsilon_1=0.01$ and $\varepsilon_1=10^{-3}$
-respectively. As it can be noticed, in terms of error scaling the results are
-pretty similar to the ones obtained with the collision estimator. As regards
-the number of histories as well, used to reach a prescribed accuracy, the
-orders of magnitude are the same for both the collision estimator and the
-expected value one. However the expected value estimator requires most of the
-time a smaller number permutations. This behavior is enhances as the value of
-the threshold decreases, making the computation heavier and heavier.
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
-\hline
-1d shifted Laplacian & $0.0463$ & $1000$\\
-\hline
-2d Laplacian & $0.1004$ & $900$\\
-\hline
-ifiss advection diffusion & $0.0661$  & $1300$\\
-\hline
-Marshak equation & $0.0526$ & $2000$\\
-\hline
-\end{tabular}
-\caption{Expected value estimator - Adjoint Monte Carlo. Adaptive selection of
-histories.
-$\varepsilon_1=0.1$}
-\label{tab:Adj_adapt4}
-\end{table}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
-\hline
-1d shifted Laplacian & $0.004$ & $100,600$\\
-\hline
-2d Laplacian & $0.0094$ & $83,700$\\
-\hline
-ifiss advection diffusion & $0.0088$  & $124,400$\\
-\hline
-Marshak equation & $0.0056$ & $166,000$\\
-\hline
-\end{tabular}
-\caption{Expected value - Adjoint Monte Carlo. Adaptive selection of histories.
-$\varepsilon_1=0.01$}
-\label{tab:Adj_adapt5}
-\end{table}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|}
-\hline
-\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
-\hline
-1d shifted Laplacian & $0.004$ & $10,063,300$\\
-\hline
-2d Laplacian & $ 9.31 \cdot 10^{-4} $ & $ 8,377,500 $\\
-\hline
-ifiss advection diffusion & $ 0.0013 $  & $ 12,435,000 $\\
-\hline
-Marshak equation & $7.79 \cdot 10^{-4}$ & $16,537,000$\\
-\hline
-\end{tabular}
-\caption{Expected value - Adjoint Monte Carlo. Adaptive selection of histories.
-$\varepsilon_1=10^{-3}$}
-\label{tab:Adj_adapt6}
-\end{table}
-
-\subsection{Preconditioning Approaches}
-
-This subsection is focused on showing the effect of different preconditioners
-for the values attained by the spectral radii $\rho(H)$ and $\rho(\hat{H})$.
-For this purpose, we consider the two dimensional laplacian
-discretized with centered finite differences and the "ifiss advection
-diffusion" matrix. The properties of both the matrices have already been
-explained in previous sections.
-Results in terms of spectral radii when the diagonal preconditioning is applied
-have been shown earlier as well.
-
-Here the analysis is focused on the evaluating the effects of the
-block diagonal preconditioner and the factorized approximate sparse inverse.
-As concerns the matrix $\hat{H}$, it has been built accordingly to te
-procedure for the Adjoint Monte Carlo method. In
-the first case the size of the blocks is a tuning parameter, whilst in the
-second case the tuning parameter is the drop tolerance to manage the fill-in
-effect of the factors.
-
-As regards the employment for the block diagonal preconditiner, in Tables
-\ref{tab:lap2d_block} and \ref{tab:ifiss_block} there is a counterintuitive
-common trend. Indeed in both the cases an increase of $\rho(H)$ and
-$\rho(\hat{H})$  occurs as the size of the blocks diminishes.
-
-In Tables \ref{tab:lap2d_ainv} and \ref{tab:ifiss_ainv} the values of
-the spectral radii are shown for the case of the factorized AINV
-preconditioner. It is interesting to point out that for the
-two-dimensional
-laplacian, a drop tolerance $\tau=0.05$ entails $\rho(H)<1$ but the same does
-not hold for $\hat{H}$. Both the convergence conditions are restored by
-decreasing even more the drop tolerance, causing a massive fill-in effect
-though.
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|c|}
-\hline
-\textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
-&$\rho(\hat{H})$\\
-\hline
- 5 & 2.6164 & $0.9915$ & $0.9837$\\
-\hline
- 10 & 5.6027 & $0.9907$ & $0.9861$\\
-\hline
- 30 & 16.3356 & $0.9898$  & $0.9886$\\
-\hline
-\end{tabular}
-\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the two dimensional
-laplacian with block diagonal preconditioning.}
-\label{tab:lap2d_block}
-\end{table}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|c|}
-\hline
-\textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
-&$\rho(\hat{H})$\\
-\hline
- 3 & 1.4352 & $0.9804$ & $0.9531$\\
-\hline
- 9 & 3.0220 & $0.9790$ & $0.9674$\\
-\hline
- 33 & 9.8164 & $0.9783$  & $0.9774$\\
-\hline
-\end{tabular}
-\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the ifiss advection
-diffusion matrix with block diagonal preconditioning.}
-\label{tab:ifiss_block}
-\end{table}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|c|}
-\hline
-\textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
-&$\rho(\hat{H})$\\
-\hline
- 0.05 & 8.2797 & $0.9610$ & $1.0231$\\
-\hline
- 0.01 & 33.1390 & $0.8668$ & $0.8279$\\
-\hline
-\end{tabular}
-\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the two dimensional
-laplacian with factorized AINV preconditioning.}
-\label{tab:lap2d_ainv}
-\end{table}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|c|c|c|c|}
-\hline
-\textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
-&$\rho(\hat{H})$\\
-\hline
- 0.05 & 3.8392 & $0.9396$ & $0.9069$\\
-\hline
- 0.01 & 15.1798 & $0.7964$ & $0.6361$\\
-\hline
-\end{tabular}
-\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the ifiss advection
-diffusion matrix with factorized AINV preconditioning.}
-\label{tab:ifiss_ainv}
-\end{table}
-
-\subsection{Hybrid Methods}
-
-\subsubsection{Poisson Problem}
-
-In the current subsection a Poisson problem is taken into account:
-\begin{equation}
-\begin{cases}
- -\Delta u = f \quad \text{in}\quad \Omega \\
- u=0\quad \text{on} \quad \partial\Omega
- \end{cases}
-\end{equation}
-
-where $\Omega=(0,1)\times (0,1)$, $\sigma=1$.
-A finite difference scheme is employed to discretize the problem.
-As a right hand side, a
-sinusoidal distribution in both $x$ and $y$ directions is considered, so that
-\[
- f(x_i, y_i)=sin\bigg(\frac{\pi x}{N-1}\bigg)sin\bigg(\frac{\pi y}{N-1}\bigg).
-\]
-N is the number of node considered on each direction, in our case $N=32$. The
-resulting linear system has 900 d.o.f.'s and $h=\frac{1}{29}$.
-By applying a left diagonal preconditioning on the matrix, it is possible to
-reformulate the problem as a fixed point scheme.
-The iteration matrix $H$ is characterized by the fact that $\lVert
-H\rVert_1=1$, which is not sufficient to guarantee the convergence of
-any
-stochastic scheme analyzed in this work. However, by theoretical results, it is
-well known that applying a left diagonal preconditioning to the problem at hand
-produces an iteration matrix $H$ such that
-\[
- \lim_{h\rightarrow 0}\rho(H)=1^{-}
-\]
-where $h$ is the spatial discretization step.
-This is enough to have the guarantee that the deterministic fixed point scheme
-converges without the explicit computation of the spectral radius $\rho(H)$.
-However, the smaller is the discretization step $h$, the higher is the number of
-iterations required to reach a prescribed convergence criterion. For our
-discretization setting, the iteration matrix $H$ has a
-spectral radius of $\rho(H)\approx 0.9949$.
-
-In order to have the guarantee of the convergence of the Adjoint Monte Carlo
-method, it is necessary to look at the spectral radius of $\hat{H}$ too. If an
-almost optimal probability is used, this implies that
-the Adjoint Monte Carlo method has a $\hat{H}$ matrix such that
-\[
- \lVert \hat{H}\rVert_1\le (\rVert H\rVert_1)^2=1.
-\]
-This condition by itself is not enough to guarantee that
-$\rho(\hat{H})<1$.
-
-However, it must be recalled that the iteration matrix $H$ has zero
-valued entries on the main diagonal and it has:
-\begin{itemize}
- \item four entries equal to $\frac{-1}{4}$ on the rows associated
-with a node which is not adjacent to the boundary
-\item two entries equal to $\frac{-1}{4}$ on the rows associated with nodes
-adjacent to the corner of the square domain
-\item three entries equal to $\frac{-1}{4}$ on the rows associated with nodes
-adjacent to the boundary on an edge of the square domain.
-\end{itemize}
-
-Recalling the definition of $\hat{H}$ entrywise in terms of the
-iteration matrix $H$ and the transition probability $P$, we get that
-
-\[
- \hat{H}=\tilde{D}H
-\]
-
-where $\tilde{D}$ is a diagonal matrix with diagonal entries equal to
-1, $\frac{1}{2}$ or $\frac{3}{4}$.
-In particular $\tilde{d}_i=diag(\tilde D)_i=1$ if the $i$-th row of the
-discretized laplacian is associated with a node which is not adjacent to the
-boundary, $\tilde{d}_i=diag(\tilde D)_i=\frac{1}{2}$ if the row is associated
-with a node of the grid adjacent to the corner of the boundary,
-$\tilde{d}_i=diag(\tilde D)_i=\frac{3}{4}$ if the associated node of the grid
-is adjacent to the boundary of the domain far from a corner.
-
-Since $H$ is an irreducible nonnegative matrix and $\tilde{D}$ is a
-positive definite diagonal matrix, we resort to a result presented in
-\cite{FK1975} to say that
-\[
- \rho(\hat{H})=\rho(\tilde{D}H)\le \rho(\tilde D)\rho(H).
-\]
-
-Since $\rho(\tilde{D})=1$, then $\rho(\hat{H})\le \rho(H)<1$.
-Therefore the diagonal preconditioning always guarantees a convergent
-setting for the Monte Carlo linear solver, once it is applied on a
-two-dimensional Laplace
-operator discretized with finite differences. By a similar reasoning, the same
-results can be obtained for a $n$-dimensional laplace operator.
-
-The computation of the solution is realized both with the deterministic
-Synthetic Acceleration and with Adjoint MCSA to compare the effectiveness of
-the
-stochastic technique. Results are reported in Table
-\ref{Poisson_results}. For the sake of completeness, results obtained with
-the Sequential Monte Carlo are shown as well. An evident discrepancy in terms
-of numerical iterations
-is already detected with the Sequential Monte Carlo outperforming the
-preconditioned fixed point. Results are improved even more
-when
-switching to the MCSA, which employs less histories than the Sequential MC at
-each iteration to
-reach convergence. In fact the average number of histories employed by the
-Sequential
-Monte Carlo per iteration is almost one order of magnitude higher than for
-MCSA. This
-might be explained by the fact that the first row of the iterative algorithm in
-the MCSA refines the accuracy of the solution coming from the previous
-iteration,
-before using it as the input for the MC solver at the current iterate.
-In particular, since the first line of the algorithm is a deterministic step of
-the Synthetic Acceleration, its effect is to dampen the statistical noise
-produced by the Monte Carlo linear solver to estimate the update $\delta
-x^{n + \frac{1}{2}}$ at the previous iteration. Therefore
-refinement provided by the Synthetic Acceleration step decreases the number of
-random walks
-to employ for a prescribed accuracy. This is validated even more by the fact
-that for the first iteration the Sequential Monte Carlo and the MCSA resort to
-the same number of histories, while their behaviours start differing from the
-second iteration on, when the statistical noise is introduced in the estimation
-of the solution to \ref{linsys_prec} (see Figures \ref{SEQ_poisson} and
-\ref{MCSA_poisson}).
-
-\begin{table}[!h]
-\centering
-\hspace*{-0.8cm}
-\begin{tabular}{|c|c|c|c|}
-\hline
-algorithm & relative err.& \# iterations & average \# histories per iteration\\
-\hline
-Synthetic Acceleration & $9.8081\cdot 10^{-8}$ & 3133 & - \\
-\hline
-Sequential MC & $7.9037 \cdot 10^{-8}$ & 9 & 8.264,900\\
-\hline
- Adjoint MCSA & $8.0872\cdot 10^{-8}$ & 8 & 1,738,250\\
-\hline
-\end{tabular}
-\caption{Numerical results for the Poisson problem.}
-\label{Poisson_results}
-\end{table}
-
-\begin{figure}[h!]
-  \centering
-    \includegraphics[width=\textwidth]{SEQ_poisson.jpg}
-    \caption{Sequential MC - Poisson problem. Number of random walks employed
-at each numerical iteration.}
-\label{SEQ_poisson}
-\end{figure}
-
-\begin{figure}[h!]
-  \centering
-    \includegraphics[width=\textwidth]{MCSA_poisson.jpg}
-      \caption{MCSA - Poisson problem. Number of random walks employed at each
-numerical iteration.}
-\label{MCSA_poisson}
-\end{figure}
-
-\subsubsection{Reaction-Diffusion Problem}
-
-By modifying the equation of the previous test case, we consider now
-an advection reaction problem
-
-\begin{equation}
-\begin{cases}
- -\Delta u +\sigma u= f \quad \text{in}\quad \Omega \\
- u=0\quad \text{on} \quad \partial\Omega
- \end{cases}
-\end{equation}
-where $\Omega=(0,1)\times (0,1)$, $\sigma=0.1$ and $f=1$.
-A finite difference scheme is applied to discretize the problem.
-The number of nodes
-on each direction of the domain is 100, so that $h\approx 0.01$. The
-discretized problem has 9604 d.o.f's. A left
-diagonal preconditioning is still applied to
-the stiffness matrix obtained from the discretization. The 1-norm of the
-iteration matrix is $\lVert H\rVert_1\approx 0.9756$. This automatically
-guarantees the convergence of the Adjoint Monte Carlo linear solver. The
-computation of the solution to the partial differential problem is still
-accomplished with the MCSA algorithm. A threshold of $\varepsilon =10^{-8}$ is
-used as a stop criterion for the relative residual. The threshold
-for the adaptive selection of the random walks instead is set
-to $\varepsilon_1=0.1$.
-In Table \ref{DR_results} a comparison between the deterministic Synthetic
-Acceleration
-scheme, Sequential MC and the MCSA is provided.
-The employment of the MCSA provides a significant reduction of the number of
-histories necessary to satisfy the convergence criterion, in line with results
-coming from the previous test case.
-
-\begin{table}[!h]
-\centering
-\hspace*{-0.8cm}
-\begin{tabular}{|c|c|c|c|}
-\hline
-algorithm & relative err.& \# iterations & average \# histories per iteration\\
-\hline
- Synthetic acceleration & $9.073\cdot 10^{-8}$ & 634 & - \\
-\hline
- Sequential MC & $8.415 \cdot 10^{-8}$ &  8 & 12,391,375\\
- \hline
- Adjoint MCSA & $6.633 \cdot 10^{-8}$ &  7 & 3,163,700\\
-\hline
-\end{tabular}
-\caption{Numerical results for the diffusion reaction problem.}
-\label{DR_results}
-\end{table}
-
-\begin{figure}[h!]
-  \centering
-    \includegraphics[width=\textwidth]{SEQ_diffreac.jpg}
-      \caption{Sequential MC - Diffusion reaction problem. Number of random
-walks
-employed at each
-numerical iteration.}
-\label{SEQ_diffreac}
-\end{figure}
-
-\begin{figure}[h!]
-  \centering
-    \includegraphics[width=\textwidth]{MCSA_diffreac.jpg}
-      \caption{MCSA - Diffusion reaction problem. Number of random walks
-employed at each
-numerical iteration.}
-\label{MCSA_diffreac}
-\end{figure}
-
-\subsubsection{Parabolic Problem}
-
-In this subsection we restrict our treatise to the analysis of parabolic
-partial
-differential problems of the form
-\begin{equation}
-\begin{cases}
- \frac{\partial u }{\partial t} + \mathcal{L}(u)=f \quad \text{in} \Omega \\
-\text{B.C.} \quad \text{on}\quad \delta \Omega.
-\end{cases}
-\label{parabolic}
-\end{equation}
-
-The partial derivative in time is discretized with finite differences, whereas
-quadrilateral linear finite elements are employed for the space discretization.
-Naming $N_h$ the number of d.o.f.'s associated with the finite element
-discretization, the problem \ref{parabolic} turn into a fully discretized
-problem such as:
-
-\[
- \bigg (\frac{1}{\Delta t}M+A\bigg )(\mathbf{u}^{k+1}-\mathbf{u}^k)+A[\theta
-\mathbf{u}^{k+1}+(1-\theta)\mathbf{u}^k]=\theta
-\mathbf{f}^{k+1}+(1-\theta)\mathbf{f}^k,
-\]
-
-where $k$ is an index that goes with the time dimension.
-
-By tuning the value of the parameter $\theta$ different numerical schemes can
-be obtained with different properties of numerical stability.
-However our discussion is not involved in this kind of issues.
-In particular we restrict our attention to a single generic time step
-associated with an Implicit Euler time discretization (corresponding to
-$\theta=1$). The vector of the right hand side is chosen so that the exact
-solution to the linear system for the specific time step chosen is a unit
-vector $\tilde{\mathbf{u}}=\mathbbm{1}^{N_h}$.
-
-By referring to $h$ as the space discretization step, we pick
-\[
- \Delta t \le C h.
-\]
-
-Give an open and bounded set $\Omega=(0,1)\times(0,1)$, we consider the
-following problem
-\begin{equation}
- \begin{cases}\frac{\partial u}{\partial t} -\mu \Delta u
-+\boldsymbol{\beta}(\mathbf{x})\cdot \nabla u=0, \;\; \quad \quad
-\mathbf{x}\in \Omega,\quad t\in(0,T] \\
-u(\mathbf{x},0) = u_0, \qquad \qquad \qquad \qquad \quad \mathbf{x}\in
-\Omega \\
-u(\mathbf{x},t)=u_D(\mathbf{x}),  \; \, \, \qquad \qquad \qquad \quad
-\mathbf{x}\in \partial
-\Omega, \quad t\in(0,T],
- \end{cases}
-\end{equation}
-
-where $\mu=\frac{3}{200}$, $\boldsymbol{\beta}(\mathbf{x})=[2y(1-x^2),\;
--2x(1-y^2)]^T$, \\$u_D=0$ on $\{\{x=0\}\times (0,1)\}$, $\{(0,1)\times
-\{y=0\}\}$, $\{(0,1)\times \{y=1\}\}$.\newline
-
-The discretization of the problem is realized with the \texttt{IFISS} toolbox.
-The value of the discretization step is $h=2^{-8}$, generating a discretized
-linear system with 66,049 d.o.f.'s.
-As concerns the time discretization, the time step is set to $\Delta t =
-10h$.\newline
-
-The sparse approximate inverse is employed for
-the linear system at hand as a right preconditioner, with a drop tolerance of
-$\tau=0.05$ for both the factors.
-
-In this situation it is not possible to
-resort to any of the sufficient conditions to verify the convergence. Therefore
-the only viable option is the explicit computation of the spectral radii of $H$
-and $\hat{H}$.
-The spectral radius of the iteration matrix is $\rho(H)\approx
-0.9218$ and the spectral radius of $\hat{H}$ for the Adjoint Monte Carlo is
-$\rho(\hat{H})\approx 0.9148$. The transition probability employed is the
-Almost
-Optimal one. Resorting to a Uniform Probability in this case would have impeded
-the
-convergence, since in this case $\rho(\hat{H})\approx 1.8401$.
-Therefore
-this is an example demonstrating that the Almost Optimal Probability
-facilitates the performance of the stochastic algorithm, outperforming the
-Uniform one.
-
-The fill-in effect is
-\[
- \frac{nnz(H)}{nnz(A)}=4.26,
-\]
-therefore the relative number of nonzero elements in $H$ is still acceptable
-in terms of sparsity pattern and memory storage.
-
-The threshold for the check on the relative residual is set to $\varepsilon
-=10^{-8}$ is, whilst the threshold
-for the adaptive selection of the random walks is set
-to $\varepsilon_1=0.1$.
-In order to test the effectiveness of employing the MCSA with respect to a
-purely deterministic scheme, a comparison with the Richardson algorithm and
-the Sequential Monte Carlo is
-accomplished. The results are shown in Table \ref{parabolic_results}. As you
-can see, the adoption of both Sequential Monte Carlo and MCSA reduces
-dramatically the number of iterations with
-respect to the fixed point method. Of course this
-advantage has the
-increase of the computational time as a payoff. Nevertheless we put off this
-discussion which is of
-interest for a parallelized implementation of the method. Even more the number
-of random walks employed by the Sequential MC is higher than the one associated
-with the MCSA, in accordance with the previous results.
-
-\begin{table}[!h]
-\centering
-\hspace*{-0.8cm}
-\begin{tabular}{|c|c|c|c|}
-\hline
-algorithm & relative err.& \# iterations & average \# histories per iteration\\
-\hline
- Richardson & $6.277\cdot 10^{-7}$ & 178 & - \\
- \hline
- Sequential MC & $1.918 \cdot 10^{-9}$ & 8 & 51,535,375\\
-\hline
- Adjoint MCSA & $1.341\cdot 10^{-9}$ & 6 & 16,244,000\\
-\hline
-\end{tabular}
-\caption{Numerical results for the parabolic problem.}
-\label{parabolic_results}
-\end{table}
-
-\begin{figure}[h!]
-  \centering
-    \includegraphics[width=\textwidth]{SEQ_parabolic.jpg}
-      \caption{Sequential Monte Carlo - Parabolic problem. Number of random
-walks
-employed at each
-numerical iteration.}
-\label{MCSA_parabolic}
-\end{figure}
-
-\begin{figure}[h!]
-  \centering
-    \includegraphics[width=\textwidth]{MCSA_parabolic.jpg}
-      \caption{MCSA - Parabolic problem. Number of random walks
-employed at each
-numerical iteration.}
-\label{MCSA_parabolic}
-\end{figure}
-
-\section{Conclusions and Future Work}
-
-The work presented in this paper has been accomplished to provide an overview
-about MC linear solvers. In particular the goal was to combine the main
-theoretical results found in literature with additional novel contributions on
-both the theoretical and empirical sides. This made it possible to understand
-the actual limitations of the presented algorithms.
-
-Necessary and sufficient conditions to ensure convergence are too costly for
-most of the applicative problems coming fro engineering and Computational
-Physics. On the other hand, sufficient criteria are currently viable for a
-restricted set of matrices (generalized diagonally dominant).
-
-An empirical analysis has been carries out on parabolic problems and it has
-been shown that a proper selection of the time discretization step enables the
-employment of stochastic solvers.
-
-We reserve for future works the analysis of the plausible connections between
-$\rho(H)$, $\rho(\hat{H})$ and the number of numerical iterations necessary to
-reach convergence. Moreover a study about the resilience of the problem will be
-accomplished, resorting to fault injection techniques.
+\input{intro}
+\input{mcls}
+\input{convergence}
+\input{results}
+\input{conclusion}
 

### 2015-09-16 MAX_reports/MC_solvers_article/conclusion.tex
Breaking paper into multiple files for easier editing/collaborating.

\section{Conclusions and Future Work}

The work presented in this paper has been accomplished to provide an overview
about MC linear solvers. In particular the goal was to combine the main
theoretical results found in literature with additional novel contributions on
both the theoretical and empirical sides. This made it possible to understand
the actual limitations of the presented algorithms.

Necessary and sufficient conditions to ensure convergence are too costly for
most of the applicative problems coming fro engineering and Computational
Physics. On the other hand, sufficient criteria are currently viable for a
restricted set of matrices (generalized diagonally dominant).

An empirical analysis has been carries out on parabolic problems and it has
been shown that a proper selection of the time discretization step enables the
employment of stochastic solvers.

We reserve for future works the analysis of the plausible connections between
$\rho(H)$, $\rho(\hat{H})$ and the number of numerical iterations necessary to
reach convergence. Moreover a study about the resilience of the problem will be
accomplished, resorting to fault injection techniques.
i

### 2015-09-16 MAX_reports/MC_solvers_article/convergence.tex
Breaking paper into multiple files for easier editing/collaborating.

\section{Convergence Behavior of Stochastic Methods}

The requests imposed on the expected value and second moment of the estimators
can be reformulated in a purely deterministic setting by looking at the
spectral radii of some properly defined matrices.

For instance the condition of finiteness of the expected value can be
reformulated by requiring
\begin{equation}
 \rho(H)<1
 \label{rho_h},
\end{equation}

where $H$ is the iteration matrix of the fixed point scheme.
In fact both equations (\ref{dir_mean}) and (\ref{adj_mean}) can be considered
as
power series in terms of $H$ and the spectral radius of $H$ smaller
than 1 is a necessary and sufficient condition for the Neumann series to
converge.

By analogy we aim at reproducing a similar reasoning for the second moment of
the estimator, since we want it to be finite as well.
By looking at the equations of the variance (\ref{dir_var}) and (\ref{adj_var})
for
the Forward and the Adjoint method respectively, we notice that the second
moment can be reinterpreted as a power series with respect to the matrices
defined as follows:

\[
 \hat{H}_{i,j}=\frac{H_{i,j}^2}{P_{i,j}} \quad \text{- Forward Method}.
\]

or

\[
 \hat{H}_{i,j}=\frac{H_{j,i}^2}{P_{i,j}} \quad \text{- Adjoint Method}.
\]

Accordingly to the necessary and sufficient condition for the power series to
converge, we require
\begin{equation}
 \rho(\hat{H})<1.
 \label{rho_star}
\end{equation}

Condition (\ref{rho_h}) must be required for a generic fixed point scheme to
reach convergence, whereas the extra condition (\ref{rho_star}) is typical of
the stochastic schemes presented in this work. Having two conditions to
verify makes the numerical treatment of the linear system via MC very
difficult. We will discuss this in details later in the treatise. \newline

\subsection{Necessary and Sufficient Conditions}

Here below some results presented in
\cite{MASC2013} about necessary conditions and sufficient
conditions for
convergence are enunciated. In particular a suitable choice for constructing a
transition probability is discussed to guarantee and speed-up convergence.

We remember that the construction of the transition probability must satisfy
some constraints such as

\[
\begin{cases}
  P_{i,j}\ge 0 \\
 \sum_{j=1}^N P_{i,j}=1
\end{cases}
\]

plus one more restriction which varies depending on the particular Monte Carlo
approach chosen

\begin{align*}
& \text{Forward Method:} \quad H_{i,j}\ne 0 \Rightarrow P_{i,j}\ne 0 \\ &
\text{Adjoint Method:} \quad H_{j,i}\ne 0 \Rightarrow P_{i,j}\ne 0
\end{align*}.

The previous restrictions on the attainable values for the entries of the
transition probability are called \textit{transition conditions}.

At first an introductory theoretical result is shown, necessary to motivate the
choices made afterwards for the choice of the transition probability.

\begin{thm}
 Consider a vector $a=(a_1,a_2,\cdots,a_N)^T$ where at least one element is
nonzero, $a_k\ne0$ for some $k\in\{1,\cdots,N\}$.
\begin{itemize}
 \item For a probability vector $p=(p_1,p_2,\cdots,p_N)^T$ satisfying the
transition conditions, the lower bound of $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is $\bigg(\sum_{k=1}^N \lvert a_k\rvert\bigg)^2$
\item There always exists a probability vector such that $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}\ge c\ge 1$ for all $c>1$.
\end{itemize}
\label{lemma}
\end{thm}

It might be observed that the minimum of the quantity $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is attained when the probability vector is
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
a_k\rvert}$. It implies that the $\infty$-norm of $\hat{H}$ is minimized as
much
as possible. \newline

For the sake of simplicity let us introduce now a generic random walk
truncated at a certain $k$-th step
\[
 \gamma_k:\; r_0\rightarrow r_1 \rightarrow r_2 \rightarrow \cdots \rightarrow
r_k
\]
and the statistical estimator
\[
 X(\gamma_k)=\frac{H_{r_0,r_1}H_{r_1,r_2}\cdots
H_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots P_{r_{k-1},r_k}}f_{r_k}
\]
which is associated with the Forward Method.

Then the following theorem holds
\begin{thm}\textit{(Suited to the Forward Method)}

Let $H\in \mathbb{R}^{n\times n}$ such that $\lVert H\rVert_{\infty}<1$.
Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the
$k$-th step. Then,
there always exists a
transition matrix $P$ such that
$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and
$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
\label{for_thm}
\end{thm}

If we introduce the estimator
\[
 X(\gamma_k)=\frac{H^T_{r_0,r_1}H^T_{r_1,r_2}\cdots
H^T_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots
P_{r_{k-1},r_k}}sign(f_{r_0})\lVert \mathbf{f}\rVert_1
\]
which is associated with the Adjoint Method, then we can state a
theorem specular to \ref{for_thm}.

\begin{thm}\textit{(Suited to the Adjoint Method)}

 Let $H\in \mathbb{R}^{n\times n}$ with $\lVert H\rVert_{1}<1$.
Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the
$k$-th step. Then,
there always exists a
transition matrix $P$ such that
$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and
$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
\label{adj_thm}
\end{thm}

In particular, relying on what guaranteed by Theorem \ref{lemma}, in both
Theorems \ref{for_thm} and \ref{adj_thm} the thesis holds by picking the MAO
transition probability.

These theoretical results represent sufficient conditions for the convergence
of the
Forward and Adjoint Monte Carlo and they can be easily checked.
However requiring $\lVert
H\rVert_{\infty}<1$ or $\lVert H\rVert_1<1$ may be too demanding.
Indeed, even if there exist some preconditioners able to guarantee this
condition (e.g. approximate inverse preconditioners with a proper value of
dropping
tolerance), it may entail to compromise the sparsity of the
iteration
matrix.

Thusly, in situations where the hypotheses of Theorems \ref{for_thm} and
\ref{adj_thm} are too demanding to be verified, the necessary and sufficient
condition
on the
spectral radius of $\hat{H}$ is the only criterion to take as reference.
Nonetheless this constraint consists in
computing the spectral radii of two matrices and the calculation
might be very costly. Moreover the computation of the spectral radii is
affected by resilience issues as well as
a deterministic solver for the resolution of
a linear system. Therefore the original difficulty is transferred without being
actually tackled.

\subsubsection{Forward MC}

\subsubsection{Adjoint MC with Collision Estimator}

\subsubsection{Adjoint MC with Expected-Value Estimator}

\subsection{Construction of Transition Probabilities}

The way the transition probability is defined is extremely important. In fact
in some circumstances it might determine whether the stochastic scheme
converges or not. Two very simple ways of defining it are listed here
below. These are the approaches adopted typically in literature.

\subsubsection{Uniform probabilities}
With this approach transition matrix $P$ is such that all the non-zero elements
in each row have equal probability of occurring.
\[
\text{Forward}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{i,j}=0 \\
\frac{1}{\#(\text{non-zeros in the row)}} \quad \text{if} \quad H_{i,j}\ne 0
\end{cases}\quad
\]
\[
\text{Adjoint}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{j,i}=0 \\
\frac{1}{\#(\text{non-zeros in the column)}} \quad \text{if} \quad H_{j,i}\ne 0
\end{cases}
\]
The Monte Carlo approach resorting to this definition of the transition matrix,
in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).

\subsubsection{Weighted probabilities}
An alternative way of defining the transition matrix consists in attributing
probabilities to all the entries of the transition probability accordingly
to the magnitude of the elements. For instance we may employ the following
definition
\[
\text{Forward}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{i,j}\rvert^p}{\sum_{k=1}^n
\lvert H_{i,k}\rvert^p}
\]

\[
\text{Adjoint}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{j,i}\rvert^p}{\sum_{k=1}^n
\lvert H_{k,i}\rvert^p}
\]

where $p\in \mathbb{N}$.
This way of proceeding, for $p=1$, still basing on the work \cite{AADBTW2005},
is called
\textit{Monte Carlo Almost Optimal} (MAO).
The reason why the almost optimal probability is called this way is associated
with a notice from Theorem \ref{lemma}. Indeed from
the mentioned lemma we know that the quantity $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is minimized when the probability vector is
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
a_k\rvert}$. This implies that the almost optimal probability minimizes the
$\infty$-norm of $\hat{H}$ in the Forward method and the $1$-norm of $\hat{H}$
in the
Adjoint method. This is very important to be pointed out since these two norms
are so far the best practical upper bounds for $\rho(\hat{H})$ that we can
easily
handle. In the Section dedicated to the numerical results it will be proved
that the Almost Optimal Probability eases the convergence of the MC solver,
even in cases when the employment of the Uniform Probability does not provide
convergence at all.

\subsection{Classes of Matrices with Guaranteed Convergence}

As we have already seen, sufficient conditions for
the convergence of the Monte Carlo linear solvers are very restrictive (see
\cite{MASC2013}), whereas the necessary and sufficient condition requires the
computation of
$\rho(H)$and $\rho(\hat{H})$, which is very expensive in general. Once a
proper
preconditioner is picked (e.g. Approximate Inverse), $\rho(H)<1$ may
be always guaranteed, whilst $\rho(\hat{H})<1$ is more
problematic to be verified.

In order to bypass the cost of these checks, it is useful to pursue for sets of
matrices for
which the Monte Carlo linear solver always converges, maybe with the use of an
ad-hoc preconditioner.

\subsubsection{Strictly Diagonally Dominant}

 One of the these sets is
represented by strictly diagonally dominant matrices.

\begin{defn}
 A matrix $A\in\mathbb{R}^{n\times n}$is strictly diagonally dominant (s.d.d)
by rows if
 \begin{equation}
    \lvert a_{ij}\rvert>\sum_{\substack{i=1\\ i\ne j}}^{i=n}\lvert
a_{ij}\rvert
\label{sddr}
 \end{equation}
\end{defn}

\begin{defn}
 A matrix $A\in\mathbb{R}^{n\times n}$is strictly diagonally dominant (s.d.d)
by columns if
 \begin{equation}
    \lvert a_{ij}\rvert>\sum_{\substack{j=1\\ j\ne i}}^{j=n}\lvert
a_{ij}\rvert
\label{sddc}
 \end{equation}
\end{defn}

When (\ref{sddr}) holds we can resort to a left diagonal preconditioning and we
get an iteration matrix $H=I-D^{-1}A$ such that $\lVert H \rVert_{\infty}<1$.

Introducing a MAO transition probability for the Forward Method we get
\[
 P_{ij}=\frac{\lvert H_{ij}\rvert}{\sum_{k=1}^n\lvert H_{ik}\rvert}
\]
therefore the entries of $\hat{H}$ are defined as follows
\[
 \hat{H}_{ij}=\frac{H^2_{ij}}{P_{ij}}=\lvert H_{ij}\rvert\bigg
(\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ).
\]
Therefore
\[
 \sum_{j=1}^n\lvert \hat{H}_{ij} \rvert = \sum_{j=1}^n \hat{H}_{ij} =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg )\bigg (\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ) =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg ) ^2 <1 \quad \forall i=1,\cdots, n.
\]
This automatically implies that $\rho(\hat{H})<=\lVert
\hat{H}\rVert_{\infty}<1$. Thus the Forward Monte Carlo converges.
However nothing a priori can be
stated about the behavior of the Adjoint method.\newline

Instead if (\ref{sddc}) holds we can resort to a right diagonal preconditioning
and we
get an iteration matrix $H=I-AD^{-1}$ such that $\lVert H \rVert_{1}<1$.
In this case, by following a similar reasoning to the one made before, we
conclude that the Adjoint Method convergences thanks to $\lVert
\hat{H}\rVert_1<1$,
whereas nothing can be said in advance as concerns the Forward method.

\subsubsection{Generalized Diagonally Dominant}

 Another set of matrices for which the convergence of
the MC solvers is guaranteed is made of \textit{generalized diagonally
dominant matrices}.

\begin{defn}
A square matrix $A\in\mathbb{R}^{n\times n}$ is said to be generalized
diagonally dominant if
\[
 \lvert a_{ii}\rvert x_i \ge \sum_{\substack{j=1\\j\ne i}}^n \lvert
a_{ij}\rvert
x_j, \quad i=1,\cdots,n
\]
for some positive vector $\mathbf{x}=(x_1,\cdots,x_n)^T$.
\end{defn}

A proper subset of the generalized diagonally dominant matrices is represented
by $M$- matrices (see \cite{Ax1996}).
There are many definitions of $M$-matrices that may be found on books and all
of them are equivalent. The one presented in \cite{Saad} states the
following

\begin{defn}
A matrix $A\in\mathbb{R}^{n\times n}$ is said to be an $M$-matrix if it
satisfies the following four properties:
\begin{itemize}
 \item $a_{i,i}>0$ for $i=1,\cdots,n$
 \item $a_{i,j}\le 0$ for $i\ne j$, $i,j=1,\cdots,n$
 \item $A$ is nonsingular
 \item $A^{-1}\ge 0$
\end{itemize}
\end{defn}

A very important result about $M$-matrices enables to come up with a convergent
splitting and, therefore, a convergent fixed point scheme.

In particular Jacobi, Block Jacobi and Gauss Seidel  are
convergent
splitting for an $M$-matrix.
However, the construction of a splitting like this is not enough to ensure the
convergence of
the Monte Carlo linear solver. In fact none of these conditions is sufficient
to guarantee $\rho(\hat{H})<1$. Because of this obstacle, other options must
be considered.\newline

For this purpose we
appeal to a result in \cite{Li2002}. In this paper the author presents an
algorithm to transform a generalized diagonally dominant matrix with
nonzero diagonal entries into strictly diagonally dominant. The
algorithm works for a generic complex matrix $A\in\mathbb{C}^{n\times n}$.

Here below we report the algorithm at hand\newline

\begin{algorithm}[H]
 For a given complex matrix $A$, $a_{ii}\ne 0$, $i=1,\cdots, n$;\newline
 \begin{enumerate}
  \item Compute $S_{i}=\sum_{\substack{j=1 \\ j\ne i}^{n}}\lvert
a_{ij}\rvert$,
$i=1,2,\cdots,n$
\item Set $t=0$. For $i=1,2,\cdots, n$, if $\lvert a_{ii}\rvert>S_i$, then set
$t=t+1$
\item If $t=0$, then print "A is not a GDDM": END
\item If $t=n$, then print "A is a GDDM": END
\item \For{i=1,n}{$d_{i}=\frac{S_i+\varepsilon}{\lvert
a_{ii}\rvert+\varepsilon} \quad
\varepsilon>0,\quad j=1,2,\cdots,n$\;
$a_{ji}=a_{ji}\cdot d_i$}
\item Go to step 1.
 \end{enumerate}
 \caption{Algorithm to turn a GDDM matrix into a s.d.d by rows.}
\end{algorithm}

This approach turns a generalized diagonally dominant matrix into a
strictly diagonally dominant matrix by rows. By substituting
$\displaystyle S_{i}=\sum^{n}_{\substack{j=1 \\ j\ne i}}\lvert a_{ij}\rvert$
at step 1 with
$\displaystyle S_{i}=\sum^{n}_{\substack{i=1 \\ i\ne j}}\lvert a_{ij}\rvert$
and by replacing
$a_{ji}=a_{ji}\cdot d_i$ with $a_{ji}=a_{ji}\cdot d_j$ we obtain the algorithm
that turns a GDDM matrix into a s.d.d. by columns.\newline

Once we have applied this transformation to the matrix at hand into a s.d.d.,
we can use the Monte Carlo linear solver which is ensured to converge.

\subsubsection{Block Diagonally Dominant}

 In this section we wonder in which situations a block diagonal preconditioning
might come up with a convergent Monte Carlo linear solver.
By mimicking the computations we already showed previously, the iteration
matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal
preconditioning takes the form

\[
 H=I-D^{-1}A=\begin{bmatrix}0_{n_1\times n_1} & -[A_{11}]^{-1}A_{12} & \cdots &
\cdots & -[A_{11}]^{-1}A_{1p} \\
-[A_{22}]^{-1}A_{21} & 0_{n_2\times n_2} & -[A_{22}]^{-1}A_{23} &
\cdots & -[A_{22}]^{-1}A_{2p}\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
-[A_{pp}]^{-1}A_{p1} &  \cdots & \cdots&
-[A_{pp}]^{-1}A_{p,p-1} & 0_{n_p \times n_p}
\end{bmatrix}.
\]

For the sake of clarity, we notify that the symbol "$\%$" used in the following
subsections represents the modulus function between two integer numbers, while
the floor function is represented with pretty standard symbol "$\lfloor
\cdot \rfloor$".

\paragraph{Forward method}
By assuming that all $n_i$'s have the same value we may define
\[
 m=n_i=\text{size of a block}=\frac{n}{p}.
\]

The MAO transition probability matrix becomes:
\[
 P_{i,j}=\frac{\lvert H_{i,j} \rvert}{\sum_{k=1}^n\lvert H_{i,k}
\rvert}=\frac{\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor]
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg
\rvert}{\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg
([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert}
\]

Consequently, the $\hat{H}$ matrix is defined such that
\[
\hat{H}_{i,j} = \lvert H_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert H_{ik}\rvert\bigg)=
\bigg \lvert \bigg ([A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg
\rvert
\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert
\]

By computing the sum over a generic row of $\hat{H}$ we get:
\[
 \sum_{j=1}^n \lvert \hat{H}_{i,j}\rvert=\sum_{j=1}^n \hat{H}_{i,j} =
 \bigg ( \sum_{\substack{j=1\\j\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert \bigg ) ^2.
\]

If we focus on the norm $\lVert \hat{H}\rVert_{\infty}$, then the following
equivalence condition holds:
\[
 \lVert \hat{H}\rVert_{\infty}<1 \Leftrightarrow \sum_{\substack{j=1\\j\ne
i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert <1 \quad \forall
i=1,\cdots,n
\]
 A sufficient condition for this to happen is
 \begin{equation}
  \sum_{\substack{j=1\\j\ne i}}^p \lVert [A_{ii}]^{-1}A_{ij}\rVert_{\infty}<1.
    \label{block_cs}\quad \forall i=1,\ldots,p.
 \end{equation}

By defining a matrix $\tilde{H}\in \mathbb{R}^{p\times p}$ such that
\[
 \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{\infty} &
\cdots &
\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{\infty} \\
\lVert [A_{22}]^{-1}A_{21}\rVert_{\infty} & 0 & \lVert
[A_{22}]^{-1}A_{23}\rVert_{\infty} &
\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{\infty} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert [A_{pp}]^{-1}A_{p1}\rVert_{\infty} &  \cdots & \cdots&
\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{\infty} & 0
\end{bmatrix}
\]

We can use the $\tilde{H}$ matrix just defined in order to introduce a
sufficient condition for the convergence of the Forward Monet Carlo method with
a Block Diagonal preconditioning.

\begin{equation}
 \lVert \tilde{H} \rVert_{\infty}<1 \Rightarrow \lVert \hat{H}
\rVert_{\infty}<1.
\end{equation}

\paragraph{Adjoint method}
Analogously to the Forward method, if we define
\[
(\hat{H})^T_{i,j} = \lvert H^T_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert
H^T_{ik}\rvert\bigg)
\]
we can formulate a sufficient condition for the convergence of the Adjoint
Monte
Carlo method by introducing a matrix $\tilde{H}$ which in this case is such that
\[
  \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{1} & \cdots
&
\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{1} \\
\lVert [A_{22}]^{-1}A_{21}\rVert_{1} & 0 & \lVert
[A_{22}]^{-1}A_{23}\rVert_{1} &
\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{1} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert [A_{pp}]^{-1}A_{p1}\rVert_{1} &  \cdots & \cdots&
\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{1} & 0
\end{bmatrix}.
\]
The sufficient condition assumes the form
\begin{equation}
 \lVert \tilde{H} \rVert_{1}<1 \Rightarrow \lVert \hat{H} \rVert_{1}<1.
\end{equation}

\subsection{Adaptive Methods}

In formulas (\ref{dir_mean}) and (\ref{adj_mean}) the estimation of the
solution to
the linear system (\ref{linsys_prec}) involves infinite sums. Of course this is
not viable for
computational issues, therefore it is necessary to resort to truncation of
the sums through proper criteria. In details, the truncation implies the
employment of a finite number of permutations and a finite number of
steps for each one of the permutations computed. Therefore it is necessary to
truncate at two different levels to decide both how many permutations to employ
and how many steps each one of them should be kept running.

\subsubsection{History Length}

At first we focus on the definition of a weight cut-off
in order to decide where to terminate the histories. It may be reinterpreted as
a quantity guiding us to decide how many terms of the Neumann
series to consider. \newline

The goal is to apply the cutoff via an automatic procedure, without requiring
the intervention of the user directly.
Therefore we are looking for an integer $m$ such that
\[
\tilde{\theta}_i=E\bigg[\sum_{\ell=0}^m W_{\ell}
f_{k_{\ell}}\bigg]=x_i=\sum_{\ell=0}^m
\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n
P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},
k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots
w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}
\]

and

\[
\tilde{\theta}_j=E\bigg[\sum_{\ell=0}^m W_{\ell}\delta_{k_{\ell},
j}\bigg]=\sum_{\ell=0}^{m}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}
\]

are good approximations of (\ref{dir_mean}) and (\ref{adj_mean}) respectively.

In \cite{Slattery2013} there is a criterion for the cutoff of the random walk
applicable to both the Forward and the Adjoint method. It requires the set up
of a relative cutoff threshold $W_c$ and to look for a step $m$ such that
\begin{equation}
W_m \le W_c W_0.
\label{cutoff}
\end{equation}
$W_0$ is the value of the weight at the initial step of the random walk and
$W_m$ is the value of the weight after $m$ steps.

\subsubsection{Number of Random Walks}

We now focus on quantities that might lead to a proper
number of permutations to compute, in order to get a prescribed order of
accuracy.
As regards the Forward method, we know that the expression for the variance
is defined in formula (\ref{dir_var}).

In this context, a reasonable criterion to determine the number $\tilde{N}_i$
of random walks to be run is setting a threshold $\varepsilon_1$ and determine
\begin{equation}
\tilde{N}_i \quad s.t.\quad \frac{\sqrt{Var[\theta_i]}}{\lvert
E[\theta_i]\rvert}<\varepsilon_1, \quad i=1,\cdots,n.
\label{forward_adapt}
\end{equation}
The dependence of $Var[\theta_i]$ and $E[\theta_i]$ on $\tilde{N}_i$, which
seems
to be absent in the previous formula, is highlighted by the fact that
$\theta_i$
is estimated by fixing a finite number of histories.
Therefore we are controlling the relative standard deviation requiring it not
to be too large. In other words we are pursuing a statistical setting where the
uncertainty factor is not dominating over the expected value.
This simple adaptive approach can be applied for the estimation of each
component $x_i$. Hence different number of histories may be employed to
compute different entries of the solution vector. \newline

As concerns the Adjoint method, the estimation of the variance for each entry
is reproduced by Formula (\ref{adj_var}).

A possible adaptive selection of $\tilde{N}$, in this situation, is
\begin{equation}
\tilde{N} \quad s.t. \quad \frac{\lVert
\boldsymbol{\sigma}_{\tilde{N}}\rVert_1}{\lVert
\mathbf{x}\rVert_1}<\varepsilon_1,
\label{adjoint_adapt}
\end{equation}
where $\boldsymbol{\sigma}$ is a vector whose entries are
$\boldsymbol{\sigma}_{\tilde{N},i}=Var[\theta_i]$.

What introduced above can be exploited in order to build
an a posteriori adaptive algorithm, capable to identify the minimal value of
$\tilde{N}$ that verifies (\ref{forward_adapt}) or (\ref{adjoint_adapt})
respectively.

Below there are the pseudo-codes associated with both Forward and Adjoint Monte
Carlo.

\begin{algorithm}[H]
 \KwData{$N$, $\varepsilon_1$}
 \KwResult{$\tilde{N}_i$, $\sigma_i$, $x_i$}
 For each entry of the solution vector (for $i=1,\cdots,n$) \;
 $\tilde{N}_i=N$\;
 compute $\theta_i$\;
 \While{$\frac{\sigma_i}{\lvert
E[\theta_i]\rvert}<\varepsilon_1$}{
\vspace{0.1cm}
  $\tilde{N}=\tilde{N}+N$\;
  compute $E[\theta_i]=x_i$\;
 }
 return $\theta_i$, $x_i$, $\sigma_i$\;
 \caption{A posteriori adaptive Forward Monte Carlo}
\end{algorithm}

\begin{algorithm}[H]
 \KwData{$N$, $\varepsilon_1$}
 \KwResult{$\tilde{N}$, $\boldsymbol{\sigma_i}$, $\mathbf{x}$}
 $\tilde{N}=N$\;
 compute $\theta$\;
 \While{$\frac{\lVert\boldsymbol{\sigma} \rVert}{\lVert
E[\theta]\rVert}<\varepsilon_1$}{
\vspace{0.1cm}
  $\tilde{N}=\tilde{N}+N$\;
  compute $E[\theta]=\mathbf{x}$\;
 }
 return $\theta$, $\mathbf{x}$, $\boldsymbol{\sigma}$\;
 \caption{A posteriori adaptive Adjoint Monte Carlo}
\end{algorithm}

\subsection{Preconditioning}

In this section we focus on the pursuit of preconditioners that facilitate the
convergence of the stochastic schemes introduced so far. A priori, all of the
following techniques are good candidates to improve the numerical setting.
However issues about the fill-in effect and the verification of convergence
conditions limit their employment for our purposes.

\subsubsection*{Block Preconditioning}
Consider a matrix $A\in\mathbb{R}^{n\times n}$ of the form
\[
 A=\begin{bmatrix}A_{11} & A_{12} & \cdots & A_{1p}\\
    A_{21} & A_{22} & \cdots & A_{2p}\\
    \vdots & \vdots & \ddots & \vdots \\
    A_{p1} & \cdots & \cdots & A_{pp}
   \end{bmatrix}
\]
where $p$ is a divisor of $n$ and such that $A_{ii} \in \mathbb{R}^{n_i
\times n_i}$ are nonsingular square matrices.
By defining a block diagonal matrix such as
\[
 D=\begin{bmatrix}A_{11} & 0_{12} & \cdots & 0_{1p}\\
    0_{21} & A_{22} & \cdots & 0_{2p}\\
    \vdots & \vdots & \ddots & \vdots \\
    0_{p1} & \cdots & \cdots & A_{pp}
   \end{bmatrix}
\]
it is possible to compute its inverse.
$D^{-1}$  can be used as a preconditioner for the original matrix $A$.
In particular
\[
 D^{-1}A=\begin{bmatrix}I_{n_1\times n_1} & A_{11}^{-1}A_{12} & \cdots &
\cdots & A_{11}^{-1}A_{1p} \\
A_{22}^{-1}A_{21} & I_{n_2\times n_2} & A_{22}^{-1}A_{23} &
\cdots & A_{22}^{-1}A_{2p}\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
A_{pp}^{-1}A_{p1} &  \cdots & \cdots&
A_{pp}^{-1}A_{p,p-1} & I_{n_p \times n_p}
\end{bmatrix}.
\]
The ultimate goal is to resort to this preconditioning technique in order to
achieve the condition $\rho(D^{-1}A)<1$. In fact this is necessary in order to
reformulate \ref{linsys} into \ref{fixedpoint}.

Convergence of \ref{fixedpoint} with block diagonal preconditioning, from the
deterministic point of view, is guaranteed when $A$ is an irreducibly
diagonally dominant matrix or when $A$ is an $M$-matrix. In fact in situation
such as these, theoretical results guarantee that $A=D-N$ is a regular
splitting (see \cite{Ax1996}). However the same theoretical results cannot be
exploited to guarantee the convergence of the MC solvers, because of the
additional check on $\rho(\hat{H})<1$ to be considered.

\subsubsection*{Approximate Inverse of a Sparse Matrix (AINV)}

This stimulated our interests into other types of preconditioners. At first we
concentrated on Approximate Inverse preconditioners (see \cite{Saad} and
\cite{Benzi2002}).

There are many practical techniques that may be applied in order to build the
AINV preconditioner.
One of these is described in \cite{Tuma1996} and \cite{Tuma1998}.
This is the actual approach we applied to the construction of the
preconditioner. The technique constructs the preconditioner via a
bi-conjugation fashion.
The idea is to compute an approximate factorization of the form $W^T A Z=D$.
$W$ and $Z$ are unit upper triangular matrices and $D$ is a diagonal matrix.

In order to compute the AINV preconditioner, a \texttt{FORTRAN} code
provided by Miroslav Tuma (Academy of Sciences, Prague) is used.
We report as follows the algorithm employed for the construction of the so
called \textit{Right-looking Factored AINV}. The algorithm provide the user
with a dropping tolerance $\tau$ which can be used in order to tune the fill-in
effect
of the factored preconditioner. \newline

\begin{algorithm}[H]
 \KwData{$p=q=(0,\cdots,0)\in \mathbb{R}^n$, $Z=[z_1,\cdots,z_n]=I_n$,
$W=[w_1,\cdots,w_n]=I_n$}
 \KwResult{$Z$, $W$}
 \For{$k=1,\cdots, n$}{
 $p_k=w_k^T A e_k$, $q_k=e_k^T A z_k$ \;
     \For{$i=k+1,\cdots,n$}{
         $p_i=(w_k^T A e_k)/p_k$, $q_i=(e_k^T A z_i)/q_k$\;
         Apply a dropping rule to $p_i$, $q_i$\;
         $w_i=w_i - w_k p_i$, $z_i=z_i - z_k q_i$\\
         Apply a dropping rule to $w_{j,i}$ and $z_{j,i}$ for $j=1,\cdots,i$\;
     }\;
  }\;
 Choose diagonal entries of $D$ as the component of $p$ or $q$\;
 return $Z$, $W$\;
 \caption{Right-looking Factored AINV}
\end{algorithm}

\subsubsection*{Incomplete LU Factorization Preconditioners (ILU)}
This set of preconditioners has the goal of computing a sparse lower triangular
matrix $L$ and a sparse upper triangular matrix $U$ such that the residual
matrix $R=LU-A$ respects some constraints. One of these constraints may be the
fact that some entries must be set to zero.
For instance we can introduce a zero pattern set
\[
 P\subset \{(i,j)\lvert i\ne j; 1\le i,j\le n\}
\]
and use this set in order to impose some constraints on the shape of the
preconditioner built.\newline

Different variants of the
Incomplete LU Factorization algorithm are available (see \cite{Saad}).
However it is
possible to prove that if all the formulation are well defined, then they are
all equivalent.
One of these variants is the following

\begin{algorithm}[H]
 \KwData{$P$}
 \KwResult{$A$}
 For each $(i,j)\in P$ set $a_{ij}=0$\;
 \For{$k=1,\cdot,n-1$}{
 \For{$i=k+1,n$}{
 \If{$(i,k)\notin P$}{
 $a_{ik}=\frac{a_{ik}}{a_{kk}}$
 \For{$j=k+1,\cdots, n$ and for $(i,j)\notin P$}{
  $a_{ij}=a_{ij}-a_{ik}*a_{kj}$
 }
 }
 }
 }
 return $Z$, $W$\;
 \caption{General Static Pattern ILU}
\end{algorithm}

This kind of algorithm is guaranteed to terminate without any breakdown just
for $M$-matrices.

Assuming that the zero pattern $P$ coincides with the zero pattern of $A$ leads
to the $ILU(0)$. \newline

\begin{algorithm}[H]
 \KwData{$P$}
 \KwResult{$A$}
 \For{$i=2,\cdot,n$}{
 \For{$k=i,i-1$ and for $(i,k)\in NZ(A)$}{
  $a_{ik}=\frac{a_{ik}}{a_{kk}}$
  \For{$j=k+1,\cdots,n$ and for $(i,j)\in NZ(A)$}{
  $a_{ij}=a_{ij}-a_{ik}a_{kj}$}
 }
 }
 \caption{ILU(0).}
\end{algorithm}

More sophisticated variants of the algorithm enable to enlighten the constraint
of the sparsity pattern, finding a compromise between sparsity and accuracy in
approximating the inverse of $A$. We refer to \cite{Saad} and
\cite{Benzi2002} for further information about it.
i

### 2015-09-16 MAX_reports/MC_solvers_article/intro.tex
Breaking paper into multiple files for easier editing/collaborating.

\section{Introduction}

For a long period the
pursuit of a reliable solver for linear and nonlinear algebraic systems has
been one of the most important issues in Numerical Analaysis. The
mathematical properties of the model (e.g. likely sparsity pattern of the
matrix) have urged scientists to look for efficient and pattern-driven
solvers, capable to cope with curse of dimensionality and high scalability
requests.

As scalability computing move towards exascale facilities (which stands for
machines computing up to $10^{18}$ flops), new numerical schemes suitable for
this
kind of architectures are strongly demanded. In fact the purpose is to combine
high
fidelity of the results with an optimized use of hardware resources at hand.

Recently new algorithms that combine statistical and numerical techniques have
been developed. Even though \textit{numerical efficiency} has been sacrificed
sometimes, the advantages in terms of \textit{robustness} of these methods make
it worth carrying on studies in this direction.
A class of methods that may be employed for this purpose is represented by
Monte Carlo linear solvers.

These algorithms are the main topic covered in the work presented here,
starting from what was developed so far by some of
the authors who gave significant contributions (see \cite{Hal1962},
\cite{Hal1994},
\cite{DA1998}, \cite{DVA2001}, \cite{AADBTW2005},\cite{ESW2013} and
\cite{EMSH2014}). \newline

As underlined in \cite{DA1998}, Monte Carlo methods may be split into two
classes: \textit{direct methods} (\cite{DA1998}, \cite{DVA2001}) and
\textit{iterative methods} (\cite{Hal1962},
\cite{Hal1994}, \cite{ESW2013}
and \cite{EMSH2014}). The first are characterized by a merely stochastic
scheme,
therefore the provided error with respect to the exact solution is made of
just a stochastic component. The second ones, instead, combine numerical and
statistical schemes generating two types of error: a
\textit{stochastic} one and a \textit{systematic} one. It does not
mean
that
it will be always simple to recognize them separately. However it is important
to
be
aware of this intrinsic structure, in order to target what is the part of the
scheme that requires a refinement (e.g. increase of numerical iterations
rather than random walks).\newline

The paper is organized as follows. In Section 2 we describe all the
viable stochastic settings for the definition of the Markovian process. In
Section 3 we reinterpret probabilistic facts in terms of spectral properties of
the linear system to be solved. Section 4 is dedicated to necessary conditions
and sufficient conditions for the convergence of the algorithms. In Section 5
different adaptive approaches are presented, in order to truncate the random
walks and to decide how many samplings to use. Section 6 regards the study
of matrices for which the convergence of the stochastic schemes is facilitated.
Then in Section 7 a hybrid scheme combining numerical and probabilistic
techniques is described: the \textit{Monte Carlo Synthetic Acceleration}.
Numerical results are presented in Section 8 and 9, while Section 10 is
dedicated to
conclusive remarks and goals for the future.

### 2015-09-16 MAX_reports/MC_solvers_article/mcls.tex
Breaking paper into multiple files for easier editing/collaborating.

\section{Stochastic Linear Solvers}

We consider the solution of systems of linear equations of the form
\begin{equation}
A \mathbf{x}=\mathbf{b},
\label{linsys}
\end{equation}
where $A\in \mathbb{R}^{n\times n}$, $\mathbf{x}$, $\mathbf{b} \in
\mathbb{R}^n$.

By applying a left preconditioning, the system (\ref{linsys}) gets the
following shape:
\begin{equation}
P^{-1}A \mathbf{x}=P^{-1}\mathbf{b},
\label{linsys_prec}
\end{equation}
(\ref{linsys_prec}) can be
reinterpreted as a fixed point scheme
\begin{equation}
 \mathbf{x}=H\mathbf{x}+\mathbf{f}.
 \label{fixedpoint}
\end{equation}
where $H=I-P^{-1}A$ and $\mathbf{f}=P^{-1}\mathbf{b}$.

Assuming that the spectral radius $\rho(H)<1$, the solution to
(\ref{fixedpoint}) can be written in terms of a power series of
$H$ (Neumann series)
\[
\mathbf{x}=\sum_{i=0}^\infty H^i\mathbf{f}.
\]
Therefore the fixed point scheme generates a sequence of approximate solutions
$\{\mathbf{x}^{(k)}\}_{k=0}^{\infty}$ which converges to the exact solution
regardless of the initial guess $\mathbf{x}_0$.

By restricting the attention to a single component of $\mathbf{x}$ we
have
\begin{equation}
x_i=\sum_{\ell=0}^\infty \sum_{k_1}^n\sum_{k_2}^n\cdots \sum_{k_{\ell}=1}^n
H_{k_0,k_1}H_{k_1,k_2}\cdots H_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{forward}
\end{equation}
The last equation can be reinterpreted as the realization of an estimator
defined on a random walk.

Let us start considering a random walk whose
state space $S$ is characterized by the set of indices of the forcing term
$\mathbf{f}$:
\[
S=\{1,2,\cdots, n\} \subset \mathbb{N}.
\]
Each $i$-th step of the random walk has a random variable
$k_i$ associated with it. The realization of $k_i$ represents the index of the
component of $\mathbf{f}$
which is visited in the current step of the random walk.

The way the transition probability is built and the way the initial state of
the random walk is chosen give birth to two different approaches we are
going to use: the \textit{Forward} and \textit{Adjoint} methods.

\subsection{Forward Method}

The goal is to evaluate a functional such as
\[
J(\mathbf{x})=(\mathbf{h},\mathbf{x})=\sum_{i=1}^n h_i x_i.
\]
where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in
$\mathbb{R}^n$ of the functional $J$.
We can use it to build the initial probability $\tilde{p}:
S\rightarrow [0,1]$ of the random walk such that
\[
\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n \lvert
h_i\rvert}.
\]
It is important to stress out that the role of vector $\mathbf{h}$ is
restricted to the construction of the initial probability. Once that is done,
it is
not used anymore in the definition of the stochastic process.
A possible choice for the transition probability $P$ can be
\[
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert H_{i,j}\rvert}{\sum_{k=1}^n
\lvert H_{i,k}\rvert}.
\]
where $\tilde{p}(\cdot,i):S\rightarrow [0,1]$ $\forall i\in S$.
A related sequence of random variables $w_{i,j}$ can be defined
such that
\[
w_{i,j}=\frac{H_{i,j}}{P_{i,j}}.
\]
The probability distribution of the random variables $w_{i,j}$ is represented
by the transition matrix that rules the stochastic process. The $w_{i,j}$'s
just introduced can be used to build one more sequence
of random variables.
At first we introduce quantities $W_j$
\[
W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_j=W_{j-1} w_{i,j}, \quad
j=1,\cdots, i.
\]
By defining
\[
X(\nu)=\sum_{m=0}^k W_m f_{i_m}
\]
as the random variable associated with a specific permutation $\nu$, we can
define the estimator $\theta_i$ such as
\[
\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu).
\]
The integer $n$ represents the size of the solution vector to (\ref{linsys})
and
the
index
$i$
is referred to the component of the solution vector we want to compute.
$P_{\nu}$ is the probability associated with a specific permutation of the
random walk.
It can be proved that
\[
E[W_i f_{k_i}]=(\mathbf{h},H^i\mathbf{f}), \quad i=0,1,2,\cdots
\]
and
\[
\theta_i=E\bigg[\sum_{i=0}^\infty W_i f_{k_i}\bigg]=(\mathbf{h},\mathbf{x}).
\]

A possible choice for $\mathbf{h}$ is a vector of the standard basis. This
would correspond in setting manually the initial state of the random walk,
which turns the related initial probability into a Kronecker delta
\[
\tilde{p}(k_0=i)=\delta_{i,j}.
\]
By doing so, we have
\begin{equation}
\theta_i=E\bigg[\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]=x_i=\sum_{l=0}^\infty
\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n
P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},
k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots
w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{dir_mean}
\end{equation}

As regards the variance, we remember that the following relation holds:
\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_{\ell}}^2\bigg] - \bigg (E\bigg[\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]\bigg )^2.
\label{dir_var}
\end{equation}

Hence the variance can be computed as the difference between the second
moment of the random variable and the square of the first moment.\newline

In order to apply the Central Limit Theorem (CLT) to the estimators defined
above, we must require that
the estimators have both finite expected value and variance. This is
equivalent in
checking the finiteness of the expected value and of the second moment.
Therefore we have to impose the following conditions:

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell} f_{k_{\ell}}\bigg]<\infty
\end{equation}

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_{\ell}}^2\bigg]<\infty
\end{equation}

The Forward method presented above, however, has the limitation of employing an
entire set of permutations to estimate just a single entry of
the solution at a time. Therefore, in order to estimate the entire solution
vector for \ref{linsys_prec}, we have to employ as many sets of permutation as
the number of entries composing the vector. Of course this might turn to be
extremely expensive both in terms of time and resources for most of the
challenging problems coming from applications. This is essentially the reason
why usually it is preferred to adopt another stochastic algorithm which is
explained here following: the \textit{Adjoint} Monte Carlo method.

\subsection{Adjoint Method}

A second Monte Carlo method can be derived by considering the linear system
adjoint to (\ref{linsys_prec})
\begin{equation}
(P^{-1}A)^T\mathbf{y}=\mathbf{d},
\label{adj}
\end{equation}
where $\mathbf{y}$ and $\mathbf{d}$ are the adjoint solution and source term.

By reformulating the fixed point scheme, introducing initial probability,
transition probability and weights
sequences in the same fashion as done before, the expected value for the
estimator becomes
\begin{equation}
\theta_j=E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell},
j}\bigg]=\sum_{\ell=0}^{\infty}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}.
\label{adj_mean}
\end{equation}
This estimator is known in literature as \textit{collision} estimator.

We can notice now that the component of the source vector, in each
permutation of the random walk, is associated with the initial step. The
Kronecker delta
at the end of the series stands for a filter. It means that if the goal is to
estimate the $j$-th component of the solution vector, just a subset of all the
permutations gives a contribution to this. The only permutations considered are
those ones that currently reside on the index associated with the entry to be
estimated.

The variance is

\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
f_{k_0}\delta_{k_{\ell},j}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_0}^2\delta_{k_{\ell},j}\bigg ] - \bigg (E\bigg[\sum_{\ell=0}^\infty
W_{\ell}
f_{k_0}\delta_{k_{\ell},j}\bigg]\bigg )^2\quad j=1,\cdots,n
\label{adj_var}.
\end{equation}

On the same lead of what done for the Forward method, there is need to impose
finiteness of the expected value and second moment. Therefore the following
conditions must be verified:

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell},
j}\bigg]<\infty \quad j=1,\cdots,n
\end{equation}
and
\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_0}^2\delta_{k_{\ell},j}\bigg]<\infty \quad j=1,\cdots,n.
\end{equation}

The main advantage of this method, compared to the Forward one, consists in the
fact that a single set of permutation is used to estimate the entire solution
vector. Therefore, in general, this algorithm has to be preferred to the
previous one since it enables to reduce dramatically the computational burden.

In literature another estimator is employed along with the Adjoint Monte Carlo
method, the so called \textit{expected value} estimator. Its
formulations is as follows:

\begin{equation}
\theta_j=E\bigg[f_j + \sum_{\ell=0}^\infty
W_{\ell}H_{k_{\ell}, j}^T\bigg]=f_j
+ \sum_{\ell=0}^{\infty}\sum_{k_1}^n\sum_{k_2} ^n\cdots\sum_ { k_ { \ell}}^n
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
w_{k_{\ell-1},k_{\ell}}H_{k_{\ell},j}^T.
\label{adj_mean1}
\end{equation}

Differently from the collision estimator, the expected value estimator averages
the deterministic contribution of the iteration matrix over all the potential
states $j$ that might be reached from the current state $\ell$. The variance
in this case becomes:

\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
f_{k_0}H_{k_{\ell},j}^T\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_0}^2 {H_{k_{\ell},j}^T}^T\bigg ] - \bigg (E\bigg[\sum_{\ell=0}^\infty
W_{\ell}
f_{k_0}H_{k_{\ell},j}^T\bigg]\bigg )^2\quad j=1,\cdots,n
\label{adj_var}.
\end{equation}

In the sections dedicated to the numerical tests, both the collision estimator
and the expected value are employed, in order to test their effectiveness.

\subsection{Hybrid Stochastic/Deterministic Methods}

Even if the standard Monte Carlo, in both its Forward and Adjoint formulation,
has the advantage of being embarrassingly parallelizable (because of the
independence of the samplings employed), it has a drawback as
well. In fact its slow rate of convergence, associated with the
$\frac{1}{\sqrt{N}}$ behavior predicted but the CLT, prevents its use for
problems of actual interest in applied linear algebra (with many d.o.f's).

Moreover, when the spectral radius of the iteration matrix is close to the
unit
value, Monte Carlo direct methods reach convergence after an excessive number
of steps for each
permutation. In particular this aspect will be pointed out more by some
numerical tests analyzed in the upcoming sections of this work.
This motivation makes the standard Mote Carlo paradigm even more
unpractical for solving linear systems associated with realistic applications.
\newline

To cope with this, in the early 2000's Evans et al. introduced a new method:
the
\textit{Monte Carlo Synthetic Acceleration (MCSA)}
(\cite{ESW2013} and \cite{EMSH2014}). Earlier contributions on these direction
are attributed also to other scientists (e.g. \textit{Sequential Method} by
J.H. Halton). However the MCSA is the algorithm we are going to focus on more
hereafter. If
we think about a fixed point formulation of the problem
such as $\mathbf{x}=H\mathbf{x}+\mathbf{b}$, then the Monte Carlo Synthetic
Acceleration assumes this form

\begin{algorithm}[H]
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$x_{num}$}
 $\mathbf{x}^{l}=\mathbf{x}_0$\;
 \While{not reached convergence}{
  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
  $\delta \mathbf{x}^{l+\frac{1}{2}}=(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$\;
  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta
\mathbf{x}^{l+\frac{1}{2}}$\;
 }
 $x_{num}=x^{l+1}$\;
 \caption{Monte Carlo Synthetic Acceleration}
\end{algorithm}
The Monte Carlo method is used to compute the updating contribution $\delta
\mathbf{x}^{l+\frac{1}{2}}$.

The role of the first line of the iterative process has the goal to compress
the residual of the previous iteration. This is the actual difference
between the MCSA and the Sequential Monte Carlo. Once this done, the corrected
residual
is employed to compute an update $\delta x^{1+\frac{1}{2}}$ to the solution via
one of the Monte Carlo techniques introduced before. The utility of compressing
the residual at the very beginning of the iteration will be shown in the two
last test cases for numerical validations.
The intent of MCSA is to designate most of the time employed by the
computation to the accomplishment of the stochastic part of the algorithm.
This
is useful to decrease the probability of fault occurrence in the non-resilient
steps, which is represented by the non-stochastic component of the numerical
scheme presented above. In particular, the higher is the percentage of time
spent on the computation of $\delta x^{1+\frac{1}{2}}$ via Monte Carlo, the
better it should be in terms of resilience.

\subsubsection{Sequential Monte Carlo}

\subsubsection{Monte Carlo Synthetic Acceleration}

### 2015-09-16 MAX_reports/MC_solvers_article/results.tex
Breaking paper into multiple files for easier editing/collaborating.

\section{Numerical Results}

\subsection{Adaptive Methods}

In this subsection we show some numerical tests to validate the utility of the
adaptive approaches for the selection of the number of permutations to employ.
For this purpose, we restrict our attention to standard Monte Carlo linear
solvers.

At first we take into account the Forward Monte Carlo linear solver.
In particular, the adaptive threshold associated with formulas
(\ref{forward_adapt}) and
(\ref{adjoint_adapt}) is set to $\varepsilon_1=0.1$ and to
$\varepsilon_1=0.01$. \newline

A set of matrices associated with problems of various nature has been collected.
All the problems treated have a small number of d.o.f's to control the
massive computational cost of MC solvers on a standard laptop. All the analyzed
matrices are derived from the
discretization of other differential problems.
One of these matrices is produced by the finite volume discretization of
a thermal equation (Marshak problem), another one is a 1D diffusion reaction
problem discretized with finite differences, a
2D laplacian discretized with finite differences and an advection
diffusion
problem discretized with quadrilateral linear finite elements (the
\texttt{ifiss} package is used in this case).
For all the problems at hand a left diagonal preconditioning is applied.
Details about all these matrices are gathered in Table
\ref{table_data}.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{d.o.f.'s} &
$\rho(H)$&Forward $\rho(\hat{H})$&Adjoint $\rho(\hat{H})$\\
\hline
1d shifted Laplacian & 50 & $0.4991$ & $0.9827$ & $0.9827$\\
\hline
2d Laplacian & 900 & $0.9949$ & $0.994$ & $0.9945$\\
\hline
ifiss & 1089 &$0.9836$ & $0.983$ & $0.983$ \\
\hline
Marshak Equation & 1600 & $0.6009$ & $0.3758$ & $ 0.3815$ \\
\hline
\end{tabular}
\caption{Properties of the matrices employed for the checks on adaptivity.}
\label{table_data}
\end{table}

For all the numerical tests results are presented both by using the Forward
and the Adjoint Monte Carlo.

As concerns the Forward method, a safety threshold for maximal number of
histories per
entry is set to $10^7$. Results are presented for
$\varepsilon_1=0.1$ and $\varepsilon_1=0.01$ in Tables \ref{tab:For_adapt} and
\ref{tab:For_adapt2} respectively. A batch size of two is used at each
adaptive check to verify the magnitude of the apparent relative standard
deviation. As expected, results are aligned with the convergence rate predicted
by the Central Limit Theorem. Indeed, decreasing of a factor of 10 the
tolerance $\varepsilon_1$, the relative error decreases of the same order,
employing in average one hundred times more histories.\newline

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.003$ & $5,220$\\
\hline
2d Laplacian & $0.1051$ & $262,350$\\
\hline
ifiss advection diffusion & $0.022$  & $1,060,770$\\
\hline
Marshak equation & $0.0012$ & $1,456,000$\\
\hline
\end{tabular}
\caption{Forward Monte Carlo. Adaptive selection of histories.
$\varepsilon_1=0.1$}
\label{tab:For_adapt}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $4\cdot 10^{-4}$ & $512,094$\\
\hline
2d Laplacian & $0.0032$ & $108,551,850$\\
\hline
ifiss advection diffusion & $0.0023$  & $105,476,650$\\
\hline
Marshak equation & $5.8 \cdot 10^{-4}$ & $144,018,700$\\
\hline
\end{tabular}
\caption{Forward Monte Carlo. Adaptive selection of histories.
$\varepsilon_1=0.01$}
\label{tab:For_adapt2}
\end{table}

As regards the Adjoint Monte Carlo, at each adaptive check the number of
random walks employed is increased by ten.

A maximal number of histories equal to $10^10$ is employed as a safety
threshold.

Here below Tables \ref{tab:Adj_adapt}, \ref{tab:Adj_adapt2} and
\ref{tab:Adj_adapt3} show results
for the different test cases
considered using $\varepsilon_1=0.1$, $\varepsilon_1=0.01$ and
$\varepsilon_1=10^{-3}$ respectively.
By comparing the two tables, it is noticed that a decrease of the threshold
induces a reduction of the relative error of the same order of magnitude.
Therefore this validates
the effectiveness of the adaptive selection of histories with an error
reduction goal. The total number of histories employed increases on average by
two orders of magnitude and this aspect is in line too with what expected by
theoretical results.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.08$ & $1820$\\
\hline
2d Laplacian & $0.136$ & $570$\\
\hline
ifiss advection diffusion & $0.08$  & $2400$\\
\hline
Marshak equation & $0.288$ & $880$\\
\hline
\end{tabular}
\caption{Adjoint Monte Carlo. Adaptive selection of histories.
$\varepsilon_1=0.1$}
\label{tab:Adj_adapt}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.0082$ & $185,400$\\
\hline
2d Laplacian & $0.0122$ & $126,800$\\
\hline
ifiss advection diffusion & $0.0093$  & $219,293$\\
\hline
Marshak equation & $0.0105$ & $650,040$\\
\hline
\end{tabular}
\caption{Adjoint Monte Carlo. Adaptive selection of histories.
$\varepsilon_1=0.01$}
\label{tab:Adj_adapt2}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $9.56 \cdot 10^{-4}$ & 15,268,560 \\
\hline
2d Laplacian & $0.001$ & $12,600,000$\\
\hline
ifiss advection diffusion & $0.0011$  & $23,952,000$\\
\hline
Marshak equation & $0.0011$ &  $80,236,000$\\
\hline
\end{tabular}
\caption{Adjoint Monte Carlo. Adaptive selection of histories.
$\varepsilon_1=10^{-3}$}
\label{tab:Adj_adapt3}
\end{table}

The same results can be reproduced by resorting to the expected value
estimator. Results are shown in the Tables \ref{tab:Adj_adapt4},
\ref{tab:Adj_adapt5} and \ref{tab:Adj_adapt6} for the threshold values of
$\varepsilon_1=0.1$, $\varepsilon_1=0.01$ and $\varepsilon_1=10^{-3}$
respectively. As it can be noticed, in terms of error scaling the results are
pretty similar to the ones obtained with the collision estimator. As regards
the number of histories as well, used to reach a prescribed accuracy, the
orders of magnitude are the same for both the collision estimator and the
expected value one. However the expected value estimator requires most of the
time a smaller number permutations. This behavior is enhances as the value of
the threshold decreases, making the computation heavier and heavier.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.0463$ & $1000$\\
\hline
2d Laplacian & $0.1004$ & $900$\\
\hline
ifiss advection diffusion & $0.0661$  & $1300$\\
\hline
Marshak equation & $0.0526$ & $2000$\\
\hline
\end{tabular}
\caption{Expected value estimator - Adjoint Monte Carlo. Adaptive selection of
histories.
$\varepsilon_1=0.1$}
\label{tab:Adj_adapt4}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.004$ & $100,600$\\
\hline
2d Laplacian & $0.0094$ & $83,700$\\
\hline
ifiss advection diffusion & $0.0088$  & $124,400$\\
\hline
Marshak equation & $0.0056$ & $166,000$\\
\hline
\end{tabular}
\caption{Expected value - Adjoint Monte Carlo. Adaptive selection of histories.
$\varepsilon_1=0.01$}
\label{tab:Adj_adapt5}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1d shifted Laplacian & $0.004$ & $10,063,300$\\
\hline
2d Laplacian & $ 9.31 \cdot 10^{-4} $ & $ 8,377,500 $\\
\hline
ifiss advection diffusion & $ 0.0013 $  & $ 12,435,000 $\\
\hline
Marshak equation & $7.79 \cdot 10^{-4}$ & $16,537,000$\\
\hline
\end{tabular}
\caption{Expected value - Adjoint Monte Carlo. Adaptive selection of histories.
$\varepsilon_1=10^{-3}$}
\label{tab:Adj_adapt6}
\end{table}

\subsection{Preconditioning Approaches}

This subsection is focused on showing the effect of different preconditioners
for the values attained by the spectral radii $\rho(H)$ and $\rho(\hat{H})$.
For this purpose, we consider the two dimensional laplacian
discretized with centered finite differences and the "ifiss advection
diffusion" matrix. The properties of both the matrices have already been
explained in previous sections.
Results in terms of spectral radii when the diagonal preconditioning is applied
have been shown earlier as well.

Here the analysis is focused on the evaluating the effects of the
block diagonal preconditioner and the factorized approximate sparse inverse.
As concerns the matrix $\hat{H}$, it has been built accordingly to te
procedure for the Adjoint Monte Carlo method. In
the first case the size of the blocks is a tuning parameter, whilst in the
second case the tuning parameter is the drop tolerance to manage the fill-in
effect of the factors.

As regards the employment for the block diagonal preconditiner, in Tables
\ref{tab:lap2d_block} and \ref{tab:ifiss_block} there is a counterintuitive
common trend. Indeed in both the cases an increase of $\rho(H)$ and
$\rho(\hat{H})$  occurs as the size of the blocks diminishes.

In Tables \ref{tab:lap2d_ainv} and \ref{tab:ifiss_ainv} the values of
the spectral radii are shown for the case of the factorized AINV
preconditioner. It is interesting to point out that for the
two-dimensional
laplacian, a drop tolerance $\tau=0.05$ entails $\rho(H)<1$ but the same does
not hold for $\hat{H}$. Both the convergence conditions are restored by
decreasing even more the drop tolerance, causing a massive fill-in effect
though.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
&$\rho(\hat{H})$\\
\hline
 5 & 2.6164 & $0.9915$ & $0.9837$\\
\hline
 10 & 5.6027 & $0.9907$ & $0.9861$\\
\hline
 30 & 16.3356 & $0.9898$  & $0.9886$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the two dimensional
laplacian with block diagonal preconditioning.}
\label{tab:lap2d_block}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
&$\rho(\hat{H})$\\
\hline
 3 & 1.4352 & $0.9804$ & $0.9531$\\
\hline
 9 & 3.0220 & $0.9790$ & $0.9674$\\
\hline
 33 & 9.8164 & $0.9783$  & $0.9774$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the ifiss advection
diffusion matrix with block diagonal preconditioning.}
\label{tab:ifiss_block}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
&$\rho(\hat{H})$\\
\hline
 0.05 & 8.2797 & $0.9610$ & $1.0231$\\
\hline
 0.01 & 33.1390 & $0.8668$ & $0.8279$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the two dimensional
laplacian with factorized AINV preconditioning.}
\label{tab:lap2d_ainv}
\end{table}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
&$\rho(\hat{H})$\\
\hline
 0.05 & 3.8392 & $0.9396$ & $0.9069$\\
\hline
 0.01 & 15.1798 & $0.7964$ & $0.6361$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the ifiss advection
diffusion matrix with factorized AINV preconditioning.}
\label{tab:ifiss_ainv}
\end{table}

\subsection{Hybrid Methods}

\subsubsection{Poisson Problem}

In the current subsection a Poisson problem is taken into account:
\begin{equation}
\begin{cases}
 -\Delta u = f \quad \text{in}\quad \Omega \\
 u=0\quad \text{on} \quad \partial\Omega
 \end{cases}
\end{equation}

where $\Omega=(0,1)\times (0,1)$, $\sigma=1$.
A finite difference scheme is employed to discretize the problem.
As a right hand side, a
sinusoidal distribution in both $x$ and $y$ directions is considered, so that
\[
 f(x_i, y_i)=sin\bigg(\frac{\pi x}{N-1}\bigg)sin\bigg(\frac{\pi y}{N-1}\bigg).
\]
N is the number of node considered on each direction, in our case $N=32$. The
resulting linear system has 900 d.o.f.'s and $h=\frac{1}{29}$.
By applying a left diagonal preconditioning on the matrix, it is possible to
reformulate the problem as a fixed point scheme.
The iteration matrix $H$ is characterized by the fact that $\lVert
H\rVert_1=1$, which is not sufficient to guarantee the convergence of
any
stochastic scheme analyzed in this work. However, by theoretical results, it is
well known that applying a left diagonal preconditioning to the problem at hand
produces an iteration matrix $H$ such that
\[
 \lim_{h\rightarrow 0}\rho(H)=1^{-}
\]
where $h$ is the spatial discretization step.
This is enough to have the guarantee that the deterministic fixed point scheme
converges without the explicit computation of the spectral radius $\rho(H)$.
However, the smaller is the discretization step $h$, the higher is the number of
iterations required to reach a prescribed convergence criterion. For our
discretization setting, the iteration matrix $H$ has a
spectral radius of $\rho(H)\approx 0.9949$.

In order to have the guarantee of the convergence of the Adjoint Monte Carlo
method, it is necessary to look at the spectral radius of $\hat{H}$ too. If an
almost optimal probability is used, this implies that
the Adjoint Monte Carlo method has a $\hat{H}$ matrix such that
\[
 \lVert \hat{H}\rVert_1\le (\rVert H\rVert_1)^2=1.
\]
This condition by itself is not enough to guarantee that
$\rho(\hat{H})<1$.

However, it must be recalled that the iteration matrix $H$ has zero
valued entries on the main diagonal and it has:
\begin{itemize}
 \item four entries equal to $\frac{-1}{4}$ on the rows associated
with a node which is not adjacent to the boundary
\item two entries equal to $\frac{-1}{4}$ on the rows associated with nodes
adjacent to the corner of the square domain
\item three entries equal to $\frac{-1}{4}$ on the rows associated with nodes
adjacent to the boundary on an edge of the square domain.
\end{itemize}

Recalling the definition of $\hat{H}$ entrywise in terms of the
iteration matrix $H$ and the transition probability $P$, we get that

\[
 \hat{H}=\tilde{D}H
\]

where $\tilde{D}$ is a diagonal matrix with diagonal entries equal to
1, $\frac{1}{2}$ or $\frac{3}{4}$.
In particular $\tilde{d}_i=diag(\tilde D)_i=1$ if the $i$-th row of the
discretized laplacian is associated with a node which is not adjacent to the
boundary, $\tilde{d}_i=diag(\tilde D)_i=\frac{1}{2}$ if the row is associated
with a node of the grid adjacent to the corner of the boundary,
$\tilde{d}_i=diag(\tilde D)_i=\frac{3}{4}$ if the associated node of the grid
is adjacent to the boundary of the domain far from a corner.

Since $H$ is an irreducible nonnegative matrix and $\tilde{D}$ is a
positive definite diagonal matrix, we resort to a result presented in
\cite{FK1975} to say that
\[
 \rho(\hat{H})=\rho(\tilde{D}H)\le \rho(\tilde D)\rho(H).
\]

Since $\rho(\tilde{D})=1$, then $\rho(\hat{H})\le \rho(H)<1$.
Therefore the diagonal preconditioning always guarantees a convergent
setting for the Monte Carlo linear solver, once it is applied on a
two-dimensional Laplace
operator discretized with finite differences. By a similar reasoning, the same
results can be obtained for a $n$-dimensional laplace operator.

The computation of the solution is realized both with the deterministic
Synthetic Acceleration and with Adjoint MCSA to compare the effectiveness of
the
stochastic technique. Results are reported in Table
\ref{Poisson_results}. For the sake of completeness, results obtained with
the Sequential Monte Carlo are shown as well. An evident discrepancy in terms
of numerical iterations
is already detected with the Sequential Monte Carlo outperforming the
preconditioned fixed point. Results are improved even more
when
switching to the MCSA, which employs less histories than the Sequential MC at
each iteration to
reach convergence. In fact the average number of histories employed by the
Sequential
Monte Carlo per iteration is almost one order of magnitude higher than for
MCSA. This
might be explained by the fact that the first row of the iterative algorithm in
the MCSA refines the accuracy of the solution coming from the previous
iteration,
before using it as the input for the MC solver at the current iterate.
In particular, since the first line of the algorithm is a deterministic step of
the Synthetic Acceleration, its effect is to dampen the statistical noise
produced by the Monte Carlo linear solver to estimate the update $\delta
x^{n + \frac{1}{2}}$ at the previous iteration. Therefore
refinement provided by the Synthetic Acceleration step decreases the number of
random walks
to employ for a prescribed accuracy. This is validated even more by the fact
that for the first iteration the Sequential Monte Carlo and the MCSA resort to
the same number of histories, while their behaviours start differing from the
second iteration on, when the statistical noise is introduced in the estimation
of the solution to \ref{linsys_prec} (see Figures \ref{SEQ_poisson} and
\ref{MCSA_poisson}).

\begin{table}[!h]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
Synthetic Acceleration & $9.8081\cdot 10^{-8}$ & 3133 & - \\
\hline
Sequential MC & $7.9037 \cdot 10^{-8}$ & 9 & 8.264,900\\
\hline
 Adjoint MCSA & $8.0872\cdot 10^{-8}$ & 8 & 1,738,250\\
\hline
\end{tabular}
\caption{Numerical results for the Poisson problem.}
\label{Poisson_results}
\end{table}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{SEQ_poisson.jpg}
    \caption{Sequential MC - Poisson problem. Number of random walks employed
at each numerical iteration.}
\label{SEQ_poisson}
\end{figure}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{MCSA_poisson.jpg}
      \caption{MCSA - Poisson problem. Number of random walks employed at each
numerical iteration.}
\label{MCSA_poisson}
\end{figure}

\subsubsection{Reaction-Diffusion Problem}

By modifying the equation of the previous test case, we consider now
an advection reaction problem

\begin{equation}
\begin{cases}
 -\Delta u +\sigma u= f \quad \text{in}\quad \Omega \\
 u=0\quad \text{on} \quad \partial\Omega
 \end{cases}
\end{equation}
where $\Omega=(0,1)\times (0,1)$, $\sigma=0.1$ and $f=1$.
A finite difference scheme is applied to discretize the problem.
The number of nodes
on each direction of the domain is 100, so that $h\approx 0.01$. The
discretized problem has 9604 d.o.f's. A left
diagonal preconditioning is still applied to
the stiffness matrix obtained from the discretization. The 1-norm of the
iteration matrix is $\lVert H\rVert_1\approx 0.9756$. This automatically
guarantees the convergence of the Adjoint Monte Carlo linear solver. The
computation of the solution to the partial differential problem is still
accomplished with the MCSA algorithm. A threshold of $\varepsilon =10^{-8}$ is
used as a stop criterion for the relative residual. The threshold
for the adaptive selection of the random walks instead is set
to $\varepsilon_1=0.1$.
In Table \ref{DR_results} a comparison between the deterministic Synthetic
Acceleration
scheme, Sequential MC and the MCSA is provided.
The employment of the MCSA provides a significant reduction of the number of
histories necessary to satisfy the convergence criterion, in line with results
coming from the previous test case.

\begin{table}[!h]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
 Synthetic acceleration & $9.073\cdot 10^{-8}$ & 634 & - \\
\hline
 Sequential MC & $8.415 \cdot 10^{-8}$ &  8 & 12,391,375\\
 \hline
 Adjoint MCSA & $6.633 \cdot 10^{-8}$ &  7 & 3,163,700\\
\hline
\end{tabular}
\caption{Numerical results for the diffusion reaction problem.}
\label{DR_results}
\end{table}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{SEQ_diffreac.jpg}
      \caption{Sequential MC - Diffusion reaction problem. Number of random
walks
employed at each
numerical iteration.}
\label{SEQ_diffreac}
\end{figure}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{MCSA_diffreac.jpg}
      \caption{MCSA - Diffusion reaction problem. Number of random walks
employed at each
numerical iteration.}
\label{MCSA_diffreac}
\end{figure}

\subsubsection{Parabolic Problem}

In this subsection we restrict our treatise to the analysis of parabolic
partial
differential problems of the form
\begin{equation}
\begin{cases}
 \frac{\partial u }{\partial t} + \mathcal{L}(u)=f \quad \text{in} \Omega \\
\text{B.C.} \quad \text{on}\quad \delta \Omega.
\end{cases}
\label{parabolic}
\end{equation}

The partial derivative in time is discretized with finite differences, whereas
quadrilateral linear finite elements are employed for the space discretization.
Naming $N_h$ the number of d.o.f.'s associated with the finite element
discretization, the problem \ref{parabolic} turn into a fully discretized
problem such as:

\[
 \bigg (\frac{1}{\Delta t}M+A\bigg )(\mathbf{u}^{k+1}-\mathbf{u}^k)+A[\theta
\mathbf{u}^{k+1}+(1-\theta)\mathbf{u}^k]=\theta
\mathbf{f}^{k+1}+(1-\theta)\mathbf{f}^k,
\]

where $k$ is an index that goes with the time dimension.

By tuning the value of the parameter $\theta$ different numerical schemes can
be obtained with different properties of numerical stability.
However our discussion is not involved in this kind of issues.
In particular we restrict our attention to a single generic time step
associated with an Implicit Euler time discretization (corresponding to
$\theta=1$). The vector of the right hand side is chosen so that the exact
solution to the linear system for the specific time step chosen is a unit
vector $\tilde{\mathbf{u}}=\mathbbm{1}^{N_h}$.

By referring to $h$ as the space discretization step, we pick
\[
 \Delta t \le C h.
\]

Give an open and bounded set $\Omega=(0,1)\times(0,1)$, we consider the
following problem
\begin{equation}
 \begin{cases}\frac{\partial u}{\partial t} -\mu \Delta u
+\boldsymbol{\beta}(\mathbf{x})\cdot \nabla u=0, \;\; \quad \quad
\mathbf{x}\in \Omega,\quad t\in(0,T] \\
u(\mathbf{x},0) = u_0, \qquad \qquad \qquad \qquad \quad \mathbf{x}\in
\Omega \\
u(\mathbf{x},t)=u_D(\mathbf{x}),  \; \, \, \qquad \qquad \qquad \quad
\mathbf{x}\in \partial
\Omega, \quad t\in(0,T],
 \end{cases}
\end{equation}

where $\mu=\frac{3}{200}$, $\boldsymbol{\beta}(\mathbf{x})=[2y(1-x^2),\;
-2x(1-y^2)]^T$, \\$u_D=0$ on $\{\{x=0\}\times (0,1)\}$, $\{(0,1)\times
\{y=0\}\}$, $\{(0,1)\times \{y=1\}\}$.\newline

The discretization of the problem is realized with the \texttt{IFISS} toolbox.
The value of the discretization step is $h=2^{-8}$, generating a discretized
linear system with 66,049 d.o.f.'s.
As concerns the time discretization, the time step is set to $\Delta t =
10h$.\newline

The sparse approximate inverse is employed for
the linear system at hand as a right preconditioner, with a drop tolerance of
$\tau=0.05$ for both the factors.

In this situation it is not possible to
resort to any of the sufficient conditions to verify the convergence. Therefore
the only viable option is the explicit computation of the spectral radii of $H$
and $\hat{H}$.
The spectral radius of the iteration matrix is $\rho(H)\approx
0.9218$ and the spectral radius of $\hat{H}$ for the Adjoint Monte Carlo is
$\rho(\hat{H})\approx 0.9148$. The transition probability employed is the
Almost
Optimal one. Resorting to a Uniform Probability in this case would have impeded
the
convergence, since in this case $\rho(\hat{H})\approx 1.8401$.
Therefore
this is an example demonstrating that the Almost Optimal Probability
facilitates the performance of the stochastic algorithm, outperforming the
Uniform one.

The fill-in effect is
\[
 \frac{nnz(H)}{nnz(A)}=4.26,
\]
therefore the relative number of nonzero elements in $H$ is still acceptable
in terms of sparsity pattern and memory storage.

The threshold for the check on the relative residual is set to $\varepsilon
=10^{-8}$ is, whilst the threshold
for the adaptive selection of the random walks is set
to $\varepsilon_1=0.1$.
In order to test the effectiveness of employing the MCSA with respect to a
purely deterministic scheme, a comparison with the Richardson algorithm and
the Sequential Monte Carlo is
accomplished. The results are shown in Table \ref{parabolic_results}. As you
can see, the adoption of both Sequential Monte Carlo and MCSA reduces
dramatically the number of iterations with
respect to the fixed point method. Of course this
advantage has the
increase of the computational time as a payoff. Nevertheless we put off this
discussion which is of
interest for a parallelized implementation of the method. Even more the number
of random walks employed by the Sequential MC is higher than the one associated
with the MCSA, in accordance with the previous results.

\begin{table}[!h]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
 Richardson & $6.277\cdot 10^{-7}$ & 178 & - \\
 \hline
 Sequential MC & $1.918 \cdot 10^{-9}$ & 8 & 51,535,375\\
\hline
 Adjoint MCSA & $1.341\cdot 10^{-9}$ & 6 & 16,244,000\\
\hline
\end{tabular}
\caption{Numerical results for the parabolic problem.}
\label{parabolic_results}
\end{table}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{SEQ_parabolic.jpg}
      \caption{Sequential Monte Carlo - Parabolic problem. Number of random
walks
employed at each
numerical iteration.}
\label{MCSA_parabolic}
\end{figure}

\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{MCSA_parabolic.jpg}
      \caption{MCSA - Parabolic problem. Number of random walks
employed at each
numerical iteration.}
\label{MCSA_parabolic}
\end{figure}

### 2015-09-16 MAX_reports/MC_solvers_article/intro.tex
Editing work on mcls paper.

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 therefore the provided error with respect to the exact solution is made of
-just a stochastic component. The second ones, instead, combine numerical and
-statistical schemes generating two types of error: a
+just a stochastic component. The iterative Monte Carlo methods utilize more
+traditional iterative algorithms alongside the stochastic approach,
+generating two types of error: a
 \textit{stochastic} one and a \textit{systematic} one. It does not
@@
 aware of this intrinsic structure, in order to target what is the part of the
-scheme that requires a refinement (e.g. increase of numerical iterations
-rather than random walks).\newline
+scheme that requires a refinement (e.g. increasing the number of iterations
+rather than the number of random walks).
 
-The paper is organized as follows. In Section 2 we describe all the
-viable stochastic settings for the definition of the Markovian process. In
-Section 3 we reinterpret probabilistic facts in terms of spectral properties of
-the linear system to be solved. Section 4 is dedicated to necessary conditions
-and sufficient conditions for the convergence of the algorithms. In Section 5
-different adaptive approaches are presented, in order to truncate the random
-walks and to decide how many samplings to use. Section 6 regards the study
-of matrices for which the convergence of the stochastic schemes is facilitated.
-Then in Section 7 a hybrid scheme combining numerical and probabilistic
-techniques is described: the \textit{Monte Carlo Synthetic Acceleration}.
-Numerical results are presented in Section 8 and 9, while Section 10 is
-dedicated to
-conclusive remarks and goals for the future.
+The paper is organized as follows.
+In Section~\ref{sec:mcls} we provide an overview of existing Monte Carlo
+linear solver algorithms.
+In Section~\ref{sec:convergence} we will discuss the convergence behavior
+of stochastic solvers, including a discussion of classes of matrices for
+which convergence can be guaranteed.
+Section~\ref{sec:results} provides some numerical results illustrating
+pertinent properties of the various approaches and Section~\ref{sec:conclusion}
+will provide some concluding remarks and areas for future investigation.

### 2015-09-16 MAX_reports/MC_solvers_article/results.tex
Editing work on mcls paper.

@@
 \section{Numerical Results}
+\label{sec:results}
 
@@
 second iteration on, when the statistical noise is introduced in the estimation
-of the solution to \ref{linsys_prec} (see Figures \ref{SEQ_poisson} and
+of the solution to \ref{linsys} (see Figures \ref{SEQ_poisson} and
 \ref{MCSA_poisson}).

### 2016-01-25 MAX_reports/MC_solvers_article/convergence.tex

@@
 \section{Convergence Behavior of Stochastic Methods}
+\label{sec:convergence}
 
-The requests imposed on the expected value and second moment of the estimators
-can be reformulated in a purely deterministic setting by looking at the
-spectral radii of some properly defined matrices.
-
-For instance the condition of finiteness of the expected value can be
+The convergence requirements imposed by the Monte Carlo estimator and
+the corresponding variance can be reformulated in a purely deterministic
+setting by considering certain properly defined matrices.
+For instance, the condition of finiteness of the expected value can be
 reformulated by requiring
@@
 \end{equation}
-
 where $H$ is the iteration matrix of the fixed point scheme.
@@
 
-By analogy we aim at reproducing a similar reasoning for the second moment of
-the estimator, since we want it to be finite as well.
-By looking at the equations of the variance (\ref{dir_var}) and (\ref{adj_var})
+We wish to achieve a similar result based on the requirement that the
+second moment of the estimator must be finite as well.
+Examination of the equations of the variance (\ref{dir_var}) and (\ref{adj_var})
 for
-the Forward and the Adjoint method respectively, we notice that the second
+the forward and the adjoint method respectively, we notice that the second
 moment can be reinterpreted as a power series with respect to the matrices
 defined as follows:
-
 \[
@@
 \]
-
 or
-
 \[
@@
 \]
-
 Accordingly to the necessary and sufficient condition for the power series to
@@
 \end{equation}
-
 Condition (\ref{rho_h}) must be required for a generic fixed point scheme to
 reach convergence, whereas the extra condition (\ref{rho_star}) is typical of
-the stochastic schemes presented in this work. Having two conditions to
-verify makes the numerical treatment of the linear system via MC very
-difficult. We will discuss this in details later in the treatise. \newline
+the stochastic schemes presented in this work.
 
@@
 
-Here below some results presented in
-\cite{MASC2013} about necessary conditions and sufficient
-conditions for
-convergence are enunciated. In particular a suitable choice for constructing a
-transition probability is discussed to guarantee and speed-up convergence.
+Here we briefly reproduce some results introduced in \cite{MASC2013}
+concerning necessary conditions and sufficient conditions for convergence.
+In particular, a suitable choice for constructing a
+transition probability is discussed to guarantee convergence.
 
-We remember that the construction of the transition probability must satisfy
-some constraints such as
-
+The construction of the transition probability must satisfy
 \[
@@
   P_{i,j}\ge 0 \\
- \sum_{j=1}^N P_{i,j}=1
+ \sum_{j=1}^N P_{i,j}\le 1 \,.
 \end{cases}
 \]
-
-plus one more restriction which varies depending on the particular Monte Carlo
-approach chosen
-
+One additional constraint relates the sparsity pattern of $H$ to that
+of the transition probabilities:
 \begin{align*}
@@
 \end{align*}.
-
-The previous restrictions on the attainable values for the entries of the
+The previous constraints on the values for the entries of the
 transition probability are called \textit{transition conditions}.
 
-At first an introductory theoretical result is shown, necessary to motivate the
-choices made afterwards for the choice of the transition probability.
+An introductory theoretical result is provided to motivate the possible
+choices for the transition probability.
 
@@
 much
-as possible. \newline
+as possible.
 
@@
 \]
-which is associated with the Forward Method.
+which is associated with the forward method.
 
@@
 \]
-which is associated with the Adjoint Method, then we can state a
+which is associated with the adjoint method, then we can state a
 theorem specular to \ref{for_thm}.
@@
 
-In particular, relying on what guaranteed by Theorem \ref{lemma}, in both
-Theorems \ref{for_thm} and \ref{adj_thm} the thesis holds by picking the MAO
-transition probability.
-
-These theoretical results represent sufficient conditions for the convergence
-of the
-Forward and Adjoint Monte Carlo and they can be easily checked.
-However requiring $\lVert
-H\rVert_{\infty}<1$ or $\lVert H\rVert_1<1$ may be too demanding.
-Indeed, even if there exist some preconditioners able to guarantee this
-condition (e.g. approximate inverse preconditioners with a proper value of
-dropping
-tolerance), it may entail to compromise the sparsity of the
-iteration
-matrix.
-
-Thusly, in situations where the hypotheses of Theorems \ref{for_thm} and
-\ref{adj_thm} are too demanding to be verified, the necessary and sufficient
-condition
-on the
-spectral radius of $\hat{H}$ is the only criterion to take as reference.
-Nonetheless this constraint consists in
-computing the spectral radii of two matrices and the calculation
-might be very costly. Moreover the computation of the spectral radii is
-affected by resilience issues as well as
-a deterministic solver for the resolution of
-a linear system. Therefore the original difficulty is transferred without being
-actually tackled.
-
-\subsubsection{Forward MC}
-
-\subsubsection{Adjoint MC with Collision Estimator}
-
-\subsubsection{Adjoint MC with Expected-Value Estimator}
+These theoretical results represent sufficient (but not necessary) conditions
+for the convergence
+of the forward and adjoint Monte Carlo and can be easily verified.
+However, in many cases the conditions $\lVert H\rVert_{\infty}<1$ or
+$\lVert H\rVert_1<1$ may be restrictive.
 
@@
 
-The way the transition probability is defined is extremely important. In fact
-in some circumstances it might determine whether the stochastic scheme
-converges or not. Two very simple ways of defining it are listed here
-below. These are the approaches adopted typically in literature.
+The method for determining the transition probability has a significant impact
+on the properties of the resulting approach, and in many circumstances the
+choice can be the difference between convergence or divergence of the stochastic
+scheme.  Two primary approaches have been considered in the literature:
+\textit{uniform probabilities} and \textit{weighted probabilities}.
 
 \subsubsection{Uniform probabilities}
-With this approach transition matrix $P$ is such that all the non-zero elements
-in each row have equal probability of occurring.
+
+With this approach, the transition matrix $P$ is such that all the non-zero elements
+in each row have equal probability of occurring:
 \[
@@
 \subsubsection{Weighted probabilities}
+
 An alternative way of defining the transition matrix consists in attributing
-probabilities to all the entries of the transition probability accordingly
+probabilities to the entries of the transition probability according
 to the magnitude of the elements. For instance we may employ the following
@@
 \]
-
+and
 \[
@@
 H_{j,i}\rvert^p}{\sum_{k=1}^n
-\lvert H_{k,i}\rvert^p}
+\lvert H_{k,i}\rvert^p} \,,
 \]
-
 where $p\in \mathbb{N}$.
-This way of proceeding, for $p=1$, still basing on the work \cite{AADBTW2005},
-is called
+This selection, for $p=1$, has been termed
 \textit{Monte Carlo Almost Optimal} (MAO).
-The reason why the almost optimal probability is called this way is associated
-with a notice from Theorem \ref{lemma}. Indeed from
-the mentioned lemma we know that the quantity $\displaystyle
-\sum_{k=0}^N\frac{a_k^2}{p_k}$ is minimized when the probability vector is
+The reason for the ``almost optimal'' designation can been seem from
+Theorem~\ref{lemma}, as the quantity
+$\displaystyle \sum_{k=0}^N\frac{a_k^2}{p_k}$
+is minimized when the probability vector is
 defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
 a_k\rvert}$. This implies that the almost optimal probability minimizes the
-$\infty$-norm of $\hat{H}$ in the Forward method and the $1$-norm of $\hat{H}$
-in the
-Adjoint method. This is very important to be pointed out since these two norms
-are so far the best practical upper bounds for $\rho(\hat{H})$ that we can
-easily
-handle. In the Section dedicated to the numerical results it will be proved
-that the Almost Optimal Probability eases the convergence of the MC solver,
-even in cases when the employment of the Uniform Probability does not provide
+$\infty$-norm of $\hat{H}$ in the forward method and the $1$-norm of $\hat{H}$
+in the adjoint method. This is very important, since these two norms
+provide computable upper bounds for $\rho(\hat{H})$.
+In Section~\ref{sec:results}, it will be shown
+that the almost optimal probability enables convergence of the MC solver,
+even in cases when the employment of the uniform probability does not provide
 convergence at all.
@@
 \cite{MASC2013}), whereas the necessary and sufficient condition requires the
-computation of
-$\rho(H)$and $\rho(\hat{H})$, which is very expensive in general. Once a
-proper
-preconditioner is picked (e.g. Approximate Inverse), $\rho(H)<1$ may
-be always guaranteed, whilst $\rho(\hat{H})<1$ is more
-problematic to be verified.
-
-In order to bypass the cost of these checks, it is useful to pursue for sets of
-matrices for
-which the Monte Carlo linear solver always converges, maybe with the use of an
-ad-hoc preconditioner.
+computation of $\rho(H)$and $\rho(\hat{H})$, which is very expensive in general.
+Many instances of matrices (perhaps with appropriate preconditioning) such
+that $\rho(H)<1$ are available, but verifying $\rho(\hat{H})<1$ is generally
+more problematic.  We now with to identify classes of matrices for which
+the Monte Carlo linear solver is guaranteed to converge.
 
 \subsubsection{Strictly Diagonally Dominant}
+\label{sec:sdd}
 
- One of the these sets is
-represented by strictly diagonally dominant matrices.
+One of the these sets is represented by strictly diagonally dominant matrices.
 
@@
 
-Introducing a MAO transition probability for the Forward Method we get
+Introducing a MAO transition probability for the forward method we get
 \[
@@
 This automatically implies that $\rho(\hat{H})<=\lVert
-\hat{H}\rVert_{\infty}<1$. Thus the Forward Monte Carlo converges.
+\hat{H}\rVert_{\infty}<1$. Thus the forward Monte Carlo converges.
 However nothing a priori can be
-stated about the behavior of the Adjoint method.\newline
+stated about the behavior of the adjoint method.
 
@@
 In this case, by following a similar reasoning to the one made before, we
-conclude that the Adjoint Method convergences thanks to $\lVert
+conclude that the adjoint method convergences thanks to $\lVert
 \hat{H}\rVert_1<1$,
-whereas nothing can be said in advance as concerns the Forward method.
+whereas nothing can be said in advance as concerns the forward method.
 
 \subsubsection{Generalized Diagonally Dominant}
+\label{sec:gdd}
 
@@
 following
-
 \begin{defn}
@@
 splitting and, therefore, a convergent fixed point scheme.
-
 In particular Jacobi, Block Jacobi and Gauss Seidel  are
@@
 to guarantee $\rho(\hat{H})<1$. Because of this obstacle, other options must
-be considered.\newline
+be considered.
 
@@
 
-Here below we report the algorithm at hand\newline
+Here below we report the algorithm at hand
 
@@
 $a_{ji}=a_{ji}\cdot d_i$ with $a_{ji}=a_{ji}\cdot d_j$ we obtain the algorithm
-that turns a GDDM matrix into a s.d.d. by columns.\newline
+that turns a GDDM matrix into a s.d.d. by columns.
 
@@
 \subsubsection{Block Diagonally Dominant}
+\label{sec:bdd}
 
- In this section we wonder in which situations a block diagonal preconditioning
-might come up with a convergent Monte Carlo linear solver.
-By mimicking the computations we already showed previously, the iteration
+In the section we consider situations in which block diagonal preconditioning
+can result in a convergent Monte Carlo linear solver.
+By mirroring the computations we already showed previously, the iteration
 matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal
 preconditioning takes the form
-
 \[
@@
 
-For the sake of clarity, we notify that the symbol "$\%$" used in the following
+For the sake of clarity, we note that the symbol "$\%$" is used in the following
 subsections represents the modulus function between two integer numbers, while
-the floor function is represented with pretty standard symbol "$\lfloor
+the floor function is represented with the standard symbol "$\lfloor
 \cdot \rfloor$".
 
-\paragraph{Forward method}
+We first consider the forward method.
 By assuming that all $n_i$'s have the same value we may define
@@
 \]
-
 The MAO transition probability matrix becomes:
@@
 \]
-
 Consequently, the $\hat{H}$ matrix is defined such that
@@
 \]
-
 By computing the sum over a generic row of $\hat{H}$ we get:
@@
 \]
-
 If we focus on the norm $\lVert \hat{H}\rVert_{\infty}$, then the following
@@
  \end{equation}
-
 By defining a matrix $\tilde{H}\in \mathbb{R}^{p\times p}$ such that
@@
 \]
-
 We can use the $\tilde{H}$ matrix just defined in order to introduce a
-sufficient condition for the convergence of the Forward Monet Carlo method with
+sufficient condition for the convergence of the forward Monet Carlo method with
 a Block Diagonal preconditioning.
-
 \begin{equation}
@@
 
-\paragraph{Adjoint method}
-Analogously to the Forward method, if we define
+We now turn our attention to the adjoint method.
+Analogously to the forward method, if we define
 \[
@@
 \]
-we can formulate a sufficient condition for the convergence of the Adjoint
+we can formulate a sufficient condition for the convergence of the adjoint
 Monte
@@
 
-In formulas (\ref{dir_mean}) and (\ref{adj_mean}) the estimation of the
+In formulas (\ref{dir_mean}) and (\ref{adj_mean}), the estimation of the
 solution to
-the linear system (\ref{linsys_prec}) involves infinite sums. Of course this is
-not viable for
-computational issues, therefore it is necessary to resort to truncation of
-the sums through proper criteria. In details, the truncation implies the
-employment of a finite number of permutations and a finite number of
-steps for each one of the permutations computed. Therefore it is necessary to
-truncate at two different levels to decide both how many permutations to employ
-and how many steps each one of them should be kept running.
+the linear system (\ref{linsys}) involves infinite sums. Of course this is
+not viable in a computational setting,
+therefore it is necessary to truncate both the number of steps taken in
+a single random walk as well as the number of random walks that
+are performed.
 
@@
 
-At first we focus on the definition of a weight cut-off
-in order to decide where to terminate the histories. It may be reinterpreted as
-a quantity guiding us to decide how many terms of the Neumann
-series to consider. \newline
+We first consider criteria to terminate an individual random walk, effectively
+deciding how many terms of the Neumann series will be considered.
+One possibility is to set a predetermined history length at which point all
+histories are terminated.  The approach, however, presents two difficulties.
+First, it is difficult to determine a priori how many steps on average will be
+necessary to achieve a specified tolerance.  Second, due to the stochastic
+nature of the random walks, some histories will retain important information
+longer than others.  Truncating histories at a predetermined step runs the
+risk of either prematurely truncating important histories, leading to larger
+errors, or continuing unimportant histories longer than necessary,
+leading to computational inefficiency.
 
-The goal is to apply the cutoff via an automatic procedure, without requiring
+Our goal is to apply a cutoff via an automatic procedure, without requiring
 the intervention of the user directly.
@@
 \]
-
 and
-
 \[
@@
 \]
-
 are good approximations of (\ref{dir_mean}) and (\ref{adj_mean}) respectively.
 
-In \cite{Slattery2013} there is a criterion for the cutoff of the random walk
-applicable to both the Forward and the Adjoint method. It requires the set up
-of a relative cutoff threshold $W_c$ and to look for a step $m$ such that
+In \cite{Slattery2013}, a criterion was used for the cutoff of the random walk
+applicable to both the forward and the adjoint method. It requires the set up
+of a relative weight cutoff threshold $W_c$ and to look for a step $m$ such that
 \begin{equation}
-W_m \le W_c W_0.
+W_m \le W_c W_0 \,,
 \label{cutoff}
 \end{equation}
-$W_0$ is the value of the weight at the initial step of the random walk and
-$W_m$ is the value of the weight after $m$ steps.
+where $W_0$ is the value of the weight at the initial step of the random walk
+and $W_m$ is the value of the weight after $m$ steps.
 
@@
 
-We now focus on quantities that might lead to a proper
-number of permutations to compute, in order to get a prescribed order of
-accuracy.
-As regards the Forward method, we know that the expression for the variance
-is defined in formula (\ref{dir_var}).
+We now consider the selection of the number of random walks that should be
+performed to achieve a given accuracy.  Unlike the termination of histories,
+this is a subject that has not been discussed in the Monte Carlo linear
+solver literature, as all previous studies have considered the simulation
+of a prescribed number of histories.
 
+The expression for the variance of the forward method is given by
+formula (\ref{dir_var}).
 In this context, a reasonable criterion to determine the number $\tilde{N}_i$
@@
 is estimated by fixing a finite number of histories.
-Therefore we are controlling the relative standard deviation requiring it not
-to be too large. In other words we are pursuing a statistical setting where the
-uncertainty factor is not dominating over the expected value.
+Therefore, we are controlling the relative standard deviation requiring it not
+to be too large. In other words, we are pursuing a statistical setting where the
+uncertainty factor is small relative to the expected value.
 This simple adaptive approach can be applied for the estimation of each
 component $x_i$. Hence different number of histories may be employed to
-compute different entries of the solution vector. \newline
+compute different entries of the solution vector.
 
-As concerns the Adjoint method, the estimation of the variance for each entry
-is reproduced by Formula (\ref{adj_var}).
-
+As concerns the adjoint method, the estimation of the variance for each entry
+is reproduced by formula (\ref{adj_var}).
 A possible adaptive selection of $\tilde{N}$, in this situation, is
@@
 
-What introduced above can be exploited in order to build
-an a posteriori adaptive algorithm, capable to identify the minimal value of
+The criteria introduced above can be exploited in order to build
+an a posteriori adaptive algorithm, capable of identifying the minimal value of
 $\tilde{N}$ that verifies (\ref{forward_adapt}) or (\ref{adjoint_adapt})
 respectively.
-
-Below there are the pseudo-codes associated with both Forward and Adjoint Monte
-Carlo.
+Algorithms \ref{alg:adaptive_for} and \ref{alg:adaptive_adj} describe
+the Monte Carlo approaches with the adaptive criteria.
 
@@
  return $\theta_i$, $x_i$, $\sigma_i$\;
- \caption{A posteriori adaptive Forward Monte Carlo}
+ \caption{A posteriori adaptive Forward Monte Carlo \label{alg:adaptive_for}}
 \end{algorithm}
@@
  return $\theta$, $\mathbf{x}$, $\boldsymbol{\sigma}$\;
- \caption{A posteriori adaptive Adjoint Monte Carlo}
+ \caption{A posteriori adaptive Adjoint Monte Carlo \label{alg:adaptive_adj}}
 \end{algorithm}
@@
 
-In this section we focus on the pursuit of preconditioners that facilitate the
-convergence of the stochastic schemes introduced so far. A priori, all of the
-following techniques are good candidates to improve the numerical setting.
-However issues about the fill-in effect and the verification of convergence
-conditions limit their employment for our purposes.
+Preconditioning can be incorporated into the Monte Carlo linear solver
+algorithm by simply substituting $A \leftarrow P^{-1}A$ and
+$b \leftarrow P^{-1}b$ in Eq.~\eqref{linsys}.  The Monte Carlo process, however,
+introduced some constraints on the choice of preconditioners used.  Most
+significantly, because the transition probabilities are built based on the
+values of the iteration matrix, it is necessary to have access to the
+entries of the preconditioned matrix $P^{-1}A$.
+Therefore, we must focus on preconditioner selections that enable explicitly
+forming the preconditioned matrix while retaining some of the sparsity of the
+original matrix.
 
-\subsubsection*{Block Preconditioning}
-Consider a matrix $A\in\mathbb{R}^{n\times n}$ of the form
-\[
- A=\begin{bmatrix}A_{11} & A_{12} & \cdots & A_{1p}\\
-    A_{21} & A_{22} & \cdots & A_{2p}\\
-    \vdots & \vdots & \ddots & \vdots \\
-    A_{p1} & \cdots & \cdots & A_{pp}
-   \end{bmatrix}
-\]
-where $p$ is a divisor of $n$ and such that $A_{ii} \in \mathbb{R}^{n_i
-\times n_i}$ are nonsingular square matrices.
-By defining a block diagonal matrix such as
-\[
- D=\begin{bmatrix}A_{11} & 0_{12} & \cdots & 0_{1p}\\
-    0_{21} & A_{22} & \cdots & 0_{2p}\\
-    \vdots & \vdots & \ddots & \vdots \\
-    0_{p1} & \cdots & \cdots & A_{pp}
-   \end{bmatrix}
-\]
-it is possible to compute its inverse.
-$D^{-1}$  can be used as a preconditioner for the original matrix $A$.
-In particular
-\[
- D^{-1}A=\begin{bmatrix}I_{n_1\times n_1} & A_{11}^{-1}A_{12} & \cdots &
-\cdots & A_{11}^{-1}A_{1p} \\
-A_{22}^{-1}A_{21} & I_{n_2\times n_2} & A_{22}^{-1}A_{23} &
-\cdots & A_{22}^{-1}A_{2p}\\
-\vdots & \vdots & \ddots & \vdots & \vdots\\
-\vdots & \vdots & \vdots &\ddots & \vdots \\
-A_{pp}^{-1}A_{p1} &  \cdots & \cdots&
-A_{pp}^{-1}A_{p,p-1} & I_{n_p \times n_p}
-\end{bmatrix}.
-\]
-The ultimate goal is to resort to this preconditioning technique in order to
-achieve the condition $\rho(D^{-1}A)<1$. In fact this is necessary in order to
-reformulate \ref{linsys} into \ref{fixedpoint}.
-
-Convergence of \ref{fixedpoint} with block diagonal preconditioning, from the
+One possible preconditioning approach involves either diagonal or block
+diagonal preconditioning.  Convergence of \ref{fixedpoint} with block
+diagonal preconditioning, from the
 deterministic point of view, is guaranteed when $A$ is an irreducibly
@@
 additional check on $\rho(\hat{H})<1$ to be considered.
+From the discussions in Section~\ref{sec:sdd}
+and \ref{sec:bdd}, selecting a diagonal or block diagonal preconditioner can
+guarantee convergence for matrices that are strictly diagonally dominant
+or block diagonally dominant, respectively.
 
-\subsubsection*{Approximate Inverse of a Sparse Matrix (AINV)}
+Other standard preconditioning approaches can also be used in an attempt
+to reduce $\rho(\hat{H})$ while still retaining the ability to explicitly
+form the preconditioned matrix.  One possibility is the use of incomplete
+LU factorizations \cite{Saad,Benzi2002}.  If $P=LU$ is the preconditioner
+with sparse triangular factors $L$ and $U$, then $P^{-1}A$ can easly be
+formed by back substitution.
+Variants of the algorithm set constraints on the amount of fill-in
+of the sparsity pattern of the triangular factors, allowing the user to
+find a compromise between sparsity and accuracy in
+approximating the inverse of $A$.
 
-This stimulated our interests into other types of preconditioners. At first we
-concentrated on Approximate Inverse preconditioners (see \cite{Saad} and
-\cite{Benzi2002}).
+Another class of preconditioners that are viable for use with Monte Carlo
+linear solvers are approximate inverse preconditioners
+\cite{Saad,Benzi2002,Tuma1996,Tuma1998}.  In these algorithms, an approximation
+to the inverse of $A$ is generated directly, making computation of the
+preconditioned matrix a straightforward task.  As with ILU factorizations,
+multiple versions of approximate inverse preconditioning exist which may
+have different behavior in terms of effectiveness of the preconditioner
+versus the resulting reduction in sparsity.
 
-There are many practical techniques that may be applied in order to build the
-AINV preconditioner.
-One of these is described in \cite{Tuma1996} and \cite{Tuma1998}.
-This is the actual approach we applied to the construction of the
-preconditioner. The technique constructs the preconditioner via a
-bi-conjugation fashion.
-The idea is to compute an approximate factorization of the form $W^T A Z=D$.
-$W$ and $Z$ are unit upper triangular matrices and $D$ is a diagonal matrix.
-
-In order to compute the AINV preconditioner, a \texttt{FORTRAN} code
-provided by Miroslav Tuma (Academy of Sciences, Prague) is used.
-We report as follows the algorithm employed for the construction of the so
-called \textit{Right-looking Factored AINV}. The algorithm provide the user
-with a dropping tolerance $\tau$ which can be used in order to tune the fill-in
-effect
-of the factored preconditioner. \newline
-
-\begin{algorithm}[H]
- \KwData{$p=q=(0,\cdots,0)\in \mathbb{R}^n$, $Z=[z_1,\cdots,z_n]=I_n$,
-$W=[w_1,\cdots,w_n]=I_n$}
- \KwResult{$Z$, $W$}
- \For{$k=1,\cdots, n$}{
- $p_k=w_k^T A e_k$, $q_k=e_k^T A z_k$ \;
-     \For{$i=k+1,\cdots,n$}{
-         $p_i=(w_k^T A e_k)/p_k$, $q_i=(e_k^T A z_i)/q_k$\;
-         Apply a dropping rule to $p_i$, $q_i$\;
-         $w_i=w_i - w_k p_i$, $z_i=z_i - z_k q_i$\\
-         Apply a dropping rule to $w_{j,i}$ and $z_{j,i}$ for $j=1,\cdots,i$\;
-     }\;
-  }\;
- Choose diagonal entries of $D$ as the component of $p$ or $q$\;
- return $Z$, $W$\;
- \caption{Right-looking Factored AINV}
-\end{algorithm}
-
-\subsubsection*{Incomplete LU Factorization Preconditioners (ILU)}
-This set of preconditioners has the goal of computing a sparse lower triangular
-matrix $L$ and a sparse upper triangular matrix $U$ such that the residual
-matrix $R=LU-A$ respects some constraints. One of these constraints may be the
-fact that some entries must be set to zero.
-For instance we can introduce a zero pattern set
-\[
- P\subset \{(i,j)\lvert i\ne j; 1\le i,j\le n\}
-\]
-and use this set in order to impose some constraints on the shape of the
-preconditioner built.\newline
-
-Different variants of the
-Incomplete LU Factorization algorithm are available (see \cite{Saad}).
-However it is
-possible to prove that if all the formulation are well defined, then they are
-all equivalent.
-One of these variants is the following
-
-\begin{algorithm}[H]
- \KwData{$P$}
- \KwResult{$A$}
- For each $(i,j)\in P$ set $a_{ij}=0$\;
- \For{$k=1,\cdot,n-1$}{
- \For{$i=k+1,n$}{
- \If{$(i,k)\notin P$}{
- $a_{ik}=\frac{a_{ik}}{a_{kk}}$
- \For{$j=k+1,\cdots, n$ and for $(i,j)\notin P$}{
-  $a_{ij}=a_{ij}-a_{ik}*a_{kj}$
- }
- }
- }
- }
- return $Z$, $W$\;
- \caption{General Static Pattern ILU}
-\end{algorithm}
-
-This kind of algorithm is guaranteed to terminate without any breakdown just
-for $M$-matrices.
-
-Assuming that the zero pattern $P$ coincides with the zero pattern of $A$ leads
-to the $ILU(0)$. \newline
-
-\begin{algorithm}[H]
- \KwData{$P$}
- \KwResult{$A$}
- \For{$i=2,\cdot,n$}{
- \For{$k=i,i-1$ and for $(i,k)\in NZ(A)$}{
-  $a_{ik}=\frac{a_{ik}}{a_{kk}}$
-  \For{$j=k+1,\cdots,n$ and for $(i,j)\in NZ(A)$}{
-  $a_{ij}=a_{ij}-a_{ik}a_{kj}$}
- }
- }
- \caption{ILU(0).}
-\end{algorithm}
-
-More sophisticated variants of the algorithm enable to enlighten the constraint
-of the sparsity pattern, finding a compromise between sparsity and accuracy in
-approximating the inverse of $A$. We refer to \cite{Saad} and
-\cite{Benzi2002} for further information about it.
-i
+A downside with the use of ILU or approximate inverse preconditioning is the
+the quality of preconditioner needed to achieve $\rho(\hat{A})<1$ is
+difficult to determine.  In fact, in some situations it may happen that
+modifying the preconditioner to reduce $\rho(A)$ may actually lead to an
+increase in $\rho(\hat{A})$, decreasing the effectiveness of the Monte
+Carlo process on the system.

### 2016-01-25 MAX_reports/MC_solvers_article/mcls.tex

@@
 \section{Stochastic Linear Solvers}
+\label{sec:mcls}
 
@@
 \end{equation}
-where $A\in \mathbb{R}^{n\times n}$, $\mathbf{x}$, $\mathbf{b} \in
+where $A\in \mathbb{R}^{n\times n}$ and $\mathbf{x}$, $\mathbf{b} \in
 \mathbb{R}^n$.
 
-By applying a left preconditioning, the system (\ref{linsys}) gets the
-following shape:
-\begin{equation}
-P^{-1}A \mathbf{x}=P^{-1}\mathbf{b},
-\label{linsys_prec}
-\end{equation}
-(\ref{linsys_prec}) can be
-reinterpreted as a fixed point scheme
+Equation~(\ref{linsys}) can be reinterpreted as a fixed point scheme
 \begin{equation}
@@
 \end{equation}
-where $H=I-P^{-1}A$ and $\mathbf{f}=P^{-1}\mathbf{b}$.
+where $H=I-A$ and $\mathbf{f}=\mathbf{b}$.
 
@@
 The last equation can be reinterpreted as the realization of an estimator
-defined on a random walk.
-
-Let us start considering a random walk whose
+defined on a random walk.  Let us start considering a random walk whose
 state space $S$ is characterized by the set of indices of the forcing term
@@
 which is visited in the current step of the random walk.
-
-The way the transition probability is built and the way the initial state of
-the random walk is chosen give birth to two different approaches we are
-going to use: the \textit{Forward} and \textit{Adjoint} methods.
+The method of building the transition probabilities and the selection of
+the initial state of the random walk gives rise to two different approaches:
+the \textit{forward} and \textit{adjoint} methods.
 
 \subsection{Forward Method}
+\label{subsec:forward}
 
@@
 It is important to stress out that the role of vector $\mathbf{h}$ is
-restricted to the construction of the initial probability. Once that is done,
-it is
-not used anymore in the definition of the stochastic process.
+restricted to the construction of the initial probability and is not
+used further in the definition of the stochastic process.
 A possible choice for the transition probability $P$ can be
@@
 
-The Forward method presented above, however, has the limitation of employing an
+The forward method presented above, however, has the limitation of employing an
 entire set of permutations to estimate just a single entry of
 the solution at a time. Therefore, in order to estimate the entire solution
-vector for \ref{linsys_prec}, we have to employ as many sets of permutation as
-the number of entries composing the vector. Of course this might turn to be
-extremely expensive both in terms of time and resources for most of the
-challenging problems coming from applications. This is essentially the reason
-why usually it is preferred to adopt another stochastic algorithm which is
-explained here following: the \textit{Adjoint} Monte Carlo method.
+vector for Eq.~\eqref{linsys}, we have to employ a separate set of
+permutations for each entry in the solution vector.
 
 \subsection{Adjoint Method}
+\label{subsec:adjoint}
 
 A second Monte Carlo method can be derived by considering the linear system
-adjoint to (\ref{linsys_prec})
+adjoint to (\ref{linsys})
 \begin{equation}
-(P^{-1}A)^T\mathbf{y}=\mathbf{d},
+A^T\mathbf{y}=\mathbf{d},
 \label{adj}
@@
 
+\textcolor{red}{Provide some more details relating the adjoint to original problem.}
+
 By reformulating the fixed point scheme, introducing initial probability,
-transition probability and weights
+transition probability and weight
 sequences in the same fashion as done before, the expected value for the
@@
 
-We can notice now that the component of the source vector, in each
-permutation of the random walk, is associated with the initial step. The
-Kronecker delta
-at the end of the series stands for a filter. It means that if the goal is to
-estimate the $j$-th component of the solution vector, just a subset of all the
-permutations gives a contribution to this. The only permutations considered are
-those ones that currently reside on the index associated with the entry to be
-estimated.
+The forward method adds a contribution to the component of the solution
+vector where the random walk began based on the value of the source vector
+in the state in which the walk currently resides.  The adjoint method,
+on the other hand, adds a contribution to the component of the solution
+vector where the random walk currently resides based on the value of the
+source vector in the state in which the walk began.
+The Kronecker delta at the end of the series represents a filter, indicating
+that only a subset of permutations contribute to the $j$-th component
+of the solution vector.
 
-The variance is
-
+The variance is given by
 \begin{equation}
@@
 
-On the same lead of what done for the Forward method, there is need to impose
-finiteness of the expected value and second moment. Therefore the following
+Along the same lines as the development for the forward method, we must
+impose finiteness of the expected value and second moment.
+Therefore the following
 conditions must be verified:
@@
 
-The main advantage of this method, compared to the Forward one, consists in the
-fact that a single set of permutation is used to estimate the entire solution
-vector. Therefore, in general, this algorithm has to be preferred to the
-previous one since it enables to reduce dramatically the computational burden.
+The main advantage of this method, compared to the forward one, consists in the
+fact that a single set of permutations is used to estimate the entire solution
+vector.  Unless only a small portion of the problem is of interest, this
+property often leads to the adjoint method being favored over the forward method.
 
@@
 formulations is as follows:
-
 \begin{equation}
@@
 
-Differently from the collision estimator, the expected value estimator averages
+Thus, the expected value estimator averages
 the deterministic contribution of the iteration matrix over all the potential
@@
 in this case becomes:
-
 \begin{equation}
@@
 
-In the sections dedicated to the numerical tests, both the collision estimator
-and the expected value are employed, in order to test their effectiveness.
-
 \subsection{Hybrid Stochastic/Deterministic Methods}
 
-Even if the standard Monte Carlo, in both its Forward and Adjoint formulation,
-has the advantage of being embarrassingly parallelizable (because of the
-independence of the samplings employed), it has a drawback as
-well. In fact its slow rate of convergence, associated with the
-$\frac{1}{\sqrt{N}}$ behavior predicted but the CLT, prevents its use for
-problems of actual interest in applied linear algebra (with many d.o.f's).
+The direct methods described in Sections~\ref{subsec:foward} and
+\ref{subsec:adjoint} suffer from a slow rate of convergence due to the
+$\frac{1}{\sqrt{N}}$ behavior dictated by the central limit theorem ($N$ here
+is the number of random walks used to estimate the solution).  Furthermore,
+when the spectral radius of the iteration matrix is close to unity, each
+individual random walk may require a large number of transitions to
+approximate the corresponding components in the Neumann series.
+To offset the slow convergence of the central limit theorem, schemes have
+been proposed which combine traditional fixed-point iterative methods with
+the stochastic solvers.  The first such method, due to Halton, was termed
+the Sequential Monte Carlo (SMC) method, and can be written as:
+\begin{algorithm}[H]
+ \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
+ \KwResult{$x_{num}$}
+ $\mathbf{x}^{l}=\mathbf{x}_0$\;
+ \While{not reached convergence}{
+  $\mathbf{r}^{l}=\mathbf{b}-A\mathbf{x}^{l}$\;
+  $\delta \mathbf{x}^{l+1}=(I-H)^{-1}\mathbf{r}^{l}$\;
+  $\mathbf{x}^{l+1}=\mathbf{x}^{l}+\delta \mathbf{x}^{l+1}$\;
+ }
+ $x_{num}=x^{l+1}$\;
+ \caption{Sequential Monte Carlo}
+\end{algorithm}
+The Monte Carlo linear solver method is used to compute the update $\delta
+\mathbf{x}^{l}$.  This algorithm is equivalent to a preconditioned Richardson
+iteration in which the Monte Carlo process acts as the preconditioner.
+Because Monte Carlo is only applied within a single iteration, the central
+limit theorem is only applicable within that iteration rather than to
+the overall convergence behavior of the algorithm.
 
-Moreover, when the spectral radius of the iteration matrix is close to the
-unit
-value, Monte Carlo direct methods reach convergence after an excessive number
-of steps for each
-permutation. In particular this aspect will be pointed out more by some
-numerical tests analyzed in the upcoming sections of this work.
-This motivation makes the standard Mote Carlo paradigm even more
-unpractical for solving linear systems associated with realistic applications.
-\newline
-
-To cope with this, in the early 2000's Evans et al. introduced a new method:
-the
-\textit{Monte Carlo Synthetic Acceleration (MCSA)}
-(\cite{ESW2013} and \cite{EMSH2014}). Earlier contributions on these direction
-are attributed also to other scientists (e.g. \textit{Sequential Method} by
-J.H. Halton). However the MCSA is the algorithm we are going to focus on more
-hereafter. If
-we think about a fixed point formulation of the problem
-such as $\mathbf{x}=H\mathbf{x}+\mathbf{b}$, then the Monte Carlo Synthetic
-Acceleration assumes this form
-
+A further extension of Halton's method, termed
+\textit{Monte Carlo Synthetic Acceleration} (MCSA), has been recently
+introduced (\cite{ESW2013} and \cite{EMSH2014}).  The MCSA algorithm can be written as
 \begin{algorithm}[H]
@@
 \end{algorithm}
-The Monte Carlo method is used to compute the updating contribution $\delta
-\mathbf{x}^{l+\frac{1}{2}}$.
-
-The role of the first line of the iterative process has the goal to compress
-the residual of the previous iteration. This is the actual difference
-between the MCSA and the Sequential Monte Carlo. Once this done, the corrected
-residual
-is employed to compute an update $\delta x^{1+\frac{1}{2}}$ to the solution via
-one of the Monte Carlo techniques introduced before. The utility of compressing
-the residual at the very beginning of the iteration will be shown in the two
-last test cases for numerical validations.
-The intent of MCSA is to designate most of the time employed by the
-computation to the accomplishment of the stochastic part of the algorithm.
-This
-is useful to decrease the probability of fault occurrence in the non-resilient
-steps, which is represented by the non-stochastic component of the numerical
-scheme presented above. In particular, the higher is the percentage of time
-spent on the computation of $\delta x^{1+\frac{1}{2}}$ via Monte Carlo, the
-better it should be in terms of resilience.
-
-\subsubsection{Sequential Monte Carlo}
-
-\subsubsection{Monte Carlo Synthetic Acceleration}
+As with SMC, a Monte Carlo linear solver is used to compute the updating
+contribution $\delta \mathbf{x}^{l+\frac{1}{2}}$.  In this approach, an
+extra step of unpreconditioned Richardson iteration is added to smooth out
+some of the high-frequency noise introduced by the Monte Carlo process.
+This way, the deterministic and stochastic components of the algorithm
+act in a complementary fashion.

### 2016-01-25 doc/NCState_seminar_2014/ncstate_seminar.tex
adding nc state seminar tex file

--- 2015-03-10 doc/siam_cse_2015/siam_cse_2015.tex
@@
 %%---------------------------------------------------------------------------%%
-%% SIAM CSE 2015 Presentation
+%% Shift Roadmap Presentation
 %%---------------------------------------------------------------------------%%
@@
 
-\title[Parallel Monte Carlo Solvers]{Parallel Algorithms for Monte Carlo
-  Linear Solvers}
-\author[Slattery, Hamilton, Evans]{Stuart Slattery
-  \\ Steven Hamilton \\ Tom Evans}
-\date[3/17/15]{March 17, 2015}
+\title[Monte Carlo Linear Solvers]{Monte Carlo Linear Solvers}
+\author[Slattery, Evans, Hamilton]{Stuart Slattery \\ Tom Evans
+  \\ Steven Hamilton}
+\date[11/4/14]{November 4, 2014}
 \institute[ORNL]{\small Oak Ridge National Laboratory}
@@
   to increase dramatically
-  \vfill
+  \vfil
   \begin{itemize}
@@
 \end{frame}
-
 %%---------------------------------------------------------------------------%%
@@
           \begin{itemize}
-            \item Soft failures a component of the tally variance
-            \item Hard failures potentially mitigated by replication
+            \item Soft failures buried within the tally variance
+            \item Hard failures mitigated by replication
             \item Concurrency enabled by several levels of parallelism
@@
   \vfill
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Outline}
-
-  \begin{itemize}
-  \item Monte Carlo Linear Solvers
-    \vfill
-  \item Domain Decomposition and Replication
-    \vfill
-  \item Scaling Studies
-    \vfill
-  \item Matrix-Free and Stochastic Approximate Inverse Algorithms
-  \end{itemize}
-
 \end{frame}
@@
 %%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Monte Carlo}
+\begin{itemize}
+  \item Choose $\mathbf{P}$ and $\mathbf{W}$ such that
+    $\mathbf{H}^{T} = \mathbf{P} \circ \mathbf{W}$
+  \item Typical choice (Monte Carlo Almost-Optimal):
+    \begin{equation*}
+      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ji}| }
+      {\sum_{i=1}^{N} | \mathbf{H}_{ji} |}
+    \end{equation*}
+  \item To estimate solution:
+    \begin{itemize}
+      \item Start a history in random state $i$ by sampling from distribution
+        determined by $\mathbf{b}$
+      \item Transition to new state $j$ based probabilities determined by
+        $\mathbf{P}_i$
+      \item Modify history weight based on corresponding entry in
+        $\mathbf{W}_{ij}$
+      \item Add contribution to $\mathbf{x}_j$ based on current history weight
+    \end{itemize}
+  \item A given random walk can contribute to many different components of
+    the solution vector
+\end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Solving the Heat Equation: Adjoint Method}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics<1>[width=4in]{adjoint_1.png}
+      \includegraphics<2>[width=4in]{adjoint_10.png}
+      \includegraphics<3>[width=4in]{adjoint_100.png}
+      \includegraphics<4>[width=4in]{adjoint_1000.png}
+      \includegraphics<5>[width=4in]{adjoint_10000.png}
+      \includegraphics<6>[width=4in]{adjoint_100000.png}
+      \includegraphics<7>[width=4in]{adjoint_1000000.png}
+      \includegraphics<8>[width=4in]{adjoint_10000000.png}
+    \end{center}
+    \caption{
+      \only<1>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{0} total histories.} }
+      \only<2>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{1} total histories.} }
+      \only<3>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{2} total histories.} }
+      \only<4>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{3} total histories.} }
+      \only<5>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{4} total histories.} }
+      \only<6>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{5} total histories.} }
+      \only<7>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{6} total histories.} }
+      \only<8>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{7} total histories.} }
+    }
+  \end{figure}
+\end{frame}
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Limitations of Monte Carlo}
+\begin{itemize}
+  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
+    is generally intractable
+    \begin{itemize}
+      \item Central limit theorem is barrier to accurate solutions
+    \end{itemize}
+\end{itemize}
+\begin{figure}
+  \centering
+  \includegraphics[width=3.5in]{Ifiss_ConvDiff}
+\end{figure}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}
+
+  \center Residual Monte Carlo Methods
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Sequential Monte Carlo}
+  \begin{itemize}
+  \item Devised by John Halton in the 1960's as a residual method
+\vfill
+    \item Instead of directly solving $\mathbf{Ax}=\mathbf{b}$ with Monte
+      Carlo, apply Monte Carlo to residual equation
+      $\mathbf{A}\delta = \mathbf{r}$
+    \vfill
+    \item Preconditioned Richardson iteration using Monte Carlo as
+      preconditioner:
+      \begin{align*}
+        \mathbf{r}^{k} &= \mathbf{b} - \mathbf{Ax}^{k} \\
+        \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k} \\
+        \mathbf{x}^{k+1} &= \mathbf{x}^k + \delta
+      \end{align*}
+    \vfill
+    \item Exponential convergence achieved by decoupling Monte Carlo
+      error from solution error
+  \end{itemize}
+\end{frame}
+%%---------------------------------------------------------------------------%%
 \begin{frame}{Monte Carlo Synthetic Acceleration}
@@
     \vfill
-  \item Can be abstracted as a general linear solver with Monte Carlo as a
-    preconditioner, $M_{MC}$
+  \item Can be abstracted as a general linear solver
     \vfill
@@
       \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} \\
-      \mathbf{\delta} &= \mathbf{M_{MC}} \mathbf{r}^{k+1/2} \\
+      \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k+1/2} \\
       \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta
     \end{align*}
+  \end{itemize}
+\end{frame}
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Preconditioned MCSA}
+\vspace*{-0.2in}
+  {
+    \small
+    \[
+    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k})
+    \]
+    \[
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
+    \]
+    \[
+    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k+1/2})
+    \]
+    \[
+    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
+    \]
+    \[
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
+    \]
+  }
+  \vspace*{-0.1in}
+  \begin{itemize}
+    \item Requires explicit construction of preconditioned matrix
+      \begin{itemize}
+        \item This has generally limited preconditioner selection to diagonal,
+          block diagonal, sparse approximate inverse approaches
+      \end{itemize}
+    \vfill
+    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
+      in $\mathbf{M^{-1}A}$
+    \vfill
+    \item Need to investigate preconditioning strategies that lead to sparse
+      preconditioned systems (Michele Benzi, Emory)
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Residual Algorithm Comparison}
+  \begin{figure}
+    \includegraphics[width=3.5in]{seq_conv_100}
+  \end{figure}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{3D Equilibrium Thermal Radiation Diffusion}
+\vspace*{-0.1in}
+
+{ \tiny
+  T. Evans, S. Mosher, S. Slattery, S. Hamilton, ``A Monte Carlo
+  synthetic-acceleration method for solving the thermal radiation
+  diffusion equation,'' Journal of Computational Physics \textbf{258},
+  pp.~338--358 (2014).  
+}
+\vspace*{-0.1in}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[ht!]
+        \centerline{ \includegraphics[width=2in,clip]{mesh.pdf}}
+      \end{figure}
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[ht!]
+        \centerline{ \includegraphics[width=2in,clip]{T_multi_mat.pdf}}
+      \end{figure}
+    \end{column}
+  \end{columns}
+
+\vspace*{-0.15in}
+{\small
+  \begin{table}
+    \centering
+    \begin{tabular}{cccc}
+      \toprule
+      \multirow{2}{*}{\bfseries Solver} &
+      \bfseries Total &
+      \bfseries Relative Time &
+      \bfseries Relative Total \\
+      & \bfseries Iterations &
+      \bfseries Per Iteration &
+      \bfseries Time \\
+      \midrule
+      CG (Jacobi) & 42,881 &  2.93 & 1.02 \\
+      CG (ML) & 14,097 & 26.60 & 1.94 \\
+      MCSA (Jacobi) & 90,129 & 1.0 & 1.0 \\
+      \bottomrule
+    \end{tabular}
+  \end{table}
+}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Path to Harder Problems}
+  \begin{itemize}
+    \item Experience with radiation diffusion suggests that for problems
+      where spectral radius is low ($\lesssim 0.8$),
+      MCSA can be competitive with leading methods
+      \begin{itemize}
+        \item Very low cost per iteration to approximate Neumann series
+      \end{itemize}
+    \vfill
+    \item What about more challenging problems?
+      \begin{itemize}
+        \item Neumann series converges more slowly, so histories last longer
+        \item More histories may be required
+        \item What if Neumann series is not convergent?
+      \end{itemize}
   \end{itemize}
@@
 
-  \center Domain Decomposition and Replication
+  \center Polynomial Formulation
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Stopping Criteria -- Weight Cutoff}
+  \begin{itemize}
+    \item Histories are usually terminated when the relative weight drops
+      below a specified cutoff (i.e. $\frac{w}{w_0} < \tau$)
+    \vfill
+    \item Adjoint MCSA on JPWH\_991 matrix with diagonal preconditioning
+      ($\rho ( \mathbf{I-D^{-1}A} ) \approx 0.98$):
+  \end{itemize}
+\begin{figure}
+  \centering
+  \includegraphics[width=2.5in]{WtCutoff}
+\end{figure}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Weight vs. History Length Cutoff}
+\begin{figure}
+  \centering
+  \includegraphics[width=3in]{WtVsLength}
+\end{figure}
+\begin{itemize}
+  \item Unlike weight cutoff, history length cutoff ``saturates'' at
+    some number of histories -- fixed length Neumann series has been
+    effectively reproduced
+  \item Saturation point is higher for longer history lengths
+\end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo as Approximate Polynomial Preconditioning}
+  \begin{itemize}
+    \item Using history length cutoff, Monte Carlo process is approximating
+      a fixed length Neumann series polynomial
+      \[
+      x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+      \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}
+      \]
+    \vfill
+    \item As number of histories grows, iteration count becomes identical to
+      using ``true'' polynomial
+    \vfill
+    \item Why limit ourselves to the Neumann series polynomial?
+      \begin{itemize}
+        \item Chebyshev or GMRES polynomials are viable alternatives
+      \end{itemize}
+  \end{itemize}
+\end{frame}
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Neumann Series Polynomial}
+\begin{table}
+\caption{Adjoint MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+Values are MCSA iteration counts (timing in milliseconds)
+\label{tab:lap_adjoint_neumann}}
+\centering
+\begin{tabular}{cccc}
+\toprule
+Histories per & \multicolumn{3}{c}{Polynomial Order} \\
+\cmidrule(lr){2-4}
+Iteration & 2 & 4 & 6 \\
+\midrule
+250 &  775(100) & 651(96) & 664(110) \\
+500 &  725(122) & 502(104) & 394(95) \\
+1000 & 707(174) & 482(179) & 366(144) \\
+2000 & 703(280) & 471(259) & 356(251) \\
+4000 & 698(494) & 467(497) & 350(458) \\
+8000 & 697(923) & 464(905) & 350(873) \\
+16000 & 695(1796) & 462(1768) & 347(1711) \\
+\bottomrule
+\end{tabular}
+\end{table}
+\end{frame}
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Chebyshev Polynomial}
+\begin{table}
+\caption{Adjoint MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+Values are MCSA iteration counts (timing in milliseconds)
+\label{tab:lap_adjoint_cheby}}
+\centering
+\begin{tabular}{cccc}
+\toprule
+Histories per & \multicolumn{3}{c}{Polynomial Order} \\
+\cmidrule(lr){2-4}
+Iteration & 1 & 2 & 3 \\
+\midrule
+250 & - & - & - \\
+500 & - & - & - \\
+1000 & - & - & - \\
+2000 & - & 328(134) & - \\
+4000 & - & 296(210) & - \\
+8000 & - & 288(380) & 262(423) \\
+16000 & 1132(1866) & 283(721) & 175(550) \\
+\bottomrule
+\end{tabular}
+\end{table}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Polynomial Methods -- Summary}
+  \begin{itemize}
+    \item Using history length cutoff rather than weight cutoff can lead to
+      large reductions in iteration counts
+      \vfill
+    \item A good approximation to a few terms in the Neumann series
+      performs better than a statistically noisy approximation to
+      full series
+      \vfill
+    \item Significant reduction in iteration counts are possible using
+      alternate polynomials, but generally outweighed by increase in number
+      of histories required
+      \begin{itemize}
+        \item May be very beneficial from resiliency and parallel efficiency
+          standpoints
+        \item Will be re-evaluated in future efforts
+      \end{itemize}
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}
+
+  \center Parallelism
 
@@
       \item Domain decomposition determined by the input system
-        \bigskip
-      \item Load balancing straightforward for the forward method
       \end{itemize}
@@
 
-      \small
       \begin{itemize}
@@
       \item Asynchronous nearest neighbor communication of histories
-      \item System-tunable communciation parameters of buffer size and check
-        frequency (performance impact)
-      \item Need an asynchronous strategy for exiting the transport loop
-        without a collective
+      \item Binary asynchronous communication tree for completing
+        transport
       \end{itemize}
 
-      \vspace{-0.25in}
       \begin{figure}[htpb!]
@@
 
-      \vspace{-0.3in}
+      \begin{itemize}
+      \item Extensible to problems where histories may be created
+        (i.e. variance reduction)
+      \end{itemize}
+
       \begin{figure}[htpb!]
         \begin{center}
-          \scalebox{0.85}{ \input{transport_loop.pdftex_t} }
+          \scalebox{0.45}{ \input{binary_comm_tree.pdftex_t} }
         \end{center}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Stopping the Asynchronous Kernel}
-
-  \vspace{-0.25in}
-
-  \begin{figure}[htpb!]
-    \begin{center}
-      \scalebox{0.4}{ \input{master_comm_tree.pdftex_t} }
-    \end{center}
-    \caption{Master/Slave}
-  \end{figure}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-      \vspace{-0.75in}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.4}{ \input{binary_comm_tree.pdftex_t} }
-        \end{center}
-        \caption{Binary Tree}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-      \vspace{-0.5in}
-
-      \begin{figure}[t!]
-        \begin{center}
-          \includegraphics[width=0.99\textwidth]{titan_weak_bvsm.pdf}
-        \end{center}
-        \caption{Weak scaling absolute efficiency}
-        \label{fig:titan_weak_bvsm}
-      \end{figure}
-
-    \end{column}
-
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Replication}
-
-  Different batches of Monte Carlo samplescan be combined in summation via
-  superposition if they have different random number streams. For two
-  different batches:
-
-  \[
-  \mathbf{M_{MC}} \mathbf{x} =
-  \frac{1}{2}(\mathbf{M_1}+\mathbf{M_2})\mathbf{x} 
-  \]
-
-  Consider each of these batches independent \textit{subsets} of a Monte Carlo
-  operator where now the operator can be formed as a general additive
-  decomposition of $N_S$ subsets:
-
-  \[
-  \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
-  \]
-
-  We replicate the linear problem and form each subset on a different group of
-  parallel processes. Applying the subsets to a vector requires an AllReduce
-  to form the sum. Each subset is domain decomposed.
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}
-
-  \center Scaling Studies
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel Test Application -- Nuclear Reactor Analysis}
+\begin{frame}{Parallel Application -- Nuclear Reactor Analysis}
   \small
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}
-
-  \center Matrix-Free and Stochastic Approximate Inverse Algorithms
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Alternative Parallelism -- Additive Schwarz}
@@
 %%---------------------------------------------------------------------------%%
+\begin{frame}
+
+  \center Conclusions
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Additional Thoughts}
+  \begin{itemize}
+    \item Our current implementations rely on performing fixed number of
+      histories per Monte Carlo solve
+      \begin{itemize}
+        \item How can we dynamically select/adapt the number of histories
+          that should be performed?  Use variance-based stopping criteria?
+      \end{itemize}
+    \vfill
+    \item The ``almost optimal'' selection of the probability and weight
+      matrices is arbitrary (as long as
+      $\mathbf{H}/\mathbf{H}^T=\mathbf{P} \circ \mathbf{W}$)
+      \begin{itemize}
+        \item If $\mathbf{P}$ is taken to have a uniform distribution within
+          each row, then sampling from the distribution can be done in
+          constant (rather than logarithmic) time
+      \end{itemize}
+    \vfill
+    \item Exploring Anderson acceleration of MCSA (Tim Kelley)
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Linear Solvers Library (MCLS)}
+
+  \begin{itemize}
+  \item Designed to be easily incorporated with production physics
+    codes
+    \bigskip
+  \item General asynchronous MSOD MCSA implementation
+    \begin{itemize}
+    \item Forward and adjoint Monte Carlo with method of expected
+      values
+    \item Parallel row matrix/vector interface
+    \item General fixed point iteration strategy
+    \item Explicit algebraic preconditioner suite
+    \item Planned extensions to threaded/hybrid parallelism
+    \end{itemize}
+    \medskip
+  \item Implemented in C++
+    \bigskip
+  \item Heavy use of the Trilinos scientific computing libraries
+    \bigskip
+  \item Open-source BSD 3-clause license
+    \bigskip
+  \item \textbf{\url{https://github.com/sslattery/MCLS}}
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Profugus Neutronics Mini-Application}
+
+  A mini-application provides open-source kernels that effectively
+  capture the algorithmic features of full applications
+
+  \begin{columns}
+    \begin{column}{0.3\textwidth}
+      \vspace*{-0.4in}
+      \begin{figure}
+        \centering
+        \includegraphics[width=1.5in]{exnihilo.png}
+      \end{figure}
+    \end{column}
+    \begin{column}{0.7\textwidth}
+      \begin{itemize}
+      \item Extracts core computational kernels from the Exnihilo transport
+        code suite
+        \begin{itemize}
+        \item Denovo: deterministic transport solvers
+        \item Shift: Monte Carlo transport
+        \item Insilico: reactor neutronics (cross sections, geometry,
+          input/output)
+        \end{itemize}
+      \item Exnihilo is export-controlled, Profugus is not
+      \end{itemize}
+    \end{column}
+  \end{columns}
+
+  \vspace*{0.3in}
+  Profugus includes:
+  \begin{itemize}
+  \item Fixed source and eigenvalue problems
+  \item Trilinos solvers/preconditioners
+  \item Generic material and cross section libraries
+  \item \textbf{\url{https://github.com/ORNL-CEES/Profugus}}
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 \begin{frame}{Conclusions}

### 2016-01-25 doc/ecp/pres.tex
Added slides for MCREX DOE presentation.

--- 2015-03-10 doc/siam_cse_2015/siam_cse_2015.tex
@@
 %%---------------------------------------------------------------------------%%
-%% SIAM CSE 2015 Presentation
+%% Beamer ORNL Presentation Template
 %%---------------------------------------------------------------------------%%
@@
 \documentclass{beamer}
-\usefonttheme[onlymath]{serif}
-\usepackage{colortbl}
-\usepackage{longtable,ltcaption,array}
 %% The amssymb package provides various useful mathematical symbols
@@
 \usepackage{multirow}
-\usepackage{caption}
+\usepackage[compatibility=false]{caption}
 \usepackage[labelformat=empty]{subcaption}
-\usepackage{tabularx}
+\usepackage{tikz}
+\usepackage{ragged2e}
+
+\usetikzlibrary{arrows}
 
@@
 \setbeamercolor{item}{fg=MSUgreen}
+\beamertemplatenavigationsymbolsempty
 \setbeamertemplate{headline}{}
-\setbeamertemplate{navigation symbols}{}
+\setbeamertemplate{footline}{}
+\setbeamertemplate{title page}
+{
+  \begin{tabular}{cr}
+    \begin{minipage}{5.5cm}
+      {\RaggedRight \bf \large \textcolor{MSUgreen}{\inserttitle}}\\
+      \vspace{\baselineskip}\\
+      \insertauthor\\
+      \vspace{0.5\baselineskip}\\
+      \insertinstitute\\
+      \insertdate
+    \end{minipage}
+  \end{tabular}
+}
 
@@
 %%---------------------------------------------------------------------------%%
-%% MATH STUFF
-%%---------------------------------------------------------------------------%%
-
-\newcommand{\Pn}{\textit{P}$\negthinspace_N$}
-\newcommand{\SPn}{\textit{SP}$\negthinspace_N$}
-
-\newcommand{\vOmega}{\ensuremath{\ve{\Omega}}}
-\newcommand{\hOmega}{\ensuremath{\hat{\ve{\Omega}}}}
-
-\newcommand{\sigs}{\ensuremath{\sigma_{\text{s}}}}
-\newcommand{\sigf}{\ensuremath{\sigma_{\text{f}}}}
-\newcommand{\sigsm}{\ensuremath{\sigma_{\text{s}m}}}
-\newcommand{\sigsn}{\ensuremath{\sigma_{\text{s}n}}}
-
-\newcommand{\phig}{\ensuremath{\Phi}}
-\newcommand{\Sigmag}{\ensuremath{\mathbf{\Sigma}}}
-\newcommand{\Dg}{\ensuremath{\mathbb{D}}}
-\newcommand{\Fg}{\ensuremath{\mathbb{F}}}
-\newcommand{\Qg}{\ensuremath{\mathbb{Q}}}
-\newcommand{\ug}{\ensuremath{\mathbb{U}}}
-\newcommand{\Ag}{\ensuremath{\mathbb{A}}}
-\newcommand{\Bg}{\ensuremath{\mathbb{B}}}
-\newcommand{\Cg}{\ensuremath{\mathbb{C}}}
-\newcommand{\Jg}{\ensuremath{\mathbb{J}}}
-\newcommand{\sg}{\ensuremath{\mathbb{S}}}
-
-\newcommand{\aphig}{\ensuremath{\Phi^{\dagger}}}
-\newcommand{\aSigmag}{\ensuremath{\mathbf{\Sigma}^{\dagger}}}
-
-\newcommand{\apsi}[1]{\ensuremath{\psi^{\dagger\,#1}}}
-\newcommand{\aphi}[1]{\ensuremath{\phi^{\dagger\,#1}}}
-\newcommand{\aq}[1]{\ensuremath{q^{\dagger\,#1}}}
-\newcommand{\ak}{\ensuremath{k^{\dagger}}}
-
-\newcommand{\normal}{\ensuremath{\hat{\ve{n}}}}
-
-% Phantom minus sign for helping with alignment of matrices
-\newcommand{\phmin}{\ensuremath{\phantom{-}}}
-
-\DeclareMathOperator{\diag}{diag}
-
-%%---------------------------------------------------------------------------%%
 %% TITLE
 
-\title[Parallel Monte Carlo Solvers]{Parallel Algorithms for Monte Carlo
-  Linear Solvers}
-\author[Slattery, Hamilton, Evans]{Stuart Slattery
-  \\ Steven Hamilton \\ Tom Evans}
-\date[3/17/15]{March 17, 2015}
-\institute[ORNL]{\small Oak Ridge National Laboratory}
-\titlegraphic{\includegraphics[width=1.5in]{ORNL_Balls.pdf}}
-
-\defbeamertemplate{footline}{titlefoot}{
-    \vspace{-0.5cm}
-    \hspace*{0.25cm}
-    \includegraphics[width=2cm]{doe_logo}
-    \hfill
-    \includegraphics[width=3.25cm]{WordMarkLeaf}
-    \hspace*{0.5cm}
-}
-
-\setbeamertemplate{title page}
-{
-  \begin{tabular}{cr}
-    \begin{minipage}{4.7cm}
-      {\bf \Large\textcolor{MSUgreen}{\inserttitle}}\\
-
-      \vspace{2\baselineskip}
-      \insertauthor\\
-
-      \insertinstitute\\
-
-      \insertdate
-    \end{minipage}
-    &
-    \begin{minipage}{5cm}
-      \raggedright
-      \vspace{-0.75cm}
-      \includegraphics[height=9.8cm]{combined_background}
-    \end{minipage}
-  \end{tabular}
-}
+\title[MCREX]{Using Monte Carlo Algorithms to Achieve Resiliency \& Performance
+  at Scale for Linear and Nonlinear Solver Applications}
+\author{Thomas Evans \and Steven Hamilton}
+\date[DOE, 2016]{DOE RX-Solvers \\ January 29, 2016}
+\institute[ORNL]{Oak Ridge National Lab}
 
@@
 \begin{document}
-
 %%---------------------------------------------------------------------------%%
@@
 {
+  \usebackgroundtemplate{
+    \includegraphics[width=\paperwidth,height=\paperheight]{title_template}}
   \setbeamertemplate{logo}{}
-\setbeamertemplate{footline}[titlefoot]
   \begin{frame}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Acknowledgements}
+\section*{Outline}
+%%---------------------------------------------------------------------------%%
 
-  This material is based upon work supported by the U.S. Department of
-  Energy, Office of Science, Advanced Scientific Computing Research
-  program.
+\begin{frame}
+  \tableofcontents[pausesections]
+\end{frame}
 
-  \vfill
+%%---------------------------------------------------------------------------%%
+\section{Project Status}
+%%---------------------------------------------------------------------------%%
+\subsection{Project Goals and Target Applications}
+%%---------------------------------------------------------------------------%%
 
-  This research used resources of the Oak Ridge Leadership Computing
-  Facility at the Oak Ridge National Laboratory, which is supported by the
-  Office of Science of the U.S. Department of Energy under Contract No.
-  DE-AC05-00OR22725.
+\begin{frame}{Project Objectives}
+
+  The MCREX project has 3 objectives for the development of Monte Carlo solver
+  methods for linear systems:
+  \begin{itemize}
+  \item Analyze and improve the iterative performance of Monte Carlo
+    algorithms for linear systems
+  \item Analyze and characterize the resilient aspects of Monte Carlo
+    algorithms for linear systems
+  \item Develop and model highly scalable parallel algorithms for the
+    implementation of Monte Carlo algorithms for linear systems
+ \end{itemize}
+ The end goal is to develop a set of resilient, scalable algorithms for the
+ solution of linear systems.  We have made significant progress in each of
+ these areas as summarized below.
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Motivation}
-  \vfill
-\begin{itemize}
-  \item As we move towards exascale computing, the rate of errors is expected
-  to increase dramatically
-  \vfill
-  \begin{itemize}
-  \item The probability that a compute node will fail during the course
-    of a large scale calculation may be near 1
-  \end{itemize}
-  \vfill
-\item Algorithms need to not only have increased concurrency/scalability
-  but have the ability to recover from hardware faults
-  \begin{itemize}
-  \item Lightweight machines
-  \item Heterogeneous machines
-  \item Both characterized by low power and high concurrency
-  \end{itemize}
-  \vfill
-\end{itemize}
-\vfill
-\end{frame}
 
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Towards Exascale Concurrency and Resiliency}
-  \begin{itemize}
-    \item Two basic strategies:
-      \vfill
-      \begin{enumerate}
-        \item State with current ``state of the art'' methods and make
-          incremental modifications to improve scalability and fault
-          tolerance
-          \begin{itemize}
-            \item Many efforts are heading in this direction, attempting
-              to find additional concurrency to exploit
-          \end{itemize}
-        \vfill
-        \item Start with methods having natural scalability and resiliency
-          aspects and work at improving performance (e.g. Monte Carlo)
-          \begin{itemize}
-            \item Soft failures a component of the tally variance
-            \item Hard failures potentially mitigated by replication
-            \item Concurrency enabled by several levels of parallelism
-          \end{itemize}
-      \end{enumerate}
-  \end{itemize}
-  \vfill
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Outline}
+\begin{frame}{Target Applications}
 
   \begin{itemize}
-  \item Monte Carlo Linear Solvers
-    \vfill
-  \item Domain Decomposition and Replication
-    \vfill
-  \item Scaling Studies
-    \vfill
-  \item Matrix-Free and Stochastic Approximate Inverse Algorithms
+
+  \item Models defined by PDEs that result in sparse linear systems
+    \begin{itemize}
+    \item fluid flow
+    \item radiation transport
+    \item electrostatics
+    \end{itemize}
+
+  \item We have begun to analyze eigenvalue systems that are important in
+    network systems analysis and data mining
+
+  \item Much of our analysis work has used and idealized
+    advection-diffusion-reaction system that has properties common to these
+    application areas
+
+  \item Extension to non-linear Navier-Stokes dropped due to budget
+    constraints of funded proposal
   \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}
+\subsection{Overview and Status}
+%%---------------------------------------------------------------------------%%
 
-  \center Monte Carlo Methods
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Monte Carlo for Linear Systems}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Forward Monte Carlo}
+
+\begin{frame}{Solving the Heat Equation}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics<1>[width=4in]{../NCState_seminar_2014/adjoint_1}
+      \includegraphics<2>[width=4in]{../NCState_seminar_2014/adjoint_10}
+      \includegraphics<3>[width=4in]{../NCState_seminar_2014/adjoint_100}
+      \includegraphics<4>[width=4in]{../NCState_seminar_2014/adjoint_1000}
+      \includegraphics<5>[width=4in]{../NCState_seminar_2014/adjoint_10000}
+      \includegraphics<6>[width=4in]{../NCState_seminar_2014/adjoint_100000}
+      \includegraphics<7>[width=4in]{../NCState_seminar_2014/adjoint_1000000}
+      \includegraphics<8>[width=4in]{../NCState_seminar_2014/adjoint_10000000}
+    \end{center}
+    \caption{
+      \only<1>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{0} total histories.} }
+      \only<2>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{1} total histories.} }
+      \only<3>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{2} total histories.} }
+      \only<4>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{3} total histories.} }
+      \only<5>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{4} total histories.} }
+      \only<6>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{5} total histories.} }
+      \only<7>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{6} total histories.} }
+      \only<8>{\textbf{Adjoint solution.}
+        \textit{\sn{1}{7} total histories.} }
+    }
+  \end{figure}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Limitations of Monte Carlo}
   \begin{itemize}
-  \item Choose a row-stochastic matrix $\mathbf{P}$ and weight matrix
-    $\mathbf{W}$ such that $\mathbf{H} = \mathbf{P} \circ \mathbf{W}$
-  \item Typical choice (Monte Carlo Almost-Optimal):
-    \begin{equation*}
-      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ij}| }
-      {\sum_{j=1}^{N} | \mathbf{H}_{ij} | }
-    \end{equation*}
-  \item To compute solution component $\mathbf{x}_i$:
+  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
+    is generally intractable
     \begin{itemize}
-      \item Start a history in state $i$ (with initial weight of 1)
-      \item Transition to new state $j$ based probabilities determined by
-        $\mathbf{P}_i$
-      \item Modify history weight based on corresponding entry in
-        $\mathbf{W}_{ij}$
-      \item Add contribution to $\mathbf{x}_i$ based on current history weight
-        and value of $\mathbf{b}_j$
+    \item Central limit theorem is barrier to accurate solutions
     \end{itemize}
-  \item A given random walk can only contribute to a single component of
-    the solution vector
+    \begin{figure}
+      \centering
+      \includegraphics[width=1.75in]{../NCState_seminar_2014/Ifiss_ConvDiff}
+    \end{figure}
+  \item There is a potentially restrictive limit on the spectral properties of
+    $\mathbf{H}$
+    \begin{equation}
+      \rho(\mathbf{H}) < 1\:, \rho(\hat{\mathbf{H}}) < 1\:, \hat{H}_{ij} =
+      \frac{H_{ij}^2}{P_{ij}}
+    \end{equation}
   \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Sampling Example (Forward Monte Carlo)}
-  \begin{itemize}
-    \item Suppose
-  \begin{equation*}
-    \mathbf{A} = \begin{bmatrix}
-      \phmin 1.0 & -0.2 & -0.6 \\
-      -0.4 & \phmin 1.0 & -0.4 \\
-      -0.1 & -0.4 & \phmin 1.0 \end{bmatrix} \to
-    \mathbf{H} \equiv (\mathbf{I-A}) = \begin{bmatrix}
-       0.0 &  0.2 &  0.6 \\
-       0.4 &  0.0 &  0.4 \\
-       0.1 &  0.4 &  0.0 \end{bmatrix}
-  \end{equation*}
-    then
-  \begin{equation*}
-    \mathbf{P} = \begin{bmatrix}
-       0.0 & 0.25 & 0.75 \\
-       0.5 &  0.0 & 0.5 \\
-       0.2 &  0.8 & 0.0 \end{bmatrix}, \quad
-    \mathbf{W} = \begin{bmatrix}
-       0.0 &  0.8 &  0.8 \\
-       0.8 &  0.0 &  0.8 \\
-       0.5 &  0.5 &  0.0 \end{bmatrix}
-  \end{equation*}
-    \vfill
-    \item If a history is started in state $3$, there is a $20\%$ chance of
-      it transitioning to state $1$ and an $80\%$ chance of moving to state
-      $2$
-  \end{itemize}
-\end{frame}
 
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Solving the Heat Equation: Forward Method}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics<1>[width=4in]{direct_1.png}
-      \includegraphics<2>[width=4in]{direct_10.png}
-      \includegraphics<3>[width=4in]{direct_100.png}
-      \includegraphics<4>[width=4in]{direct_1000.png}
-    \end{center}
-    \caption{
-      \only<1>{\textbf{Forward solution.}
-        \textit{\sn{2.5}{3} total histories.} }
-      \only<2>{\textbf{Forward solution.}
-        \textit{\sn{2.5}{4} total histories.} }
-      \only<3>{\textbf{Forward solution.}
-        \textit{\sn{2.5}{5} total histories.} }
-      \only<4>{\textbf{Forward solution.}
-        \textit{\sn{2.5}{6} total histories.} }
-    }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Monte Carlo Synthetic Acceleration}
   \begin{itemize}
-  \item Devised by Evans and Mosher in the 2000's as an acceleration
-    scheme for radiation diffusion problems (LANL)
+  \item Address issue of convergence constrained by CLF
     \vfill
-  \item Can be abstracted as a general linear solver with Monte Carlo as a
-    preconditioner, $M_{MC}$
+  \item Can be abstracted as a general linear solver
     \vfill
@@
       \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} \\
-      \mathbf{\delta} &= \mathbf{M_{MC}} \mathbf{r}^{k+1/2} \\
+      \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k+1/2} \\
       \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta
     \end{align*}
+  \item We still must address issue of convergence constraint by spectral radius
   \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}
 
-  \center Domain Decomposition and Replication
+\begin{frame}{Preconditioned MCSA}
+\vspace*{-0.2in}
+  {
+    \small
+    \[
+    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k})
+    \]
+    \[
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
+    \]
+    \[
+    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k+1/2})
+    \]
+    \[
+    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
+    \]
+    \[
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
+    \]
+  }
+  \vspace*{-0.1in}
+  \begin{itemize}
+    \item Requires explicit construction of preconditioned matrix
+      \begin{itemize}
+        \item This has generally limited preconditioner selection to diagonal,
+          block diagonal, sparse approximate inverse approaches
+      \end{itemize}
+    \vfill
+    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
+      in $\mathbf{M^{-1}A}$
+    \vfill
+    \item Need to investigate preconditioning strategies that lead to sparse
+      preconditioned systems
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Monte Carlo as Approximate Polynomial Preconditioning}
+  \begin{itemize}
+    \item Using history length cutoff, Monte Carlo process is approximating
+      a fixed length Neumann series polynomial
+      \[
+      x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+      \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}
+      \]
+    \vfill
+    \item As number of histories grows, iteration count becomes identical to
+      using ``true'' polynomial
+    \vfill
+    \item Why limit ourselves to the Neumann series polynomial?
+      \begin{itemize}
+        \item Chebyshev or GMRES polynomials are viable alternatives
+      \end{itemize}
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Neumann Series Polynomial}
+  \begin{table}
+    \caption{Adjoint MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+      Values are MCSA iteration counts (timing in milliseconds)
+      \label{tab:lap_adjoint_neumann}}
+    \centering
+    \begin{tabular}{cccc}
+      \toprule
+      Histories per & \multicolumn{3}{c}{Polynomial Order} \\
+      \cmidrule(lr){2-4}
+      Iteration & 2 & 4 & 6 \\
+      \midrule
+      250 &  775(100) & 651(96) & 664(110) \\
+      500 &  725(122) & 502(104) & 394(95) \\
+      1000 & 707(174) & 482(179) & 366(144) \\
+      2000 & 703(280) & 471(259) & 356(251) \\
+      4000 & 698(494) & 467(497) & 350(458) \\
+      8000 & 697(923) & 464(905) & 350(873) \\
+      16000 & 695(1796) & 462(1768) & 347(1711) \\
+      \bottomrule
+    \end{tabular}
+  \end{table}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Chebyshev Polynomial}
+  \begin{table}
+    \caption{Adjoint MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
+      Values are MCSA iteration counts (timing in milliseconds)
+      \label{tab:lap_adjoint_cheby}}
+    \centering
+    \begin{tabular}{cccc}
+      \toprule
+      Histories per & \multicolumn{3}{c}{Polynomial Order} \\
+      \cmidrule(lr){2-4}
+      Iteration & 1 & 2 & 3 \\
+      \midrule
+      250 & - & - & - \\
+      500 & - & - & - \\
+      1000 & - & - & - \\
+      2000 & - & 328(134) & - \\
+      4000 & - & 296(210) & - \\
+      8000 & - & 288(380) & 262(423) \\
+      16000 & 1132(1866) & 283(721) & 175(550) \\
+      \bottomrule
+    \end{tabular}
+  \end{table}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Polynomial Methods -- Summary}
+  \begin{itemize}
+    \item Using history length cutoff rather than weight cutoff can lead to
+      large reductions in iteration counts
+      \vfill
+    \item A good approximation to a few terms in the Neumann series
+      performs better than a statistically noisy approximation to
+      full series
+      \vfill
+    \item Significant reduction in iteration counts are possible using
+      alternate polynomials, but generally outweighed by increase in number
+      of histories required
+      \begin{itemize}
+        \item May be very beneficial from resiliency and parallel efficiency
+          standpoints
+        \item Will be re-evaluated in future efforts
+      \end{itemize}
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Sparse Approximate Inverse Preconditioning}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Domain Decomposed Monte Carlo}
+\subsection{Software}
+%%---------------------------------------------------------------------------%%
 
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
+\begin{frame}{MCSA Implementations}
+
   \begin{itemize}
-      \item Each parallel process owns a piece of the domain (linear
-        system)
-        \bigskip
-      \item Random walks must be transported between adjacent domains
-        through parallel communication
-        \bigskip
-      \item Domain decomposition determined by the input system
-        \bigskip
-      \item Load balancing straightforward for the forward method
+  \item {\bf MCLS} (\underline{github.com/sslattery/MCLS})
+    \begin{itemize}
+    \item Massively parallel (MPI) implementation of domain decomposed MCSA
+      algorithm
+    \item Implemented in Trilinos framework and callable through Trilinos'
+      abstract numerical algorithms (Thyra, Stratimikos) interface
+    \item Works with any Tpetra row matrix
     \end{itemize}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.75}{ \input{ddnu_example.pdftex_t} }
-        \end{center}
-        \caption{\small Domain decomposition example illustrating
-          how domain-to-domain transport creates communication costs.}
-      \end{figure}
-    \end{column}
-  \end{columns}
+  \item {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
+    \begin{itemize}
+    \item GPU-accelerated on-node MC solve
+    \item Domain-local MC solves implemented for distributed systems as an
+      additive Schwartz method
+    \end{itemize}
+  \end{itemize}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Asynchronous Monte Carlo Transport Kernel}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-
-      \small
-      \begin{itemize}
-      \item General extension of the Milagro algorithm (LANL)
-      \item Asynchronous nearest neighbor communication of histories
-      \item System-tunable communciation parameters of buffer size and check
-        frequency (performance impact)
-      \item Need an asynchronous strategy for exiting the transport loop
-        without a collective
-      \end{itemize}
-
-      \vspace{-0.25in}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.45}{ \input{domain_to_domain.pdftex_t} }
-        \end{center}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-      \vspace{-0.3in}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.85}{ \input{transport_loop.pdftex_t} }
-        \end{center}
-      \end{figure}
-
-    \end{column}
-
-  \end{columns}
-
-\end{frame}
+\section{Transition and Roadmap}
+%%---------------------------------------------------------------------------%%
 
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Stopping the Asynchronous Kernel}
 
-  \vspace{-0.25in}
-
-  \begin{figure}[htpb!]
-    \begin{center}
-      \scalebox{0.4}{ \input{master_comm_tree.pdftex_t} }
-    \end{center}
-    \caption{Master/Slave}
-  \end{figure}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-      \vspace{-0.75in}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.4}{ \input{binary_comm_tree.pdftex_t} }
-        \end{center}
-        \caption{Binary Tree}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-      \vspace{-0.5in}
-
-      \begin{figure}[t!]
-        \begin{center}
-          \includegraphics[width=0.99\textwidth]{titan_weak_bvsm.pdf}
-        \end{center}
-        \caption{Weak scaling absolute efficiency}
-        \label{fig:titan_weak_bvsm}
-      \end{figure}
-
-    \end{column}
-
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Replication}
-
-  Different batches of Monte Carlo samplescan be combined in summation via
-  superposition if they have different random number streams. For two
-  different batches:
-
-  \[
-  \mathbf{M_{MC}} \mathbf{x} =
-  \frac{1}{2}(\mathbf{M_1}+\mathbf{M_2})\mathbf{x} 
-  \]
-
-  Consider each of these batches independent \textit{subsets} of a Monte Carlo
-  operator where now the operator can be formed as a general additive
-  decomposition of $N_S$ subsets:
-
-  \[
-  \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
-  \]
-
-  We replicate the linear problem and form each subset on a different group of
-  parallel processes. Applying the subsets to a vector requires an AllReduce
-  to form the sum. Each subset is domain decomposed.
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}
-
-  \center Scaling Studies
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel Test Application -- Nuclear Reactor Analysis}
-  \small
-  The simplified $P_N$ ($SP_N$) equations are an approximation
-  to the Boltzmann neutron transport equation used to simulate nuclear
-  reactors
-
-  \vspace*{-0.1in}
-
-  \begin{multline}
-    \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-    \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-    \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-    \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-    q(\vec{r},\hat{\Omega},E)
-  \end{multline}
-
-  \begin{multline}
-    -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-      \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-      \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-      \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-      \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-    0,2,4,\cdots,N
-  \end{multline}
-
-  \begin{equation}
-  \hspace*{-0.75in}
-    -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-    \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
-    \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4
-  \end{equation}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{$SP_N$ Assembly Problem}
-\small
-  \begin{columns}
-    \begin{column}{0.4\textwidth}
-      Test problem -- $3 \times 3$ array of fuel assemblies with
-      control rod in center location (Profugus)
-      \begin{itemize}
-      \item 23 energy groups, 2 angular moments, 25M degrees of freedom
-      \item 1,000 computational cores via domain decomposition
-      \item We are interested in solving generalized eigenvalue problem,
-        for this study we use Arnoldi as the eigensolver and compare
-        different methods for solving linear systems
-      \end{itemize}
-    \end{column}
-    \begin{column}{0.6\textwidth}
-      \vspace*{-0.5in}
-      \begin{figure}
+\vfill
 \centering
-        \includegraphics[width=3in]{prob4}
-      \end{figure}
-    \end{column}
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel $SP_N$ Results}
-\begin{table}
-\centering
-\begin{tabular}{cccc}
-\toprule
-%Method & Total GMRES Iterations & Setup Time (s) & Solve Time (s) \\
-\multirow{2}{*}{\bfseries Method} &
-\bfseries Total Linear & \bfseries Setup & \bfseries Solve \\
-& \bfseries Iteration & \bfseries Time (s) & \bfseries Time (s) \\
-\midrule
-GMRES-ILUT      & 1675 & 0.7  & 18.4 \\
-GMRES-AMG       & 626  & 0.7  & 46.0 \\
-GMRES-MGE       & 498  & 1.5  & 33.7 \\
-Richardson-AINV & 5208 & 20.6 & 52.0 \\
-MCSA-AINV       & 1268 & 25.5 & 46.6 \\
-\bottomrule
-\end{tabular}
-\end{table}
-\begin{itemize}
-  \item ILUT preconditioning is winner here, but known to have issues
-    with parallel scaling on large core counts
-  \item Solve times for MCSA are competitive, but setup times are very large
-    due to construction of sparse approximate inverse
-\end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}
-
-  \center Matrix-Free and Stochastic Approximate Inverse Algorithms
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Alternative Parallelism -- Additive Schwarz}
-  \begin{itemize}
-    \item Instead of performing Monte Carlo on full problem, another
-      possibility is to apply Monte Carlo as an additive Schwarz approach
+\huge Questions?
 \vfill
-    \item Decompose problem into (possibly overlapping) domains
-    \vfill
-    \item Perform Monte Carlo on individual subdomains
-      \begin{itemize}
-        \item No communication costs in Monte Carlo problem!
-      \end{itemize}
-    \vfill
-    \item With domain decomposed Monte Carlo, iteration counts are effectively
-      independent of the number of processors
-    \vfill
-    \item In an additive Schwarz approach, the preconditioner will become
-      less effective as processor counts grow -- algorithmic scalability
-      may be an issue
-      \vfill
-    \item \textbf{Replication for resiliency and performance}
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{More Parallelism -- Threading}
-  \begin{itemize}
-    \item Within a Monte Carlo solve, every history is independent of other
-      histories -- great potential for highly concurrent hardware
-      (GPU, Xeon Phi)
-    \vfill
-    \item Polynomial formulation enables a priori determination of
-      operation counts per thread
-      \vfill
-    \item Memory locality an issue due to random access via random
-      walks (block formulation?)
-      \vfill
-    \item Early experiments using the Trilinos Kokkos library show promising
-      performance for multi-core CPUs
-    \vfill
-    \item Team members recently took part in OLCF ``Hackathon'' in
-      late October to begin implementing computation kernels in
-      OpenACC to allow for GPU capability on Titan with early results
-      indicating \textbf{1.3-9.4x} speedup for MCSA (largely dependent
-      on random number generation)
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Conclusions}
-  \begin{itemize}
-  \item Monte Carlo methods offer great potential for both resilient and
-    highly parallel solvers
-    \vfill
-  \item For certain classes of problems, Monte Carlo methods can be
-    competitive with leading modern solvers
-    \vfill
-    \item Extending methods to broader problem areas is significant challenge
-      and an attractive area for continued research
-      \vfill
-    \item Performance modeling and resiliency simulations this FY
-  \end{itemize}
 \end{frame}

### 2016-01-25 doc/ecp/pres.tex
Adding preconditioning slides.

@@
     \item Significant reduction in iteration counts are possible using
-      alternate polynomials, but generally outweighed by increase in number
+      alternate polynomials, but partially offset by increase in number
       of histories required
@@
 
-\begin{frame}{Sparse Approximate Inverse Preconditioning}
+\begin{frame}{Preconditioning}
+  \begin{itemize}
+    \item Preconditioning for Monte Carlo introduces nonstandard considerations
+      \begin{itemize}
+        \item Elements of preconditioned matrix must be explicitly available
+        \item Goal of preconditioning is to reduce $\rho(\hat{\mathbf{H}})$,
+          may not follow ``standard'' trends
+      \end{itemize}
+  \end{itemize}
+  \vfill
+
+  \begin{table}[!h]
+  \centering
+  \begin{tabular}{|c|c|c|c|}
+  \hline
+  \textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
+  &$\rho(\hat{H})$\\
+  \hline
+   3 & 1.4352 & $0.9804$ & $0.9531$\\
+  \hline
+   9 & 3.0220 & $0.9790$ & $0.9674$\\
+  \hline
+   33 & 9.8164 & $0.9783$  & $0.9774$\\
+  \hline
+  \end{tabular}
+  \caption{$\rho(H)$ and $\rho(\hat{H})$ for advection--diffusion
+  matrix with block diagonal preconditioning.}
+  \label{tab:ifiss_block}
+  \end{table}
+\end{frame}
+
+\begin{frame}{Preconditioning}
+  \begin{itemize}
+    \item Several options have been analyzed
+      \begin{itemize}
+        \item (Block) diagonal preconditioners, suitable for (block)
+          diagonally dominant systems
+        \item ILU-style preconditioners are possible, but require additional
+          filtering to maintain sparsity in preconditioned system
+        \item Approximate inverse preconditioning is attractive, forming
+          preconditioned matrix only requires sparse matrix-matrix product
+      \end{itemize}
+  \end{itemize}
+
+  \begin{table}[!h]
+  \centering
+  \begin{tabular}{|c|c|c|c|}
+  \hline
+  \textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
+  &$\rho(\hat{H})$\\
+  \hline
+   0.05 & 8.2797 & $0.9610$ & $1.0231$\\
+  \hline
+   0.01 & 33.1390 & $0.8668$ & $0.8279$\\
+  \hline
+  \end{tabular}
+  \caption{$\rho(H)$ and $\rho(\hat{H})$ for 2D
+  Laplacian with AINV preconditioning.}
+  \label{tab:lap2d_ainv}
+  \end{table}
 

### 2016-01-25 doc/ecp/pres.tex
Updated first section draft.

@@
 \setbeamertemplate{headline}{}
-\setbeamertemplate{footline}{}
+\setbeamertemplate{footline}{
+  \insertframenumber/\inserttotalframenumber}
 \setbeamertemplate{title page}
@@
 \begin{frame}
-  \tableofcontents[pausesections]
+  \tableofcontents
 \end{frame}
@@
 
+%%---------------------------------------------------------------------------%%
+
 \begin{frame}{Preconditioning}
@@
 
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Domain Decomposed Monte Carlo}
+
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \begin{itemize}
+      \item Each parallel process owns a piece of the domain (linear
+        system)
+        \bigskip
+      \item Random walks must be transported between adjacent domains
+        through parallel communication
+        \bigskip
+      \item Domain decomposition determined by the input system
+        \bigskip
+      \item Load balancing not yet addressed
+      \end{itemize}
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.75}{ \input{ddnu_example.pdftex_t} }
+        \end{center}
+        \caption{\small Domain decomposition example illustrating
+          how domain-to-domain transport creates communication costs.}
+      \end{figure}
+    \end{column}
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Asynchronous Monte Carlo Transport Kernel}
+
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+      \vspace{-0.1in}
+      \small
+      \begin{itemize}
+      \item General extension of the Milagro algorithm (LANL)
+      \item Asynchronous nearest neighbor communication of histories
+      \item System-tunable communication parameters of buffer size and check
+        frequency (performance impact)
+      \item Need an asynchronous strategy for exiting the transport loop
+        without a collective (running sum)
+      \end{itemize}
+
+      \vspace{-0.15in}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.4}{ \input{domain_to_domain.pdftex_t} }
+        \end{center}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+
+      \vspace{-0.25in}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.825}{ \input{transport_loop.pdftex_t} }
+        \end{center}
+      \end{figure}
+
+    \end{column}
+
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Parallel Test - Simplified $P_N$ ($SP_N$) Assembly Problem}
+\small
+\begin{columns}
+
+    \begin{column}{0.55\textwidth}
+
+      \vspace{-0.2in}
+      \begin{figure}
+        \centering
+        \includegraphics[width=2.0in]{../siam_cse_2015/presentation/prob4}
+        \vspace{-0.2in}
+        \caption{$SP_N$ solution example}
+      \end{figure}
+
+      \vspace{-0.1in}
+
+      The ($SP_N$) equations are an approximation to the Boltzmann neutron
+      transport equation used to simulate nuclear reactors
+
+      \vspace{-0.2in}
+
+      \[
+        \small
+        -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
+        \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
+        \mathbb{F}_{nm} \mathbb{U}_m
+      \]
+
+    \end{column}
+
+    \begin{column}{0.45\textwidth}
+      Scaling problem -- $1 \times 1$ to $17 \times 17$ array of fuel
+      assemblies with 289 pins each resolved by a $2 \times 2$ spatial mesh
+      and 200 axial zones
+
+      \vfill
+
+      \begin{itemize}
+      \item 7 energy groups, 1 angular moment, 1.6M to 273.5M degrees of
+        freedom
+      \item 64 to 10,800 computational cores via domain decomposition
+      \item We are usually interested in solving generalized eigenvalue
+        problem - we use the operator from these problems to test the kernel
+        scaling
+      \end{itemize}
+    \end{column}
+
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Domain Decomposition Scaling}
+
+  \vspace{-0.2in}
+
+  \begin{table}[htb!]
+    \tiny
+    \begin{center}
+      \begin{tabular}{rrrrrrr}
+        \toprule
+        \multicolumn{1}{r}{Cores} &
+        \multicolumn{1}{r}{DOFs} &
+        \multicolumn{1}{r}{DOFs/Core} &
+        \multicolumn{1}{r}{Time Min (s)} &
+        \multicolumn{1}{r}{Time Max (s)} &
+        \multicolumn{1}{r}{Time Ave (s)} &
+        \multicolumn{1}{r}{Efficiency}
+        \\ \midrule
+        %%
+        256 & 273\,509\,600 & 1\,068\,397 & 515.51 & 515.52 & 515.52 & 1.00 \\
+        1\,024 & 273\,509\,600 & 267\,099 & 122.76 & 122.77 & 122.76 & 1.05 \\ %CACHE
+        4\,096 & 273\,509\,600 & 66\,775 & 27.96 & 27.97 & 27.96 & 1.15 \\
+        7\,744 & 273\,509\,600 & 35\,319 & 17.72 & 17.72 & 17.72 & 0.96 \\
+        10\,816 & 273\,509\,600 & 25\,288 & 13.72 & 13.72 & 13.72 & 0.89 \\
+        %%
+        \bottomrule
+      \end{tabular}
+    \end{center}
+    \vspace{-0.07in}
+    \caption{\small Strong Scaling}
+  \end{table}
+
+  \vspace{-0.4in}
+
+  \begin{table}[htb!]
+    \tiny
+    \begin{center}
+      \begin{tabular}{rrrrrrr}
+        \toprule
+        \multicolumn{1}{r}{Cores} &
+        \multicolumn{1}{r}{DOFs} &
+        \multicolumn{1}{r}{DOFs/Core} &
+        \multicolumn{1}{r}{Time Min (s)} &
+        \multicolumn{1}{r}{Time Max (s)} &
+        \multicolumn{1}{r}{Time Ave (s)} &
+        \multicolumn{1}{r}{Efficiency}
+        \\ \midrule
+        %%
+        64 & 1\,618\,400 & 25\,288 & 12.65 & 12.65 & 12.65 & 1.00 \\
+        256 & 6\,473\,600 & 25\,288 & 12.86 & 12.86 & 12.86 & 0.98 \\
+        1\,024 & 25\,894\,400 & 25\,288 & 14.59 & 14.59 & 14.59 & 0.87 \\
+        4\,096 & 103\,577\,600 & 25\,288 & 14.25 & 14.25 & 14.25 & 0.89 \\
+        7\,744 & 195\,826\,400 & 25\,288 & 14.75 & 14.78 & 14.75 & 0.86 \\
+        10\,816 & 273\,509\,600 & 25\,288 & 13.72 & 13.72 & 13.72 & 0.92 \\
+        %
+        \bottomrule
+      \end{tabular}
+    \end{center}
+        \vspace{-0.07in}
+    \caption{\small Weak Scaling}
+  \end{table}
+
+  \vspace{-0.25in}
+
+  \begin{itemize}
+    \small
+  \item Titan Cray XK7 was used with no GPUs - used Eos parameters
+  \item Results are consistent with transport literature
+  \item History length of 15 and 3 histories per DOF
+  \item Consider these effective operator apply times of 15 realizations of
+    $\ve{H}$
+  \item MC dominates run times in solver applications
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Replication}
+
+  \vspace{-0.1in}
+
+  Different batches of Monte Carlo samples can be combined in summation via
+  superposition if they have different random number streams.  Consider each
+  of these batches independent \textit{subsets} of a Monte Carlo operator
+  where now the operator can be formed as a general additive decomposition of
+  $N_S$ subsets:
+
+  \[
+  \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
+  \]
+
+  We replicate the linear problem and form each subset on a different group of
+  parallel processes. Applying the subsets to a vector requires an AllReduce
+  to form the sum. Each subset is domain decomposed.
+  \vfill
+  This provides a mechanism to support \textit{resiliency}; the replicated
+  component provides {\bf useful} work, not just redundancy
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Replication Scaling}
+
+  \begin{table}[htb!]
+    \tiny
+    \begin{center}
+      \begin{tabular}{rrrrrrr}
+        \toprule
+        \multicolumn{1}{r}{Subsets} &
+        \multicolumn{1}{r}{Cores} &
+        \multicolumn{1}{r}{DOFs} &
+        \multicolumn{1}{r}{Time Min (s)} &
+        \multicolumn{1}{r}{Time Max (s)} &
+        \multicolumn{1}{r}{Time Ave (s)} &
+        \multicolumn{1}{r}{Efficiency}
+        \\ \midrule
+        %%
+        1 & 256 & 6\,473\,600 & 12.65 & 12.65 & 12.65 & 1.00 \\
+        2 & 512 & 6\,473\,600 & 6.62 & 6.80 & 6.72 & 0.93 \\
+        3 & 768 & 6\,473\,600 & 4.50 & 4.73 & 4.66 & 0.89 \\
+        4 & 1\,024 & 6\,473\,600 & 3.62 & 3.81 & 3.71 & 0.83 \\
+        %
+        \bottomrule
+      \end{tabular}
+    \end{center}
+    \caption{\small Fixed algorithmic scaling}
+  \end{table}
+
+  \vfill
+
+  \begin{itemize}
+    \small
+  \item Timings are Monte Carlo time plus AllReduce time
+  \item 256 cores per subset with domain decomposition
+  \item $1/N_S$ samples per subset gives flat algorithmic performance
+  \item Two performance issues
+    \begin{itemize}
+    \item Dividing up samples creates work starvation
+    \item AllReduce buffer size is constant regardless of work starvation
+    \end{itemize}
+  \item Research is needed for a more performant subset combination
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{GPU Optimization}
+
+  We have implemented a version of MCSA that uses GPUs to perform the MC
+  linear solve.
+
+  Distributed parallelism is enabled through an additive Schwartz method, all
+  MC solves
+  \begin{itemize}
+  \item All MC random walks are local
+  \item No inter-domain communication of histories
+  \end{itemize}
+
+  \begin{table}[h]
+    \begin{center}
+      \caption{2D parabolic problem with AINV preconditioning.}
+      \begin{tabular}{|c|c|c|c|}\hline
+        Node & Iterations & Time (sec) & Relative Residual \\\hline
+        CPU & 6 & 992.23 & $1.23\times 10^{-7}$ \\\hline
+        GPU & 6 & 29.83 & $1.53\times 10^{-7}$ \\\hline
+      \end{tabular}
+    \end{center}
+  \end{table}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\subsection{Research Requirements}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}
+
+  The following important questions need to be answered before MC methods can
+  be applied to linear systems as a ``standard'' solver:
+
+  \begin{itemize}
+  \item Automatically determine history termination criteria
+    \begin{itemize}
+    \item We have investigated adaptive schemes based on variance
+    \item However, variance is not directly correlated with error
+    \end{itemize}
+  \item Identify classes of matrices and preconditioners that satisfy
+    convergence constraints
+    \begin{itemize}
+    \item This has been done for SDD and GDD
+    \end{itemize}
+  \item Demonstrate resiliency to hard MPI errors through replicated domain
+    parallelism
+  \item Further investigate non-linear, time-dependent problems
+    \begin{itemize}
+    \item This was part of the original proposal
+    \item Non-linear, sub-cycle time-stepping can result in strongly diagonally
+      dominant problems for which MCSA has been shown to be effective
+    \end{itemize}
+  \end{itemize}
 \end{frame}
@@
 
-  \begin{itemize}
-  \item {\bf MCLS} (\underline{github.com/sslattery/MCLS})
+  {\bf MCLS} (\underline{github.com/sslattery/MCLS})
   \begin{itemize}
@@
   \end{itemize}
-  \item {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
+  \vfill
+  {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
   \begin{itemize}
@@
     additive Schwartz method
-    \end{itemize}
   \end{itemize}

### 2016-01-26 doc/ecp/pres.tex
Added slide content.

@@
 
-  \item Models defined by PDEs that result in sparse linear systems
+  \item Models defined by PDEs that result in sparse linear systems, examples:
     \begin{itemize}
@@
     \item radiation transport
-    \item electrostatics
+    \item gas dynamics
     \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
+
+\begin{frame}{Summary}
+
+  Much of the first 2 years of the project were attempting to resolve
+  \textit{algorithmic} challenges related to solver performance and efficiency
+  \begin{itemize}
+  \item Iterative schemes to improve efficiency
+  \item Analysis to determine convergence
+  \item Development and application of effective preconditioners
+  \end{itemize}
+
+  \vfill
+
+  We have also developed scalable parallel algorithms that can take advantage
+  of the benefits of replicating Monte Carlo random walks
+  \begin{itemize}
+  \item Developed multilevel parallel strategies for domain decomposed Monte
+    Carlo that allows redundancy
+  \item Developed concurrent on-node random walk algorithms for GPU
+    optimization
+  \end{itemize}
+
+  \vfill
+
+  We have developed two open source software packages that contain
+  production-quality versions of these solvers (MPI and GPU).
+
+\end{frame}
 
@@
     $\mathbf{H}$
-    \begin{equation}
+    \begin{equation*}
       \rho(\mathbf{H}) < 1\:, \rho(\hat{\mathbf{H}}) < 1\:, \hat{H}_{ij} =
       \frac{H_{ij}^2}{P_{ij}}
-    \end{equation}
+    \end{equation*}
   \end{itemize}
@@
   \item We still must address issue of convergence constraint by spectral radius
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Preconditioned MCSA}
-\vspace*{-0.2in}
-  {
-    \small
-    \[
-    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k})
-    \]
-    \[
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
-    \]
-    \[
-    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k+1/2})
-    \]
-    \[
-    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
-    \]
-    \[
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
-    \]
-  }
-  \vspace*{-0.1in}
-  \begin{itemize}
-    \item Requires explicit construction of preconditioned matrix
-      \begin{itemize}
-        \item This has generally limited preconditioner selection to diagonal,
-          block diagonal, sparse approximate inverse approaches
-      \end{itemize}
-    \vfill
-    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
-      in $\mathbf{M^{-1}A}$
-    \vfill
-    \item Need to investigate preconditioning strategies that lead to sparse
-      preconditioned systems
   \end{itemize}
@@
   \begin{itemize}
+    \item Preconditioning needed to address convergence constraints due to
+      spectral radius conditions
     \item Preconditioning for Monte Carlo introduces nonstandard considerations
       \begin{itemize}
-        \item Elements of preconditioned matrix must be explicitly available
+      \item Elements of preconditioned matrix must be explicitly available,
+        sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity in
+        $\mathbf{M^{-1}A}$
         \item Goal of preconditioning is to reduce $\rho(\hat{\mathbf{H}})$,
@@
   \end{table}
+\end{frame}
 
@@
       \item Load balancing not yet addressed
+        \bigskip
+      \item Uses modified version of the Milagro algorithm (LANL)
       \end{itemize}
@@
     \end{column}
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Asynchronous Monte Carlo Transport Kernel}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-      \vspace{-0.1in}
-      \small
-      \begin{itemize}
-      \item General extension of the Milagro algorithm (LANL)
-      \item Asynchronous nearest neighbor communication of histories
-      \item System-tunable communication parameters of buffer size and check
-        frequency (performance impact)
-      \item Need an asynchronous strategy for exiting the transport loop
-        without a collective (running sum)
-      \end{itemize}
-
-      \vspace{-0.15in}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.4}{ \input{domain_to_domain.pdftex_t} }
-        \end{center}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-      \vspace{-0.25in}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.825}{ \input{transport_loop.pdftex_t} }
-        \end{center}
-      \end{figure}
-
-    \end{column}
-
   \end{columns}
@@
 
-\begin{frame}
+\begin{frame}{Remaining Research}
 
@@
   \end{itemize}
+  \vfill
+  All software is BSD open source.
+  \vfill
+  Uses TriBITS (Trilinos) open source build system
+  \vfill
+  Both software packages contain a complete suite of unit-tests integrated
+  through GoogleTest (Alea) or TeuchosTest (MCLS)
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Publications and Presentations}
+
+  \underline{Publications}
+  \begin{itemize}
+  \item M. Benzi, T.M. Evans, S.P. Hamilton, M.L. Pasini, and
+    S.R. Slattery. Analysis of Monte Carlo Accelerated Iterative Methods for
+    Sparse Linear Systems. \textit{in preparation}, 2016.
+  \item S.R. Slattery, T.M. Evans, P.P.H. Wilson. A Spectral Analysis of the
+    Domain Decomposed Monte Carlo Method for Linear Systems. \textit{Nuclear
+      Engineering Design},  {\bf 295}, 632--638, 2015.
+  \item T.M. Evans, S.W. Mosher, S.R. Slattery, and S.P. Hamilton. A Monte Carlo
+    Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion
+    Equation. \textit{J. Comp. Phys.}, {\bf 258}, 338--358, 2014.
+  \end{itemize}
+
+  \underline{Presentations}
+  \begin{itemize}
+  \item Monte Carlo Synthetic Acceleration Methods for Sparse Linear Systems
+    (M.L. Pasini, M. Benzi, T. Evans, S. Hamilton, S. Slattery), \textit{SIAM
+      ALA}, October 30, 2015.
+  \item Parallel Algorithms for the Monte Carlo Synthetic Acceleration Linear
+    Solver Method (S. Slattery, T. Evans, S. Hamilton), \textit{SIAM CSE},
+    Salt Lake City, 2015.
+  \item Iterative Performance of Monte Carlo Linear Solver Methods
+    (M.L. Pasini), \textit{SIAM CSE}, Salt Lake City, 2015.
+  \item A multilevel Monte Carlo method for linear systems (S. Slattery,
+    T. Evans, S. Hamilton), \textit{Copper Mountain Conference on Iterative
+      Solvers}, Copper MT, CO, 2014.
+  \item Monte Carlo Synthetic Acceleration as approximate polynomial
+    preconditioning (S. Hamilton, T. Evans, S. Slattery), \textit{Copper
+      Mountain Conference on Iterative Solvers}, Copper MT, CO, 2014.
+  \item Monte Carlo Linear Solvers (S. Slattery, T. Evans, S. Hamilton),
+    \textit{NC State Seminar}, Nov. 4, 2014.
+  \item Monte Carlo Linear Solvers (S. Hamilton, T. Evans, S. Slattery),
+    \textit{Emory Mathematics Dept. Seminar}, Sept. 19, 2014.
+  \item Monte Carlo Methods for Linear Systems (T. Evans), \textit{MIT Nuclear
+      Engineering Dept Seminar Series}, Sept. 30, 2014.
+  \end{itemize}
 
@@
 %%---------------------------------------------------------------------------%%
+\subsection{ECP Application Requirements}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Connection to ECP Requirements}
+
+  \begin{itemize}
+
+  \item Inter-Operability
+    \begin{itemize}
+    \item MCLS built using Trilinos solver API
+    \item Utilizes Hypre/ParaSAILS for AINV preconditioning
+    \item Enables testing of MC solver algorithms in a wide variety of
+      applications, ie. anyone who uses Trilinos
+    \end{itemize}
+
+  \item Scalability
+    \begin{itemize}
+    \item Multilevel inter-node domain decomposition methods show excellent
+      scalability
+    \item GPU-enabled on-node MC kernel yields measures speedups of 30 -
+      50$\times$
+    \end{itemize}
+
+  \item Resiliency
+    \begin{itemize}
+    \item Replicated subsets enable useful work to occur at the same time as
+      providing redundancy
+    \item Yields protection against hard errors
+    \end{itemize}
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Inter-Operatibility Example}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{ECP Coordinated Reasearch}
+
+  \underline{\bf Goal}: Extend the general algorithmic MC exascale methods
+  w.r.t. scalability and resiliency into a broader application space
+
+  \vfill
+
+  MC methods are used in many DOE application areas:
+  \begin{itemize}
+  \item high-energy density physics
+  \item astrophysics
+  \item nuclear engineering
+  \item particle accelerators and neutron scattering
+  \item quantum chemistry
+  \item visualization
+  \item fluid dynamics
+  \item computational biology
+  \end{itemize}
+
+  \vfill
+
+  Much of the algorithmic research performed in MCREX is directly applicable
+  to the use of MC methods in these other application areas
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Continued Research for MC Linear Solvers}
+
+  Investigate use in time-dependent applications, e.g. non-linear radiation
+  hydrodynamics for high-energy density physics applications:
+  \begin{itemize}
+  \item Used as inner linear solver in a non-linear (Newton) iteration
+  \item Time-dependent applications generally have strong diagonal-dominance
+    that makes MCSA competitive with Krylov methods
+  \item MCSA has resiliency and parallel efficiency benefits over subspace
+    methods
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Building on ASCR Funding}
+
+  MCREX has leveraged and built on work derived from several ASCR projects:
+  \begin{itemize}
+  \item Trilinos
+  \item XPress
+  \item xSim
+  \item Ideas
+  \end{itemize}
+
+\end{frame}
 
@@
 %%---------------------------------------------------------------------------%%
+\section*{Extra Slides}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Preconditioned MCSA}
+\vspace*{-0.2in}
+  {
+    \small
+    \[
+    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k})
+    \]
+    \[
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
+    \]
+    \[
+    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k+1/2})
+    \]
+    \[
+    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
+    \]
+    \[
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
+    \]
+  }
+  \vspace*{-0.1in}
+  \begin{itemize}
+    \item Requires explicit construction of preconditioned matrix
+      \begin{itemize}
+        \item This has generally limited preconditioner selection to diagonal,
+          block diagonal, sparse approximate inverse approaches
+      \end{itemize}
+    \vfill
+    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
+      in $\mathbf{M^{-1}A}$
+    \vfill
+    \item Need to investigate preconditioning strategies that lead to sparse
+      preconditioned systems
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Asynchronous Monte Carlo Transport Kernel}
+
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+      \vspace{-0.1in}
+      \small
+      \begin{itemize}
+      \item General extension of the Milagro algorithm (LANL)
+      \item Asynchronous nearest neighbor communication of histories
+      \item System-tunable communication parameters of buffer size and check
+        frequency (performance impact)
+      \item Need an asynchronous strategy for exiting the transport loop
+        without a collective (running sum)
+      \end{itemize}
+
+      \vspace{-0.15in}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.4}{ \input{domain_to_domain.pdftex_t} }
+        \end{center}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+
+      \vspace{-0.25in}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.825}{ \input{transport_loop.pdftex_t} }
+        \end{center}
+      \end{figure}
+
+    \end{column}
+
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 \end{document}

### 2016-01-26 doc/ecp/pres.tex
Updated presentation.

@@
 \usepackage{multirow}
+\usepackage{subfigure}
 \usepackage[compatibility=false]{caption}
-\usepackage[labelformat=empty]{subcaption}
 \usepackage{tikz}
 \usepackage{ragged2e}
+\usepackage{algorithm,algorithmic}
 
@@
       \begin{itemize}
+        \setlength{\itemsep}{-0.4\baselineskip}
       \item Each parallel process owns a piece of the domain (linear
@@
 
-\begin{frame}{Remaining Research}
+\begin{frame}{Continued Research is Necessary}
 
@@
 
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Publications and Presentations}
+
   \underline{Presentations}
   \begin{itemize}
+    \footnotesize
+    \setlength{\itemsep}{-0.1\baselineskip}
   \item Monte Carlo Synthetic Acceleration Methods for Sparse Linear Systems
@@
 
+  Using the MCLS interface to Trilinos, it is straightforward to implement
+  MCSA in other applications
+  \vfill
+  We implemented MCSA in the production CFD code, Drekar (SNL) using a
+  Newton-MCSA algorithm:
+  \begin{algorithm}[H]
+    \footnotesize
+    \begin{algorithmic}[1]
+      \STATE $k := 0$
+      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
+        ||\ve{F}(\ve{u}^{0})||$}
+      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
+      \COMMENT{Automatic differentiation - Panzer}
+      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
+      \COMMENT{Solve for the Newton correction with MCSA}
+      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
+      \STATE $k \leftarrow k+1$
+      \ENDWHILE
+    \end{algorithmic}
+    \caption{FANM}
+  \end{algorithm}
+  \vfill
+  Because MCLS was built with Trilinos abstract numerical algorithms this
+  integration was straightforward and we can test MCLS in a whole new problem
+  space
+  \vfill
+  We are also able to utilize all of Trilinos' preconditioners
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Thermal Convection Cavity Results}
+
+  \begin{columns}
+    \begin{column}{0.6\textwidth}
+
+      \small{
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lccc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}&
+              \multicolumn{1}{c}{NK}&
+              \multicolumn{1}{c}{FANM}&
+              \multicolumn{1}{c}{NR}\\
+              \hline
+              Ra=\sn{1}{3} & 5 & 5 & 5\\
+              Ra=\sn{1}{4} & 7 & 7 & 7\\
+              Ra=\sn{1}{5} & 9 & 10 & 9\\
+              Ra=\sn{1}{6} & 11 & 11 & 11\\
+              %%
+              \hline\hline
+            \end{tabular}`
+          \end{center}
+          \caption{Nonlinear iterations.}
+        \end{table}
+
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lccc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}&
+              \multicolumn{1}{c}{GMRES}&
+              \multicolumn{1}{c}{MCSA}&
+              \multicolumn{1}{c}{Richardson}\\
+              \hline
+              Ra=\sn{1}{3} & 32 & 18 & 38\\
+              Ra=\sn{1}{4} & 23 & 17 & 34\\
+              Ra=\sn{1}{5} & 25 & 20 & 34\\
+              Ra=\sn{1}{6} & 39 & 25 & 48\\
+              %%
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Total linear solver iterations.}
+        \end{table}
+      }
+
+    \end{column}
+
+    \begin{column}{0.4\textwidth}
+
+      \small{
+        \begin{itemize}
+        \item Weaker preconditioning at Ra=\sn{1}{5} case
+          \medskip
+        \item Constant forcing term for Ra=\sn{1}{6} case
+        \end{itemize}
+      }
+
+      \bigskip
+
+      \tiny{
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}&
+              \multicolumn{1}{c}{NK Speedup}\\
+              \hline
+              Ra=\sn{1}{3} & 338 \\
+              Ra=\sn{1}{4} & 336 \\
+              Ra=\sn{1}{5} & 346 \\
+              Ra=\sn{1}{6} & 465 \\
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Newton-Krylov speedup over FANM.}
+        \end{table}
+      }
+
+    \end{column}
+  \end{columns}
+
 \end{frame}
@@
 
-  \vfill
-
   MC methods are used in many DOE application areas:
   \begin{itemize}
+    \setlength{\itemsep}{-0.2\baselineskip}
   \item high-energy density physics
@@
   \item nuclear engineering
-  \item particle accelerators and neutron scattering
+  \item particle accelerators and neutron spallation source
   \item quantum chemistry
@@
   \item fluid dynamics
+  \item combustion
   \item computational biology
   \end{itemize}
-
-  \vfill
 
@@
 
-\begin{frame}{Continued Research for MC Linear Solvers}
+\begin{frame}{ECP R\&D I: Time-dependent Problems}
 
@@
 
+\begin{frame}{ECP R\&D II: Monte Carlo Particle Transport}
+
+  MC particle transport has applications in many areas across the DOE complex
+  \begin{itemize}
+    \setlength{\itemsep}{-0.2\baselineskip}
+  \item high-energy density physics
+  \item neutron spallation source
+  \item nuclear engineering
+  \item astrophysics
+  \item combustion
+  \end{itemize}
+
+  \vfill
+
+  There are significant challenges in porting random walk algorithms to modern
+  heterogeneous computing devices
+  \begin{enumerate}
+  \item Random walks are dominated by branching statements $\rightarrow$
+    thread divergence
+  \item Runtime is often dominated by sampling and interpolating large fields
+    of data (keeping the GPUs fed)
+  \end{enumerate}
+
+  \vfill
+
+  Supporting research into the efficient parallelization and instrumentation
+  of random walk algorithms for particle transport will benefit many
+  applications in DOE
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{ECP R\&D III: Particle Methods}
+
+  Non-stochastic particle methods share a great deal of algorithmic features
+  with stochastic particle methods
+  \begin{itemize}
+  \item Particles must be ordered and sorted efficiently
+  \item Particles must be moved through a combinatorial and mesh-based
+    geometries
+  \item Data structures must be optimized for efficient vectorization (over
+    32-thread warps on GPUs)
+  \item Inter-node communication of particle data structures are required for
+    domain decomposition parallel topologies
+  \item Physical parameters must be mapped between particles and an underlying
+    data representation (grid)
+  \end{itemize}
+
+  \vfill
+
+  A supporting technolgies research effort on a generalized set of libraries
+  for managing particle-based data structures on heterogeneous computing
+  platforms will benefit many applications in DOE
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
 \begin{frame}{Building on ASCR Funding}
 
-  MCREX has leveraged and built on work derived from several ASCR projects:
-  \begin{itemize}
-  \item Trilinos
-  \item XPress
-  \item xSim
-  \item Ideas
-  \end{itemize}
+  \underline{\bf Trilinos}
+  \begin{itemize}
+  \item Provides abstract solver interface for MCLS
+  \item Provides build system for Alea and MCLS
+  \item Provides data structures and utilities for Alea and MCLS
+  \end{itemize}
+
+  \vfill
+
+  \underline{\bf xSIM}
+  \begin{itemize}
+  \item Simulator to test for hard error detection and remediation
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Building on ASCR Funding}
+
+  \underline{\bf XPRESS}
+  \begin{itemize}
+  \item We are currently implementing a CUDA transport kernel with scheduling
+    optimized by HPX as part of XPRESS
+  \item We have already compared OpenMP versus HPX for on-node MC particle
+    transport algorithms
+  \end{itemize}
+
+  \begin{figure}
+    \centering
+    \begin{tabular}{cc}
+      \subfigure[time]{\includegraphics[width=2in]{time}}
+      &
+      \subfigure[efficiency]{\includegraphics[width=2in]{tallies}}
+    \end{tabular}
+  \end{figure}
 

### 2016-01-26 doc/ecp/pres.tex
Updated presentation.

@@
   at Scale for Linear and Nonlinear Solver Applications}
-\author{Thomas Evans \and Steven Hamilton}
-\date[DOE, 2016]{DOE RX-Solvers \\ January 29, 2016}
-\institute[ORNL]{Oak Ridge National Lab}
+\author{Thomas Evans (PI, ORNL)\\
+  Michele Benzi (Co-PI, Emory)\\
+  Steven Hamilton (ORNL)\\
+  Stuart Slattery (ORNL)\\
+  Christian Engelmann (ORNL)\\
+  Wayne Joubert (ORNL)\\
+  Massimiliano Lupo Pasini (Emory)
+}
+\date[DOE, 2016]{DOE RX-Solvers, 1/29/2016}
+%\institute[ORNL]{Oak Ridge National Lab}
 
@@
 %%---------------------------------------------------------------------------%%
+\subsection{ECP R\&D}
+%%---------------------------------------------------------------------------%%
 
@@
 %%---------------------------------------------------------------------------%%
+\subsection{Building on ASCR Funding}
+%%---------------------------------------------------------------------------%%
 
@@
   \begin{itemize}
+    \setlength{\itemsep}{-0.1\baselineskip}
   \item We are currently implementing a CUDA transport kernel with scheduling
@@
     transport algorithms
-  \end{itemize}
-
     \begin{figure}
@@
       \begin{tabular}{cc}
-      \subfigure[time]{\includegraphics[width=2in]{time}}
+        \subfigure[time]{\includegraphics[width=1.8in]{time}}
         &
-      \subfigure[efficiency]{\includegraphics[width=2in]{tallies}}
+        \subfigure[efficiency]{\includegraphics[width=1.8in]{tallies}}
       \end{tabular}
     \end{figure}
+  \item All code implementations are available in the open source Profugus
+    mini-app (\underline{github.com/ORNL-CEES/Profugus})
+  \end{itemize}
 

### 2016-01-26 doc/ecp/pres.tex
Added scaling plots.

@@
 
+\begin{frame}{Replication}
+
+  \vspace{-0.1in}
+
+  Different batches of Monte Carlo samples can be combined in summation via
+  superposition if they have different random number streams.  Consider each
+  of these batches independent \textit{subsets} of a Monte Carlo operator
+  where now the operator can be formed as a general additive decomposition of
+  $N_S$ subsets:
+
+  \[
+  \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
+  \]
+
+  We replicate the linear problem and form each subset on a different group of
+  parallel processes. Applying the subsets to a vector requires an AllReduce
+  to form the sum. Each subset is domain decomposed.
+  \vfill
+  This provides a mechanism to support \textit{resiliency}; the replicated
+  component provides {\bf useful} work, not just redundancy
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
 \begin{frame}{Parallel Test - Simplified $P_N$ ($SP_N$) Assembly Problem}
@@
       \end{itemize}
+    \end{column}
+
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Scaling - DD}
+
+  \begin{figure}
+    \centering
+    \begin{tabular}{cc}
+      \subfigure[Strong time]{\includegraphics[width=1.8in]{strong_dd_time}} &
+      \subfigure[Strong efficiency]{\includegraphics[width=1.8in]{strong_dd_eff}} \\
+      %%
+      \subfigure[Weak time]{\includegraphics[width=1.8in]{weak_dd_time}} &
+      \subfigure[Weak efficiency]{\includegraphics[width=1.8in]{weak_dd_eff}} \\
+    \end{tabular}
+  \end{figure}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Scaling - Replication}
+
+  \begin{figure}
+    \centering
+    \begin{tabular}{cc}
+      \subfigure[Strong time]{\includegraphics[width=2in]{strong_rep_time}} &
+      \subfigure[Strong efficiency]{\includegraphics[width=2in]{strong_rep_eff}} \\
+      %%
+    \end{tabular}
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{GPU Optimization}
+
+  We have implemented a version of MCSA that uses GPUs to perform the MC
+  linear solve.
+
+  Distributed parallelism is enabled through an additive Schwartz method, all
+  MC solves
+  \begin{itemize}
+  \item All MC random walks are local
+  \item No inter-domain communication of histories
+  \end{itemize}
+
+  \begin{table}[h]
+    \begin{center}
+      \caption{2D parabolic problem with AINV preconditioning.}
+      \begin{tabular}{|c|c|c|c|}\hline
+        Node & Iterations & Time (sec) & Relative Residual \\\hline
+        CPU & 6 & 992.23 & $1.23\times 10^{-7}$ \\\hline
+        GPU & 6 & 29.83 & $1.53\times 10^{-7}$ \\\hline
+      \end{tabular}
+    \end{center}
+  \end{table}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\subsection{Research Requirements}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Continued Research is Necessary}
+
+  The following important questions need to be answered before MC methods can
+  be applied to linear systems as a ``standard'' solver:
+
+  \begin{itemize}
+  \item Automatically determine history termination criteria
+    \begin{itemize}
+    \item We have investigated adaptive schemes based on variance
+    \item However, variance is not directly correlated with error
+    \end{itemize}
+  \item Identify classes of matrices and preconditioners that satisfy
+    convergence constraints
+    \begin{itemize}
+    \item This has been done for SDD and GDD
+    \end{itemize}
+  \item Demonstrate resiliency to hard MPI errors through replicated domain
+    parallelism
+  \item Further investigate non-linear, time-dependent problems
+    \begin{itemize}
+    \item This was part of the original proposal
+    \item Non-linear, sub-cycle time-stepping can result in strongly diagonally
+      dominant problems for which MCSA has been shown to be effective
+    \end{itemize}
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\subsection{Software}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{MCSA Implementations}
+
+  {\bf MCLS} (\underline{github.com/sslattery/MCLS})
+  \begin{itemize}
+  \item Massively parallel (MPI) implementation of domain decomposed MCSA
+    algorithm
+  \item Implemented in Trilinos framework and callable through Trilinos'
+    abstract numerical algorithms (Thyra, Stratimikos) interface
+  \item Works with any Tpetra row matrix
+  \end{itemize}
+  \vfill
+  {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
+  \begin{itemize}
+  \item GPU-accelerated on-node MC solve
+  \item Domain-local MC solves implemented for distributed systems as an
+    additive Schwartz method
+  \end{itemize}
+  \vfill
+  All software is BSD open source.
+  \vfill
+  Uses TriBITS (Trilinos) open source build system
+  \vfill
+  Both software packages contain a complete suite of unit-tests integrated
+  through GoogleTest (Alea) or TeuchosTest (MCLS)
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Publications and Presentations}
+
+  \underline{Publications}
+  \begin{itemize}
+  \item M. Benzi, T.M. Evans, S.P. Hamilton, M.L. Pasini, and
+    S.R. Slattery. Analysis of Monte Carlo Accelerated Iterative Methods for
+    Sparse Linear Systems. \textit{in preparation}, 2016.
+  \item S.R. Slattery, T.M. Evans, P.P.H. Wilson. A Spectral Analysis of the
+    Domain Decomposed Monte Carlo Method for Linear Systems. \textit{Nuclear
+      Engineering Design},  {\bf 295}, 632--638, 2015.
+  \item T.M. Evans, S.W. Mosher, S.R. Slattery, and S.P. Hamilton. A Monte Carlo
+    Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion
+    Equation. \textit{J. Comp. Phys.}, {\bf 258}, 338--358, 2014.
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Publications and Presentations}
+
+  \underline{Presentations}
+  \begin{itemize}
+    \footnotesize
+    \setlength{\itemsep}{-0.1\baselineskip}
+  \item Monte Carlo Synthetic Acceleration Methods for Sparse Linear Systems
+    (M.L. Pasini, M. Benzi, T. Evans, S. Hamilton, S. Slattery), \textit{SIAM
+      ALA}, October 30, 2015.
+  \item Parallel Algorithms for the Monte Carlo Synthetic Acceleration Linear
+    Solver Method (S. Slattery, T. Evans, S. Hamilton), \textit{SIAM CSE},
+    Salt Lake City, 2015.
+  \item Iterative Performance of Monte Carlo Linear Solver Methods
+    (M.L. Pasini), \textit{SIAM CSE}, Salt Lake City, 2015.
+  \item A multilevel Monte Carlo method for linear systems (S. Slattery,
+    T. Evans, S. Hamilton), \textit{Copper Mountain Conference on Iterative
+      Solvers}, Copper MT, CO, 2014.
+  \item Monte Carlo Synthetic Acceleration as approximate polynomial
+    preconditioning (S. Hamilton, T. Evans, S. Slattery), \textit{Copper
+      Mountain Conference on Iterative Solvers}, Copper MT, CO, 2014.
+  \item Monte Carlo Linear Solvers (S. Slattery, T. Evans, S. Hamilton),
+    \textit{NC State Seminar}, Nov. 4, 2014.
+  \item Monte Carlo Linear Solvers (S. Hamilton, T. Evans, S. Slattery),
+    \textit{Emory Mathematics Dept. Seminar}, Sept. 19, 2014.
+  \item Monte Carlo Methods for Linear Systems (T. Evans), \textit{MIT Nuclear
+      Engineering Dept Seminar Series}, Sept. 30, 2014.
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\section{Transition and Roadmap}
+%%---------------------------------------------------------------------------%%
+\subsection{ECP Application Requirements}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Connection to ECP Requirements}
+
+  \begin{itemize}
+
+  \item Inter-Operability
+    \begin{itemize}
+    \item MCLS built using Trilinos solver API
+    \item Utilizes Hypre/ParaSAILS for AINV preconditioning
+    \item Enables testing of MC solver algorithms in a wide variety of
+      applications, ie. anyone who uses Trilinos
+    \end{itemize}
+
+  \item Scalability
+    \begin{itemize}
+    \item Multilevel inter-node domain decomposition methods show excellent
+      scalability
+    \item GPU-enabled on-node MC kernel yields measures speedups of 30 -
+      50$\times$
+    \end{itemize}
+
+  \item Resiliency
+    \begin{itemize}
+    \item Replicated subsets enable useful work to occur at the same time as
+      providing redundancy
+    \item Yields protection against hard errors
+    \end{itemize}
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Inter-Operatibility Example}
+
+  Using the MCLS interface to Trilinos, it is straightforward to implement
+  MCSA in other applications
+  \vfill
+  We implemented MCSA in the production CFD code, Drekar (SNL) using a
+  Newton-MCSA algorithm:
+  \begin{algorithm}[H]
+    \footnotesize
+    \begin{algorithmic}[1]
+      \STATE $k := 0$
+      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
+        ||\ve{F}(\ve{u}^{0})||$}
+      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
+      \COMMENT{Automatic differentiation - Panzer}
+      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
+      \COMMENT{Solve for the Newton correction with MCSA}
+      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
+      \STATE $k \leftarrow k+1$
+      \ENDWHILE
+    \end{algorithmic}
+    \caption{FANM}
+  \end{algorithm}
+  \vfill
+  Because MCLS was built with Trilinos abstract numerical algorithms this
+  integration was straightforward and we can test MCLS in a whole new problem
+  space
+  \vfill
+  We are also able to utilize all of Trilinos' preconditioners
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Thermal Convection Cavity Results}
+
+  \begin{columns}
+    \begin{column}{0.6\textwidth}
+
+      \small{
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lccc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}&
+              \multicolumn{1}{c}{NK}&
+              \multicolumn{1}{c}{FANM}&
+              \multicolumn{1}{c}{NR}\\
+              \hline
+              Ra=\sn{1}{3} & 5 & 5 & 5\\
+              Ra=\sn{1}{4} & 7 & 7 & 7\\
+              Ra=\sn{1}{5} & 9 & 10 & 9\\
+              Ra=\sn{1}{6} & 11 & 11 & 11\\
+              %%
+              \hline\hline
+            \end{tabular}`
+          \end{center}
+          \caption{Nonlinear iterations.}
+        \end{table}
+
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lccc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}&
+              \multicolumn{1}{c}{GMRES}&
+              \multicolumn{1}{c}{MCSA}&
+              \multicolumn{1}{c}{Richardson}\\
+              \hline
+              Ra=\sn{1}{3} & 32 & 18 & 38\\
+              Ra=\sn{1}{4} & 23 & 17 & 34\\
+              Ra=\sn{1}{5} & 25 & 20 & 34\\
+              Ra=\sn{1}{6} & 39 & 25 & 48\\
+              %%
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Total linear solver iterations.}
+        \end{table}
+      }
+
+    \end{column}
+
+    \begin{column}{0.4\textwidth}
+
+      \small{
+        \begin{itemize}
+        \item Weaker preconditioning at Ra=\sn{1}{5} case
+          \medskip
+        \item Constant forcing term for Ra=\sn{1}{6} case
+        \end{itemize}
+      }
+
+      \bigskip
+
+      \tiny{
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}&
+              \multicolumn{1}{c}{NK Speedup}\\
+              \hline
+              Ra=\sn{1}{3} & 338 \\
+              Ra=\sn{1}{4} & 336 \\
+              Ra=\sn{1}{5} & 346 \\
+              Ra=\sn{1}{6} & 465 \\
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Newton-Krylov speedup over FANM.}
+        \end{table}
+      }
+
+    \end{column}
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\subsection{ECP R\&D}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{ECP Coordinated Reasearch}
+
+  \underline{\bf Goal}: Extend the general algorithmic MC exascale methods
+  w.r.t. scalability and resiliency into a broader application space
+
+  MC methods are used in many DOE application areas:
+  \begin{itemize}
+    \setlength{\itemsep}{-0.2\baselineskip}
+  \item high-energy density physics
+  \item astrophysics
+  \item nuclear engineering
+  \item particle accelerators and neutron spallation source
+  \item quantum chemistry
+  \item visualization
+  \item fluid dynamics
+  \item combustion
+  \item computational biology
+  \end{itemize}
+
+  Much of the algorithmic research performed in MCREX is directly applicable
+  to the use of MC methods in these other application areas
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{ECP R\&D I: Time-dependent Problems}
+
+  Investigate use in time-dependent applications, e.g. non-linear radiation
+  hydrodynamics for high-energy density physics applications:
+  \begin{itemize}
+  \item Used as inner linear solver in a non-linear (Newton) iteration
+  \item Time-dependent applications generally have strong diagonal-dominance
+    that makes MCSA competitive with Krylov methods
+  \item MCSA has resiliency and parallel efficiency benefits over subspace
+    methods
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{ECP R\&D II: Monte Carlo Particle Transport}
+
+  MC particle transport has applications in many areas across the DOE complex
+  \begin{itemize}
+    \setlength{\itemsep}{-0.2\baselineskip}
+  \item high-energy density physics
+  \item neutron spallation source
+  \item nuclear engineering
+  \item astrophysics
+  \item combustion
+  \end{itemize}
+
+  \vfill
+
+  There are significant challenges in porting random walk algorithms to modern
+  heterogeneous computing devices
+  \begin{enumerate}
+  \item Random walks are dominated by branching statements $\rightarrow$
+    thread divergence
+  \item Runtime is often dominated by sampling and interpolating large fields
+    of data (keeping the GPUs fed)
+  \end{enumerate}
+
+  \vfill
+
+  Supporting research into the efficient parallelization and instrumentation
+  of random walk algorithms for particle transport will benefit many
+  applications in DOE
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{ECP R\&D III: Particle Methods}
+
+  Non-stochastic particle methods share a great deal of algorithmic features
+  with stochastic particle methods
+  \begin{itemize}
+  \item Particles must be ordered and sorted efficiently
+  \item Particles must be moved through a combinatorial and mesh-based
+    geometries
+  \item Data structures must be optimized for efficient vectorization (over
+    32-thread warps on GPUs)
+  \item Inter-node communication of particle data structures are required for
+    domain decomposition parallel topologies
+  \item Physical parameters must be mapped between particles and an underlying
+    data representation (grid)
+  \end{itemize}
+
+  \vfill
+
+  A supporting technolgies research effort on a generalized set of libraries
+  for managing particle-based data structures on heterogeneous computing
+  platforms will benefit many applications in DOE
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\subsection{Building on ASCR Funding}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Building on ASCR Funding}
+
+  \underline{\bf Trilinos}
+  \begin{itemize}
+  \item Provides abstract solver interface for MCLS
+  \item Provides build system for Alea and MCLS
+  \item Provides data structures and utilities for Alea and MCLS
+  \end{itemize}
+
+  \vfill
+
+  \underline{\bf xSIM}
+  \begin{itemize}
+  \item Simulator to test for hard error detection and remediation
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Building on ASCR Funding}
+
+  \underline{\bf XPRESS}
+  \begin{itemize}
+    \setlength{\itemsep}{-0.1\baselineskip}
+  \item We are currently implementing a CUDA transport kernel with scheduling
+    optimized by HPX as part of XPRESS
+  \item We have already compared OpenMP versus HPX for on-node MC particle
+    transport algorithms
+    \begin{figure}
+      \centering
+      \begin{tabular}{cc}
+        \subfigure[time]{\includegraphics[width=1.8in]{time}}
+        &
+        \subfigure[efficiency]{\includegraphics[width=1.8in]{tallies}}
+      \end{tabular}
+    \end{figure}
+  \item All code implementations are available in the open source Profugus
+    mini-app (\underline{github.com/ORNL-CEES/Profugus})
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}
+\vfill
+\centering
+\huge Questions?
+\vfill
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\section*{Extra Slides}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Preconditioned MCSA}
+\vspace*{-0.2in}
+  {
+    \small
+    \[
+    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k})
+    \]
+    \[
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
+    \]
+    \[
+    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k+1/2})
+    \]
+    \[
+    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
+    \]
+    \[
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
+    \]
+  }
+  \vspace*{-0.1in}
+  \begin{itemize}
+    \item Requires explicit construction of preconditioned matrix
+      \begin{itemize}
+        \item This has generally limited preconditioner selection to diagonal,
+          block diagonal, sparse approximate inverse approaches
+      \end{itemize}
+    \vfill
+    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
+      in $\mathbf{M^{-1}A}$
+    \vfill
+    \item Need to investigate preconditioning strategies that lead to sparse
+      preconditioned systems
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Asynchronous Monte Carlo Transport Kernel}
+
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+      \vspace{-0.1in}
+      \small
+      \begin{itemize}
+      \item General extension of the Milagro algorithm (LANL)
+      \item Asynchronous nearest neighbor communication of histories
+      \item System-tunable communication parameters of buffer size and check
+        frequency (performance impact)
+      \item Need an asynchronous strategy for exiting the transport loop
+        without a collective (running sum)
+      \end{itemize}
+
+      \vspace{-0.15in}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.4}{ \input{domain_to_domain.pdftex_t} }
+        \end{center}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+
+      \vspace{-0.25in}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.825}{ \input{transport_loop.pdftex_t} }
+        \end{center}
+      \end{figure}
+
     \end{column}
@@
 
-\begin{frame}{Replication}
-
-  \vspace{-0.1in}
-
-  Different batches of Monte Carlo samples can be combined in summation via
-  superposition if they have different random number streams.  Consider each
-  of these batches independent \textit{subsets} of a Monte Carlo operator
-  where now the operator can be formed as a general additive decomposition of
-  $N_S$ subsets:
-
-  \[
-  \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
-  \]
-
-  We replicate the linear problem and form each subset on a different group of
-  parallel processes. Applying the subsets to a vector requires an AllReduce
-  to form the sum. Each subset is domain decomposed.
-  \vfill
-  This provides a mechanism to support \textit{resiliency}; the replicated
-  component provides {\bf useful} work, not just redundancy
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
 \begin{frame}{Replication Scaling}
@@
 %%---------------------------------------------------------------------------%%
-
-\begin{frame}{GPU Optimization}
-
-  We have implemented a version of MCSA that uses GPUs to perform the MC
-  linear solve.
-
-  Distributed parallelism is enabled through an additive Schwartz method, all
-  MC solves
-  \begin{itemize}
-  \item All MC random walks are local
-  \item No inter-domain communication of histories
-  \end{itemize}
-
-  \begin{table}[h]
-    \begin{center}
-      \caption{2D parabolic problem with AINV preconditioning.}
-      \begin{tabular}{|c|c|c|c|}\hline
-        Node & Iterations & Time (sec) & Relative Residual \\\hline
-        CPU & 6 & 992.23 & $1.23\times 10^{-7}$ \\\hline
-        GPU & 6 & 29.83 & $1.53\times 10^{-7}$ \\\hline
-      \end{tabular}
-    \end{center}
-  \end{table}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\subsection{Research Requirements}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Continued Research is Necessary}
-
-  The following important questions need to be answered before MC methods can
-  be applied to linear systems as a ``standard'' solver:
-
-  \begin{itemize}
-  \item Automatically determine history termination criteria
-    \begin{itemize}
-    \item We have investigated adaptive schemes based on variance
-    \item However, variance is not directly correlated with error
-    \end{itemize}
-  \item Identify classes of matrices and preconditioners that satisfy
-    convergence constraints
-    \begin{itemize}
-    \item This has been done for SDD and GDD
-    \end{itemize}
-  \item Demonstrate resiliency to hard MPI errors through replicated domain
-    parallelism
-  \item Further investigate non-linear, time-dependent problems
-    \begin{itemize}
-    \item This was part of the original proposal
-    \item Non-linear, sub-cycle time-stepping can result in strongly diagonally
-      dominant problems for which MCSA has been shown to be effective
-    \end{itemize}
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\subsection{Software}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{MCSA Implementations}
-
-  {\bf MCLS} (\underline{github.com/sslattery/MCLS})
-  \begin{itemize}
-  \item Massively parallel (MPI) implementation of domain decomposed MCSA
-    algorithm
-  \item Implemented in Trilinos framework and callable through Trilinos'
-    abstract numerical algorithms (Thyra, Stratimikos) interface
-  \item Works with any Tpetra row matrix
-  \end{itemize}
-  \vfill
-  {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
-  \begin{itemize}
-  \item GPU-accelerated on-node MC solve
-  \item Domain-local MC solves implemented for distributed systems as an
-    additive Schwartz method
-  \end{itemize}
-  \vfill
-  All software is BSD open source.
-  \vfill
-  Uses TriBITS (Trilinos) open source build system
-  \vfill
-  Both software packages contain a complete suite of unit-tests integrated
-  through GoogleTest (Alea) or TeuchosTest (MCLS)
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Publications and Presentations}
-
-  \underline{Publications}
-  \begin{itemize}
-  \item M. Benzi, T.M. Evans, S.P. Hamilton, M.L. Pasini, and
-    S.R. Slattery. Analysis of Monte Carlo Accelerated Iterative Methods for
-    Sparse Linear Systems. \textit{in preparation}, 2016.
-  \item S.R. Slattery, T.M. Evans, P.P.H. Wilson. A Spectral Analysis of the
-    Domain Decomposed Monte Carlo Method for Linear Systems. \textit{Nuclear
-      Engineering Design},  {\bf 295}, 632--638, 2015.
-  \item T.M. Evans, S.W. Mosher, S.R. Slattery, and S.P. Hamilton. A Monte Carlo
-    Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion
-    Equation. \textit{J. Comp. Phys.}, {\bf 258}, 338--358, 2014.
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Publications and Presentations}
-
-  \underline{Presentations}
-  \begin{itemize}
-    \footnotesize
-    \setlength{\itemsep}{-0.1\baselineskip}
-  \item Monte Carlo Synthetic Acceleration Methods for Sparse Linear Systems
-    (M.L. Pasini, M. Benzi, T. Evans, S. Hamilton, S. Slattery), \textit{SIAM
-      ALA}, October 30, 2015.
-  \item Parallel Algorithms for the Monte Carlo Synthetic Acceleration Linear
-    Solver Method (S. Slattery, T. Evans, S. Hamilton), \textit{SIAM CSE},
-    Salt Lake City, 2015.
-  \item Iterative Performance of Monte Carlo Linear Solver Methods
-    (M.L. Pasini), \textit{SIAM CSE}, Salt Lake City, 2015.
-  \item A multilevel Monte Carlo method for linear systems (S. Slattery,
-    T. Evans, S. Hamilton), \textit{Copper Mountain Conference on Iterative
-      Solvers}, Copper MT, CO, 2014.
-  \item Monte Carlo Synthetic Acceleration as approximate polynomial
-    preconditioning (S. Hamilton, T. Evans, S. Slattery), \textit{Copper
-      Mountain Conference on Iterative Solvers}, Copper MT, CO, 2014.
-  \item Monte Carlo Linear Solvers (S. Slattery, T. Evans, S. Hamilton),
-    \textit{NC State Seminar}, Nov. 4, 2014.
-  \item Monte Carlo Linear Solvers (S. Hamilton, T. Evans, S. Slattery),
-    \textit{Emory Mathematics Dept. Seminar}, Sept. 19, 2014.
-  \item Monte Carlo Methods for Linear Systems (T. Evans), \textit{MIT Nuclear
-      Engineering Dept Seminar Series}, Sept. 30, 2014.
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\section{Transition and Roadmap}
-%%---------------------------------------------------------------------------%%
-\subsection{ECP Application Requirements}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Connection to ECP Requirements}
-
-  \begin{itemize}
-
-  \item Inter-Operability
-    \begin{itemize}
-    \item MCLS built using Trilinos solver API
-    \item Utilizes Hypre/ParaSAILS for AINV preconditioning
-    \item Enables testing of MC solver algorithms in a wide variety of
-      applications, ie. anyone who uses Trilinos
-    \end{itemize}
-
-  \item Scalability
-    \begin{itemize}
-    \item Multilevel inter-node domain decomposition methods show excellent
-      scalability
-    \item GPU-enabled on-node MC kernel yields measures speedups of 30 -
-      50$\times$
-    \end{itemize}
-
-  \item Resiliency
-    \begin{itemize}
-    \item Replicated subsets enable useful work to occur at the same time as
-      providing redundancy
-    \item Yields protection against hard errors
-    \end{itemize}
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Inter-Operatibility Example}
-
-  Using the MCLS interface to Trilinos, it is straightforward to implement
-  MCSA in other applications
-  \vfill
-  We implemented MCSA in the production CFD code, Drekar (SNL) using a
-  Newton-MCSA algorithm:
-  \begin{algorithm}[H]
-    \footnotesize
-    \begin{algorithmic}[1]
-      \STATE $k := 0$
-      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
-        ||\ve{F}(\ve{u}^{0})||$}
-      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
-      \COMMENT{Automatic differentiation - Panzer}
-      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
-      \COMMENT{Solve for the Newton correction with MCSA}
-      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
-      \STATE $k \leftarrow k+1$
-      \ENDWHILE
-    \end{algorithmic}
-    \caption{FANM}
-  \end{algorithm}
-  \vfill
-  Because MCLS was built with Trilinos abstract numerical algorithms this
-  integration was straightforward and we can test MCLS in a whole new problem
-  space
-  \vfill
-  We are also able to utilize all of Trilinos' preconditioners
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Thermal Convection Cavity Results}
-
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-
-      \small{
-        \begin{table}[h!]
-          \begin{center}
-            \begin{tabular}{lccc}\hline\hline
-              \multicolumn{1}{l}{Benchmark}&
-              \multicolumn{1}{c}{NK}&
-              \multicolumn{1}{c}{FANM}&
-              \multicolumn{1}{c}{NR}\\
-              \hline
-              Ra=\sn{1}{3} & 5 & 5 & 5\\
-              Ra=\sn{1}{4} & 7 & 7 & 7\\
-              Ra=\sn{1}{5} & 9 & 10 & 9\\
-              Ra=\sn{1}{6} & 11 & 11 & 11\\
-              %%
-              \hline\hline
-            \end{tabular}`
-          \end{center}
-          \caption{Nonlinear iterations.}
-        \end{table}
-
-        \begin{table}[h!]
-          \begin{center}
-            \begin{tabular}{lccc}\hline\hline
-              \multicolumn{1}{l}{Benchmark}&
-              \multicolumn{1}{c}{GMRES}&
-              \multicolumn{1}{c}{MCSA}&
-              \multicolumn{1}{c}{Richardson}\\
-              \hline
-              Ra=\sn{1}{3} & 32 & 18 & 38\\
-              Ra=\sn{1}{4} & 23 & 17 & 34\\
-              Ra=\sn{1}{5} & 25 & 20 & 34\\
-              Ra=\sn{1}{6} & 39 & 25 & 48\\
-              %%
-              \hline\hline
-            \end{tabular}
-          \end{center}
-          \caption{Total linear solver iterations.}
-        \end{table}
-      }
-
-    \end{column}
-
-    \begin{column}{0.4\textwidth}
-
-      \small{
-        \begin{itemize}
-        \item Weaker preconditioning at Ra=\sn{1}{5} case
-          \medskip
-        \item Constant forcing term for Ra=\sn{1}{6} case
-        \end{itemize}
-      }
-
-      \bigskip
-
-      \tiny{
-        \begin{table}[h!]
-          \begin{center}
-            \begin{tabular}{lc}\hline\hline
-              \multicolumn{1}{l}{Benchmark}&
-              \multicolumn{1}{c}{NK Speedup}\\
-              \hline
-              Ra=\sn{1}{3} & 338 \\
-              Ra=\sn{1}{4} & 336 \\
-              Ra=\sn{1}{5} & 346 \\
-              Ra=\sn{1}{6} & 465 \\
-              \hline\hline
-            \end{tabular}
-          \end{center}
-          \caption{Newton-Krylov speedup over FANM.}
-        \end{table}
-      }
-
-    \end{column}
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\subsection{ECP R\&D}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{ECP Coordinated Reasearch}
-
-  \underline{\bf Goal}: Extend the general algorithmic MC exascale methods
-  w.r.t. scalability and resiliency into a broader application space
-
-  MC methods are used in many DOE application areas:
-  \begin{itemize}
-    \setlength{\itemsep}{-0.2\baselineskip}
-  \item high-energy density physics
-  \item astrophysics
-  \item nuclear engineering
-  \item particle accelerators and neutron spallation source
-  \item quantum chemistry
-  \item visualization
-  \item fluid dynamics
-  \item combustion
-  \item computational biology
-  \end{itemize}
-
-  Much of the algorithmic research performed in MCREX is directly applicable
-  to the use of MC methods in these other application areas
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{ECP R\&D I: Time-dependent Problems}
-
-  Investigate use in time-dependent applications, e.g. non-linear radiation
-  hydrodynamics for high-energy density physics applications:
-  \begin{itemize}
-  \item Used as inner linear solver in a non-linear (Newton) iteration
-  \item Time-dependent applications generally have strong diagonal-dominance
-    that makes MCSA competitive with Krylov methods
-  \item MCSA has resiliency and parallel efficiency benefits over subspace
-    methods
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{ECP R\&D II: Monte Carlo Particle Transport}
-
-  MC particle transport has applications in many areas across the DOE complex
-  \begin{itemize}
-    \setlength{\itemsep}{-0.2\baselineskip}
-  \item high-energy density physics
-  \item neutron spallation source
-  \item nuclear engineering
-  \item astrophysics
-  \item combustion
-  \end{itemize}
-
-  \vfill
-
-  There are significant challenges in porting random walk algorithms to modern
-  heterogeneous computing devices
-  \begin{enumerate}
-  \item Random walks are dominated by branching statements $\rightarrow$
-    thread divergence
-  \item Runtime is often dominated by sampling and interpolating large fields
-    of data (keeping the GPUs fed)
-  \end{enumerate}
-
-  \vfill
-
-  Supporting research into the efficient parallelization and instrumentation
-  of random walk algorithms for particle transport will benefit many
-  applications in DOE
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{ECP R\&D III: Particle Methods}
-
-  Non-stochastic particle methods share a great deal of algorithmic features
-  with stochastic particle methods
-  \begin{itemize}
-  \item Particles must be ordered and sorted efficiently
-  \item Particles must be moved through a combinatorial and mesh-based
-    geometries
-  \item Data structures must be optimized for efficient vectorization (over
-    32-thread warps on GPUs)
-  \item Inter-node communication of particle data structures are required for
-    domain decomposition parallel topologies
-  \item Physical parameters must be mapped between particles and an underlying
-    data representation (grid)
-  \end{itemize}
-
-  \vfill
-
-  A supporting technolgies research effort on a generalized set of libraries
-  for managing particle-based data structures on heterogeneous computing
-  platforms will benefit many applications in DOE
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\subsection{Building on ASCR Funding}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Building on ASCR Funding}
-
-  \underline{\bf Trilinos}
-  \begin{itemize}
-  \item Provides abstract solver interface for MCLS
-  \item Provides build system for Alea and MCLS
-  \item Provides data structures and utilities for Alea and MCLS
-  \end{itemize}
-
-  \vfill
-
-  \underline{\bf xSIM}
-  \begin{itemize}
-  \item Simulator to test for hard error detection and remediation
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Building on ASCR Funding}
-
-  \underline{\bf XPRESS}
-  \begin{itemize}
-    \setlength{\itemsep}{-0.1\baselineskip}
-  \item We are currently implementing a CUDA transport kernel with scheduling
-    optimized by HPX as part of XPRESS
-  \item We have already compared OpenMP versus HPX for on-node MC particle
-    transport algorithms
-    \begin{figure}
-      \centering
-      \begin{tabular}{cc}
-        \subfigure[time]{\includegraphics[width=1.8in]{time}}
-        &
-        \subfigure[efficiency]{\includegraphics[width=1.8in]{tallies}}
-      \end{tabular}
-    \end{figure}
-  \item All code implementations are available in the open source Profugus
-    mini-app (\underline{github.com/ORNL-CEES/Profugus})
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}
-\vfill
-\centering
-\huge Questions?
-\vfill
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\section*{Extra Slides}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Preconditioned MCSA}
-\vspace*{-0.2in}
-  {
-    \small
-    \[
-    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k})
-    \]
-    \[
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
-    \]
-    \[
-    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k+1/2})
-    \]
-    \[
-    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
-    \]
-    \[
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
-    \]
-  }
-  \vspace*{-0.1in}
-  \begin{itemize}
-    \item Requires explicit construction of preconditioned matrix
-      \begin{itemize}
-        \item This has generally limited preconditioner selection to diagonal,
-          block diagonal, sparse approximate inverse approaches
-      \end{itemize}
-    \vfill
-    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
-      in $\mathbf{M^{-1}A}$
-    \vfill
-    \item Need to investigate preconditioning strategies that lead to sparse
-      preconditioned systems
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Asynchronous Monte Carlo Transport Kernel}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-      \vspace{-0.1in}
-      \small
-      \begin{itemize}
-      \item General extension of the Milagro algorithm (LANL)
-      \item Asynchronous nearest neighbor communication of histories
-      \item System-tunable communication parameters of buffer size and check
-        frequency (performance impact)
-      \item Need an asynchronous strategy for exiting the transport loop
-        without a collective (running sum)
-      \end{itemize}
-
-      \vspace{-0.15in}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.4}{ \input{domain_to_domain.pdftex_t} }
-        \end{center}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-      \vspace{-0.25in}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.825}{ \input{transport_loop.pdftex_t} }
-        \end{center}
-      \end{figure}
-
-    \end{column}
-
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \end{document}

### 2016-01-26 doc/ecp/generate_scaling.py

@@ -1,0 +1,93 @@
+###############################################################################
+# File  : ecp/generate_scaling.py
+# Author: Thomas M. Evans
+# Date  : Mon Jan 25 22:55:47 2016
+###############################################################################
+from __future__ import (division, absolute_import, print_function, )
+#-----------------------------------------------------------------------------#
+import numpy as np
+import matplotlib.pyplot as plt
+
+import sys
+sys.path.append('/codes/Exnihilo/Exnihilo/python')
+
+import exnihilotools.matplotlib as ext
+ext.screen_style()
+###############################################################################
+
+def main():
+
+    str_cores = [256, 1024, 4096, 7744, 10816]
+    str_time  = [515.52, 122.76, 27.96, 17.72, 13.72]
+    str_eff   = [1.00, 1.05, 1.15, 0.96, 0.89]
+
+    wk_cores = [64, 256, 1024, 4096, 7744, 10816]
+    wk_time  = [12.65, 12.86, 14.59, 14.25, 14.75, 13.72]
+    wk_eff   = [1.00, 0.98, 0.87, 0.89, 0.86, 0.92]
+
+    # Strong Scaling - time
+    plt.loglog(str_cores, str_time, 'o-')
+    plt.grid()
+    plt.xlabel("cores")
+    plt.ylabel("$t$ (sec)")
+    plt.savefig('strong_dd_time.pdf', bbox_inches='tight')
+    plt.close()
+
+    # Strong Scaling - efficiency
+    plt.plot(str_cores, str_eff, 'o-')
+    plt.grid()
+    plt.xlabel("cores")
+    plt.ylabel("$\epsilon$")
+    plt.ylim(0.5, 1.5)
+    plt.savefig('strong_dd_eff.pdf', bbox_inches='tight')
+    plt.close()
+
+    # Weak Scaling - time
+    plt.plot(wk_cores, wk_time, 'o-')
+    plt.grid()
+    plt.xlabel("cores")
+    plt.ylabel("$t$ (sec)")
+    plt.ylim(6,18)
+    plt.savefig('weak_dd_time.pdf', bbox_inches='tight')
+    plt.close()
+
+    # Weak Scaling - efficiency
+    plt.plot(wk_cores, wk_eff, 'o-')
+    plt.grid()
+    plt.xlabel("cores")
+    plt.ylabel("$\epsilon$")
+    plt.ylim(0.5, 1.5)
+    plt.savefig('weak_dd_eff.pdf', bbox_inches='tight')
+    plt.close()
+
[…]

### 2016-01-26 doc/ecp/pres.tex
Modifying polynomial MCSA slides.

@@
       a fixed length Neumann series polynomial
-      \[
-      x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-      \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}
-      \]
-    \vfill
-    \item As number of histories grows, iteration count becomes identical to
-      using ``true'' polynomial
     \vfill
@@
   \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Neumann Series Polynomial}
+
   \begin{table}
-    \caption{Adjoint MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
-      Values are MCSA iteration counts (timing in milliseconds)
-      \label{tab:lap_adjoint_neumann}}
+    \caption{Polynomial MCSA - Shifted Laplacian}
     \centering
@@
       \toprule
-      Histories per & \multicolumn{3}{c}{Polynomial Order} \\
-      \cmidrule(lr){2-4}
-      Iteration & 2 & 4 & 6 \\
+      Polynomial & Polynomial & MCSA & Histories per \\
+      Type & Order & Iterations & Iteration \\
       \midrule
-      250 &  775(100) & 651(96) & 664(110) \\
-      500 &  725(122) & 502(104) & 394(95) \\
-      1000 & 707(174) & 482(179) & 366(144) \\
-      2000 & 703(280) & 471(259) & 356(251) \\
-      4000 & 698(494) & 467(497) & 350(458) \\
-      8000 & 697(923) & 464(905) & 350(873) \\
-      16000 & 695(1796) & 462(1768) & 347(1711) \\
-      \bottomrule
-    \end{tabular}
-  \end{table}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Chebyshev Polynomial}
-  \begin{table}
-    \caption{Adjoint MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
-      Values are MCSA iteration counts (timing in milliseconds)
-      \label{tab:lap_adjoint_cheby}}
-    \centering
-    \begin{tabular}{cccc}
-      \toprule
-      Histories per & \multicolumn{3}{c}{Polynomial Order} \\
-      \cmidrule(lr){2-4}
-      Iteration & 1 & 2 & 3 \\
-      \midrule
-      250 & - & - & - \\
-      500 & - & - & - \\
-      1000 & - & - & - \\
-      2000 & - & 328(134) & - \\
-      4000 & - & 296(210) & - \\
-      8000 & - & 288(380) & 262(423) \\
-      16000 & 1132(1866) & 283(721) & 175(550) \\
+      Neumann & 4 & 502 & 500 \\
+      Chebyshev & 4 & 246 & 16000 \\
+      GMRES & 4 & 260 & 2000 \\
       \bottomrule

### 2016-01-26 doc/ecp/pres.tex
Changing CLF to CLT.
stuart: updating ecb slides

--- 2016-01-26 doc/ecp/pres.tex
@@
 
-  \item Much of our analysis work has used and idealized
+  \item Much of our analysis work has used an idealized
     advection-diffusion-reaction system that has properties common to these
@@
 
-\begin{frame}{Scaling - DD}
+\begin{frame}{Scaling - Domain Decomposition}
 
@@
 
-\begin{frame}{Scaling - Replication}
+\begin{frame}{Scaling - Domain Replication}
+
+  \vspace{-0.3in}
 
@@
 
+  \vfill
+  \begin{itemize}
+  \item A single subset consisted of a domain decomposed problem on 256 cores
+  \item Each additional subset replicates the original domain decomposed
+    subset creating a heirarchy of replication and decomposition
+  \end{itemize}
+
 \end{frame}
@@
   linear solve.
-
+  \vfill
   Distributed parallelism is enabled through an additive Schwartz method, all
@@
       providing redundancy
-    \item Yields protection against hard errors
+    \item Yields algorithmic protection against hard errors
     \end{itemize}
@@
 
+  \vfill
+
   MC methods are used in many DOE application areas:
@@
     that makes MCSA competitive with Krylov methods
-  \item MCSA has resiliency and parallel efficiency benefits over subspace
-    methods
+  \item MCSA has resiliency and parallel efficiency benefits over conventional
+    subspace methods
   \end{itemize}
@@
   \item Provides data structures and utilities for Alea and MCLS
+  \item Provides testing harness for MCLS
   \end{itemize}

### 2016-01-26 doc/ecp/pres.tex

--- 2016-01-26 doc/ecp/pres.tex
@@
   \begin{itemize}
-  \item Address issue of convergence constrained by CLF
+  \item Address issue of convergence constrained by CLT
     \vfill

### 2016-01-26 doc/ecp/pres.tex
Presentation update.

@@
 \usepackage{multirow}
-\usepackage{subfigure}
+\usepackage{subcaption}
 \usepackage[compatibility=false]{caption}
@@
 
+% Phantom minus sign for helping with alignment of matrices
+\newcommand{\phmin}{\ensuremath{\phantom{-}}}
+
 %%---------------------------------------------------------------------------%%
@@
  The end goal is to develop a set of resilient, scalable algorithms for the
- solution of linear systems.  We have made significant progress in each of
- these areas as summarized below.
+ solution of linear systems.
+
+ \vfill
+
+ \underline{Budget}: 300K/year (ORNL), 70K/year (Emory)
 
@@
 %%---------------------------------------------------------------------------%%
-\subsection{Overview and Status}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Summary}
-
-  Much of the first 2 years of the project were attempting to resolve
+
+\begin{frame}{Project Achievements}
+
+  Much of the first 2 years of the project was spent resolving
   \textit{algorithmic} challenges related to solver performance and efficiency
@@
   We have developed two open source software packages that contain
-  production-quality versions of these solvers (MPI and GPU).
-
-\end{frame}
+  parallel, production-quality versions of these solvers (MPI and GPU).
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\subsection{Overview and Status}
+%%---------------------------------------------------------------------------%%
 
 \begin{frame}{Monte Carlo for Linear Systems}
+
+  First reference, G.E. Forsythe and R.A. Leibler. Matrix Inverstion by a
+  Monte Carlo Method. \textit{Math. Tables and Comp.}, {\bf 4}(31), 127--129,
+  1950 (credited to von Neumann and Ulam, 1940s).
+  \vfill
   \begin{itemize}
@@
   \end{itemize}
-  \vspace*{-0.1in}
   \[
@@
 
-\begin{frame}{Limitations of Monte Carlo}
+\begin{frame}{Limitations of Monte Carlo---Efficiency}
   \begin{itemize}
@@
     is generally intractable
-    \begin{itemize}
   \item Central limit theorem is barrier to accurate solutions
-    \end{itemize}
     \begin{figure}
       \centering
-      \includegraphics[width=1.75in]{../NCState_seminar_2014/Ifiss_ConvDiff}
+      \includegraphics[width=3in]{../NCState_seminar_2014/Ifiss_ConvDiff}
     \end{figure}
-  \item There is a potentially restrictive limit on the spectral properties of
-    $\mathbf{H}$
-    \begin{equation*}
-      \rho(\mathbf{H}) < 1\:, \rho(\hat{\mathbf{H}}) < 1\:, \hat{H}_{ij} =
-      \frac{H_{ij}^2}{P_{ij}}
-    \end{equation*}
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Monte Carlo Synthetic Acceleration}
-  \begin{itemize}
-  \item Address issue of convergence constrained by CLT
-    \vfill
-  \item Can be abstracted as a general linear solver
-    \vfill
-  \item Combine with Richardson iteration as a ``smoother'' in between
-    Monte Carlo steps:
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Limitations of Monte Carlo---Convergence}
+
+  \begin{itemize}
+  \item Because we are using MC to estimate terms in the Neumann series, the
+    spectral radius of the iteration matrix must be
+    \[
+    \rho(\mathbf{H}) < 1
+    \]
+
+  \item Additionally, because we are estimating these terms stochastically,
+    the relationship of the variance of the estimates to the error imposes an
+    additional restriction 
+    \[
+    \rho(\hat{\mathbf{H}}) < 1\:, 
+    \]
+    where 
+    \[
+    \hat{H}_{ij} = \frac{H_{ij}^2}{P_{ij}}
+    \]
+    and the probabillities come from the decomposition of $\mathbf{H}$
+    \[
+    \mathbf{H} = \mathbf{P}\circ\mathbf{W}
+    \]
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Can Monte Carlo Convergence be Improved?}
+
+  \underline{Monte Carlo Synthetic Acceleration}
+  \vfill
+  \begin{itemize}
+  \item Building on Halton's Sequential Monte Carlo, we obtain Monte Carlo as
+    a corrective term as opposed to a direct solver to achieve exponential
+    convergence
+  \item Form iterative method in which Richardson iteration is a ``smoother''
+    in between Monte Carlo steps:
     \begin{align*}
-      \mathbf{r}^k &= \mathbf{b} - \mathbf{Ax}^k \\
-      \mathbf{x}^{k+1/2} &= \mathbf{x}^k + \mathbf{r}^k \\
-      \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} \\
-      \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k+1/2} \\
-      \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta
+      \mathbf{x}^{k+1/2} &= (\ve{I} - \ve{A})\mathbf{x}^k + \ve{b} 
+      && \text{Richardson Iteration}\\
+      %% 
+      \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} && 
+      \text{compute residual}\\
+      %% 
+      \hat{\mathbf{A}} \mathbf{\delta}_{MC} &= \mathbf{r}^{k+1/2}
+      && \text{solve using MC}\\
+      %%
+      \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta_{MC} 
+      && \text{update} 
+      %% 
     \end{align*}
-  \item We still must address issue of convergence constraint by spectral radius
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Monte Carlo as Approximate Polynomial Preconditioning}
-  \begin{itemize}
-    \item Using history length cutoff, Monte Carlo process is approximating
-      a fixed length Neumann series polynomial
-    \vfill
-    \item Why limit ourselves to the Neumann series polynomial?
+  \item This scheme is still constrained by the spectral limits on $\ve{H}$
+  \end{itemize}
+  \vfill\centering
+  \fbox{MCSA is a general linear solver}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Why limit to the Neumann series polynomial?}
   \begin{itemize}
   \item Chebyshev or GMRES polynomials are viable alternatives
-      \end{itemize}
-  \end{itemize}
+  \vfill
+  \item A good approximation to a few terms in the polynomial
+    performs better than a statistically noisy approximation to
+    full series
+  \vfill
+  \item Using history length cutoff rather than weight cutoff can lead to
+    large reductions in iteration counts
+  \vfill
+  \item Significant reduction in iteration counts are possible using
+    alternate polynomials, but partially offset by increase in number
+    of histories required
+  \end{itemize}
+  \vfill
 
@@
   \end{table}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Polynomial Methods -- Summary}
-  \begin{itemize}
-    \item Using history length cutoff rather than weight cutoff can lead to
-      large reductions in iteration counts
-      \vfill
-    \item A good approximation to a few terms in the Neumann series
-      performs better than a statistically noisy approximation to
-      full series
-      \vfill
-    \item Significant reduction in iteration counts are possible using
-      alternate polynomials, but partially offset by increase in number
-      of histories required
-      \begin{itemize}
-        \item May be very beneficial from resiliency and parallel efficiency
-          standpoints
-        \item Will be re-evaluated in future efforts
-      \end{itemize}
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Preconditioning}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Enabling Convergence through Preconditioning}
   \begin{itemize}
@@
 
-\begin{frame}{Domain Decomposed Monte Carlo}
+\begin{frame}{Monte Carlo Scalability---Domain Decomposed Monte Carlo}
 
@@
 
-\begin{frame}{Replication}
-
-  \vspace{-0.1in}
-
-  Different batches of Monte Carlo samples can be combined in summation via
-  superposition if they have different random number streams.  Consider each
-  of these batches independent \textit{subsets} of a Monte Carlo operator
-  where now the operator can be formed as a general additive decomposition of
-  $N_S$ subsets:
-
+\begin{frame}{Combining History Decomposition with Domain Decomposition}
+
+  \begin{itemize}
+  \item Different batches of Monte Carlo samples can be combined in summation
+    via superposition if they have different random number streams
+    \vfill
+  \item The new operator can be formed as a general additive
+    decomposition of $N_S$ subsets:
     \[
@@
     \]
-
-  We replicate the linear problem and form each subset on a different group of
-  parallel processes. Applying the subsets to a vector requires an AllReduce
-  to form the sum. Each subset is domain decomposed.
-  \vfill
-  This provides a mechanism to support \textit{resiliency}; the replicated
-  component provides {\bf useful} work, not just redundancy
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Parallel Test - Simplified $P_N$ ($SP_N$) Assembly Problem}
+    \vfill
+  \item We replicate the linear problem and form each subset on a different
+    group of parallel processes.  Each subset is domain decomposed.  
+    \vfill
+  \item This provides a mechanism to support \textit{resiliency}; the
+    replicated component provides {\bf useful} work, not just redundancy
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Application to Nuclear Reactor Analysis}
 \small
@@
 
-\begin{frame}{Scaling - Domain Decomposition}
+\begin{frame}{Scaling---Domain Decomposition}
 
@@
     \centering
-    \begin{tabular}{cc}
-      \subfigure[Strong time]{\includegraphics[width=1.8in]{strong_dd_time}} &
-      \subfigure[Strong efficiency]{\includegraphics[width=1.8in]{strong_dd_eff}} \\
+    \begin{subfigure}{0.35\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{strong_dd_time}
+      \caption{Strong scaling time}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.35\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{strong_dd_eff}
+      \caption{Strong scaling efficiency}
+    \end{subfigure}\\
     %%
-      \subfigure[Weak time]{\includegraphics[width=1.8in]{weak_dd_time}} &
-      \subfigure[Weak efficiency]{\includegraphics[width=1.8in]{weak_dd_eff}} \\
-    \end{tabular}
+    \begin{subfigure}{0.35\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{weak_dd_time}
+      \caption{Weak scaling time}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.35\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{weak_dd_eff}
+      \caption{Weak scaling efficiency}
+    \end{subfigure}
   \end{figure}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Scaling - Domain Replication}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Scaling---History Replication + Domain Decomposition}
 
@@
     \centering
-    \begin{tabular}{cc}
-      \subfigure[Strong time]{\includegraphics[width=2in]{strong_rep_time}} &
-      \subfigure[Strong efficiency]{\includegraphics[width=2in]{strong_rep_eff}} \\
-      %%
-    \end{tabular}
+    \begin{subfigure}{0.4\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{strong_rep_time}
+      \caption{Strong scaling time}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.4\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{strong_rep_eff}
+      \caption{Strong scaling efficiency}
+    \end{subfigure}
   \end{figure}
@@
 
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{GPU Optimization}
+\begin{frame}{Monte Carlo GPU Algorithm for Linear Systems}
 
@@
 %%---------------------------------------------------------------------------%%
-\subsection{Research Requirements}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Continued Research is Necessary}
-
-  The following important questions need to be answered before MC methods can
-  be applied to linear systems as a ``standard'' solver:
-
-  \begin{itemize}
-  \item Automatically determine history termination criteria
-    \begin{itemize}
-    \item We have investigated adaptive schemes based on variance
-    \item However, variance is not directly correlated with error
-    \end{itemize}
-  \item Identify classes of matrices and preconditioners that satisfy
-    convergence constraints
-    \begin{itemize}
-    \item This has been done for SDD and GDD
-    \end{itemize}
-  \item Demonstrate resiliency to hard MPI errors through replicated domain
-    parallelism
-  \item Further investigate non-linear, time-dependent problems
-    \begin{itemize}
-    \item This was part of the original proposal
-    \item Non-linear, sub-cycle time-stepping can result in strongly diagonally
-      dominant problems for which MCSA has been shown to be effective
-    \end{itemize}
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \subsection{Software}
@@
 
-\begin{frame}{MCSA Implementations}
+\begin{frame}{Evolution of MC Linear Solver Software Products}
 
@@
   through GoogleTest (Alea) or TeuchosTest (MCLS)
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\subsection{Research Requirements}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{What research remains for DOE ECP integration?}
+
+  The following important questions need to be answered before MC methods can
+  be applied to linear systems as a ``standard'' solver in ECP applications
+
+  \begin{itemize}
+  \item Automatically determine history termination criteria
+    \begin{itemize}
+    \item We have investigated adaptive schemes based on variance
+    \item However, variance is not directly correlated with error
+    \end{itemize}
+  \item Identify classes of matrices and preconditioners that satisfy
+    convergence constraints
+    \begin{itemize}
+    \item This has been done for SDD and GDD
+    \end{itemize}
+  \item Demonstrate resiliency to hard MPI errors through replicated domain
+    parallelism
+  \item Identify optimal, tunable parameter values for performance in
+    different applications
+  \end{itemize}
 \end{frame}
@@
 
-\begin{frame}{Connection to ECP Requirements}
+\begin{frame}{Ties to Identified Requirements of ECP Applications}
 
@@
 
-\begin{frame}{Inter-Operatibility Example}
-
-  Using the MCLS interface to Trilinos, it is straightforward to implement
-  MCSA in other applications
-  \vfill
-  We implemented MCSA in the production CFD code, Drekar (SNL) using a
-  Newton-MCSA algorithm:
+\begin{frame}{Inter-Operability---Connecting to Production Applications}
+
+  MCLS implements Trilinos ANA interfaces making it straightforward to
+  implement MCSA in other applications
+  \begin{itemize}
+  \item Exnihilo neutronics (shown before)
+  \item Drekar (SNL) CFD
+  \end{itemize}
+  \vfill
+  We implemented MCSA in Drekar using a Newton-MCSA algorithm via the Trilinos
+  package NOX:
   \begin{algorithm}[H]
@@
       \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
-      \COMMENT{Automatic differentiation - Panzer}
+      \COMMENT{Automatic differentiation - Trilinos FAD}
       \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
@@
   \vfill
-  Because MCLS was built with Trilinos abstract numerical algorithms this
-  integration was straightforward and we can test MCLS in a whole new problem
-  space
-  \vfill
-  We are also able to utilize all of Trilinos' preconditioners
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Thermal Convection Cavity Results}
+  \centering
+  \fbox{Trilinos ANA enables application integration}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Drekar-MCLS Thermal Convection Cavity Results}
 
@@
     \end{column}
-
+    %%
     \begin{column}{0.4\textwidth}
-
-      \small{
-        \begin{itemize}
-        \item Weaker preconditioning at Ra=\sn{1}{5} case
-          \medskip
-        \item Constant forcing term for Ra=\sn{1}{6} case
-        \end{itemize}
-      }
-
-      \bigskip
-
-      \tiny{
-        \begin{table}[h!]
-          \begin{center}
-            \begin{tabular}{lc}\hline\hline
-              \multicolumn{1}{l}{Benchmark}&
-              \multicolumn{1}{c}{NK Speedup}\\
-              \hline
-              Ra=\sn{1}{3} & 338 \\
-              Ra=\sn{1}{4} & 336 \\
-              Ra=\sn{1}{5} & 346 \\
-              Ra=\sn{1}{6} & 465 \\
-              \hline\hline
-            \end{tabular}
-          \end{center}
-          \caption{Newton-Krylov speedup over FANM.}
-        \end{table}
-      }
-
+      \vspace{-0.4in}
+      \begin{figure}
+        \centering
+        \includegraphics[width=0.9\textwidth]{convection_isotherms}
+        \caption{Steady state isotherms.}
+      \end{figure}
     \end{column}
+
   \end{columns}
@@
     \begin{figure}
-      \centering
-      \begin{tabular}{cc}
-        \subfigure[time]{\includegraphics[width=1.8in]{time}}
-        &
-        \subfigure[efficiency]{\includegraphics[width=1.8in]{tallies}}
-      \end{tabular}
+      \begin{subfigure}{0.4\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{time}
+      \end{subfigure}
+      ~
+      \begin{subfigure}{0.4\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{tallies}
+      \end{subfigure}
     \end{figure}
@@
 \section*{Extra Slides}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Forward Monte Carlo}
+\begin{itemize}
+  \item Choose a row-stochastic matrix $\mathbf{P}$ and weight matrix
+    $\mathbf{W}$ such that $\mathbf{H} = \mathbf{P} \circ \mathbf{W}$
+  \item Typical choice (Monte Carlo Almost-Optimal):
+    \begin{equation*}
+      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ij}| }
+      {\sum_{j=1}^{N} | \mathbf{H}_{ij} | }
+    \end{equation*}
+  \item To compute solution component $\mathbf{x}_i$:
+    \begin{itemize}
+      \item Start a history in state $i$ (with initial weight of 1)
+      \item Transition to new state $j$ based probabilities determined by
+        $\mathbf{P}_i$
+      \item Modify history weight based on corresponding entry in
+        $\mathbf{W}_{ij}$
+      \item Add contribution to $\mathbf{x}_i$ based on current history weight
+        and value of $\mathbf{b}_j$
+    \end{itemize}
+  \item A given random walk can only contribute to a single component of
+    the solution vector
+\end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Sampling Example (Forward Monte Carlo)}
+  \begin{itemize}
+    \item Suppose
+  \begin{equation*}
+    \mathbf{A} = \begin{bmatrix}
+      \phmin 1.0 & -0.2 & -0.6 \\
+      -0.4 & \phmin 1.0 & -0.4 \\
+      -0.1 & -0.4 & \phmin 1.0 \end{bmatrix} \to
+    \mathbf{H} \equiv (\mathbf{I-A}) = \begin{bmatrix}
+       0.0 &  0.2 &  0.6 \\
+       0.4 &  0.0 &  0.4 \\
+       0.1 &  0.4 &  0.0 \end{bmatrix}
+  \end{equation*}
+    then
+  \begin{equation*}
+    \mathbf{P} = \begin{bmatrix}
+       0.0 & 0.25 & 0.75 \\
+       0.5 &  0.0 & 0.5 \\
+       0.2 &  0.8 & 0.0 \end{bmatrix}, \quad
+    \mathbf{W} = \begin{bmatrix}
+       0.0 &  0.8 &  0.8 \\
+       0.8 &  0.0 &  0.8 \\
+       0.5 &  0.5 &  0.0 \end{bmatrix}
+  \end{equation*}
+    \vfill
+    \item If a history is started in state $3$, there is a $20\%$ chance of
+      it transitioning to state $1$ and an $80\%$ chance of moving to state
+      $2$
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Adjoint Monte Carlo}
+\begin{itemize}
+  \item Choose $\mathbf{P}$ and $\mathbf{W}$ such that
+    $\mathbf{H}^{T} = \mathbf{P} \circ \mathbf{W}$
+  \item Typical choice (Monte Carlo Almost-Optimal):
+    \begin{equation*}
+      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ji}| }
+      {\sum_{i=1}^{N} | \mathbf{H}_{ji} |}
+    \end{equation*}
+  \item To estimate solution:
+    \begin{itemize}
+      \item Start a history in random state $i$ by sampling from distribution
+        determined by $\mathbf{b}$
+      \item Transition to new state $j$ based probabilities determined by
+        $\mathbf{P}_i$
+      \item Modify history weight based on corresponding entry in
+        $\mathbf{W}_{ij}$
+      \item Add contribution to $\mathbf{x}_j$ based on current history weight
+    \end{itemize}
+  \item A given random walk can contribute to many different components of
+    the solution vector
+\end{itemize}
+\end{frame}
+
 %%---------------------------------------------------------------------------%%
@@
   \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Monte Carlo as Approximate Polynomial Preconditioning}
+  \begin{itemize}
+    \item Using history length cutoff, Monte Carlo process is approximating
+      a fixed length Neumann series polynomial
+    \vfill
+    \item Why limit ourselves to the Neumann series polynomial?
+      \begin{itemize}
+        \item Chebyshev or GMRES polynomials are viable alternatives
+      \end{itemize}
+  \end{itemize}
+
+  \begin{table}
+    \caption{Polynomial MCSA - Shifted Laplacian}
+    \centering
+    \begin{tabular}{cccc}
+      \toprule
+      Polynomial & Polynomial & MCSA & Histories per \\
+      Type & Order & Iterations & Iteration \\
+      \midrule
+      Neumann & 4 & 502 & 500 \\
+      Chebyshev & 4 & 246 & 16000 \\
+      GMRES & 4 & 260 & 2000 \\
+      \bottomrule
+    \end{tabular}
+  \end{table}
 \end{frame}

### 2016-01-26 doc/ecp/pres.tex
Table formatting.
Almost finished.

@@
 
-  \begin{table}[!h]
-  \centering
-  \begin{tabular}{|c|c|c|c|}
-  \hline
+  \begin{table}
+  \centering
+  \begin{tabular}{cccc}
+  \toprule
   \textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
   &$\rho(\hat{H})$\\
-  \hline
+  \midrule
    3 & 1.4352 & $0.9804$ & $0.9531$\\
-  \hline
    9 & 3.0220 & $0.9790$ & $0.9674$\\
-  \hline
    33 & 9.8164 & $0.9783$  & $0.9774$\\
-  \hline
+  \bottomrule
   \end{tabular}
@@
 
-  \begin{table}[!h]
-  \centering
-  \begin{tabular}{|c|c|c|c|}
-  \hline
+  \begin{table}
+  \centering
+  \begin{tabular}{cccc}
+  \toprule
   \textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
   &$\rho(\hat{H})$\\
-  \hline
-   0.05 & 8.2797 & $0.9610$ & $1.0231$\\
-  \hline
-   0.01 & 33.1390 & $0.8668$ & $0.8279$\\
-  \hline
+  \midrule
+   0.05 & 8.2797 & $0.9610$ & $1.0231$  \\
+   0.01 & 33.1390 & $0.8668$ & $0.8279$ \\
+  \bottomrule
   \end{tabular}
@@
       \caption{2D parabolic problem with AINV preconditioning.}
-      \begin{tabular}{|c|c|c|c|}\hline
-        Node & Iterations & Time (sec) & Relative Residual \\\hline
-        CPU & 6 & 992.23 & $1.23\times 10^{-7}$ \\\hline
-        GPU & 6 & 29.83 & $1.53\times 10^{-7}$ \\\hline
+      \begin{tabular}{cccc}
+        \toprule
+        Node & Iterations & Time (sec) & Relative Residual \\
+        \midrule
+        CPU & 6 & 992.23 & $1.23\times 10^{-7}$ \\
+        GPU & 6 & 29.83 & $1.53\times 10^{-7}$ \\
+        \bottomrule
       \end{tabular}
@@
       \small{
-        \begin{table}[h!]
+        \begin{table}
           \begin{center}
-            \begin{tabular}{lccc}\hline\hline
+            \begin{tabular}{lccc}
+              \toprule
               \multicolumn{1}{l}{Benchmark}&
@@
               \multicolumn{1}{c}{NR}\\
-              \hline
+              \midrule
               Ra=\sn{1}{3} & 5 & 5 & 5\\
@@
               %%
-              \hline\hline
+              \bottomrule
             \end{tabular}`
@@
 
-        \begin{table}[h!]
+        \begin{table}
           \begin{center}
-            \begin{tabular}{lccc}\hline\hline
+            \begin{tabular}{lccc}
+              \toprule
               \multicolumn{1}{l}{Benchmark}&
@@
               \multicolumn{1}{c}{Richardson}\\
-              \hline
+              \midrule
               Ra=\sn{1}{3} & 32 & 18 & 38\\
@@
               %%
-              \hline\hline
+              \bottomrule
             \end{tabular}
@@
 %%---------------------------------------------------------------------------%%
-\subsection{ECP R\&D}
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{ECP Coordinated Reasearch}
+\subsection{R\&D in ECP}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Monte Carlo R\&D in ECP}
 
@@
     \setlength{\itemsep}{-0.2\baselineskip}
+  \item fluid dynamics
+  \item combustion
+  \item quantum chemistry
+  \item visualization
+  \item nuclear engineering
   \item high-energy density physics
   \item astrophysics
-  \item nuclear engineering
   \item particle accelerators and neutron spallation source
-  \item quantum chemistry
-  \item visualization
-  \item fluid dynamics
-  \item combustion
   \item computational biology
@@
 
-\begin{frame}{ECP R\&D I: Time-dependent Problems}
-
-  Investigate use in time-dependent applications, e.g. non-linear radiation
-  hydrodynamics for high-energy density physics applications:
+\begin{frame}{Monte Carlo for Time-dependent Applications in ECP}
+
+  Investigate use in time-dependent applications, e.g. fluid flow, compustion,
+  non-linear radiation hydrodynamics for high-energy density physics
+  applications:
+  \vfill
   \begin{itemize}
   \item Used as inner linear solver in a non-linear (Newton) iteration
+  \vfill
   \item Time-dependent applications generally have strong diagonal-dominance
     that makes MCSA competitive with Krylov methods
+  \vfill
   \item MCSA has resiliency and parallel efficiency benefits over conventional
@@
 
-\begin{frame}{ECP R\&D II: Monte Carlo Particle Transport}
-
-  MC particle transport has applications in many areas across the DOE complex
+\begin{frame}{Monte Carlo for Particle Transport Applications in ECP}
+
+  MC particle transport has applications across the DOE complex
   \begin{itemize}
     \setlength{\itemsep}{-0.2\baselineskip}
+  \item combustion
+  \item nuclear engineering
+  \item rarefied gas dynamics
+  \item astrophysics
   \item high-energy density physics
-  \item neutron spallation source
-  \item nuclear engineering
-  \item astrophysics
-  \item combustion
   \end{itemize}
@@
   \begin{enumerate}
+    \setlength{\itemsep}{-0.2\baselineskip}
   \item Random walks are dominated by branching statements $\rightarrow$
@@
 
-\begin{frame}{ECP R\&D III: Particle Methods}
+\begin{frame}{General Particle Applications in ECP}
 

### 2016-01-26 doc/ecp/pres.tex
Finished all but timeline.

@@
 
-\begin{frame}{Building on ASCR Funding}
-
-  \underline{\bf Trilinos}
-  \begin{itemize}
-  \item Provides abstract solver interface for MCLS
-  \item Provides build system for Alea and MCLS
+\begin{frame}{Trilinos Enables Inter-Operability}
+
+  \begin{itemize}
+  \item Provides abstract numerical algorithms
+    \begin{itemize}
+    \item solver interfaces
+    \item operator/vector interfaces
+    \item preconditioner interfaces
+    \end{itemize}
+    \vfill
+  \item Provides build system (TriBITS) for Alea and MCLS
+    \vfill
   \item Provides data structures and utilities for Alea and MCLS
+    \vfill
   \item Provides testing harness for MCLS
@@
 
-  \vfill
-
-  \underline{\bf xSIM}
-  \begin{itemize}
-  \item Simulator to test for hard error detection and remediation
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
-\begin{frame}{Building on ASCR Funding}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{xSIM Development and Fault Analysis Supported by MCREX}
+
+  \begin{itemize}
+  \item Use xSIM to model communication performance of MPI messaging
+    \begin{itemize}
+    \item Latency, bandwidth, contention, routing, and rendevous protocol
+    \item Star, ring, mesh, torus, twisted torus, and tree topologies
+    \item Hierarchical combinations of topologies, e.g. on-chip, on-node,
+      off-node
+    \end{itemize}
+  \item Also used to simulate hard errors to test for detection and recovery
+  \end{itemize}
+  \vfill
+  \begin{figure}
+    \centering
+    \includegraphics[height=.4\textwidth]{xsim}
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Using XSTACK Technologies Developed by XPRESS}
 

### 2016-01-28 doc/ecp/pres.tex
Final draft.
Fixed spelling error.

@@
 
-  Investigate use in time-dependent applications, e.g. fluid flow, compustion,
+  Apply MCLS to time-dependent applications, e.g. fluid flow, compustion,
   non-linear radiation hydrodynamics for high-energy density physics
@@
 
-  A supporting technolgies research effort on a generalized set of libraries
+  A supporting technologies research effort on a generalized set of libraries
   for managing particle-based data structures on heterogeneous computing
@@
 %%---------------------------------------------------------------------------%%
+\subsection{ECP Roadmap}
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{A Monte Carlo and Particle Methods Roadmap for ECP}
+
+  \centering
+  \includegraphics[width=1\textwidth]{timeline}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 
@@
 \huge Questions?
+\vfill
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}
+\vfill
+\centering
+\huge Extra Slides
 \vfill

### 2016-01-28 doc/ecp/pres.tex
Final presentation.

@@
 
-\logo{\includegraphics[width=2cm]{new_logo}}
+\logo{\includegraphics[width=2cm]{faded_logo}}
 
@@
 \beamertemplatenavigationsymbolsempty
-\setbeamertemplate{headline}{}
-\setbeamertemplate{footline}{
-  \insertframenumber/\inserttotalframenumber}
+%\setbeamertemplate{headline}{}
+%\setbeamertemplate{footline}{
+%  \insertframenumber/\inserttotalframenumber}
+
 \setbeamertemplate{title page}
@@
       \insertauthor\\
-      \vspace{0.5\baselineskip}\\
+      %\vspace{0.5\baselineskip}\\
       \insertinstitute\\
@@
 
-\title[MCREX]{Using Monte Carlo Algorithms to Achieve Resiliency \& Performance
-  at Scale for Linear and Nonlinear Solver Applications}
-\author{Thomas Evans (PI, ORNL)\\
+\title[MCREX]{MCREX: Using Monte Carlo Algorithms to Achieve Resiliency \&
+  Performance at Scale for Linear and Nonlinear Solver Applications}
+\author[Evans et al.]{Thomas Evans (PI, ORNL)\\
   Michele Benzi (Co-PI, Emory)\\
@@
   Wayne Joubert (ORNL)\\
-  Massimiliano Lupo Pasini (Emory)
-}
+  Massimiliano Lupo Pasini (Emory) }
 \date[DOE, 2016]{DOE RX-Solvers, 1/29/2016}
@@
 {
+  \setbeamertemplate{headline}{}
+  \setbeamertemplate{footline}{}
   \usebackgroundtemplate{
@@
   \setbeamertemplate{logo}{}
-  \begin{frame}
+  \begin{frame}[plain]
     \titlepage
@@
     Carlo that allows redundancy
-  \item Developed concurrent on-node random walk algorithms for GPU
-    optimization
+  \item Developed on-node random walk algorithms for GPU optimization
   \end{itemize}
@@
 
-  First reference, G.E. Forsythe and R.A. Leibler. Matrix Inverstion by a
-  Monte Carlo Method. \textit{Math. Tables and Comp.}, {\bf 4}(31), 127--129,
-  1950 (credited to von Neumann and Ulam, 1940s).
-  \vfill
+  First reference: G.E. Forsythe and R.A. Leibler. Matrix Inversion by
+    a Monte Carlo Method. \textit{Math. Tables and Comp.}, {\bf 4}(31),
+    127--129, 1950 (credited to von Neumann and Ulam, 1940s).
   \begin{itemize}
@@
     \caption{
-      \only<1>{\textbf{Adjoint solution.}
-        \textit{\sn{1}{0} total histories.} }
-      \only<2>{\textbf{Adjoint solution.}
-        \textit{\sn{1}{1} total histories.} }
-      \only<3>{\textbf{Adjoint solution.}
-        \textit{\sn{1}{2} total histories.} }
-      \only<4>{\textbf{Adjoint solution.}
-        \textit{\sn{1}{3} total histories.} }
-      \only<5>{\textbf{Adjoint solution.}
-        \textit{\sn{1}{4} total histories.} }
-      \only<6>{\textbf{Adjoint solution.}
-        \textit{\sn{1}{5} total histories.} }
-      \only<7>{\textbf{Adjoint solution.}
-        \textit{\sn{1}{6} total histories.} }
-      \only<8>{\textbf{Adjoint solution.}
-        \textit{\sn{1}{7} total histories.} }
+      \only<1>{\textit{\sn{1}{0} total histories.} }%
+      \only<2>{\textit{\sn{1}{1} total histories.} }%
+      \only<3>{\textit{\sn{1}{2} total histories.} }%
+      \only<4>{\textit{\sn{1}{3} total histories.} }%
+      \only<5>{\textit{\sn{1}{4} total histories.} }%
+      \only<6>{\textit{\sn{1}{5} total histories.} }%
+      \only<7>{\textit{\sn{1}{6} total histories.} }%
+      \only<8>{\textit{\sn{1}{7} total histories.} }%
     }
@@
   \end{itemize}
-  \vfill
-
+  \vspace{-0.6\baselineskip}
   \begin{table}
@@
   \centering
+  \caption{Advection--diffusion matrix with block diagonal preconditioning.}
   \begin{tabular}{cccc}
@@
   \end{tabular}
-  \caption{$\rho(H)$ and $\rho(\hat{H})$ for advection--diffusion
-  matrix with block diagonal preconditioning.}
-  \label{tab:ifiss_block}
   \end{table}
@@
   \centering
+  \caption{2D Laplacian with AINV preconditioning.}
   \begin{tabular}{cccc}
@@
   \end{tabular}
-  \caption{$\rho(H)$ and $\rho(\hat{H})$ for 2D
-  Laplacian with AINV preconditioning.}
-  \label{tab:lap2d_ainv}
   \end{table}
@@
       \begin{itemize}
-        \setlength{\itemsep}{-0.4\baselineskip}
+        %\setlength{\itemsep}{-0.4\baselineskip}
       \item Each parallel process owns a piece of the domain (linear
         system)
-        \bigskip
       \item Random walks must be transported between adjacent domains
         through parallel communication
-        \bigskip
       \item Domain decomposition determined by the input system
-        \bigskip
       \item Load balancing not yet addressed
-        \bigskip
       \item Uses modified version of the Milagro algorithm (LANL)
@@
 
-\begin{frame}{Publications and Presentations}
-
-  \underline{Publications}
+\begin{frame}{Publications}
+
   \begin{itemize}
@@
     Sparse Linear Systems. \textit{in preparation}, 2016.
+  \item C. Engelmann and T. Naughton. A New Deadlock Resolution Protocol and
+    Message Matching Algorithm for the Extreme-scale
+    Simulator. \textit{Concurrency and Computation: Practice and Experience,
+      2016}. John Wiley \& Sons, Inc.. ISSN 1532-0634, 2016.
   \item S.R. Slattery, T.M. Evans, P.P.H. Wilson. A Spectral Analysis of the
@@
 
-\begin{frame}{Publications and Presentations}
-
-  \underline{Presentations}
+\begin{frame}{Conferences}
+
   \begin{itemize}
@@
     \setlength{\itemsep}{-0.1\baselineskip}
+  \item C. Engelmann and T. Naughton. Supporting the Development of Soft-Error
+    Resilient Message Passing Applications using Simulation. In
+    \textit{Proceedings of the 13th IASTED International Conference on
+      Parallel and Distributed Computing and Networks (PDCN) 2016}, Innsbruck,
+    Austria, February 15-16, 2016. ACTA Press, Calgary, AB, Canada.
   \item Monte Carlo Synthetic Acceleration Methods for Sparse Linear Systems
@@
     Salt Lake City, 2015.
+  \item C. Engelmann and T. Naughton. A Network Contention Model for the
+    Extreme-scale Simulator. In \textit{Proceedings of the 34th IASTED
+      International Conference on Modelling, Identification and Control (MIC)
+      2015}, Innsbruck, Austria, February 17-18, 2015. ACTA Press, Calgary,
+    AB, Canada.
   \item Iterative Performance of Monte Carlo Linear Solver Methods
     (M.L. Pasini), \textit{SIAM CSE}, Salt Lake City, 2015.
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Conferences}
+
+  \begin{itemize}
+    \footnotesize
+    \setlength{\itemsep}{-0.1\baselineskip}
   \item A multilevel Monte Carlo method for linear systems (S. Slattery,
@@
       Mountain Conference on Iterative Solvers}, Copper MT, CO, 2014.
+  \item C. Engelmann and T. Naughton. Improving the Performance of the
+    Extreme-scale Simulator. In \textit{Proceedings of the 18th IEEE/ACM
+      International Symposium on Distributed Simulation and Real Time
+      Applications (DS-RT) 2014}, pages 198-207, Toulouse, France, October
+    1-3, 2014. IEEE Computer Society, Los Alamitos, CA, USA. ISBN
+    978-1-4799-6143-6. ISSN 1550-6525.
+  \item T. Naughton, C. Engelmann, G. Vallée, and S. Böhm. Supporting the
+    Development of Resilient Message Passing Applications using Simulation. In
+    \textit{Proceedings of the 22nd Euromicro International Conference on
+      Parallel, Distributed, and network-based Processing (PDP) 2014}, pages
+    271-278, Turin, Italy, February 12-14, 2014. IEEE Computer Society, Los
+    Alamitos, CA, USA. ISSN 1066-6192. Acceptance rate 32.6\% (73/224).
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+
+\begin{frame}{Presentations}
+
+  \begin{itemize}
   \item Monte Carlo Linear Solvers (S. Slattery, T. Evans, S. Hamilton),
@@
         \end{table}
-
+        \vspace{-0.5\baselineskip}
         \begin{table}
@@
     \centering
-    \includegraphics[height=.4\textwidth]{xsim}
+    \includegraphics[height=.35\textwidth]{xsim}
   \end{figure}

### 2016-01-29 manuscript1/paper.tex
manuscript1 folder added

%
% This draft dated January 29, 2016  [MB]
%

\documentclass[final,leqno,onefignum,onetabnum]{siamltex1213}
\usepackage{amssymb,amsmath,bm}
\usepackage{color}
\usepackage[lined,boxruled]{algorithm2e}
\usepackage{comment}
\usepackage{calrsfs}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{rotating}
\usepackage{multirow}
\DeclareMathAlphabet{\pcal}{OMS}{zplm}{m}{n}

 \newtheorem{thm}{Theorem}
% \newtheorem{lemma}{Lemma}
 \newtheorem{defn}{Definition}
 \newtheorem{prop}{Proposition}
\newtheorem{rem}{Remark}

\hoffset 0.6in

%opening
\title{Analysis of Monte Carlo accelerated iterative methods for
sparse linear systems}
\author{Michele Benzi\thanks{Department of Mathematics and Computer
Science, Emory University, Atlanta, Georgia 30322, USA
(benzi@mathcs.emory.edu). The work of this author was supported
by Department of Energy (Office of Science) grant ERKJ247.}
\and
Thomas M.~Evans\thanks{Oak Ridge National Laboratory, 1 Bethel Valley Rd., 
Oak Ridge, TN 37831, USA (evanstm@ornl.gov).}
\and Steven P. Hamilton\thanks{Oak Ridge National Laboratory, 
1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (hamiltonsp@ornl.gov).}
\and 
\linebreak Massimiliano Lupo  Pasini\thanks{Department of Mathematics and Computer
Science, Emory University, Atlanta, Georgia 30322, USA (mlupopa@emory.edu).}
\and 
Stuart R.~Slattery\thanks{Oak Ridge National Laboratory,
1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (slatterysr@ornl.gov).}
}
\date{}

\begin{document}
\bibliographystyle{siam}

\pagestyle{myheadings}
\markboth{{\sc M.~Benzi, T.~M.~Evans, S.~P.~Hamilton, M.~Lupo Pasini, and S.~Slattery}}
{Monte Carlo solvers for sparse linear systems}

\maketitle

\begin{abstract}
We consider hybrid deterministic-stochastic iterative algorithms for the solution
of large, sparse linear systems. Starting from a convergent splitting of the 
coefficient matrix, we analyze various types of Monte Carlo acceleration schemes
applied to the original preconditioned Richardson (stationary) iteration. These
methods are expected to have considerable potential for resiliency to faults
when implemented on massively parallel machines.

We establish sufficient conditions for the convergence of the hybrid schemes, and
we investigate different types of preconditioners including sparse approximate
inverses. Numerical experiments on linear systems arising from the discretization
of partial differential equations are presented.
\end{abstract}

\begin{keywords}
sparse linear systems, iterative methods, Richardson iteration, preconditioning,
Monte Carlo methods, sparse approximate inverses, resilience
\end{keywords}

\begin{AMS}
65C05, 65F50, 65F08, 65F10
\end{AMS}

\section{Introduction}
\label{sec:intro}

The next generation of computational science applications will require numerical
solvers that are both reliable and capable of high performance on projected exascale
platforms. In order to meet these goals, solvers must be resilient to soft and hard 
system failures, provide high concurrency on heterogeneous hardware configurations,
and retain numerical accuracy and efficiency. In this paper we focus on the
solution of large sparse systems of linear equations, for
example of the kind arising from the discretization of partial differential
rquations (PDEs).  A possible approach is to try to adapt existing solvers
(such as preconditioned Krylov subspace or multigrid methods) to the new
computational environments, and indeed several efforts are under way in this
direction; see, e.g., \cite{Agullo,FLRU15,Heroux,Rizzi,Stoyanov} and references therein. 
An alternative approach is to investigate
new algorithms that can address issues of resiliency, particularly fault
tolerance and hard processor failures, naturally. An example is provided
by the recently proposed {\em Monte Carlo Synthetic Acceleration Methods}
(MCSA), see \cite{EMSH2014,Slattery2013}. 
In these methods, an underlying (deterministic) stationary Richardson iterative
method is combined with a stochastic, Monte Carlo-based ``acceleration" scheme.
Ideally, the accelerated scheme will converge to the solution of the linear
system in far fewer (outer) iterations than the basic scheme without Monte Carlo
acceleration, with the added advantage that most of the computational effort
is now relegated to the Monte Carlo portion of the algorithm, which is 
highly parallel and offers a more straightforward path to resiliency
than standard, deterministic solvers. In addition,
a careful combination of the Richardson
and Monte Carlo parts of the algorithm allows to circumvent the well known
problem of slow Monte Carlo error reduction; see \cite{EMSH2014}.

Numerical evidence presented in \cite{EMSH2014} suggests
that MCSA can be competitive, for certain classes of problems,
with established deterministic solvers such as preconditioned 
conjugate gradients and GMRES. So far, however, no theoretical analysis
of the convergence properties of these solvers has been carried out. In
particular, it is not clear a priori whether the method, applied to a
particular linear system, will converge. 
Indeed, the convergence of the underlying preconditioned Richardson
iteration is not sufficient, in general, to guarantee the convergence
of the MCSA-accelerated iteration. In other words, it is quite possible
that the stochastic ``acceleration" part of the algorithm may actually
cause the hybrid method to diverge or stagnate.

In this paper we address this fundamental issue,
discussing both necessary and sufficient conditions for convergence.
We also discuss the choice of splitting, or preconditioner, and illustrate
our findings by means of numerical experiments. 
We do not specifically
consider in this paper the resiliency issue, which will be addressed 
elsewhere.

The paper is organized as follows.
In Section~\ref{sec:mcls} we provide an overview of existing Monte Carlo
linear solver algorithms.
In Section~\ref{sec:convergence} we will discuss the convergence behavior
of stochastic solvers, including a discussion of classes of matrices for
which convergence can be guaranteed.
Section~\ref{sec:results} provides some numerical results illustrating
properties of the various approaches and in Section~\ref{sec:conclusion} we
give our conclusions.

\section{Stochastic linear solvers}
\label{sec:mcls}

Linear solvers based on stochastic techniques have a
long history, 
going back to the famed 1928 paper by Courant, Friedrichs, and Lewy
on finite difference schemes for PDEs \cite{CFL}. 
Many authors have considered linear solvers based on Monte Carlo
techniques, with important early contributions by
Curtiss \cite{Curtiss} and by Forsythe and Leibler \cite{FL50}. 
More recent works include \cite{AADBTW2005,Hal1994},
and \cite{DA1998}, among others.
Until recently, 
these methods have had mixed success at best,
due to their generally inferior performance when compared to state-of-the-art
deterministic solvers like multigrid or preconditioned Krylov methods.
Current interest in resilient solvers, where some performance may be
traded off for increased robustness in the presence of faults, has prompted a fresh 
look at methods incorporating Monte Carlo ideas 
\cite{EMSH2014,ESW2013,Slattery2013}.

As mentioned in \cite{DA1998}, Monte Carlo methods may be divided into two broad
classes: \textit{direct methods}, such as those described in \cite{DA1998,DVA2001}, and
\textit{iterative methods}, which refer to techniques such as those presented
in \cite{Hal1962,Hal1994}; see also \cite{EMSH2014,ESW2013}.
The first type consists of purely stochastic schemes,
therefore the resulting error with respect to the exact solution is made of
just a stochastic component. In contrast, the iterative Monte Carlo methods utilize more
traditional iterative algorithms alongside the stochastic approach,
generating two types of error: a
\textit{stochastic} one and a \textit{systematic} one. 
In practice, it may be difficult to separate the two components;
nevertheless, awareness of this intrinsic structure is useful, as it allows 
algorithm designers some flexibility in the choice of what part of the algorithm
to target for refinement (e.g., trading off convergence speed for resilience by
balancing the number of ``deterministic" outer iterations
against the number of random walks to be used within each iteration).

Consider a system of linear equations of the form
\begin{equation}
A \mathbf{x}=\mathbf{b},
\label{linsys}
\end{equation}
where $A\in \mathbb{R}^{n\times n}$ and $\mathbf{x}$, $\mathbf{b} \in
\mathbb{R}^n$.
Equation~(\ref{linsys}) can be recast as a fixed point problem:
\begin{equation}
 \mathbf{x}=H\mathbf{x}+\mathbf{f},
 \label{fixedpoint}
\end{equation}
where $H=I-A$ and $\mathbf{f}=\mathbf{b}$.
Assuming that the spectral radius $\rho(H)<1$, the solution to
\eqref{fixedpoint} can be written in terms of a power series in
$H$ (Neumann series):
\[
\mathbf{x}=\sum_{\ell=0}^\infty H^\ell\mathbf{f}.
\]
Denoting the $k$th partial sum by $\mathbf{x}^{(k)}$, the sequence of approximate solutions
$\{\mathbf{x}^{(k)}\}_{k=0}^{\infty}$ converges to the exact solution
regardless of the initial guess $\mathbf{x}_0$.

By restricting the attention to a single component of $\mathbf{x}$ we
obtain
\begin{equation}
x_i=f_i + \sum_{\ell=1}^\infty \sum_{k_1=1}^n\sum_{k_2=1}^n\cdots 
\sum_{k_{\ell}=1}^n
H_{i,k_1}H_{k_1,k_2}\cdots H_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{forward}
\end{equation}
The last equation can be reinterpreted as the realization of an estimator
defined on a random walk.  Let us start considering a random walk whose
state space $S$ is labeled by the set of indices of the forcing term
$\mathbf{f}$:
\[
S=\{1,2,\ldots, n\} \subset \mathbb{N}.
\]
Each $i$th step of the random walk has a random variable
$k_i$ associated with it. The realization of $k_i$ represents the index of the
component of $\mathbf{f}$
which is visited in the current step of the random walk.
The transition probabilities and the selection of
the initial state of the random walk can be accomplished according to 
different modalities, leading to two distinct methods:
the \textit{forward} and \textit{adjoint} methods.
These methods are described next.

\subsection{Forward method}
\label{subsec:forward}

The general goal is to evaluate a functional as
\[
J(\mathbf{x})=\left \langle\mathbf{h},\mathbf{x}\right\rangle=\sum_{i=1}^n h_i 
x_i,
\]
where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in
$\mathbb{R}^n$ of the functional $J$.
Such a representative can be used to build the initial probability 
$\tilde{p}:
S\rightarrow [0,1]$ of the random walk as
\[
\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n 
\lvert h_i\rvert}.
\]
It is important to highlight that the role of vector $\mathbf{h}$ is
confined to the construction of the initial probability, and that $\mathbf{h}$
is not
used afterwards in the stochastic process.
A possible choice for the transition probability $P$ can be
\[
p(k_{\ell}=j \;\lvert\;k_{\ell-1}=i )=P_{ij}=\frac{\lvert 
H_{ij}\rvert}{\sum_{k=1}^n
\lvert H_{ik}\rvert}.
\]
where 
\[
p(\cdot,i):S\rightarrow [0,1], \qquad \forall i\in S
\]
and $k_{\ell}\in S$ 
represents the state reached at a generic $\ell$th step of the random walk.
A related sequence of random variables $w_{ij}$ can be defined
as
\[
w_{ij}=\frac{H_{ij}}{P_{ij}}.
\]
The probability distribution of the random variables $w_{ij}$ is represented
by the transition matrix that governs the stochastic process. The $w_{ij}$
quantities just introduced can be used to build one more sequence
of random variables.
At first we introduce quantities $W_\ell$
\[
W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_{\ell}=W_{\ell-1} 
w_{k_{\ell-1},k_{\ell}}.
\]
By introducing $\nu$ as a generic permutation that represents the realization 
of a random walk, we define
\[
X(\nu)=\sum_{\ell=0}^\infty W_{\ell} f_{k_{\ell}}
\]
as the random variable associated with a specific permutation $\nu$. 
We can thus
define the estimator $\theta$ as
\[
\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu),
\]
where $\nu$ ranges over all possible realizations.
$P_{\nu}$ is the probability associated with a specific permutation of the
random walk.
It can be proved (see \cite{Hal1962} and \cite{Hal1994}) that
\[
E[W_{\ell} 
f_{k_{\ell}}]=\left\langle\mathbf{h},H^{\ell}\mathbf{f}\right\rangle, \quad 
\ell=0,1,2,\ldots
\]
and
\[
\theta=E\bigg[\sum_{\ell=0}^\infty W_{\ell} 
f_{k_{\ell}}\bigg]=\left\langle\mathbf{h},\mathbf{x}\right\rangle.
\]

A possible choice for $\mathbf{h}$ is a vector of the standard basis, 
$\mathbf{h} = \mathbf{e}_i$. This
would correspond to setting the related initial probability to a Kronecker 
delta:
\[
\tilde{p}(k_0=j)=\delta_{ij}.
\]
By doing so, we have $k_0 = i$ and 
\begin{equation}
%\theta= E\bigg[\sum_{\ell=0}^\infty W_{\ell}
%f_{k_{\ell}}\bigg]= 
\theta = x_i= 
f_i + \sum_{l=1}^\infty
\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n
P_{i,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},
k_{\ell}}w_{i,k_1}w_{k_1,k_2}\cdots
w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{dir_mean}
\end{equation}

As regards the variance, we recall the following relation:
\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_{\ell}}^2\bigg] - \bigg (E\bigg[\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]\bigg )^2.
\label{dir_var}
\end{equation}

Hence, the variance can be computed as the difference between the second
moment of the random variable and the square of its first moment.\newline

In order to apply the Central Limit Theorem (CLT) to the estimators defined
above, we must require that
the estimators have both finite expected value and finite variance. This is
equivalent to
checking the finiteness of the expected value and second moment.
Therefore, we have to impose the following conditions:

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell} f_{k_{\ell}}\bigg]<\infty,
\end{equation}

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_{\ell}}^2\bigg]<\infty .
\end{equation}

The forward method presented above, however, has the limitation of employing an
entire set of permutations to estimate just a single entry of
the solution at a time. Hence, in order to estimate the entire solution
vector for Eq.~\eqref{linsys}, we have to employ a separate set of
permutations for each entry in the solution vector. This limitation
can be circumvented by the adjoint method which we describe below.

\begin{rem}
It is important to note that in order to construct the random walks, 
access to the individual entries of $H$ is required. Hence, $H$ needs to be formed explicitly
and therefore must be sparse in order to have a practical algorithm.
\end{rem}

\subsection{Adjoint method}
\label{subsec:adjoint}

A second Monte Carlo method can be derived by considering the linear system
adjoint to \eqref{linsys}:
\begin{equation}
A^T\mathbf{y}=\mathbf{d},
\label{adj}
\end{equation}
where $\mathbf{y}$ and $\mathbf{d}$ are the adjoint solution and source term.
Equation \eqref{adj} can be recast in a manner similar to \eqref{fixedpoint}:
\[
\mathbf{y} = H^T \mathbf{y} + \mathbf{d}.
\]
Note that $\rho(H^T) = \rho(H) <1$, hence convergence of the Neumann series
(fixed point iteration) for (\ref{linsys}) guarantees convergence
for the adjoint system (\ref{adj}).

Exploiting the following inner product equivalence:
\[
\left\langle 
A^T\mathbf{y},\mathbf{x}\right\rangle=\left\langle\mathbf{y},A\mathbf{x} 
\right\rangle ,
\]
it follows that
\begin{equation}
\left\langle\mathbf{x},\mathbf{d}\right\rangle=\left\langle\mathbf{y},\mathbf{f}
\right\rangle.
\label{adj_relationship}
\end{equation}

By writing the Neumann series for the solution to \eqref{adj} we have:
\[
\mathbf{y}=\sum_{\ell=0}^{\infty} (H^T)^\ell \mathbf{d},
\]
and focusing on a single entry of the solution vector:
\[
y_i = d_i + \sum_{\ell=1}^{\infty}\sum_{k_1=1}^{n}\sum_{k_2=1}^n 
\cdots \sum_{k_{\ell}=1}^n H^T_{i,k_1} H^T_{k_1,k_2}\cdots 
H^T_{k_{\ell-1},k_{\ell}} d_{k_\ell}.
\]

The undetermined quantities in the dual problem \eqref{adj} are $\mathbf{y}$ 
and 
$\mathbf{d}$. Therefore, two constraints are required: the first constraint is 
Eq.~\eqref{adj_relationship} and as a second constraint we select
$\mathbf{d}$ to be one of the standard basis vectors.
Applying this choice of $\mathbf{d}$ to (\ref{adj_relationship}) we get:
\[
\left\langle\mathbf{y},\mathbf{f}\right\rangle=\left\langle\mathbf{x},\mathbf{d}
\right\rangle=x_i.
\]

In order to give a stochastic interpretation of the adjoint method similar to 
the one obtained for the forward method, we introduce the initial probability:
\[
\tilde{p}(k_0=i)=\frac{\lvert f_i\rvert}{\lVert 
\mathbf{f}\rVert_1}
\]
and the initial weight:
\[
W_0 = \lVert \mathbf{f}\rVert_1\frac{f_i}{\lvert f_i \rvert}.
\]
The transition probability is defined as 
\[
p(k_\ell = j \lvert k_{\ell-1}=i)=P_{ij}=\frac{\lvert 
H^T_{ij}\rvert}{\sum_{k=1}^n \lvert H^T_{ik}\rvert} = \frac{\lvert 
H_{ji}\rvert}{\sum_{k=1}^n \lvert H_{ki}\rvert}
\]
and the sequence of weights as follows:
\[
w_{ij}=\frac{H_{ji}}{P_{ij}}.
\]

By reformulating the fixed point scheme in its statistical interpretation, the 
following formula
holds for the estimator of the solution vector associated with the adjoint 
method: it is the vector $\boldsymbol{\theta}\in \mathbb{R}^n$ such that
\begin{equation}
%\[
\begin{array}{rl}
\theta_i& =E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell},
i}\bigg]\\
&={\displaystyle \sum_{\ell=0}^{\infty}\sum_{k_0=1}^n\sum_{k_1=1}^n\sum_{k_2=1}
^n\cdots\sum_ { k_ { \ell }=1 } ^n
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots 
P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},i}.}
\label{adj_mean}
\end{array}
%\]
\end{equation}
This estimator is known in literature as \textit{collision} estimator.

The forward method adds a contribution to the component of the solution
vector where the random walk began, based on the value of the source vector
in the state in which the walk currently resides.  The adjoint method,
on the other hand, adds a contribution to the component of the solution
vector where the random walk currently resides based on the value of the
source vector in the state in which the walk began.
The Kronecker delta at the end of the series \eqref{adj_mean} represents a 
filter, indicating
that only a subset of permutations contribute to the $j$th component
of the solution vector.

The variance is given by
\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
\delta_{k_{\ell},i}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
\delta_{k_{\ell},i}\bigg ] - \bigg 
(E\bigg[\sum_{\ell=0}^\infty
W_{\ell}
\delta_{k_{\ell},i}\bigg]\bigg )^2, \qquad i=1,\ldots,n
\label{adj_var}.
\end{equation}

Along the same lines as the development for the forward method, we must
impose finiteness of the expected value and second moment.
Therefore, the following
conditions must be verified:

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell},
i}\bigg]<\infty \qquad i=1,\ldots,n
\end{equation}
and
\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
\delta_{k_{\ell},i}\bigg]<\infty, \qquad i=1,\ldots,n.
\end{equation}

The main advantage of this method, compared to the forward one, consists in the
fact that a single set of permutations is used to estimate the entire solution
vector.  Unless only a small portion of the problem is of interest, this
property often leads to the adjoint method being favored over the forward
method. 
 In other terms, the adjoint method should be preferred when approximating
the solution globallly over the entire computational domain, while the forward
method is especially useful when approximating the solution locally.

In literature another estimator is employed along with the adjoint Monte Carlo
method, the so called \textit{expected value} estimator. Its
formulation is as follows: it is the vector
$\boldsymbol{\theta}\in \mathbb{R}^n$ such that
\begin{equation}
\begin{array}{rl}
\theta_i & =E\bigg[f_i + \sum_{\ell=0}^\infty
W_{\ell}H_{k_{\ell}, i}^T\bigg]\\
&=f_i 
+{\displaystyle \sum_{\ell=0}^{\infty}\sum_{k_0=1}^n\sum_{k_1=1}^n\sum_{k_2=1} ^n\cdots\sum_ 
{ k_ { \ell}=1}^n
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots 
P_{k_{\ell-1},k_{\ell}}w_{k_0,k_1}\cdots
w_{k_{\ell-1},k_{\ell}}H_{k_{\ell},i}^T.}
\label{adj_mean1}
\end{array}
\end{equation}

Hence, the expected value estimator averages
the deterministic contribution of the iteration matrix over all the potential
states $j$ that might be reached from the current state $\ell$. The variance
in this case becomes:
\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
H_{k_{\ell},i}^T\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
{H_{k_{\ell},i}^T}^T\bigg ] - \bigg 
(E\bigg[\sum_{\ell=0}^\infty
W_{\ell}
H_{k_{\ell},i}^T\bigg]\bigg )^2, \qquad i=1,\ldots,n
\label{adj_var1}.
\end{equation}

\subsection{Hybrid stochastic/deterministic methods}

The direct methods described in Sections~\ref{subsec:forward} and
\ref{subsec:adjoint} suffer from a slow rate of convergence due to the
$\frac{1}{\sqrt{N}}$ behavior dictated by the central limit theorem ($N$ here
is the number of random walks used to estimate the solution).  Furthermore,
when the spectral radius of the iteration matrix is close to unity, each
individual random walk may require a large number of transitions to
approximate the corresponding components in the Neumann series.
To offset the slow convergence of the central limit theorem, schemes have
been proposed which combine traditional fixed point iterative methods with
the stochastic solvers.  The first such method, due to Halton, was termed
the Sequential Monte Carlo (SMC) method, and can be written as follows.

\begin{algorithm}[H]
 %\SetAlgoLined
 %\BlankLine
 \LinesNumbered
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$\mathbf{x}_{num}$}
 $l=0$\;
 \While{not reached convergence}{
  $\mathbf{r}^{l}=\mathbf{b}-A\mathbf{x}^{l}$\;
  $\delta \mathbf{x}^{l+1}\approx(I-H)^{-1}\mathbf{r}^{l}$; \qquad 
\text \% {Computed via Standard MC}
  $\mathbf{x}^{l+1}=\mathbf{x}^{l}+\delta \mathbf{x}^{l+1}$\;
  $l=l+1$\;
 }
 $\mathbf{x}_{num}=\mathbf{x}^{l+1}$\;
 \caption{Sequential Monte Carlo}
\end{algorithm}

The Monte Carlo linear solver method is used to compute the update $\delta
\mathbf{x}^{l}$.  This algorithm is equivalent to a Richardson iteration
accelerated by a correction obtained by approximately solving the error-residual
equation
\begin{equation}
(I - H)\delta \mathbf{x}^{l+1} = \mathbf{r}^{l}.
\label{eq:corr}
\end{equation}
If this equation were to be solved exactly, the corresponding approximation
$\mathbf{x}^{l+1}=\mathbf{x}^{l}+\delta \mathbf{x}^{l+1}$ would be the exact
solution to the linear system. This is of course impractical, since solving
(\ref{eq:corr}) is equivalent to solving the original linear system
(\ref{linsys}). Instead, the correction is obtained by
solving (\ref{eq:corr}) only approximately, using a Monte Carlo method.
Because Monte Carlo is only applied within a single iteration, the central
limit theorem is only applicable within that iteration rather than to
the overall convergence behavior of the algorithm. This allows a trade-off
between the amount of time and effort spent on the inner (stochastic) and
outer (deterministic) iterations, which can take account the competing goal
of reliability and rapid convergence.

A further extension of Halton's method, termed
\textit{Monte Carlo Synthetic Acceleration} (MCSA), has been recently
introduced in \cite{ESW2013} and \cite{EMSH2014}.  The MCSA algorithm can be 
written as:

\begin{algorithm}[H]
\LinesNumbered
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$\mathbf{x}_{num}$}
 $l=0$\;
 \While{not reached convergence}{
  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
 $\delta\mathbf{x}^{l+\frac{1}{2}}\approx(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$; 
\qquad
\% \text{Computed via Standard MC}
  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta
\mathbf{x}^{l+\frac{1}{2}}$\;
 $l=l+1$\;
 }
 $\mathbf{x}_{num}=\mathbf{x}^{l+1}$\;
 \caption{Monte Carlo Synthetic Acceleration}
\end{algorithm}

As with SMC, a Monte Carlo linear solver is used to compute the updating
contribution $\delta \mathbf{x}^{l+\frac{1}{2}}$.  In this approach, an
extra step of Richardson iteration is added to smooth out
some of the high-frequency noise introduced by the Monte Carlo process.
This way, the deterministic and stochastic components of the algorithm
act in a complementary fashion.

Obviously, a minimum requirement is that the linear system can be written
in the form (\ref{fixedpoint}) with $\rho(H) < 1$. This is typically achieved by
preconditioning. That is, we find an invertible matrix $P$ such that $H = I - P^{-1}A$
satisfies $\rho(H) < 1$, and we apply the method to the fixed point problem
(\ref{fixedpoint}) where $H = I - P^{-1}A$ and $\mathbf{f} = P^{-1}\mathbf{b}$. In
other words, the underlying deterministic iteration is a preconditioned
Richardson iteration. Various choices of the preconditioner are possible;
a detailed discussion of this issue is deferred until Section \ref{sec:prec}.
Here we note only that because we need explicit 
knowledge of the entries of $H$, not all preconditioning choices are
viable; in particular, $P$ needs to be such that $H = I - P^{-1}A$ retains
a high degree of sparsity. Unless otherwise specified, below we
assume that the transformation of the original linear system (\ref{linsys})
to the fixed point form (\ref{fixedpoint}) with $\rho(H)<1$ has already
been carried out.

\section{Convergence behavior of stochastic methods}
\label{sec:convergence}

Interestingly,
the convergence requirements imposed by the Monte Carlo estimator and
the corresponding variance can be reformulated in a purely deterministic
setting.
For instance, the condition of finiteness of the expected value turns out to 
be equivalent to requiring
\begin{equation}
 \rho(H)<1
 \label{rho_h},
\end{equation}
where $H$ is the iteration matrix of the fixed point scheme.
Indeed, we can see from (\ref{dir_mean}) and (\ref{adj_mean}) that the
expected value is expressed in terms of
power series of $H$, and the condition $\rho(H)<1$ 
is a necessary and sufficient condition for the Neumann series to
converge.

Next, we address the finiteness requirement for the second moment.
Equations \eqref{dir_var} and \eqref{adj_var}
for
the forward and the adjoint method, respectively, show that the second
moment can be reinterpreted as a power series with respect to the matrices
defined as follows:
\[
 \hat{H}_{ij}=\frac{H_{ij}^2}{P_{ij}} \quad \text{- Forward Method}
\]
and
\[
 \hat{H}_{ij}=\frac{H_{ji}^2}{P_{ij}} \quad \text{- Adjoint Method}.
\]
In order for the corresponding power series to
converge, we must require
\begin{equation}
 \rho(\hat{H})<1.
 \label{rho_star}
\end{equation}
Hence, condition \eqref{rho_h} is required for a generic fixed point scheme to
reach convergence, whereas the extra condition \eqref{rho_star} is typical of
the stochastic schemes studied in this work.
Moreover, since the finiteness of the variance automatically entails the 
finiteness of the expected value, we can state that \eqref{rho_star} 
implicitly entails \eqref{rho_h}, whereas the converse is not true in 
general.

\subsection{Necessary and sufficient conditions}

Here we report some results presented in 
\cite{Srin2010} and \cite{MASC2013},
concerning necessary conditions and sufficient conditions for convergence.
In particular, these papers discuss suitable choices for constructing the
transition probability matrix, $P$.

The construction of the transition probability must obviously satisfy the following 
constraints (called \textit{transition conditions}):
\[
\begin{cases}
  P_{ij}\ge 0 \\
 \sum_{j=1}^N P_{ij}= 1 \,.
\end{cases}
\]
One additional requirement relates the sparsity pattern of $H$ to that
of the transition probabilities:
\begin{align*}
& \text{Forward Method:} \quad H_{ij}\ne 0 \Rightarrow P_{ij}\ne 0, \\ &
\text{Adjoint Method:} \quad H_{ji}\ne 0 \Rightarrow P_{ij}\ne 0.
\end{align*} 
The following auxiliary result can be found in \cite{MASC2013}.

\begin{lemma}
 Consider a generic vector 
$\mathbf{g}=(g_1,g_2,\ldots,g_N)^T$ where at least 
one element is
nonzero, $g_k\ne0$ for some $k\in\{1,\ldots,N\}$. %$N$ represents the 
%arbitrary length of the vector $\mathbf{g}$. 
Then, the following statements hold:
\begin{enumerate}
 \item[(i)] for any probability distribution vector 
$\boldsymbol{\beta}=(\beta_1,\beta_2,\ldots,
\beta_N)^T$ satisfying the
transition conditions, $\displaystyle
\sum_{k=1}^N\frac{g_k^2}{\beta_k}\ge \bigg(\sum_{k=1}^N 
\lvert g_k\rvert\bigg)^2$; moreover, the lower bound is attained for
the probability vector defined by $\displaystyle \beta_k=\frac{\lvert
g_k\rvert}{\sum_{k=1}^N \lvert g_k\rvert}$;
\item[(ii)] there always exists a probability vector 
$\boldsymbol{\beta}$ such that $\displaystyle
\sum_{k=1}^N\frac{g_k^2}{\beta_k}\ge c$, for all 
$c>1$.
\end{enumerate}
\label{lemma1}
\end{lemma}

Consider now a generic realization of a random walk, truncated at a certain $k$th step:
\[
% \gamma_k:\; r_0\rightarrow r_1 \rightarrow r_2 \rightarrow \cdots \rightarrow r_k,
 \nu_k:\; r_0\rightarrow r_1 \rightarrow r_2 \rightarrow \cdots \rightarrow r_k,
\]
and the corresponding statistical estimator associated with the Forward Monte 
Carlo method:
\[
% X(\gamma_k)=\frac{H_{r_0,r_1}H_{r_1,r_2}\cdots
 X(\nu_k)=\frac{H_{r_0,r_1}H_{r_1,r_2}\cdots
H_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots P_{r_{k-1},r_k}}f_{r_k}.
\]
Then, the following result holds (see \cite{MASC2013}).
\begin{thm}\textit{(Forward method version)}
Let $H\in \mathbb{R}^{n\times n}$ be such that $\lVert H\rVert_{\infty}<1$.
%Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the
Consider $\nu_k$ as the realization of a random walk $\nu$ truncated at the
$k$th step. Then,
there always exists a
transition matrix $P$ such that
$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and
$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
\label{for_thm}
\end{thm}

If we introduce the estimator associated with the Adjoint Monte Carlo:
\[
% X(\gamma_k)=\frac{H^T_{r_0,r_1}H^T_{r_1,r_2}\cdots
 X(\nu_k)=\frac{H^T_{r_0,r_1}H^T_{r_1,r_2}\cdots
H^T_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots
P_{r_{k-1},r_k}}\text{sign}(f_{r_0})\lVert \mathbf{f}\rVert_1,
\]
then we can state a
theorem analogous to \ref{for_thm} (see \cite{MASC2013}).

\begin{thm}\textit{(Adjoint method version)}
 Let $H\in \mathbb{R}^{n\times n}$ with $\lVert H\rVert_{1}<1$.
%Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the
Consider $\nu_k$ as the realization of a random walk $\nu$ truncated at the
$k$th step. Then,
there always exists a
transition matrix $P$ such that
$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and
$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
\label{adj_thm}
\end{thm}

These results represent sufficient (but not necessary) conditions
for the convergence
of the forward and adjoint Monte Carlo and can be easily verified if $H$ is explicitly
available.
However, in many cases the conditions $\lVert H\rVert_{\infty}<1$ or
$\lVert H\rVert_1<1$ may be too restrictive.

The connection between Lemma \ref{lemma1} and Theorems \ref{for_thm}-\ref{adj_thm}
will be explained in 
the next section, dedicated to the definition of transition probabilities.

\subsection{Construction of transition probabilities}

The way the transition probability is defined has a significant impact
on the properties of the resulting algorithm, and in many circumstances the
choice can make the difference between convergence or divergence of the stochastic
scheme.  Two main approaches have been considered in the literature:
\textit{uniform probabilities} and \textit{weighted probabilities}.
We discuss these next.

\subsubsection{Uniform probabilities}

With this approach, the transition matrix $P$ is such that all the 
transitions corresponding to each row have equal probability of occurring:
\[
\text{Forward}:\;P_{ij}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{ij}=0, \\
\frac{1}{\#(\text{non-zeros in row $i$ of $H$)}} \quad \text{if} \quad H_{ij}\ne 0;
\end{cases}\quad
\]
\[
\text{Adjoint}:\;P_{ij}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{ji}=0, \\
\frac{1}{\#(\text{non-zeros in column $i$ of $H$)}} \quad \text{if} \quad H_{ji}\ne 0.
\end{cases}
\]
The Monte Carlo approach resorting to this definition of the transition matrix,
in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).

\subsubsection{Weighted probabilities}

An alternative definition of transition matrices aims to associate nonzero 
probability to the nonzero entries of $H$ accordingly to their magnitude. For 
instance, we may employ the following definition:
\[
\text{Forward}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{ij}=\frac{\lvert
H_{ij}\rvert^p}{\sum_{k=1}^n
\lvert H_{ik}\rvert^p,}
\]
and
\[
\text{Adjoint}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{ij}=\frac{\lvert
H_{ji}\rvert^p}{\sum_{k=1}^n
\lvert H_{ki}\rvert^p} \,,
\]
where $p\in \mathbb{N}$.
The case $p=1$ is called
\textit{Monte Carlo Almost Optimal} (MAO).
The reason for the ``almost optimal'' designation can be understood looking 
at
Lemma~\ref{lemma1}, as the quantity
%$\displaystyle \sum_{k=1}^N\frac{\mathbf{g}_k^2}{\boldsymbol{\beta}_k}$
$\sum_{k=1}^N\frac{g_k^2}{\beta_k}$
is minimized when the probability vector is
defined as $\displaystyle \beta_k=\frac{\lvert 
g_k\rvert}{\sum_{k=1}^N \lvert
g_k\rvert}$. Indeed, Lemma~\ref{lemma1} implies that the almost optimal probability 
minimizes the
$\infty$-norm of $\hat{H}$ for the forward method and the $1$-norm of $\hat{H}$
for the adjoint method, since the role of $\mathbf{g}$ in Lemma~\ref{lemma1} is played 
by the rows of $H$ in the former case and by the columns of $H$ in the latter one. 
This observation provides us with easily computable upper bounds for 
$\rho(\hat{H})$.

\subsection{Classes of matrices with guaranteed convergence}

On the one hand, sufficient conditions for
convergence of Monte Carlo linear solvers are very restrictive; see, e.g.,
\cite{Srin2010} and \cite{MASC2013}. On the other hand, the necessary and 
sufficient 
condition in \cite{MASC2013} requires knowledge of $\rho(\hat{H})$, which 
is not readily available. Note that explicit computation of $\rho(\hat{H})$
is quite expensive, comparable to the cost of solving the original linear system.
While ensuring that $\rho(H)<1$ (by means with appropriate preconditioning)
is in many cases possible, guaranteeing that $\rho(\hat{H})<1$ is generally
much more problematic.

Here we identify matrix types for which both conditions can be satisfied
by an appropriate choice of splitting, so that convergence of the Monte Carlo scheme
is guaranteed.

%\subsubsection{Strictly Diagonally Dominant}
\subsubsection{SDD matrices}
\label{sec:sdd}

One of these categories is represented by strictly diagonally dominant 
(SDD) matrices. We investigate under which conditions diagonal preconditioning
is enough to ensure convergence.
We recall the following definitions.

\begin{defn}
 A matrix $A\in\mathbb{R}^{n\times n}$ is strictly diagonally dominant 
by rows if
 \begin{equation}
    \lvert a_{ij}\rvert>\sum_{\substack{i=1\\ i\ne j}}^{n}\lvert
a_{ij}\rvert .
\label{sddr}
 \end{equation}
\end{defn}

\begin{defn}
 A matrix $A\in\mathbb{R}^{n\times n}$ is strictly diagonally dominant
by columns if $A^T$ is strictly diagonally dominant by rows, i.e., 
 \begin{equation}
    \lvert a_{ij}\rvert>\sum_{\substack{j=1\\ j\ne i}}^{n}\lvert
a_{ij}\rvert .
\label{sddc}
 \end{equation}
\end{defn}

Suppose $A$ is SDD by rows. Then we can apply left diagonal (Jacobi) preconditioning, 
obtaining an iteration matrix $H=I-P^{-1}A$ such that $\lVert H 
\rVert_{\infty}<1$.
Introducing a MAO transition probability for the forward method: 
\[
 P_{ij}=\frac{\lvert H_{ij}\rvert}{\sum_{k=1}^n\lvert H_{ik}\rvert} 
\]
we have that the entries of $\hat{H}$ are defined as follows:
\[
 \hat{H}_{ij}=\frac{H^2_{ij}}{P_{ij}}=\lvert H_{ij}\rvert\bigg
(\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ).
\]
Consequently,
\[
 \sum_{j=1}^n\lvert \hat{H}_{ij} \rvert = \sum_{j=1}^n \hat{H}_{ij} =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg )\bigg (\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ) =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg ) ^2 <1, \qquad \forall i=1,\cdots, n.
\]
This implies that $\rho(\hat{H})\le \lVert
\hat{H}\rVert_{\infty}<1$, guaranteeing the forward Monte Carlo converges.
However, nothing can be said {\em a priori}
about the convergence of the adjoint method.

On the other hand, if \eqref{sddc} holds, we can apply right diagonal 
(Jacobi) preconditioning, which
results in an iteration matrix $H=I-AP^{-1}$ such that $\lVert H 
\rVert_{1}<1$.
In this case, by a similar reasoning we
conclude that the adjoint method converges, owing to $\lVert
\hat{H}\rVert_1<1$; however,
nothing can be said \textit{a priori} about the forward method.

Finally, it is clear that if $A$ is SDD by rows {\em and} by columns, then 
a (left or right) diagonal preconditioning will result in the convergence
of both the forward and the adjoint Monte Carlo schemes.

\subsubsection{GDD matrices}
\label{sec:gdd}

Another class of matrices for which the convergence of
MC solvers is ensured is that of generalized diagonally
dominant (GDD) matrices. We recall the following definition.

\begin{defn}
A square matrix $A\in\mathbb{C}^{n\times n}$ is said to be {\em generalized
diagonally dominant} if
\[
 \lvert a_{ii}\rvert d_i \ge \sum_{\substack{j=1\\j\ne i}}^n \lvert
a_{ij}\rvert
d_j, \quad i=1,\ldots,n,
\]
for some positive vector $\mathbf{d}=(d_1,\ldots,d_n)^T$.
\end{defn}

Note that this means that there exists a diagonal matrix  $\Delta$
such that $A\Delta$ is SDD.
A proper subclass of the class of GDD matrices is represented
by the nonsingular $M$-matrices.
Recall that $A$ is a nonsingular $M$-matrix if it is of the form
$A=r I - B$ where $B$ is nonnegative and $r>\rho(B)$.
It can be shown (see, e.g., \cite{BP1979}) that  a matrix
$A\in\mathbb{R}^{n\times n}$ is a nonsingular $M$-matrix if and only
if there exists a positive diagonal matrix $\Delta$ such that
$A\Delta$ is SDD by rows. Clearly, every 
nonsingular $M$-matrix is GDD.

It is well known (see, e.g., \cite{Ax1994}) that the classical
Jacobi, Block Jacobi and Gauss--Seidel splittings are
convergent if $A$ is a nonsingular $M$-matrix.
However, this is not enough to ensure the
convergence of MC schemes based on the corresponding fixed point
(preconditioned Richardson) iteration, since in general
we cannot expect that $\rho(\hat{H})<1$.

Nevertheless, if $A$ is an $M$-matrix there exist efficient methods
to determine a diagonal scaling of $A$ so that the resulting matrix
is SDD by rows. Note that the scaled matrix is still an $M$-matrix, 
therefore applying left Jacobi preconditioning to this matrix will
guarantee that both $\rho(H)<1$ and $\rho(\hat{H})<1$.

In \cite{Li2002}, the author presents a procedure to determine
whether a given matrix $A\in\mathbb{C}^{n\times n}$
is GDD (in which case the diagonal scaling that
makes $A$ GDD is produced), or not.
The algorithm can be described as follows.

\begin{algorithm}[H]
\LinesNumbered
 \KwData{matrix $A$, $a_{ii}\ne 0$, $i=1,\ldots, n$}
 \KwResult{$d_i$, $i=1,\ldots,n$}

 Compute $\displaystyle S_{i}=\sum_{\substack{j=1 \\ j\ne i}}^n\lvert
a_{ij}\rvert$,
$i=1,\ldots,n$\\
Set $t=0$\\ 
\For{$i=1,\ldots, n$}{
\If{$\lvert a_{ii}\rvert>S_i$}{$t=t+1$}
}
\uIf{$t=0$}{print ``A is not GDD"}
\uElseIf{$t=n$}{print ``A is GDD"}
\Else{
\For{$i=1,\ldots, n$}{$d_{i}=\frac{S_i+\varepsilon}{\lvert
a_{ii}\rvert+\varepsilon} \quad
\varepsilon>0,\quad j=1,\ldots,n$\\
$a_{ji}=a_{ji}\cdot d_i$}
Go to step 1
}
 \caption{Algorithm to determine whether a matrix is GDD}
\end{algorithm}

This procedure, which in practice converges very fast, turns a generalized 
diagonally dominant matrix (in particular, a nonsingular $M$-matrix) into a
strictly diagonally dominant matrix by rows. By replacing
$\displaystyle S_{i}=\sum^{n}_{\substack{j=1 \\ j\ne i}}\lvert a_{ij}\rvert$
at step 1 with
$\displaystyle S_{j}=\sum^{n}_{\substack{i=1 \\ i\ne j}}\lvert a_{ij}\rvert$
and by replacing
$a_{ji}=a_{ji}\cdot d_i$ with $a_{ji}=a_{ji}\cdot d_j$, we obtain the algorithm
that turns a GDD matrix into a matrix that is SDD by columns.

Once we have applied this transformation to the original matrix $A$
the Monte Carlo scheme combined with diagonal preconditioning
is ensured to converge.

\subsubsection{Block diagonally dominant matrices}
\label{sec:bdd}

In this section we analyze situations in which block diagonal preconditioning
can produce a convergent Monte Carlo linear solver.
Assume that $A$ has been partitioned into $p\times p$ block form, and that 
each diagonal block has size $n_i$ with $n_1 +\cdots +n_p=n$. Assume further
that each diagonal block $A_{ii}$ is nonsingular.
The iteration
matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal left
preconditioning is
\[
 H=\begin{bmatrix}0_{n_1\times n_1} & -A_{11}^{-1}A_{12} & \cdots &
\cdots & -A_{11}^{-1}A_{1p} \\
-A_{22}^{-1}A_{21} & 0_{n_2\times n_2} & -A_{22}^{-1}A_{23} &
\cdots & -A_{22}^{-1}A_{2p}\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
-A_{pp}^{-1}A_{p1} &  \cdots & \cdots&
-A_{pp}^{-1}A_{p,p-1} & 0_{n_p \times n_p}
\end{bmatrix}.
\]

Below, we denote with ``$i|m$" the 
modulo operation applied to the integers $i$ and $m$. The 
symbol ``$\lfloor \cdot \rfloor$" stands for the floor function, as usual.

Consider first the forward method. Assuming (for ease of notation)
that all the blocks have the same size $m=n/p$,
the entries of the MAO transition probability matrix become:
\[
 P_{ij}=\frac{\lvert H_{ij} \rvert}{\sum_{k=1}^n\lvert H_{ik}
\rvert}=\frac{\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor 
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i|m,j|m}\bigg
\rvert}{\displaystyle \sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg
([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{k}{m}\rfloor}\bigg )_{i|m,k|m}\bigg \rvert}.
\]
Consequently, the entries of $\hat{H}$ are given by
\[
\begin{array}{rl}
\hat{H}_{ij} & = \lvert H_{ij}\rvert\bigg(\sum_{k=1}^n\lvert H_{ik}\rvert\bigg)\\
& =
\bigg \lvert \bigg ([A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i|m,j|m}\bigg
\rvert
{\displaystyle
\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{k}{m}\rfloor}\bigg )_{i|m,k|m}\bigg \rvert.
}
\end{array}
\]
Computing the sum over a generic row of $\hat{H}$, we obtain 
\[
 \sum_{j=1}^n \lvert \hat{H}_{ij}\rvert=\sum_{j=1}^n \hat{H}_{ij} =
 \bigg ( \sum_{\substack{j=1\\j\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i|m,j|m}\bigg \rvert \bigg ) ^2.
\]
Consider now the quantity $\lVert \hat{H}\rVert_{\infty}$. Clearly,
\[
 \lVert \hat{H}\rVert_{\infty}<1 \Leftrightarrow \sum_{\substack{j=1\\j\ne
i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i|m,j|m}\bigg \rvert <1, \qquad \forall
i=1,\ldots,n.
\]
 A sufficient condition for this to happen is that
 \begin{equation}
  \sum_{\substack{k=1\\k\ne h}}^p \lVert A_{hh}^{-1}A_{hk}\rVert_{\infty}<1,
    \label{block_cs}\qquad \forall h=1,\ldots,p.
 \end{equation}
Introducing the matrix $\tilde{H}\in \mathbb{R}^{p\times p}$ defined as
\[
 \tilde{H}=\begin{bmatrix}0 & \lVert A_{11}^{-1}A_{12}\rVert_{\infty} &
\cdots &
\cdots & \lVert A_{11}^{-1}A_{1p}\rVert_{\infty} \\
\lVert A_{22}^{-1}A_{21}\rVert_{\infty} & 0 & \lVert
A_{22}^{-1}A_{23}\rVert_{\infty} &
\dots & \lVert A_{22}^{-1}A_{2p}\rVert_{\infty} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert A_{pp}^{-1}A_{p1}\rVert_{\infty} &  \cdots & \cdots&
\lVert A_{pp}^{-1}A_{p,p-1}\rVert_{\infty} & 0
\end{bmatrix}
\]
we can formulate a
sufficient condition on the convergence of the forward Monte
Carlo scheme:
\begin{equation}
 \lVert \tilde{H} \rVert_{\infty}<1 \Rightarrow \lVert \hat{H}
\rVert_{\infty}<1.
\end{equation}
Note that (\ref{block_cs}) also implies that $\|H\|_{\infty} < 1$.

We now turn our attention to the adjoint method.
Analogously to the forward method, we can define
\[
(\hat{H})^T_{ij} = \lvert H^T_{ij}\rvert\bigg(\sum_{k=1}^n\lvert
H^T_{ik}\rvert\bigg) = \lvert H_{ji}\rvert\bigg(\sum_{k=1}^n\lvert
H_{ki}\rvert\bigg).
\]
This allows us to formulate a sufficient condition for the convergence of 
the adjoint Monte Carlo method with block diagonal preconditioning.
Letting
\[
  \tilde{H}=\begin{bmatrix}0 & \lVert A_{11}^{-1}A_{12}\rVert_{1} & \cdots
&
\cdots & \lVert A_{11}^{-1}A_{1p}\rVert_{1} \\
\lVert A_{22}^{-1}A_{21}\rVert_{1} & 0 & \lVert
A_{22}^{-1}A_{23}\rVert_{1} &
\cdots & \lVert A_{22}^{-1}A_{2p}\rVert_{1} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert A_{pp}^{-1}A_{p1}\rVert_{1} &  \cdots & \cdots&
\lVert A_{pp}^{-1}A_{p,p-1}\rVert_{1} & 0
\end{bmatrix},
\]
a sufficient condition is that $\lVert \tilde{H} \rVert_{1}<1$, i.e.,
 \begin{equation}
  \sum_{\substack{h=1\\h\ne k}}^p \lVert A_{hh}^{-1}A_{hk}\rVert_1<1,
    \label{block_cs_2}\qquad \forall k=1,\ldots,p.
 \end{equation}
Again, this condition also implies that $\|H\|_1 < 1$.

We say that $A$ is {\em strictly block diagonally dominant by rows (columns) 
with respect to a given block partition}
if condition (\ref{block_cs}) (respectively, (\ref{block_cs_2})) is satisfied 
relative to that particular block partition.
Note that a matrix may be strictly block diagonally dominant with
respect to one partition and not to another.
We note that these definitions of block diagonal dominance are different from 
those found in, e.g., \cite{FV62}, and they are easier to check in practice
since they do not require computing the 2-norm of the blocks of $H$.

\subsection{Adaptive methods} \label{sec:adapt}

In formulas (\ref{dir_mean}) and (\ref{adj_mean}), the estimation of the
solution to
the linear system (\ref{linsys}) involves infinite sums, which in actual
computation have to be truncated. In the following we discuss criteria
to decide the number of steps to be taken in
a single random walk as well as the number of random walks that
need to be performed at each Richardson iteration.

\subsubsection{History length}

We first consider criteria to terminate an individual random walk, effectively
deciding how many terms of the Neumann series will be considered.
One possibility is to set a predetermined history length, at which point all
histories are terminated.  This approach, however, presents two difficulties.
First, it is difficult to determine a priori how many steps on average will be
necessary to achieve a specified tolerance.  Second, due to the stochastic
nature of the random walks, some histories will retain important information
longer than others.  Truncating histories at a predetermined step runs the
risk of either prematurely truncating important histories, leading to larger
errors, or continuing unimportant histories longer than necessary,
leading to computational inefficiency.

Our goal is to apply a cutoff via an automatic procedure, without requiring
any user intervention. We would like to determine
an integer $m$ such that
\[
\begin{array}{rl}
\tilde{\theta}&=E\bigg[\sum_{\ell=0}^m W_{\ell}
f_{k_{\ell}}\bigg]\\
& = f_i + {\displaystyle \sum_{\ell=1}^m
\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n
P_{i,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},
k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots
w_{k_{\ell-1}, k_{\ell}} f_{k_{\ell}}
}
\end{array}
\]
and %$\tilde{\boldsymbol{\theta}}\in \mathbb{R}^n$ s.t.
\[
\begin{array}{rl}
\tilde{\theta}_i&=E\bigg[\sum_{\ell=0}^m W_{\ell}\delta_{k_{\ell},
j}\bigg]\\
& ={\displaystyle \sum_{\ell=0}^{m}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots 
P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},i}
}
\end{array}
\]
are good approximations of (\ref{dir_mean}) and (\ref{adj_mean}), respectively.

In \cite{Slattery2013} a criterion was given which is applicable to both 
forward and adjoint methods. It requires to set up a relative weight cutoff 
threshold $W_c$ and to look for a step $m$ such that
\begin{equation}
W_m \le W_c W_0 \,.
\label{cutoff}
\end{equation}
In (\ref{cutoff}), $W_0$ is the value of the weight at the initial step of the random walk
and $W_m$ is the value of the weight after $m$ steps.
We will adopt a similar strategy in this work.

\subsubsection{Number of random walks}

We now consider the selection of the number of random walks that should be
performed to achieve a given accuracy. Unlike the termination of histories,
this is a subject that has not been discussed in the Monte Carlo linear
solver literature, as all previous studies have considered the simulation
of a prescribed number of histories.

The expression for the variance of the forward method is given by
formula (\ref{dir_var}).
In this context, a reasonable criterion to determine the number $\tilde{N}_i$
of random walks to be run is to set a threshold $\varepsilon_1$ and determine
$\tilde{N}_i$ such that
\begin{equation}
\frac{\sqrt{Var[\theta_i]}}{\lvert
E[\theta_i]\rvert}<\varepsilon_1, \quad i=1,\ldots,n.
\label{forward_adapt}
\end{equation}
The dependence of $Var[\theta_i]$ and $E[\theta_i]$ on $\tilde{N}_i$
is due to the fact that
$\theta_i$ is estimated by performing a fixed  number of histories.
Therefore, we are controlling the relative standard deviation, requiring it not
to be too large. In other words, we are aiming for an approximate solution in which 
the uncertainty factor is small relative to the expected value.
This simple adaptive approach can be applied for the estimation of each
component $x_i$. Hence, a different number of histories may be employed to
compute different entries of the solution vector.

As concerns the adjoint method, the estimation of the variance is 
given in formula \eqref{adj_var}.
A possible criterion for the adaptive selection of the number $\tilde{N}$ of random
walk, in this situation, is that it satisfies the condition
\begin{equation}
\frac{\lVert
\boldsymbol{\sigma}^{\tilde{N}}\rVert_1}{\lVert
\mathbf{x}\rVert_1}<\varepsilon_1,
\label{adjoint_adapt}
\end{equation}
where $\boldsymbol{\sigma}$ is a vector whose entries are
$\sigma^{\tilde{N}}_i=Var[\theta_i]$ and $\mathbf{x}$ is a vector whose entries 
are $x_i = E[\theta_i]$.

The criteria introduced above can be exploited to build
an a posteriori adaptive algorithm, capable of identifying the minimal value of
$\tilde{N}$ that verifies \eqref{forward_adapt} or \eqref{adjoint_adapt},
respectively.
Algorithms \ref{alg:adaptive_for} and \ref{alg:adaptive_adj} describe
the Monte Carlo approaches with the adaptive criteria.

\begin{algorithm}[H]
\LinesNumbered
 \KwData{$N$, $\varepsilon_1$}
 \KwResult{$\tilde{N}_i$, $\sigma_i$, $x_i$}
 \For{$i=1,\ldots,n$}
 {
 $\tilde{N}_i=N$\;
 compute $x_i$\;
 compute $\sigma_i$\;
 \While{$\frac{\sigma_i}{\lvert
x_i\rvert}<\varepsilon_1$}{
\vspace{0.1cm}
  $\tilde{N}=\tilde{N}+N$\;
  compute $x_i$\;
  compute $\sigma_i$\;
 }
 }
 \caption{A posteriori adaptive Forward Monte Carlo \label{alg:adaptive_for}}
\end{algorithm}

\begin{algorithm}[H]
\LinesNumbered
 \KwData{$N$, $\varepsilon_1$}
 \KwResult{$\tilde{N}$, $\boldsymbol{\sigma}$, $\mathbf{x}$}
 $\tilde{N}=N$\;
 compute $\mathbf{x}$\;
 compute $\boldsymbol{\sigma}$\;
 \While{$\frac{\lVert\boldsymbol{\sigma} \rVert_1}{\lVert
\mathbf{x}\rVert_1}<\varepsilon_1$}{
\vspace{0.1cm}
  $\tilde{N}=\tilde{N}+N$\;
  compute $\mathbf{x}$\;
  compute $\boldsymbol{\sigma}$\;
 }
 \caption{A posteriori adaptive Adjoint Monte Carlo \label{alg:adaptive_adj}}
\end{algorithm}

The use of the adaptive approach for the selection of the number of histories 
has a dual purpose. First, it guarantees that the update computed with the 
Monte Carlo step is accurate enough to preserve convergence. Second, it 
provides the user with a tuning parameter to distribute the computation 
between the deterministic and the stochastic part of the algorithm. 
Lowering the value of the threshold for the relative standard deviation 
increases the number of histories per iteration. This results in a more accurate
stochastic updating and reduces the iterations necessary to converge. 
While guessing an a priori fixed number of histories may lead to 
a smaller number of Monte Carlo histories overall, it might either hinder 
the convergence or distribute too much computation on the deterministic side 
of the scheme (or both). 
Generally speaking, the adaptive approach is more robust and more
useful, especially when $\rho(H)$ is close to 1, 
since in this case the Richardson step is less effective in dampening the uncertainty 
coming from the previous iterations.

\subsection{Preconditioning} \label{sec:prec}

As noted at the end of Section \ref{sec:mcls}, left
preconditioning can be incorporated into any Monte Carlo linear solver
algorithm by simply substituting $A$ with $P^{-1}A$ and
${\bf b}$ with $P^{-1}{\bf b}$ in (\ref{linsys}); i.e., 
we set $H = I - P^{-1}A$   
and ${\bf f} = P^{-1}{\bf b}$ in (\ref{fixedpoint}).
Right preconditioning can also be incorporated by rewriting 
(\ref{linsys}) as $AP^{-1}{\bf y} = {\bf b}$, with ${\bf y} = P{\bf x}$;
i.e., we set $H=I-AP^{-1}$ and replace $\bf x$ by $\bf y$ in (\ref{fixedpoint}).  
The solution $\bf x$ to the original system (\ref{linsys}) is then 
given by ${\bf x} = P^{-1}{\bf y}$. Likewise, split (left and right)
preconditioning can also be used. The Monte Carlo process, however,
imposes some constraints on the choice of preconditioner.  Most
significantly, because the transition probabilities are built based on the
values of the iteration matrix $H$, it is necessary to have access to the
entries of the preconditioned matrix $P^{-1}A$ (or $AP^{-1}$).
Therefore, we are limited to preconditioners that enable explicitly
forming the preconditioned matrix while retaining some of the sparsity of the
original matrix.

One possible preconditioning approach involves either diagonal or block
diagonal preconditioning (with blocks of small or moderate size). Diagonal
preconditioning does not alter the sparsity of the original coefficient matrix,
while block diagonal preconditioning will incur a moderate amount of
fill-in in the preconditioned matrix if the blocks are not too large.
From the discussions in Section~\ref{sec:sdd}
and \ref{sec:bdd}, selecting a diagonal or block diagonal preconditioner 
guarantees convergence of the Monte Carlo schemes
for matrices that are strictly diagonally dominant
or block diagonally dominant, respectively. In addition, $M$-matrices
that are not strictly or block diagonally dominant can also be dealt with by
first rescaling $A$ so that it becomes strictly diagonally dominant, as
discussed in Section~\ref{sec:sdd}.

In principle,
other standard preconditioning approaches can also be used in an attempt
to achieve both $\rho(H)<1$ and $\rho(\hat{H})<1$ while still retaining 
sparsity in the preconditioned matrix.  One possibility is the use of incomplete
LU factorizations \cite{Benzi2002,Saad}.  If $P=LU$ is the preconditioner
with sparse triangular factors $L$ and $U$, then $P^{-1}A$ can in principle
be formed explicitly
provided that the sparsity in $L$, $U$ and $A$
is carefully exploited in the forward and back substitutions
needed to form $(LU)^{-1}=U^{-1}(L^{-1}A)$. In general, 
however, this results in a
rather full preconditioned matrix; sparsity needs to be preserved by dropping
small entries in
the resulting matrix.

Another class of preconditioners that are potentially of interest 
for use with Monte Carlo
linear solvers are approximate inverse preconditioners 
\cite{Benzi2002}.  In these algorithms, an approximation
to the inverse of $A$ is generated directly and the computation of the
preconditioned matrix reduces to one or more sparse matrix-matrix products, 
a relatively
straightforward task.  As with ILU factorizations,
multiple versions of approximate inverse preconditioning exist which may
have different behavior in terms of effectiveness of the preconditioner
versus the resulting reduction in sparsity.

A downside of the use of ILU or approximate inverse preconditioning is that
the quality of preconditioner needed to achieve both $\rho(H)<1$
and $\rho(\hat{H})<1$ is difficult to determine.
Indeed, in some situations it may happen that
modifying the preconditioner so as to reduce $\rho(H)$ may actually lead to an
increase in $\rho(\hat{H})$, decreasing the effectiveness of the Monte
Carlo process on the system or even causing it to diverge.
In other words, for both ILU and sparse approximate inverse
preconditioning it seems to be very difficult to guarantee convergence
of the Monte Carlo linear solvers a priori.

\section{Numerical results}
\label{sec:results}

In this section we discuss the results of numerical experiments. The main goal 
of these experiments is to gain some insight into the behavior of adaptive
techniques, and to test different preconditioning options within the 
Monte Carlo approach.

\subsection{Numerical tests with adaptive methods}

In this subsection we study experimentally the 
adaptive approaches discussed in Section \ref{sec:adapt}.  
For this purpose, we restrict our attention to standard Monte Carlo linear
solvers.

For these tests we limit ourselves
to small matrices, primarily because the numerical experiments are
being computed on a standard laptop and the computational cost of
Monte Carlo methods rapidly becomes prohibitive on such machines.
The smallest of these matrices represents a finite
difference discretization of a 1D reaction-diffusion equation, 
the second one is a discrete 2D Laplacian with zero Dirichlet boundary
conditions, the third a steady 2D advection-diffusion operator discretized
by quadrilateral linear finite elements using the IFISS package \cite{ifiss},
and the fourth one 
results from a finite volume discretization of
a thermal radiation diffusion equation (Marshak problem). The first and the last of these
matrices are strictly diagonally dominant by both rows and columns.
For all the problems, left diagonal preconditioning is applied.
Details about these matrices are given in Table
\ref{table_data}.

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Type of problem} & $n$ &
$\rho(H)$&Forward $\rho(\hat{H})$&Adjoint $\rho(\hat{H})$\\
\hline
1D reaction-diffusion & 50 & $0.4991$ & $0.9827$ & $0.9827$\\
\hline
2D Laplacian & 900 & $0.9949$ & $0.994$ & $0.9945$\\
\hline
2D advection-diffusion & 1089 &$0.9836$ & $0.983$ & $0.983$ \\
\hline
Marshak problem & 1600 & $0.6009$ & $0.3758$ & $ 0.3815$ \\
\hline
\end{tabular}
\caption{Properties of the matrices employed for the checks on adaptivity.}
\label{table_data}
\end{table}

We present results for both the Forward
and the Adjoint Monte Carlo methods.
As concerns the Forward method, we set the maximum number of
histories per entry to $10^7$. In Tables \ref{tab:For_adapt} and \ref{tab:For_adapt2} 
we show results for values of the adaptive threshold
(\ref{forward_adapt}) equal to
$\varepsilon_1=0.1$ and $\varepsilon_1=0.01$, respectively.
A batch size of two is used at each
adaptive check to verify the magnitude of the apparent relative standard
deviation. As expected, results are aligned with the convergence rate predicted
by the Central Limit Theorem. Indeed, decreasing by a factor of 10 the
tolerance $\varepsilon_1$ we see that the relative error undergoes a
decrease of the same order, requiring roughly
one hundred times more histories. In the case of the 2D Laplacian we actually
have an increase of a factor close to 400 in the number of histories, but the
relative error is decreased by more than thirty times. For this particular
example, the forward method overestimates the number of histories
needed to satisfy a prescribed reduction on the standard deviation.

As regards the Adjoint Monte Carlo, at each adaptive check the number of
random walks employed is increased by ten.
A maximum number of histories equal to $10^{10}$ is set.
Tables \ref{tab:Adj_adapt}, \ref{tab:Adj_adapt2} and
\ref{tab:Adj_adapt3} show results
for the different test cases using adaptive thresholds (\ref{adjoint_adapt})
with values $\varepsilon_1=0.1$, $\varepsilon_1=0.01$ and
$\varepsilon_1=0.001$, respectively.
By comparing the reported errors, it is
clear that a decrease in the value of the threshold
induces a reduction of the relative error of the same order of magnitude.
This confirms
the effectiveness of the adaptive selection of histories with an error
reduction goal. Each decrease in the error by an order of magnitude
requires an increase in the total number of histories employed by
two orders of magnitude, as expected.

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1D reaction-diffusion & $0.003$ & $5,220$\\
\hline
2D Laplacian & $0.1051$ & $262,350$\\
\hline
2D advection-diffusion & $0.022$  & $1,060,770$\\
\hline
Marshak problem & $0.0012$ & $1,456,000$\\
\hline
\end{tabular}
\caption{Forward Monte Carlo. Adaptive selection of histories,
$\varepsilon_1=0.1$.}
\label{tab:For_adapt}
\end{table}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1D reaction-diffusion & $4\cdot 10^{-4}$ & $512,094$\\
\hline
2D Laplacian & $0.0032$ & $108,551,850$\\
\hline
2D advection-diffusion & $0.0023$  & $105,476,650$\\
\hline
Marshak problem & $5.8 \cdot 10^{-4}$ & $144,018,700$\\
\hline
\end{tabular}
\caption{Forward Monte Carlo. Adaptive selection of histories,
$\varepsilon_1=0.01$.}
\label{tab:For_adapt2}
\end{table}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1D reaction-diffusion & $0.08$ & $1820$\\
\hline
2D Laplacian & $0.136$ & $570$\\
\hline
2D advection-diffusion & $0.08$  & $2400$\\
\hline
Marshak problem & $0.288$ & $880$\\
\hline
\end{tabular}
\caption{Collision estimator -- Adjoint Monte Carlo. Adaptive selection of histories,
$\varepsilon_1=0.1$.}
\label{tab:Adj_adapt}
\end{table}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1D reaction-diffusion & $0.0082$ & $185,400$\\
\hline
2D Laplacian & $0.0122$ & $126,800$\\
\hline
2D advection-diffusion & $0.0093$  & $219,293$\\
\hline
Marshak problem & $0.0105$ & $650,040$\\
\hline
\end{tabular}
\caption{Collision estimator -- Adjoint Monte Carlo. Adaptive selection of histories,
$\varepsilon_1=0.01$.}
\label{tab:Adj_adapt2}
\end{table}

%\vspace{-0.1in}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1D reaction-diffusion & $9.56 \cdot 10^{-4}$ & 15,268,560 \\
\hline
2D Laplacian & $0.001$ & $12,600,000$\\
\hline
2D advection-diffusion & $0.0011$  & $23,952,000$\\
\hline
Marshak problem & $0.0011$ &  $80,236,000$\\
\hline
\end{tabular}
\caption{Collision estimator -- Adjoint Monte Carlo. Adaptive selection of histories,
$\varepsilon_1=0.001$.}
\label{tab:Adj_adapt3}
\end{table}

%\vspace{-0.1in}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1D reaction-diffusion & $0.0463$ & $1000$\\
\hline
2D Laplacian & $0.1004$ & $900$\\
\hline
2D advection-diffusion & $0.0661$  & $1300$\\
\hline
Marshak problem & $0.0526$ & $2000$\\
\hline
\end{tabular}
\caption{Expected value estimator -- Adjoint Monte Carlo. Adaptive selection of
histories,
$\varepsilon_1=0.1$}
\label{tab:Adj_adapt4}
\end{table}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1D reaction-diffusion & $0.004$ & $100,600$\\
\hline
2D Laplacian & $0.0094$ & $83,700$\\
\hline
2D advection-diffusion & $0.0088$  & $124,400$\\
\hline
Marshak problem & $0.0056$ & $166,000$\\
\hline
\end{tabular}
\caption{Expected value estimator -- Adjoint Monte Carlo. Adaptive selection of histories,
$\varepsilon_1=0.01.$}
\label{tab:Adj_adapt5}
\end{table}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{Relative Error} &\textbf{Nb. Histories}\\
\hline
1D reaction-diffusion & $0.004$ & $10,063,300$\\
\hline
2D Laplacian & $ 9.31 \cdot 10^{-4} $ & $ 8,377,500 $\\
\hline
2D advection-diffusion & $ 0.0013 $  & $ 12,435,000 $\\
\hline
Marshak problem & $7.79 \cdot 10^{-4}$ & $16,537,000$\\
\hline
\end{tabular}
\caption{Expected value estimator -- Adjoint Monte Carlo. Adaptive selection of histories,
$\varepsilon_1=0.001.$}
\label{tab:Adj_adapt6}
\end{table}

The same simulations can be run resorting to the expected value
estimator. Results are shown in the Tables \ref{tab:Adj_adapt4},
\ref{tab:Adj_adapt5} and \ref{tab:Adj_adapt6} for the threshold values of
$\varepsilon_1=0.1$, $\varepsilon_1=0.01$ and $\varepsilon_1=0.001$
respectively. As it can be noticed, in terms of error scaling the results are
quite similar to the ones obtained with the collision estimator. As regards
the number of histories needed to reach a prescribed accuracy, the
orders of magnitude are the same for both the collision and the
expected value estimators. However, the expected value estimator requires  
in most cases a smaller number of realizations. This behavior becomes more
pronounced as the value of
the threshold decreases, making the computation increasingly cost-effective.

\subsection{Preconditioning approaches}

In this subsection we examine the effect of different preconditioners
on the values attained by the spectral radii $\rho(H)$ and $\rho(\hat{H})$.
For this purpose, we focus on the 2D discrete Laplacian and the 
2D discrete advection
diffusion operator from the previous section.

The values of the two spectral radii with
diagonal preconditioning have already been shown in Table
\ref{tab:Adj_adapt}.
Here we consider the effect of 
block diagonal preconditioning for different block sizes,
and the use of the factorized sparse approximate inverse 
preconditioner AINV \cite{BMT1996,BT1998}
for different values of the drop tolerance (which controls the
sparsity in the approximate inverse factors). 
Intuitively, with these two types of preconditioners both $\rho(H)$ and $\rho(\hat{H})$
should approach zero for increasing block size and decreasing
drop tolerance, respectively; however, the convergence need
not be monotonic in general, particularly for $\rho(\hat{H})$.
This somewhat counterintuitive
behavior is shown in Tables \ref{tab:lap2d_block} and \ref{tab:ifiss_block},
where an increase in the size of the blocks used for the block diagonal
preconditioner results in an increase of $\rho(\hat{H})$ for both test problems. 
Note also the very slow rate of decrease in $\rho(H)$ for increasing block
size, which is more than offset by the rapid increase in the density of $H$,
which of course implies much higher costs.
We mention that for a block size of 30 the 2D discrete Laplacian is
block diagonally dominant, but not for smaller block sizes. The 2D
discrete advection-diffusion operator is not block diagonally dominant
for any of the three reported block sizes.

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
&$\rho(\hat{H})$\\
\hline
 5 & 2.6164 & $0.9915$ & $0.9837$\\
\hline
 10 & 5.6027 & $0.9907$ & $0.9861$\\
\hline
 30 & 16.3356 & $0.9898$  & $0.9886$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the
2D Laplacian with block diagonal preconditioning.}
\label{tab:lap2d_block}
\end{table}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Block size} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
&$\rho(\hat{H})$\\
\hline
 3 & 1.4352 & $0.9804$ & $0.9531$\\
\hline
 9 & 3.0220 & $0.9790$ & $0.9674$\\
\hline
 33 & 9.8164 & $0.9783$  & $0.9774$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the 
2D advection-diffusion problem with block diagonal preconditioning.}
\label{tab:ifiss_block}
\end{table}

In Tables \ref{tab:lap2d_ainv} and \ref{tab:ifiss_ainv} the values of
the spectral radii are shown for the AINV
preconditioner with two different values of the drop tolerance
\cite{BMT1996,BT1998}. It is interesting to point out that, for the
two-dimensional
Laplacian, a drop tolerance $\tau=0.05$ entails $\rho(H)<1$ but the same does
not hold for $\hat{H}$. Both convergence conditions are satisfied by
reducing the drop tolerance, but at the price of very high fill-in in the
preconditioned matrix.

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
&$\rho(\hat{H})$\\
\hline
 0.05 & 8.2797 & $0.9610$ & $1.0231$\\
\hline
 0.01 & 33.1390 & $0.8668$ & $0.8279$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the 
2D Laplacian with AINV preconditioning.}
\label{tab:lap2d_ainv}
\end{table}

\begin{table}[!t]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
&$\rho(\hat{H})$\\
\hline
 0.05 & 3.8392 & $0.9396$ & $0.9069$\\
\hline
 0.01 & 15.1798 & $0.7964$ & $0.6361$\\
\hline
\end{tabular}
\caption{Behavior of $\rho(H)$ and $\rho(\hat{H})$ for the 2D 
advection-diffusion problem with AINV preconditioning.}
\label{tab:ifiss_ainv}
\end{table}

In summary, we conclude that it is generally very challenging to guarantee 
the convergence of Monte Carlo linear solvers a priori. Simple (block) diagonal 
preconditioners may work even if $A$ is not strictly (block)
diagonally dominant, but it is hard to know beforehand if a method will converge,
especially due to lack of a priori bounds on $\rho(\hat{H})$.
Moreover, the choice of the block sizes in the block diagonal case
is not an easy matter. Sparse approximate inverses are a possibility
but the amount of fill-in required to satisfy the
convergence conditions could be unacceptably high. These observations
suggest that it is difficult to use Monte Carlo linear solvers in the
case of linear systems arising from the discretization of steady-state
PDEs, which typically are not strictly diagonally dominant.
As we shall see later, the situation is more favorable in the case of
time-dependent problems.

\subsection{Hybrid methods}
Next, we present results for hybrid methods, combining a deterministic
Richardson iteration with Monte Carlo acceleration. We recall that the
convergence conditions are the same as for the direct stochastic
approaches, therefore the concluding observations from the previous section
still apply.

\subsubsection{Poisson problem}

Consider the standard 2D Poisson model problem:
\begin{equation}
\begin{cases}
 -\Delta u = f \quad \text{in}\quad \Omega, \\
 u=0\quad \text{on} \quad \partial\Omega ,
 \end{cases}
\label{eq:Poisson}
\end{equation}
where $\Omega=(0,1)\times (0,1)$. % and $\sigma=1$.
For the numerical experiments we use as the right-hand side
a sinusoidal distribution in both $x$ and $y$ directions:
\[
 f(x, y)=\sin ({\pi x})\sin ({\pi y}).
\]
We discretize problem (\ref{eq:Poisson}) using standard 5-point finite
differences. For $N=32$ nodes in each direction, we obtain the $900\times 900$
linear system already used in the previous section.

Consider the iteration matrix $H$ corresponding to (left) diagonal
preconditioning. It is well known that $\rho(H) <1$, and indeed we
can see from Table \ref{table_data} that $\rho(H)\approx 0.9949$. 
It is also easy to see that $\|H\|_1 = 1$. 
In order for the Adjoint Monte Carlo
method to converge, it is necessary to 
have $\rho(\hat{H}) < 1$, too. If an
almost optimal probability is used,
the Adjoint Monte Carlo method leads to a $\hat{H}$ matrix such that
\[
 \lVert \hat{H}\rVert_1\le (\rVert H\rVert_1)^2=1.
\]
This condition by itself is not enough to guarantee that
$\rho(\hat{H})<1$.
However, the iteration matrix $H$ has zero-valued
entries on the main diagonal and it has:
\begin{itemize}
 \item four entries equal to $\frac{1}{4}$ on the rows associated
with a node which is not adjacent to the boundary;
\item two entries equal to $\frac{1}{4}$ on the rows associated with nodes
adjacent to the corner of the square domain;
\item three entries equal to $\frac{1}{4}$ on the rows associated with nodes
adjacent to the boundary on an edge of the square domain.
\end{itemize}

Recalling the definition of $\hat{H}$ in terms of the entries of the
iteration matrix $H$ and the transition probability $P$, we see that
\[
 \hat{H}=\tilde{D}H,
\]
where $\tilde{D}$ is a diagonal matrix with diagonal entries equal to
1, $\frac{1}{2}$ or $\frac{3}{4}$.
In particular, $\tilde{d}_i=\text{diag}(\tilde D)_i=1$ if the $i$th row of the
discretized Laplacian is associated with a node which is not adjacent to the
boundary, $\tilde{d}_i=\text{diag}(\tilde D)_i=\frac{1}{2}$ if the row is associated
with a node of the grid adjacent to the corner of the boundary,
and $\tilde{d}_i=\text{diag}(\tilde D)_i=\frac{3}{4}$ if the associated node of the grid
is adjacent to the boundary of the domain far from a corner.
Since $H$ is an irreducible nonnegative matrix and $\tilde{D}$ is a
positive definite diagonal matrix, we can invoke a result in 
\cite{FK1975} to conclude that
\[
 \rho(\hat{H})=\rho(\tilde{D}H)\le \rho(\tilde D)\rho(H).
\]

Since $\rho(\tilde{D})=1$, then $\rho(\hat{H})\le \rho(H)<1$.
Therefore, diagonal preconditioning always guarantees convergence
for the Monte Carlo linear solver applied to any 2D Laplace
operator discretized with 5-point finite differences. 
Similar arguments apply to the $d$-dimensional Laplacian, for any $d$.

Results of numerical experiments are reported in Table \ref{Poisson_results}.
We compare the purely deterministic Richardson iteration (Jacobi's method
in this case) with two hybrid
methods, Halton's
sequential Monte Carlo and MCSA using the Adjoint method.
The threshold
for the adaptive selection of the random walks is set
to $\varepsilon_1=0.1$.
Convergence is attained when the initial residual has been reduced
by at least eight orders of magnitude, starting from a zero initial guess.
Note that, as expected, the diagonally preconditioned Richardson iteration
converges very slowly, since the spectral radius of $H$ is very close to 1.
Halton's Sequential Monte Carlo performs quite well on this problem, but
even better results are obtained by the Adjoint MCSA method, which converges
in one iteration less than Sequential Monte Carlo while requiring far less
Monte Carlo histories per iteration. 
 This
might be explained by the fact that within each outer
iteration, the first relaxation step
in the MCSA algorithm refines the accuracy of the solution coming from the previous
iteration,
before using it as the input for the MC solver at the current iterate.
In particular, 
its effect is to dampen the statistical noise
produced by the Monte Carlo linear solver in the estimation of the update $\delta
{\bf x}^{l + \frac{1}{2}}$ performed at the previous iteration. Therefore, the refinement, 
or smoothing, accomplished by the Richardson relaxation decreases the number of
random walks needed
for a prescribed accuracy. This hypothesis is validated by the fact
that at the first iteration both Sequential Monte Carlo and MCSA use
the same number of histories. Their behaviors start differing from the
second iteration on, when the statistical noise is introduced in the estimation
of the correction to the current iterate; 
see Figures \ref{SEQ_poisson} and
\ref{MCSA_poisson} (note the semilogarithmic scale on the vertical axis). 
It can be seen that
in all the numerical tests presented in this section there is a sharp increase 
in the number of histories when going from the first to subsequent iterations. 
This is due to the introduction of statistical noise coming from the Monte Carlo 
updating of the solution. Indeed, the stochastic noise, introduced from the 
second iteration on, increases the uncertainty associated with the estimate 
of the solution. Therefore, the adaptive criterion forces the algorithm 
to perform a higher number of histories 
to achieve a prescribed accuracy.

\begin{table}[!t]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
Richardson & $9.8081\cdot 10^{-8}$ & 3133 & - \\
\hline
Sequential MC & $7.9037 \cdot 10^{-8}$ & 9 & 8.264,900\\
\hline
 Adjoint MCSA & $8.0872\cdot 10^{-8}$ & 8 & 1,738,250\\
\hline
\end{tabular}
\caption{Numerical results for the Poisson problem.}
\label{Poisson_results}
\end{table}

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{SEQ_poisson.eps}
\vspace{-0.3in}
    \caption{Sequential MC - Poisson problem. Number of random walks employed
at each iteration.}
\label{SEQ_poisson}
\end{figure}

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{MCSA_poisson.eps}
\vspace{-0.3in}
      \caption{MCSA - Poisson problem. Number of random walks employed at each
iteration.}
\label{MCSA_poisson}
\end{figure}

\subsubsection{Reaction-diffusion problem}

Here we consider the simple reaction-diffusion problem

\begin{equation}
\begin{cases}
 -\Delta u +\sigma u= f \quad \text{in}\quad \Omega, \\
 u=0\quad \text{on} \quad \partial\Omega,
 \end{cases}
\end{equation}
where $\Omega=(0,1)\times (0,1)$, $\sigma=0.1$ and $f\equiv 1$.
A 5-point finite difference scheme is applied to discretize the problem.
The number of nodes
on each direction of the domain is 100, so that $h\approx 0.01$. The
discretized problem is $n\times n$ with $n=9604$. A left
diagonal preconditioning is again applied to
the coefficient matrix obtained from the discretization, which
is strictly diagonally dominant. The 1-norm of the
iteration matrix is $\lVert H\rVert_1\approx 0.9756$. This automatically
guarantees the convergence of the Adjoint Monte Carlo linear solver. 
In Table \ref{DR_results} a comparison between the deterministic
Richardson iteration, Sequential Monte Carlo and MCSA is provided.
The results are similar to those for the Poisson equation.
The number of histories per iteration for Sequential Monte Carlo and MCSA 
is shown in Figures \ref{SEQ_diffreac} and \ref{MCSA_diffreac}.

\begin{table}[!t]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
 Richardson & $9.073\cdot 10^{-8}$ & 634 & - \\
\hline
 Sequential MC & $8.415 \cdot 10^{-8}$ &  8 & 12,391,375\\
 \hline
 Adjoint MCSA & $6.633 \cdot 10^{-8}$ &  7 & 3,163,700\\
\hline
\end{tabular}
\caption{Numerical results for the diffusion reaction problem.}
\label{DR_results}
\end{table}

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{SEQ_diffreac.eps}
      \caption{Sequential MC - Reaction-diffusion problem. Number of random
walks
employed at each
 iteration.}
\label{SEQ_diffreac}
\end{figure}

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{MCSA_diffreac.eps}
      \caption{MCSA - Reaction-diffusion problem. Number of random walks
employed at each
iteration.}
\label{MCSA_diffreac}
\end{figure}

\subsubsection{Parabolic problem}

Here we consider the 
following time-dependent problem:
\begin{equation}
 \begin{cases}\frac{\partial u}{\partial t} -\mu \Delta u
+\boldsymbol{\beta}(\mathbf{x})\cdot \nabla u=0, \;\; \quad \quad
\mathbf{x}\in \Omega,\quad t\in(0,T] \\
u(\mathbf{x},0) = u_0, \qquad \qquad \qquad \qquad \quad \mathbf{x}\in
\Omega \\
u(\mathbf{x},t)=u_D(\mathbf{x}),  \; \, \, \qquad \qquad \qquad \quad
\mathbf{x}\in \partial
\Omega, \quad t\in(0,T],
 \end{cases}
\end{equation}
where $\Omega = (0,1)\times (0,1)$, $T>0$,
$\mu=\frac{3}{200}$, $\boldsymbol{\beta}(\mathbf{x})=[2y(1-x^2),\;
-2x(1-y^2)]^T$, and $u_D=0$ on $\{\{x=0\}\times (0,1)\}$, $\{(0,1)\times
\{y=0\}\}$, $\{(0,1)\times \{y=1\}\}$.\newline

Implicit discretization in time (backward Euler scheme) with time step
$\Delta t$ and
a spatial discretization with quadrilateral linear finite elements using
the IFISS toolbox \cite{ifiss} leads
to a sequence of linear systems of the form
$$ \left (\frac{1}{\Delta t} M + A\right )\mathbf{u}^{k+1} = 
\mathbf{f}^k, \quad k=0,1,\ldots $$

Here we restrict our attention to a single generic time step.
The right-hand side is chosen so that the exact
solution to the linear system for the specific time step chosen is 
the vector of all ones. For the experiments we use a uniform discretization
with mesh size $h=2^{-8}$ and we let $\Delta t = 10h$.
The resulting linear system has $n=66,049$ unknowns.

We use the factorized
sparse approximate inverse AINV \cite{BT1998} as a right preconditioner,
with drop tolerance $\tau = 0.05$ for both inverse factors.
With this choice,
the spectral radius of the iteration matrix $H=I-AP^{-1}$ is $\rho(H)\approx
0.9218$ and the spectral radius of $\hat{H}$ for the Adjoint Monte Carlo is
$\rho(\hat{H})\approx 0.9148$. The MAO transition probability is employed.
 Resorting to a uniform probability in this case would have impeded
the
convergence, since in this case $\rho(\hat{H})\approx 1.8401$.
This is an example demonstrating how the MAO probability can 
improve the behavior of the stochastic algorithm, outperforming the
uniform one.

The fill-in ratio is given by $\frac{nnz(H)}{nnz(A)}=4.26$,
therefore the relative number of nonzero elements in $H$ is still acceptable
in terms of storage and computational costs.

As before,
the threshold for the check on the relative residual is set to $\varepsilon
=10^{-8}$, while the threshold
for the adaptive selection of the random walks is set
to $\varepsilon_1=0.1$.
The results for all three methods are shown in Table \ref{parabolic_results}. As 
one can see, both Sequential Monte Carlo and MCSA 
dramatically reduce the number of iterations with
respect to the purely deterministic
 preconditioned Richardson iteration, with MCSA oupterforming
Sequential Monte Carlo. 
Of course each iteration is now more expensive due to the Monte Carlo
calculations required at each Richardson iteration, but we stress
that Monte Carlo is an embarrassingly parallel method.
Monte Carlo calculations are also expected to be more robust in the
presence of faults, which is one of the main motivations for the
present work.

\begin{table}[!t]
\centering
\hspace*{-0.8cm}
\begin{tabular}{|c|c|c|c|}
\hline
algorithm & relative err.& \# iterations & average \# histories per iteration\\
\hline
 Richardson & $6.277\cdot 10^{-7}$ & 178 & - \\
 \hline
 Sequential MC & $1.918 \cdot 10^{-9}$ & 8 & 51,535,375\\
\hline
 Adjoint MCSA & $1.341\cdot 10^{-9}$ & 6 & 16,244,000\\
\hline
\end{tabular}
\caption{Numerical results for the parabolic problem.}
\label{parabolic_results}
\end{table}

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{SEQ_parabolic.eps}
      \caption{Sequential Monte Carlo - Parabolic problem. Number of random
walks
employed at each
iteration.}
\label{SEQ_parabolic}
\end{figure}

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{MCSA_parabolic.eps}
      \caption{MCSA - Parabolic problem. Number of random walks
employed at each
iteration.}
\label{MCSA_parabolic}
\end{figure}

Finally, Figures \ref{SEQ_parabolic} and \ref{MCSA_parabolic} 
show the number of Monte Carlo histories per iteration 
for Sequential Monte Carlo
and for MCSA, respectively.

\section{Conclusions and future work}
\label{sec:conclusion}

In this paper we have reviewed known convergence conditions for Monte Carlo linear
solvers and established a few new sufficient conditions. In particular, we
have determined classes of matrices for which the method  is guaranteed
to converge. The main focus has
been on the recently proposed MCSA algorithm, which clearly outperforms
previous approaches. This method combines a deterministic fixed point
iteration (preconditioned Richardson method) with a Monte Carlo 
acceleration scheme, typically an Adjoint Monte Carlo estimator. Various
types of preconditioners have been tested, including diagonal, block
diagonal, and sparse approximate inverse preconditioning.

Generally speaking, it is difficult to ensure a priori that the
hybrid solver will converge. In particular, convergence of the underlying
preconditioned Richardson iteration is necessary, but not sufficient. 
The application of hybrid solvers to non-diagonally dominant, steady-state
problems presents a challenge and may require some trial-and-error in the
choice of tuning parameters, such as the block size or drop tolerances; 
convergence can be guaranteed for some standard model problems but in 
general it is difficult to enforce.
This is an inherent limitation of hybrid deterministic-stochastic
approaches of the kind considered in this paper.

On a positive note, numerical experiments show that these methods 
are quite promising for solving strictly diagonally dominant linear systems 
arising from time-dependent simulations, such as unsteady diffusion and
advection-diffusion type equations. Problems of this type are quite
important in practice, as they are often the most time-consuming part
of many large-scale CFD and radiation transport simulations. 
Linear systems with such properties also arise in other application areas,
such as network science and data mining.

In this paper we have not attempted to analyze the algorithmic scalability 
of the hybrid solvers. A difficulty is the fact that these methods contain
a number of tuneable parameters, each one of which can have great impact on 
performance and convergence behavior:
the choice of preconditioner, the stopping
criteria used for the Richardson iteration, the criteria for the number
and length of Monte Carlo histories to be run at each iteration, the 
particular estimator used, and
possibly others. While the cost per iteration is linear in the
number $n$ of unknowns, it is not clear how to predict the rate of convergence of
the outer iterations, since it depends strongly on the amount of work
done in the Monte Carlo acceleration phase, which is also not known a
priori except for some rather conservative upper bounds. Clearly,
the scaling behavior of hybrid methods with respect to problem size
needs to be further investigated.

Future work should also focus on testing hybrid methods on 
large parallel architectures and on evaluating their resiliency in the presence
of simulated faults. Efforts in this direction are currently under way.

\section*{Acknowledgments}
We are indebted to Miroslav T\accent23 uma for help with the AINV code used
in some of the numerical experiments.

\begin{thebibliography}{9}

  \bibitem{Agullo}
  {\sc E.~Agullo, L.~Giraud, A.~Guermouche, J.~Roman, and M.~Zounon},
  {\em Towards resilient parallel linear Krylov solvers: recover-restart
  strategies},
  Research Report N. 8324, June 2013, Project-Teams HiePACS, INRIA.

 \bibitem{AADBTW2005}
 {\sc V.~Alexandrov, E.~Atanassov, I.~T.~Dimov, S.~Branford, A.~Thandavan, and 
C.~Weihrauch},
{\em Parallel hybrid Monte Carlo algorithms for matrix computations 
problems},
Lecture Notes in Computer Science,
3516 (2005), pp.~752--759.

 \bibitem{Ax1994}
 {\sc O.~Axelsson},
 {\em Iterative Solution Methods},
 Cambridge University Press, 1994.

 \bibitem{Benzi2002}
 {\sc M.~Benzi}, 
 {\em Preconditioning techniques for large linear systems: a survey},
 Journal of Computational Physics,
 182 (2002), pp.~417--477.

 \bibitem{BMT1996}
 {\sc M.~Benzi, C.~D.~Meyer, and M.~Tuma},
 {\em A sparse approximate inverse preconditioner for the 
 conjugate gradient method},
  SIAM Journal on Scientific Computing,
  17 (1996), pp.~1135--1149.

 \bibitem{BT1998}
 {\sc M.~Benzi and M.~T\accent23uma},
 {\em A sparse approximate inverse preconditioner for nonsymmetric linear 
systems},
  SIAM Journal on Scientific Computing,
  19 (1998), pp.~968--994.

\bibitem{BP1979}
{\sc A.~Berman and R.~J.~Plemmons},
{\em Nonnegative Matrices in the Mathematical Sciences}, 
Academic Press, New York, 1979.

\bibitem{CFL}
{\sc R.~Courant, K.~Friedrichs, and H.~Lewy},
{\em \"Uber die partiellen Differenzengleichungen der mathematischen
Physik}, Mathematische Annalen, 100 (1928), pp.~32--74.

\bibitem{Curtiss}
{\sc J.~H.~Curtiss},
{\em Sampling methods applied to differential and difference equations},
Proc.~Seminar on Scientific Computation, Nov.~1949, IBM, New York, 1950, pp.~87--109.

 \bibitem{DA1998}
 {\sc I.~T.~Dimov and V.~N.~Alexandrov}, 
 {\em A new highly convergent Monte Carlo method for matrix computations},
 Mathematics and Computers in Simulation, 
 47 (1998), pp.~165--181.

 \bibitem{DVA2001}
 {\sc I.~T.~Dimov, V.~Alexandrov, and A.~Karaivanova},
 {\em Parallel resolvent Monte Carlo algorithms for linear algebra 
 problems},
 Mathematics and Computers in Simulation,
 55 (2001), pp.~25--35.

\bibitem{ifiss}
{\sc H.~C.~Elman, A.~Ramage, and D.~J.~Silvester}, 
{\em IFISS: A computational laboratory for investigating incompressible flow problems},
 SIAM Review, 56 (2014), pp.~261--273.

 \bibitem{EMSH2014}
 {\sc T.~M.~Evans, S.~W.~Mosher, S.~R.~Slattery, and S.~P.~Hamilton},
 {\em A Monte Carlo synthetic-acceleration method for solving the thermal 
 radiation diffusion equation},
 Journal of Computational Physics, 
 258 (2014), pp.~338--358.

 \bibitem{ESW2013}
 {\sc T.~M.~Evans, S.~R.~Slattery, and P.~P.~H.~Wilson}, 
 {\em A spectral analysis of the domain decomposed Monte Carlo method for 
linear systems},
 International Conference on Mathematics and Computational Methods 
Applied to Nuclear Science \& Engineering, 2013.

\bibitem{FLRU15}
 {\sc M.~Fasi, J.~Langou, Y.~Robert, and B.~U\c car},
 {\em A backward/forward recovery approach for the preconditioned
  conjugate gradient method}, arXiv:1511.04478v1, 13 November 2015.

\bibitem{FV62}
{\sc D.~F.~Feingold and R.~S.~Varga},
  {\em Block diagonally dominant matrices and generalizations
   of the Gerschgorin circle theorem},
Pacific Journal of Mathematics, 4 (1962), pp.~1241--1250.

\bibitem{FL50}
{\sc G.~E.~Forsythe and R.~A.~Leibler},
{\em Matrix inversion by a Monte Carlo method},
Math.~Tables Other Aids Comput., 6 (1952), pp.~78--81.

  \bibitem{FK1975}
  {\sc S.~Friedland and S.~Karlin},
  {\em Some inequalities for the spectral radius of non-negative matrices 
  and applications},
  Duke Mathematical Journal, 
  42 (1975), pp.~459--490.

 \bibitem{Hal1962}
 {\sc J.~H.~Halton},
 {\em Sequential Monte Carlo},
 Mathematical Proceeding of the Cambridge Philosophical Society,
 58 (1962), pp.~57--58.

 \bibitem{Hal1994}
 {\sc J.~H.~Halton}, 
 {\em Sequential Monte Carlo techniques for the solution of linear 
systems},
 Journal of Scientific Computing,
 9 (1994), pp.~213--257.

 \bibitem{MASC2013}
 {\sc J.~Hao, M.~Mascagni, and Y.~Li},
 {\em Convergence analysis of Markov chain Monte Carlo linear solvers using 
Ulam-von Neumann algorithm},
 SIAM Journal on Numerical Analysis,
 51 (2013), pp.~2107--2122.

  \bibitem{Heroux}
  {\sc M.~Hoemmen and M.~Heroux},
  {\em Fault-tolerant iterative methods via selective reliability},
  Proceedings of the 2011 International Conference for High Performance
Computing, Networking, Storage and Analysis (SC), vol.~3, IEEE Computer Society,
2011.

  \bibitem{Horn1994}
  {\sc R.~A.~Horn and C.~R.~Johnson},
  {\em Topics in Matrix Analysis},
  Cambridge University Press, 1994.

  \bibitem{Li2002}
  {\sc L.~Li},
  {\em On the iterative criterion for generalized diagonally dominant 
matrices},
  SIAM Journal on Matrix Analysis and Applications,
  24 (2002), pp.~17--24.

 \bibitem{Saad}
 {\sc Y.~Saad}, 
 {\em Iterative Methods for Sparse Linear Systems},
 Society for Industrial and Applied Mathematics, Philadelphia, 2003.

  \bibitem{Rizzi}
  {\sc K.~Sargsyan, F.~Rizzi, P.~Mycek, C.~Safta, K.~Morris, H.~Najm,
  O.~Le Maitre, O.~Knio, and B.~Debusschere},
  {\em Fault resilient domain decomposition preconditioner for PDEs},
  SIAM Journal on Scientific Computing,
  37 (2015), pp.~A2317--A2345.

 \bibitem{Slattery2013}
 {\sc S.~Slattery},
 {\em Parallel Monte Carlo Synthetic Acceleration Methods For Discrete 
Transport Problems},
 Ph.D.~Thesis, University of Wisconsin-Madison, 2013.

 \bibitem{Srin2010}
 {\sc A.~Srinivasan}, 
 {\em Monte Carlo linear solvers with non-diagonal splitting},
 Mathematics and Computer in Simulation,
 80 (2010), pp. 1133-1143.

 \bibitem{Stoyanov}
 {\sc M.~Stoyanov and C.~Webster},
 {\em Numerical analysis of fixed point algorithms in the presence of hardware
  faults},
  SIAM Journal on Scientific Computing,
  37 (2015), pp.~C532--C553.

\end{thebibliography}

\end{document}

### 2016-02-03 manuscript1/paper.tex
Added acknowledgement.

@@
 \title{Analysis of Monte Carlo accelerated iterative methods for
-sparse linear systems}
+sparse linear systems \tnoteref{ornl-cr}}
 \author{Michele Benzi\thanks{Department of Mathematics and Computer
@@
 }
+\tnotetext[ornl-cr]{Notice: This manuscript
+  has been authored by UT-Battelle, LLC, under contract DE-AC05-00OR22725
+  with the U.S. Department of Energy. The United States Government retains
+  and the publisher, by accepting the article for publication, acknowledges
+  that the United States Government retains a non-exclusive, paid-up,
+  irrevocable, world-wide license to publish or reproduce the published form
+  of this manuscript, or allow others to do so, for United States Government
+  purposes.}
 \date{}

### 2017-03-06 doc/nla/mcsa.tex
Added NLA article sources.

--- 2016-01-29 manuscript1/paper.tex
@@
-%
-% This draft dated January 29, 2016  [MB]
-%
-
-\documentclass[final,leqno,onefignum,onetabnum]{siamltex1213}
+\documentclass[times]{nlaauth}
+
+\usepackage{moreverb}
+
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 \usepackage{amssymb,amsmath,bm}
 \usepackage{color}
-\usepackage[lined,boxruled]{algorithm2e}
+\usepackage[ruled]{algorithm2e}
 \usepackage{comment}
@@
  \newtheorem{thm}{Theorem}
-% \newtheorem{lemma}{Lemma}
+ \newtheorem{lemma}{Lemma}
  \newtheorem{defn}{Definition}
@@
 
-\hoffset 0.6in
+%\hoffset 0.6in
 
 %opening
-\title{Analysis of Monte Carlo accelerated iterative methods for
+%\title{Analysis of Monte Carlo accelerated iterative methods for
+%sparse linear systems}
+%\author{Michele Benzi\thanks{Department of Mathematics and Computer
+%Science, Emory University, Atlanta, Georgia 30322, USA
+%(benzi@mathcs.emory.edu). The work of this author was supported
+%by Department of Energy (Office of Science) grant ERKJ247.}
+%\and
+%Thomas M.~Evans\thanks{Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
+%Oak Ridge, TN 37831, USA (evanstm@ornl.gov).}
+%\and Steven P. Hamilton\thanks{Oak Ridge National Laboratory,
+%1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (hamiltonsp@ornl.gov).}
+%\and
+%\linebreak Massimiliano Lupo  Pasini\thanks{Department of Mathematics and Computer
+%Science, Emory University, Atlanta, Georgia 30322, USA (mlupopa@emory.edu).}
+%\and
+%Stuart R.~Slattery\thanks{Oak Ridge National Laboratory,
+%1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (slatterysr@ornl.gov).}
+%}
+
+\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
+T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+\def\volumeyear{2017}
+\def\DOI{nla.2008}
+
+\begin{document}
+
+\runningheads{M.~Benzi et al.}{Monte Carlo solvers for sparse linear systems}
+
+\title{Analysis of Monte Carlo accelerated iterative methods\\ for
 sparse linear systems}
-\author{Michele Benzi\thanks{Department of Mathematics and Computer
-Science, Emory University, Atlanta, Georgia 30322, USA
-(benzi@mathcs.emory.edu). The work of this author was supported
-by Department of Energy (Office of Science) grant ERKJ247.}
-\and
-Thomas M.~Evans\thanks{Oak Ridge National Laboratory, 1 Bethel Valley Rd., 
-Oak Ridge, TN 37831, USA (evanstm@ornl.gov).}
-\and Steven P. Hamilton\thanks{Oak Ridge National Laboratory, 
-1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (hamiltonsp@ornl.gov).}
-\and 
-\linebreak Massimiliano Lupo  Pasini\thanks{Department of Mathematics and Computer
-Science, Emory University, Atlanta, Georgia 30322, USA (mlupopa@emory.edu).}
-\and 
-Stuart R.~Slattery\thanks{Oak Ridge National Laboratory,
-1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (slatterysr@ornl.gov).}
+
+\author{Michele Benzi\affil{1}, Thomas M.~Evans\affil{2}, Steven P. Hamilton\affil{3},
+\linebreak Massimiliano Lupo  Pasini\affil{4}, and Stuart R.~Slattery\affil{5}
 }
-\date{}
-
-\begin{document}
-\bibliographystyle{siam}
+
+\address{\affilnum{1}Department of Mathematics and Computer
+Science, Emory University, Atlanta, Georgia 30322, USA (benzi@mathcs.emory.edu).\break
+\affilnum{2}Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
+Oak Ridge, TN 37831, USA (evanstm@ornl.gov).\break
+\affilnum{3}Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
+Oak Ridge, TN 37831, USA (hamiltonsp@ornl.gov).\break
+\affilnum{4}Department of Mathematics and Computer
+Science, Emory University, Atlanta, Georgia 30322, USA (mlupopa@emory.edu).\break
+\affilnum{5}Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
+Oak Ridge, TN 37831, USA (slatterysr@ornl.gov).}
+
+\cgsn{\affilnum{1}This manuscript has been authored by UT-Battelle, LLC under
+  Contract No. DE-AC05-00OR22725 with the U.S. Department of Energy. The
+  United States Government retains and the publisher, by accepting the article
+  for publication, acknowledges that the United States Government retains a
+  non-exclusive, paid-up, irrevocable, world-wide license to publish or
+  reproduce the published form of this manuscript, or allow others to do so,
+  for United States Government purposes. The Department of Energy will provide
+  public access to these results of federally sponsored research in accordance
+  with the DOE Public Access Plan
+  (http://energy.gov/downloads/doe-public-access-plan).}{Dept. of Energy
+  (Office of Science) ERKJ247}
 
 \pagestyle{myheadings}
-\markboth{{\sc M.~Benzi, T.~M.~Evans, S.~P.~Hamilton, M.~Lupo Pasini, and S.~Slattery}}
-{Monte Carlo solvers for sparse linear systems}
-
-\maketitle
+%\markboth{{\sc M.~Benzi, T.~M.~Evans, S.~P.~Hamilton, M.~Lupo Pasini, and S.~R.~Slattery}}
+%{Monte Carlo solvers for sparse linear systems}
 
@@
 
-\begin{keywords}
-sparse linear systems, iterative methods, Richardson iteration, preconditioning,
-Monte Carlo methods, sparse approximate inverses, resilience
-\end{keywords}
-
-\begin{AMS}
-65C05, 65F50, 65F08, 65F10
-\end{AMS}
+\keywords{sparse linear systems; iterative methods; Richardson iteration; preconditioning;
+Monte Carlo methods; sparse approximate inverses; resilience}
+
+\maketitle
+
+%\begin{AMS}
+%65C05, 65F50, 65F08, 65F10
+%\end{AMS}
 
@@
 with established deterministic solvers such as preconditioned
-conjugate gradients and GMRES. So far, however, no theoretical analysis
+conjugate gradients and Generalized Minimum Residual (GMRES). So far, however, no theoretical analysis
 of the convergence properties of these solvers has been carried out. In
@@
 which is visited in the current step of the random walk.
+ The construction of random walks is accomplished considering the directed graph associated with the matrix $H$.
+The nodes of this graph are labeled 1 through $n$, and there is a directed edge from node $i$ to node $j$
+if and only if $H_{i,j}\ne 0$. Starting from a given node, the random walk consists of a sequence of nodes obtained by
+jumping from one node to the next along directed edges, choosing the next node at random according to a transition
+probability distribution matrix constructed from $H$ or from $H^T$, see below. Note that it may happen that a row of $H$
+(or of $H^T$) is all zero; this happens
+when there are no out-going (respectively, in-coming) edges from (respectively, to)
+the corresponding node. In this case, that particular random walk is terminated and another
+node is selected as the starting point for the next walk.
 The transition probabilities and the selection of
@@
 
-The general goal is to evaluate a functional as
+ Given a functional $J$, the goal is to compute its action on a vector $\mathbf{x}$ by
+ constructing a statistical estimator whose expected value equals $J(\mathbf{x})$. Each statistical sampling
+is represented by a random walk and it contributes to build an estimate of the expected value.
+Towards this goal, it is necessary to introduce an initial probability distribution
+and a transition matrix so that random walks are well defined.
+Recalling Riesz's representation theorem one can write
 \[
@@
 used afterwards in the stochastic process.
-A possible choice for the transition probability $P$ can be
+A possible choice for the transition probability matrix $P$ can be
 \[
@@
 \]
-By introducing $\nu$ as a generic permutation that represents the realization 
+In probability theory, a random walk is itself envisioned as a random variable that can assume multiple
+values consisting of different realizations. Indeed, given a starting point, there are in general many choices
+(one for each nonzero in the corresponding row of $P$) to select the second state and from there on recursively.
+The actual feasibility of a path and the frequency with which it is selected depend on the initial probability and on
+the transition matrix.
+By introducing $\nu$ as a realization
 of a random walk, we define
@@
 \]
-as the random variable associated with a specific permutation $\nu$. 
+as the random variable associated with a specific realization $\nu$.
 We can thus
@@
 where $\nu$ ranges over all possible realizations.
-$P_{\nu}$ is the probability associated with a specific permutation of the
+$P_{\nu}$ is the probability associated with a specific realization of the
 random walk.
@@
 The forward method presented above, however, has the limitation of employing an
-entire set of permutations to estimate just a single entry of
+entire set of realizations to estimate just a single entry of
 the solution at a time. Hence, in order to estimate the entire solution
 vector for Eq.~\eqref{linsys}, we have to employ a separate set of
-permutations for each entry in the solution vector. This limitation
+realizations for each entry in the solution vector. This limitation
 can be circumvented by the adjoint method which we describe below.
@@
 filter, indicating
-that only a subset of permutations contribute to the $j$th component
+that only a subset of realizations contribute to the $j$th component
 of the solution vector.
@@
 The main advantage of this method, compared to the forward one, consists in the
-fact that a single set of permutations is used to estimate the entire solution
+fact that a single set of realizations is used to estimate the entire solution
 vector.  Unless only a small portion of the problem is of interest, this
@@
  In other terms, the adjoint method should be preferred when approximating
-the solution globallly over the entire computational domain, while the forward
+the solution globally over the entire computational domain, while the forward
 method is especially useful when approximating the solution locally.
@@
 H_{k_{\ell},i}^T\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
-{H_{k_{\ell},i}^T}^T\bigg ] - \bigg 
+{H_{k_{\ell},i}^T}\bigg ] - \bigg
 (E\bigg[\sum_{\ell=0}^\infty
@@
   $\delta \mathbf{x}^{l+1}\approx(I-H)^{-1}\mathbf{r}^{l}$; \qquad
-\text \% {Computed via Standard MC}
+\text \% {Computed via Standard MC}\;
   $\mathbf{x}^{l+1}=\mathbf{x}^{l}+\delta \mathbf{x}^{l+1}$\;
@@
 between the amount of time and effort spent on the inner (stochastic) and
-outer (deterministic) iterations, which can take account the competing goal
+outer (deterministic) iterations, which can take into account the competing goals
 of reliability and rapid convergence.
@@
 \qquad
-\% \text{Computed via Standard MC}
+\% \text{Computed via Standard MC}\;
   $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta
@@
 is quite expensive, comparable to the cost of solving the original linear system.
-While ensuring that $\rho(H)<1$ (by means with appropriate preconditioning)
+While ensuring that $\rho(H)<1$ (by means of appropriate preconditioning)
 is in many cases possible, guaranteeing that $\rho(\hat{H})<1$ is generally
@@
 
-Note that this means that there exists a diagonal matrix  $\Delta$
-such that $A\Delta$ is SDD.
 A proper subclass of the class of GDD matrices is represented
@@
 
-Nevertheless, if $A$ is an $M$-matrix there exist efficient methods
+Nevertheless, if $A$ is an $M$-matrix, there exist efficient methods
 to determine a diagonal scaling of $A$ so that the resulting matrix
@@
 is GDD (in which case the diagonal scaling that
-makes $A$ GDD is produced), or not.
+makes $A$ SDD is produced), or not.
 The algorithm can be described as follows.
@@
 preconditioning does not alter the sparsity of the original coefficient matrix,
-while block diagonal preconditioning will incur a moderate amount of
+whereas block diagonal preconditioning will incur a moderate amount of
 fill-in in the preconditioned matrix if the blocks are not too large.
@@
 of the Monte Carlo linear solvers a priori.
+
+\subsection{Considerations about computational complexity} \label{sec:complex}
+
+Providing an analysis of the computational complexity for the aforementioned algorithms is not entirely
+straightforward because of their stochastic nature. Indeed, different statistical samplings can produce estimates with
+different uncertainty levels, requiring a proper tuning of the number of samplings computed to reach a prescribed accuracy.
+Moreover, we already mentioned that the asymptotic analysis of MC convergence assumes random walks with infinitely many steps
+and $N\rightarrow \infty$, where $N$ is the number of random walks. However, in practice each history must be truncated to a
+finite number of steps and the number of statistical samplings must be finite as well. The actual value of $N$ and the length
+of the histories affect the accuracy of the statistical estimation, thus influencing the number of iterations in a hybrid algorithm,
+since the uncertainty propagates to subsequent iterations.
+Therefore, here  we can only provide a tentative analysis of the complexity of the forward and adjoint Monte Carlo
+methods, assuming a specific history length and a fixed number $N$ of statistical samplings.
+
+Recalling formula \eqref{dir_mean} for the entry-wise estimate for the solution with the forward method, the cost of
+reconstructing the entire solution vector is
+\begin{equation}
+\text{Forward:}\quad \mathcal{O}(N\cdot k_{\ell}\cdot n),
+\label{for_cost}
+\end{equation}
+where $N$ is the number of histories, $k_{\ell}$ is the length of each random walk, and $n$ is the number of unknowns.
+As concerns the adjoint method, formula \eqref{adj_mean} leads to
+\begin{equation}
+\text{Adjoint:}\quad \mathcal{O}(N\cdot k_{\ell}).
+\label{adj_cost}
+\end{equation}
+The operation count for the hybrid schemes can thus be obtained by combining the cost of the Richardson scheme with the
+complexity of standard MC techniques.
+
+Regarding the Sequential Monte Carlo algorithm, the standard MC scheme is combined with the computation of the residual
+at each iteration. The cost of the residual computation is essentially that of a sparse matrix-vector product, which is
+$\mathcal{O}(n)$ for a sparse system. Therefore, the complexity of a single Sequential MC iteration is
+\begin{equation}
+\text{Sequential MC with forward MC update:}\quad \mathcal{O}((n + 1) \cdot N\cdot k_{\ell})
+\label{seqfor_cost}
+\end{equation}
+using the forward MC as an inner solver, and
+\begin{equation}
+\text{Sequential MC with adjoint MC update:}\quad \mathcal{O}(n + N\cdot k_{\ell})
+\label{seqadj_cost}
+\end{equation}
+using the adjoint MC as an inner solver.
+
+A single iteration of MCSA requires computing the residual, applying a matrix-vector product using $H$, and applying an MC update.
+The complexity of an MCSA iteration using the forward MC is therefore
+\begin{equation}
+\text{MCSA with forward MC update:}\quad \mathcal{O}((2n + 1) \cdot N\cdot k_{\ell}),
+\label{mcsafor_cost}
+\end{equation}
+whereas using the adjoint MC we find
+\begin{equation}
+\text{MCSA with adjoint MC update:}\quad \mathcal{O}(2n + N\cdot k_{\ell}).
+\label{mcsaadj_cost}
+\end{equation}
+Note that these estimates require $nnz(H)\approx nnz(A)$, in the sense that both $H$ and $A$ contain $\mathcal{O}(n)$ nonzeros.
+The actual values attained by $N$ and $k_{\ell}$ depend on the thresholds employed to truncate a single history and to determine
+the number of random walks to use. In general, the higher $N$ and $k_{\ell}$, the lower the number of outer iterations to achieve
+a prescribed accuracy.
+Finally, the total cost will depend also on the number of iterations, which is difficult to predict in practice.
 
@@
 \[
- \lVert \hat{H}\rVert_1\le (\rVert H\rVert_1)^2=1.
+ \lVert \hat{H}\rVert_1\le \rVert H\rVert_1^2=1.
 \]
@@
 see Figures \ref{SEQ_poisson} and
-\ref{MCSA_poisson} (note the semilogarithmic scale on the vertical axis). 
+\ref{MCSA_poisson}.
 It can be seen that
 in all the numerical tests presented in this section there is a sharp increase
+(from $O(10^3)$--$O(10^5)$ to $O(10^6)$--$O(10^7)$)
 in the number of histories when going from the first to subsequent iterations.
@@
   \centering
-    \includegraphics[width=\textwidth]{SEQ_poisson.eps}
+    \includegraphics[width=\textwidth]{MC_graphs/SEQ_poisson}
 \vspace{-0.3in}
     \caption{Sequential MC - Poisson problem. Number of random walks employed
-at each iteration.}
+at each iteration. The number for the first iteration is 2000.}
 \label{SEQ_poisson}
@@
   \centering
-    \includegraphics[width=\textwidth]{MCSA_poisson.eps}
+    \includegraphics[width=\textwidth]{MC_graphs/MCSA_poisson}
 \vspace{-0.3in}
       \caption{MCSA - Poisson problem. Number of random walks employed at each
-iteration.}
+iteration. The number for the first iteration is 2000.}
 \label{MCSA_poisson}
@@
   \centering
-    \includegraphics[width=\textwidth]{SEQ_diffreac.eps}
+    \includegraphics[width=\textwidth]{MC_graphs/SEQ_diffreac}
       \caption{Sequential MC - Reaction-diffusion problem. Number of random
@@
 employed at each
- iteration.}
+ iteration. The number first the first iteration is 36,000.}
 \label{SEQ_diffreac}
@@
   \centering
-    \includegraphics[width=\textwidth]{MCSA_diffreac.eps}
+    \includegraphics[width=\textwidth]{MC_graphs/MCSA_diffreac}
       \caption{MCSA - Reaction-diffusion problem. Number of random walks
 employed at each
-iteration.}
+iteration. The number first the first iteration is 36,000.}
 \label{MCSA_diffreac}
@@
   \centering
-    \includegraphics[width=\textwidth]{SEQ_parabolic.eps}
+    \includegraphics[width=\textwidth]{MC_graphs/SEQ_parabolic}
       \caption{Sequential Monte Carlo - Parabolic problem. Number of random
@@
 employed at each
-iteration.}
+iteration. The number for the first iteration is 505,000.}
 \label{SEQ_parabolic}
@@
   \centering
-    \includegraphics[width=\textwidth]{MCSA_parabolic.eps}
+    \includegraphics[width=\textwidth]{MC_graphs/MCSA_parabolic}
       \caption{MCSA - Parabolic problem. Number of random walks
 employed at each
-iteration.}
+iteration. The number for the first iteration is 505,000.}
 \label{MCSA_parabolic}
@@
 This is an inherent limitation of hybrid deterministic-stochastic
-approaches of the kind considered in this paper.
+approaches of the kind considered in this paper. It is of course possible in many cases to obtain convergence by
+combining the MCSA algorithm with a flexible Krylov subspace method like
+Flexible GMRES. However, this entails additional costs, decreased parallelism and
+increased storage requirements, as well as a possible reduction in the
+resilience of the algorithm due to the need for orthogonalization and the
+attendant additional communication needed.
 
@@
 important in practice, as they are often the most time-consuming part
-of many large-scale CFD and radiation transport simulations. 
+of many large-scale Computational Fluid Dynamics and radiation transport simulations.
 Linear systems with such properties also arise in other application areas,
@@
 
-\section*{Acknowledgments}
-We are indebted to Miroslav T\accent23 uma for help with the AINV code used
+%\section*{Acknowledgments}
+\acks
+ This manuscript
+  has been authored by UT-Battelle, LLC, under contract DE-AC05-00OR22725
+  with the U.S. Department of Energy. The United States Government retains
+  and the publisher, by accepting the article for publication, acknowledges
+  that the United States Government retains a non-exclusive, paid-up,
+  irrevocable, world-wide license to publish or reproduce the published form
+  of this manuscript, or allow others to do so, for United States Government
+  purposes.
+
+ We would also like to thank Miroslav T\accent23 uma for
+providing the AINV code used
 in some of the numerical experiments.
 
-\begin{thebibliography}{9}
+\begin{thebibliography}{99}
 
@@
  {\em Iterative Solution Methods},
- Cambridge University Press, 1994.
+ Cambridge University Press, Cambridge, UK, 1994.
 
@@
 
- \bibitem{ESW2013}
- {\sc T.~M.~Evans, S.~R.~Slattery, and P.~P.~H.~Wilson}, 
- {\em A spectral analysis of the domain decomposed Monte Carlo method for 
-linear systems},
- International Conference on Mathematics and Computational Methods 
-Applied to Nuclear Science \& Engineering, 2013.
-
 \bibitem{FLRU15}
@@
  {\em A backward/forward recovery approach for the preconditioned
-  conjugate gradient method}, arXiv:1511.04478v1, 13 November 2015.
+  conjugate gradient method}, Journal of Computational Science, 17 (2016), pp.~522--534
 
@@
   {\em Fault-tolerant iterative methods via selective reliability},
-  Proceedings of the 2011 International Conference for High Performance
-Computing, Networking, Storage and Analysis (SC), vol.~3, IEEE Computer Society,
+  Proceedings of the International Conference for High Performance
+Computing, Networking, Storage and Analysis (SC), Seattle, WA, vol.~3, IEEE Computer Society,
 2011.
-
-  \bibitem{Horn1994}
-  {\sc R.~A.~Horn and C.~R.~Johnson},
-  {\em Topics in Matrix Analysis},
-  Cambridge University Press, 1994.
 
@@
  \bibitem{Slattery2013}
- {\sc S.~Slattery},
+ {\sc S.~R.~Slattery},
  {\em Parallel Monte Carlo Synthetic Acceleration Methods For Discrete
 Transport Problems},
- Ph.D.~Thesis, University of Wisconsin-Madison, 2013.
+ Ph.D.~Thesis, University of Wisconsin-Madison, 2013,
+http://search.proquest.com/ebrary/docview/1449206072.
+
+ \bibitem{ESW2013}
+ {\sc S.~R.~Slattery, T.~M.~Evans, and P.~P.~H.~Wilson},
+ {\em A spectral analysis of the domain decomposed Monte Carlo method for
+linear systems},
+ International Conference on Mathematics and Computational Methods
+Applied to Nuclear Science \& Engineering, Sun Valley Resort, ID, 2013.
 
